# Preface {.unnumbered}

```{r}
#| label: setup-preface
#| include: false
source("R/_setup.R")
```

This book is the mathematical companion to the `statmodels7` stack. Its purpose
is to collect in one place, with complete derivations, the mathematics that the
packages implement: the link functions with their derivatives to fourth order,
the probability distributions with their scores and information matrices, and
the operations that build new distributions out of existing ones — changes of
variables, zero-inflation, zero-adjustment, truncation.

The style of the book follows a single rule: **nothing is cited away**. Whenever
a formula appears, the derivation that produces it from the definitions appears
with it, with the intermediate steps written out. A reader who wants to know why
the score of a truncated distribution is the parent's score recentred at its
truncated mean, or why the expected information on the link scale is a simple
congruence transform while the observed one is not, will find the argument on
the page rather than a reference to it. The mathematics is elementary throughout
— calculus, a little probability, the occasional exchange of derivative and
integral — and the aim has been to make every step easy to follow, so that the
book can be read as a course in the likelihood theory of parametric families as
much as a description of one particular software stack.

The book is generated with Quarto directly from the source code of the packages,
which are loaded from the working tree each time it is rendered. The numerical
examples scattered through the chapters are therefore executed by the very code
they illustrate, and the output shown is the output produced at build time.

## How the book is organised {.unnumbered}

@sec-introduction explains what the framework is for and why it is built the way
it is. It contains no mathematics and can be read on its own.

@sec-links is devoted to link functions: what they are required to satisfy, how
the derivatives of a link and of its inverse determine one another, and why the
stack carries them to fourth order. It ends with a catalogue of the sixteen link
classes, each with its formulas written out in full.

@sec-distributions develops the likelihood theory the stack rests on — score,
observed and expected information, the Bartlett identities, derivatives on the
unconstrained scale via Faà di Bruno's formula — and then presents the fourteen
distributions, with particular care for the places where the parametrisation
differs from the one base R uses.

@sec-transformations treats the operations that turn one distribution into
another. This is the heart of the book: the wrappers compose with any parent, so
their formulas are derived once, in general, and the chapter proves every one of
them — the invariance of likelihood derivatives under a change of variables, the
score and information of the zero-inflated and hurdle models, the mixed
distributions that arise from zero-adjusting a continuous family, and the
derivatives of a truncated log-likelihood.

@sec-notation collects the notation used throughout.

## Building the book {.unnumbered}

From the `book/` directory of the stack:

``` bash
quarto render
```

```{r}
#| label: provenance
#| echo: false
prov <- data.frame(
  item = c("linkfunctions7", "distributions7", "R", "built"),
  value = c(
    as.character(utils::packageVersion("linkfunctions7")),
    as.character(utils::packageVersion("distributions7")),
    paste(R.version$major, R.version$minor, sep = "."),
    format(Sys.time(), "%Y-%m-%d")
  )
)
knitr::kable(prov, col.names = c("", ""), caption = "This build")
```
