# Preface {.unnumbered}

```{r}
#| label: setup-preface
#| include: false
source("R/_setup.R")
```

This book is the mathematical companion to the `statmodels7` toolkit. Its purpose
is to collect in one place the mathematics that the packages implement: the link
functions with their derivatives to fourth order, the probability distributions
with their scores and information matrices, the operations that build new
distributions out of existing ones, and the optimisation methods that fit them.

It is written to be taught from as much as to be consulted, and that has settled
two questions of style. The first is what to show. Every formula the toolkit
actually evaluates appears on the page, in the form the code uses, together with
enough of the reasoning to make it intelligible: where a quantity comes from,
what it means, and which condition it depends on. What does not appear is the
formal proof. A reader who wants the argument in full is pointed to the paper or
the textbook where it is given properly, and the appendix collects those
pointers; the space this saves is spent instead on explaining the results and on
the places where the implementation had to make a choice. The second question is
how to say it. The mathematics here is elementary throughout — calculus, a
little probability, the occasional exchange of derivative and integral — and the
prose aims to be continuous and unhurried rather than clipped, so that the
chapters can be read straight through.

The mathematics is not new, and no attempt is made to pretend otherwise. What is
particular to this book is that the formulas shown are the formulas the packages
run, which is a claim about software rather than about mathematics, and it is
worth as much as the care taken to keep it true.

The book is generated with Quarto directly from the source code of the packages,
which are loaded from the working tree each time it is rendered. The numerical
examples scattered through the chapters are therefore executed by the very code
they illustrate, and the output shown is the output produced at build time.

## How the book is organised {.unnumbered}

After the introduction there is **one chapter per package**, so that the book
grows the way the toolkit does.

@sec-introduction explains what the toolkit is for and why it is built the way
it is. It contains no mathematics and can be read on its own.

@sec-links is `linkfunctions7`: what a link is required to satisfy, how the
derivatives of a link and of its inverse determine one another, and why they are
carried to fourth order. It ends with a catalogue of the sixteen link classes,
each with its formulas written out in full.

@sec-distributions7 is `distributions7`, and divides into four.
@sec-distributions develops the likelihood theory the toolkit rests on — score,
observed and expected information, the Bartlett identities, derivatives on the
unconstrained scale via Faà di Bruno's formula, derivatives of the distribution
function — and presents the fourteen families, with particular care where the
parametrisation differs from the one base R uses. @sec-transformations treats the
operations that turn one distribution into another, and is the most demanding
part of the book, since the wrappers compose with any parent and their formulas
are therefore stated once, in general, at every order the toolkit supports.
@sec-fitting fits a distribution to a sample by maximum likelihood, which is the
smallest problem that exercises all of the preceding machinery at once.
@sec-fallbacks describes what a distribution receives when it implements nothing
but its density.

@sec-optimizers7 is `optimizers7`, and divides into four as well.
@sec-descent asks what a step is and what a line search along it must guarantee;
@sec-curvature takes the methods in order of how much curvature they store, from
the exact Hessian down to a single number; @sec-nonsmooth deals with objectives
that have a kink in them, where the gradient does not become small at the
solution and a method built on the assumption that it does arrives at the answer
and reports failure; and @sec-constraints covers box constraints, removed by a
change of variable, and Adam, whose steps are not descent steps at all.

@sec-notation collects the notation used throughout, and the references follow
it.

## Building the book {.unnumbered}

From the `book/` directory of the toolkit:

``` bash
quarto render
```

```{r}
#| label: provenance
#| echo: false
prov <- data.frame(
  item = c("linkfunctions7", "distributions7", "optimizers7", "R", "built"),
  value = c(
    as.character(utils::packageVersion("linkfunctions7")),
    as.character(utils::packageVersion("distributions7")),
    as.character(utils::packageVersion("optimizers7")),
    paste(R.version$major, R.version$minor, sep = "."),
    format(Sys.time(), "%Y-%m-%d")
  )
)
knitr::kable(prov, col.names = c("", ""), caption = "This build")
```
