# Preface {.unnumbered}

```{r}
#| label: setup-preface
#| include: false
source("R/_setup.R")
```

This book is the mathematical companion to the `statmodels7` toolkit. Its purpose
is to collect in one place, with complete derivations, the mathematics that the
packages implement: the link functions with their derivatives to fourth order,
the probability distributions with their scores and information matrices, and
the operations that build new distributions out of existing ones — changes of
variables, zero-inflation, zero-adjustment, truncation.

The style of the book follows a single rule: **nothing is cited away**. Whenever
a formula appears, the derivation that produces it from the definitions appears
with it, with the intermediate steps written out. A reader who wants to know why
the score of a truncated distribution is the parent's score recentred at its
truncated mean, or why the expected information on the link scale is a simple
congruence transform while the observed one is not, will find the argument on
the page rather than a reference to it. The mathematics is elementary throughout
— calculus, a little probability, the occasional exchange of derivative and
integral — and the aim has been to make every step easy to follow, so that the
book can be read as a course in the likelihood theory of parametric families as
much as a description of one particular software toolkit.

The book is generated with Quarto directly from the source code of the packages,
which are loaded from the working tree each time it is rendered. The numerical
examples scattered through the chapters are therefore executed by the very code
they illustrate, and the output shown is the output produced at build time.

## How the book is organised {.unnumbered}

After the introduction there is **one chapter per package**, so that the book
grows the way the toolkit does.

@sec-introduction explains what the toolkit is for and why it is built the way
it is. It contains no mathematics and can be read on its own.

@sec-links is `linkfunctions7`: what a link is required to satisfy, how the
derivatives of a link and of its inverse determine one another, and why they are
carried to fourth order. It ends with a catalogue of the sixteen link classes,
each with its formulas written out in full.

@sec-distributions7 is `distributions7`, and divides into four.
@sec-distributions develops the likelihood theory the toolkit rests on — score,
observed and expected information, the Bartlett identities, derivatives on the
unconstrained scale via Faà di Bruno's formula, derivatives of the distribution
function — and presents the fourteen families, with particular care where the
parametrisation differs from the one base R uses. @sec-transformations treats the
operations that turn one distribution into another, and is the most demanding
part of the book: the wrappers compose with any parent, so their formulas are
derived once, in general, and every one of them is proved. @sec-fitting fits a
distribution to a sample by maximum likelihood, which is the smallest problem
that exercises all of the preceding machinery at once. @sec-fallbacks describes
what a distribution receives when it implements nothing but its density.

@sec-notation collects the notation used throughout.

## Building the book {.unnumbered}

From the `book/` directory of the toolkit:

``` bash
quarto render
```

```{r}
#| label: provenance
#| echo: false
prov <- data.frame(
  item = c("linkfunctions7", "distributions7", "R", "built"),
  value = c(
    as.character(utils::packageVersion("linkfunctions7")),
    as.character(utils::packageVersion("distributions7")),
    paste(R.version$major, R.version$minor, sep = "."),
    format(Sys.time(), "%Y-%m-%d")
  )
)
knitr::kable(prov, col.names = c("", ""), caption = "This build")
```
