# Preface {.unnumbered}

```{r}
#| label: setup-preface
#| include: false
source("R/_setup.R")
```

This book is the mathematical companion to the `statmodels7` toolkit. Its purpose is to collect in one place the mathematics that the packages
implement. That covers the link functions with their derivatives to fourth
order, the probability distributions with their scores and information
matrices, the operations that build new distributions out of existing ones,
and the optimization methods that fit them.

It is written to be taught from as much as to be consulted, and that has settled
two questions of style. The first is what to show. Every formula the toolkit
actually evaluates appears on the page, in the form the code uses, together with
enough of the reasoning to make it intelligible: where a quantity comes from,
what it means, and which condition it depends on. What does not appear is the
formal proof. A reader who wants the argument in full is pointed to the paper or the
textbook where it is given properly, and the appendix collects those
references. The space this saves is spent on explaining the results and the
places where the implementation had to make a choice. The second question is
how to say it. The mathematics here is elementary throughout — calculus, a
little probability, the occasional exchange of derivative and integral — and the
prose aims to be continuous and unhurried rather than clipped, so that the
chapters can be read straight through.

The mathematics is not new, and no attempt is made to pretend otherwise. What is
particular to this book is that the formulas shown are the formulas the packages
run, which is a claim about software rather than
about mathematics, and whose value rests on the care taken to keep it true.

The book is generated with Quarto directly from the source code of the packages,
which are loaded from the working tree each time it is rendered. The numerical
examples scattered through the chapters are therefore executed by the very code
they illustrate, and the output shown is the output produced at build time.

## How the book is organized {.unnumbered}

After the introduction there is **one chapter per package**, so that the book
grows the way the toolkit does.

@sec-introduction explains what the toolkit is for and why it is built the way
it is. It contains no mathematics and can be read on its own.

@sec-links is `linkfunctions7`: what a link is required to satisfy, how the
derivatives of a link and of its inverse determine one another, and why they are
carried to fourth order. It ends with a catalog of the sixteen link classes,
each with its formulas written out in full.

@sec-distributions7 is `distributions7`, and divides into five.
@sec-distributions develops the likelihood theory the toolkit rests on —
score, observed and expected information, the Bartlett identities,
derivatives on the unconstrained scale via Faà di Bruno's formula,
derivatives of the distribution function — and presents the univariate
families, thirty-eight of them, under the numbering convention that gives a
family with several parametrizations one name per parametrization.
Particular care is taken where the parametrization differs from the
one used by base R. @sec-transformations treats the
operations that turn one distribution into another, and is the most demanding
part of the book, since the wrappers compose with any parent and their formulas
are therefore stated once, in general, at every order the toolkit supports.
@sec-fitting fits a distribution to a sample by maximum likelihood, which is the
smallest problem that exercises all of the preceding machinery at once.
@sec-fallbacks describes what a distribution receives when it implements nothing
but its density. @sec-multivariate treats families whose observations are
vectors, where the likelihood theory is unchanged and only the parametrization
of the matrix has to be arranged.

@sec-optimizers7 is `optimizers7`, and divides into four as well.
@sec-descent defines descent directions and the conditions a line search
must guarantee. @sec-curvature takes the methods in order of how much
curvature they store, from the exact Hessian down to a single number.
@sec-nonsmooth deals with objectives that have a kink, where the gradient
does not become small at the solution and a gradient-based method reaches the
answer and reports failure. @sec-constraints covers box constraints, removed
by a change of variable, and Adam, whose steps are not descent steps. Finally
@sec-starting covers the choice of the starting point, which reuses the same
change of variable, and the question of how many runs to make.

@sec-basis7 is `basis7`, and divides into five. @sec-basis-expansions defines
a basis expansion and the four quantities the package computes from it.
@sec-basis-families presents B-splines, the Fourier basis and the Legendre
polynomials, each generated by a recurrence that also produces its derivatives
and its integral. @sec-basis-gram is about the matrix of inner products, which
a roughness penalty is a quadratic form in and which is computed exactly rather
than approximated on a grid. @sec-basis-transforms derives the three linear
transformations of a basis — orthonormalization, a linear constraint, and the
Demmler-Reinsch construction — from the property each is asked to produce.
@sec-basis-tensor builds a basis of several variables from marginal ones.

@sec-parameters7 is `parameters7`, and divides into three.
@sec-parameter-contract sets out what a constrained parameter is and why it is
not a link. @sec-parameter-families presents the families: the two charts on
the positive definite cone, the diagonal families that reuse the link
functions of @sec-links, the fixed matrix carried by a single scale, the
correlation matrix and the structured covariances, and the sets that are not
matrices at all. @sec-parameter-rank treats matrices that are not of full
rank, which is what a penalty is, and the pseudo-determinant and rank
determination they require.

@sec-numericals7 is `numericals7`, the numerical layer at the bottom of the
toolkit, and divides into four. @sec-numerical-jets presents jets, the
truncated Taylor algebra by which a map written in plain R differentiates
itself to fourth order. @sec-numerical-stencils presents the single
finite-difference stencil library. @sec-numerical-quadrature presents
quadrature and series summation vectorized over the parameters, one matrix
evaluation for many parameter values. @sec-numerical-special presents the
special functions the distributions are written in, each with the overflow
discipline that was learned on it.

@sec-notation collects the notation used throughout, and the references follow
it.

## Building the book {.unnumbered}

From the `book/` directory of the toolkit:

``` bash
quarto render
```

```{r}
#| label: provenance
#| echo: false
prov <- data.frame(
  item = c("numericals7", "linkfunctions7", "distributions7", "optimizers7",
           "basis7", "parameters7", "R", "built"),
  value = c(
    as.character(utils::packageVersion("numericals7")),
    as.character(utils::packageVersion("linkfunctions7")),
    as.character(utils::packageVersion("distributions7")),
    as.character(utils::packageVersion("optimizers7")),
    as.character(utils::packageVersion("basis7")),
    as.character(utils::packageVersion("parameters7")),
    paste(R.version$major, R.version$minor, sep = "."),
    format(Sys.time(), "%Y-%m-%d")
  )
)
knitr::kable(prov, col.names = c("", ""), caption = "This build")
```
