# Introduction {#sec-introduction}

```{r}
#| label: setup-01
#| include: false
source("R/_setup.R")
```

## One Gamma, written many times {#sec-rewritten}

Open almost any R modelling package and somewhere in its internals you will find
a Gamma distribution — not *the* Gamma distribution, but that package's own
private copy of it, written as a handful of helper functions: a `switch` on a
character string that returns a density, a closure that computes a score, a
block inside a fitting routine that produces whatever second derivative the
routine happens to need. The same is true of link functions. Every package that
fits something positive carries its own log link, and every package that fits a
probability carries its own logit.

None of these objects can be reused, because none of them is really an object.
They are code paths, private to the package that owns them, and they contain
only what their author's routine happened to call for — usually the density and
the score, sometimes the Hessian, almost never anything beyond. A package that
needs a third derivative writes its own Gamma from scratch; a package that wants
derivatives with respect to an unconstrained parameter does the same. The result
is that the R ecosystem contains dozens of Gamma distributions, each partial,
each written under deadline, and each capable of being wrong in its own way.

The observation the stack is built on is that all of this is unnecessary,
because a Gamma distribution is a fixed mathematical object. It does not vary
from package to package. Its density, its score, its information, its third and
fourth derivatives, its derivatives with respect to the response, its behaviour
under any reparametrisation — all of these are determined once and for all by
the definition, and they can therefore be written down once, correctly and
completely, for everyone to build on.

That is what `statmodels7` is: infrastructure meant to be reused, rather than
another modelling package. The measure of its success is not how well it fits
models but whether somebody else's modelling package can be built on top of it.

## What "completely" means {#sec-completely}

A distribution in this stack supplies, or can derive, everything a modelling
routine might reasonably ask of it: the density or mass function together with
the cdf, the quantile function and a random generator; the score and the
observed information; the expected information; third and fourth derivatives of
the log-likelihood, both observed and expected; derivatives with respect to the
response as well as the parameters; and every one of these quantities with
respect to the unconstrained parameters as well as the natural ones.

The last two items are the ones usually missing from other implementations, and
they are precisely the ones a modelling framework depends on. Derivatives in the
response are what quantile-type residuals and smoothness penalties are made of.
Derivatives on the unconstrained scale are what an optimiser wants: a Gamma
variance must stay positive, and an algorithm forced to respect that constraint
is doing constrained optimisation, whereas an algorithm working with
$\eta = \log \sigma^2$ moves freely over the whole real line. Third and fourth
derivatives, finally, are the raw material of higher-order asymptotics and bias
correction, and their presence is the reason the change of scale developed in
@sec-link-scale has to be carried to fourth order rather than stopping, as the
chain rule usually does in practice, at the second.

## The stack {#sec-stack}

The name encodes a convention: every package is a plural noun followed by `7`,
the `7` being the S7 object system everything is built on.

| Package | What it provides |
|---|---|
| `linkfunctions7` | 16 link classes with exact analytical derivatives to fourth order, in both directions |
| `distributions7` | 14 distributions with exact score, information and third/fourth derivatives, plus wrappers, transformations and maximum likelihood |
| `modelterms7` | *planned* — model terms usable directly in formulas |
| `basis7` | *planned* — spline and other bases |
| `penalties7` | *planned* — penalties |
| `statmodels7` | *planned* — the modelling framework itself |

: The stack {.striped}

The destination is the last row: a framework in the spirit of GAMLSS in which
every parameter of a distribution can be modelled rather than just the mean, the
optimiser is pluggable, estimation can be frequentist or Bayesian, and users can
define new model terms and drop them straight into formulas. Everything below it
exists to make that possible, and one design pressure has shaped it all: *exact
derivatives, on the unconstrained scale, for components the framework has never
seen before*.

## Only the density is required {#sec-extension}

A stack that worked only for the fourteen distributions it ships with would have
solved nothing; it would simply be one more closed catalogue, larger than the
others. What matters is what happens when a user brings a distribution the
authors never considered, and here the contract is deliberately minimal: **a new
distribution needs nothing but its density**. Everything else has a fallback —
the cdf is computed by quadrature, the quantile function by root-finding, the
random generator by a ratio-of-uniforms scheme that needs only the density, and
every derivative by finite differences. Whatever the user later supplies in
closed form simply replaces the corresponding fallback, with no registration
step and no separate interface for the "advanced" case.

To see what this means in practice, here is a complete definition of the
Rayleigh distribution, which has density
$f(y;\sigma) = (y/\sigma^2)\, e^{-y^2/2\sigma^2}$ on $(0, \infty)$:

```{r}
#| label: minimal-distribution
Rayleigh <- S7::new_class("Rayleigh", parent = continuous_distrib, package = NULL)

S7::method(distrib_pdf, Rayleigh) <- function(distrib, y, theta, log = FALSE) {
  s <- theta[[1]]
  ld <- log(y) - 2 * log(s) - y^2 / (2 * s^2)
  if (log) ld else exp(ld)
}

rayleigh <- Rayleigh(
  distrib_name = "rayleigh", dimension = "univariate", bounds = c(0, Inf),
  params = "sigma", params_interpretation = c(sigma = "scale"), n_params = 1,
  params_bounds = list(sigma = c(0, Inf)),
  link_params = list(sigma = log_link())
)
```

Nothing above mentions a cdf, a quantile function, a generator or a derivative,
and yet all of them are immediately available. The median of a Rayleigh is
$\sigma\sqrt{2\log 2}$, its score is $y^2/\sigma^3 - 2/\sigma$, and its expected
information is $4/\sigma^2$; the fallbacks recover each of these from the
density alone:

```{r}
#| label: minimal-distribution-use
theta <- list(sigma = 2)

distrib_quantile(rayleigh, 0.5, theta)     # by root-finding on the numerical cdf
2 * sqrt(2 * log(2))                       # the closed form

distrib_gradient(rayleigh, c(1, 3), theta) # by finite differences
c(1, 3)^2 / 2^3 - 2 / 2                    # the closed form

distrib_expected_hessian(rayleigh, 1, theta)$sigma_sigma   # -4/sigma^2
```

The twenty lines above are a first-class citizen of the stack: the object can be
zero-inflated, truncated, transformed, and fitted by Fisher scoring exactly like
the distributions the package ships with.

## Checking one's own work {#sec-validators}

Because users are expected to write their own distributions and links, the
packages ship the tools to validate them. `check_link()` verifies that a link is
invertible in both directions, strictly monotone, consistent with the inverse
function theorem, and that each analytical derivative agrees with a numerical
one; `check_distrib()` runs the analogous battery on a distribution — the
density integrates to one, the cdf and the quantile function invert each other
and agree with the density, the random generator matches the cdf, every
derivative matches a finite-difference computation, and the expected information
matches a Monte Carlo estimate of the variance of the score.

```{r}
#| label: check-rayleigh
res <- check_distrib(rayleigh, theta = list(sigma = 2), orders = 1:2,
                     nsim = 5e4, verbose = FALSE)
res[, c("check", "status")]
```

These validators run on any object, including one written five minutes ago, and
they are used systematically on everything the packages contain. The chapters
that follow present the mathematics behind all of it.
