# Introduction {#sec-introduction}

```{r}
#| label: setup-01
#| include: false
source("R/_setup.R")
```

## One Gamma, written many times {#sec-rewritten}

Open almost any R modelling package and somewhere in its internals you will find
a Gamma distribution — not *the* Gamma distribution, but that package's own
private copy of it, written as a handful of helper functions: a `switch` on a
character string that returns a density, a closure that computes a score, a
block inside a fitting routine that produces whatever second derivative the
routine happens to need. The same is true of link functions. Every package that
fits something positive carries its own log link, and every package that fits a
probability carries its own logit.

None of these objects can be reused, because none of them is really an object.
They are code paths, private to the package that owns them, and they contain
only what their author's routine happened to call for — usually the density and
the score, sometimes the Hessian, almost never anything beyond. A package that
needs a third derivative writes its own Gamma from scratch; a package that wants
derivatives with respect to an unconstrained parameter does the same. The result
is that the R ecosystem contains dozens of Gamma distributions, each partial,
each written under deadline, and each capable of being wrong in its own way.

The observation the toolkit is built on is that all of this is unnecessary,
because a Gamma distribution is a fixed mathematical object. It does not vary
from package to package. Its density, its score, its information, its third and
fourth derivatives, its derivatives with respect to the response, its behaviour
under any reparametrisation — all of these are determined once and for all by
the definition, and they can therefore be written down once, correctly and
completely, for everyone to build on.

That is what `statmodels7` is: infrastructure meant to be reused, rather than
another modelling package. The measure of its success is not how well it fits
models but whether somebody else's modelling package can be built on top of it.

## What "completely" means {#sec-completely}

The principle applies to both kinds of object the toolkit currently provides,
and it is worth stating for each, because what counts as complete differs.

A **link function** is a map between a constrained parameter and an
unconstrained one, and what a modelling routine can ask of it is its value and
its derivatives — in both directions. Almost every implementation in R carries
the first derivative of the inverse link and stops there, which is exactly what a
GLM fitted by iterative least squares needs and nothing more. A link in this
toolkit carries four derivatives of the forward map and four of the inverse, so
that it can serve a routine working to any order up to the fourth, and it knows
its own domain so that the routine does not have to.

A **distribution** supplies, or can derive, the density or mass function
together with the cdf, the quantile function and a random generator; the score
and the observed information; the expected information; third and fourth
derivatives of the log-likelihood, both observed and expected; derivatives with
respect to the response as well as the parameters; derivatives of the
distribution function, which is what a censored observation contributes; and
every one of these with respect to the unconstrained parameters as well as the
natural ones.

The items at the end of each list are the ones usually missing elsewhere, and
they are precisely the ones a modelling framework depends on. Derivatives in the
response are what quantile-type residuals and smoothness penalties are made of.
Derivatives on the unconstrained scale are what an optimiser wants: a Gamma
variance must stay positive, and an algorithm forced to respect that constraint
is doing constrained optimisation, whereas an algorithm working with
$\eta = \log \sigma^2$ moves freely over the whole real line. Third and fourth
derivatives, finally, are the raw material of higher-order asymptotics and bias
correction.

This is also where the two packages meet, and why neither would be much use
alone. Transporting a fourth derivative of a log-likelihood from the natural
scale to the unconstrained one is an application of Faà di Bruno's formula, and
it consumes exactly the four derivatives of the inverse link that
`linkfunctions7` carries — @sec-link-scale works this out in detail. A link
stopping at second order would silently cap the whole toolkit at second-order
methods, however much the distributions knew about themselves.

## The toolkit {#sec-toolkit}

The name encodes a convention: every package is a plural noun followed by `7`,
the `7` being the S7 object system everything is built on.

| Package | What it provides |
|---|---|
| `linkfunctions7` | 16 link classes with exact analytical derivatives to fourth order, in both directions |
| `distributions7` | 14 distributions with exact score, information and third/fourth derivatives, plus wrappers, transformations and maximum likelihood |
| `optimizers7` | 9 optimisation algorithms as objects, with composable stopping rules, safeguards that report themselves, and box constraints removed by reparameterisation |
| `modelterms7` | *planned* — model terms usable directly in formulas |
| `basis7` | *planned* — spline and other bases |
| `penalties7` | *planned* — penalties |
| `statmodels7` | *planned* — the modelling framework itself |

: The toolkit {.striped}

The destination is the last row: a framework in the spirit of GAMLSS in which
every parameter of a distribution can be modelled rather than just the mean, the
optimiser is pluggable, estimation can be frequentist or Bayesian, and users can
define new model terms and drop them straight into formulas. Everything below it
exists to make that possible, and one design pressure has shaped it all: *exact
derivatives, on the unconstrained scale, for components the framework has never
seen before*.

## Extension is the point {#sec-extension}

A toolkit that worked only for the sixteen links and fourteen distributions it
ships with would have solved nothing; it would simply be one more closed
catalogue, larger than the others. What matters is what happens when a user
brings an object the authors never considered, and in both packages the answer is
that it is treated exactly like the ones that came with them — there is no
registration step, and no separate interface for the "advanced" case.

For a link the requirement is only that the ten generics be implemented, and
since a link is a formula and its derivatives, that is a dozen lines. Nothing
distinguishes the result from a built-in one: it can be handed to any
distribution as the link for any parameter, and the fourth-order machinery will
use it without knowing where it came from.

```{r}
#| label: minimal-link
NegLog <- S7::new_class("NegLog", parent = link, package = NULL)

S7::method(linkfun,   NegLog) <- function(x, theta) -log(theta)
S7::method(linkinv,   NegLog) <- function(x, eta)   exp(-eta)
S7::method(dlinkfun,  NegLog) <- function(x, theta) -1 / theta
S7::method(d2linkfun, NegLog) <- function(x, theta)  1 / theta^2
S7::method(dlinkinv,  NegLog) <- function(x, eta)   -exp(-eta)
S7::method(d2linkinv, NegLog) <- function(x, eta)    exp(-eta)

neglog <- NegLog(link_name = "neglog", link_bounds = c(0, Inf), link_params = NULL)
linkinv(neglog, linkfun(neglog, c(0.5, 2)))
```

For a distribution the contract is even smaller, and it is the one that decides
whether the package is infrastructure or a catalogue: **a new distribution needs
nothing but its density**. Everything else has a fallback —
the cdf is computed by quadrature, the quantile function by root-finding, the
random generator by a ratio-of-uniforms scheme that needs only the density, and
every derivative by finite differences. Whatever the user later supplies in
closed form simply replaces the corresponding fallback, with no registration
step and no separate interface for the "advanced" case.

To see what this means in practice, here is a complete definition of the
Rayleigh distribution, which has density
$f(y;\sigma) = (y/\sigma^2)\, e^{-y^2/2\sigma^2}$ on $(0, \infty)$:

```{r}
#| label: minimal-distribution
Rayleigh <- S7::new_class("Rayleigh", parent = continuous_distrib, package = NULL)

S7::method(distrib_pdf, Rayleigh) <- function(distrib, y, theta, log = FALSE) {
  s <- theta[[1]]
  ld <- log(y) - 2 * log(s) - y^2 / (2 * s^2)
  if (log) ld else exp(ld)
}

rayleigh <- Rayleigh(
  distrib_name = "rayleigh", dimension = "univariate", bounds = c(0, Inf),
  params = "sigma", params_interpretation = c(sigma = "scale"), n_params = 1,
  params_bounds = list(sigma = c(0, Inf)),
  link_params = list(sigma = log_link())
)
```

Nothing above mentions a cdf, a quantile function, a generator or a derivative,
and yet all of them are immediately available. The median of a Rayleigh is
$\sigma\sqrt{2\log 2}$, its score is $y^2/\sigma^3 - 2/\sigma$, and its expected
information is $4/\sigma^2$; the fallbacks recover each of these from the
density alone:

```{r}
#| label: minimal-distribution-use
theta <- list(sigma = 2)

distrib_quantile(rayleigh, 0.5, theta)     # by root-finding on the numerical cdf
2 * sqrt(2 * log(2))                       # the closed form

distrib_gradient(rayleigh, c(1, 3), theta) # by finite differences
c(1, 3)^2 / 2^3 - 2 / 2                    # the closed form

distrib_expected_hessian(rayleigh, 1, theta)$sigma_sigma   # -4/sigma^2
```

The twenty lines above are a first-class citizen of the toolkit: the object can be
zero-inflated, truncated, transformed, and fitted by Fisher scoring exactly like
the distributions the package ships with.

## Checking one's own work {#sec-validators}

Because users are expected to write their own distributions and links, the
packages ship the tools to validate them. `check_link()` verifies that a link is
invertible in both directions, strictly monotone, consistent with the inverse
function theorem, and that each analytical derivative agrees with a numerical
one; `check_distrib()` runs the analogous battery on a distribution — the
density integrates to one, the cdf and the quantile function invert each other
and agree with the density, the random generator matches the cdf, every
derivative matches a finite-difference computation, and the expected information
matches a Monte Carlo estimate of the variance of the score.

Both run on any object, including one written five minutes ago — here, on the
two defined above:

```{r}
#| label: check-neglog
invisible(check_link(neglog))
```

```{r}
#| label: check-rayleigh
res <- check_distrib(rayleigh, theta = list(sigma = 2), orders = 1:2,
                     nsim = 5e4, verbose = FALSE)
res[, c("check", "status")]
```

The link above declares only two derivative orders of the four, and the
validator says so rather than passing what it was not given.

They are used systematically on everything the packages contain. The three
chapters that follow present the mathematics behind all of it, one per package:
@sec-links for `linkfunctions7`, @sec-distributions7 for `distributions7`, and
@sec-optimizers7 for `optimizers7`.
