# Introduction {#sec-introduction}

```{r}
#| label: setup-01
#| include: false
source("R/_setup.R")
```

## Duplicated infrastructure {#sec-rewritten}

Almost every R modeling package contains, somewhere in its internals, its
own private copy of the Gamma distribution, written as a handful of helper
functions. Typically there is a `switch` on a character string that returns a
density, a closure that computes a score, and a block inside a fitting
routine that produces the one second derivative that routine needs. The same is true of link functions. Every package that
fits something positive carries its own log link, and every package that fits a
probability carries its own logit.

None of these objects can be reused, because none of them is really an object.
They are code paths, private to the package that owns them, and they contain
only what their author's routine happened to call for — usually the density and
the score, sometimes the Hessian, almost never anything beyond. A package that
needs a third derivative writes its own Gamma from scratch; a package that wants
derivatives with respect to an unconstrained parameter does the same. The result
is that the R ecosystem contains dozens of Gamma implementations, each
partial, and each a separate opportunity for error.

The observation the toolkit is built on is that all of this is unnecessary,
because a Gamma distribution is a fixed mathematical object. It does not vary
from package to package. Its density, score, information, third and fourth derivatives, derivatives
with respect to the response and behavior under reparametrization are all
determined by the definition, once and for all. They can therefore be written
down once, correctly and completely, for everyone to build on.

That is what `statmodels7` is: infrastructure meant to be reused, not another
modeling package. Its success is measured by whether another modeling package
can be built on top of it.

## The contents of a complete implementation {#sec-completely}

The principle applies to both kinds of object the toolkit currently provides,
and it takes a different form for each, because what counts as complete
differs.

A **link function** is a map between a constrained parameter and an
unconstrained one, and what a modeling routine can ask of it is its value and
its derivatives — in both directions. Almost every implementation in R carries
the first derivative of the inverse link and stops there, which is all that a
GLM fitted by iteratively reweighted least squares requires. A link in this toolkit carries four derivatives of the forward map and four
of the inverse, so it can serve a routine working at any order up to the
fourth, and it carries its own domain, so that the calling routine need not
encode it.

A **distribution** supplies, or can derive, the following. It has the density
or mass function, the cdf, the quantile function and a random generator. It
has the score, the observed information and the expected information, plus
third and fourth derivatives of the log-likelihood, both observed and
expected. It has derivatives with respect to the response as well as the
parameters, and derivatives of the distribution function, which is what a
censored observation contributes. Finally, it provides every one of these
with respect to the unconstrained parameters as well as the natural ones.

The items at the end of each list are the ones usually missing elsewhere, and
they are precisely the ones a modeling framework depends on. Derivatives in the
response are what quantile-type residuals and smoothness penalties are made of.
Derivatives on the unconstrained scale are what an optimizer wants: a Gamma
variance must stay positive, and an algorithm forced to respect that constraint
is doing constrained optimization, whereas an algorithm working with
$\eta = \log \sigma^2$ moves freely over the whole real line. Third and fourth
derivatives, finally, are the raw material of higher-order asymptotics and bias
correction.

This is also where the two packages meet, and why neither would be much use
alone. Transporting a fourth derivative of a log-likelihood from the natural
scale to the unconstrained one is an application of Faà di Bruno's formula, and
it consumes exactly the four derivatives of the inverse link that
`linkfunctions7` carries — @sec-link-scale works this out in detail. A link
stopping at second order would silently cap the whole toolkit at second-order
methods, however much the distributions knew about themselves.

## The toolkit {#sec-toolkit}

The name encodes a convention: every package name ends in `7`, the `7` being
the S7 object system everything is built on.

| Package | What it provides |
|---|---|
| `numericals7` | the numerical layer everything above consumes: one finite-difference stencil library, quadrature and series vectorized over the parameters, the combinatorial enumerations behind the chain rules, and the special functions with their overflow discipline |
| `linkfunctions7` | 16 link classes with exact analytical derivatives to fourth order, in both directions |
| `distributions7` | 40 univariate and 4 multivariate distributions with exact score, information and third/fourth derivatives, one name per parametrization where a family has several, plus wrappers, transformations, closed-form moments and maximum likelihood |
| `optimizers7` | 11 optimization algorithms as objects, with composable stopping rules, safeguards that report themselves, starting values generated on the unconstrained scale, and box constraints removed by reparametrization |
| `basis7` | bases as objects: evaluation, derivatives of any order, the integral anchored at the lower endpoint, and exact inner products; B-splines, Fourier and Legendre, products of them for several variables, and the linear transformations that orthonormalize, constrain and diagonalize |
| `parameters7` | constrained parameters: a map from an unconstrained vector to a matrix, a simplex or a transition matrix, exact to fourth order, with the log-determinant and the solves; rank-deficient matrices admitted, since that is what a penalty is |
| `penalties7` | penalties as objects: the value with its normalizing constant, exact derivatives in the coefficients and the hyperparameters, the mixed block, and the rank, null space and log pseudo-determinant a marginal criterion consumes |
| `modelterms7` | *planned* — model terms usable directly in formulas |
| `statmodels7` | the meta-package that installs and attaches the members; later also the modeling framework itself |

: The toolkit {.striped}

The destination is the last row: a framework in the spirit of the generalized
additive models for location, scale and shape of @rigby2005, described at length
in @stasinopoulos2017. In that framework every parameter of a distribution can
be modeled rather than just the mean, the optimizer is pluggable, and
estimation can be frequentist or Bayesian. Users can also define new model
terms and drop them straight into formulas. Everything below it
exists to make that possible, and one design pressure has shaped it all: *exact
derivatives, on the unconstrained scale, for components the framework has never
seen before*.

## Extending the toolkit {#sec-extension}

A toolkit restricted to the sixteen links and forty-four distributions it
ships with would be one more closed catalog. The relevant test is what happens
when a user brings an object the authors never considered, and the answer is
that the object is treated like the ones that came with the packages: there is
no registration step and no separate interface for the user-defined case.

For a link the requirement is only that the ten generics be implemented, and
since a link is a formula and its derivatives, that is a dozen lines. Nothing
distinguishes the result from a built-in one: it can be handed to any
distribution as the link for any parameter, and the fourth-order machinery
treats it identically to a shipped link.

```{r}
#| label: minimal-link
NegLog <- S7::new_class("NegLog", parent = link, package = NULL)

S7::method(linkfun,   NegLog) <- function(x, theta) -log(theta)
S7::method(linkinv,   NegLog) <- function(x, eta)   exp(-eta)
S7::method(dlinkfun,  NegLog) <- function(x, theta) -1 / theta
S7::method(d2linkfun, NegLog) <- function(x, theta)  1 / theta^2
S7::method(dlinkinv,  NegLog) <- function(x, eta)   -exp(-eta)
S7::method(d2linkinv, NegLog) <- function(x, eta)    exp(-eta)

neglog <- NegLog(link_name = "neglog", link_bounds = c(0, Inf), link_params = NULL)
linkinv(neglog, linkfun(neglog, c(0.5, 2)))
```

For a distribution the contract is even smaller, and it is the one that decides
whether the package is infrastructure or a catalog: **a new distribution needs
nothing but its density**. Everything else has a fallback —
the cdf is computed by quadrature, the quantile function by root-finding, the
random generator by a ratio-of-uniforms scheme that needs only the density, and
every derivative by finite differences. Whatever the user later supplies in
closed form replaces the corresponding fallback through ordinary method
dispatch.

To see what this means in practice, here is a complete definition of the
Rayleigh distribution, which has density
$f(y;\sigma) = (y/\sigma^2)\, e^{-y^2/2\sigma^2}$ on $(0, \infty)$:

```{r}
#| label: minimal-distribution
Rayleigh <- S7::new_class("Rayleigh", parent = continuous_distrib, package = NULL)

S7::method(distrib_pdf, Rayleigh) <- function(distrib, y, theta, log = FALSE) {
  s <- theta[[1]]
  ld <- log(y) - 2 * log(s) - y^2 / (2 * s^2)
  if (log) ld else exp(ld)
}

rayleigh <- Rayleigh(
  distrib_name = "rayleigh", dimension = "univariate", bounds = c(0, Inf),
  params = "sigma", params_interpretation = c(sigma = "scale"), n_params = 1,
  params_bounds = list(sigma = c(0, Inf)),
  link_params = list(sigma = log_link())
)
```

Nothing above mentions a cdf, a quantile function, a generator or a derivative,
and yet all of them are immediately available. The median of a Rayleigh is
$\sigma\sqrt{2\log 2}$, its score is $y^2/\sigma^3 - 2/\sigma$, and its expected
information is $4/\sigma^2$; the fallbacks recover each of these from the
density alone:

```{r}
#| label: minimal-distribution-use
theta <- list(sigma = 2)

distrib_quantile(rayleigh, 0.5, theta)     # by root-finding on the numerical cdf
2 * sqrt(2 * log(2))                       # the closed form

distrib_gradient(rayleigh, c(1, 3), theta) # by finite differences
c(1, 3)^2 / 2^3 - 2 / 2                    # the closed form

distrib_expected_hessian(rayleigh, 1, theta)$sigma_sigma   # -4/sigma^2
```

The twenty lines above are a first-class citizen of the toolkit: the object can be
zero-inflated, truncated, transformed, and fitted by Fisher scoring exactly like
the distributions the package ships with.

## Validators {#sec-validators}

Because users are expected to write their own distributions and links, the
packages ship the tools to validate them. `check_link()` verifies that a link is invertible in both directions,
strictly monotone and consistent with the inverse function theorem, and that
each analytical derivative agrees with a numerical one. `check_distrib()`
runs the analogous battery on a distribution. It checks that the density
integrates to one, that the cdf and the quantile function invert each other
and agree with the density, and that the random generator matches the cdf.
It also checks every derivative against a finite-difference computation, and
the expected information against a Monte Carlo estimate of the variance of
the score.

Both run on any object, including one just defined — here, on the two objects
above:

```{r}
#| label: check-neglog
invisible(check_link(neglog))
```

```{r}
#| label: check-rayleigh
res <- check_distrib(rayleigh, theta = list(sigma = 2), orders = 1:2,
                     nsim = 5e4, verbose = FALSE)
res[, c("check", "status")]
```

The link above declares only two derivative orders of the four, and the
validator reports the remaining orders as unchecked rather than as passed.

`basis7` ships `check_basis()`, which does the same for a basis and reports a
quantity it cannot check independently as unchecked rather than as passed.

The validators are used systematically on everything the packages contain. The
seven chapters that follow present the mathematics behind all of it, one per
package: @sec-links for `linkfunctions7`, @sec-distributions7 for
`distributions7`, @sec-optimizers7 for `optimizers7`, @sec-basis7 for
`basis7`, @sec-parameters7 for `parameters7`, @sec-numericals7 for
`numericals7`, and @sec-penalties7 for `penalties7`.
