# A first model {#sec-model}

```{r}
#| label: setup-01
#| include: false
source("R/_setup.R")
source("R/model-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x, d = 1) paste0(fmt(100 * x, d), "%")
sci <- function(x, d = 1) {
  e <- floor(log10(abs(x)))
  sprintf("%.*f \\times 10^{%d}", d, x / 10^e, e)
}
```

This chapter fits one model and reads everything that `statmod()` returns
for it. We fit a smooth curve to a data set with a single covariate, print
the fit and its summary, extract the estimates and their variances, look at
the residuals, and compute predictions. Then we add a second equation for
the spread of the data, compare the two models, check that the fit
converged, and simulate new data from the fitted model.

## The data {#sec-model-data}

The data come from a simulated motorcycle accident that was used to test
crash helmets. They are measurements of the acceleration of the head after
the impact. For each observation we know the time after the impact, in
milliseconds, and the acceleration of the head, in units of $g$.
@silverman1985 analyzed these data, and they are distributed with the `MASS`
package [@venables2002].

```{r}
#| label: data
data(mcycle, package = "MASS")
nrow(mcycle)
head(mcycle, n = 3)
mcycle[which.min(mcycle$accel), ]
```

```{r}
#| label: data-numbers
#| include: false
i_min <- which.min(mcycle$accel)
t_imp <- min(mcycle$times[mcycle$accel < -10])
```

@fig-mcycle-data shows the `r nrow(mcycle)` observations. For the first
`r fmt(t_imp, 1)` milliseconds the acceleration stays within a few units of
zero. Then it drops sharply, reaches its minimum of
$`r fmt(mcycle$accel[i_min], 0)`\,g$ at `r fmt(mcycle$times[i_min], 1)`
milliseconds, rises above zero, and finally levels off. A polynomial of low
degree cannot follow this shape. The spread of the observations around the
mean also changes with time: it is small at the start, large during the
drop and the rebound, and smaller again at the end.

```{r}
#| label: fig-mcycle-data
#| fig-cap: "Head acceleration against time after impact in the `mcycle` data."
#| echo: false
ggplot(mcycle, aes(x = times, y = accel)) +
  geom_point(colour = book_grey, size = 1.3) +
  labs(x = "time after impact (ms)", y = "acceleration (g)")
```

## The first fit {#sec-model-fit}

We start with a model in which the acceleration $y_i$ at time $t_i$ has a
gaussian distribution. The mean changes smoothly with time, and the standard
deviation is the same for all observations:
$$
y_i \sim \mathrm{N}(\mu_i, \sigma), \qquad
\mu_i = \beta_0 + f(t_i).
$$
We write $\mathrm{N}(\mu, \sigma)$ for the gaussian distribution with mean
$\mu$ and standard deviation $\sigma$, the two parameters of
`gaussian1_distrib()`, and we use this notation throughout the book.
We write the function $f$ as a combination of $k = 20$ cubic B-splines,
$f(t) = \sum_{r=1}^{20} b_r B_r(t)$. To keep $f$ smooth, we subtract from the
log-likelihood a penalty on its curvature. We measure the curvature by the
integral of the squared second derivative:
$$
\tfrac{1}{2}\lambda \int f''(t)^2\,dt = \tfrac{1}{2}\lambda\, b^\top P b ,
$$
where $P$ is the matrix with entries $P_{rs} = \int B_r''(t)\,B_s''(t)\,dt$.
The smoothing parameter $\lambda$ controls how much the curve can bend. A
large $\lambda$ gives a curve close to a straight line, and a small $\lambda$
lets the curve follow the data closely.

For a given $\lambda$, `statmod()` estimates the coefficients by maximizing
the penalized log-likelihood, which is the log-likelihood minus the penalty.
It estimates $\lambda$ by restricted maximum likelihood (REML). For REML we
read the penalty as a prior distribution on the coefficients, integrate the
coefficients out of the likelihood, and maximize the result over $\lambda$
[@wood2011; @wood2017]. In `statmodels7` we write the model like this:

```{r}
#| label: fit0
library(statmodels7)
fit0 <- statmod(accel ~ s(times, bspline_smooth(k = 20)),
                distrib = gaussian1_distrib(), data = mcycle)
```

The term `s(times, bspline_smooth(k = 20))` is a penalized smooth function of
`times`. Its second argument, `bspline_smooth(k = 20)`, defines how the
function is built: from a basis of 20 cubic B-splines, with a penalty on the second
derivative. The distribution `gaussian1_distrib()` is the gaussian
distribution with two parameters, the mean `mu` and the standard deviation
`sigma`. The formula has an equation only for the mean, so `statmod()`
estimates `sigma` as a single constant. Printing the fit gives a short
description of it:

```{r}
#| label: print-fit0
fit0
```

The printout lists the equation of each parameter of the distribution and
the terms in it. The equation of `mu` has two terms: `linpar`, which holds
the intercept, and the smooth. The smooth has 19 coefficients, one fewer than
the number of B-splines, and we explain why in the next section. The number
`edf` is its effective degrees of freedom, which we define in
@sec-model-compare. The equation of `sigma` contains only an intercept. The
line that starts with `REML` gives the value of the REML criterion at the
estimate and the number of values of $\lambda$ that the search evaluated.
The last line gives the time the fit took.

## The summary {#sec-model-summary}

The function `summary()` prints the estimates one equation at a time.

```{r}
#| label: summary-fit0
summary(fit0)
```

```{r}
#| label: summary-numbers
#| include: false
cf0 <- confint(fit0, "sigma")
```

The block of each equation starts with the name of the parameter and its
link function. The mean uses the identity link, so its equation describes
$\mu_i$ directly. The standard deviation uses the log link, so its equation
describes $\log\sigma$.

The table `Parametric terms` holds the coefficients without a penalty, here
the intercepts. For each of them it gives the estimate, the standard error,
the Wald statistic $z$ (the estimate divided by its standard error), the
p-value of the test that the coefficient is zero, and a 95% confidence
interval. The intercept of `sigma` is on the log scale: $\log\hat\sigma =
`r fmt(cf0$estimate, 3)`$, so $\hat\sigma = `r fmt(exp(cf0$estimate))`$.

Each smooth term has a block of its own. The heading of the block gives the
number of coefficients of the term and its effective degrees of freedom. The
smooth has 19 coefficients, one fewer than the 20 B-splines. The B-splines
add up to one at every time, so together they contain the constant function,
and the intercept of the equation already represents the constant. For this
reason `statmod()` constrains the smooth to sum to zero over the observed
times, and this removes one coefficient.

The first row of the block is the smoothing parameter. The label `[reml]`
means that REML estimated it. Its interval is computed on the log scale and
then transformed back, so it contains only positive values. The second row,
`lin`, is the linear part of the smooth. The penalty on the second
derivative is zero for a straight line, so this coefficient is not
penalized, and we can test it like any other coefficient. It multiplies the
time standardized to mean zero and standard deviation one, so its value is
the slope of the linear part of $f$ per standard deviation of time:

```{r}
#| label: sd-times
sd(mcycle$times)
```

The other 18 coefficients of the smooth are not printed. Their values depend
on how the basis is written, so one by one they have no meaning.

The lines at the bottom are about the fit as a whole. The log-likelihood is
the *conditional* log-likelihood, which is computed at the estimated
coefficients. cAIC is minus twice this value plus twice the effective
degrees of freedom [@vaida2005], and cBIC uses $\log n$ in place of the
factor 2. @sec-model-compare computes both.

The lines `inner` and `outer` describe the point where the fit stopped. We
call curvature the negative of the matrix of second derivatives. The line
`inner` is about the coefficients. Its first number is the largest component
of the gradient of the penalized log-likelihood, where each component is
divided by the standard error that its coefficient would have if the other
coefficients were known. Its second number is the smallest eigenvalue of the
curvature of the penalized log-likelihood, after the curvature is rescaled to
have ones on its diagonal. The line `outer` gives the same two numbers for
the REML criterion as a function of the smoothing parameter. At a maximum the
first number is close to zero and the second is positive. In this fit both
lines meet these two conditions. @sec-model-certificate checks the same thing
with another function.

The last line gives the command that prints the notes of the summary. For
this fit there is one note. It states that the interval of the smoothing
parameter is built on the log scale and mapped back, and that the other
estimates are conditional on the estimated value of $\lambda$.

## Coefficients and their variances {#sec-model-coef}

The function `coef()` returns a list with one element for each parameter of
the distribution. Each element holds the coefficients of that parameter's
equation, on the scale of its link function.

```{r}
#| label: coef-fit0
coef(fit0)
```

The coefficients named `s(times).z1` to `s(times).z18` are the penalized
coefficients of the smooth. As we said above, they have no meaning one by
one. The function `vcov()` returns the variance matrix of all the
coefficients, here a matrix with `r nrow(vcov(fit0))` rows and columns. We
print the rows and columns of the two intercepts and of the linear part of
the smooth:

```{r}
#| label: vcov-fit0
V <- vcov(fit0)
dim(V)
keep <- c("mu:(Intercept)", "mu:s(times).lin", "sigma:(Intercept)")
V[keep, keep]
```

The covariances between the coefficients of `mu` and the coefficient of
`sigma` are exactly zero. For the gaussian distribution the mean and the
standard deviation carry separate information, so the block of the Fisher
information that connects the two equations is zero, and so is the
corresponding block of the variance matrix. The covariance between the
intercept of `mu` and the linear part of the smooth is zero up to rounding,
because the linear column has mean zero.

The function `confint()` returns a data frame with one row for each
coefficient: the estimate, the standard error and the two ends of a 95%
interval. With the argument `parm` we choose the coefficients:

```{r}
#| label: confint-fit0
ci0 <- confint(fit0, parm = keep)
ci0
```

The standard errors are the square roots of the diagonal of `V`, and each
interval is the estimate plus or minus 1.96 standard errors:

```{r}
#| label: confint-by-hand
sqrt(diag(V[keep, keep]))
ci0$estimate - qnorm(p = 0.975) * ci0$se
ci0$estimate + qnorm(p = 0.975) * ci0$se
```

## The smoothing parameter {#sec-model-hyper}

The smoothing parameter is not a coefficient of an equation, so `coef()`
does not return it. The function `hyper()` returns the hyperparameters of the
model, which here are the one smoothing parameter.

```{r}
#| label: hyper-fit0
hyper(fit0)
```

The column `estimate` holds the value, and the column `source` records how
it was estimated, here by REML. The column `held` is `TRUE` for a value that we
fix in the formula instead of estimating it. For example,
`s(times, bspline_smooth(k = 20), hyper = c(lambda = 1))` holds $\lambda$ at
1. The column `id` names a group of terms that share one hyperparameter, and
it is empty here.

## Fitted values and residuals {#sec-model-resid}

The function `fitted()` returns the fitted mean $\hat\mu_i$ of each
observation:

```{r}
#| label: fitted-fit0
head(fitted(fit0))
```

The function `residuals()` returns quantile residuals [@dunn1996]. For
observation $i$ with fitted distribution function $\hat F_i$, the quantile
residual is
$$
r_i = \Phi^{-1}\big(\hat F_i(y_i)\big),
$$
where $\Phi$ is the standard gaussian distribution function. If the model is
right, $\hat F_i(y_i)$ is close to a uniform variable, so $r_i$ is close to
a standard gaussian variable, whatever the distribution of the response. For
a gaussian response $r_i$ equals $(y_i - \hat\mu_i)/\hat\sigma_i$:

```{r}
#| label: residuals-fit0
r0 <- residuals(fit0)
head(r0)
s0 <- exp(coef(fit0)$sigma)
head((mcycle$accel - fitted(fit0)) / s0)
```

If the model is right, the residuals should have a standard deviation close
to 1 in every part of the time range. We compute it separately for the early
phase before 15 milliseconds, for the drop and rebound between 15 and 40
milliseconds, and for the late phase:

```{r}
#| label: residuals-phases
early <- mcycle$times < 15
middle <- mcycle$times >= 15 & mcycle$times < 40
late <- mcycle$times >= 40
sd(r0[early])
sd(r0[middle])
sd(r0[late])
sum(abs(r0) > 2)
```

The standard deviation is `r fmt(sd(r0[early]))` in the early phase,
`r fmt(sd(r0[middle]))` in the middle and `r fmt(sd(r0[late]))` at the end.
The model uses one value of $\sigma$ for all times, so it overstates the
spread at the start and understates it in the middle. @fig-resid0 shows the
same thing: the residuals are small at the start, large in the middle and
moderate at the end. Of the `r nrow(mcycle)` residuals, `r sum(abs(r0) > 2)`
lie outside $\pm 2$, and they all fall in the middle phase.

```{r}
#| label: fig-resid0
#| fig-cap: "Quantile residuals of the model with a constant standard deviation, against time."
#| echo: false
ref0 <- data.frame(y = c(0, -2, 2),
                   line = factor(c("zero", "-2 and 2", "-2 and 2"),
                                 levels = c("zero", "-2 and 2")))
ggplot(data.frame(times = mcycle$times, r = r0), aes(x = times, y = r)) +
  geom_point(aes(shape = "quantile residual"), colour = book_grey,
             size = 1.3) +
  geom_hline(aes(yintercept = y, linetype = line), data = ref0,
             colour = rust) +
  scale_shape_manual(values = 16) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  guides(shape = guide_legend(order = 1),
         linetype = guide_legend(order = 2)) +
  labs(x = "time after impact (ms)", y = "quantile residual")
```

## Predictions {#sec-model-predict}

The function `predict()` computes a parameter of the fitted distribution at
new values of the covariates. Its second argument names the parameter, and
`newdata` holds the covariates. With `se = TRUE` it also gives a standard
error and a 95% interval:

```{r}
#| label: predict-fit0
new <- data.frame(times = c(5, 20, 30, 50))
predict(fit0, what = "mu", newdata = new, se = TRUE)
```

For the mean, the interval is the prediction plus or minus 1.96 standard
errors. For a parameter with a link function, the interval is computed on
the scale of the link and then transformed back. For $\sigma$ this is the
log scale, so the interval for $\sigma$ contains only positive values. In
this model $\sigma$ is the same at every time:

```{r}
#| label: predict-sigma-fit0
predict(fit0, what = "sigma", newdata = new, se = TRUE)
```

@fig-band0 shows the fitted mean on a grid of 200 times, with two bands. The
dark band is the 95% interval of the mean from `predict(se = TRUE)`. It
describes the uncertainty of the estimated curve. The light band is
$\hat\mu(t) \pm 2\hat\sigma$, the range where about 95% of the observations
should fall if the model is right.

```{r}
#| label: band-fit0
grid <- data.frame(times = seq(from = min(mcycle$times),
                               to = max(mcycle$times), length.out = 200))
p0 <- predict(fit0, what = "mu", newdata = grid, se = TRUE)
head(p0, n = 3)
```

```{r}
#| label: fig-band0
#| fig-cap: "The model with a constant standard deviation. The line is the fitted mean, the dark band is its 95% confidence interval, and the light band is the fitted mean plus or minus two fitted standard deviations."
#| echo: false
bands0 <- data.frame(
  times = rep(grid$times, times = 2),
  lower = c(p0$lower, p0$fit - 2 * s0),
  upper = c(p0$upper, p0$fit + 2 * s0),
  band = factor(rep(c("95% interval of the mean",
                      "mean plus or minus 2 sigma"), each = nrow(grid)),
                levels = c("95% interval of the mean",
                           "mean plus or minus 2 sigma")))
ggplot(mcycle, aes(x = times)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = band), data = bands0) +
  geom_point(aes(y = accel, shape = "data"), colour = book_grey, size = 1.2) +
  geom_line(aes(y = fit, colour = "fitted mean"),
            data = data.frame(times = grid$times, fit = p0$fit),
            linewidth = 0.8) +
  scale_shape_manual(values = 16) +
  scale_colour_manual(values = chalk) +
  scale_fill_manual(values = c(adjustcolor(chalk, 0.5),
                               adjustcolor(chalk, 0.15))) +
  guides(shape = guide_legend(order = 1), colour = guide_legend(order = 2),
         fill = guide_legend(order = 3)) +
  labs(x = "time after impact (ms)", y = "acceleration (g)")
```

The fitted mean follows the data closely, but the light band has the same
width everywhere. It is far too wide in the first milliseconds and too narrow
during the drop and the rebound. Changing $\lambda$ does not help, because
the problem is the assumption of a constant standard deviation.

## An equation for the standard deviation {#sec-model-sigma}

To fix this, we let the standard deviation change smoothly with time as
well:
$$
y_i \sim \mathrm{N}(\mu_i, \sigma_i), \qquad
\mu_i = \beta_0 + f(t_i), \qquad
\log \sigma_i = \gamma_0 + g(t_i).
$$
Here $g$ is a second penalized smooth function, with its own smoothing
parameter. We model $\log\sigma_i$ rather than $\sigma_i$ so that $\sigma_i$
is always positive. A model with an equation for each parameter of the
distribution is called a generalized additive model for location, scale and
shape [@rigby2005]. In `statmodels7` we write it like this:

```{r}
#| label: fit1
fit1 <- statmod(accel ~ s(times, bspline_smooth(k = 20)) |
                  sigma ~ s(times, bspline_smooth(k = 10)),
                distrib = gaussian1_distrib(), data = mcycle)
fit1
```

The formula now contains two equations, separated by a vertical bar. The
equation before the bar has the response on the left, and it describes the
first parameter of the distribution, which is `mu`. The equation after the
bar has the name of a parameter on the left, `sigma`, and it describes that
parameter. In general there is one equation for each parameter that we want
to model. If the formula has no equation for a parameter, `statmod()`
estimates that parameter as a constant, as it did for `sigma` in the first
model. The link functions are those of the distribution, so the right-hand
side of the second equation describes $\log\sigma_i$.

The summary now has a smooth block in each equation:

```{r}
#| label: summary-fit1
summary(fit1)
```

The two smoothing parameters are estimated by REML together. The linear
part of the smooth of $\log\sigma$ has a positive estimate with a p-value
well below 0.001, so the standard deviation grows with time on average. The
smooth adds curvature to this trend. We compute $\hat\sigma(t)$ at four
times with its interval:

```{r}
#| label: predict-sigma-fit1
predict(fit1, what = "sigma", newdata = new, se = TRUE)
```

```{r}
#| label: sigma-numbers
#| include: false
ps <- predict(fit1, what = "sigma", newdata = grid, se = TRUE)
sv <- predict(fit1, what = "sigma", newdata = new)
smax <- max(ps$fit)
tmax <- grid$times[which.max(ps$fit)]
```

@fig-sigma1 shows $\hat\sigma(t)$ over the whole time range. It is
`r fmt(sv[1])` $g$ at `r new$times[1]` milliseconds, in the early flat phase.
It reaches a maximum of `r fmt(smax, 1)` $g$ at `r fmt(tmax, 1)` milliseconds
and falls to `r fmt(sv[4], 1)` $g$ at `r new$times[4]` milliseconds. The
constant estimate of the first model, `r fmt(s0)` $g$, is too large at the
start and at the end, and too small in the middle.

```{r}
#| label: fig-sigma1
#| fig-cap: "The fitted standard deviation of the second model with its pointwise 95% interval, and the constant standard deviation of the first model."
#| echo: false
sig1 <- data.frame(
  times = rep(grid$times, times = 2),
  sigma = c(ps$fit, rep(s0, times = nrow(grid))),
  curve = factor(rep(c("sigma(t), second model", "sigma, first model"),
                     each = nrow(grid)),
                 levels = c("sigma(t), second model", "sigma, first model")))
ggplot(sig1, aes(x = times)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = "95% interval"),
              data = data.frame(times = grid$times, lower = ps$lower,
                                upper = ps$upper)) +
  geom_line(aes(y = sigma, colour = curve, linetype = curve),
            linewidth = 0.8) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_fill_manual(values = adjustcolor(chalk, 0.25)) +
  expand_limits(y = 0) +
  guides(colour = guide_legend(order = 1),
         linetype = guide_legend(order = 1),
         fill = guide_legend(order = 2)) +
  labs(x = "time after impact (ms)", y = "standard deviation (g)")
```

The quantile residuals of the second model divide each observation by its
own $\hat\sigma_i$. We compute their standard deviation in the same three
phases as before:

```{r}
#| label: residuals-fit1
r1 <- residuals(fit1)
sd(r1[early])
sd(r1[middle])
sd(r1[late])
sum(abs(r1) > 2)
```

The three values are now `r fmt(sd(r1[early]))`, `r fmt(sd(r1[middle]))` and
`r fmt(sd(r1[late]))`, all close to 1, and `r sum(abs(r1) > 2)` residuals
lie outside $\pm 2$, against the `r fmt(nrow(mcycle) * 2 * pnorm(-2), 1)`
that a standard gaussian sample of `r nrow(mcycle)` values gives on average.
@fig-resid01 shows the residuals of the two models side by side.

```{r}
#| label: fig-resid01
#| fig-cap: "Quantile residuals against time. Left: the model with a constant standard deviation. Right: the model with an equation for the standard deviation."
#| echo: false
#| fig-height: 3.8
#| fig-width: 9
res01 <- data.frame(
  times = rep(mcycle$times, times = 2),
  r = c(r0, r1),
  model = factor(rep(c("constant sigma", "sigma ~ s(times)"),
                     each = nrow(mcycle)),
                 levels = c("constant sigma", "sigma ~ s(times)")))
ggplot(res01, aes(x = times, y = r)) +
  geom_point(aes(shape = "quantile residual"), colour = book_grey,
             size = 1.1) +
  geom_hline(aes(yintercept = y, linetype = "-2 and 2"),
             data = data.frame(y = c(-2, 2)), colour = rust) +
  facet_wrap(facets = vars(model)) +
  scale_shape_manual(values = 16) +
  scale_linetype_manual(values = "dashed") +
  guides(shape = guide_legend(order = 1),
         linetype = guide_legend(order = 2)) +
  labs(x = "time after impact (ms)", y = "quantile residual")
```

## Comparing the two models {#sec-model-compare}

A penalized coefficient does not count as a whole parameter, because the
penalty limits how far it can move. So we need another way to count the
parameters of a penalized fit. Let $\tilde\beta$ be the estimate without the
penalty, let $H$ be the Fisher information of the coefficients at the fit,
and let $S$ be the matrix of second derivatives of the penalty. One Newton
step from $\tilde\beta$ towards the penalized estimate gives
$$
\hat\beta = (H+S)^{-1} H \tilde\beta .
$$
The matrix $F = (H+S)^{-1}H$ turns the unpenalized estimate into the
penalized one. Without a penalty, $S = 0$ and $F$ is the identity matrix, so
its trace is the number of coefficients. As the smoothing parameters grow,
the trace falls towards the number of unpenalized coefficients. For this
reason we use the trace of $F$ as the number of parameters that the fit
uses, and we call it the effective degrees of freedom. The share of one term
is the trace of the diagonal block of $F$ that belongs to that term. The fit
stores the shares in the slot `edf`:

```{r}
#| label: edf
fit0@edf
fit1@edf
```

```{r}
#| label: edf-numbers
#| include: false
e0 <- sum(fit0@edf$edf)
e1 <- sum(fit1@edf$edf)
```

The smooth of the mean uses `r fmt(fit0@edf$edf[2])` effective degrees of
freedom in the first model and `r fmt(fit1@edf$edf[2])` in the second, where
the smooth of the standard deviation uses `r fmt(fit1@edf$edf[4])` more. The
share of the smooth of the mean changes between the two models because the
information of the mean weights each observation by $1/\sigma_i^2$. In the
first model all the weights are equal, and in the second they change with
time.

The function `logLik()` returns the conditional log-likelihood, with the
total effective degrees of freedom as its attribute `df`. The functions
`AIC()` and `BIC()` return cAIC and cBIC:

```{r}
#| label: loglik
logLik(fit0)
logLik(fit1)
AIC(fit0)
AIC(fit1)
BIC(fit0)
BIC(fit1)
```

cAIC is $-2\ell + 2\tau$, where $\ell$ is the conditional log-likelihood and
$\tau$ the total effective degrees of freedom, and cBIC uses $\log n$ in
place of 2. We compute both for the second model:

```{r}
#| label: aic-by-hand
l1 <- as.numeric(logLik(fit1))
tau1 <- sum(fit1@edf$edf)
-2 * l1 + 2 * tau1
-2 * l1 + log(nrow(mcycle)) * tau1
```

When we model the standard deviation, the conditional log-likelihood
increases by `r fmt(as.numeric(logLik(fit1)) - as.numeric(logLik(fit0)))`,
and the model uses `r fmt(e1 - e0)` more effective degrees of freedom. cAIC
goes down from `r fmt(AIC(fit0), 1)` to `r fmt(AIC(fit1), 1)` and cBIC from
`r fmt(BIC(fit0), 1)` to `r fmt(BIC(fit1), 1)`, so by both criteria the
second model is better.

The effective degrees of freedom also give the estimate of $\sigma$ in the
first model. REML estimates $\sigma$ as $\sqrt{\mathrm{RSS}/(n - \tau_\mu)}$,
where RSS is the residual sum of squares and $\tau_\mu$ is the effective
degrees of freedom of the equation of the mean. This is the value that
`lme()` and `mgcv` report. The maximum likelihood estimate
$\sqrt{\mathrm{RSS}/n}$ is smaller, because the residuals of a fitted mean
are smaller than the errors:

```{r}
#| label: sigma-reml
rss <- sum((mcycle$accel - fitted(fit0))^2)
tau_mu <- sum(fit0@edf$edf[fit0@edf$parameter == "mu"])
sqrt(rss / (nrow(mcycle) - tau_mu))
sqrt(rss / nrow(mcycle))
s0
```

## Did the fit converge? {#sec-model-certificate}

An optimizer reports whether its own stopping rule was met. This is a
statement about the search, not about the returned point. The function
`statmod_certificate()` checks the returned point itself.

```{r}
#| label: certificate-fit0
statmod_certificate(fit0)
```

```{r}
#| label: certificate-numbers
#| include: false
c0 <- statmod_certificate(fit0)
c1 <- statmod_certificate(fit1)
```

The certificate reads two quantities. The first, `mode_error`, is about the
coefficients. Let $g$ be the gradient of the penalized log-likelihood at the
returned coefficients and let $K$ be its curvature. The quantity
$\tfrac{1}{2} g^\top K^{-1} g$ is the increase of the penalized
log-likelihood that one more Newton step would give, so it measures, in
log-likelihood units, how far the coefficients are from their maximum. Here
it is $`r sci(c0$mode_error)`$. The second, `decrement`, is the same
quantity for the REML criterion as a function of the smoothing parameter.
Here it is $`r sci(c0$decrement)`$. The field `curvature` is `analytic`,
so the curvature of the criterion was computed from exact derivatives. The fields
`inner` and `outer` are the two lines of the summary. The fields `boundary`
and `reason` are empty: no smoothing parameter sits at the edge of its
range, and nothing prevented the check.

The field `state` summarizes the result. It is `converged` when the
decrement is below its tolerance. It is `boundary` when the same holds with
a smoothing parameter at the edge of its range, where the criterion no
longer changes with it. It is `not converged` when the criterion would still
increase, and `unknown` when the model has no criterion with derivatives to
check. For the second model we print only the state:

```{r}
#| label: certificate-fit1
statmod_certificate(fit1)$state
```

## Simulating from the fitted model {#sec-model-simulate}

The fitted model gives, for each observation, a distribution for the
response: here a gaussian distribution with mean $\hat\mu_i$ and standard
deviation $\hat\sigma_i$. If the model describes the data well, data drawn
from these distributions should look like the observed data. The function
`rstatmod()` draws data from a model that we write, with a formula, a
distribution and the covariates. The argument `par` gives the coefficients of
each equation. When we pass the estimated coefficients, the model is the
fitted one:

```{r}
#| label: rstatmod
set.seed(1)
sim <- rstatmod(accel ~ s(times, bspline_smooth(k = 20)) |
                  sigma ~ s(times, bspline_smooth(k = 10)),
                distrib = gaussian1_distrib(), data = mcycle,
                par = coef(fit1), n_sim = 3)
sim
```

The result holds the simulated data sets in `sim$data`, one data frame for
each of the `n_sim = 3` replicates, and the true value of every parameter at
every observation in `sim$theta`. Here the true values are the fitted ones:

```{r}
#| label: rstatmod-truth
head(sim$data[[1]])
all.equal(target = sim$theta[[1]]$sigma,
          current = predict(fit1, what = "sigma", newdata = mcycle))
```

The function `simulate()` draws from a fitted model directly, and it returns
the simulated responses as the columns of a data frame:

```{r}
#| label: simulate
set.seed(1)
sims <- simulate(fit1, nsim = 3)
head(sims, n = 3)
```

With `rstatmod()` we can also change the model. For example, we can keep the
estimated coefficients of the mean and write other coefficients for the
standard deviation, or leave out `par` and let `rstatmod()` draw every
coefficient. A simulation study with a known truth starts in this way.

```{r}
#| label: fig-sims
#| fig-cap: "The observed data (top left) and three data sets simulated from the second model with `rstatmod()`."
#| echo: false
#| fig-height: 5.6
panels <- c("observed", "simulated 1", "simulated 2", "simulated 3")
sim_df <- data.frame(
  times = rep(mcycle$times, times = 4),
  accel = c(mcycle$accel, sim$data[[1]]$accel, sim$data[[2]]$accel,
            sim$data[[3]]$accel),
  panel = factor(rep(panels, each = nrow(mcycle)), levels = panels))
sim_df$kind <- ifelse(sim_df$panel == "observed", "observed", "simulated")
ggplot(sim_df, aes(x = times, y = accel, colour = kind)) +
  geom_point(size = 1.1) +
  facet_wrap(facets = vars(panel), ncol = 2) +
  scale_colour_manual(values = c(observed = rust, simulated = book_grey),
                      guide = "none") +
  labs(x = "time (ms)", y = "acceleration (g)")
```

In @fig-sims the simulated data look like the observed data. They are flat
in the first milliseconds, they drop and come back up at the same times, and
their spread changes with time in the same way.

```{r}
#| label: gate-01
#| include: false
assert_model_ok(fit0, fit1, mcycle,
                claims = list(ci0 = ci0, keep = keep, p0 = p0, grid = grid,
                              new = new, sim = sim, sims = sims))
```
