# Beyond the basics {#sec-beyond}

```{r}
#| label: setup-02
#| include: false
source("R/_setup.R")
source("R/beyond-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x, d = 1) paste0(fmt(100 * x, d), "%")
sci <- function(x, d = 1) {
  e <- floor(log10(abs(x)))
  sprintf("%.*f \\times 10^{%d}", d, x / 10^e, e)
}
data(mcycle, package = "MASS")
```

@sec-model fitted one model and read what `statmod()` returns for it. This
chapter goes further in nine directions. We compare the three variance
matrices that a penalized fit offers, and we explain the two
log-likelihoods of a model with random effects. We test a single
coefficient with four different statistics, invert these tests into
confidence intervals, and compare two nested models with a likelihood ratio
test. Then we predict for a group that the data do not contain, start a fit
from the estimates of another fit, change the criterion that estimates the
hyperparameters, and use prior weights, offsets and parameters with a known
value.

We start again from the two models of @sec-model for the motorcycle data:
a smooth mean with a constant standard deviation, and a smooth mean with a
smooth standard deviation.

```{r}
#| label: fits-mcycle
library(statmodels7)
fit0 <- statmod(accel ~ s(times, bspline_smooth(k = 20)),
                distrib = gaussian1_distrib(), data = mcycle)
fit1 <- statmod(accel ~ s(times, bspline_smooth(k = 20)) |
                  sigma ~ s(times, bspline_smooth(k = 10)),
                distrib = gaussian1_distrib(), data = mcycle)
```

## Three variance matrices {#sec-beyond-variances}

The summary of @sec-model ends with the line "95% intervals, bayesian
variance". Every standard error of a penalized model depends on a choice of
variance matrix, and `statmodels7` offers three of them. As in @sec-model,
let $H$ be the Fisher information of the coefficients at the fit, and let
$S$ be the matrix of second derivatives of the penalty. The first two
matrices are
$$
V_b = (H + S)^{-1}, \qquad V_f = (H + S)^{-1} H (H + S)^{-1} .
$$
$V_f$ is the sampling variance of the penalized estimator when the
smoothing parameters are fixed. $V_b$ is the posterior variance of the
coefficients when we read the penalty as a gaussian prior, as REML does.
$V_b$ is larger than $V_f$. The difference accounts for the bias that the
penalty introduces: a penalized curve is pulled towards a straight line, so
an interval built from $V_f$ covers the true curve less often than its
nominal level. Intervals built from $V_b$ have close to their nominal
coverage when we average over the points of the curve
[@wahba1983; @nychka1988; @marra2012]. This is why $V_b$ is the default.

Both matrices treat the smoothing parameters as known, although REML
estimated them from the same data. The third matrix adds the uncertainty of
the smoothing parameters [@woodpya2016]:
$$
V_u = V_b + J V_\theta J^\top, \qquad
J = \frac{\partial \hat\beta}{\partial \theta},
$$
where $\theta$ contains the logarithms of the smoothing parameters and
$V_\theta$ is their variance, which we read from the curvature of the REML
criterion. The matrix $J$ measures how far the estimated coefficients move
when a smoothing parameter moves.

We choose among the three matrices with the argument `type` of `vcov()`,
`confint()`, `summary()` and `predict()`. We print the same three rows and
columns as in @sec-model under each matrix:

```{r}
#| label: vcov-types
keep <- c("mu:(Intercept)", "mu:s(times).lin", "sigma:(Intercept)")
vcov(fit1, type = "bayesian")[keep, keep]
vcov(fit1, type = "frequentist")[keep, keep]
vcov(fit1, type = "unconditional")[keep, keep]
```

The first two matrices have a zero block between `mu` and `sigma`, as in
@sec-model. The third matrix does not. The smoothing parameters of the two
equations are estimated together, so a change of one of them moves the
coefficients of both equations, and the term $J V_\theta J^\top$ connects
them. The function `confint()` takes the same argument:

```{r}
#| label: confint-types
confint(fit1, parm = keep, type = "unconditional")
```

We compare the three standard errors of the fitted mean on a grid of 200
times:

```{r}
#| label: se-types
grid <- data.frame(times = seq(from = min(mcycle$times),
                               to = max(mcycle$times), length.out = 200))
pb <- predict(fit1, what = "mu", newdata = grid, se = TRUE, type = "bayesian")
pf <- predict(fit1, what = "mu", newdata = grid, se = TRUE,
              type = "frequentist")
pu <- predict(fit1, what = "mu", newdata = grid, se = TRUE,
              type = "unconditional")
head(pb, n = 3)
head(pf, n = 3)
head(pu, n = 3)
```

```{r}
#| label: se-numbers
#| include: false
rf <- pf$se / pb$se
ru <- pu$se / pb$se
t_ru <- grid$times[which.max(ru)]
```

The three predictions `fit` are the same, and only the standard errors
differ. @fig-se-types shows them over the whole time range. The frequentist
standard error is smaller than the Bayesian one at every time, by
`r pct(1 - median(rf))` in the median and by up to `r pct(1 - min(rf))`. In
the first 15 milliseconds the two are almost equal, because there the curve
is nearly flat and the penalty moves it very little. The unconditional
standard error is larger than the Bayesian one at every time, by
`r pct(median(ru) - 1)` in the median and by `r pct(max(ru) - 1)` at
`r fmt(t_ru, 1)` milliseconds. The two curves differ where the curve bends
most, because there the fitted mean depends most on the smoothing
parameter.

```{r}
#| label: fig-se-types
#| fig-cap: "Top: the standard error of the fitted mean of the second motorcycle model under the three variance matrices. Bottom: the frequentist and the unconditional standard errors as a percentage difference from the Bayesian one."
#| echo: false
#| fig-height: 5.2
se_types <- c("bayesian", "frequentist", "unconditional")
se_scales <- c("standard error (g)", "% from bayesian")
se_df <- data.frame(
  times = rep(grid$times, times = 6),
  value = c(pb$se, pf$se, pu$se,
            rep(0, times = nrow(grid)), 100 * (rf - 1), 100 * (ru - 1)),
  type = factor(rep(rep(se_types, each = nrow(grid)), times = 2),
                levels = se_types),
  scale = factor(rep(se_scales, each = 3 * nrow(grid)), levels = se_scales))
ggplot(se_df, aes(x = times, y = value, colour = type, linetype = type)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(facets = vars(scale), ncol = 1, scales = "free_y",
             strip.position = "left") +
  scale_colour_manual(values = c(chalk, grey, rust)) +
  scale_linetype_manual(values = c("solid", "dashed", "solid")) +
  labs(x = "time after impact (ms)", y = NULL) +
  theme(strip.placement = "outside", strip.background = element_blank())
```

Which of the three intervals covers the true curve at its nominal rate? We
answer with a small simulation study. The function `rstatmod()` of
@sec-model draws data from a model that we write, and when we do not give
the coefficients it draws them: here it draws a smooth function of $x$ from
the prior that the penalty defines, with a smoothing parameter drawn as
well. We draw one such function on 100 equally spaced points and 40 data
sets from it. We fit each data set and count how many of the 100 true means
each pointwise 95% interval contains.

```{r}
#| label: coverage
set.seed(2)
xd <- data.frame(x = seq(from = 0, to = 1, length.out = 100))
sim <- rstatmod(y ~ s(x), distrib = gaussian1_distrib(), data = xd,
                n_sim = 40)
cover <- matrix(NA, nrow = 40, ncol = 3,
                dimnames = list(NULL, c("bayesian", "frequentist",
                                        "unconditional")))
for (k in 1:40) {
  fit <- statmod(y ~ s(x), distrib = gaussian1_distrib(),
                 data = sim$data[[k]])
  truth <- sim$theta[[k]]$mu
  p_b <- predict(fit, what = "mu", newdata = xd, se = TRUE, type = "bayesian")
  p_f <- predict(fit, what = "mu", newdata = xd, se = TRUE,
                 type = "frequentist")
  p_u <- predict(fit, what = "mu", newdata = xd, se = TRUE,
                 type = "unconditional")
  cover[k, "bayesian"] <- mean(p_b$lower <= truth & truth <= p_b$upper)
  cover[k, "frequentist"] <- mean(p_f$lower <= truth & truth <= p_f$upper)
  cover[k, "unconditional"] <- mean(p_u$lower <= truth & truth <= p_u$upper)
}
colMeans(cover)
```

```{r}
#| label: coverage-numbers
#| include: false
cv <- colMeans(cover)
```

Averaged over the points and the data sets, the Bayesian intervals cover
the true mean `r pct(cv[["bayesian"]])` of the time, the frequentist ones
`r pct(cv[["frequentist"]])` and the unconditional ones
`r pct(cv[["unconditional"]])`, against a nominal 95%. The frequentist
intervals cover less because $V_f$ does not include the bias of the
penalized curve. This agrees with @marra2012, and it is the reason for the
default.

## Which likelihood {#sec-beyond-likelihood}

A model with random effects has two log-likelihoods. We use a classic data
set to explain them. The data frame `Orthodont` from the `nlme` package gives
the distance, in millimeters, from the pituitary to the pterygomaxillary
fissure of 27 children, measured every two years from age 8 to age 14
[@potthoff1964; @pinheiro2000]. We center the age at 11 years and give each
child its own intercept and its own slope in age:
$$
y_{ij} \sim \mathrm{N}(\mu_{ij}, \sigma), \qquad
\mu_{ij} = \beta_0 + \beta_1 a_{ij} + b_{0i} + b_{1i} a_{ij}, \qquad
(b_{0i}, b_{1i}) \sim \mathrm{N}(0, \Sigma_b),
$$
where $a_{ij}$ is the centered age of child $i$ at its $j$-th measurement.
We center the age inside the formula, with `I(age - 11)`. The term
`random(~ I(age - 11) | Subject)` gives each child the two random effects,
and by default their covariance matrix $\Sigma_b$ is unstructured, so the
two effects can be correlated.

```{r}
#| label: fit-ortho
data(Orthodont, package = "nlme")
reml1 <- statmod(distance ~ I(age - 11) + random(~ I(age - 11) | Subject),
                 distrib = gaussian1_distrib(), data = Orthodont)
summary(reml1)
```

`statmodels7` reads the random effects as coefficients with a penalty. The
penalty is minus the logarithm of the gaussian density of the random
effects, and its hyperparameters describe $\Sigma_b$. The summary reports
them as the two standard deviations and the correlation, estimated by REML.
The function `lme()` of the `nlme` package fits the same model:

```{r}
#| label: fit-lme
library(nlme)
lme1 <- lme(distance ~ I(age - 11), random = ~ I(age - 11) | Subject,
            data = Orthodont)
fixef(lme1)
VarCorr(lme1)
```

The two packages give the same estimates of the fixed effects, the same
standard deviations and the same correlation. The residual standard
deviation of `lme()` is $\exp(\hat\gamma_0)$, the exponential of the
intercept of `sigma`:

```{r}
#| label: sigma-ortho
exp(coef(reml1)$sigma)
```

The two log-likelihoods of the model differ in how they treat the random
effects $b$. The *conditional* log-likelihood is computed at the estimated
random effects $\hat b$:
$$
\ell_C = \log f(y \mid \hat b).
$$
It is the one that `logLik()` returns by default, and the one that cAIC uses.
The *marginal* log-likelihood integrates the random effects out:
$$
\ell_M = \log \int f(y \mid b)\, \pi(b)\, db ,
$$
where $\pi$ is the gaussian density of $b$. `statmodels7` computes the
integral by the Laplace method, which is exact for a gaussian response.
REML also integrates the unpenalized coefficients $\beta_0$ and $\beta_1$
[@patterson1971]. The argument `type = "marginal"` of `logLik()` returns the
criterion that the fit maximized: the REML criterion by default, and the
marginal log-likelihood when the fit used `outer_criterion = ml()`.

```{r}
#| label: loglik-ortho
logLik(reml1)
logLik(reml1, type = "marginal")
logLik(lme1)
```

The marginal value is the restricted log-likelihood that `lme()` reports.
Its attribute `df` counts the two fixed effects, the three parameters of
$\Sigma_b$ and $\sigma$. The conditional value is larger, because $f(y \mid
\hat b)$ uses the random effects of each child, and its `df` is the total
effective degrees of freedom of @sec-model. The two values answer different
questions. The marginal log-likelihood describes a new child drawn from the
population, and the conditional one describes the 27 children that we
observed [@vaida2005].

## Testing one coefficient {#sec-beyond-tests}

The column `z` of the summary is a Wald statistic, which tests that a
coefficient is zero. Three other statistics test a coefficient $\beta_j$
against any value $\beta_{j0}$. We write $\ell$ for the log-likelihood,
$\hat\theta$ for the estimate of all the coefficients, and $\tilde\theta$
for the estimate when $\beta_j$ is held at $\beta_{j0}$. Let $U(\theta)$ be
the gradient of $\ell$ and $I(\theta)$ the Fisher information. The four
statistics are
$$
\begin{aligned}
W &= (\hat\beta_j - \beta_{j0})^2 / \mathrm{se}(\hat\beta_j)^2
  && \text{(Wald)}, \\
L &= 2\{\ell(\hat\theta) - \ell(\tilde\theta)\}
  && \text{(likelihood ratio)}, \\
R &= U(\tilde\theta)^\top I(\tilde\theta)^{-1} U(\tilde\theta)
  && \text{(Rao score)}, \\
G &= U_j(\tilde\theta)\, (\hat\beta_j - \beta_{j0})
  && \text{(Terrell gradient)}.
\end{aligned}
$$
If $\beta_j = \beta_{j0}$, each of them has approximately a $\chi^2$
distribution with one degree of freedom [@coxhinkley1974; @terrell2002].
$W$ needs only the fit that we have. The other three need the model fitted
again with $\beta_j$ held at $\beta_{j0}$. For a penalized model the three
use the penalized log-likelihood, with the smoothing parameters held at
their estimates.

We use a nonlinear model. The data frame `Puromycin` gives the initial rate
of an enzymatic reaction, in counts per minute per minute, against the
concentration of the substrate, in parts per million [@bates1988]. We use
the `r sum(Puromycin$state == "treated")` measurements on cells treated with
the antibiotic puromycin. The Michaelis–Menten model describes the rate as
$$
y_i \sim \mathrm{N}(\mu_i, \sigma), \qquad
\mu_i = \frac{V_m\, c_i}{K + c_i},
$$
where $c_i$ is the concentration. The parameter $V_m$ is the largest rate,
and $K$ is the concentration at which the rate reaches half of $V_m$. The
term `nl()` takes the formula of the mean as a function of its parameters,
and a later chapter describes it in full. The formula of the model starts
with `0` because the curve has no intercept.

```{r}
#| label: fit-puro
puro <- Puromycin[Puromycin$state == "treated", ]
fit_p <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc)),
                 distrib = gaussian1_distrib(), data = puro)
coef(fit_p)
```

The function `statmod_test()` computes one of the four statistics for one
coefficient. Its arguments name the parameter of the distribution, the
coefficient, the value under the null hypothesis and the statistic. We test
$K = 0.05$ with each of the four:

```{r}
#| label: test-puro
statmod_test(fit_p, param = "mu", coefname = "nl.K", value = 0.05,
             test = "wald")
statmod_test(fit_p, param = "mu", coefname = "nl.K", value = 0.05,
             test = "lr")
statmod_test(fit_p, param = "mu", coefname = "nl.K", value = 0.05,
             test = "score")
statmod_test(fit_p, param = "mu", coefname = "nl.K", value = 0.05,
             test = "gradient")
```

```{r}
#| label: test-numbers
#| include: false
t_w <- statmod_test(fit_p, param = "mu", coefname = "nl.K", value = 0.05,
                    test = "wald")
t_l <- statmod_test(fit_p, param = "mu", coefname = "nl.K", value = 0.05,
                    test = "lr")
t_s <- statmod_test(fit_p, param = "mu", coefname = "nl.K", value = 0.05,
                    test = "score")
t_g <- statmod_test(fit_p, param = "mu", coefname = "nl.K", value = 0.05,
                    test = "gradient")
```

The four statistics range from `r fmt(t_s@statistic)` for the score test to
`r fmt(t_l@statistic)` for the likelihood ratio, with p-values between
`r fmt(t_l@p.value, 3)` and `r fmt(t_s@p.value, 3)`. With
`r nrow(puro)` observations they do not agree closely, and none of them
rejects $K = 0.05$ at the 5% level.

We can compute two of them by hand. The Wald statistic comes from the
estimate and the standard error that `confint()` reports:

```{r}
#| label: wald-by-hand
ci_k <- confint(fit_p, parm = "mu:nl.K")
(ci_k$estimate - 0.05)^2 / ci_k$se^2
```

For the likelihood ratio, `statmod_test()` compares the maximum
log-likelihood of the full model with the maximum log-likelihood of the
model with $K$ held at 0.05, with $\sigma$ estimated by maximum likelihood
in both. The function `nls()` fits the full model by least squares. When
$K$ is held at 0.05 the mean is $V_m$ times the known variable
$c_i/(0.05 + c_i)$, so `lm()` without an intercept fits the restricted
model. The function `logLik()` of both returns the maximum log-likelihood:

```{r}
#| label: lr-by-hand
nls_p <- nls(rate ~ Vm * conc / (K + conc), data = puro,
             start = list(Vm = 200, K = 0.05))
puro$x05 <- puro$conc / (0.05 + puro$conc)
lm_05 <- lm(rate ~ 0 + x05, data = puro)
2 * (as.numeric(logLik(nls_p)) - as.numeric(logLik(lm_05)))
```

The argument `test` of `summary()` replaces the Wald statistic of every
row with another of the four statistics, each testing that its coefficient
is zero. The column then holds the signed square root of the statistic,
which has approximately a standard gaussian distribution, and its heading is
`r` instead of `z`:

```{r}
#| label: summary-lr
summary(fit_p, test = "lr")
```

Each row needs one restricted fit, so `summary(test = "lr")` costs a few
more fits than `summary()`. Only the statistic and the p-value change. The
interval of each row is still the Wald interval, and the next section
explains how to replace it.

## Intervals by inversion {#sec-beyond-invert}

A 95% confidence interval for $\beta_j$ contains the values $\beta_{j0}$
that a test at the 5% level does not reject. The Wald interval
$\hat\beta_j \pm 1.96\,\mathrm{se}$ is the set of values where
$W \le \chi^2_{1,0.95} = 3.84$, and it is symmetric around the estimate. We
obtain an interval from each of the other three statistics in the same way,
by solving for the two values of $\beta_{j0}$ where the statistic equals
3.84. These intervals can be asymmetric. We choose the statistic with the
argument `test` of `confint()`.

We also set the argument `readable` to `FALSE`. Some model terms estimate
their parameters on working coordinates, which keep the estimation stable
but are hard to interpret. For example, the score-driven filter `gas()` of a
later chapter estimates the persistence of the filter through a partial
autocorrelation, and `confint()` reports by default the autoregressive
coefficient computed from it. This default is `readable = TRUE`. A test holds
one working coordinate at a value, so an inverted interval exists only for
the working coordinates, and `confint()` with `test` other than `"wald"`
requires `readable = FALSE`. In the Michaelis–Menten model the coefficients
are $V_m$ and $K$ themselves, so this argument does not change the result
here:

```{r}
#| label: invert-puro
confint(fit_p, parm = "mu:nl.K")
confint(fit_p, parm = "mu:nl.K", test = "lr", readable = FALSE)
confint(fit_p, parm = "mu:nl.K", test = "score", readable = FALSE)
confint(fit_p, parm = "mu:nl.K", test = "gradient", readable = FALSE)
```

```{r}
#| label: invert-numbers
#| include: false
ciw <- confint(fit_p, parm = "mu:nl.K")
cil <- confint(fit_p, parm = "mu:nl.K", test = "lr", readable = FALSE)
k_hat <- ciw$estimate
```

The likelihood ratio interval runs from `r fmt(cil$lower, 4)` to
`r fmt(cil$upper, 4)`. It extends `r fmt(k_hat - cil$lower, 4)` below the
estimate $\hat K = `r fmt(k_hat, 4)`$ and `r fmt(cil$upper - k_hat, 4)`
above it, while the Wald interval extends `r fmt(k_hat - ciw$lower, 4)` on
both sides. The Wald statistic is a quadratic function of $K$, so its
interval is symmetric. The likelihood ratio statistic is not, because the
log-likelihood of this model is not a quadratic function of $K$: it rises
more steeply below $\hat K$ than above it.

The function `confint()` of `nls()` gives another interval:

```{r}
#| label: confint-nls
confint(nls_p)
```

```{r}
#| label: nls-numbers
#| include: false
ci_nls <- suppressMessages(confint(nls_p))["K", ]
```

It runs from `r fmt(ci_nls[1], 4)` to `r fmt(ci_nls[2], 4)`, so it is wider.
`nls()` uses the profile $t$ statistic [@bates1988]
$$
\tau(K_0) = \mathrm{sign}(\hat K - K_0)\,
\frac{\sqrt{S(K_0) - S(\hat K)}}{s},
$$
where $S(K_0)$ is the residual sum of squares with $K$ held at $K_0$ and
$V_m$ estimated, and $s^2 = S(\hat K)/(n - 2)$. For a gaussian model the
likelihood ratio statistic is $L = n \log\{S(K_0)/S(\hat K)\}$, so
$L = n\log\{1 + \tau^2/(n-2)\}$. The two statistics are functions of each
other, and the two intervals come from the same test. They differ in the
reference distribution. `nls()` compares $\tau$ with the quantiles of the $t$
distribution with $n - 2$ degrees of freedom, which is exact for a linear
model. `statmodels7` compares $L$ with the quantile 3.84 of the $\chi^2_1$
distribution, which corresponds to a smaller cutoff for $|\tau|$:

```{r}
#| label: t-cutoff
n_p <- nrow(puro)
qt(p = 0.975, df = n_p - 2)
sqrt((n_p - 2) * (exp(qchisq(p = 0.95, df = 1) / n_p) - 1))
```

The $\chi^2_1$ distribution is the limit of the distribution of $L$ as $n$
grows. With `r nrow(puro)` observations the limit is optimistic, and the
interval of `nls()` is the better one.

## Comparing two models {#sec-beyond-lr}

The tests of the previous sections hold one coefficient at a value. To
compare two nested models we use the likelihood ratio of the two fits. We
return to the `Orthodont` data and ask whether boys and girls have
different mean distances. The second model adds the sex of the child:

```{r}
#| label: fit-ortho-ml
ml1 <- statmod(distance ~ I(age - 11) + random(~ I(age - 11) | Subject),
               distrib = gaussian1_distrib(), data = Orthodont,
               outer_criterion = ml())
ml2 <- statmod(distance ~ I(age - 11) + Sex + random(~ I(age - 11) | Subject),
               distrib = gaussian1_distrib(), data = Orthodont,
               outer_criterion = ml())
```

Both fits use `outer_criterion = ml()`, so they maximize the marginal
log-likelihood. The likelihood ratio statistic is twice the difference of
the two maximized marginal log-likelihoods, and under the smaller model it
has approximately a $\chi^2$ distribution with one degree of freedom:

```{r}
#| label: lr-ortho
l1 <- logLik(ml1, type = "marginal")
l2 <- logLik(ml2, type = "marginal")
lr <- 2 * (as.numeric(l2) - as.numeric(l1))
lr
pchisq(lr, df = 1, lower.tail = FALSE)
```

The function `anova()` of `nlme` gives the same statistic and p-value for
the two models fitted by maximum likelihood:

```{r}
#| label: lr-lme
lme1_ml <- lme(distance ~ I(age - 11), random = ~ I(age - 11) | Subject,
               data = Orthodont, method = "ML")
lme2_ml <- lme(distance ~ I(age - 11) + Sex, random = ~ I(age - 11) | Subject,
               data = Orthodont, method = "ML")
an <- anova(lme1_ml, lme2_ml)
an$logLik
an$L.Ratio
an[["p-value"]]
```

We do not compare the two models with REML. The REML criterion is the
likelihood of the residuals of the unpenalized part of the mean, that is, of
the combinations of $y$ that do not depend on the unpenalized coefficients.
When we add `Sex` to the model, these combinations change. So the two REML
criteria describe two different functions of the data, and their difference
is not a likelihood ratio statistic [@pinheiro2000, sec. 2.4]:

```{r}
#| label: reml-diff
reml2 <- statmod(distance ~ I(age - 11) + Sex + random(~ I(age - 11) | Subject),
                 distrib = gaussian1_distrib(), data = Orthodont)
2 * (as.numeric(logLik(reml2, type = "marginal")) -
       as.numeric(logLik(reml1, type = "marginal")))
```

We use `reml()` to estimate the variances of one model, and `ml()` to
compare models that differ in their unpenalized terms.

The test of @sec-beyond-tests gives a third answer for the coefficient of
`Sex`:

```{r}
#| label: test-sex
statmod_test(reml2, param = "mu", coefname = "SexFemale", value = 0,
             test = "lr")
```

```{r}
#| label: sex-numbers
#| include: false
t_sex <- statmod_test(reml2, param = "mu", coefname = "SexFemale", value = 0,
                      test = "lr")
```

This statistic is `r fmt(t_sex@statistic)`, against `r fmt(lr)` from the
marginal log-likelihoods. The two tests treat the covariance matrix
$\Sigma_b$ of the random effects in different ways. In the restricted fit
of `statmod_test()`, $\Sigma_b$ stays at the estimate of the full model
`reml2`, and only the coefficients and $\sigma$ are estimated again. In the
comparison of the marginal log-likelihoods, each model estimates its own
$\Sigma_b$. Sex does not change within a child, so without `Sex` the
difference between boys and girls becomes a difference between children,
and the variance of the random intercepts grows to absorb it. The fits of
`lme()` by maximum likelihood show this:

```{r}
#| label: varcorr-ml
VarCorr(lme1_ml)
VarCorr(lme2_ml)
```

The standard deviation of the random intercepts is larger in the model
without `Sex`. This model explains part of the effect of sex with the
random intercepts, so its marginal log-likelihood is closer to that of the
full model, and the likelihood ratio is smaller.

## Predictions with random effects {#sec-beyond-random}

A model with random effects raises one more question about predictions:
what to predict for a group that the data do not contain. The data frame
`epil` from the `MASS` package gives the number of epileptic seizures of 59
patients in four periods of two weeks, in a trial of the drug progabide
[@thall1990]. The covariates are the treatment `trt`, the logarithm of the
number of seizures in the eight weeks before the trial (`lbase`, centered),
the logarithm of the age (`lage`, centered) and an indicator of the fourth
period (`V4`). We give each patient a random intercept on the log scale:
$$
y_{ij} \sim \mathrm{Poisson}(\mu_{ij}), \qquad
\log\mu_{ij} = x_{ij}^\top\beta + b_i, \qquad b_i \sim \mathrm{N}(0, \tau).
$$

```{r}
#| label: fit-epil
data(epil, package = "MASS")
fit_e <- statmod(y ~ lbase * trt + lage + V4 + random(~ 1 | subject),
                 distrib = poisson_distrib(), data = epil)
hyper(fit_e)
```

The hyperparameter of the random intercept is its standard deviation
$\tau$. For a patient in the data, `predict()` uses the estimated random
intercept of that patient. These are the first four rows of the data,
which belong to the first patient:

```{r}
#| label: predict-known
predict(fit_e, what = "mu", newdata = epil[1:4, ], se = TRUE)
```

For a new patient $b$ is unknown. We describe a new patient on placebo with
average baseline and age, and we give it the number 0. The patients of the
data are numbered from 1 to 59, so the fit has no random intercept for this
patient. By default `predict()` signals an error:

```{r}
#| label: predict-new-error
#| error: true
new <- data.frame(lbase = 0, trt = factor("placebo", levels(epil$trt)),
                  lage = 0, V4 = 0, subject = 0)
predict(fit_e, what = "mu", newdata = new)
```

We choose how to treat $b$ with the argument `random` of `predict()`. With
`random = "zero"` we set $b = 0$, and the prediction is
$\exp(x^\top\hat\beta)$, the mean number of seizures of a typical patient.
With `random = "marginal"` we average over the distribution of $b$, and the
prediction is the mean over all patients with these covariates:
$$
\mathbb{E}\{\exp(x^\top\beta + b)\} = \exp(x^\top\beta + \tau^2/2) .
$$
The two predictions differ because the link function is not linear
[@zeger1988].

```{r}
#| label: predict-zero-marginal
pz <- predict(fit_e, what = "mu", newdata = new, se = TRUE, random = "zero")
pz
pm <- predict(fit_e, what = "mu", newdata = new, se = TRUE,
              random = "marginal")
pm
tau <- hyper(fit_e)$estimate
pz$fit * exp(tau^2 / 2)
```

With the argument `interval` we choose which uncertainty the interval
describes. With `interval = "confidence"`, the default, it describes the
uncertainty of the estimates. With `interval = "group"` it also includes the
variation of $b$, so it covers the mean $\exp(x^\top\beta + b)$ of a new
patient. On the log scale this interval is
$x^\top\hat\beta \pm 1.96\sqrt{\mathrm{se}^2 + \hat\tau^2}$, where se is the
standard error of $x^\top\hat\beta$.

With `what = "response"` and `interval = "prediction"` the interval
describes the number of seizures $Y$ that a new patient has in one period.
Three things make $Y$ uncertain: the estimate of $x^\top\beta$, the random
intercept $b$ of the patient, and the Poisson variation of $Y$ around its
mean. The predictor $u = x^\top\beta + b$ of the new patient is gaussian,
with mean $x^\top\hat\beta$ and variance $\mathrm{se}^2 + \hat\tau^2$, and
given $u$ the count has a Poisson distribution with mean $e^u$. So
$$
P(Y \le k) = \int \mathrm{Poisson}(k;\, e^u)\,
\phi\big(u;\, x^\top\hat\beta,\, \mathrm{se}^2 + \hat\tau^2\big)\, du ,
$$
where $\mathrm{Poisson}(k; m)$ is the Poisson distribution function with
mean $m$ and $\phi(u; a, v)$ is the gaussian density with mean $a$ and
variance $v$. The interval runs from the 2.5% quantile to the 97.5%
quantile of this distribution, and the prediction `fit` is its median.

```{r}
#| label: predict-intervals
pg <- predict(fit_e, what = "mu", newdata = new, se = TRUE, random = "zero",
              interval = "group")
pg
pr <- predict(fit_e, what = "response", newdata = new, se = TRUE,
              random = "zero", interval = "prediction")
pr
```

The standard deviation of the random intercepts is $\hat\tau =
`r fmt(tau)`$. For a patient on placebo with average baseline and age, the
typical patient has a mean of `r fmt(pz$fit)` seizures per period, with a
95% interval from `r fmt(pz$lower)` to `r fmt(pz$upper)`. The mean over all
such patients is larger, `r fmt(pm$fit)`, by the factor
$\exp(\hat\tau^2/2) = `r fmt(exp(tau^2 / 2), 3)`$. The mean of one new
patient lies between `r fmt(pg$lower)` and `r fmt(pg$upper)` with
probability 95%, and the number of seizures of that patient in one period
lies between `r pr$lower` and `r pr$upper`, with a median of `r pr$fit`.
@fig-epil shows the three intervals as functions of the baseline count. All
three are conditional on the estimated variance $\hat\tau^2$.

```{r}
#| label: fig-epil
#| fig-cap: "Seizures per two weeks for a new patient on placebo, against the centered logarithm of the baseline count."
#| echo: false
#| fig-height: 4.2
#| fig-width: 9
lb <- seq(min(epil$lbase), max(epil$lbase), length.out = 60)
gnew <- new[rep(1, 60), ]; gnew$lbase <- lb
gz <- predict(fit_e, what = "mu", newdata = gnew, se = TRUE, random = "zero")
gg <- predict(fit_e, what = "mu", newdata = gnew, se = TRUE, random = "zero",
              interval = "group")
gr <- predict(fit_e, what = "response", newdata = gnew, se = TRUE,
              random = "zero", interval = "prediction")
pl <- epil$trt == "placebo"
ep_bands <- data.frame(
  lbase = rep(lb, times = 2),
  lower = c(gz$lower, gg$lower),
  upper = c(gz$upper, gg$upper),
  band = rep(c("confidence interval", "group interval"),
             each = length(lb)))
ep_mean <- data.frame(lbase = lb, value = gz$fit,
                      curve = "mean of a typical patient")
ep_pred <- data.frame(lbase = rep(lb, times = 2),
                      value = c(gr$lower, gr$upper),
                      end = rep(c("lower", "upper"), each = length(lb)),
                      curve = "prediction interval")
ggplot(mapping = aes(x = lbase)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = band), data = ep_bands) +
  geom_line(aes(y = value, colour = curve, linetype = curve), data = ep_mean,
            linewidth = 0.8) +
  geom_step(aes(y = value, colour = curve, linetype = curve, group = end),
            data = ep_pred) +
  geom_point(aes(y = y, shape = "patients on placebo"), data = epil[pl, ],
             colour = book_grey, size = 0.9) +
  scale_shape_manual(values = 16) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_fill_manual(values = c(adjustcolor(chalk, 0.4),
                               adjustcolor(chalk, 0.15))) +
  coord_cartesian(ylim = c(0, max(gr$upper))) +
  guides(shape = guide_legend(order = 1), colour = guide_legend(order = 2),
         linetype = guide_legend(order = 2), fill = guide_legend(order = 3)) +
  labs(x = "centered log baseline count", y = "seizures per period")
```

## Starting from another fit {#sec-beyond-start}

`statmod()` starts from the maximum likelihood estimates of a model with
intercepts only, and then fits the full model. When we have already fitted a
similar model, we can start from its estimates with the argument `start`
and the function `start_from()`. Here we fit the second motorcycle model
again, starting from the first one:

```{r}
#| label: start-from
fit1_s <- statmod(accel ~ s(times, bspline_smooth(k = 20)) |
                    sigma ~ s(times, bspline_smooth(k = 10)),
                  distrib = gaussian1_distrib(), data = mcycle,
                  start = start_from(fit0))
fit1_s
fit1
```

`start_from()` takes the coefficients that the two models share. The
parametric coefficients are matched by name. A smooth is matched when the two
models write it in the same way; when the two bases differ, `start_from()`
projects the fitted curve of the old model onto the new basis. A coefficient
that the old model does not have starts from the intercept-only fit, and
here this is the case of the smooth of `sigma`.

```{r}
#| label: start-numbers
#| include: false
n_ev1 <- nrow(fit1@history$outer)
n_ev1s <- nrow(fit1_s@history$outer)
```

The two fits end at the same point, up to the precision of the search:

```{r}
#| label: start-same
logLik(fit1)
logLik(fit1_s)
max(abs(fitted(fit1) - fitted(fit1_s)))
```

The difference between the two fits is
$`r d_crit <- abs(fit1@criterion - fit1_s@criterion); if (d_crit == 0) "0" else sci(d_crit)`$
in the REML criterion,
`r fmt(abs(as.numeric(logLik(fit1)) - as.numeric(logLik(fit1_s))), 3)` in
the log-likelihood and at most
`r fmt(max(abs(fitted(fit1) - fitted(fit1_s))), 3)` $g$ in the fitted
means. The fit that
starts from `fit0` evaluates the REML criterion at `r n_ev1s` values of the
smoothing parameters, against `r n_ev1` for the fit that starts from the
intercepts. The saving is largest when the models are large and the start is
close.

## The outer criterion {#sec-beyond-outer}

In @sec-model the REML criterion estimated the smoothing parameter. The
argument `outer_criterion` of `statmod()` chooses this criterion. Its
default is `reml()`, and `ml()` maximizes the marginal log-likelihood of
@sec-beyond-likelihood instead. REML integrates both the penalized and the
unpenalized coefficients of the equations, and ML integrates only the
penalized ones.

The criterion also decides where the other parameters of the distribution
are estimated. The argument `marginal` of `reml()` and `ml()` names the
parameters whose unpenalized coefficients the criterion estimates. By
default it names every parameter except the first one, the location. So in
`fit0` the criterion estimated both the smoothing parameter and the
intercept of $\log\sigma$. With `marginal = "none"` the criterion estimates
only the smoothing parameter, and $\sigma$ is estimated with the mean, by
maximizing the penalized log-likelihood.

```{r}
#| label: outer-fits
fit0_none <- statmod(accel ~ s(times, bspline_smooth(k = 20)),
                     distrib = gaussian1_distrib(), data = mcycle,
                     outer_criterion = reml(marginal = "none"))
fit0_ml <- statmod(accel ~ s(times, bspline_smooth(k = 20)),
                   distrib = gaussian1_distrib(), data = mcycle,
                   outer_criterion = ml())
exp(coef(fit0)$sigma)
exp(coef(fit0_none)$sigma)
exp(coef(fit0_ml)$sigma)
hyper(fit0)$estimate
hyper(fit0_none)$estimate
hyper(fit0_ml)$estimate
```

The three smoothing parameters are close, and the three estimates of
$\sigma$ differ. For a gaussian response each estimate of $\sigma^2$ is the
residual sum of squares divided by $n$ minus the number of degrees of
freedom that the criterion counts as spent on the mean. In a linear model
without penalty the same thing happens: the maximum likelihood estimate
divides by $n$, and the REML estimate divides by $n - p$, because REML
integrates the $p$ coefficients of the mean.

Here the coefficients of the mean are of two kinds. Let $\tau_\mu$ be the
effective degrees of freedom of the equation of the mean. Two of them belong
to the unpenalized coefficients, the intercept and the linear part of the
smooth, and the other $\tau_\mu - 2$ to the penalized ones. REML integrates
both kinds, so it divides by $n - \tau_\mu$. ML integrates only the
penalized coefficients, so it divides by $n - (\tau_\mu - 2)$. The joint
mode integrates nothing and divides by $n$:

```{r}
#| label: outer-by-hand
n <- nrow(mcycle)
rss <- sum((mcycle$accel - fitted(fit0))^2)
tau_mu <- sum(fit0@edf$edf[fit0@edf$parameter == "mu"])
tau_mu
sqrt(rss / (n - tau_mu))
rss_ml <- sum((mcycle$accel - fitted(fit0_ml))^2)
tau_ml <- sum(fit0_ml@edf$edf[fit0_ml@edf$parameter == "mu"])
sqrt(rss_ml / (n - (tau_ml - 2)))
rss_none <- sum((mcycle$accel - fitted(fit0_none))^2)
sqrt(rss_none / n)
```

A criterion that integrates a coefficient accounts for the fact that the
coefficient was estimated from the same data. The joint mode does not, so
its estimate of $\sigma$ is the smallest of the three and is biased
downwards. REML is the default because it removes this bias for all the
coefficients of the mean.

We can also hold a hyperparameter at a value that we choose. We write the
value in the term with the argument `hyper`:

```{r}
#| label: held
fit0_held <- statmod(accel ~ s(times, bspline_smooth(k = 20),
                               hyper = c(lambda = 1e-3)),
                     distrib = gaussian1_distrib(), data = mcycle)
hyper(fit0_held)
fit0_held@edf
fit0@edf
```

The column `held` of `hyper()` is `TRUE`, and the column `source` is
`fixed`. The smoothing parameter 0.001 is `r fmt(1e-3 / hyper(fit0)$estimate, 0)`
times the REML estimate, so the smooth is stiffer and uses
`r fmt(fit0_held@edf$edf[2])` effective degrees of freedom instead of
`r fmt(fit0@edf$edf[2])`. @fig-held shows the two fitted curves. The
criterion still runs, because it still estimates $\sigma$.

```{r}
#| label: fig-held
#| fig-cap: "The fitted mean of the first motorcycle model with the smoothing parameter estimated by REML and held at 0.001."
#| echo: false
#| fig-height: 3.8
ph <- predict(fit0_held, what = "mu", newdata = grid)
p0 <- predict(fit0, what = "mu", newdata = grid)
fits <- c("lambda estimated by REML", "lambda held at 0.001")
held <- data.frame(times = rep(grid$times, times = 2), fit = c(p0, ph),
                   curve = factor(rep(fits, each = nrow(grid)),
                                  levels = fits))
ggplot(mcycle, aes(x = times)) +
  geom_point(aes(y = accel, shape = "data"), colour = book_grey, size = 1.1) +
  geom_line(aes(y = fit, colour = curve, linetype = curve), data = held,
            linewidth = 0.8) +
  scale_shape_manual(values = 16) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  guides(shape = guide_legend(order = 1), colour = guide_legend(order = 2),
         linetype = guide_legend(order = 2)) +
  labs(x = "time after impact (ms)", y = "acceleration (g)")
```

## Weights, offsets and known parameters {#sec-beyond-weights}

### Prior weights

The argument `weights` of `statmod()` multiplies the contribution of each
observation to the log-likelihood:
$$
\ell = \sum_i w_i \log f(y_i; \theta_i) .
$$
These are *frequency weights*: an observation with $w_i = 3$ counts as three
identical observations. We give each observation of the motorcycle data a
weight of 1, 2 or 3, fit a straight line with these weights, and fit it
again to the data with each row repeated $w_i$ times:

```{r}
#| label: weights
set.seed(1)
w <- sample(1:3, size = nrow(mcycle), replace = TRUE)
big <- mcycle[rep(seq_len(nrow(mcycle)), times = w), ]
fit_w <- statmod(accel ~ times, distrib = gaussian1_distrib(), data = mcycle,
                 weights = w)
fit_big <- statmod(accel ~ times, distrib = gaussian1_distrib(), data = big)
coef(fit_w)
coef(fit_big)
```

The two fits are the same. The function `lm()` also takes weights, but it
reads them in another way, as *precision weights*: each observation is seen
once, and its variance is $\sigma^2/w_i$. The estimate of the mean is the
same, because in both cases it minimizes $\sum_i w_i (y_i - \mu_i)^2$. The
estimate of $\sigma$ is not:

```{r}
#| label: weights-lm
lm_w <- lm(accel ~ times, data = mcycle, weights = w)
coef(lm_w)
summary(lm_w)$sigma
exp(coef(fit_w)$sigma)
```

With frequency weights the number of observations is $\sum_i w_i =
`r sum(w)`$, and with precision weights it is $n = `r nrow(mcycle)`$. So the
two estimates of $\sigma$ differ, and so do the standard errors.
`statmodels7` uses frequency weights because their definition holds for
every distribution, while a precision weight has a meaning only for a
distribution with a parameter tied to the variance. In a model with an
equation for $\sigma$, a precision weight is a part of that equation, so we
write it there. If the variance of observation $i$ is $\sigma^2/w_i$, then
$\log\sigma_i = \gamma_0 - \tfrac{1}{2}\log w_i$, and the known term $-\tfrac{1}{2}\log w_i$ is an
*offset*, a term with a coefficient fixed at 1. We write it with
`offset()`:

```{r}
#| label: weights-precision
fit_prec <- statmod(accel ~ times | sigma ~ offset(-0.5 * log(w)),
                    distrib = gaussian1_distrib(), data = mcycle)
exp(coef(fit_prec)$sigma)
confint(fit_prec, parm = c("mu:(Intercept)", "mu:times"))
sqrt(diag(vcov(lm_w)))
```

This model is the one that `lm()` fits: the same $\sigma$ and the same
standard errors.

### Offsets

An offset is common in a model for counts, when the counts come from
different amounts of exposure. The data frame `Insurance` from the `MASS`
package gives the number of claims on car insurance policies in 64 groups
of policyholders, defined by the district, the size of the car and the age
of the driver [@venables2002]. The column `Holders` is the number of
policyholders in each group. The expected number of claims is the number of
holders times the rate of claims per holder, so we write
$$
\log\mu_i = \log(\text{holders}_i) + x_i^\top\beta ,
$$
and the coefficients describe the logarithm of the rate:

```{r}
#| label: offset
data(Insurance, package = "MASS")
fit_i <- statmod(Claims ~ offset(log(Holders)) + District + Group + Age,
                 distrib = poisson_distrib(), data = Insurance)
coef(fit_i)$mu
glm_i <- glm(Claims ~ offset(log(Holders)) + District + Group + Age,
             family = poisson, data = Insurance)
coef(glm_i)
```

The coefficients are those of `glm()`. The factors `Group` and `Age` are
ordered, so their coefficients are polynomial contrasts. The coefficient
`Group.L`, for example, is the linear trend of the log-rate across the four
sizes of the car.

### A parameter with a known value

Sometimes a parameter of the distribution has a value that we know, or a
value that we want to fix. The function `fixed()` takes a distribution and
holds some of its parameters at given values. The Student $t$ distribution
`student_t1_distrib()` has three parameters, the location `mu`, the scale
`sigma` and the degrees of freedom `nu`. With `nu` fixed at 4 we get a
distribution with heavier tails than the gaussian and two free parameters:

```{r}
#| label: fixed-distrib
fixed(student_t1_distrib(), nu = 4)
```

We use it in place of the gaussian in the second motorcycle model:

```{r}
#| label: fixed-fit
fit_t4 <- statmod(accel ~ s(times, bspline_smooth(k = 20)) |
                    sigma ~ s(times, bspline_smooth(k = 10)),
                  distrib = fixed(student_t1_distrib(), nu = 4),
                  data = mcycle)
fit_t4
AIC(fit1)
AIC(fit_t4)
```

The model with $t$ errors has the same equations as `fit1`, and it has no
equation for `nu`, because `nu` is not a parameter of the fixed
distribution. Its cAIC is higher than the cAIC of the gaussian model, so the
motorcycle data do not need tails heavier than the gaussian ones once the
standard deviation changes with time.

```{r}
#| label: gate-02
#| include: false
assert_beyond_ok(
  fits = list(fit0 = fit0, fit1 = fit1, reml1 = reml1, reml2 = reml2,
              ml1 = ml1, ml2 = ml2, puro = fit_p, epil = fit_e,
              fit1_s = fit1_s, fit0_none = fit0_none, fit0_ml = fit0_ml,
              fit0_held = fit0_held, fit_w = fit_w, fit_big = fit_big,
              fit_prec = fit_prec, fit_i = fit_i, fit_t4 = fit_t4),
  ref = list(lme1 = lme1, lme1_ml = lme1_ml, lme2_ml = lme2_ml,
             nls_p = nls_p, lm_w = lm_w, glm_i = glm_i),
  data = list(mcycle = mcycle, puro = puro, new = new, w = w),
  claims = list(pb = pb, pf = pf, pu = pu, cv = cv, keep = keep,
                ciw = ciw, cil = cil, lr = lr,
                t_sex = t_sex, pz = pz, pm = pm, pg = pg, pr = pr,
                tests = list(wald = t_w, lr = t_l, score = t_s,
                             gradient = t_g)))
```
