# The linkfunctions7 package {#sec-links}

```{r}
#| label: setup-02
#| include: false
source("R/_setup.R")
source("R/link-formulas.R")
```

The parameters of a distribution are usually constrained: a scale is positive,
a probability lies between zero and one, a correlation between minus one and
one. An optimizer, on the other hand, works best with an argument it can move
freely in any direction. A link function connects the two: it maps the
constrained domain of a parameter onto an unconstrained scale, and its inverse
maps the optimizer's iterates back into the domain where the distribution is
defined. This chapter gives the precise definition, derives the relationship
between the derivatives of a link and those of its inverse, explains why the
toolkit carries both to fourth order, and ends with the catalog of the
sixteen link classes that `linkfunctions7` provides.

## Definition {#sec-link-def}

::: {#def-link}
## Link function

Let $\Theta \subseteq \mathbb{R}$ be an open interval. A **link function** on
$\Theta$ is a map $g : \Theta \to \mathbb{R}$ which is a bijection onto its
image $E = g(\Theta)$, itself an open interval; which is four times continuously
differentiable on $\Theta$; and which is *regular*, meaning that
$g'(\theta) \neq 0$ everywhere. We write $h = g^{-1} : E \to \Theta$ for the
inverse, and throughout the book $\eta = g(\theta)$ and $\theta = h(\eta)$.
:::

Three observations about this definition are useful before going on.

First, the definition does not require $g$ to be increasing. Since $g'$ is
continuous and never zero on an interval, $g$ is automatically strictly
monotone, but the direction can be either, and both directions occur in
practice: the link for a parameter bounded above, $g(\theta) = \log(u -
\theta)$, is strictly decreasing. Code that assumes an increasing link, for
example when mapping the endpoints of an interval through it, fails on this
case. The property to require and to test is therefore strict monotonicity of
either sign.

Second, the image $E$ need not be the whole real line. For the square-root
link $E = (0, \infty)$, and the same holds for the reciprocal link. In these
cases the optimizer is not fully unconstrained, because the constraint has
moved from $\theta$ to $\eta$, and an optimizer can still step outside the
domain and receive `NaN`. For this reason the catalog below states $E$
explicitly for every link.

Third, the definition requires four continuous derivatives instead of the
customary one or two. The link objects in `stats::make.link()` carry the first derivative of the
inverse and nothing else, because a generalized linear model fitted by
iteratively reweighted least squares needs nothing more. This is the setting
of @nelder1972 and @mccullagh1989, where the link function first appears
under that name. This toolkit needs more: as @sec-link-scale shows,
transporting a fourth-order derivative of a log-likelihood onto the
unconstrained scale requires $h'$, $h''$, $h'''$ and $h''''$, so a link that
stopped at second order would restrict the entire toolkit to second-order
methods.

## Derivatives of the inverse link {#sec-ift}

Every link in the catalog implements two families of derivatives: those of
$g$ with respect to $\theta$, and those of $h$ with respect to $\eta$. Each
family determines the other completely, and the formulas that connect them are
used throughout the toolkit.

::: {#thm-ift}
## Derivatives of an inverse function

Let $g$ be a link function in the sense of @def-link and let $h = g^{-1}$. Then,
with every derivative of $g$ evaluated at $\theta = h(\eta)$,

$$
h' = \frac{1}{g'}, \qquad
h'' = -\frac{g''}{(g')^{3}}, \qquad
h''' = \frac{3(g'')^{2} - g'g'''}{(g')^{5}}, \qquad
h'''' = \frac{-15(g'')^{3} + 10\,g'g''g''' - (g')^{2}g''''}{(g')^{7}}.
$$
:::

All four expressions follow from the identity $h(g(\theta)) = \theta$, which
holds for every $\theta$ in the domain because $h$ is the inverse of $g$.
Differentiating this identity once gives $h'\,g' = 1$, and hence the first
formula. Each further differentiation produces an equation that is linear in
the highest derivative of $h$, with a power of $g'$ as its coefficient; since
$g'$ never vanishes, this equation can be solved for that derivative, and the
remaining formulas follow one after another. The same procedure, carried to arbitrary order and organized through set
partitions, is the Faà di Bruno formula, whose history and combinatorics are
presented in @johnson2002 and @comtet1974. @sec-link-scale uses it in that
general form to transport likelihood derivatives of any order onto the
unconstrained scale.

A short example makes the formulas concrete and checks the signs. For the log
link, $g(\theta) = \log\theta$, the inverse $h(\eta) = e^\eta$ is its own
derivative at every order, so every $h^{(k)}$ must equal
$u = e^{\eta} = \theta$. On the forward side $g' = u^{-1}$, $g'' = -u^{-2}$,
$g''' = 2u^{-3}$ and $g'''' = -6u^{-4}$. Substituting into the fourth formula,

$$\frac{-15(-u^{-2})^{3} + 10\,u^{-1}(-u^{-2})(2u^{-3}) - u^{-2}(-6u^{-4})}{u^{-7}}
= \frac{15u^{-6} - 20u^{-6} + 6u^{-6}}{u^{-7}} = u,$$

which is the correct value.

## The role of the fourth order {#sec-why-four}

The fourth derivative of a link is genuinely used, and the formula derived in
@sec-link-scale for the fourth derivative of a log-likelihood with respect to
the linear predictor shows where:

$$
\frac{\partial^{4}\ell}{\partial\eta^{4}}
= \ell^{(\theta\theta\theta\theta)}(h')^{4}
+ 6\,\ell^{(\theta\theta\theta)}(h')^{2}h''
+ \ell^{(\theta\theta)}\left(4h'h''' + 3(h'')^{2}\right)
+ \ell^{(\theta)}\,h'''' .
$$

The last term contains $h''''$, which vanishes only for an affine link. A link
implemented only to second order would therefore make this quantity wrong, not
merely less accurate, and the error would not shrink as the sample grows.
Fourth-order likelihood derivatives are exactly what bias correction and
higher-order asymptotics require, so the four-derivative requirement in
@def-link is necessary for those applications.

## The interface {#sec-link-interface}

A link is an S7 object. The base class `link` carries the name, the domain
$\Theta$ (`link_bounds`), and any additional parameters the link needs, such
as the exponent of the power link; each concrete link is a subclass.

```{r}
#| label: link-object
lk <- logit_link()
lk
c(lk@link_name, lk@link_bounds)
```

Every link implements the same ten generics, five in each direction:

| Generic | Returns |
|---|---|
| `linkfun(x, theta)` | $g(\theta)$ |
| `linkinv(x, eta)` | $h(\eta)$ |
| `dlinkfun` … `d4linkfun` | $g'(\theta)$ … $g''''(\theta)$ |
| `dlinkinv` … `d4linkinv` | $h'(\eta)$ … $h''''(\eta)$ |

: The ten generics {.striped}

For interactive work, two routers select the order at run time:

```{r}
#| label: link-routers
theta <- c(0.2, 0.5, 0.8)
linkfun(lk, theta)
linkderiv(lk, theta, order = 2)
linkinvderiv(lk, linkfun(lk, theta), order = 2)
```

One practical remark about these routers. `linkderiv()` dispatches on the link
and then calls the order-specific generic, which dispatches again, so each
routed call pays the cost of an S7 dispatch — roughly five microseconds —
twice. The cost accumulates in a loop: `distributions7` evaluates inverse-link
derivatives once per parameter per order on every scoring iteration.
Performance-critical code therefore calls `dlinkinv()` and its siblings
directly, and the routers are intended for interactive use.

Every link also carries `print` and `plot` methods. The plot draws the two
directions of the map side by side: the forward direction stretches the
neighborhoods of the domain's endpoints out to infinity, and the inverse
compresses the whole real line back between them.

```{r}
#| label: link-plot
#| fig-height: 4
#| fig-cap: "The complementary log-log link, in both directions."
plot(cloglog_link())
```

## The catalog {#sec-link-catalog}

Fourteen constructors produce sixteen classes: `bounded_link()` returns a
lower-, an upper- or a doubly-bounded link according to which of its two
arguments are supplied, and reduces to the identity link when neither is. For
each entry the table gives the domain, the pair $g$ and $h$, and all eight
derivatives; the accompanying note records the mathematical or numerical
properties specific to that link. Links that take a parameter — the power
link, the softplus, the bounded family — are shown at a representative value.

```{r}
#| label: catalog
#| results: asis
#| echo: false
for (id in names(LINKS)) render_link_entry(id, LINKS[[id]])
```

## Validating a link {#sec-check-link}

Users can define links of their own, so the package provides `check_link()`,
which takes any link object and verifies the properties that @def-link
demands. It checks invertibility in both directions on a grid, strict
monotonicity of either sign, and the first-order inverse function theorem
$h'(\eta)\,g'(\theta) = 1$. It then compares each analytical derivative, in
both families and at every order up to the fourth, with a numerical one.

```{r}
#| label: check-link-demo
invisible(check_link(cloglog_link()))
```

The derivative checks are organized as a chain because of a limitation of
numerical differentiation itself. A fourth derivative cannot be verified by
differentiating $g$ four times numerically: each numerical differentiation
amplifies the error of the previous one, and by the third order the result is
noise. The validator instead compares the analytical derivative of order $k$
against a *single* numerical differentiation of the analytical derivative of
order $k-1$, so that every comparison involves exactly one finite-difference
step and keeps its full accuracy. The grid also stays away from the edges of the domain. The Richardson
extrapolation in `numDeriv` evaluates the function at points up to about
$8 \times 10^{-4}\,|x|$ away from $x$, so a grid point closer than that to a
boundary would be differentiated using values from outside the domain.

## Numerical fallbacks {#sec-link-fallbacks}

Of the ten generics in @sec-link-interface, only `linkfun` and `linkinv` are
compulsory. A link that supplies $g$ and $h$ and nothing else still returns
every derivative, because the base class carries numerical fallbacks for the
eight generics it was not given, and the routers reach them like any other
method.

```{r}
#| label: bare-link
BareLog <- S7::new_class("BareLog", parent = link, package = NULL)
S7::method(linkfun, BareLog) <- function(x, theta) log(theta)
S7::method(linkinv, BareLog) <- function(x, eta)   exp(eta)

bare <- BareLog(link_name = "bare log", link_bounds = c(0, Inf), link_params = NULL)

linkderiv(bare, 2, order = 3)   # third derivative, never written down
2 / 2^3                         # the closed form
```

`distributions7` offers the same arrangement to a distribution that supplies
only its density, for the same reason. The user most likely to need the
fallbacks is the one experimenting with a link nobody has implemented, and
requiring four hand-computed derivatives before anything runs would
discourage the experiment entirely.

### The finite-difference scheme {#sec-link-fallback-chain}

The fallbacks follow the same rule that @sec-check-link imposes on the
validator: a quantity is never differentiated numerically more than once in a
row. The simplest implementation would be a recursion: the second derivative as a
difference of the first, the third as a difference of the second. But each
numerical differentiation multiplies the error of the one below it, and such
a chain is useless by the time it reaches the fourth order. The identity
link shows the failure directly: its third derivative is exactly zero, while
three nested differences return a number of order one. The instability of
repeated differencing, and the choice of step that balances truncation against
rounding, are treated in @gill1981 and @press2007.

The fallbacks instead determine how many orders are actually missing. Each one
finds the highest order $m$ the link implements analytically and applies a
**single** central stencil of order $k - m$ to that derivative:

$$
f' \approx \frac{f_{+1} - f_{-1}}{2h}, \qquad
f'' \approx \frac{f_{+1} - 2f_{0} + f_{-1}}{h^{2}},
$$
$$
f''' \approx \frac{f_{+2} - 2f_{+1} + 2f_{-1} - f_{-2}}{2h^{3}}, \qquad
f'''' \approx \frac{f_{+2} - 4f_{+1} + 6f_{0} - 4f_{-1} + f_{-2}}{h^{4}},
$$

writing $f_{\pm j}$ for $f(x \pm jh)$. A link analytic to the second order
therefore reaches the third derivative with one difference and the fourth with
two, instead of with three nested differences of its own:

```{r}
#| label: fallback-accuracy
HalfLog <- S7::new_class("HalfLog", parent = link, package = NULL)
S7::method(linkfun,   HalfLog) <- function(x, theta) log(theta)
S7::method(linkinv,   HalfLog) <- function(x, eta)   exp(eta)
S7::method(dlinkfun,  HalfLog) <- function(x, theta) 1 / theta
S7::method(d2linkfun, HalfLog) <- function(x, theta) -1 / theta^2
S7::method(dlinkinv,  HalfLog) <- function(x, eta) exp(eta)
S7::method(d2linkinv, HalfLog) <- function(x, eta) exp(eta)

half <- HalfLog(link_name = "half log", link_bounds = c(0, Inf), link_params = NULL)

theta <- c(0.5, 1, 2, 5)
exact4 <- -6 / theta^4
rel <- function(a, b) max(abs(a - b) / abs(b))

c(nothing_analytic = rel(linkderiv(bare, theta, order = 4), exact4),
  two_orders_given = rel(linkderiv(half, theta, order = 4), exact4))
```

Supplying two analytical orders, a couple of lines of algebra, gains several
digits of accuracy at the fourth. The step is $\varepsilon^{1/(k+2)}\max(1, \lvert x\rvert)$ for a stencil of
order $k$, which balances truncation error against rounding error. Near the
ends of the domain the step is reduced: the domain is open, the third- and
fourth-order stencils reach out to $2h$, and that full reach has to stay
inside it.
Without this reduction, differentiating the log link at $\theta = 10^{-8}$
would evaluate a logarithm at a negative number.

### Validation of numerical orders {#sec-link-fallback-catch}

The fallbacks create a problem for validation, the same one that
@sec-fallbacks records for distributions. `check_link()` verifies an
analytical derivative by comparing it against a numerical one; when the
"analytical" derivative is itself the numerical fallback, the comparison sets
a quantity against itself, agrees perfectly, and establishes nothing.

The validator therefore does not report such an order as having passed. It
reports it as numerical, and states up to which order the link is exact.

```{r}
#| label: check-link-honest
invisible(check_link(bare))
invisible(check_link(half))
```

Neither link is wrong; both are less thoroughly checked than a plain
`[PASSED]` would suggest, and the summary states which orders were verified.
`link_fallback_orders()` returns the same information programmatically:

```{r}
#| label: fallback-orders
link_fallback_orders(half)
link_fallback_orders(logit_link())
```

Every link in the catalog reports four analytic orders in both directions.
The fallbacks exist for links that are still incomplete: they make such a link
usable rather than exact, the catalog does not rely on them, and a link
intended for regular use should eventually not need them either.

## In practice {#sec-link-practice}

The census below constructs every link and reads its name and domain off
the object, so it stays current by construction:

```{r}
#| label: lp-census
links <- list(identity_link(), log_link(), logit_link(), probit_link(),
              cloglog_link(), loglog_link(), cauchit_link(), rhobit_link(),
              sqrt_link(), inverse_link(), inverse_sq_link(), power_link(2),
              softplus_link(1), bounded_link(0, 1), bounded_link(lwr = 2),
              bounded_link(upr = 5))
knitr::kable(data.frame(
  link = sapply(links, function(l) l@link_name),
  domain = sapply(links, function(l)
    sprintf("(%g, %g)", l@link_bounds[1], l@link_bounds[2]))
), row.names = FALSE, caption = "The link classes, read off the objects.")
```

A link is a property of a distribution's parameter, so exchanging one for
another changes the scale a model works on without touching the family.
The same Gamma carries its mean through a log link or through a square
root link, and the two fits agree on the parameter scale:

```{r}
#| label: lp-swap
d_log  <- gamma2_distrib()
d_sqrt <- gamma2_distrib(link_mu = sqrt_link())
set.seed(4)
y <- distrib_rng(d_log, 400, list(mu = 3, sigma2 = 2))

f_log  <- fit_distrib(d_log, y)
f_sqrt <- fit_distrib(d_sqrt, y)
rbind(log_link = unlist(coef(f_log)), sqrt_link = unlist(coef(f_sqrt)))
rbind(log_link = unlist(coef(f_log, scale = "link")),
      sqrt_link = unlist(coef(f_sqrt, scale = "link")))
```

The derivatives behind the fit are the ten generics of
@sec-link-interface, reached one order at a time or through the routers:

```{r}
#| label: lp-derivs
lk <- sqrt_link()
theta <- c(0.5, 2, 8)
rbind(d1 = dlinkinv(lk, linkfun(lk, theta)),
      d2 = d2linkinv(lk, linkfun(lk, theta)),
      d3 = linkinvderiv(lk, linkfun(lk, theta), order = 3))
```

```{r}
#| label: consistency-links
#| include: false
# Internal consistency gate: fails the render if any formula printed in this
# chapter disagrees with the package.
assert_links_ok()
```
