# The linkfunctions7 package {#sec-links}

```{r}
#| label: setup-02
#| include: false
source("R/_setup.R")
source("R/link-formulas.R")
```

The parameters of a distribution rarely live on the whole real line. A scale
must be positive, a probability lies between zero and one, a correlation between
minus one and one. An optimiser, on the other hand, is happiest with an argument
it can move freely in any direction. A link function is the bridge between the
two worlds: it maps the natural, constrained domain of a parameter onto an
unconstrained scale, and its inverse maps the optimiser's iterates back into the
domain where the distribution makes sense. This chapter defines the objects
precisely, works out the relationship between the derivatives of a link and
those of its inverse, explains why the toolkit carries both to fourth order, and
ends with the full catalogue of the sixteen link classes that `linkfunctions7`
provides.

## Definition {#sec-link-def}

::: {#def-link}
## Link function

Let $\Theta \subseteq \mathbb{R}$ be an open interval. A **link function** on
$\Theta$ is a map $g : \Theta \to \mathbb{R}$ which is a bijection onto its
image $E = g(\Theta)$, itself an open interval; which is four times continuously
differentiable on $\Theta$; and which is *regular*, meaning that
$g'(\theta) \neq 0$ everywhere. We write $h = g^{-1} : E \to \Theta$ for the
inverse, and throughout the book $\eta = g(\theta)$ and $\theta = h(\eta)$.
:::

Each clause of this definition is a decision with consequences, so it is worth
pausing on all three.

Notice first what the definition does *not* say: it never asks $g$ to be
increasing. Monotonicity is a consequence rather than an assumption — a
continuous function whose derivative never vanishes on an interval cannot change
the sign of that derivative, so it is automatically strictly monotone — but the
direction is left free, and this freedom is used. The link for a parameter
bounded above, $g(\theta) = \log(u - \theta)$, is strictly *decreasing*, and it
is a perfectly good link. Code that quietly assumes an increasing link, for
instance when mapping quantiles through it, breaks on exactly that case, which
is why the natural property to require, and to test, is strict monotonicity of
either sign.

Second, the image $E$ need not be the whole real line. For the square-root link
$E = (0, \infty)$, and for the reciprocal link it is a half-line as well. In
such cases the optimiser is not truly unconstrained — the constraint has merely
moved from $\theta$ to $\eta$ — and it helps to know this in advance, because
these are the links where an optimiser can step outside the domain and come back
with `NaN`. The catalogue below therefore states $E$ explicitly for every link.

Third, and most characteristically for this toolkit, the definition demands four
continuous derivatives rather than the customary one or two. The link objects in
`stats::make.link()` carry the first derivative of the inverse and nothing else,
which is all a generalized linear model fitted by iteratively reweighted least
squares ever needs — the setting of @nelder1972 and @mccullagh1989, where the
link first appears under that name. Here the ambitions are higher: as
@sec-link-scale will show in detail, transporting a fourth-order derivative of a
log-likelihood onto the unconstrained scale requires $h'$, $h''$, $h'''$ and
$h''''$, so a link that stopped at second order would silently cap the entire
toolkit at second-order methods.

## Derivatives of the inverse link {#sec-ift}

Every link in the catalogue implements two families of derivatives: those of $g$
with respect to $\theta$, and those of $h$ with respect to $\eta$. The two
families are not independent, since each determines the other completely, and
the formulas that connect them are used throughout the toolkit.

::: {#thm-ift}
## Derivatives of an inverse function

Let $g$ be a link function in the sense of @def-link and let $h = g^{-1}$. Then,
with every derivative of $g$ evaluated at $\theta = h(\eta)$,

$$
h' = \frac{1}{g'}, \qquad
h'' = -\frac{g''}{(g')^{3}}, \qquad
h''' = \frac{3(g'')^{2} - g'g'''}{(g')^{5}}, \qquad
h'''' = \frac{-15(g'')^{3} + 10\,g'g''g''' - (g')^{2}g''''}{(g')^{7}}.
$$
:::

These four expressions all come from one identity. Because $h$ undoes $g$
we have $h(g(\theta)) = \theta$ for every $\theta$ in the domain, and each
differentiation of that identity produces a relation between the derivatives of
$h$ and those of $g$ which is linear in the highest derivative of $h$, with a
power of $g'$ for its coefficient. Since $g'$ never vanishes, that relation can
always be solved for the derivative one does not yet know, and the four formulas
follow one after another. The same mechanism, carried to arbitrary order and
organised by set partitions rather than by hand, is the Faà di Bruno formula,
whose history and combinatorics are surveyed in @johnson2002 and @comtet1974;
@sec-link-scale uses it in that general form to move a likelihood derivative of
any order onto the unconstrained scale.

A quick example makes the formulas concrete, and catches sign errors better
than any amount of staring. Take the log link, $g(\theta) = \log\theta$, whose
inverse $h(\eta) = e^\eta$ has the agreeable property of being its own
derivative to every order, so every $h^{(k)}$ should equal
$u = e^{\eta} = \theta$. On the forward side $g' = u^{-1}$, $g'' = -u^{-2}$,
$g''' = 2u^{-3}$ and $g'''' = -6u^{-4}$. Feeding these into the fourth formula,

$$\frac{-15(-u^{-2})^{3} + 10\,u^{-1}(-u^{-2})(2u^{-3}) - u^{-2}(-6u^{-4})}{u^{-7}}
= \frac{15u^{-6} - 20u^{-6} + 6u^{-6}}{u^{-7}} = u,$$

exactly as it should.

## The role of the fourth order {#sec-why-four}

It is fair to ask whether a fourth derivative of a link function is ever really
used, or whether it is completeness for its own sake. The answer comes from the
formula that will be derived in @sec-link-scale for the fourth derivative of a
log-likelihood with respect to the linear predictor:

$$
\frac{\partial^{4}\ell}{\partial\eta^{4}}
= \ell^{(\theta\theta\theta\theta)}(h')^{4}
+ 6\,\ell^{(\theta\theta\theta)}(h')^{2}h''
+ \ell^{(\theta\theta)}\left(4h'h''' + 3(h'')^{2}\right)
+ \ell^{(\theta)}\,h'''' .
$$

The last term contains $h''''$, and it does not vanish for any link except an
affine one. A link truncated at second order would therefore not merely make
this quantity less accurate — it would make it wrong, by an amount that has
nothing to do with any small parameter and does not shrink as the sample grows.
Since fourth-order likelihood derivatives are exactly what bias correction and
higher-order asymptotics consume, the four-derivative requirement in
@def-link is a load-bearing part of the design.

## The interface {#sec-link-interface}

A link is an S7 object. The base class `link` carries the name, the domain
$\Theta$ (`link_bounds`), and any extra mathematical parameters the link needs,
such as the exponent of the power link; each concrete link is a subclass.

```{r}
#| label: link-object
lk <- logit_link()
lk
c(lk@link_name, lk@link_bounds)
```

Every link implements the same ten generics, five in each direction:

| Generic | Returns |
|---|---|
| `linkfun(x, theta)` | $g(\theta)$ |
| `linkinv(x, eta)` | $h(\eta)$ |
| `dlinkfun` … `d4linkfun` | $g'(\theta)$ … $g''''(\theta)$ |
| `dlinkinv` … `d4linkinv` | $h'(\eta)$ … $h''''(\eta)$ |

: The ten generics {.striped}

For interactive work, two routers select the order at run time:

```{r}
#| label: link-routers
theta <- c(0.2, 0.5, 0.8)
linkfun(lk, theta)
linkderiv(lk, theta, order = 2)
linkinvderiv(lk, linkfun(lk, theta), order = 2)
```

One practical remark about these routers. `linkderiv()` dispatches on the link
and then calls the order-specific generic, which dispatches again; since an S7
dispatch costs roughly five microseconds against one for a plain call, the
second dispatch accounts for about a third of the total. In a hot loop this
adds up — `distributions7` evaluates inverse-link derivatives once per
parameter per order on every scoring iteration — so performance-critical code
calls `dlinkinv()` and its siblings directly, and the routers are best regarded
as a convenience for interactive use.

## The catalogue {#sec-link-catalogue}

Fourteen constructors produce sixteen classes: `bounded_link()` returns a
lower-, an upper- or a doubly-bounded link according to which of its two
arguments are supplied, and reduces to the identity link when neither is. For
each entry the table gives the domain, the pair $g$ and $h$, and all eight
derivatives; the accompanying note records whatever is worth knowing about the
mathematics or the numerics of that particular link. Links that take a
parameter — the power link, the softplus, the bounded family — are shown at a
representative value.

```{r}
#| label: catalogue
#| results: asis
#| echo: false
for (id in names(LINKS)) render_link_entry(id, LINKS[[id]])
```

## Validating a link {#sec-check-link}

Since users can define links of their own, the package provides `check_link()`,
which takes any link object and verifies the properties that @def-link demands:
invertibility in both directions on a grid, strict monotonicity of either sign,
the first-order inverse function theorem $h'(\eta)\,g'(\theta) = 1$, and the
agreement of each analytical derivative, in both families and at every order up
to the fourth, with a numerical one.

```{r}
#| label: check-link-demo
invisible(check_link(cloglog_link()))
```

The derivative checks are organised as a chain, and the reason deserves a
paragraph, because it reflects a genuine limitation of numerical
differentiation rather than an implementation detail. One cannot verify a
fourth derivative by differentiating $g$ itself four times numerically: each
numerical differentiation amplifies the error of the previous one, and by the
third order the result is noise. The validator instead compares the analytical
derivative of order $k$ against a *single* numerical differentiation of the
analytical derivative of order $k-1$, so that every comparison involves exactly
one finite-difference step and retains its full accuracy. The grid, moreover,
keeps away from the edges of the domain: the Richardson extrapolation in
`numDeriv` evaluates the function at points up to about
$8 \times 10^{-4}\,|x|$ away from $x$, and a grid point closer than that to a
boundary would be differentiated using values from outside the domain.

## Numerical fallbacks {#sec-link-fallbacks}

The ten generics of @sec-link-interface are what a link *may* implement. Only
two of them are compulsory. A link that supplies $g$ and $h$ and stops there
still answers every derivative, because the base class carries numerical
fallbacks for the eight it was not given, and the routers reach them like any
other method.

```{r}
#| label: bare-link
BareLog <- S7::new_class("BareLog", parent = link, package = NULL)
S7::method(linkfun, BareLog) <- function(x, theta) log(theta)
S7::method(linkinv, BareLog) <- function(x, eta)   exp(eta)

bare <- BareLog(link_name = "bare log", link_bounds = c(0, Inf), link_params = NULL)

linkderiv(bare, 2, order = 3)   # third derivative, never written down
2 / 2^3                         # the closed form
```

This is the same bargain `distributions7` offers a distribution that supplies
only its density, and it exists for the same reason: the person most likely to
need it is the one experimenting with a transformation nobody has implemented,
and asking them to differentiate it by hand four times before anything runs is
asking them not to bother.

### The finite-difference scheme {#sec-link-fallback-chain}

How a fallback computes its answer matters more here than one might expect,
because it follows the same rule that @sec-check-link imposes on the validator:
a quantity is never differentiated numerically more than once in a row. The
simplest thing to write would be a recursion, taking the second derivative as a
difference of the first and the third as a difference of the second, but each
numerical differentiation multiplies the error of the one below it, so such a
chain is useless by the time it reaches the top. The identity link makes the
point without any arithmetic at all, since its third derivative is exactly zero
while three nested differences return a number of order one. The dangers of
repeated differencing, and the choice of step that balances truncation against
rounding, are treated at length in @gill1981 and @press2007.

What the fallbacks do instead is ask how many orders are actually missing. Each
one finds the highest order $m$ the link implements analytically, and applies a
**single** central stencil of order $k - m$ to that derivative:

$$
f' \approx \frac{f_{+1} - f_{-1}}{2h}, \qquad
f'' \approx \frac{f_{+1} - 2f_{0} + f_{-1}}{h^{2}},
$$
$$
f''' \approx \frac{f_{+2} - 2f_{+1} + 2f_{-1} - f_{-2}}{2h^{3}}, \qquad
f'''' \approx \frac{f_{+2} - 4f_{+1} + 6f_{0} - 4f_{-1} + f_{-2}}{h^{4}},
$$

writing $f_{\pm j}$ for $f(x \pm jh)$. A link analytic to the second order
therefore reaches the third with one difference and the fourth with two, rather
than standing on three differences of its own — and the difference this makes is
not marginal:

```{r}
#| label: fallback-accuracy
HalfLog <- S7::new_class("HalfLog", parent = link, package = NULL)
S7::method(linkfun,   HalfLog) <- function(x, theta) log(theta)
S7::method(linkinv,   HalfLog) <- function(x, eta)   exp(eta)
S7::method(dlinkfun,  HalfLog) <- function(x, theta) 1 / theta
S7::method(d2linkfun, HalfLog) <- function(x, theta) -1 / theta^2
S7::method(dlinkinv,  HalfLog) <- function(x, eta) exp(eta)
S7::method(d2linkinv, HalfLog) <- function(x, eta) exp(eta)

half <- HalfLog(link_name = "half log", link_bounds = c(0, Inf), link_params = NULL)

theta <- c(0.5, 1, 2, 5)
exact4 <- -6 / theta^4
rel <- function(a, b) max(abs(a - b) / abs(b))

c(nothing_analytic = rel(linkderiv(bare, theta, order = 4), exact4),
  two_orders_given = rel(linkderiv(half, theta, order = 4), exact4))
```

Two analytical orders, which is a couple of lines of algebra, buy several digits
at the fourth. The step itself is $\varepsilon^{1/(k+2)}\max(1, \lvert x\rvert)$
for a stencil of order $k$, balancing truncation against rounding, and it is
shrunk near the ends of the domain: a link's domain is open, and the third- and
fourth-order stencils reach $2h$, so it is that reach and not $h$ that has to
fit inside it. Differentiating the log link at $\theta = 10^{-8}$ would
otherwise evaluate a logarithm at a negative number.

### Validation of numerical orders {#sec-link-fallback-catch}

A convenience of this kind comes with a hazard, and it is the same one
@sec-fallbacks records for distributions. `check_link()` verifies an analytical
derivative by comparing it against a numerical one. If the analytical derivative
*is* the numerical one, the comparison is between a quantity and itself: it
agrees perfectly, and it has established nothing.

The validator therefore does not report such an order as having passed; it
reports it as numerical, and says how far the exactness actually goes.

```{r}
#| label: check-link-honest
invisible(check_link(bare))
invisible(check_link(half))
```

Neither link is wrong; both are simply less checked than a reader of a bare
`[PASSED]` would assume, and the summary says so. `link_fallback_orders()`
answers the same question programmatically, which is the form a script wants:

```{r}
#| label: fallback-orders
link_fallback_orders(half)
link_fallback_orders(logit_link())
```

Everything in the catalogue reports four and four. The fallbacks exist for links
that do not yet, and the honest description of them is that they make an
incomplete link *usable* rather than *exact* — which is why the catalogue does
not rely on them, and why a link intended for serious use should end up not
needing them either.

```{r}
#| label: consistency-links
#| include: false
# Internal consistency gate: fails the render if any formula printed in this
# chapter disagrees with the package.
assert_links_ok()
```
