# Link functions {#sec-links}

```{r}
#| label: setup-02
#| include: false
source("R/_setup.R")
source("R/link-formulas.R")
```

The parameters of a distribution rarely live on the whole real line. A scale
must be positive, a probability lies between zero and one, a correlation between
minus one and one. An optimiser, on the other hand, is happiest with an argument
it can move freely in any direction. A link function is the bridge between the
two worlds: it maps the natural, constrained domain of a parameter onto an
unconstrained scale, and its inverse maps the optimiser's iterates back into the
domain where the distribution makes sense. This chapter defines the objects
precisely, works out the relationship between the derivatives of a link and
those of its inverse, explains why the stack carries both to fourth order, and
ends with the full catalogue of the sixteen link classes that `linkfunctions7`
provides.

## What a link function is {#sec-link-def}

::: {#def-link}
## Link function

Let $\Theta \subseteq \mathbb{R}$ be an open interval. A **link function** on
$\Theta$ is a map $g : \Theta \to \mathbb{R}$ which is a bijection onto its
image $E = g(\Theta)$, itself an open interval; which is four times continuously
differentiable on $\Theta$; and which is *regular*, meaning that
$g'(\theta) \neq 0$ everywhere. We write $h = g^{-1} : E \to \Theta$ for the
inverse, and throughout the book $\eta = g(\theta)$ and $\theta = h(\eta)$.
:::

Each clause of this definition is a decision with consequences, so it is worth
pausing on all three.

Notice first what the definition does *not* say: it never asks $g$ to be
increasing. Monotonicity is a consequence rather than an assumption — a
continuous function whose derivative never vanishes on an interval cannot change
the sign of that derivative, so it is automatically strictly monotone — but the
direction is left free, and this freedom is used. The link for a parameter
bounded above, $g(\theta) = \log(u - \theta)$, is strictly *decreasing*, and it
is a perfectly good link. Code that quietly assumes an increasing link, for
instance when mapping quantiles through it, breaks on exactly that case, which
is why the natural property to require, and to test, is strict monotonicity of
either sign.

Second, the image $E$ need not be the whole real line. For the square-root link
$E = (0, \infty)$, and for the reciprocal link it is a half-line as well. In
such cases the optimiser is not truly unconstrained — the constraint has merely
moved from $\theta$ to $\eta$ — and it helps to know this in advance, because
these are the links where an optimiser can step outside the domain and come back
with `NaN`. The catalogue below therefore states $E$ explicitly for every link.

Third, and most characteristically for this stack, the definition demands four
continuous derivatives rather than the customary one or two. The link objects in
`stats::make.link()` carry the first derivative of the inverse and nothing else,
which is all a GLM fitted by IRLS ever needs. Here the ambitions are higher: as
@sec-link-scale will show in detail, transporting a fourth-order derivative of a
log-likelihood onto the unconstrained scale requires $h'$, $h''$, $h'''$ and
$h''''$, so a link that stopped at second order would silently cap the entire
stack at second-order methods.

## The derivatives of an inverse function {#sec-ift}

Every link in the catalogue implements two families of derivatives: those of $g$
with respect to $\theta$, and those of $h$ with respect to $\eta$. The two
families are not independent — each determines the other completely — and the
relationship between them is worth deriving carefully, both because the formulas
are used throughout the stack and because the derivation is a pleasant exercise
in the disciplined use of the chain rule.

::: {#thm-ift}
## Derivatives of an inverse function

Let $g$ be a link function in the sense of @def-link and let $h = g^{-1}$. Then,
with every derivative of $g$ evaluated at $\theta = h(\eta)$,

$$
h' = \frac{1}{g'}, \qquad
h'' = -\frac{g''}{(g')^{3}}, \qquad
h''' = \frac{3(g'')^{2} - g'g'''}{(g')^{5}}, \qquad
h'''' = \frac{-15(g'')^{3} + 10\,g'g''g''' - (g')^{2}g''''}{(g')^{7}}.
$$
:::

::: {.proof}
The whole argument flows from one identity. Since $h$ undoes $g$, we have
$h(g(\theta)) = \theta$ for every $\theta$ in the domain, and we may
differentiate this identity as many times as we like; each differentiation
yields a relation between the derivatives of $h$ and those of $g$, from which
the next unknown derivative of $h$ can be isolated. Throughout, $h^{(k)}$ is
shorthand for $h^{(k)}(g(\theta))$.

Differentiating once, the chain rule gives $h'\,g' = 1$. Since $g'$ never
vanishes, we may divide by it, and the first formula, $h' = 1/g'$, is already in
hand.

Now differentiate $h'g' = 1$ again. The left side is a product whose two
factors both depend on $\theta$ — the first through the composition with $g$ —
so the product rule together with the chain rule gives

$$h''\,(g')^{2} + h'\,g'' = 0 .$$

Solving for $h''$ and substituting $h' = 1/g'$,

$$h'' = -\frac{h'\,g''}{(g')^{2}} = -\frac{g''}{(g')^{3}},$$

which is the second formula. Note the pattern that will repeat: each new
differentiation produces an equation that is *linear* in the highest derivative
of $h$, with coefficient a power of $g'$, so isolating it is always a division.

Differentiating $h''(g')^{2} + h'g'' = 0$ once more, and collecting the terms
carefully,

$$h'''\,(g')^{3} + 3\,h''\,g'\,g'' + h'\,g''' = 0 . \tag{$\ast$}$$

Substituting the expressions already obtained for $h''$ and $h'$, the middle
term becomes $-3(g'')^{2}/(g')^{2}$ and the last becomes $g'''/g'$, so that

$$h''' = -\frac{1}{(g')^{3}}\left(-\frac{3(g'')^{2}}{(g')^{2}} + \frac{g'''}{g'}\right)
       = \frac{3(g'')^{2} - g'\,g'''}{(g')^{5}} .$$

One more round. Differentiating each of the three terms of $(\ast)$,

$$
\begin{aligned}
\frac{d}{d\theta}\!\left[h'''(g')^{3}\right] &= h''''(g')^{4} + 3h'''(g')^{2}g'', \\
\frac{d}{d\theta}\!\left[3h''g'g''\right] &= 3h'''(g')^{2}g'' + 3h''(g'')^{2} + 3h''g'g''', \\
\frac{d}{d\theta}\!\left[h'g'''\right] &= h''g'g''' + h'g'''' ,
\end{aligned}
$$

and adding the three lines and setting the sum to zero,

$$h''''(g')^{4} + 6h'''(g')^{2}g'' + 3h''(g'')^{2} + 4h''g'g''' + h'g'''' = 0 .$$

It remains to substitute the three lower-order results and simplify. Putting
everything except the first term over the common denominator $(g')^{3}$, the
sum of the known terms works out to

$$\frac{15(g'')^{3} - 10\,g'g''g''' + (g')^{2}g''''}{(g')^{3}},$$

and dividing by $-(g')^{4}$ yields the stated expression for $h''''$. $\square$
:::

A quick example makes the formulas concrete, and catches sign errors better
than any amount of staring. Take the log link, $g(\theta) = \log\theta$, whose
inverse $h(\eta) = e^\eta$ has the agreeable property of being its own
derivative to every order, so every $h^{(k)}$ should equal
$u = e^{\eta} = \theta$. On the forward side $g' = u^{-1}$, $g'' = -u^{-2}$,
$g''' = 2u^{-3}$ and $g'''' = -6u^{-4}$. Feeding these into the fourth formula,

$$\frac{-15(-u^{-2})^{3} + 10\,u^{-1}(-u^{-2})(2u^{-3}) - u^{-2}(-6u^{-4})}{u^{-7}}
= \frac{15u^{-6} - 20u^{-6} + 6u^{-6}}{u^{-7}} = u,$$

exactly as it should.

## Why the fourth order matters {#sec-why-four}

It is fair to ask whether a fourth derivative of a link function is ever really
used, or whether it is completeness for its own sake. The answer comes from the
formula that will be derived in @sec-link-scale for the fourth derivative of a
log-likelihood with respect to the linear predictor:

$$
\frac{\partial^{4}\ell}{\partial\eta^{4}}
= \ell^{(\theta\theta\theta\theta)}(h')^{4}
+ 6\,\ell^{(\theta\theta\theta)}(h')^{2}h''
+ \ell^{(\theta\theta)}\left(4h'h''' + 3(h'')^{2}\right)
+ \ell^{(\theta)}\,h'''' .
$$

The last term contains $h''''$, and it does not vanish for any link except an
affine one. A link truncated at second order would therefore not merely make
this quantity less accurate — it would make it wrong, by an amount that has
nothing to do with any small parameter and does not shrink as the sample grows.
Since fourth-order likelihood derivatives are exactly what bias correction and
higher-order asymptotics consume, the four-derivative requirement in
@def-link is a load-bearing part of the design.

## The interface {#sec-link-interface}

A link is an S7 object. The base class `link` carries the name, the domain
$\Theta$ (`link_bounds`), and any extra mathematical parameters the link needs,
such as the exponent of the power link; each concrete link is a subclass.

```{r}
#| label: link-object
lk <- logit_link()
lk
c(lk@link_name, lk@link_bounds)
```

Every link implements the same ten generics, five in each direction:

| Generic | Returns |
|---|---|
| `linkfun(x, theta)` | $g(\theta)$ |
| `linkinv(x, eta)` | $h(\eta)$ |
| `dlinkfun` … `d4linkfun` | $g'(\theta)$ … $g''''(\theta)$ |
| `dlinkinv` … `d4linkinv` | $h'(\eta)$ … $h''''(\eta)$ |

: The ten generics {.striped}

For interactive work, two routers select the order at run time:

```{r}
#| label: link-routers
theta <- c(0.2, 0.5, 0.8)
linkfun(lk, theta)
linkderiv(lk, theta, order = 2)
linkinvderiv(lk, linkfun(lk, theta), order = 2)
```

One practical remark about these routers. `linkderiv()` dispatches on the link
and then calls the order-specific generic, which dispatches again; since an S7
dispatch costs roughly five microseconds against one for a plain call, the
second dispatch accounts for about a third of the total. In a hot loop this
adds up — `distributions7` evaluates inverse-link derivatives once per
parameter per order on every scoring iteration — so performance-critical code
calls `dlinkinv()` and its siblings directly, and the routers are best regarded
as a convenience for interactive use.

## The catalogue {#sec-link-catalogue}

Fourteen constructors produce sixteen classes: `bounded_link()` returns a
lower-, an upper- or a doubly-bounded link according to which of its two
arguments are supplied, and reduces to the identity link when neither is. For
each entry the table gives the domain, the pair $g$ and $h$, and all eight
derivatives; the accompanying note records whatever is worth knowing about the
mathematics or the numerics of that particular link. Links that take a
parameter — the power link, the softplus, the bounded family — are shown at a
representative value.

```{r}
#| label: catalogue
#| results: asis
#| echo: false
for (id in names(LINKS)) render_link_entry(id, LINKS[[id]])
```

## Validating a link {#sec-check-link}

Since users can define links of their own, the package provides `check_link()`,
which takes any link object and verifies the properties that @def-link demands:
invertibility in both directions on a grid, strict monotonicity of either sign,
the first-order inverse function theorem $h'(\eta)\,g'(\theta) = 1$, and the
agreement of each analytical derivative, in both families and at every order up
to the fourth, with a numerical one.

```{r}
#| label: check-link-demo
invisible(check_link(cloglog_link()))
```

The derivative checks are organised as a chain, and the reason deserves a
paragraph, because it reflects a genuine limitation of numerical
differentiation rather than an implementation detail. One cannot verify a
fourth derivative by differentiating $g$ itself four times numerically: each
numerical differentiation amplifies the error of the previous one, and by the
third order the result is noise. The validator instead compares the analytical
derivative of order $k$ against a *single* numerical differentiation of the
analytical derivative of order $k-1$, so that every comparison involves exactly
one finite-difference step and retains its full accuracy. The grid, moreover,
keeps away from the edges of the domain: the Richardson extrapolation in
`numDeriv` evaluates the function at points up to about
$8 \times 10^{-4}\,|x|$ away from $x$, and a grid point closer than that to a
boundary would be differentiated using values from outside the domain.

```{r}
#| label: consistency-links
#| include: false
# Internal consistency gate: fails the render if any formula printed in this
# chapter disagrees with the package.
assert_links_ok()
```
