# Distributions {#sec-distributions}

```{r}
#| label: setup-03
#| include: false
source("R/_setup.R")
source("R/link-formulas.R")
source("R/distribution-catalogue.R")
```

This chapter develops the likelihood theory that the stack implements and then
presents the fourteen distributions that `distributions7` ships. The theory
comes first because it is shared: the score, the two informations, the Bartlett
identities and the change of scale are defined once, at the level of the base
class, and every distribution — including one a user wrote this morning —
inherits them. The catalogue comes second, with particular attention to the
places where the parametrisation chosen here differs from the one base R uses,
since those are the places where a reader is most likely to be surprised.

## The object and its contract {#sec-distrib-object}

A distribution is an S7 object carrying its name, its dimensionality, the
bounds of its support, the names and interpretations of its parameters, their
individual domains, the link function attached to each of them, and a
per-parameter flag recording whether the log-likelihood is differentiable in
that parameter — a flag whose purpose will become clear in @sec-nonregular. Two
subclasses, `continuous_distrib` and `discrete_distrib`, provide the defaults
appropriate to each kind.

```{r}
#| label: object-anatomy
d <- gamma_distrib()
d@params
d@params_bounds
d@link_params$sigma2
```

Parameters travel through the API as a named list called `theta`, and every
generic normalises this list before doing anything else. The normalisation
reorders the entries by name, so that `list(sigma2 = 2, mu = 3)` and
`list(mu = 3, sigma2 = 2)` mean the same thing rather than silently meaning
different things; it strips names from the *values*, because a parameter that
has passed through a link function comes back carrying its own name, which
would otherwise leak into the results and label a density `"sigma2"`; and it
validates every value against the parameter's domain.

The domains are treated as **open** intervals, and the openness is a deliberate
choice rather than a pedantic one. A Gaussian requires $\sigma > 0$, not
$\sigma \geq 0$, and a Bernoulli requires $0 < \mu < 1$: at the closed boundary
the log-likelihood and its derivatives cease to exist, and accepting the
boundary value would only trade an informative error now for an unexplained
`NaN` several calls later.

```{r}
#| label: bounds-error
#| error: true
distrib_pdf(gaussian_distrib(), 1, list(mu = 0, sigma = 0))
```

The generics fall into four natural groups:

| Group | Generics |
|---|---|
| Probability | `distrib_pdf`, `distrib_cdf`, `distrib_quantile`, `distrib_rng`, `distrib_atoms` |
| Parameter derivatives | `distrib_gradient`, `distrib_hessian`, `distrib_expected_hessian`, `distrib_deriv3`, `distrib_deriv4` |
| Response derivatives | `distrib_grad_y`, `distrib_hess_y` |
| Moments | `expectation`, `mean`, `variance`, `std_dev`, `skewness`, `kurtosis`, `moment` |

: The generics {.striped}

Of all of these, only `distrib_pdf` has no default; everything else can be
derived from the density alone, as the Rayleigh example of @sec-extension
showed.

## Score and information {#sec-likelihood}

Fix a distribution with density or mass function $f(y;\theta)$, where
$\theta = (\theta_1, \dots, \theta_p)^\top$, and write
$\ell(\theta; y) = \log f(y;\theta)$ for the log-likelihood of a single
observation. Subscripts on $\ell$ will always denote differentiation with
respect to parameters, so that

$$
\ell_i = \frac{\partial \ell}{\partial \theta_i}, \qquad
\ell_{ij} = \frac{\partial^{2} \ell}{\partial \theta_i \partial \theta_j}, \qquad
\ell_{ijk} = \frac{\partial^{3} \ell}{\partial \theta_i \partial \theta_j \partial \theta_k},
$$

and so on. The vector $(\ell_1, \dots, \ell_p)^\top$ is the **score**; the
matrix $-[\ell_{ij}]$ is the **observed information**; and the **expected
information**, or Fisher information, is
$\mathcal{I}(\theta) = -\mathbb{E}[\ell_{ij}]$, the expectation being taken over
$Y \sim f(\cdot\,;\theta)$ at the same parameter value.

A word on sign conventions, because both conventions circulate and a sign error
here is invisible inside a symmetric matrix. The package returns *derivatives*,
not informations: `distrib_hessian()` gives $[\ell_{ij}]$, and
`distrib_expected_hessian()` gives $\mathbb{E}[\ell_{ij}]$, each with the sign
of the derivative, so the informations are the negatives of what the functions
return.

Since mixed partial derivatives are symmetric, only one representative of each
multi-index is stored, and components are returned as named lists keyed in a
canonical order — `"mu_sigma2"` stands for
$\ell_{\mu\sigma^2} = \ell_{\sigma^2\mu}$:

```{r}
#| label: derivatives-demo
d <- gamma_distrib()
th <- list(mu = 3, sigma2 = 2)
y <- c(1, 2.5, 7)

str(distrib_gradient(d, y, th))
str(distrib_hessian(d, y, th))
```

Everything in this section and the next two assumes the model is *regular* at
$\theta$: the support of $f$ does not depend on $\theta$, the log-likelihood is
four times differentiable, and derivatives may be exchanged with the integral
sign at the orders involved. Thirteen of the fourteen distributions in the
catalogue satisfy these conditions everywhere; the exception is the Laplace,
whose location parameter sits at a kink of the log-likelihood, and
@sec-nonregular is devoted to what happens then.

## The Bartlett identities {#sec-bartlett}

Most readers will know the identity
$\mathbb{E}[\ell_{ij}] = -\mathbb{E}[\ell_i \ell_j]$, the information matrix
equality. What is less often written down is that this is only the second
member of an infinite family of identities, one for each order of
differentiation, and that the family taken as a whole allows an expected
derivative of *any* order to be computed from expectations of products of
lower-order ones. `distributions7` implements the family in full generality,
so this section derives it.

The starting point could hardly be simpler: a density integrates to one, and it
does so *for every value of the parameter*,

$$\int f(y;\theta)\, dy = 1 \qquad \text{for all } \theta \in \Theta.$$

The right-hand side is constant in $\theta$, so every derivative of the
left-hand side must vanish, and regularity lets us take those derivatives under
the integral sign:

$$\int \frac{\partial^{k} f}{\partial\theta_{i_1}\cdots\partial\theta_{i_k}}\,dy = 0.
\tag{3.1}$$

Everything now depends on expressing the derivatives of $f$ through the
derivatives of $\ell = \log f$, and the expression turns out to have a beautiful
combinatorial structure.

::: {#lem-exp-deriv}
## Derivatives of $e^{\ell}$

For every $k \geq 1$,

$$
\frac{\partial^{k} e^{\ell}}{\partial\theta_{i_1}\cdots\partial\theta_{i_k}}
= e^{\ell} \sum_{\pi \in \Pi_k} \prod_{B \in \pi} \ell_B ,
$$

where $\Pi_k$ is the set of all partitions of the index set
$\{i_1, \dots, i_k\}$ into non-empty blocks, and $\ell_B$ is the derivative of
$\ell$ with respect to the parameters indexed by the block $B$.
:::

::: {.proof}
We argue by induction on $k$. For $k = 1$ the claim reads
$\partial_i e^{\ell} = e^{\ell}\ell_i$, which is the chain rule, and indeed
$\Pi_1$ contains the single partition $\{\{i\}\}$.

Suppose the formula holds at order $k$, and differentiate it with respect to a
new index $\theta_{i_{k+1}}$. By the product rule, the derivative acts either
on the factor $e^{\ell}$ in front — producing an extra factor
$\ell_{i_{k+1}}$ — or on one of the factors $\ell_B$ inside a partition,
turning it into $\ell_{B \cup \{i_{k+1}\}}$. In the first case, the new index
ends up in a block of its own; in the second, it joins an existing block. But
these two cases are precisely the two ways of building a partition of the
$(k{+}1)$-element set from a partition of the $k$-element set, and every
partition of the larger set arises in exactly one of the two ways, from exactly
one partition of the smaller set. The sum therefore runs over $\Pi_{k+1}$ with
each term appearing once, which is the claim at order $k+1$. $\square$
:::

Substituting the lemma into (3.1), and noticing that $e^{\ell} = f$ is exactly
the density needed to turn the integral into an expectation, we arrive at the
result.

::: {#thm-bartlett}
## Bartlett identities

Under the regularity conditions above, for every $k \geq 1$ and every choice of
indices,

$$\sum_{\pi \,\in\, \Pi_k} \mathbb{E}\!\left[\;\prod_{B \in \pi} \ell_B \right] = 0.$$
:::

To use the identity, single out the partition consisting of one big block: its
contribution is $\mathbb{E}[\ell_{i_1 \cdots i_k}]$, the expected derivative of
top order, and moving everything else to the other side,

$$
\mathbb{E}\!\left[\ell_{i_1\cdots i_k}\right]
= -\sum_{\substack{\pi \in \Pi_k \\ |\pi| > 1}}
   \mathbb{E}\!\left[\prod_{B\in\pi}\ell_B\right].
\tag{3.2}$$

The first three cases, written out, show the family growing out of familiar
ground:

$$
\begin{aligned}
k = 1: \quad & \mathbb{E}[\ell_i] = 0, \\[4pt]
k = 2: \quad & \mathbb{E}[\ell_{ij}] = -\mathbb{E}[\ell_i\ell_j], \\[4pt]
k = 3: \quad & \mathbb{E}[\ell_{ijk}] = -\Big(\mathbb{E}[\ell_{ij}\ell_k]
   + \mathbb{E}[\ell_{ik}\ell_j] + \mathbb{E}[\ell_{jk}\ell_i]
   + \mathbb{E}[\ell_i\ell_j\ell_k]\Big).
\end{aligned}
$$

The case $k = 1$ is the zero-mean property of the score; the case $k = 2$ is
the information matrix equality, and read from right to left it says that the
outer-product-of-gradients estimator of the information is not a separate idea
but the second Bartlett identity itself. The implementation follows the
mathematics directly: it enumerates the set partitions of the index set and
sums products of quantities the distribution already provides. The number of
terms is the Bell number $B_k$ — one, two, five and fifteen for
$k = 1, \dots, 4$ — so the computation stays small at every order the stack
supports.

### Three ways to take an expectation {#sec-approx}

When a distribution has no closed-form expected derivative, the expectation has
to be approximated, and the derivative generics accept an `approx` argument
choosing among three strategies. The strategy `"bartlett"` (also spelled
`"opg"`) applies (3.2); at second order it needs nothing but the score, which
makes it the cheapest option, and it is the default there. The strategy
`"integrate"` computes $\mathbb{E}[\partial^k \ell]$ literally, by quadrature
for continuous distributions and by summation for discrete ones; when the
observed derivative is available in closed form this is usually the most
accurate route, and it is the default at orders three and four. The strategy
`"mc"` simulates from the distribution and averages the observed derivative
over the draws; it is robust when quadrature struggles, at the price of Monte
Carlo noise.

The difference between the first strategy and the other two is deeper than a
matter of cost. `"integrate"` and `"mc"` estimate the expected derivative
itself, whereas `"bartlett"` estimates the quantity that the identity says the
expected derivative *equals under regularity*. When regularity fails, the two
quantities part company, and the one that still deserves the name
"information" is the Bartlett one — a point @sec-nonregular makes concrete.

::: {.callout-note}
## A cautionary tale about quadrature

The `expectation()` engine splits its integral at quantiles of the
distribution, and it uses three knots rather than nine. More knots would extend
the range of densities it can integrate at all, but they would also convert
loud failures into silent wrong answers: at Gamma shape $0.02$, a nine-knot
rule returns $-54.9$ for a quantity that is exactly zero, while the three-knot
rule simply reports that the integral could not be computed. In the ordinary
range three knots are also the more accurate choice, since every extra panel
contributes its own quadrature error. An error is worth more than a plausible
number.
:::

## Derivatives on the link scale {#sec-link-scale}

The parameters of a distribution are constrained, and the linear predictors of
a model are not, so every derivative generic accepts an argument
`scale = c("parameter", "link")`. With `scale = "link"` the derivatives are
taken with respect to

$$\eta_i = g_i(\theta_i), \qquad \theta_i = h_i(\eta_i), \qquad h_i = g_i^{-1},$$

where $g_i$ is the link that the distribution carries for its $i$-th parameter.
The structural fact on which the whole section rests is that each parameter has
*its own* link and the map is applied coordinate by coordinate, so the Jacobian
of $\theta$ with respect to $\eta$ is **diagonal**. Diagonality is what makes
the general multivariate chain rule collapse into a product of univariate ones,
and we now work this out in stages, starting with a single parameter.

### One parameter {#sec-faa-univariate}

Composing $\ell$ with $h$ and differentiating repeatedly is the territory of
Faà di Bruno's formula, which organises the combinatorial explosion of the
chain rule through the *partial Bell polynomials* $B_{m,j}$.

::: {#thm-faa}
## Faà di Bruno

If $\ell$ is $m$ times differentiable in $\theta$ and $h$ is $m$ times
differentiable in $\eta$, with $\theta = h(\eta)$, then

$$
\frac{\partial^{m}}{\partial\eta^{m}}\,\ell(h(\eta))
= \sum_{j=1}^{m} \ell^{(j)}(\theta)\;
  B_{m,j}\!\left(h', h'', \dots, h^{(m-j+1)}\right).
$$
:::

For the orders the stack uses, the Bell polynomials are small enough to list in
full,

$$
\begin{array}{llll}
B_{1,1} = h' \\
B_{2,1} = h'' & B_{2,2} = (h')^{2} \\
B_{3,1} = h''' & B_{3,2} = 3h'h'' & B_{3,3} = (h')^{3} \\
B_{4,1} = h'''' & B_{4,2} = 4h'h''' + 3(h'')^{2} & B_{4,3} = 6(h')^{2}h'' & B_{4,4} = (h')^{4}
\end{array}
$$

and substituting them into the theorem gives the four formulas that the package
applies:

$$
\begin{aligned}
\frac{\partial \ell}{\partial \eta} &= \ell_{\theta}\,h', \\[4pt]
\frac{\partial^{2} \ell}{\partial \eta^{2}} &= \ell_{\theta\theta}\,(h')^{2} + \ell_{\theta}\,h'', \\[4pt]
\frac{\partial^{3} \ell}{\partial \eta^{3}} &= \ell_{\theta\theta\theta}\,(h')^{3} + 3\,\ell_{\theta\theta}\,h'h'' + \ell_{\theta}\,h''', \\[4pt]
\frac{\partial^{4} \ell}{\partial \eta^{4}} &= \ell_{\theta\theta\theta\theta}\,(h')^{4} + 6\,\ell_{\theta\theta\theta}\,(h')^{2}h'' + \ell_{\theta\theta}\left(4h'h''' + 3(h'')^{2}\right) + \ell_{\theta}\,h''''.
\end{aligned}
$$

The first two lines are easily checked by hand — the first is the chain rule,
and the second follows from differentiating it as a product — and they exhibit
the property that makes the whole construction generic: each link-scale
derivative is a *linear combination* of the parameter-scale derivatives of
order up to $m$, with coefficients that depend only on the link. The
transformation never needs to know which distribution it is transforming, which
is why it can be written once, in the body of the generic, for all
distributions at once.

### Several parameters {#sec-faa-multivariate}

With several parameters, a mixed derivative such as
$\partial^4 \ell / \partial\eta_a^2\,\partial\eta_b^2$ must be handled, and here
diagonality earns its keep.

::: {#prp-faa-diag}
## Faà di Bruno with a diagonal Jacobian

Let the multi-index involve the distinct parameters $p_1, \dots, p_r$ with
multiplicities $m_1, \dots, m_r$, and let $k = \sum_t m_t$. Then

$$
\frac{\partial^{k} \ell}{\partial \eta_{p_1}^{m_1} \cdots \partial \eta_{p_r}^{m_r}}
= \sum_{j_1=1}^{m_1} \cdots \sum_{j_r=1}^{m_r}
   \ell_{\,p_1^{j_1} \cdots p_r^{j_r}}
   \prod_{t=1}^{r} B_{m_t, j_t}\!\left(h_{p_t}', h_{p_t}'', \dots\right),
$$

where $\ell_{\,p_1^{j_1}\cdots p_r^{j_r}}$ denotes the parameter-scale
derivative taken $j_t$ times with respect to $\theta_{p_t}$.
:::

::: {.proof}
The key observation is that $\theta_{p_t} = h_{p_t}(\eta_{p_t})$ depends on its
own linear predictor and on nothing else, so the operator
$\partial^{m_t}/\partial\eta_{p_t}^{m_t}$ sees $\ell$ as a function of the
single variable $\theta_{p_t}$, all other parameters held fixed. We may
therefore apply @thm-faa in the variable $\eta_{p_1}$ alone, obtaining

$$
\frac{\partial^{m_1}\ell}{\partial\eta_{p_1}^{m_1}}
= \sum_{j_1=1}^{m_1} \ell_{\,p_1^{j_1}}\, B_{m_1,j_1}\!\left(h_{p_1}', \dots\right),
$$

in which the coefficients $B_{m_1, j_1}$ involve only the derivatives of
$h_{p_1}$ and therefore behave as constants under differentiation with respect
to any *other* linear predictor. Applying the theorem again in
$\eta_{p_2}$, then $\eta_{p_3}$, and so on — the operators commute, acting as
they do on disjoint variables — each application multiplies in its own Bell
polynomial factor and deepens the sum by one index, and after $r$ rounds the
stated nested sum is reached. $\square$
:::

Two special cases are worth extracting because they cover most of what happens
in practice. A purely mixed second derivative across two different parameters
involves $m_1 = m_2 = 1$, each inner sum has a single term, and the formula
collapses to a rescaling,

$$
\frac{\partial^{2}\ell}{\partial\eta_a \partial\eta_b} = \ell_{ab}\,h_a'\,h_b'
\qquad (a \neq b).
$$

And the number of terms in the general sum is $\prod_t m_t$ — at most eight for
the orders the stack supports — so the computation remains trivial even at
fourth order.

### The expected Hessian transforms more simply {#sec-link-scale-expected}

For expected second derivatives something pleasant happens: the awkward
first-order term drops out on its own.

::: {#cor-congruence}
## Congruence

$$\mathbb{E}\!\left[\frac{\partial^{2}\ell}{\partial\eta_a\partial\eta_b}\right]
= h_a'\,h_b'\;\mathbb{E}\!\left[\ell_{ab}\right],$$

or in matrix form
$\mathbb{E}[H_\eta] = \operatorname{diag}(h')\,\mathbb{E}[H_\theta]\,\operatorname{diag}(h')$.
:::

::: {.proof}
For $a = b$, the univariate formula gives
$\partial^2\ell/\partial\eta_a^2 = \ell_{aa}(h_a')^2 + \ell_a h_a''$. Take
expectations: the first Bartlett identity says $\mathbb{E}[\ell_a] = 0$, so the
second term vanishes and only the congruence term survives. For $a \neq b$
there was no first-order term to begin with. $\square$
:::

The corollary matters beyond elegance. A congruence transform of a positive
semi-definite matrix is positive semi-definite, so the link-scale Fisher
information inherits the definiteness of the parameter-scale one
automatically. Had one instead transformed the *observed* Hessian and then
averaged numerically, the $\ell_a h_a''$ term would have survived and that
guarantee would have been lost.

### Where the transformation lives {#sec-link-scale-impl}

The change of scale is implemented once, in the body of each derivative
generic, between the dispatch and the return: a method always computes
parameter-scale values, and the generic transforms them if `scale = "link"` was
requested.

``` r
distrib_gradient <- S7::new_generic(
  "distrib_gradient", "distrib",
  function(distrib, y, theta, scale = c("parameter", "link"), ...) {
    ...
    res <- S7::S7_dispatch()                    # parameter scale, always
    if (scale == "link") to_link_scale(...) else res
  })
```

This placement has one visible consequence for users who write their own
derivative methods: since S7 requires a method's formal arguments to include
the generic's named arguments, such a method must declare
`scale = c("parameter", "link"), ...` in its signature even though it never
touches `scale` itself. Omitting the declaration makes dispatch fail.

## Derivatives with respect to the response {#sec-response-derivatives}

Besides derivatives in the parameters, continuous distributions provide
$\partial\ell/\partial y$ and $\partial^{2}\ell/\partial y^{2}$ through
`distrib_grad_y()` and `distrib_hess_y()`; for a lattice variable there is
nothing to differentiate, so discrete distributions have no such methods. These
quantities feed quantile residuals and smoothness criteria further up the
stack. The default implementation is a central finite difference of the
log-density with a step that shrinks near the boundary of the support, so that
$y \pm \delta$ never leaves it.

```{r}
#| label: grad-y-demo
d <- gamma_distrib()
th <- list(mu = 3, sigma2 = 2)
distrib_grad_y(d, c(1, 3, 6), th)
# closed form: with shape a = mu^2/sigma2 and rate b = mu/sigma2,
# dl/dy = (a - 1)/y - b
(3^2 / 2 - 1) / c(1, 3, 6) - 3 / 2
```

## The catalogue {#sec-distrib-catalogue}

The table below is generated from the distribution objects themselves.

```{r}
#| label: catalogue-table
#| echo: false
knitr::kable(catalogue_table(), align = "lllll", row.names = FALSE,
             caption = "The fourteen distributions")
```

The individual entries follow. Several of the parametrisations are not the ones
the corresponding `d*` function in base R uses — the Gamma is parametrised by
mean and variance rather than shape and rate, the Beta by mean and precision,
the negative binomial in its NB2 form — and in each such case the entry states
the map between the two explicitly, since this is where misreadings are
cheapest to prevent. All fourteen distributions implement their observed
derivatives to fourth order in closed form, as compiled Rcpp kernels; the
continuous-integration matrix builds them with several compilers on several
platforms, which is the only real evidence one can have about compiled
arithmetic.

```{r}
#| label: catalogue-entries
#| results: asis
#| echo: false
for (id in names(DISTRIBS)) render_distrib_entry(id, DISTRIBS[[id]])
```

## What the fallbacks provide {#sec-fallbacks}

It is worth describing what a distribution receives for free when it implements
nothing but its density, both because users rely on it and because two of the
fallbacks contain ideas of independent interest.

The cdf is obtained by quadrature for continuous distributions and by summation
over the support for discrete ones, and the quantile function by root-finding
on the cdf. Derivatives fall back to finite differences, each order
differencing the analytical implementation of the order below when one exists.

The random generator deserves a longer story. The obvious fallback, inverse
transform sampling, costs one root-finding call per draw when the quantile
function is itself numerical, which is unusably slow. The package instead uses
a generalized **ratio-of-uniforms** scheme, which needs only the density, with
two devices that make it reliable: the kernel is recentred at the mode of the
density — without this, a density centred at $\mu = 1000$ produces a degenerate
bounding box — and normalised to a maximum of one. So equipped, the method
handles far more than one might expect: bimodal densities, a Student $t$ with
half a degree of freedom, even a Pareto with infinite mean. Its one genuine
failure mode is a density that diverges at an edge of its support, and that
case is transformed away rather than given up on: if
$f(y) \sim |y-a|^{\alpha-1}$ near the edge $a$, then the transformed variable
$X = |Y-a|^{1/\lambda}$ has a density behaving like $x^{\lambda\alpha-1}$,
which is bounded as soon as $\lambda\alpha > 1$. Pleasantly, the exponent
$\alpha$ never has to be searched for: walking toward the edge in powers of ten
raises the log-density by $(1-\alpha)\log 10$ per step, so the same probe that
detects the divergence also measures it. For a Gamma with shape $0.4$ this
transformation brought the cost of a draw from 27 milliseconds down to less
than a microsecond, with the draws still exact.

One caveat belongs here rather than in a footnote. A distribution that relies
on the numerical fallbacks will pass the derivative checks of
`check_distrib()` trivially, because the analytical and the numerical value are
then the same computation; those checks become informative exactly when a
closed form has been supplied.

## When the model is not regular {#sec-nonregular}

The Laplace distribution, with density
$f(y) = (2b)^{-1} e^{-|y-\mu|/b}$, is the catalogue's standing counterexample
to regularity: its log-likelihood has a kink at $y = \mu$. Away from the kink
one computes

$$\ell_\mu = \frac{\operatorname{sign}(y-\mu)}{b}, \qquad \ell_{\mu\mu} = 0,$$

so the observed information in $\mu$ vanishes identically — the log-likelihood
is piecewise linear in the location — and yet the Fisher information certainly
exists. The second Bartlett identity delivers it directly: since
$\operatorname{sign}(y-\mu)^2 = 1$ almost surely,

$$\mathcal{I}_{\mu\mu} = \mathbb{E}\left[\ell_\mu^{2}\right]
= \mathbb{E}\left[\frac{\operatorname{sign}(y-\mu)^{2}}{b^{2}}\right]
= \frac{1}{b^{2}}.$$

The distribution the package ships is not affected by any of this in daily
use, because `laplace_distrib()` implements the expected Hessian in closed
form — the score-variance value $-1/b^{2}$ just computed — and, as always when
an analytical method exists, the `approx` argument is simply ignored:

```{r}
#| label: laplace-nonregular
lap <- laplace_distrib()
th <- list(mu = 0, b = 2)
c(bartlett  = distrib_expected_hessian(lap, 0, th, approx = "bartlett")$mu_mu,
  integrate = distrib_expected_hessian(lap, 0, th, approx = "integrate")$mu_mu,
  exact     = -1 / th$b^2)
```

To see the three strategies of @sec-approx genuinely part company, one has to
look at a distribution that carries *no* closed form for the expectation — the
situation a user-defined non-regular family finds itself in. Here is a bare
Laplace, supplying only the density, the score and the observed Hessian:

```{r}
#| label: laplace-bare
BareLaplace <- S7::new_class("BareLaplace", parent = continuous_distrib, package = NULL)
S7::method(distrib_pdf, BareLaplace) <- function(distrib, y, theta, log = FALSE) {
  ld <- -log(2 * theta[[2]]) - abs(y - theta[[1]]) / theta[[2]]
  if (log) ld else exp(ld)
}
S7::method(distrib_gradient, BareLaplace) <- function(distrib, y, theta,
                                                      scale = c("parameter", "link"), ...) {
  r <- y - theta[[1]]; b <- theta[[2]]
  list(mu = sign(r) / b, b = (abs(r) / b - 1) / b)
}
S7::method(distrib_hessian, BareLaplace) <- function(distrib, y, theta,
                                                     scale = c("parameter", "link"), ...) {
  r <- y - theta[[1]]; b <- theta[[2]]; n <- length(y)
  list(mu_mu = rep(0, n), b_b = (b - 2 * abs(r)) / b^3, mu_b = -sign(r) / b^2)
}
bare <- BareLaplace(
  distrib_name = "bare laplace", dimension = "univariate", bounds = c(-Inf, Inf),
  params = c("mu", "b"), params_interpretation = c(mu = "location", b = "scale"),
  n_params = 2, params_bounds = list(mu = c(-Inf, Inf), b = c(0, Inf)),
  link_params = list(mu = identity_link(), b = log_link()),
  params_smooth = c(mu = FALSE, b = TRUE)
)

set.seed(1)
c(bartlett  = distrib_expected_hessian(bare, 0, th, approx = "bartlett")$mu_mu,
  integrate = distrib_expected_hessian(bare, 0, th, approx = "integrate")$mu_mu,
  mc        = distrib_expected_hessian(bare, 0, th, approx = "mc")$mu_mu,
  exact     = -1 / th$b^2)
```

Now the strategies disagree, and the pattern of their disagreement tells the
whole story. The Bartlett route recovers $-1/b^{2}$ exactly, because it works
from the score alone and never touches the second derivative. The other two
both return zero — and for the same reason: each of them averages the observed
$\ell_{\mu\mu}$, by quadrature in one case and over simulated draws in the
other, and that function is identically zero away from the single point
$y = \mu$, a set the integral cannot see and the draws almost surely never hit.
Averaging the second derivative, however the average is taken, produces the
expected *derivative*, which for this family is genuinely zero; it is the
identification of that quantity with the information which breaks down, and
only the strategy built on the score survives the breakage.

The base class records which parameters are differentiable in the
`params_smooth` property, and this single flag propagates through the stack:
`fit_distrib()` works on the Laplace precisely because Fisher scoring uses the
expected information, which exists, rather than the observed one, which is
degenerate.

```{r}
#| label: params-smooth
param_smoothness(laplace_distrib())
```

The flag has a second, subtler use. Comparing an analytical derivative against
a finite difference is invalid at a kink — a central difference straddling
$y = \mu$ converges to nothing, so it is the *reference* that is wrong, not the
value under test. When a distribution declares a non-smooth parameter, the
validation machinery recomputes each finite-difference reference with the step
halved and discards the observations where the two estimates disagree, judging
the disagreement relative to the estimates' own magnitude; observations near
the kink are thereby excluded from the comparison, while for a smooth
log-likelihood nothing is ever discarded and the check keeps its full strength.

```{r}
#| label: consistency-distributions
#| include: false
# Internal consistency gate: fails the render if any density printed in this
# chapter disagrees with the package, if a Bartlett identity fails, if the
# link-scale machinery disagrees with finite differences in eta, or if
# check_distrib() fails anywhere in the catalogue.
assert_distributions_ok()
```
