# Generalized linear models {#sec-glm}

```{r}
#| label: setup-04
#| include: false
source("R/_setup.R")
source("R/glm-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

A generalized linear model has a response with a distribution from the
exponential family, a mean that is a function of a linear predictor, and a
variance that is a function of the mean. The function `glm()` fits these
models, and `statmod()` fits them too, for these and for other
distributions. This chapter takes the usual cases one by one: counts,
proportions and binary responses, positive measurements, and counts that vary
more than the Poisson distribution allows. For each case we fit the model with
`statmod()` and compare the result with `glm()` or, for the negative binomial
distribution, with `MASS::glm.nb()`. Two differences appear in every
comparison, and we state them here once.

Not every model of the chapter is a generalized linear model in the strict
sense. The Poisson, binomial, Bernoulli, gamma and inverse gaussian
distributions belong to the exponential family, and with them `statmod()` fits
the models of `glm()`. The negative binomial distribution belongs to the
family only when $\theta$ is known, and `glm.nb()` treats it as a
generalized linear model with $\theta$ estimated in an outer loop. The
distributions `negbin1_distrib()`, `pig1_distrib()`, `pig2_distrib()` and
`betabinom1_distrib()` are not in the exponential family for any value of
their dispersion, and a model with an equation for the dispersion, as in
@sec-glm-dispersion, is an extension of the generalized linear model as well.
We keep these models in this chapter because they answer the same question,
the variance of counts and proportions, and because they use the same
formulas and the same link functions. The text says each time when a model
goes beyond `glm()`.

First, some distributions have a second parameter, the dispersion. `glm()`
estimates it by a moment estimator, dividing the Pearson statistic by the
residual degrees of freedom. `statmod()` estimates it as a parameter of the
distribution, by maximizing a criterion, and the estimate is a little
different. Second, `statmod()` evaluates the log-likelihood at its own
estimates, so the log-likelihood and the AIC that it reports can differ from
those of `glm()`. When the distribution has no dispersion, as the Poisson and
the binomial distributions, the two functions give the same numbers.

## Counts and rates {#sec-glm-counts}

The data frame `ships` from the `MASS` package records damage incidents to
cargo ships [@mccullagh1989]. Each row is one combination of three factors:
the type of the ship, from A to E, the period of construction and the period
of operation. The response `incidents` is the number of damage incidents in
the row, and `service` is the number of months of service of the ships in the
row.

```{r}
#| label: ships-data
data(ships, package = "MASS")
ships <- subset(ships, service > 0)
ships$year <- factor(ships$year)
ships$period <- factor(ships$period)
head(ships, n = 4)
```

We keep the `r nrow(ships)` rows with some months of service. When the
factors are the same, a row with twice the months of service has on average
twice as many incidents. So we model the rate of incidents per month of
service. Let $y_i$ be the number of incidents in row $i$ and $t_i$ its months
of service. The model is
$$
y_i \sim \mathrm{Poisson}(\mu_i), \qquad
\log \mu_i = \log t_i + x_i^\top \beta ,
$$
where $x_i$ holds the intercept and the indicators of the three factors. The
term $\log t_i$ has a coefficient fixed at one, and it is called an offset.
It turns the model for the mean count $\mu_i$ into a model for the rate
$\mu_i / t_i$. We wrote an offset in @sec-beyond-weights, and here we use it
as in `glm()`:

```{r}
#| label: fit-ships
fit_ships <- statmod(incidents ~ type + year + period + offset(log(service)),
                     distrib = poisson_distrib(), data = ships)
glm_ships <- glm(incidents ~ type + year + period + offset(log(service)),
                 family = poisson, data = ships)
summary(fit_ships)
```

The distribution `poisson_distrib()` has one parameter, the mean `mu`, with
the log link. The model has no dispersion, so the estimates are those of
`glm()`:

```{r}
#| label: ships-glm
coef(fit_ships)$mu
coef(glm_ships)
all.equal(target = unname(vcov(glm_ships)), current = unname(vcov(fit_ships)),
          tolerance = 1e-3)
logLik(fit_ships)
logLik(glm_ships)
AIC(fit_ships)
AIC(glm_ships)
```

The coefficients, the variance matrix, the log-likelihood and the AIC agree
up to the precision of the two optimizers. The $z$ values and the p-values
also agree: for a Poisson response `glm()` compares the Wald statistic with
the standard gaussian distribution, as `statmod()` does.

### Rate ratios

A coefficient of a log-linear model is the logarithm of a ratio of rates.
Each coefficient of a factor compares one level with the first level of the
factor, at equal values of the other factors. We exponentiate the estimates
and the ends of the intervals of `confint()`:

```{r}
#| label: ships-ratios
ci_ships <- confint(fit_ships)
ci_ships <- ci_ships[ci_ships$coefficient != "(Intercept)", ]
ratios <- data.frame(ratio = exp(ci_ships$estimate),
                     lower = exp(ci_ships$lower),
                     upper = exp(ci_ships$upper),
                     row.names = ci_ships$coefficient)
ratios
```

```{r}
#| label: fig-ships
#| fig-cap: "Rate ratios of damage incidents with 95% intervals, on a log scale. Each ratio compares a level with the first level of its factor: type A, construction in 1960--64 and operation in 1960--74."
#| echo: false
#| fig-height: 3.4
lab <- c(typeB = "type B", typeC = "type C", typeD = "type D",
         typeE = "type E", year65 = "built 1965-69", year70 = "built 1970-74",
         year75 = "built 1975-79", period75 = "operated 1975-79")
rr <- data.frame(ratios,
                 term = factor(lab[rownames(ratios)],
                               levels = rev(lab[rownames(ratios)])))
ggplot(rr, aes(x = ratio, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = book_grey) +
  geom_linerange(aes(xmin = lower, xmax = upper), colour = chalk,
                 linewidth = 0.8) +
  geom_point(colour = chalk, size = 2) +
  scale_x_log10() +
  labs(x = "rate ratio", y = NULL)
```

Type C has the smallest rate, `r fmt(ratios["typeC", "ratio"])` times the
rate of type A, and type E has the largest, `r fmt(ratios["typeE", "ratio"])`
times. Ships built in 1965--69 have `r fmt(ratios["year65", "ratio"])` times
the rate of ships built in 1960--64, and operation in 1975--79 multiplies the
rate by `r fmt(ratios["period75", "ratio"])`.

### Is the variance equal to the mean?

The Poisson distribution has variance equal to its mean. The Pearson
statistic $X^2 = \sum_i (y_i - \hat\mu_i)^2/\hat\mu_i$ compares the squared
residuals with this variance. If the model is right, $X^2$ divided by the
residual degrees of freedom is close to one:

```{r}
#| label: ships-pearson
mu_ships <- predict(fit_ships, what = "mu", newdata = ships)
x2_ships <- sum((ships$incidents - mu_ships)^2 / mu_ships)
df_ships <- nrow(ships) - length(coef(fit_ships)$mu)
x2_ships
df_ships
x2_ships / df_ships
```

The ratio is `r fmt(x2_ships / df_ships)`. A ratio above one means that the
counts vary more than the Poisson distribution allows: here they vary about
`r fmt(x2_ships / df_ships, 1)` times as much, and the standard errors of
the Poisson model are too small by about the square root of this ratio,
`r fmt(sqrt(x2_ships / df_ships))`. The next sections use distributions with a
variance that is larger than the mean, and a data set where the ratio is
much larger.

## Proportions and binary responses {#sec-glm-binary}

### A number of successes out of a number of trials

In a dose-response experiment a number of subjects receive a dose, and we
count how many of them react. The table below is the classic experiment on
the tobacco budworm [@collett1991; @venables2002]. Batches of 20 moths of
each sex received one of six doses of a poison, and the response is the
number of dead moths after three days. The dose is coded by its base-2
logarithm, `ldose`.

```{r}
#| label: budworm-data
budworm <- data.frame(
  ldose = rep(0:5, times = 2),
  sex = factor(rep(c("M", "F"), each = 6)),
  numdead = c(1, 4, 9, 13, 18, 20, 0, 2, 6, 10, 12, 16),
  n = 20)
budworm
```

Let $y_i$ be the number of dead moths in batch $i$, out of $m_i = 20$. The
model is
$$
y_i \sim \mathrm{Binomial}(m_i, p_i), \qquad
\mathrm{logit}\,p_i = \beta_0 + \beta_1 s_i + \beta_2 d_i + \beta_3 s_i d_i ,
$$
where $s_i$ is 1 for males and $d_i$ is the log-dose. In `statmod()` the
parameter `mu` of `binomial_distrib()` is the probability $p_i$. The response
is the two-column matrix of successes and failures, as in `glm()`, and the
number of trials of each row is the sum of its two entries:

```{r}
#| label: fit-budworm
fit_bw <- statmod(cbind(numdead, n - numdead) ~ sex * ldose,
                  distrib = binomial_distrib(), data = budworm)
glm_bw <- glm(cbind(numdead, n - numdead) ~ sex * ldose, family = binomial,
              data = budworm)
summary(fit_bw)
coef(glm_bw)
logLik(fit_bw)
logLik(glm_bw)
```

The estimates agree with those of `glm()`. The coefficient of `ldose` is the
increase of the log odds of death for one more unit of log-dose in the
reference sex, females: `r fmt(coef(fit_bw)$mu[["ldose"]], 3)`. The slope of
the males is larger by the coefficient of `sexM:ldose`,
`r fmt(coef(fit_bw)$mu[["sexM:ldose"]], 3)`, but this interaction is not
significant (its p-value is `r fmt(summary(fit_bw)@tables$mu[[1]]$table$p_value[4], 2)`), so the data do not show a
difference between the slopes of the two sexes. We drop the interaction in the
next comparison, where we look at the link.

### The link

The logit link is not the only link that maps a probability onto the real
line. The probit link $\Phi^{-1}(p)$ uses the quantile function of the
standard gaussian distribution, and the complementary log-log link
$\log(-\log(1 - p))$ is asymmetric: it approaches 0 slowly and 1 quickly. The
distribution takes the link as the argument `link_mu`. We fit the additive
model with the three links:

```{r}
#| label: fit-links
fit_bw_logit <- statmod(cbind(numdead, n - numdead) ~ sex + ldose,
                        distrib = binomial_distrib(link_mu = logit_link()),
                        data = budworm)
fit_bw_probit <- statmod(cbind(numdead, n - numdead) ~ sex + ldose,
                         distrib = binomial_distrib(link_mu = probit_link()),
                         data = budworm)
fit_bw_cloglog <- statmod(cbind(numdead, n - numdead) ~ sex + ldose,
                          distrib = binomial_distrib(link_mu = cloglog_link()),
                          data = budworm)
link_table <- data.frame(
  link = c("logit", "probit", "cloglog"),
  logLik = c(as.numeric(logLik(fit_bw_logit)),
             as.numeric(logLik(fit_bw_probit)),
             as.numeric(logLik(fit_bw_cloglog))),
  AIC = c(AIC(fit_bw_logit), AIC(fit_bw_probit), AIC(fit_bw_cloglog)))
link_table
```

The three models have the same number of parameters, so their log-likelihoods
compare directly. The probit link has the largest log-likelihood, but the
differences are small: with only 12 batches the data cannot separate the
links well. @fig-links shows why. The three curves are close in the middle of
the range of doses and differ at the ends.

```{r}
#| label: fig-links
#| fig-cap: "Proportion of dead male moths against the log-dose, with the fitted probability of the additive model under three links."
#| echo: false
#| fig-height: 3.8
grid <- data.frame(ldose = seq(from = 0, to = 5, length.out = 100),
                   sex = factor("M", levels = levels(budworm$sex)),
                   numdead = 0, n = 20)
p_logit <- predict(fit_bw_logit, what = "mu", newdata = grid)
p_probit <- predict(fit_bw_probit, what = "mu", newdata = grid)
p_cloglog <- predict(fit_bw_cloglog, what = "mu", newdata = grid)
males <- budworm$sex == "M"
links <- c("logit", "probit", "cloglog")
bw_obs <- data.frame(ldose = budworm$ldose[males],
                     prop = budworm$numdead[males] / budworm$n[males])
bw_fit <- data.frame(ldose = rep(grid$ldose, times = 3),
                     prop = c(p_logit, p_probit, p_cloglog),
                     link = factor(rep(links, each = nrow(grid)),
                                   levels = links))
ggplot(bw_obs, aes(x = ldose, y = prop)) +
  geom_point(aes(shape = "observed proportion"), colour = book_grey,
             size = 1.8) +
  geom_line(aes(colour = link, linetype = link), data = bw_fit,
            linewidth = 0.8) +
  scale_shape_manual(values = 16) +
  scale_colour_manual(values = c(chalk, rust, "grey15")) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 1)) +
  guides(shape = guide_legend(order = 1), colour = guide_legend(order = 2),
         linetype = guide_legend(order = 2)) +
  labs(x = "log-dose", y = "proportion dead")
```

### A binary response

When every row is one subject, the response is 0 or 1 and the distribution
is the Bernoulli distribution, which is the binomial distribution with one
trial. The data frame `birthwt` from the `MASS` package has the birth weight
of 189 babies, and `low` is 1 when the weight is below 2.5 kg
[@hosmer1989]. The covariates are the age and the weight of the mother at
the last menstrual period, `smoke` for smoking during the pregnancy and `ht`
for a history of hypertension:

```{r}
#| label: fit-birthwt
data(birthwt, package = "MASS")
fit_bwt <- statmod(low ~ age + lwt + smoke + ht,
                   distrib = bernoulli_distrib(), data = birthwt)
glm_bwt <- glm(low ~ age + lwt + smoke + ht, family = binomial,
               data = birthwt)
summary(fit_bwt)
coef(glm_bwt)
```

The coefficients are those of `glm()`. Each coefficient is a log odds ratio,
so we exponentiate the estimates and the ends of the intervals:

```{r}
#| label: birthwt-or
ci_bwt <- confint(fit_bwt)
ci_bwt <- ci_bwt[ci_bwt$coefficient != "(Intercept)", ]
odds <- data.frame(odds_ratio = exp(ci_bwt$estimate),
                   lower = exp(ci_bwt$lower),
                   upper = exp(ci_bwt$upper),
                   row.names = ci_bwt$coefficient)
odds
```

Smoking multiplies the odds of a low birth weight by
`r fmt(odds["smoke", "odds_ratio"])`, and a history of hypertension by
`r fmt(odds["ht", "odds_ratio"])`. The intervals are Wald intervals, built
on the log scale and mapped back, so they are not symmetric around the odds
ratio. The function `MASS::confint.glm()` computes profile intervals, which
differ from the Wald ones when the sample is small. @sec-beyond-invert gave
the intervals of `statmod()` that invert the likelihood ratio.

The Bernoulli distribution takes the same links. We fit the four links that
the `linkfunctions7` package provides for a probability, with the cauchit
link, the inverse of the Cauchy distribution function, in addition to the
three of the previous section:

```{r}
#| label: birthwt-links
fit_bwt_probit <- statmod(low ~ age + lwt + smoke + ht,
                          distrib = bernoulli_distrib(link_mu = probit_link()),
                          data = birthwt)
fit_bwt_cloglog <- statmod(low ~ age + lwt + smoke + ht,
                           distrib = bernoulli_distrib(link_mu = cloglog_link()),
                           data = birthwt)
fit_bwt_cauchit <- statmod(low ~ age + lwt + smoke + ht,
                           distrib = bernoulli_distrib(link_mu = cauchit_link()),
                           data = birthwt)
bwt_links <- data.frame(
  link = c("logit", "probit", "cloglog", "cauchit"),
  logLik = c(as.numeric(logLik(fit_bwt)), as.numeric(logLik(fit_bwt_probit)),
             as.numeric(logLik(fit_bwt_cloglog)),
             as.numeric(logLik(fit_bwt_cauchit))),
  AIC = c(AIC(fit_bwt), AIC(fit_bwt_probit), AIC(fit_bwt_cloglog),
          AIC(fit_bwt_cauchit)))
bwt_links
```

The four log-likelihoods differ by less than one unit, so the data do not
choose among them. The logit link is the usual choice because its
coefficients are log odds ratios.

## Positive responses {#sec-glm-positive}

The data frame `GAGurine` from the `MASS` package gives the concentration of
glycosaminoglycans (GAG) in the urine of `r nrow(MASS::GAGurine)` children
aged from 0 to 17 years [@venables2002]. The concentration is positive, and
its spread is larger where its mean is larger. The gamma distribution has
mean $\mu$ and variance $\phi\mu^2$, so its standard deviation is
proportional to the mean, and $\phi$ is the squared coefficient of variation.
We model the mean with a log link, as a straight line in the age:
$$
y_i \sim \mathrm{Gamma}(\mu_i, \phi), \qquad
\log\mu_i = \beta_0 + \beta_1 a_i .
$$

```{r}
#| label: fit-gamma
data(GAGurine, package = "MASS")
fit_gam <- statmod(GAG ~ Age, distrib = gamma1_distrib(), data = GAGurine)
glm_gam <- glm(GAG ~ Age, family = Gamma(link = "log"), data = GAGurine)
summary(fit_gam)
coef(glm_gam)
```

The distribution `gamma1_distrib()` has two parameters, the mean `mu` and
the dispersion `phi`, both with the log link. The formula has no equation for
`phi`, so `statmod()` estimates it as a constant. The coefficients of the
mean are those of `glm()`. The dispersion is not:

```{r}
#| label: gamma-phi
n_gag <- nrow(GAGurine)
fit_gam_ml <- statmod(GAG ~ Age, distrib = gamma1_distrib(), data = GAGurine,
                      outer_criterion = ml())
phi_table <- data.frame(
  phi = c(exp(coef(fit_gam)$phi), exp(coef(fit_gam_ml)$phi),
          summary(glm_gam)$dispersion, glm_gam$deviance / n_gag,
          1 / MASS::gamma.shape(glm_gam)$alpha),
  row.names = c("statmod, default", "statmod, ml()", "glm, Pearson",
                "glm, deviance / n", "MASS::gamma.shape"))
phi_table
```

Four of the estimates are close to each other, and the Pearson estimate of
`glm()` is larger. The maximum likelihood estimate is
`r fmt(phi_table["statmod, ml()", "phi"], 4)`. It is the value of
`MASS::gamma.shape()`, and it is the estimate of `statmod()` with the
criterion `ml()`. The default criterion of `statmod()` is REML. It
integrates the two coefficients of the mean out of the likelihood, and for a
gamma response its estimate is close to the maximum likelihood estimate times
$n/(n - 2)$, as the two numbers below show:

```{r}
#| label: gamma-reml
phi_table["statmod, ml()", "phi"] * n_gag / (n_gag - 2)
phi_table["statmod, default", "phi"]
```

The estimate of `glm()` is different in kind. It is the moment estimate, the
Pearson statistic divided by $n - 2$, and it is
`r fmt(phi_table["glm, Pearson", "phi"], 4)`, larger than the others. The standard errors of the coefficients of the mean are proportional
to $\sqrt\phi$, so `glm()` reports larger standard errors than `statmod()`,
in the ratio of the square roots of the two estimates:

```{r}
#| label: gamma-se
se_gam <- sqrt(diag(vcov(fit_gam)))[["mu:Age"]]
se_glm <- summary(glm_gam)$coefficients["Age", "Std. Error"]
c(statmod = se_gam, glm = se_glm, ratio = se_glm / se_gam)
sqrt(phi_table["glm, Pearson", "phi"] / phi_table["statmod, default", "phi"])
```

The log-likelihood depends on the dispersion too. `logLik()` of `statmod()`
evaluates the gamma density at its estimate of $\phi$. `logLik()` of `glm()`
evaluates it at the deviance divided by $n$, which approximates the maximum
likelihood estimate. Both count three parameters, the two coefficients and
$\phi$:

```{r}
#| label: gamma-loglik
logLik(fit_gam)
logLik(fit_gam_ml)
logLik(glm_gam)
```

### The inverse gaussian distribution

When the standard deviation grows faster with the mean, the inverse gaussian
distribution, with variance $\phi\mu^3$, is an alternative. The two models
describe the same response, so we compare them through the AIC:

```{r}
#| label: fit-ig
fit_ig <- statmod(GAG ~ Age, distrib = invgauss1_distrib(), data = GAGurine)
c(gamma = AIC(fit_gam), inverse_gaussian = AIC(fit_ig))
```

The gamma distribution has the smaller AIC, by `r fmt(AIC(fit_ig) - AIC(fit_gam), 1)`
units. The linear log-mean is a restriction that we can relax. A smooth
function of the age gives a much better fit:

```{r}
#| label: fit-gamma-smooth
fit_gam_s <- statmod(GAG ~ s(Age, bspline_smooth(k = 10)),
                     distrib = gamma1_distrib(), data = GAGurine)
c(linear = AIC(fit_gam), smooth = AIC(fit_gam_s))
```

The effective number of parameters of the smooth model is
`r fmt(sum(fit_gam_s@edf$edf), 1)`, against 3 for the linear model.
@fig-gag shows that the curve is steeper for the youngest children than a
straight line on the log scale can be. The smooth terms are the subject of a
later chapter.

```{r}
#| label: fig-gag
#| fig-cap: "GAG concentration against age, with three fitted means: gamma with a linear log-mean, inverse gaussian with a linear log-mean, and gamma with a smooth log-mean."
#| echo: false
#| fig-height: 3.8
ga <- data.frame(Age = seq(from = 0, to = 17, length.out = 200))
m_gam <- predict(fit_gam, what = "mu", newdata = ga)
m_ig <- predict(fit_ig, what = "mu", newdata = ga)
m_gs <- predict(fit_gam_s, what = "mu", newdata = ga)
gag_models <- c("gamma, linear", "inverse gaussian, linear", "gamma, smooth")
gag_fit <- data.frame(Age = rep(ga$Age, times = 3),
                      GAG = c(m_gam, m_ig, m_gs),
                      model = factor(rep(gag_models, each = nrow(ga)),
                                     levels = gag_models))
ggplot(GAGurine, aes(x = Age, y = GAG)) +
  geom_point(aes(shape = "data"), colour = book_grey, size = 1.1) +
  geom_line(aes(colour = model, linetype = model), data = gag_fit,
            linewidth = 0.8) +
  scale_shape_manual(values = 16) +
  scale_colour_manual(values = c(chalk, rust, "grey15")) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
  guides(shape = guide_legend(order = 1), colour = guide_legend(order = 2),
         linetype = guide_legend(order = 2)) +
  labs(x = "age (years)", y = "GAG")
```

## Counts that vary too much {#sec-glm-overdispersion}

The data frame `quine` from the `MASS` package records the days of absence
from school in one year for `r nrow(MASS::quine)` children in a town in New
South Wales [@aitkin1978; @venables2002]. The covariates are the ethnic
background of the child (`Eth`), the sex, the age group and the learner
status (`Lrn`). The mean number of days is
`r fmt(mean(MASS::quine$Days), 1)` and the variance is
`r fmt(var(MASS::quine$Days), 1)`, so the counts vary far more than a
Poisson distribution allows.

```{r}
#| label: quine-data
data(quine, package = "MASS")
formula_q <- Days ~ Eth + Sex + Age + Lrn
fit_q_pois <- statmod(formula_q, distrib = poisson_distrib(), data = quine)
```

### Four distributions for the variance

Four distributions describe counts with a variance that is larger than the
mean. Each has a mean $\mu$ and one dispersion parameter, and each gives the
variance as a different function of the mean:

| distribution in `statmod()` | parameter | variance |
|:----------------------------|:----------|:---------------------------|
| `poisson_distrib()` | none | $\mu$ |
| `negbin2_distrib()` | $\theta$ | $\mu + \mu^2/\theta$ |
| `negbin1_distrib()` | $\theta$ | $\mu(1 + \theta)$ |
| `pig1_distrib()`, `pig2_distrib()` | $\sigma$ | $\mu + \sigma\mu^2$ |

Of these, only the Poisson distribution and the negative binomial
distribution with a known $\theta$ are in the exponential family. The other
three are extensions of the generalized linear model: the variance function of
`negbin1_distrib()` is linear in the mean, like that of the quasi-Poisson
model of `glm()` but with a proper distribution and a likelihood, and the
Poisson-inverse gaussian distribution, `pig`, is a mixture of Poisson
distributions with an inverse gaussian mean. It has the same variance function
as the first negative binomial distribution, with a longer right tail. The two
versions of it differ in the parametrization of the dispersion.
We fit the four distributions:

```{r}
#| label: fit-quine-families
fit_q_nb2 <- statmod(formula_q, distrib = negbin2_distrib(), data = quine)
fit_q_nb1 <- statmod(formula_q, distrib = negbin1_distrib(), data = quine)
fit_q_pig1 <- statmod(formula_q, distrib = pig1_distrib(), data = quine)
fit_q_pig2 <- statmod(formula_q, distrib = pig2_distrib(), data = quine)
```

To see whether the variance function of a fitted model is right, we divide
the Pearson statistic by the residual degrees of freedom. For a model with
effective number of parameters $\tau$ the statistic is
$$
\frac{1}{n - \tau} \sum_{i=1}^{n}
\frac{(y_i - \hat\mu_i)^2}{\widehat{\mathrm{Var}}(y_i)} ,
$$
and its value is close to one when the variance function is right
[@mccullagh1989, sec. 4.5]. The function `variance()` returns the variance
of a distribution at given values of its parameters. The function
`pearson_ratio()` below collects the fitted values of the parameters with
`predict()` and applies it:

```{r}
#| label: quine-pearson
pearson_ratio <- function(fit, distrib, data, y) {
  theta <- list()
  for (p in distrib@params) {
    theta[[p]] <- predict(fit, what = p, newdata = data)
  }
  v <- variance(distrib, theta = theta)
  tau <- sum(fit@edf$edf)
  sum((y - theta$mu)^2 / v) / (length(y) - tau)
}
quine_fits <- data.frame(
  distribution = c("Poisson", "negbin2", "negbin1", "pig1", "pig2"),
  AIC = c(AIC(fit_q_pois), AIC(fit_q_nb2), AIC(fit_q_nb1), AIC(fit_q_pig1),
          AIC(fit_q_pig2)),
  pearson = c(
    pearson_ratio(fit_q_pois, poisson_distrib(), quine, quine$Days),
    pearson_ratio(fit_q_nb2, negbin2_distrib(), quine, quine$Days),
    pearson_ratio(fit_q_nb1, negbin1_distrib(), quine, quine$Days),
    pearson_ratio(fit_q_pig1, pig1_distrib(), quine, quine$Days),
    pearson_ratio(fit_q_pig2, pig2_distrib(), quine, quine$Days)))
quine_fits
```

The Poisson model has a ratio of `r fmt(quine_fits$pearson[1], 1)`: the counts
vary about `r fmt(quine_fits$pearson[1], 0)` times as much as the Poisson
variance. The first negative binomial distribution brings the ratio to
`r fmt(quine_fits$pearson[2])`, and it has the smallest AIC. We may compare
the AIC of the five models, although the distributions are different,
because every log-likelihood is the logarithm of the probability that its
model gives to the same observed counts [@akaike1974]. The ratios of the
`pig` models are below one, because their variance function overstates the
variance of these counts.

@fig-quine compares the observed number of children in six classes of
absence with the numbers that the Poisson and the first negative binomial
model expect. For each class we sum over the children the probability that
the fitted distribution gives to the class:

```{r}
#| label: quine-expected
brk <- c(-0.5, 0.5, 5.5, 10.5, 20.5, 40.5, Inf)
cls <- c("0", "1-5", "6-10", "11-20", "21-40", "41+")
observed <- as.numeric(table(cut(quine$Days, breaks = brk)))
mu_pois <- predict(fit_q_pois, what = "mu", newdata = quine)
mu_nb2 <- predict(fit_q_nb2, what = "mu", newdata = quine)
th_nb2 <- predict(fit_q_nb2, what = "theta", newdata = quine)
exp_pois <- numeric(6)
exp_nb2 <- numeric(6)
for (j in 1:6) {
  exp_pois[j] <- sum(ppois(q = brk[j + 1], lambda = mu_pois) -
                       ppois(q = brk[j], lambda = mu_pois))
  exp_nb2[j] <- sum(pnbinom(q = brk[j + 1], size = th_nb2, mu = mu_nb2) -
                      pnbinom(q = brk[j], size = th_nb2, mu = mu_nb2))
}
data.frame(class = cls, observed = observed, Poisson = exp_pois,
           negbin2 = exp_nb2)
```

```{r}
#| label: fig-quine
#| fig-cap: "Number of children in six classes of days of absence: observed (grey), expected under the Poisson model (rust) and expected under the first negative binomial model (green)."
#| echo: false
#| fig-height: 3.4
counts <- c("observed", "Poisson", "negative binomial")
quine_bars <- data.frame(class = factor(rep(cls, times = 3), levels = cls),
                         children = c(observed, exp_pois, exp_nb2),
                         source = factor(rep(counts, each = length(cls)),
                                         levels = counts))
ggplot(quine_bars, aes(x = class, y = children, fill = source)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.85) +
  scale_fill_manual(values = c(book_grey, rust, chalk)) +
  labs(x = "days of absence", y = "children")
```

The Poisson model expects `r fmt(exp_pois[1], 2)` children with no absence,
and there are `r observed[1]`. It expects `r fmt(exp_pois[6], 1)` children
with more than 40 days, and there are `r observed[6]`. The negative binomial
model expects `r fmt(exp_nb2[1], 1)` and `r fmt(exp_nb2[6], 1)` children in
these two classes.

### The dispersion of the negative binomial

The function `MASS::glm.nb()` fits the same negative binomial model. It
alternates the estimation of the coefficients with the maximum likelihood
estimation of $\theta$. The default criterion of `statmod()` is REML, as for
the gamma distribution, and the estimate of $\theta$ is a little smaller.
With the criterion `ml()` the two functions agree:

```{r}
#| label: quine-glmnb
glm_q_nb <- MASS::glm.nb(formula_q, data = quine)
fit_q_nb2_ml <- statmod(formula_q, distrib = negbin2_distrib(), data = quine,
                        outer_criterion = ml())
c(reml = exp(coef(fit_q_nb2)$theta), ml = exp(coef(fit_q_nb2_ml)$theta),
  glm.nb = glm_q_nb$theta)
max(abs(coef(fit_q_nb2_ml)$mu - coef(glm_q_nb)))
max(abs(coef(fit_q_nb2)$mu - coef(glm_q_nb)))
c(statmod_ml = as.numeric(logLik(fit_q_nb2_ml)),
  glm.nb = as.numeric(logLik(glm_q_nb)))
```

The REML estimate of the dispersion removes the bias that comes from
estimating the seven coefficients of the mean, in the same way as in
@sec-lm-linear, where REML gave the residual variance $\mathrm{RSS}/(n - p)$.
The coefficients of the mean move by less than $10^{-3}$.

### Proportions that vary too much

A proportion has the same problem as a count, and the remedy is again an
extension of the generalized linear model and not a member of it: the
beta-binomial distribution is a mixture of binomial distributions, and it is
not in the exponential family. The binomial distribution
with $m$ trials has variance $m\,p(1 - p)$, and a group of trials that share
a common cause varies more than this. The beta-binomial distribution
`betabinom1_distrib()` has the probability $\mu$ and a dispersion $\sigma$,
and its variance is
$$
\mathrm{Var}(Y) = m\mu(1 - \mu)\left(1 + (m - 1)\frac{\sigma}{1 + \sigma}\right) ,
$$
which is the binomial variance multiplied by a factor above one. The factor
$\sigma/(1 + \sigma)$ is the correlation between two trials of the same
group. We draw 200 groups of 20 trials from this distribution with
`rstatmod()`, with the true values $\mu = \mathrm{logit}^{-1}(0.2 + x)$ and
$\sigma = 0.5$, and we fit the binomial and the beta-binomial models:

```{r}
#| label: betabinom-sim
set.seed(11)
xs <- data.frame(x = runif(n = 200, min = -1, max = 1))
sim_bb <- rstatmod(y ~ x | sigma ~ 1, distrib = betabinom1_distrib(size = 20),
                   data = xs, par = list(mu = c(0.2, 1), sigma = log(0.5)))
fit_b_bin <- statmod(y ~ x, distrib = binomial_distrib(size = 20),
                     data = sim_bb$data)
fit_b_bb <- statmod(y ~ x, distrib = betabinom1_distrib(size = 20),
                    data = sim_bb$data)
summary(fit_b_bb)
```

The estimate of $\log\sigma$ is `r fmt(coef(fit_b_bb)$sigma[[1]], 2)` against
the true value $\log 0.5 = `r fmt(log(0.5), 2)`$. The binomial model fits
the same mean with the same kind of coefficients, but its standard errors are
too small, because it treats the 20 trials of a group as independent:

```{r}
#| label: betabinom-se
se_bin <- sqrt(diag(vcov(fit_b_bin)))
se_bb <- sqrt(diag(vcov(fit_b_bb)))[c("mu:(Intercept)", "mu:x")]
cbind(binomial = se_bin, beta_binomial = se_bb, ratio = se_bb / se_bin)
c(binomial = AIC(fit_b_bin), beta_binomial = AIC(fit_b_bb))
```

The ratio of the two standard errors is about 2.6. Its square, 6.6, is of the
same size as the Pearson ratio of the binomial fit, and as the factor
$1 + 19 \cdot 0.5/1.5$ by which the true variance exceeds the binomial one.
We compute the three numbers:

```{r}
#| label: betabinom-pearson
mu_b <- predict(fit_b_bin, what = "mu", newdata = sim_bb$data)
x2_b <- sum((sim_bb$data$y - 20 * mu_b)^2 / (20 * mu_b * (1 - mu_b)))
x2_b / (nrow(sim_bb$data) - 2)
1 + 19 * 0.5 / 1.5
(se_bb / se_bin)^2
```

## A dispersion with covariates {#sec-glm-dispersion}

Every parameter of a distribution can have its own equation, as $\sigma$
had in @sec-lm-spread. The dispersion of a generalized linear model is the
second parameter, and we can let it depend on covariates. A model with a
regression for the dispersion is a joint model for the mean and the
dispersion [@smyth1989], an extension of the generalized linear model that
`glm()` does not fit. In the `GAGurine`
data the variation of the concentration may change with age. We give the
dispersion $\phi$ of the gamma model a linear equation in the age:

```{r}
#| label: fit-gamma-phi
fit_gam_phi <- statmod(GAG ~ Age | phi ~ Age, distrib = gamma1_distrib(),
                       data = GAGurine)
summary(fit_gam_phi)
```

The coefficient of the age in the equation of $\log\phi$ is
`r fmt(coef(fit_gam_phi)$phi[["Age"]], 3)`, so the dispersion grows by a
factor $e^{`r fmt(coef(fit_gam_phi)$phi[["Age"]], 3)`} =
`r fmt(exp(coef(fit_gam_phi)$phi[["Age"]]), 3)`$ for each year of age. We
compute the dispersion at three ages with `predict()`:

```{r}
#| label: gamma-phi-predict
ages <- data.frame(Age = c(1, 9, 17))
predict(fit_gam_phi, what = "phi", newdata = ages)
```

The criterion of both fits integrates the coefficients of the mean out of the
likelihood. For a gamma or a negative binomial response the integral is
approximated by the Laplace method. The two models have the same equation for
the mean, so they integrate the same coefficients, and the two criteria are
marginal likelihoods of the same data. Their difference is a legitimate
likelihood ratio statistic, as for the standard deviation of `cars` in
@sec-lm-spread. The model with a constant dispersion is the one with a zero
coefficient of the age:

```{r}
#| label: gamma-phi-lr
l0_gam <- logLik(fit_gam, type = "marginal")
l1_gam <- logLik(fit_gam_phi, type = "marginal")
lr_gam <- 2 * (as.numeric(l1_gam) - as.numeric(l0_gam))
lr_gam
pchisq(lr_gam, df = 1, lower.tail = FALSE)
c(constant = AIC(fit_gam), age = AIC(fit_gam_phi))
```

The p-value is `r fmt(pchisq(lr_gam, df = 1, lower.tail = FALSE), 4)`, and
the AIC falls by `r fmt(AIC(fit_gam) - AIC(fit_gam_phi), 1)`, so the
dispersion depends on the age.

The same can be done for a count model. In the negative binomial model of
`quine` the dispersion $\theta$ may depend on the ethnic background and on the
learner status:

```{r}
#| label: fit-quine-theta
fit_q_theta <- statmod(Days ~ Eth + Sex + Age + Lrn | theta ~ Eth + Lrn,
                       distrib = negbin2_distrib(), data = quine)
summary(fit_q_theta)
l0_q <- logLik(fit_q_nb2, type = "marginal")
l1_q <- logLik(fit_q_theta, type = "marginal")
lr_q <- 2 * (as.numeric(l1_q) - as.numeric(l0_q))
lr_q
pchisq(lr_q, df = 2, lower.tail = FALSE)
c(constant = AIC(fit_q_nb2), covariates = AIC(fit_q_theta))
```

Here the test does not reject a constant dispersion. The statistic
`r fmt(lr_q, 2)` has a p-value of `r fmt(pchisq(lr_q, df = 2, lower.tail = FALSE), 2)`
on two degrees of freedom, and the AIC of the larger model is larger. The
data contain no evidence that the two covariates change the dispersion, and
we keep the simpler model.

## The link {#sec-glm-links}

The link of a parameter is an object of the `linkfunctions7` package, and
the distribution takes it as an argument. A link has two functions that
invert each other, `linkfun()` from the parameter to the linear predictor and
`linkinv()` from the predictor to the parameter, and it has derivatives of
both. The square root link, for example, maps a positive mean $\mu$ to
$\eta = \sqrt\mu$:

```{r}
#| label: link-object
lk <- sqrt_link()
mu_values <- c(1, 4, 9)
eta_values <- linkfun(lk, theta = mu_values)
eta_values
linkinv(lk, eta = eta_values)
dlinkinv(lk, eta = eta_values)
```

The last call gives the derivative of the mean with respect to the
predictor, $d\mu/d\eta = 2\eta$. The fitting functions use it, and the
derivatives of higher order, to compute the score and the information on the
scale of the linear predictor.

We fit the model for `quine` with the Poisson distribution and the square
root link, and compare it with `glm()`:

```{r}
#| label: fit-sqrt
fit_q_sqrt <- statmod(formula_q, distrib = poisson_distrib(link_mu = sqrt_link()),
                      data = quine)
glm_q_sqrt <- glm(formula_q, family = poisson(link = "sqrt"), data = quine)
coef(fit_q_sqrt)$mu
coef(glm_q_sqrt)
c(statmod = AIC(fit_q_sqrt), glm = AIC(glm_q_sqrt), log_link = AIC(fit_q_pois))
```

The estimates are those of `glm()`. The two links give different models, and
the log link has the smaller AIC.

### A link written by the user

A link that the package does not provide is an S7 object with two methods,
`linkfun()` and `linkinv()`. As an example we take the square of the
logistic function,
$$
p = \left(\frac{1}{1 + e^{-\eta}}\right)^2 ,
\qquad \eta = \mathrm{logit}\left(\sqrt p\right),
$$
which maps the real line onto the interval $(0, 1)$, as the links of a
probability do, and it is not symmetric. We define the class, register the
two methods, and create the object:

```{r}
#| label: user-link
SqLogit <- S7::new_class("SqLogit", parent = link)
S7::method(linkfun, SqLogit) <- function(x, theta) qlogis(sqrt(theta))
S7::method(linkinv, SqLogit) <- function(x, eta) plogis(eta)^2
sq_logit <- SqLogit(link_name = "square-logistic", link_bounds = c(0, 1),
                    link_params = NULL)
```

The derivatives are not written, so the package computes them by finite
differences. The function `link_fallback_orders()` returns the number of
orders of each direction that are analytic, here zero, and `check_link()`
tests the properties of the link:

```{r}
#| label: user-link-check
link_fallback_orders(sq_logit)
invisible(check_link(sq_logit))
dlinkinv(sq_logit, eta = c(-1, 0, 1))
```

The checks of invertibility and monotonicity pass, and the derivative rows
are marked as numerical, because a numerical derivative cannot be checked
against another numerical derivative. The link now works in any
distribution. We fit the additive budworm model with it and compare it with
the logit link:

```{r}
#| label: fit-user-link
fit_bw_sq <- statmod(cbind(numdead, n - numdead) ~ sex + ldose,
                     distrib = binomial_distrib(link_mu = sq_logit),
                     data = budworm)
coef(fit_bw_sq)$mu
c(square_logistic = as.numeric(logLik(fit_bw_sq)),
  logit = as.numeric(logLik(fit_bw_logit)))
```

```{r}
#| label: gate-04
#| include: false
assert_glm_ok(
  fits = list(ships = fit_ships, bw = fit_bw, bw_logit = fit_bw_logit,
              bw_probit = fit_bw_probit, bw_cloglog = fit_bw_cloglog,
              bwt = fit_bwt, bwt_probit = fit_bwt_probit,
              bwt_cloglog = fit_bwt_cloglog, bwt_cauchit = fit_bwt_cauchit,
              gam = fit_gam, gam_ml = fit_gam_ml, ig = fit_ig,
              gam_s = fit_gam_s, gam_phi = fit_gam_phi,
              q_pois = fit_q_pois, q_nb2 = fit_q_nb2, q_nb1 = fit_q_nb1,
              q_pig1 = fit_q_pig1, q_pig2 = fit_q_pig2,
              q_nb2_ml = fit_q_nb2_ml, q_theta = fit_q_theta,
              q_sqrt = fit_q_sqrt, b_bin = fit_b_bin, b_bb = fit_b_bb,
              bw_sq = fit_bw_sq),
  ref = list(ships = glm_ships, bw = glm_bw, bwt = glm_bwt, gam = glm_gam,
             q_nb = glm_q_nb, q_sqrt = glm_q_sqrt),
  data = list(ships = ships, budworm = budworm, birthwt = birthwt,
              gag = GAGurine, quine = quine, sim_bb = sim_bb$data),
  claims = list(ratios = ratios, odds = odds, x2_ships = x2_ships,
                df_ships = df_ships, link_table = link_table,
                bwt_links = bwt_links, phi_table = phi_table,
                quine_fits = quine_fits, observed = observed,
                exp_pois = exp_pois, exp_nb2 = exp_nb2,
                lr_gam = lr_gam, lr_q = lr_q, sq_logit = sq_logit))
```
