# Distributions {#sec-distribs}

```{r}
#| label: setup-05
#| include: false
source("R/_setup.R")
source("R/families-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

In the previous chapters the distribution of the response was gaussian,
Poisson, binomial, gamma or negative binomial. The argument `distrib` of
`statmod()` accepts every distribution of the `distributions7` package, and
the parameters of a distribution are the parameters that the formula models:
`distrib@params` lists them, and each one can have an equation, as `sigma` had
in @sec-lm-spread. This chapter shows the families that the earlier chapters
do not use. They describe responses with heavy tails, with asymmetry, with
values in the unit interval and with many zeros. After them come the wrappers
that change a family (`truncated()`, `folded()` and `transformation()`), the
wrapper that holds a parameter at a known value (`fixed()`), and a
distribution that the user defines by writing its density.

The references of this chapter maximize the likelihood, and the default
criterion of `statmod()` is REML, which estimates the dispersion parameters by
integrating out the coefficients of the means (@sec-lm-linear). We therefore
refit with `outer_criterion = ml()` whenever we compare the estimates with
those of another package.

## Heavy tails {#sec-distribs-heavy}

We return to the `GAGurine` data of @sec-glm-positive. A few children have a
concentration of glycosaminoglycans far above that of the other children of
their age, and the spread of the concentration falls with the age. We give
`sigma` a linear equation in the age, and we compare a gaussian response with
a Student $t$ response. The Student $t$ distribution has a location $\mu$, a
scale $\sigma$ and a number of degrees of freedom $\nu$. A small $\nu$ gives
heavy tails, and as $\nu$ grows the distribution approaches the gaussian
[@lange1989]. The parameter `nu` has the log link, and we leave it without
an equation, so that `statmod()` estimates it as a constant.

```{r}
#| label: fit-heavy
data(GAGurine, package = "MASS")
fit_hg <- statmod(GAG ~ Age | sigma ~ Age, distrib = gaussian1_distrib(),
                  data = GAGurine)
summary(fit_hg)
fit_ht <- statmod(GAG ~ Age | sigma ~ Age, distrib = student_t1_distrib(),
                  data = GAGurine)
summary(fit_ht)
```

The estimate of $\log\nu$ is `r fmt(coef(fit_ht)$nu[[1]], 3)`, so
$\hat\nu = `r fmt(exp(coef(fit_ht)$nu[[1]]))`$. We read the interval of
$\nu$ with `confint()`, which maps the interval of the coefficient back to the
parameter:

```{r}
#| label: heavy-nu
ci_ht <- confint(fit_ht)
ci_ht[ci_ht$parameter == "nu", ]
c(gaussian = AIC(fit_hg), student_t = AIC(fit_ht))
```

The interval of $\log\nu$ runs from `r fmt(ci_ht$lower[ci_ht$parameter == "nu"], 2)`
to `r fmt(ci_ht$upper[ci_ht$parameter == "nu"], 2)`, so $\nu$ lies between
`r fmt(exp(ci_ht$lower[ci_ht$parameter == "nu"]), 1)` and
`r fmt(exp(ci_ht$upper[ci_ht$parameter == "nu"]), 1)`. A gaussian
distribution, with $\nu = \infty$, is far outside this interval. The AIC of the
$t$ model is `r fmt(AIC(fit_hg) - AIC(fit_ht), 0)` units smaller than the AIC
of the gaussian model.

### The same model in `gamlss`

The package `gamlss` fits the same model with its family `TF`
[@rigby2005]. It maximizes the likelihood, so we compare it with the fit of
`statmod()` under `ml()`. This model has no penalty, so the criterion `ml()`
integrates no coefficient out of the likelihood, and the estimates are those of
the joint maximum likelihood. The default stopping rule of `gamlss` is loose
(the relative change of the deviance below 0.001), and we tighten it:

```{r}
#| label: heavy-gamlss
fit_ht_ml <- statmod(GAG ~ Age | sigma ~ Age, distrib = student_t1_distrib(),
                     data = GAGurine, outer_criterion = ml())
gl_ht <- gamlss::gamlss(GAG ~ Age, sigma.formula = ~ Age,
                        family = gamlss.dist::TF, data = GAGurine,
                        control = gamlss::gamlss.control(n.cyc = 200,
                                                         c.crit = 1e-10,
                                                         trace = FALSE))
cmp_ht <- rbind(
  statmod_ml = c(coef(fit_ht_ml)$mu, coef(fit_ht_ml)$sigma, coef(fit_ht_ml)$nu),
  gamlss = c(coef(gl_ht, what = "mu"), coef(gl_ht, what = "sigma"),
             coef(gl_ht, what = "nu")))
colnames(cmp_ht) <- c("mu:(Intercept)", "mu:Age", "sigma:(Intercept)",
                      "sigma:Age", "nu:(Intercept)")
cmp_ht
c(statmod_ml = as.numeric(logLik(fit_ht_ml)), gamlss = as.numeric(logLik(gl_ht)))
```

The two sets of estimates agree to five digits, and the log-likelihoods agree
to the printed digits. With the default stopping rule of `gamlss` the
agreement is about three digits, because that rule stops the iteration before
the maximum. The fits with `outer_criterion = NULL` and with
`ml(marginal = "none")` give the same estimates to about five digits, since
there is nothing to integrate out.

@fig-heavy shows the two fitted models. The band of each model runs from the
5% to the 95% quantile of the fitted distribution at each age, so it should
leave out 10% of the children. We count the children outside each band:

```{r}
#| label: heavy-coverage
mu_d <- predict(fit_ht, what = "mu", newdata = GAGurine)
sg_d <- predict(fit_ht, what = "sigma", newdata = GAGurine)
nu_d <- exp(coef(fit_ht)$nu[[1]])
out_t <- mean(GAGurine$GAG < mu_d + qt(0.05, nu_d) * sg_d |
                GAGurine$GAG > mu_d + qt(0.95, nu_d) * sg_d)
mu_dg <- predict(fit_hg, what = "mu", newdata = GAGurine)
sg_dg <- predict(fit_hg, what = "sigma", newdata = GAGurine)
out_g <- mean(GAGurine$GAG < mu_dg + qnorm(0.05) * sg_dg |
                GAGurine$GAG > mu_dg + qnorm(0.95) * sg_dg)
c(student_t = out_t, gaussian = out_g)
```

The band of the $t$ model leaves out `r fmt(100 * out_t, 1)`% of the children
and the band of the gaussian model `r fmt(100 * out_g, 1)`%. The $t$ band is
close to the 10% that the model predicts, and the gaussian band is wider than
the data need. In the figure the upper limit of the gaussian band is above
that of the $t$ band for the youngest children, and the lower limit of the $t$
band is below the gaussian one. The scale $\sigma$ of the $t$ model is not a
standard deviation, and the children far above the others are covered by the
tails of the distribution. The AIC of the gaussian model is much larger, so
the similar bands do not make the two models equal.

```{r}
#| label: fig-heavy
#| fig-cap: "GAG concentration against age, with the fitted location and the band from the 5% to the 95% quantile of the gaussian model (dashed) and of the Student $t$ model (solid)."
#| echo: false
#| fig-height: 4
ga <- data.frame(Age = seq(from = 0, to = 17, length.out = 100))
mu_t <- predict(fit_ht, what = "mu", newdata = ga)
sg_t <- predict(fit_ht, what = "sigma", newdata = ga)
mu_g <- predict(fit_hg, what = "mu", newdata = ga)
sg_g <- predict(fit_hg, what = "sigma", newdata = ga)
nu_t <- exp(coef(fit_ht)$nu[[1]])
heavy_band <- data.frame(Age = ga$Age, lower = mu_t + qt(0.05, nu_t) * sg_t,
                         upper = mu_t + qt(0.95, nu_t) * sg_t)
heavy_lines <- data.frame(
  Age = rep(ga$Age, times = 3),
  GAG = c(mu_t, mu_g + qnorm(0.95) * sg_g, mu_g + qnorm(0.05) * sg_g),
  curve = factor(rep(c(1, 2, 2), each = nrow(ga)),
                 labels = c("t: location", "gaussian: 5% and 95%")),
  line = rep(1:3, each = nrow(ga)))
ggplot(GAGurine, aes(x = Age, y = GAG)) +
  geom_ribbon(data = heavy_band,
              aes(x = Age, ymin = lower, ymax = upper, fill = "t: 5-95% band"),
              inherit.aes = FALSE, alpha = 0.2) +
  geom_point(colour = book_grey, size = 1) +
  geom_line(data = heavy_lines,
            aes(colour = curve, linetype = curve, group = line),
            linewidth = 0.8) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_fill_manual(values = chalk) +
  guides(colour = guide_legend(order = 1), linetype = guide_legend(order = 1),
         fill = guide_legend(order = 2)) +
  coord_cartesian(ylim = c(0, 45)) +
  labs(x = "age (years)", y = "GAG")
```

### The degrees of freedom with an equation

The parameter `nu` can have an equation like `mu` and `sigma`. The tails may
be heavier for some ages than for others, so we let $\log\nu$ depend on the
age:

```{r}
#| label: fit-heavy-nu
crit_e <- reml(hessian = "expected")
fit_ht0 <- statmod(GAG ~ Age | sigma ~ Age, distrib = student_t1_distrib(),
                   data = GAGurine, outer_criterion = crit_e)
summary(fit_ht0)
fit_ht1 <- statmod(GAG ~ Age | sigma ~ Age | nu ~ Age,
                   distrib = student_t1_distrib(), data = GAGurine,
                   outer_criterion = crit_e)
summary(fit_ht1)
```

The criterion of these two fits is `reml(hessian = "expected")`. The default
REML criterion evaluates the curvature of the log-likelihood with the observed
information, and for a $t$ distribution with small $\nu$ the observed
information of the mean is not positive at the observations far from the
location. The criterion then has no value at some of the points that the
search visits, and the model with an equation for `nu` ends with the state
`not converged` of @sec-model-certificate. The expected information of the
$t$ distribution is positive at every point, and the two fits above reach the
state `r statmod_certificate(fit_ht1)$state`.

The coefficient of the age in the equation of $\log\nu$ is
`r fmt(coef(fit_ht1)$nu[["Age"]], 3)`: the degrees of freedom fall with the
age, and the tails are heavier for older children. We compare the two models
with the likelihood ratio of the marginal log-likelihoods, as in
@sec-glm-dispersion. Both fits integrate the same coefficients of the mean, and
they use the same criterion:

```{r}
#| label: heavy-nu-lr
l0_ht <- logLik(fit_ht0, type = "marginal")
l1_ht <- logLik(fit_ht1, type = "marginal")
lr_ht <- 2 * (as.numeric(l1_ht) - as.numeric(l0_ht))
lr_ht
pchisq(lr_ht, df = 1, lower.tail = FALSE)
ages <- data.frame(Age = c(1, 9, 17))
predict(fit_ht1, what = "nu", newdata = ages)
```

The statistic is `r fmt(lr_ht, 1)`, with a p-value of
`r fmt(pchisq(lr_ht, df = 1, lower.tail = FALSE), 4)`. The fitted $\nu$ is
`r fmt(predict(fit_ht1, what = "nu", newdata = ages)[1], 1)` at one year of
age and `r fmt(predict(fit_ht1, what = "nu", newdata = ages)[3], 2)` at 17 years. The
value of $\nu$ for the oldest children is below one, which is a distribution
with heavier tails than the Cauchy distribution ($\nu = 1$).

### A degrees of freedom parameter at the boundary

When the data have gaussian tails, the likelihood increases with $\nu$ up to
infinity, and no finite estimate exists. We draw 400 observations from a
gaussian distribution with `rstatmod()` and we fit a Student $t$ model:

```{r}
#| label: fit-heavy-boundary
set.seed(1)
xb <- data.frame(x = runif(n = 400))
sim_b <- rstatmod(y ~ x | sigma ~ 1, distrib = gaussian1_distrib(), data = xb,
                  par = list(mu = c(1, 2), sigma = log(0.5)))
fit_bt <- statmod(y ~ x | sigma ~ 1, distrib = student_t1_distrib(),
                  data = sim_b$data)
summary(fit_bt)
fit_bg <- statmod(y ~ x | sigma ~ 1, distrib = gaussian1_distrib(),
                  data = sim_b$data)
summary(fit_bg)
exp(coef(fit_bt)$nu)
cert_bt <- statmod_certificate(fit_bt)
cert_bt$state
cert_bt$boundary
c(student_t = AIC(fit_bt), gaussian = AIC(fit_bg))
```

The estimate of $\nu$ is `r format(exp(coef(fit_bt)$nu[[1]]), scientific = FALSE, big.mark = ",")`,
the largest value that the log link can produce, and the certificate gives the
state `r cert_bt$state` and names the coefficient `r cert_bt$boundary`. The
fit is at its maximum, and the maximum is a limit: the $t$ model has become the
gaussian model. `summary()` leaves the standard error of this coefficient
empty, since it has no meaning at the boundary. The AIC of the gaussian model
is smaller by `r fmt(AIC(fit_bt) - AIC(fit_bg), 1)` units, which is the
penalty of the extra parameter.

## Asymmetry {#sec-distribs-skew}

The data frame `Cars93` from the `MASS` package has the list price in
thousands of dollars of 93 cars and their engine power in horse power. The
price has a long right tail. We compute the skewness of the residuals of a
linear regression of the price on the power, as the third central moment
divided by the cube of the standard deviation:

```{r}
#| label: skew-data
data(Cars93, package = "MASS")
cars93 <- Cars93[, c("Price", "Horsepower")]
r_lm <- residuals(lm(Price ~ Horsepower, data = cars93))
skew_lm <- mean((r_lm - mean(r_lm))^3) / mean((r_lm - mean(r_lm))^2)^1.5
skew_lm
```

The skew normal distribution of @azzalini1985 has a location, a scale and a
shape $\alpha$ that controls the asymmetry. Its skewness cannot be larger than
the limit that it reaches as $\alpha \to \infty$:
$$
\gamma_{\max} = \frac{4 - \pi}{2}\,
\frac{(2/\pi)^{3/2}}{(1 - 2/\pi)^{3/2}} .
$$

```{r}
#| label: skew-limit
gamma_max <- (4 - pi) / 2 * (2 / pi)^1.5 / (1 - 2 / pi)^1.5
gamma_max
```

The skewness of the residuals is larger than this limit. This skewness is a
moment of the sample, and the likelihood of the skew normal
regression still has its maximum at a finite shape. The package has two
parametrizations of the distribution. `skewnormal1_distrib()` has the
direct parameters of @azzalini1985: the location $\xi$ (`mu`), the scale
$\omega$ (`sigma`) and the shape $\alpha$ (`alpha`), with the identity link
for the shape. `skewnormal2_distrib()` has the centred parameters: the mean
(`mu`), the standard deviation (`sigma`) and the skewness `gamma1`, with a
link that keeps `gamma1` between $-\gamma_{\max}$ and $\gamma_{\max}$. The two
parametrizations describe the same family of distributions, so we expect the
same maximum from both:

```{r}
#| label: fit-skew-sn
fit_sn1 <- statmod(Price ~ Horsepower, distrib = skewnormal1_distrib(),
                   data = cars93)
summary(fit_sn1)
fit_ss <- statmod(Price ~ Horsepower, distrib = skewnormal2_distrib(),
                  data = cars93)
summary(fit_ss)
rbind(direct = c(slope = coef(fit_sn1)$mu[["Horsepower"]],
                 logLik = as.numeric(logLik(fit_sn1)),
                 marginal = as.numeric(logLik(fit_sn1, type = "marginal"))),
      centred = c(coef(fit_ss)$mu[["Horsepower"]], as.numeric(logLik(fit_ss)),
                  as.numeric(logLik(fit_ss, type = "marginal"))))
```

The slopes and the log-likelihoods agree. The intercepts differ, because the
intercept of the first fit is the location $\xi$ for a car with no power, and
the intercept of the second fit is the mean, which is $\xi$ plus
$\omega\sqrt{2/\pi}\,\alpha/\sqrt{1 + \alpha^2}$. The estimate of the shape is
$\hat\alpha = `r fmt(coef(fit_sn1)$alpha[[1]], 2)`$, and the estimate of the
skewness in the second fit is
$\hat\gamma_1 = `r fmt(predict(fit_ss, what = "gamma1", newdata = cars93[1, ])[[1]], 3)`$,
close to the limit $\gamma_{\max}$.

The direct parametrization is difficult to use near this limit. The skewness
of the price without the covariate is
`r fmt(mean(((cars93$Price - mean(cars93$Price)) / sd(cars93$Price))^3), 2)`,
which is also above $\gamma_{\max}$, and the model without covariates then has
its maximum at $\alpha \to \infty$, with $\xi$ at the smallest price. At that
limit the distribution is a half-normal distribution, and the curvature of
the log-likelihood in $\xi$ grows as $\alpha^2$. A fit that started there would
take steps of order $1/\alpha^2$. For this reason `statmod()` starts the
direct parametrization from the moment estimate, with the skewness of the
sample held at 0.9, when the skewness of the response is beyond that value.
The centred parametrization does not have this difficulty, since the mean and
the standard deviation stay finite as the skewness approaches $\gamma_{\max}$
[@azzalini1985].

The skew $t$ distribution of @azzalini2003 has the degrees of freedom $\nu$ as
a fourth parameter, and its skewness is not bounded by this limit. We fit the
gaussian model and the skew $t$ model, and we compare them with the skew
normal model:

```{r}
#| label: fit-skew
fit_sg <- statmod(Price ~ Horsepower, distrib = gaussian1_distrib(),
                  data = cars93)
summary(fit_sg)
fit_st <- statmod(Price ~ Horsepower, distrib = skewt_distrib(), data = cars93)
summary(fit_st)
skew_table <- data.frame(
  distribution = c("gaussian", "skew normal", "skew t"),
  parameters = c(3, 4, 5),
  logLik = c(as.numeric(logLik(fit_sg)), as.numeric(logLik(fit_ss)),
             as.numeric(logLik(fit_st))),
  AIC = c(AIC(fit_sg), AIC(fit_ss), AIC(fit_st)))
skew_table
```

The skew normal model has a much smaller AIC than the gaussian model, and the
skew $t$ model has the smallest. The estimate of the degrees of freedom of the
skew $t$ model is $\hat\nu = `r fmt(exp(coef(fit_st)$nu[[1]]))`$. The tails of
the skew $t$ distribution decay like those of a Student $t$ distribution with
$\nu$ degrees of freedom, so the same limits hold: the skew $t$ distribution
has a variance only when $\nu > 2$ and a skewness only when $\nu > 3$. The
generics `variance()` and `skewness()` of `distributions7` return these
moments at given parameters, and `NaN` where the moment does not exist. We
evaluate them at the fitted parameters for a car with 150 horse power:

```{r}
#| label: skewt-moments
th_150 <- list(mu = predict(fit_st, what = "mu", newdata = data.frame(Horsepower = 150)),
               sigma = predict(fit_st, what = "sigma", newdata = data.frame(Horsepower = 150)),
               alpha = predict(fit_st, what = "alpha", newdata = data.frame(Horsepower = 150)),
               nu = predict(fit_st, what = "nu", newdata = data.frame(Horsepower = 150)))
c(nu = th_150$nu, variance = variance(skewt_distrib(), theta = th_150),
  skewness = skewness(skewt_distrib(), theta = th_150))
th_150$nu <- 6
c(nu = th_150$nu, variance = variance(skewt_distrib(), theta = th_150),
  skewness = skewness(skewt_distrib(), theta = th_150))
```

At the fitted $\nu$ the variance is finite and the skewness does not exist. With
$\nu = 6$ and the other parameters unchanged, both exist.

### A check of the log-likelihood

The skew normal density is $f(y) = (2/\omega)\,\phi(z)\,\Phi(\alpha z)$, with
$z = (y - \xi)/\omega$. We write the log-likelihood in R, and we maximize it
with `optim()`. This is a computation that shares no code with
`distributions7`:

```{r}
#| label: skew-check
nll_sn <- function(p) {
  z <- (cars93$Price - p[1] - p[2] * cars93$Horsepower) / exp(p[3])
  -sum(log(2) - p[3] + dnorm(z, log = TRUE) + pnorm(p[4] * z, log.p = TRUE))
}
opt_sn <- optim(par = c(-2, 0.1, 0.5, 1), fn = nll_sn, method = "BFGS",
                control = list(reltol = 1e-12, maxit = 1000))
fit_ss_ml <- statmod(Price ~ Horsepower, distrib = skewnormal2_distrib(),
                     data = cars93, outer_criterion = NULL)
c(by_hand = -opt_sn$value, statmod_ml = as.numeric(logLik(fit_ss_ml)))
```

With `outer_criterion = NULL` the function `statmod()` has no criterion for
the dispersion parameters, and it maximizes the log-likelihood jointly over
all the coefficients. The maximized log-likelihoods agree. The parametrization of the
hand-written function is the direct one, $(\xi, \omega, \alpha)$, and the
model of `statmod()` uses $(\mu, \sigma, \gamma_1)$, so the parameters differ
while the maximum is the same.

@fig-skew shows the conditional median and the band from the 10% to the 90%
quantile of the gaussian and of the skew $t$ model. The gaussian band is
symmetric around the median and it extends below zero for the cheap cars. The
skew $t$ band is short below the median and long above it.

```{r}
#| label: fig-skew
#| fig-cap: "List price against engine power. The lines are the conditional median and the 10% and 90% quantiles of the gaussian model (dashed) and of the skew $t$ model (solid)."
#| echo: false
#| fig-height: 4
hp <- data.frame(Horsepower = seq(from = 55, to = 300, length.out = 40))
pars_t <- list(mu = predict(fit_st, what = "mu", newdata = hp),
               sigma = predict(fit_st, what = "sigma", newdata = hp),
               alpha = predict(fit_st, what = "alpha", newdata = hp),
               nu = predict(fit_st, what = "nu", newdata = hp))
qt_lo <- numeric(40); qt_md <- numeric(40); qt_hi <- numeric(40)
for (i in 1:40) {
  th_i <- lapply(pars_t, function(v) v[i])
  qt_lo[i] <- distrib_quantile(skewt_distrib(), 0.1, th_i)
  qt_md[i] <- distrib_quantile(skewt_distrib(), 0.5, th_i)
  qt_hi[i] <- distrib_quantile(skewt_distrib(), 0.9, th_i)
}
mu_sg <- predict(fit_sg, what = "mu", newdata = hp)
sg_sg <- predict(fit_sg, what = "sigma", newdata = hp)
skew_lines <- data.frame(
  Horsepower = rep(hp$Horsepower, times = 6),
  Price = c(qt_md, qt_lo, qt_hi, mu_sg, mu_sg + qnorm(0.1) * sg_sg,
            mu_sg + qnorm(0.9) * sg_sg),
  curve = factor(rep(c(1, 2, 2, 3, 3, 3), each = nrow(hp)),
                 labels = c("skew t: median", "skew t: 10% and 90%",
                            "gaussian: mean, 10% and 90%")),
  line = rep(1:6, each = nrow(hp)))
ggplot(cars93, aes(x = Horsepower, y = Price)) +
  geom_point(colour = book_grey, size = 1.1) +
  geom_line(data = skew_lines,
            aes(colour = curve, linetype = curve, group = line),
            linewidth = 0.8) +
  scale_colour_manual(values = c(chalk, chalk, rust)) +
  scale_linetype_manual(values = c("solid", "dotted", "dashed")) +
  coord_cartesian(ylim = c(min(mu_sg + qnorm(0.1) * sg_sg), 65)) +
  labs(x = "horse power", y = "price (thousand dollars)")
```

## Proportions {#sec-distribs-beta}

A response in the open interval $(0, 1)$, such as a rate or a share, has a
variance that is smaller near the ends of the interval. The beta distribution
has a mean $\mu$ and a precision $\phi$, and its variance is
$\mu(1 - \mu)/(1 + \phi)$ [@ferrari2004]. The parameter `mu` has the logit
link and `phi` the log link. The data frame `swiss` of the `datasets` package
has, for 47 French-speaking provinces of Switzerland in 1888, the share of
males who worked in agriculture (`Agriculture`, in percent) and the share of
army draftees with education beyond primary school (`Education`, in percent).
We model the proportion in agriculture as a function of the education:

```{r}
#| label: fit-beta
swiss_p <- swiss[, c("Agriculture", "Education")]
swiss_p$agr <- swiss_p$Agriculture / 100
fit_b0 <- statmod(agr ~ Education, distrib = beta1_distrib(), data = swiss_p)
summary(fit_b0)
```

The coefficient of the education is `r fmt(coef(fit_b0)$mu[["Education"]], 3)`
on the logit scale, so the odds of working in agriculture fall by
`r fmt(100 * (1 - exp(coef(fit_b0)$mu[["Education"]])), 1)`% for each
percent of draftees with further education. The precision is
$\hat\phi = `r fmt(exp(coef(fit_b0)$phi[[1]]), 2)`$.

The packages `gamlss` and `glmmTMB` fit the same model. The `gamlss` family
`BE` parametrizes the dispersion by $\sigma = 1/\sqrt{1 + \phi}$ on the logit
scale, and `glmmTMB` uses $\phi$ itself. Both maximize the likelihood:

```{r}
#| label: beta-refs
fit_b0_ml <- statmod(agr ~ Education, distrib = beta1_distrib(), data = swiss_p,
                     outer_criterion = ml())
gl_be <- gamlss::gamlss(agr ~ Education, family = gamlss.dist::BE,
                        data = swiss_p, trace = FALSE)
tmb_be <- glmmTMB::glmmTMB(agr ~ Education, family = glmmTMB::beta_family(),
                           data = swiss_p)
sigma_be <- plogis(coef(gl_be, what = "sigma"))
phi_table <- data.frame(
  phi = c(exp(coef(fit_b0_ml)$phi[[1]]), (1 - sigma_be^2) / sigma_be^2,
          glmmTMB::sigma(tmb_be), exp(coef(fit_b0)$phi[[1]])),
  logLik = c(as.numeric(logLik(fit_b0_ml)), as.numeric(logLik(gl_be)),
             as.numeric(logLik(tmb_be)), as.numeric(logLik(fit_b0))),
  row.names = c("statmod, ml()", "gamlss BE", "glmmTMB", "statmod, default"))
phi_table
rbind(statmod_ml = coef(fit_b0_ml)$mu, gamlss = coef(gl_be, what = "mu"),
      glmmTMB = glmmTMB::fixef(tmb_be)$cond)
```

The three packages give the same coefficients, the same precision and the same
log-likelihood. The default criterion of `statmod()` gives a precision that is
about two per cent smaller, because it integrates the two coefficients of the
mean out of the likelihood, as it did for the gamma distribution in
@sec-glm-positive.

### A precision with an equation

We give `phi` an equation in the education, and we compare the two models with
the likelihood ratio of the marginal log-likelihoods:

```{r}
#| label: fit-beta-phi
fit_b1 <- statmod(agr ~ Education | phi ~ Education, distrib = beta1_distrib(),
                  data = swiss_p)
summary(fit_b1)
l0_b <- logLik(fit_b0, type = "marginal")
l1_b <- logLik(fit_b1, type = "marginal")
lr_b <- 2 * (as.numeric(l1_b) - as.numeric(l0_b))
lr_b
pchisq(lr_b, df = 1, lower.tail = FALSE)
c(constant = AIC(fit_b0), education = AIC(fit_b1))
```

The statistic is `r fmt(lr_b, 2)` and the p-value
`r fmt(pchisq(lr_b, df = 1, lower.tail = FALSE), 2)`, so the data do not show a
change of the precision with the education, and the simpler model stays.

## Excess zeros {#sec-distribs-zeros}

A count with more zeros than its distribution expects needs a model for the
zeros. The data frame `solder` of the `rpart` package comes from an experiment
on the soldering of components on printed circuit boards [@chambers1992]. The
response `skips` is the number of visible solder skips on a board, and the
covariates are five factors: the size of the opening in the solder mask, the
thickness of the solder, the type of mask, the type of pad and the panel.

```{r}
#| label: solder-data
data(solder, package = "rpart")
fs <- skips ~ Opening + Solder + Mask + PadType + Panel
c(boards = nrow(solder), zeros = sum(solder$skips == 0))
```

The wrapper `zero_inflated()` adds to a count distribution a probability
$\zeta$ of a zero that does not come from the distribution
[@lambert1992]:
$$
P(y = 0) = \zeta + (1 - \zeta)\, f(0), \qquad
P(y = k) = (1 - \zeta)\, f(k), \quad k \ge 1 .
$$
The wrapper `zero_adjusted()` builds a hurdle model [@mullahy1986]. It gives
the probability $\pi$ of a zero, and it describes the positive counts by the
distribution truncated at zero:
$$
P(y = 0) = \pi, \qquad
P(y = k) = (1 - \pi)\,\frac{f(k)}{1 - f(0)}, \quad k \ge 1 .
$$
Each wrapper adds one parameter, `zi` for the first and `za` for the second,
with the logit link, and the parameter can have an equation like any other. We
fit the Poisson and the negative binomial models of @sec-glm-overdispersion,
the zero-inflated negative binomial model with a constant $\zeta$, and the
hurdle model in which `za` depends on the same five factors as the mean:

```{r}
#| label: fit-zeros
fit_zp <- statmod(fs, distrib = poisson_distrib(), data = solder)
summary(fit_zp)
fit_zn <- statmod(fs, distrib = negbin2_distrib(), data = solder)
summary(fit_zn)
fit_zi <- statmod(fs, distrib = zero_inflated(negbin2_distrib()), data = solder)
summary(fit_zi)
fit_za <- statmod(skips ~ Opening + Solder + Mask + PadType + Panel |
                    za ~ Opening + Solder + Mask + PadType + Panel,
                  distrib = zero_adjusted(negbin2_distrib()), data = solder)
c(Poisson = AIC(fit_zp), negbin2 = AIC(fit_zn), zero_inflated = AIC(fit_zi),
  hurdle = AIC(fit_za))
```

The zero-inflated model has almost the same AIC as the negative binomial
model. A constant $\zeta$ adds zeros in the same proportion to every
combination of the factors, and the estimate is
$\hat\zeta = `r fmt(plogis(coef(fit_zi)$zi[[1]]), 3)`$. The hurdle model has a
much smaller AIC. In it the probability of a zero depends on the factors, and
the summary of the model shows how:

```{r}
#| label: zeros-za
summary(fit_za)
```

All the coefficients of `za` for the opening and the solder are negative, so
the probability of a board with no skips falls for the larger openings and the
thinner solder, which are the conditions that produce more skips. The
zero-inflated model can also have an equation for `zi`. We give it the
opening and the solder:

```{r}
#| label: fit-zi-x
fit_zi2 <- statmod(skips ~ Opening + Solder + Mask + PadType + Panel |
                     zi ~ Opening + Solder,
                   distrib = zero_inflated(negbin2_distrib()), data = solder)
summary(fit_zi2)
c(constant = AIC(fit_zi), opening_solder = AIC(fit_zi2))
```

### The same models in `glmmTMB`

The package `glmmTMB` fits the zero-inflated and the hurdle negative binomial
models, the second with its family `truncated_nbinom2`. We compare the log-likelihoods
of the two packages, with `statmod()` under `ml()`:

```{r}
#| label: zeros-refs
fit_zi_ml <- statmod(fs, distrib = zero_inflated(negbin2_distrib()),
                     data = solder, outer_criterion = ml())
fit_za_ml <- statmod(skips ~ Opening + Solder + Mask + PadType + Panel |
                       za ~ Opening + Solder + Mask + PadType + Panel,
                     distrib = zero_adjusted(negbin2_distrib()), data = solder,
                     outer_criterion = ml())
tmb_zi <- glmmTMB::glmmTMB(fs, ziformula = ~ 1, family = glmmTMB::nbinom2,
                           data = solder)
tmb_za <- glmmTMB::glmmTMB(fs, ziformula = ~ Opening + Solder + Mask + PadType + Panel,
                           family = glmmTMB::truncated_nbinom2, data = solder)
data.frame(statmod = c(as.numeric(logLik(fit_zi_ml)), as.numeric(logLik(fit_za_ml))),
           glmmTMB = c(as.numeric(logLik(tmb_zi)), as.numeric(logLik(tmb_za))),
           row.names = c("zero-inflated", "hurdle"))
```

The log-likelihoods are the same. The parameter of the zero-inflated
model of `glmmTMB` is `zi` on the logit scale, and it is
`r fmt(glmmTMB::fixef(tmb_zi)$zi[[1]], 3)`, against
`r fmt(coef(fit_zi_ml)$zi[[1]], 3)` for `statmod()`.

## Truncation, folding and transformation {#sec-distribs-wrappers}

Three more wrappers change the support of a distribution. They take a
distribution and return another distribution that has the same parameters and
can be used in any formula.

### Truncation

A response is truncated when only the values above a limit are observed. The
density of the truncated distribution is the density of the original one
divided by the probability of the observed region:
$$
f_T(y) = \frac{f(y)}{1 - F(a)}, \qquad y \ge a .
$$
`truncated(distrib, lower =, upper =)` builds this distribution. We draw 300
observations from a gaussian distribution truncated below zero, with a mean
that increases with $x$, and we fit the truncated and the plain gaussian
models:

```{r}
#| label: fit-truncated
set.seed(5)
xt <- data.frame(x = runif(n = 300))
tg <- truncated(gaussian1_distrib(), lower = 0)
sim_t <- rstatmod(y ~ x | sigma ~ 1, distrib = tg, data = xt,
                  par = list(mu = c(0.5, 1.5), sigma = 0))
fit_tr <- statmod(y ~ x | sigma ~ 1, distrib = tg, data = sim_t$data,
                  outer_criterion = ml())
summary(fit_tr)
fit_tg <- statmod(y ~ x | sigma ~ 1, distrib = gaussian1_distrib(),
                  data = sim_t$data, outer_criterion = ml())
summary(fit_tg)
rbind(truth = c(0.5, 1.5, 0), truncated = c(coef(fit_tr)$mu, coef(fit_tr)$sigma),
      gaussian = c(coef(fit_tg)$mu, coef(fit_tg)$sigma))
```

The plain gaussian model ignores the truncation, and its intercept and slope
are biased. The truncated model recovers the true values. @fig-truncated shows
why: near $x = 0$ the observations cannot fall below zero, so the plain
gaussian line is pushed up at the left end and flattened. The lines of the
truncated model are the mean of the gaussian distribution before truncation.

```{r}
#| label: fig-truncated
#| fig-cap: "Simulated data from a gaussian distribution truncated below zero, with the true mean before truncation and the fitted lines of the truncated and of the plain gaussian model."
#| echo: false
#| fig-height: 3.8
xg <- seq(from = 0, to = 1, length.out = 100)
trunc_labels <- c("truncation at zero", "true mean before truncation",
                  "truncated model", "plain gaussian model")
trunc_lines <- data.frame(
  x = rep(xg, times = 3),
  y = c(0.5 + 1.5 * xg, coef(fit_tr)$mu[[1]] + coef(fit_tr)$mu[[2]] * xg,
        coef(fit_tg)$mu[[1]] + coef(fit_tg)$mu[[2]] * xg),
  curve = factor(rep(trunc_labels[2:4], each = length(xg)),
                 levels = trunc_labels))
trunc_zero <- data.frame(curve = factor(trunc_labels[1], levels = trunc_labels))
ggplot(sim_t$data, aes(x = x, y = y)) +
  geom_point(colour = book_grey, size = 1) +
  geom_hline(data = trunc_zero,
             aes(yintercept = 0, colour = curve, linetype = curve,
                 linewidth = curve)) +
  geom_line(data = trunc_lines,
            aes(colour = curve, linetype = curve, linewidth = curve)) +
  scale_colour_manual(values = c("black", "black", chalk, rust)) +
  scale_linetype_manual(values = c("dotted", "dotted", "solid", "dashed")) +
  scale_linewidth_manual(values = c(0.4, 0.8, 0.8, 0.8)) +
  coord_cartesian(ylim = c(-0.3, max(sim_t$data$y))) +
  labs(x = "x", y = "y")
```

The log-likelihood of the truncated model is the sum of the logarithms of the
density above, and we write it in R:

```{r}
#| label: truncated-check
nll_tr <- function(p) {
  mu <- p[1] + p[2] * sim_t$data$x
  s <- exp(p[3])
  -sum(dnorm(sim_t$data$y, mean = mu, sd = s, log = TRUE) -
         pnorm(0, mean = mu, sd = s, lower.tail = FALSE, log.p = TRUE))
}
opt_tr <- optim(par = c(0.5, 1, 0), fn = nll_tr, method = "BFGS",
                control = list(reltol = 1e-12))
c(by_hand = -opt_tr$value, statmod = as.numeric(logLik(fit_tr)))
```

### Folding

The folded distribution is the distribution of $|X|$. It describes a
measurement whose sign is lost, such as a distance or a deviation. The
folded normal density is $f(y) = \phi_{\mu,\sigma}(y) + \phi_{-\mu,\sigma}(y)$
for $y \ge 0$ [@leone1961]. The density does not change when $\mu$ is replaced
by $-\mu$, so the intercept and the slope are identified only up to a common
change of sign: the pairs $(-0.3, 1.5)$ and $(0.3, -1.5)$ describe the same
distribution, and `statmod()` returns one of the two. What the data determine
is the curve $|\mu(x)|$. We draw data from $\mu(x) = -0.3 + 1.5x$, which
crosses zero at $x = 0.2$, and we fit a model with the same linear equation for
`mu`. The model has no break-point term. The curve $|\mu(x)|$ has a corner at
the point where $\mu(x)$ crosses zero, because the absolute value changes the
sign of the line there. @fig-folded shows the data with the curve $|\mu(x)|$
of the truth and of the fit:

```{r}
#| label: fit-folded
set.seed(6)
sim_f <- rstatmod(y ~ x | sigma ~ 1, distrib = folded(gaussian1_distrib()),
                  data = xt, par = list(mu = c(-0.3, 1.5), sigma = log(0.6)))
fit_fo <- statmod(y ~ x | sigma ~ 1, distrib = folded(gaussian1_distrib()),
                  data = sim_f$data, outer_criterion = ml())
summary(fit_fo)
nll_fo <- function(p) {
  mu <- p[1] + p[2] * sim_f$data$x
  s <- exp(p[3])
  -sum(log(dnorm(sim_f$data$y, mean = mu, sd = s) +
             dnorm(sim_f$data$y, mean = -mu, sd = s)))
}
opt_fo <- optim(par = c(0, 1, 0), fn = nll_fo, method = "BFGS",
                control = list(reltol = 1e-12))
rbind(truth = c(-0.3, 1.5, log(0.6)), statmod = c(coef(fit_fo)$mu, coef(fit_fo)$sigma),
      by_hand = opt_fo$par)
c(by_hand = -opt_fo$value, statmod = as.numeric(logLik(fit_fo)))
```

```{r}
#| label: fig-folded
#| fig-cap: "Simulated data from a folded gaussian distribution, with the curve of the absolute value of the true mean and of the fitted mean."
#| echo: false
#| fig-height: 3.8
fold_lines <- data.frame(
  x = rep(xg, times = 2),
  y = c(abs(-0.3 + 1.5 * xg),
        abs(coef(fit_fo)$mu[[1]] + coef(fit_fo)$mu[[2]] * xg)),
  curve = factor(rep(1:2, each = length(xg)),
                 labels = c("|true mean|", "|fitted mean|")))
ggplot(sim_f$data, aes(x = x, y = y)) +
  geom_point(colour = book_grey, size = 1) +
  geom_line(data = fold_lines, aes(colour = curve, linetype = curve),
            linewidth = 0.8) +
  scale_colour_manual(values = c("black", chalk)) +
  scale_linetype_manual(values = c("dotted", "solid")) +
  labs(x = "x", y = "y")
```

### Transformation

`transformation(distrib, transformer)` is the distribution of $T(X)$, where $X$
has the distribution `distrib` and $T$ is a transformer. The package has
transformers for the exponential, the logarithm, the power, the logit and its
inverse, among others. The exponential of a gaussian variable is a lognormal
variable, so `transformation(gaussian1_distrib(), exp_transform())` is the
lognormal distribution of `lognormal1_distrib()`. We fit both to the `GAGurine`
data and compare the results:

```{r}
#| label: fit-transformation
fit_tx <- statmod(GAG ~ Age, distrib = transformation(gaussian1_distrib(),
                                                      exp_transform()),
                  data = GAGurine, outer_criterion = ml())
summary(fit_tx)
fit_ln <- statmod(GAG ~ Age, distrib = lognormal1_distrib(), data = GAGurine,
                  outer_criterion = ml())
summary(fit_ln)
rbind(transformation = coef(fit_tx)$mu, lognormal1 = coef(fit_ln)$mu)
c(transformation = as.numeric(logLik(fit_tx)),
  lognormal1 = as.numeric(logLik(fit_ln)))
```

The coefficients and the log-likelihoods are identical, and the two fits differ
only in the parametrization of the second parameter: `sigma` for the
transformed gaussian and `sigma2`, the variance, for `lognormal1_distrib()`.

The inverse of the logit is a transformation of the real line onto the unit
interval, so a gaussian variable transformed by `expit_transform()` is a
logit-normal variable [@aitchison1980], an alternative to the beta
distribution for a proportion. We fit it to the `swiss` data of
@sec-distribs-beta:

```{r}
#| label: fit-logitnormal
fit_ln_b <- statmod(agr ~ Education, distrib = transformation(gaussian1_distrib(),
                                                              expit_transform()),
                    data = swiss_p)
summary(fit_ln_b)
c(beta = AIC(fit_b0), logit_normal = AIC(fit_ln_b))
```

The logit-normal model has the smaller AIC by
`r fmt(AIC(fit_b0) - AIC(fit_ln_b), 1)` units. The two distributions have the
same number of parameters, and their log-likelihoods are the logarithms of the
probability that each model gives to the same observed proportions. The
comparison holds because the response of both models is $y$ itself.

## A known parameter {#sec-distribs-fixed}

A parameter may be known from outside the data. `fixed(distrib, ...)` holds
the named parameters of `distrib` at given values, and the distribution that it
returns has the remaining parameters only. In the `quine` data of
@sec-glm-overdispersion the negative binomial dispersion $\theta$ is estimated.
We fix it at 1.5 and compare the fit with `glm()` of the family
`MASS::negative.binomial(theta = 1.5)`, which takes $\theta$ as known:

```{r}
#| label: fit-fixed
data(quine, package = "MASS")
fq <- Days ~ Eth + Sex + Age + Lrn
d_fixed <- fixed(negbin2_distrib(), theta = 1.5)
d_fixed@params
fit_fx <- statmod(fq, distrib = d_fixed, data = quine, outer_criterion = ml())
summary(fit_fx)
glm_fx <- glm(fq, family = MASS::negative.binomial(theta = 1.5), data = quine)
rbind(statmod = coef(fit_fx)$mu, glm = coef(glm_fx))
c(statmod = as.numeric(logLik(fit_fx)), glm = as.numeric(logLik(glm_fx)))
```

The distribution has the parameter `mu` only. The coefficients agree to about
four digits, which is the precision of the iteration of `glm()`, and the
log-likelihoods agree to the printed digits. The same wrapper applies to any family: the equation for
`nu` of a Student $t$ model is replaced by a known degrees of freedom with
`fixed(student_t1_distrib(), nu = 4)`.

## A distribution defined by its density {#sec-distribs-own}

A distribution that the package does not provide needs one function, its
density. The log-logistic distribution has the density
$$
f(y) = \frac{(\beta/\alpha)(y/\alpha)^{\beta - 1}}
{\bigl(1 + (y/\alpha)^\beta\bigr)^2}, \qquad y > 0,
$$
with a scale $\alpha$, which is the median, and a shape $\beta$
[@bennett1983]. We define a class that inherits from `continuous_distrib`,
the method `distrib_pdf()` for it, and a constructor with the names, the bounds
and the links of the parameters:

```{r}
#| label: own-family
LogLogistic <- S7::new_class("LogLogistic", parent = continuous_distrib)
S7::method(distrib_pdf, LogLogistic) <- function(distrib, y, theta, log = FALSE,
                                                 ...) {
  a <- theta[[1]]
  b <- theta[[2]]
  z <- y / a
  ld <- log(b) - log(a) + (b - 1) * log(z) - 2 * log1p(z^b)
  if (log) ld else exp(ld)
}
loglogistic_distrib <- function(link_mu = log_link(), link_shape = log_link()) {
  LogLogistic(distrib_name = "log-logistic", dimension = "univariate",
              bounds = c(0, Inf), params = c("mu", "shape"),
              params_interpretation = c(mu = "scale", shape = "shape"),
              n_params = 2,
              params_bounds = list(mu = c(0, Inf), shape = c(0, Inf)),
              link_params = list(mu = link_mu, shape = link_shape))
}
d_ll <- loglogistic_distrib()
distrib_pdf(d_ll, y = 2, theta = list(mu = 3, shape = 2))
```

That is all the definition. The cdf, the quantile function, the random
generator, the score, the Hessian and the expected information come from
the numerical fallbacks of `distributions7`. The function `check_distrib()`
tests them against each other:

```{r}
#| label: own-check
chk_ll <- check_distrib(d_ll, theta = list(mu = 3, shape = 2))
```

The `r nrow(chk_ll)` checks in the table pass. The `r nrow(attr(chk_ll, "skipped"))` checks of the derivatives in the
parameters are not run, and the table gives the reason: the family has no
derivatives of its own, so their values are the numerical fallbacks, and a
comparison with finite differences would compare the fallbacks with
themselves. The distribution is now an argument of `statmod()`.
We fit it to the `GAGurine` data, and we compare the fit with `survreg()` of
the `survival` package, which fits the log-logistic distribution to a
response without censoring through a parametrization of the logarithm of the
response:

```{r}
#| label: own-fit
fit_ll <- statmod(GAG ~ Age, distrib = loglogistic_distrib(), data = GAGurine,
                  outer_criterion = ml())
summary(fit_ll)
sv_ll <- survival::survreg(survival::Surv(GAG) ~ Age, data = GAGurine,
                           dist = "loglogistic")
rbind(statmod = c(coef(fit_ll)$mu, coef(fit_ll)$shape),
      survreg = c(coef(sv_ll), -log(sv_ll$scale)))
c(statmod = as.numeric(logLik(fit_ll)), survreg = as.numeric(logLik(sv_ll)))
```

In `survreg()` the logarithm of the response is $X\beta + s W$, with $W$
a standard logistic variable, so $\alpha = \exp(X\beta)$ and $\beta = 1/s$.
The two fits give the same coefficients and the same log-likelihood.

```{r}
#| label: gate-05
#| include: false
assert_families_ok(
  fits = list(hg = fit_hg, ht = fit_ht, ht_ml = fit_ht_ml, ht0 = fit_ht0,
              ht1 = fit_ht1, bt = fit_bt, bg = fit_bg, sg = fit_sg,
              sn1 = fit_sn1, ss = fit_ss, st = fit_st, ss_ml = fit_ss_ml, b0 = fit_b0,
              b0_ml = fit_b0_ml, b1 = fit_b1, zp = fit_zp, zn = fit_zn,
              zi = fit_zi, za = fit_za, zi2 = fit_zi2, zi_ml = fit_zi_ml,
              za_ml = fit_za_ml, tr = fit_tr, tg = fit_tg, fo = fit_fo,
              tx = fit_tx, ln = fit_ln, ln_b = fit_ln_b, fx = fit_fx,
              ll = fit_ll),
  ref = list(gl_ht = gl_ht, gl_be = gl_be, tmb_be = tmb_be, tmb_zi = tmb_zi,
             tmb_za = tmb_za, glm_fx = glm_fx, sv_ll = sv_ll),
  data = list(gag = GAGurine, cars93 = cars93, swiss = swiss_p,
              solder = solder, sim_t = sim_t$data, sim_f = sim_f$data,
              quine = quine),
  claims = list(lr_ht = lr_ht, lr_b = lr_b, cert_bt = cert_bt,
                skew_lm = skew_lm, gamma_max = gamma_max,
                nll_sn = nll_sn, opt_sn = opt_sn, opt_tr = opt_tr,
                opt_fo = opt_fo, chk_ll = chk_ll, phi_table = phi_table,
                skew_table = skew_table, d_fixed = d_fixed, d_ll = d_ll))
```
