# Generalized additive models {#sec-gam}

```{r}
#| label: setup-06
#| include: false
source("R/_setup.R")
source("R/gam-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"; ink <- "grey15"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
dec <- function(x, s = 3) format(signif(x, s), scientific = FALSE, big.mark = ",")
```

A generalized additive model writes the predictor of a parameter as a sum of
smooth functions of the covariates [@wood2017]. In the previous chapters the
smooth term `s()` appeared with its default construction, a basis of cubic
B-splines with a penalty on the second derivative. This chapter shows the
other constructions. The second argument of `s()` is a *smoother*, an object
of the `basis7` package that fixes four things: the basis, the penalty, the
treatment of the functions that the penalty leaves free (the null space), and
the coordinates of the coefficients. The chapter first varies these choices
for the B-spline smoother, then shows the other families of smoothers and the
penalties built from differential operators. It then shows smooths that change
with another variable (`by =`), smooths of two covariates (`te()`), smooths in
the equations of the other parameters, smooths with non-gaussian responses,
and a smooth with a lasso penalty in place of the penalty on the roughness.

The package `mgcv` fits the same models [@wood2017], and we compare the two
packages in each section. We call it as `mgcv::gam()` without attaching
`mgcv`, because `mgcv` has functions called `s()` and `te()` too. Both
packages choose the smoothing parameters by REML.

## B-spline smooths {#sec-gam-bspline}

The data frame `Boston` of the `MASS` package has, for 506 census tracts of
Boston in 1970, the median value of the owner-occupied homes (`medv`, in
thousands of dollars) and the percentage of the population with a lower
status (`lstat`) [@harrison1978]. The value falls quickly with `lstat` at the
left and slowly at the right. We fit a smooth function of `lstat` with the
default smoother, and with 20 basis functions in place of the default 10:

```{r}
#| label: fit-boston
data(Boston, package = "MASS")
fit_b20 <- statmod(medv ~ s(lstat, bspline_smooth(k = 20)),
                   distrib = gaussian1_distrib(), data = Boston)
summary(fit_b20)
```

The function `bspline_smooth()` builds the smoother. Its argument `k` is the
number of B-splines, `degree` is their degree (3 by default), and `order` is
the order $m$ of the derivative in the penalty
$\lambda \int f^{(m)}(x)^2\,dx$ (2 by default). The same model in `mgcv` is
the smooth with `bs = "bs"`. We compare the effective degrees of freedom of
the mean equation, the estimate of $\sigma$ and the fitted values:

```{r}
#| label: boston-mgcv
g_b20 <- mgcv::gam(medv ~ s(lstat, bs = "bs", k = 20), data = Boston,
                   method = "REML")
edf_mu <- function(fit) sum(fit@edf$edf[fit@edf$parameter == "mu"])
c(statmod = edf_mu(fit_b20), mgcv = sum(g_b20$edf))
c(statmod = exp(coef(fit_b20)$sigma[[1]]), mgcv = sqrt(g_b20$sig2))
max(abs(fitted(fit_b20) - fitted(g_b20)))
```

The two fits are the same. They use the same basis and the same penalty, and
both estimate $\sigma$ on the REML criterion, which integrates the
coefficients of the mean (@sec-beyond-outer).

### The number of basis functions

The number `k` sets the largest number of degrees of freedom that the smooth
can have, and the smoothing parameter sets how many of them the fit uses.
We fit the model with 10, 20 and 40 basis functions:

```{r}
#| label: boston-k
fit_b10 <- statmod(medv ~ s(lstat, bspline_smooth(k = 10)),
                   distrib = gaussian1_distrib(), data = Boston)
fit_b40 <- statmod(medv ~ s(lstat, bspline_smooth(k = 40)),
                   distrib = gaussian1_distrib(), data = Boston)
k_table <- data.frame(
  k = c(10, 20, 40),
  edf = c(edf_mu(fit_b10), edf_mu(fit_b20), edf_mu(fit_b40)),
  lambda = c(hyper(fit_b10)$estimate, hyper(fit_b20)$estimate,
             hyper(fit_b40)$estimate),
  AIC = c(AIC(fit_b10), AIC(fit_b20), AIC(fit_b40)))
k_table
```

From 10 to 20 basis functions the effective degrees of freedom grow from
`r fmt(k_table$edf[1])` to `r fmt(k_table$edf[2])`, so with 10 functions the
basis limited the curve. From 20 to 40 they grow only to
`r fmt(k_table$edf[3])`: the penalty holds the curve, and a larger basis
changes little. A basis larger than the curve needs is therefore safe, and it
costs only computing time.

### The order of the penalty

The penalty $\int f^{(m)}(x)^2\,dx$ is zero for the polynomials of degree
below $m$. These polynomials form the null space of the penalty, and the
smooth estimates them without a penalty. With $m = 1$ the null space holds
the constants, and a large $\lambda$ makes $f$ flat. With $m = 2$ it holds the
straight lines, and with $m = 3$ the parabolas. The intercept of the equation
already represents the constant, so the smooth has one free coefficient for
$m = 2$ (`lin`) and two for $m = 3$ (`lin` and `poly2`):

```{r}
#| label: boston-order
fit_o1 <- statmod(medv ~ s(lstat, bspline_smooth(k = 20, order = 1)),
                  distrib = gaussian1_distrib(), data = Boston)
fit_o3 <- statmod(medv ~ s(lstat, bspline_smooth(k = 20, order = 3)),
                  distrib = gaussian1_distrib(), data = Boston)
names(coef(fit_o3)$mu)[1:4]
g_o1 <- mgcv::gam(medv ~ s(lstat, bs = "bs", k = 20, m = c(3, 1)),
                  data = Boston, method = "REML")
g_o3 <- mgcv::gam(medv ~ s(lstat, bs = "bs", k = 20, m = c(3, 3)),
                  data = Boston, method = "REML")
data.frame(order = 1:3,
           edf = c(edf_mu(fit_o1), edf_mu(fit_b20), edf_mu(fit_o3)),
           mgcv = c(sum(g_o1$edf), sum(g_b20$edf), sum(g_o3$edf)))
```

In `mgcv` the order is the second element of `m`. The fits agree again. On
these data the penalty on the first derivative gives the most effective
degrees of freedom and the most wiggly curve, and the penalty on the third
derivative the fewest. @fig-boston-order shows the three curves.

```{r}
#| label: fig-boston-order
#| fig-cap: "Median value of the homes against the percentage of lower-status population, with the smooths fitted with a penalty on the first, second and third derivative."
#| echo: false
#| fig-height: 4
lg <- data.frame(lstat = seq(from = min(Boston$lstat), to = max(Boston$lstat),
                             length.out = 200))
order_lines <- data.frame(
  lstat = rep(lg$lstat, times = 3),
  medv = c(predict(fit_o1, what = "mu", newdata = lg),
           predict(fit_b20, what = "mu", newdata = lg),
           predict(fit_o3, what = "mu", newdata = lg)),
  curve = factor(rep(1:3, each = nrow(lg)),
                 labels = c("order 1", "order 2", "order 3")))
ggplot(Boston, aes(x = lstat, y = medv)) +
  geom_point(colour = book_grey, size = 0.8) +
  geom_line(data = order_lines, aes(colour = curve, linetype = curve),
            linewidth = 0.8) +
  scale_colour_manual(values = c(rust, chalk, ink)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted")) +
  labs(x = "lower-status population (%)", y = "median value (1000 $)")
```

### The null space

The argument `null_space` decides what happens to the functions that the
penalty leaves free. With `"keep"`, the default, they have no penalty. With
`"shrink"` they get a small penalty of their own [@marra2012], so a large
$\lambda$ can remove the whole term, the straight line included. This is
useful when a model has many smooths and some covariates may have no
effect. We add to the model the number of rooms per dwelling (`rm`) and the
proportion of the town's area used by non-retail business (`indus`), and we
fit it with both settings:

```{r}
#| label: fit-shrink
fit_sk <- statmod(medv ~ s(lstat, bspline_smooth(k = 10)) +
                    s(rm, bspline_smooth(k = 10)) +
                    s(indus, bspline_smooth(k = 10)),
                  distrib = gaussian1_distrib(), data = Boston)
shr <- bspline_smooth(k = 10, null_space = "shrink")
fit_ss <- statmod(medv ~ s(lstat, shr) + s(rm, shr) + s(indus, shr),
                  distrib = gaussian1_distrib(), data = Boston)
fit_sk@edf
fit_ss@edf
c(keep = AIC(fit_sk), shrink = AIC(fit_ss))
```

With `"keep"` the smooth of `indus` keeps its straight line and uses one
effective degree of freedom. With `"shrink"` its effective degrees of
freedom fall to `r fmt(fit_ss@edf$edf[4], 3)`, so the term has left the model,
while the smooths of `lstat` and `rm` change little. The AIC is smaller with
the shrinkage.

## Other smoothers {#sec-gam-families}

The package `basis7` has five more families of smoothers:
`pspline_smooth()`, `legendre_smooth()`, `adaptive_smooth()`,
`cyclic_smooth()` and `fourier_smooth()`. Each one goes in the same place
as `bspline_smooth()`, as the second argument of `s()`.

### P-splines

A P-spline has a B-spline basis and a penalty on the differences of the
coefficients in place of the integral of a derivative [@eilers1996]:
$$
\lambda \sum_j \bigl(\Delta^d b_j\bigr)^2 ,
$$
where $\Delta^d$ is the difference of order $d$ (2 by default). The
differences are taken on the coefficients of the basis whose knots are
equally spaced also outside the range of the data, which is the construction
of @eilers1996. The penalty is zero for the coefficients that are a
polynomial of degree below $d$ in their index, and these give a polynomial of
degree below $d$ in $x$. The same model in `mgcv` has `bs = "ps"`:

```{r}
#| label: fit-pspline
fit_ps <- statmod(medv ~ s(lstat, pspline_smooth(k = 20)),
                  distrib = gaussian1_distrib(), data = Boston)
g_ps <- mgcv::gam(medv ~ s(lstat, bs = "ps", k = 20), data = Boston,
                  method = "REML")
c(statmod = edf_mu(fit_ps), mgcv = sum(g_ps$edf))
max(abs(fitted(fit_ps) - fitted(g_ps)))
c(bspline = AIC(fit_b20), pspline = AIC(fit_ps))
```

The P-spline gives almost the same fit as the B-spline smooth with the
penalty on the second derivative. The two penalties are different, but both
are zero for straight lines and both grow with the curvature of the
function.

### Global polynomials

The smoother `legendre_smooth()` has a basis of Legendre polynomials, which
are polynomials over the whole range of the covariate. Each function of a
B-spline basis is zero outside a short interval, and each Legendre polynomial
changes over the whole range. A penalized polynomial suits a function that is
smooth everywhere. On the `Boston` data the B-spline smooth has the smaller
AIC:

```{r}
#| label: fit-legendre
fit_lg <- statmod(medv ~ s(lstat, legendre_smooth(k = 8)),
                  distrib = gaussian1_distrib(), data = Boston)
c(bspline = edf_mu(fit_b20), legendre = edf_mu(fit_lg))
c(bspline = AIC(fit_b20), legendre = AIC(fit_lg))
```

### Adaptive smooths

A single smoothing parameter smooths the curve by the same amount over the
whole range of $x$. The acceleration of the `mcycle` data of @sec-model is
close to zero for the first 14 milliseconds and changes quickly afterwards. The
smoother `adaptive_smooth()` gives the differences of the P-spline penalty a
weight that changes along $x$:
$$
\sum_j w_j \bigl(\Delta^d b_j\bigr)^2, \qquad
w_j = \sum_{i=1}^{m} \lambda_i v_i(j),
$$
where $v_1, \dots, v_m$ are B-splines over the index $j$ [@wood2017]. The
smooth has $m$ smoothing parameters (5 by default), and REML estimates all of
them:

```{r}
#| label: fit-adaptive
data(mcycle, package = "MASS")
fit_ad <- statmod(accel ~ s(times, adaptive_smooth(k = 40, m = 5)),
                  distrib = gaussian1_distrib(), data = mcycle)
fit_p40 <- statmod(accel ~ s(times, pspline_smooth(k = 40)),
                   distrib = gaussian1_distrib(), data = mcycle)
hyper(fit_ad)[, c("name", "estimate")]
g_ad <- mgcv::gam(accel ~ s(times, bs = "ad", k = 40, m = 5), data = mcycle,
                  method = "REML")
c(statmod = edf_mu(fit_ad), mgcv = sum(g_ad$edf))
max(abs(fitted(fit_ad) - fitted(g_ad)))
data.frame(edf = c(edf_mu(fit_p40), edf_mu(fit_ad)),
           AIC = c(AIC(fit_p40), AIC(fit_ad)),
           row.names = c("pspline", "adaptive"))
```

The adaptive smooth agrees with `mgcv`'s `bs = "ad"`. The five smoothing
parameters span many orders of magnitude: the curve is smoothed strongly where
the weight is large and little where it is small. The adaptive smooth uses
fewer effective degrees of freedom than the P-spline and has the smaller AIC.
@fig-adaptive shows that the P-spline curve wiggles after about 40
milliseconds, where the data are noisy around zero, and the adaptive curve is
flat there.

```{r}
#| label: fig-adaptive
#| fig-cap: "Head acceleration against time, with the P-spline smooth and the adaptive smooth, both with 40 basis functions."
#| echo: false
#| fig-height: 4
mg <- data.frame(times = seq(from = min(mcycle$times), to = max(mcycle$times),
                             length.out = 300))
adapt_lines <- data.frame(
  times = rep(mg$times, times = 2),
  accel = c(predict(fit_p40, what = "mu", newdata = mg),
            predict(fit_ad, what = "mu", newdata = mg)),
  curve = factor(rep(1:2, each = nrow(mg)), labels = c("P-spline", "adaptive")))
ggplot(mcycle, aes(x = times, y = accel)) +
  geom_point(colour = book_grey, size = 1) +
  geom_line(data = adapt_lines, aes(colour = curve, linetype = curve),
            linewidth = 0.8) +
  scale_colour_manual(values = c(rust, chalk)) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = "time after impact (ms)", y = "acceleration (g)")
```

### Periodic smooths

The data frame `chicagoNMMAPS` of the `dlnm` package [@gasparrini2011] has,
for each day from 1987 to 2000 in Chicago, the mean temperature in degrees
Celsius, the number of deaths and other variables [@peng2008]. The column
`doy` is the day of the year, from 1 to 366. We keep the six years from 1995 to
2000, and we model the temperature as a smooth function of the day of the
year. The temperature on 31 December and on 1 January should be close, since
the two days are adjacent. A B-spline smooth has no constraint that joins the
two ends of the year. The smoother
`cyclic_smooth()` has a periodic basis of B-splines, whose functions and
first two derivatives take the same value at `lower` and at `upper`:

```{r}
#| label: fit-cyclic
data(chicagoNMMAPS, package = "dlnm")
chic <- chicagoNMMAPS[chicagoNMMAPS$year >= 1995, ]
fit_tb <- statmod(temp ~ s(doy, bspline_smooth(k = 10)),
                  distrib = gaussian1_distrib(), data = chic)
fit_tc <- statmod(temp ~ s(doy, cyclic_smooth(k = 10, lower = 0, upper = 366)),
                  distrib = gaussian1_distrib(), data = chic)
ends <- data.frame(doy = c(1, 366))
rbind(bspline = predict(fit_tb, what = "mu", newdata = ends),
      cyclic = predict(fit_tc, what = "mu", newdata = ends))
```

The B-spline smooth gives a temperature of
`r fmt(predict(fit_tb, what = "mu", newdata = ends)[1])` degrees on the first
day of the year and `r fmt(predict(fit_tb, what = "mu", newdata = ends)[2])`
on the last. The two ends of its curve are estimated from the observations
near each end separately, and the difference is
`r fmt(abs(diff(predict(fit_tb, what = "mu", newdata = ends))))` degrees. The
cyclic smooth joins the two ends. @fig-cyclic shows the curves near the end
of the year and near its beginning.

```{r}
#| label: fig-cyclic
#| fig-cap: "Mean daily temperature in Chicago from 1995 to 2000 against the day of the year, with the B-spline and the cyclic smooth. The right panel repeats the last 60 days of the year before the first 60 days, so the two ends of the year are adjacent."
#| echo: false
#| fig-height: 3.8
#| fig-width: 9
dg <- data.frame(doy = seq(from = 1, to = 366, length.out = 366))
pb <- predict(fit_tb, what = "mu", newdata = dg)
pc <- predict(fit_tc, what = "mu", newdata = dg)
sh <- c(dg$doy[307:366] - 366, dg$doy[1:60])
cyc_panels <- c("day of the year", "days from 1 January")
cyc_lines <- data.frame(
  day = c(dg$doy, dg$doy, sh, sh),
  temp = c(pb, pc, pb[c(307:366, 1:60)], pc[c(307:366, 1:60)]),
  curve = factor(rep(c(1, 2, 1, 2), times = c(366, 366, 120, 120)),
                 labels = c("B-spline", "cyclic")),
  panel = factor(rep(1:2, times = c(732, 240)), labels = cyc_panels))
cyc_data <- data.frame(day = chic$doy, temp = chic$temp,
                       panel = factor(cyc_panels[1], levels = cyc_panels))
cyc_zero <- data.frame(panel = factor(cyc_panels[2], levels = cyc_panels))
ggplot(cyc_lines, aes(x = day, y = temp)) +
  geom_point(data = cyc_data, colour = book_grey, size = 0.4) +
  geom_vline(data = cyc_zero, aes(xintercept = 0), linetype = "dotted",
             colour = book_grey) +
  geom_line(aes(colour = curve, linetype = curve), linewidth = 0.8) +
  facet_wrap(facets = vars(panel), scales = "free", strip.position = "bottom") +
  scale_colour_manual(values = c(rust, chalk)) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = NULL, y = "temperature (C)") +
  theme(strip.placement = "outside", strip.background = element_blank(),
        strip.text = element_text(size = rel(1)))
```

The package `mgcv` has a cyclic cubic regression spline, `bs = "cc"`. Its
functions are the same periodic cubic splines as those of `cyclic_smooth()`,
but the argument `k` counts a different thing in the two packages. In
`cyclic_smooth()`, `k` is the number of periodic basis functions, and the
knots of the B-splines divide the period into `k` intervals of equal length.
In `bs = "cc"`, `k` is the number of knots, and the first and the last knot
are the two ends of the period, so the knots divide the period into `k - 1`
intervals. The two packages then remove the constant, which the intercept of
the model carries. With `k = 10`, `cyclic_smooth()` therefore has 9
coefficients and `bs = "cc"` has 8. The penalty is the same in the two
packages: the integral of the squared second derivative over one period.
The coefficients of `bs = "cc"` are the values of the function at the knots,
and the coefficients of `cyclic_smooth()` are in the Demmler-Reinsch
coordinates, but a change of coordinates does not change the fit. With
`k = 11` and eleven knots at equal distances from 0 to 366, `bs = "cc"` has
the same functions and the same penalty as `cyclic_smooth(k = 10)`:

```{r}
#| label: cyclic-mgcv
g_cc <- mgcv::gam(temp ~ s(doy, bs = "cc", k = 11), data = chic,
                  method = "REML",
                  knots = list(doy = seq(from = 0, to = 366, length.out = 11)))
c(statmod = length(coef(fit_tc)$mu) - 1, mgcv = length(coef(g_cc)) - 1)
c(statmod = edf_mu(fit_tc), mgcv = sum(g_cc$edf))
max(abs(fitted(fit_tc) - fitted(g_cc)))
```

The two fits are the same model. Their fitted values differ only by the
precision with which REML locates the smoothing parameter. The package `mgcv`
also has a cyclic P-spline, `bs = "cp"`, which we use in @sec-gam-sigma. Its
penalty is on the differences of the coefficients, as for `pspline_smooth()`,
and it is a different smooth.

### Fourier smooths

The smoother `fourier_smooth()` has a basis of sines and cosines. With
$T$ = `upper - lower` the period and $\nu = 2\pi/T$, the basis functions are
the constant and the pairs $\sin(j\nu x)$, $\cos(j\nu x)$ for
$j = 1, 2, \dots$. The argument `k` is the number of basis functions, and
with `k = 9` the basis has the constant and the first four pairs. The smooth
removes the constant, which the intercept carries, and it has 8
coefficients. Every basis function takes the same value and has the same
derivatives at `lower` and at `upper`, so the fitted curve is periodic. Like
the Legendre polynomials, the sines and the cosines change over the whole
range of the covariate, while each periodic B-spline of `cyclic_smooth()` is
zero outside a short interval:

```{r}
#| label: fit-fourier
fit_tf <- statmod(temp ~ s(doy, fourier_smooth(k = 9, lower = 0, upper = 366)),
                  distrib = gaussian1_distrib(), data = chic)
predict(fit_tf, what = "mu", newdata = ends)
data.frame(edf = c(edf_mu(fit_tc), edf_mu(fit_tf)),
           AIC = c(AIC(fit_tc), AIC(fit_tf)),
           row.names = c("cyclic", "Fourier"))
```

The two periodic smooths have almost the same AIC. The yearly temperature
changes slowly over the year, and a few harmonics describe it well. The
Fourier smooth uses fewer effective degrees of freedom than the cyclic one. A
local basis is better for a seasonal pattern with a narrow peak, which the
first harmonics of a Fourier basis cannot follow. The penalty of
`fourier_smooth()` is not the integral of the squared second derivative. It
is built from a differential operator, which is the subject of the next
section.

## Penalties from differential operators {#sec-gam-operators}

The penalty of a smooth measures how far the function is from the functions
of its null space. The penalty on the derivative of order $m$ measures the
distance from the polynomials of degree below $m$. A linear differential
operator
$$
L f = w_0 f + w_1 f' + \dots + w_{m-1} f^{(m-1)} + f^{(m)}
$$
gives the penalty $\lambda \int (Lf)^2\,dx$, whose null space is the set of
the solutions of $Lf = 0$ [@ramsay2005]. The argument `order` of a smoother
takes such an operator, and a number `m` is the operator
`deriv_operator(m)`, with $Lf = f^{(m)}$. The package `basis7` has three other
operators:

- `harmonic_operator(period)` is $L f = \nu^2 f' + f'''$, with
  $\nu = 2\pi/T$ and $T$ the period. Its null space holds the constant,
  $\sin(\nu x)$ and $\cos(\nu x)$;
- `oscillator_operator(period)` is $L f = \nu^2 f + f''$, the equation of a
  harmonic oscillator. Its null space holds $\sin(\nu x)$ and $\cos(\nu x)$,
  without the constant;
- `linear_operator(w)` is the operator with the weights
  $w = (w_0, \dots, w_{m-1})$.

The function `operator_null()` returns the null space of an operator:

```{r}
#| label: operator-null
operator_null(deriv_operator(2))
operator_null(harmonic_operator(period = 366))
operator_null(oscillator_operator(period = 366))
```

When $\lambda$ is large the penalty forces $Lf$ close to zero, and the fitted
curve is close to a function of the null space. The functions of the null
space are free coefficients of the smooth, which the penalty does not shrink.

### The harmonic operator on a Fourier basis

The harmonic operator, with the period `upper - lower`, is the default penalty
of `fourier_smooth()`. The fit `fit_tf` has two free coefficients, `sin1` and
`cos1`, and a large $\lambda$ makes its curve the constant plus one sine wave
of the period, which is the simplest periodic function. The penalty on the
second derivative (`order = 2`) has only the constant in its null space on a
Fourier basis, because a straight line is not periodic. With this penalty a
large $\lambda$ makes the curve a constant. The argument `harmonics` of
`harmonic_operator()` gives the number of harmonics in the null space, and
with `harmonics = 2` the smooth has four free coefficients:

```{r}
#| label: fourier-operator
names(coef(fit_tf)$mu)
fit_tf2 <- statmod(temp ~ s(doy, fourier_smooth(k = 9, lower = 0, upper = 366,
                                                order = 2)),
                   distrib = gaussian1_distrib(), data = chic)
op_h2 <- harmonic_operator(period = 366, harmonics = 2)
fit_tf3 <- statmod(temp ~ s(doy, fourier_smooth(k = 9, lower = 0, upper = 366,
                                                order = op_h2)),
                   distrib = gaussian1_distrib(), data = chic)
names(coef(fit_tf3)$mu)[2:5]
data.frame(edf = c(edf_mu(fit_tf), edf_mu(fit_tf2), edf_mu(fit_tf3)),
           AIC = c(AIC(fit_tf), AIC(fit_tf2), AIC(fit_tf3)),
           row.names = c("harmonic", "second derivative", "two harmonics"))
```

The three penalties give close fits. The yearly temperature is close to one
sine wave, and the harmonic operator shrinks the curve toward that wave, while
the penalty on the second derivative shrinks it toward a constant. The
harmonic operator with one harmonic has the smallest AIC and the fewest
effective degrees of freedom.

### A yearly cycle on a B-spline basis

An operator can penalize a basis that is not periodic. We write the
temperature as a function of the day since 1 January 1995, over the six
years, and we penalize a B-spline smooth by `harmonic_operator(period =
365.25)`. The smoother builds the null space of the operator from its exact
functions: the smooth has a free sine and a free cosine of period 365.25 days
beside the B-splines, and the penalized B-splines are made orthogonal to
them. We fit the smooth with 20 and with 60 B-splines, and we compare it with
the penalty on the second derivative:

```{r}
#| label: day-operator
chic$day <- as.numeric(chic$date - min(chic$date))
op_year <- harmonic_operator(period = 365.25)
fit_d20 <- statmod(temp ~ s(day, bspline_smooth(k = 20)),
                   distrib = gaussian1_distrib(), data = chic)
fit_h20 <- statmod(temp ~ s(day, bspline_smooth(k = 20, order = op_year)),
                   distrib = gaussian1_distrib(), data = chic)
fit_d60 <- statmod(temp ~ s(day, bspline_smooth(k = 60)),
                   distrib = gaussian1_distrib(), data = chic)
fit_h60 <- statmod(temp ~ s(day, bspline_smooth(k = 60, order = op_year)),
                   distrib = gaussian1_distrib(), data = chic)
names(coef(fit_h20)$mu)[1:4]
dfs <- list(fit_d20, fit_h20, fit_d60, fit_h60)
data.frame(edf = sapply(dfs, edf_mu),
           sigma = sapply(dfs, function(f) exp(coef(f)$sigma[[1]])),
           AIC = sapply(dfs, AIC),
           row.names = c("second derivative, k = 20", "harmonic, k = 20",
                         "second derivative, k = 60", "harmonic, k = 60"))
```

Twenty B-splines over six years are fewer than four for each year. With the
penalty on the second derivative the smooth uses
`r fmt(edf_mu(fit_d20))` effective degrees of freedom of 20, so the penalty is
almost zero, and the basis alone cannot follow the yearly cycle. With the
harmonic operator the exact sine and cosine of the year are in the smooth, and
its AIC is lower by `r fmt(AIC(fit_d20) - AIC(fit_h20), 1)`. With 60
B-splines, ten for each year, the basis can follow the cycle, and the smooth
with the penalty on the second derivative has the smallest AIC of the four
fits. Each year differs a little from the average year, and this smooth
follows those differences. The harmonic operator shrinks the curve toward one
sine wave that repeats over the six years, and it removes part of the
differences between the years.

### An exponential decay

The data frame `wtloss` of the `MASS` package has the weight in kilograms of
an obese patient on 52 days of a diet of 246 days [@venables2002]. The weight
falls quickly at first and then more slowly, and @venables2002 fit it with the
model
$$
W(t) = \beta_0 + \beta_1\, 2^{-t/\theta},
$$
a decay toward the weight $\beta_0$, in which $\theta$ is the time the excess
weight takes to halve. With $a = \log(2)/\theta$ the function $2^{-t/\theta}$
is $e^{-a t}$. The operator $L f = a f' + f''$ has in its null space the
constant and $e^{-a t}$, so the operator `linear_operator(c(0, a))` shrinks
a smooth toward the curves of this model. We estimate $\theta$ with `nls()`
and we use the rate $a$ from that fit:

```{r}
#| label: wtloss-operator
data(wtloss, package = "MASS")
fit_nls <- nls(Weight ~ b0 + b1 * 2^(-Days / th), data = wtloss,
               start = c(b0 = 90, b1 = 95, th = 120))
coef(fit_nls)
a <- log(2) / coef(fit_nls)[["th"]]
op_exp <- linear_operator(w = c(0, a))
operator_null(op_exp)
fit_w2 <- statmod(Weight ~ s(Days, bspline_smooth(k = 10)),
                  distrib = gaussian1_distrib(), data = wtloss)
fit_we <- statmod(Weight ~ s(Days, bspline_smooth(k = 10, order = op_exp)),
                  distrib = gaussian1_distrib(), data = wtloss)
data.frame(edf = c(edf_mu(fit_w2), edf_mu(fit_we)),
           lambda = c(hyper(fit_w2)$estimate, hyper(fit_we)$estimate),
           AIC = c(AIC(fit_w2), AIC(fit_we)),
           row.names = c("second derivative", "exponential"))
max(abs(fitted(fit_we) - fitted(fit_nls)))
```

REML makes the smoothing parameter of the exponential operator very large, and
the smooth uses `r fmt(edf_mu(fit_we))` effective degrees of freedom, the
intercept and the coefficient of $e^{-at}$. Its fitted curve is the curve of
`nls()` to `r dec(max(abs(fitted(fit_we) - fitted(fit_nls))))` kilograms. The
data are close to the model, and REML keeps no departure from it. The smooth with the penalty on the second derivative
uses `r fmt(edf_mu(fit_w2))` effective degrees of freedom and has the larger
AIC. We took the rate from `nls()`, so the AIC of the exponential smooth does
not count the estimation of $\theta$.

### A trend with a yearly cycle

Operators can be multiplied, and the product has the null spaces of both
factors. The data set `co2` of R gives the monthly concentration of carbon
dioxide at Mauna Loa from 1959 to 1997, in parts per million. The series has
a trend and a yearly cycle. The product of `deriv_operator(2)` and
`oscillator_operator(period = 1)` has a null space with the straight lines
and the sine and cosine of period one year:

```{r}
#| label: co2-operator
op_co2 <- deriv_operator(2) * oscillator_operator(period = 1)
op_co2
co2d <- data.frame(t = as.numeric(time(co2)), y = as.numeric(co2))
fit_c1 <- statmod(y ~ s(t, bspline_smooth(k = 40)),
                  distrib = gaussian1_distrib(), data = co2d)
fit_c3 <- statmod(y ~ s(t, bspline_smooth(k = 120)),
                  distrib = gaussian1_distrib(), data = co2d)
fit_c2 <- statmod(y ~ s(t, bspline_smooth(k = 40, degree = 5, order = op_co2)),
                  distrib = gaussian1_distrib(), data = co2d)
data.frame(edf = c(edf_mu(fit_c1), edf_mu(fit_c3), edf_mu(fit_c2)),
           sigma = c(exp(coef(fit_c1)$sigma[[1]]), exp(coef(fit_c3)$sigma[[1]]),
                     exp(coef(fit_c2)$sigma[[1]])),
           AIC = c(AIC(fit_c1), AIC(fit_c3), AIC(fit_c2)),
           row.names = c("second derivative, k = 40",
                         "second derivative, k = 120", "trend and cycle"))
```

The operator has order 4, and the B-splines of degree 3 have a fourth
derivative equal to zero, so the smoother needs `degree = 5`. With the
penalty on the second derivative the smooth follows the trend and misses the
cycle, also with 120 basis functions, which are three for each year: the
cycle has a large second derivative, and this penalty removes it. With the
operator, the sine and the cosine of the year are free coefficients of the
smooth, and the residual standard deviation falls from
`r fmt(exp(coef(fit_c1)$sigma[[1]]))` to `r fmt(exp(coef(fit_c2)$sigma[[1]]))`
parts per million. @fig-co2 shows the last five years.

```{r}
#| label: fig-co2
#| fig-cap: "Monthly CO2 concentration at Mauna Loa from 1993 to 1997, with the smooth penalized on the second derivative and the smooth penalized by the operator whose null space holds a trend and a yearly cycle."
#| echo: false
#| fig-height: 3.8
last <- co2d$t >= 1993
co2_lines <- data.frame(
  t = rep(co2d$t[last], times = 2),
  y = c(fitted(fit_c1)[last], fitted(fit_c2)[last]),
  curve = factor(rep(1:2, each = sum(last)),
                 labels = c("second derivative", "trend and cycle")))
ggplot(co2d[last, ], aes(x = t, y = y)) +
  geom_point(colour = book_grey, size = 1.1) +
  geom_line(data = co2_lines, aes(colour = curve, linetype = curve),
            linewidth = 0.8) +
  scale_colour_manual(values = c(rust, chalk)) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = "year", y = "CO2 (ppm)")
```

## Smooths that change with another variable {#sec-gam-by}

### One curve for each level of a factor

The data frame `ChickWeight` gives the weight, in grams, of
`r nlevels(ChickWeight$Chick)` chicks from birth to day 21 [@crowder1990].
The chicks were weighed every second day, and each chick received one of four
diets. The argument `by = Diet` of `s()` builds one smooth curve of time for
each diet. The curves of a factor `by` are centered, so the formula also has
the main effect `Diet`. The spread of the weights grows with the age of the
chicks, so we give `sigma` a linear equation in time:

```{r}
#| label: fit-chick
fit_cs <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8), by = Diet) |
                    sigma ~ Time,
                  distrib = gaussian1_distrib(), data = ChickWeight)
fit_cl <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8), by = Diet,
                                    by_hyper = "level") |
                    sigma ~ Time,
                  distrib = gaussian1_distrib(), data = ChickWeight)
hyper(fit_cs)[, c("name", "estimate")]
hyper(fit_cl)[, c("name", "estimate")]
```

With `by_hyper = "shared"`, the default, the four curves share one smoothing
parameter. With `by_hyper = "level"` each curve has its own. In `mgcv` the
curves of a factor `by` have one smoothing parameter each, and the argument
`id` makes them share one. The family `gaussian()` of `mgcv` has a constant
$\sigma$, so we compare the two packages on the model with a constant
$\sigma$:

```{r}
#| label: chick-mgcv
fit_cs0 <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8), by = Diet),
                   distrib = gaussian1_distrib(), data = ChickWeight)
g_cs0 <- mgcv::gam(weight ~ Diet + s(Time, bs = "bs", k = 8, by = Diet, id = 1),
                   data = ChickWeight, method = "REML")
c(statmod = edf_mu(fit_cs0), mgcv = sum(g_cs0$edf))
max(abs(fitted(fit_cs0) - fitted(g_cs0)))
```

@fig-chick shows the four curves of the model with the shared smoothing
parameter and an equation for $\sigma$. On diet 3 the chicks grow fastest.

```{r}
#| label: fig-chick
#| fig-cap: "Weight of the chicks against time, with the fitted mean curve of each diet (shared smoothing parameter, sigma linear in time)."
#| echo: false
#| fig-height: 4
tg <- seq(from = 0, to = 21, length.out = 100)
chick_lines <- do.call(rbind, lapply(levels(ChickWeight$Diet), function(d) {
  nd <- data.frame(Time = tg, Diet = factor(d, levels = levels(ChickWeight$Diet)))
  data.frame(Time = tg, weight = predict(fit_cs, what = "mu", newdata = nd),
             curve = paste("diet", d))
}))
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point(colour = book_grey, size = 0.6) +
  geom_line(data = chick_lines, aes(colour = curve, linetype = curve),
            linewidth = 0.8) +
  scale_colour_manual(values = book_palette[1:4]) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash")) +
  labs(x = "day", y = "weight (g)")
```

### A coefficient that changes with a covariate

With a numeric variable $z$ as `by`, the term `s(x, by = z)` is $f(x)\,z$: the
coefficient of $z$ is a smooth function of $x$ [@hastie1993]. The term is not
centered, because the constant part of $f$ is the main effect of $z$, and its
first coefficient, `const`, has no penalty. We simulate 400 observations from
$$
y_i = \sin(2\pi x_i) + (1 + 2x_i^2)\, z_i + \varepsilon_i,
\qquad \varepsilon_i \sim \mathrm{N}(0, 0.5),
$$
with $x$ uniform on $(0, 1)$ and $z$ standard normal, and we fit the model
with a smooth of $x$ and a smooth of $x$ multiplied by $z$:

```{r}
#| label: fit-vc
set.seed(3)
vc <- data.frame(x = runif(n = 400), z = rnorm(n = 400))
vc$y <- sin(2 * pi * vc$x) + (1 + 2 * vc$x^2) * vc$z + rnorm(n = 400, sd = 0.5)
fit_vc <- statmod(y ~ s(x, bspline_smooth(k = 10)) +
                    s(x, bspline_smooth(k = 10), by = z),
                  distrib = gaussian1_distrib(), data = vc)
names(coef(fit_vc)$mu)[c(1:3, 11:13)]
g_vc <- mgcv::gam(y ~ s(x, bs = "bs", k = 10) + s(x, bs = "bs", k = 10, by = z),
                  data = vc, method = "REML")
c(statmod = edf_mu(fit_vc), mgcv = sum(g_vc$edf))
max(abs(fitted(fit_vc) - fitted(g_vc)))
```

The coefficients of the second term have the prefix `s(x):z`, the name that
`mgcv` gives the term too. The fitted coefficient of $z$ at a value of $x$ is
the difference between the predictions at $z = 1$ and at $z = 0$:

```{r}
#| label: vc-coef
xv <- data.frame(x = seq(from = min(vc$x), to = max(vc$x), length.out = 100), z = 1)
xv0 <- xv
xv0$z <- 0
beta_vc <- predict(fit_vc, what = "mu", newdata = xv) -
  predict(fit_vc, what = "mu", newdata = xv0)
```

@fig-vc shows the data against $x$, and the fitted coefficient of $z$ against
the true one.

```{r}
#| label: fig-vc
#| fig-cap: "Left: the simulated response against x. Right: the fitted coefficient of z as a function of x, and the true coefficient."
#| echo: false
#| fig-height: 3.8
#| fig-width: 9
vc_panels <- c("y", "coefficient of z")
vc_data <- data.frame(x = vc$x, value = vc$y,
                      panel = factor(vc_panels[1], levels = vc_panels))
vc_lines <- data.frame(
  x = rep(xv$x, times = 2),
  value = c(beta_vc, 1 + 2 * xv$x^2),
  curve = factor(rep(1:2, each = nrow(xv)), labels = c("fitted", "true")),
  panel = factor(vc_panels[2], levels = vc_panels))
vc_range <- data.frame(x = mean(xv$x), value = c(0.5, 3.3),
                       panel = factor(vc_panels[2], levels = vc_panels))
ggplot(vc_lines, aes(x = x, y = value)) +
  geom_point(data = vc_data, colour = book_grey, size = 0.8) +
  geom_blank(data = vc_range) +
  geom_line(aes(colour = curve, linetype = curve), linewidth = 0.8) +
  facet_wrap(facets = vars(panel), scales = "free_y", strip.position = "left") +
  scale_colour_manual(values = c(chalk, ink)) +
  scale_linetype_manual(values = c("solid", "dotted")) +
  labs(x = "x", y = NULL) +
  theme(strip.placement = "outside", strip.background = element_blank(),
        strip.text = element_text(size = rel(1)))
```

## Smooths of two covariates {#sec-gam-te}

The data frame `topo` of the `MASS` package gives the elevation, in feet, of
52 points of a piece of land [@davis1973; @venables2002]. The coordinates `x`
and `y` are in units of 50 feet. The term `te(x, y)` builds a smooth function
of the two coordinates as a tensor product: it takes a basis for $x$ and a
basis for $y$, and it uses all the products of one function from each basis
[@wood2006]. Its argument `smooths` gives the smoother of each margin,
`bspline_smooth(k = 5)` by default. The penalty is the sum of one penalty for
each direction,
$$
\lambda_1\, b^\top (P_x \otimes I)\, b + \lambda_2\, b^\top (I \otimes P_y)\, b ,
$$
where $P_x$ and $P_y$ are the penalties of the margins. With
`anisotropic = TRUE`, the default, each direction has its own smoothing
parameter, so the surface can be smoother along one direction than along the
other. With `anisotropic = FALSE` the two share one. We fit both, and a model
with a separate smooth for each coordinate:

```{r}
#| label: fit-topo
data(topo, package = "MASS")
fit_te <- statmod(z ~ te(x, y), distrib = gaussian1_distrib(), data = topo)
fit_ti <- statmod(z ~ te(x, y, anisotropic = FALSE),
                  distrib = gaussian1_distrib(), data = topo)
fit_ta <- statmod(z ~ s(x, bspline_smooth(k = 5)) + s(y, bspline_smooth(k = 5)),
                  distrib = gaussian1_distrib(), data = topo)
hyper(fit_te)[, c("name", "estimate")]
data.frame(edf = c(edf_mu(fit_te), edf_mu(fit_ti), edf_mu(fit_ta)),
           sigma = c(exp(coef(fit_te)$sigma[[1]]), exp(coef(fit_ti)$sigma[[1]]),
                     exp(coef(fit_ta)$sigma[[1]])),
           AIC = c(AIC(fit_te), AIC(fit_ti), AIC(fit_ta)),
           row.names = c("anisotropic", "isotropic", "additive"))
```

The additive model has a much larger AIC: the elevation is not a sum of a
function of $x$ and a function of $y$. The two tensor products are close.
@fig-topo shows the fitted surface of the anisotropic model as a map.

```{r}
#| label: fig-topo
#| fig-cap: "The fitted elevation of the `topo` data, in feet. The points are the 52 measured locations."
#| echo: false
#| fig-height: 5
gx <- seq(from = min(topo$x), to = max(topo$x), length.out = 60)
gy <- seq(from = min(topo$y), to = max(topo$y), length.out = 60)
gt <- expand.grid(x = gx, y = gy)
zt <- matrix(predict(fit_te, what = "mu", newdata = gt), nrow = length(gx))
topo_grid <- data.frame(gt, z = as.vector(zt))
ggplot(topo_grid, aes(x = x, y = y)) +
  geom_raster(aes(fill = z)) +
  geom_contour(aes(z = z), breaks = pretty(range(zt), n = 10),
               colour = "grey25", linewidth = 0.3) +
  geom_point(data = topo, colour = rust, size = 1.1) +
  scale_fill_gradientn(colours = hcl.colors(30, "Greens 3", rev = TRUE)) +
  coord_fixed(expand = FALSE, clip = "off") +
  labs(x = "x (50 ft)", y = "y (50 ft)")
```

### The same model in `mgcv`

The function `te()` of `mgcv` builds the tensor product from the same
margins with `bs = "bs"`, and the two surfaces are not the same:

```{r}
#| label: topo-mgcv
g_te <- mgcv::gam(z ~ te(x, y, bs = "bs", k = 5, np = FALSE), data = topo,
                  method = "REML")
c(statmod = edf_mu(fit_te), mgcv = sum(g_te$edf))
max(abs(fitted(fit_te) - fitted(g_te)))
```

The two packages use B-spline margins with the same functions and the same
penalties, and the tensor products span the same functions. The difference is
the identity matrix $I$ in the penalty above. The penalty $P_x$ is an
integral, so it does not depend on how the basis of $x$ is written, but
$P_x \otimes I$ also penalizes each coefficient of the $y$ basis with weight
one, and this weight depends on which basis of $y$ we use. The B-splines of
`basis7` have knots that repeat at the ends of the range, while those of
`mgcv` have knots equally spaced beyond the range. The two bases span the
same functions with different coefficients, so the two penalties are
different and give different surfaces. The argument `np = TRUE`, the default
of `mgcv`, writes each margin in yet another basis. Neither choice is more
correct than the other.

## Smooths in the other parameters {#sec-gam-sigma}

Every parameter of the distribution can have smooth terms, as `sigma` had in
@sec-model-sigma. The temperatures in Chicago vary more in winter than in
summer. We give `sigma` a cyclic smooth of the day of the year:

```{r}
#| label: fit-chic-sigma
fit_ts <- statmod(temp ~ s(doy, cyclic_smooth(k = 10, lower = 0, upper = 366)) |
                    sigma ~ s(doy, cyclic_smooth(k = 8, lower = 0, upper = 366)),
                  distrib = gaussian1_distrib(), data = chic)
c(constant_sigma = AIC(fit_tc), smooth_sigma = AIC(fit_ts))
days <- data.frame(doy = c(15, 105, 196, 288))
predict(fit_ts, what = "sigma", newdata = days)
```

The standard deviation is `r fmt(predict(fit_ts, what = "sigma", newdata = days)[1], 1)`
degrees in mid January and `r fmt(predict(fit_ts, what = "sigma", newdata = days)[3], 1)`
in mid July. The family `gaulss()` of `mgcv` fits a gaussian model with an
equation for the standard deviation. Its second equation is for
$\log(1/\sigma - 0.01)$, and the prediction on the scale of the response gives
$1/\sigma$:

```{r}
#| label: chic-gaulss
g_ts <- mgcv::gam(list(temp ~ s(doy, bs = "cp", k = 10), ~ s(doy, bs = "cp", k = 8)),
                  family = mgcv::gaulss(), data = chic,
                  knots = list(doy = c(0, 366)))
1 / predict(g_ts, newdata = days, type = "response")[, 2]
```

The smooths differ (the cyclic P-spline of `mgcv` has a penalty on the
differences of the coefficients), and the fitted standard
deviations agree to about 0.05 degrees. @fig-chic-sigma shows the band of two
standard deviations around the mean.

```{r}
#| label: fig-chic-sigma
#| fig-cap: "Mean daily temperature in Chicago against the day of the year, with the fitted mean and the band of plus and minus two fitted standard deviations."
#| echo: false
#| fig-height: 3.8
ms <- predict(fit_ts, what = "mu", newdata = dg)
ss <- predict(fit_ts, what = "sigma", newdata = dg)
sig_lines <- data.frame(
  doy = rep(dg$doy, times = 3),
  temp = c(ms, ms + 2 * ss, ms - 2 * ss),
  curve = factor(rep(c(1, 2, 2), each = nrow(dg)),
                 labels = c("mean", "mean plus and minus 2 sigma")),
  line = rep(1:3, each = nrow(dg)))
ggplot(chic, aes(x = doy, y = temp)) +
  geom_point(colour = book_grey, size = 0.4) +
  geom_line(data = sig_lines,
            aes(colour = curve, linetype = curve, group = line),
            linewidth = 0.8) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "day of the year", y = "temperature (C)")
```

## Smooths with other distributions {#sec-gam-counts}

The number of deaths in a day is a count. A common model of the daily deaths
has a smooth function of time for the long-term trend and the seasons, a
smooth function of the temperature, and the day of the week [@peng2008]. We
fit it with a Poisson response to the six years of data, and we compare it
with `mgcv`:

```{r}
#| label: fit-deaths
fit_dp <- statmod(death ~ s(time, bspline_smooth(k = 12)) +
                    s(temp, bspline_smooth(k = 10)) + dow,
                  distrib = poisson_distrib(), data = chic)
g_dp <- mgcv::gam(death ~ s(time, bs = "bs", k = 12) + s(temp, bs = "bs", k = 10) +
                    dow, family = poisson, data = chic, method = "REML")
c(statmod = edf_mu(fit_dp), mgcv = sum(g_dp$edf))
max(abs(fitted(fit_dp) - fitted(g_dp)))
```

The ratio of the Pearson statistic to the residual degrees of freedom
measures the overdispersion (@sec-glm-overdispersion). We compute it and fit
the negative binomial model:

```{r}
#| label: deaths-nb
mu_dp <- fitted(fit_dp)
sum((chic$death - mu_dp)^2 / mu_dp) / (nrow(chic) - edf_mu(fit_dp))
fit_dn <- statmod(death ~ s(time, bspline_smooth(k = 12)) +
                    s(temp, bspline_smooth(k = 10)) + dow,
                  distrib = negbin2_distrib(), data = chic)
g_dn <- mgcv::gam(death ~ s(time, bs = "bs", k = 12) + s(temp, bs = "bs", k = 10) +
                    dow, family = mgcv::nb(), data = chic, method = "REML")
c(statmod = exp(coef(fit_dn)$theta[[1]]), mgcv = g_dn$family$getTheta(TRUE))
c(Poisson = AIC(fit_dp), negbin2 = AIC(fit_dn))
```

The ratio is `r fmt(sum((chic$death - mu_dp)^2 / mu_dp) / (nrow(chic) - edf_mu(fit_dp)))`,
so the deaths vary more than a Poisson distribution allows. The dispersion
$\theta$ of the negative binomial model is the same in the two packages, and
the negative binomial model has the smaller AIC. @fig-deaths shows the fitted
ratio of the death rate at each temperature to the rate at 20 degrees, with
the other covariates held fixed.

```{r}
#| label: fig-deaths
#| fig-cap: "Ratio of the daily death rate at each temperature to the rate at 20 degrees Celsius, from the negative binomial model, with a 95% pointwise interval."
#| echo: false
#| fig-height: 3.8
tt <- data.frame(temp = seq(from = -20, to = 32, length.out = 100),
                 time = median(chic$time), dow = "Monday")
ref <- data.frame(temp = 20, time = median(chic$time), dow = "Monday")
pe <- predict(fit_dn, what = "link:mu", newdata = tt, se = TRUE)
e0 <- predict(fit_dn, what = "link:mu", newdata = ref)
deaths_curve <- data.frame(temp = tt$temp, ratio = exp(pe$fit - e0),
                           lower = exp(pe$lower - e0),
                           upper = exp(pe$upper - e0))
ggplot(deaths_curve, aes(x = temp)) +
  geom_hline(yintercept = 1, linetype = "dotted", colour = book_grey) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = "95% interval"),
              alpha = 0.2) +
  geom_line(aes(y = ratio, colour = "rate ratio"), linewidth = 0.8) +
  scale_colour_manual(values = chalk) +
  scale_fill_manual(values = chalk) +
  guides(colour = guide_legend(order = 1), fill = guide_legend(order = 2)) +
  labs(x = "temperature (C)", y = "rate ratio")
```

## A lasso on the coefficients of a smooth {#sec-gam-lasso}

The argument `penalty` of a smoother replaces the penalty on the roughness with
another penalty on the same coefficients. It takes a function of the number
of coefficients that returns a penalty of the `penalties7` package, such as
`penalties7::lasso_penalty`. The coefficients of the smooth are written in
the Demmler-Reinsch basis [@demmler1975], in which the roughness penalty is
diagonal and the coordinates are ordered from the smoothest function to the
most wiggly one. A lasso on these coordinates sets some of them to zero, and
so chooses how many directions of wiggliness the curve keeps. The lasso does
not act on the free coefficient `lin`. We fit it to the `mcycle` data:

```{r}
#| label: fit-lasso
fit_ls <- statmod(accel ~ s(times, bspline_smooth(k = 20,
                                                 penalty = penalties7::lasso_penalty)),
                  distrib = gaussian1_distrib(), data = mcycle)
coef(fit_ls)$mu
hyper(fit_ls)[, c("name", "estimate", "source")]
```

The lasso keeps `r sum(coef(fit_ls)$mu[-(1:2)] != 0)` of the 18 penalized
coefficients. A lasso penalty has a kink at zero, so REML does not apply, and
`statmod()` chooses $\lambda$ by BIC on a path of values. We compare the fit
with the smooth penalized on the second derivative:

```{r}
#| label: lasso-compare
fit_r20 <- statmod(accel ~ s(times, bspline_smooth(k = 20)),
                   distrib = gaussian1_distrib(), data = mcycle)
data.frame(edf = c(edf_mu(fit_r20), edf_mu(fit_ls)),
           AIC = c(AIC(fit_r20), AIC(fit_ls)),
           row.names = c("roughness", "lasso"))
statmod_certificate(fit_ls)$state
```

The effective degrees of freedom of the lasso are the number of coefficients
that it keeps, plus the intercept and `lin` [@zou2007]. On these data the
lasso smooth has the smaller AIC. The certificate is `unknown`: it reads the
derivative of the criterion in the smoothing parameter, and the BIC along the
path of a penalty with a kink has no derivative.
@fig-lasso shows the two curves.

```{r}
#| label: fig-lasso
#| fig-cap: "Head acceleration against time, with the smooth penalized on the second derivative and the smooth with a lasso penalty on its Demmler-Reinsch coefficients."
#| echo: false
#| fig-height: 4
lasso_lines <- data.frame(
  times = rep(mg$times, times = 2),
  accel = c(predict(fit_r20, what = "mu", newdata = mg),
            predict(fit_ls, what = "mu", newdata = mg)),
  curve = factor(rep(1:2, each = nrow(mg)),
                 labels = c("roughness penalty", "lasso")))
ggplot(mcycle, aes(x = times, y = accel)) +
  geom_point(colour = book_grey, size = 1) +
  geom_line(data = lasso_lines, aes(colour = curve, linetype = curve),
            linewidth = 0.8) +
  scale_colour_manual(values = c(rust, chalk)) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = "time after impact (ms)", y = "acceleration (g)")
```

```{r}
#| label: gate-06
#| include: false
assert_gam_ok(
  fits = list(b10 = fit_b10, b20 = fit_b20, b40 = fit_b40, o1 = fit_o1,
              o3 = fit_o3, sk = fit_sk, ss = fit_ss, ps = fit_ps, tb = fit_tb,
              tc = fit_tc, tf = fit_tf, tf2 = fit_tf2, tf3 = fit_tf3,
              lg = fit_lg, d20 = fit_d20, h20 = fit_h20, d60 = fit_d60,
              h60 = fit_h60, w2 = fit_w2, we = fit_we,
              ad = fit_ad, p40 = fit_p40, c1 = fit_c1, c2 = fit_c2,
              c3 = fit_c3, cs = fit_cs, cl = fit_cl, cs0 = fit_cs0,
              vc = fit_vc, te = fit_te, ti = fit_ti, ta = fit_ta,
              ts = fit_ts, dp = fit_dp, dn = fit_dn, ls = fit_ls,
              r20 = fit_r20),
  ref = list(b20 = g_b20, o1 = g_o1, o3 = g_o3, ps = g_ps, cc = g_cc,
             ad = g_ad, cs0 = g_cs0, vc = g_vc, te = g_te, ts = g_ts,
             dp = g_dp, dn = g_dn, nls = fit_nls),
  data = list(cw = ChickWeight, chic = chic),
  claims = list(op_co2 = op_co2, beta_vc = beta_vc, xv = xv$x))
```
