# Generalized linear mixed models {#sec-glmm}

```{r}
#| label: setup-08
#| include: false
source("R/_setup.R")
source("R/glmm-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"; ink <- "grey15"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

A generalized linear mixed model adds random effects to the linear predictor
of a generalized linear model [@breslow1993]. Given the random effects $b_i$
of group $i$, the observations of the group are independent, with
$$
y_{ij} \mid b_i \sim F\{\, h^{-1}(\eta_{ij})\,\}, \qquad
\eta_{ij} = x_{ij}^\top\beta + z_{ij}^\top b_i, \qquad
b_i \sim \mathrm{N}(0, \Sigma_b),
$$
where $F$ is a Poisson, Bernoulli, negative binomial or any other
distribution, and $h$ is the link of its parameter. The likelihood of the
fixed effects and of $\Sigma_b$ is an integral over the random effects of
each group:
$$
L(\beta, \Sigma_b) = \prod_i \int \prod_j f\{y_{ij} \mid
  h^{-1}(x_{ij}^\top\beta + z_{ij}^\top b)\}\, \phi(b; 0, \Sigma_b)\, db .
$$
For a gaussian response with the identity link this integral has a closed
form, which is why the models of @sec-lmm are exact. For any other
distribution it has none. `statmodels7` replaces it with the Laplace
approximation [@tierney1986]: the integrand is approximated by a gaussian
density centred at its mode $\hat b_i$, and
$$
\log \int e^{g(b)}\, db \approx g(\hat b) - \tfrac12 \log\det\{-g''(\hat b)\}
+ \tfrac{d}{2}\log(2\pi),
$$
where $g$ is the logarithm of the integrand and $d$ the dimension of $b$.
In `statmodels7` the random effects are coefficients with a penalty, so
$\hat b$ is the mode of the penalized likelihood, and the REML criterion
applies the same approximation to the fixed effects of the mean as well
(@sec-beyond-outer). The packages `glmmTMB` [@brooks2017] and `lme4`
[@bates2015] use the same approximation by default, and we compare with them
throughout the chapter.

The other parameters of the distribution can have random effects too: we
give random effects to the dispersion $\theta$ of a negative binomial and to
the probability of an extra zero in a zero-inflated model. The last section
shows what a link that is not the identity does to the predictions for a new
group.

## Counts with a random intercept {#sec-glmm-poisson}

We return to the `epil` data of @sec-beyond-random: the number of epileptic
seizures of 59 patients in four periods of two weeks, in a trial of the drug
progabide [@thall1990]. The covariates are the treatment `trt`, the centered
logarithm of the baseline count `lbase`, the centered logarithm of the age
`lage` and the indicator `V4` of the fourth period. Each patient has a random
intercept on the log scale [@breslow1993]:
$$
y_{ij} \mid b_i \sim \mathrm{Poisson}(\mu_{ij}), \qquad
\log\mu_{ij} = x_{ij}^\top\beta + b_i, \qquad b_i \sim \mathrm{N}(0, \tau).
$$

```{r}
#| label: fit-epil
data(epil, package = "MASS")
epil$subject <- factor(epil$subject)
fit_ep <- statmod(y ~ lbase * trt + lage + V4 + random(~ 1 | subject),
                  distrib = poisson_distrib(), data = epil)
summary(fit_ep)
```

A coefficient is a log rate ratio for one patient: given $b_i$, the fourth
period has $\exp(\hat\beta_{V4}) = `r fmt(exp(coef(fit_ep)$mu[["V4"]]), 3)`$
times the seizures of the other three. The standard deviation of the random
intercepts is $\hat\tau = `r fmt(hyper(fit_ep)$estimate, 3)`$, so two patients
with the same covariates can differ in their rate by a factor of
$\exp(2 \times 1.96\,\hat\tau) = `r fmt(exp(2 * 1.96 * hyper(fit_ep)$estimate), 1)`$.

With `REML = TRUE`, `glmmTMB()` integrates the fixed effects as well, and it
gives the same fit:

```{r}
#| label: tmb-epil
tmb_ep <- glmmTMB::glmmTMB(y ~ lbase * trt + lage + V4 + (1 | subject),
                           family = poisson, data = epil, REML = TRUE)
glmmTMB::fixef(tmb_ep)$cond
glmmTMB::VarCorr(tmb_ep)
logLik(tmb_ep)
logLik(fit_ep, type = "marginal")
```

### Maximum likelihood

The functions `glmer()` and `glmmTMB()` estimate by maximum likelihood by
default. With `ml()` the fixed effects of the mean are held at the joint mode
of the fixed and random effects, and the criterion integrates the random
effects alone (@sec-beyond-outer). With `ml(marginal = "all")` the fixed
effects are also chosen to maximize the Laplace approximation of
$L(\beta, \tau)$, which is what `glmer()` does:

```{r}
#| label: ml-epil
fit_ep_ml <- statmod(y ~ lbase * trt + lage + V4 + random(~ 1 | subject),
                     distrib = poisson_distrib(), data = epil,
                     outer_criterion = ml())
fit_ep_all <- statmod(y ~ lbase * trt + lage + V4 + random(~ 1 | subject),
                      distrib = poisson_distrib(), data = epil,
                      outer_criterion = ml(marginal = "all"))
glmer_ep <- lme4::glmer(y ~ lbase * trt + lage + V4 + (1 | subject),
                        family = poisson, data = epil)
data.frame(
  fit = c("ml()", "ml(marginal = \"all\")", "glmer"),
  tau = c(hyper(fit_ep_ml)$estimate, hyper(fit_ep_all)$estimate,
          attr(lme4::VarCorr(glmer_ep)$subject, "stddev")),
  trt = c(coef(fit_ep_ml)$mu[["trtprogabide"]],
          coef(fit_ep_all)$mu[["trtprogabide"]],
          lme4::fixef(glmer_ep)[["trtprogabide"]]),
  logLik = c(logLik(fit_ep_ml, type = "marginal"),
             logLik(fit_ep_all, type = "marginal"), logLik(glmer_ep)))
```

The second row, `ml(marginal = "all")`, agrees with `glmer()` in the third
row to four digits. The first row differs from both slightly: for a gaussian
response the joint mode and the maximum of $L$ give the same fixed effects, but for a Poisson response the determinant in the Laplace
approximation depends on $\beta$, so the two differ.

## Binary responses {#sec-glmm-bernoulli}

The data frame `bacteria` of the `MASS` package records the presence of the
bacterium *H. influenzae* in 50 children with otitis media, at weeks 0, 2,
4, 6 and 11 [@venables2002, sec. 10.4]. The children received a placebo, a
drug, or the drug with encouragement to take it (`drug+`). We code the
response as 0 and 1 and compare the visits after week 2 with the first two:
$$
y_{ij} \mid b_i \sim \mathrm{Bernoulli}(p_{ij}), \qquad
\mathrm{logit}\, p_{ij} = x_{ij}^\top\beta + b_i, \qquad
b_i \sim \mathrm{N}(0, \tau).
$$

```{r}
#| label: fit-bacteria
data(bacteria, package = "MASS")
bacteria$y <- as.integer(bacteria$y == "y")
bacteria$late <- as.integer(bacteria$week > 2)
fit_bc <- statmod(y ~ trt + late + random(~ 1 | ID),
                  distrib = bernoulli_distrib(), data = bacteria)
summary(fit_bc)
```

The coefficients are log odds ratios for one child. The coefficients of both
treatments are negative, and after week 2 the odds are
$\exp(\hat\beta_{\mathrm{late}}) = `r fmt(exp(coef(fit_bc)$mu[["late"]]), 3)`$
times those of the first two visits. The children differ a lot: the
standard deviation of the random intercepts is
$\hat\tau = `r fmt(hyper(fit_bc)$estimate, 3)`$ on the logit scale.

`glmer()` maximizes the Laplace approximation of the likelihood, and the fit
of `ml(marginal = "all")` is the same:

```{r}
#| label: glmer-bacteria
fit_bc_all <- statmod(y ~ trt + late + random(~ 1 | ID),
                      distrib = bernoulli_distrib(), data = bacteria,
                      outer_criterion = ml(marginal = "all"))
glmer_bc <- lme4::glmer(y ~ trt + late + (1 | ID), family = binomial,
                        data = bacteria)
coef(fit_bc_all)$mu[1:4]
lme4::fixef(glmer_bc)
c(statmod = hyper(fit_bc_all)$estimate,
  glmer = attr(lme4::VarCorr(glmer_bc)$ID, "stddev"))
logLik(fit_bc_all, type = "marginal")
logLik(glmer_bc)
```

### The accuracy of the Laplace approximation

Each child has at most five binary observations, and with so little
information on each $b_i$ the integrand of a group is far from gaussian. The
Laplace approximation then underestimates the variance of the random effects
[@breslow1995]. The function `glmer()` can compute the integral of each
group with adaptive Gauss-Hermite quadrature, which is exact as the number of
nodes grows. With 25 nodes:

```{r}
#| label: agq-bacteria
glmer_25 <- lme4::glmer(y ~ trt + late + (1 | ID), family = binomial,
                        data = bacteria, nAGQ = 25)
lme4::fixef(glmer_25)
attr(lme4::VarCorr(glmer_25)$ID, "stddev")
logLik(glmer_25)
```

The standard deviation from the quadrature is
`r fmt(attr(lme4::VarCorr(glmer_25)$ID, "stddev"), 3)`, against
`r fmt(hyper(fit_bc_all)$estimate, 3)` from the Laplace approximation, and the
fixed effects move by less than their standard errors. The conclusions do not
change here, but with fewer observations per group the difference grows. The
Laplace approximation is what `statmodels7` computes; the comparison with
`glmer(nAGQ = )` is a way to check it on a model with one random intercept.

## Negative binomial counts {#sec-glmm-negbin}

The data frame `Owls` of the `glmmTMB` package records the begging of barn
owl nestlings in 27 nests [@roulin2007; @zuur2009]. The response
`SiblingNegotiation` is the number of calls the nestlings made in the 30
seconds before a parent arrived at the nest. The covariates are the food
treatment of the nest (`Deprived` or `Satiated`), the sex of the parent that
arrived, and the number of nestlings, which we enter as an offset so that
the model describes the calls per nestling. The counts vary much more than a
Poisson distribution allows, so we use the negative binomial distribution of
@sec-glm-overdispersion, with $\mathrm{Var}(y) = \mu + \mu^2/\theta$, and a
random intercept for each nest:
$$
\log\mu_{ij} = \log n_{ij} + x_{ij}^\top\beta + b_i, \qquad
b_i \sim \mathrm{N}(0, \tau).
$$

```{r}
#| label: fit-owls
data(Owls, package = "glmmTMB")
fit_ow1 <- statmod(SiblingNegotiation ~ FoodTreatment * SexParent +
                     offset(logBroodSize) + random(~ 1 | Nest),
                   distrib = negbin2_distrib(), data = Owls)
summary(fit_ow1)
```

`glmmTMB()` with the family `nbinom2` has the same parametrization:

```{r}
#| label: tmb-owls
tmb_ow1 <- glmmTMB::glmmTMB(SiblingNegotiation ~ FoodTreatment * SexParent +
                              offset(logBroodSize) + (1 | Nest),
                            family = glmmTMB::nbinom2, data = Owls,
                            REML = TRUE)
glmmTMB::fixef(tmb_ow1)$cond
glmmTMB::sigma(tmb_ow1)
exp(coef(fit_ow1)$theta[[1]])
logLik(tmb_ow1)
logLik(fit_ow1, type = "marginal")
```

The function `sigma()` of `glmmTMB` returns $\hat\theta$ for this family.

### Random effects on the dispersion

The equation of $\theta$ can have covariates and random effects of its own.
We let the dispersion depend on the food treatment and on the nest:
$$
\log\theta_{ij} = \gamma_0 + \gamma_1 s_{ij} + c_i, \qquad
c_i \sim \mathrm{N}(0, \kappa),
$$
where $s_{ij}$ is 1 for a satiated nest and $c_i$ is independent of $b_i$.

```{r}
#| label: fit-owls-theta
fit_owf <- statmod(SiblingNegotiation ~ FoodTreatment * SexParent +
                     offset(logBroodSize) + random(~ 1 | Nest) |
                     theta ~ FoodTreatment,
                   distrib = negbin2_distrib(), data = Owls)
fit_ow2 <- statmod(SiblingNegotiation ~ FoodTreatment * SexParent +
                     offset(logBroodSize) + random(~ 1 | Nest) |
                     theta ~ FoodTreatment + random(~ 1 | Nest),
                   distrib = negbin2_distrib(), data = Owls)
summary(fit_ow2)
```

In `glmmTMB()` the equation of $\theta$ is the argument `dispformula`:

```{r}
#| label: tmb-owls-theta
tmb_ow2 <- glmmTMB::glmmTMB(SiblingNegotiation ~ FoodTreatment * SexParent +
                              offset(logBroodSize) + (1 | Nest),
                            dispformula = ~ FoodTreatment + (1 | Nest),
                            family = glmmTMB::nbinom2, data = Owls,
                            REML = TRUE)
glmmTMB::fixef(tmb_ow2)
glmmTMB::VarCorr(tmb_ow2)
logLik(tmb_ow2)
logLik(fit_ow2, type = "marginal")
```

The three models have the same fixed effects in the mean, so we compare them
with their restricted log-likelihoods:

```{r}
#| label: lr-owls
l1 <- as.numeric(logLik(fit_ow1, type = "marginal"))
lf <- as.numeric(logLik(fit_owf, type = "marginal"))
l2 <- as.numeric(logLik(fit_ow2, type = "marginal"))
lr_food <- 2 * (lf - l1)
lr_nest <- 2 * (l2 - lf)
c(lr_food = lr_food, lr_nest = lr_nest)
pchisq(lr_food, df = 1, lower.tail = FALSE)
```

The food treatment changes the dispersion: in a satiated nest
$\hat\theta$ is $\exp(\hat\gamma_1) = `r fmt(exp(coef(fit_ow2)$theta[["FoodTreatmentSatiated"]]), 3)`$
times that of a deprived nest, so the counts vary more. The nests differ in
their dispersion too. Under the smaller model the variance $\kappa^2$ is
zero, on the boundary of its range, so `lr_nest` does not have the $\chi^2_1$
distribution [@self1987]; its p-value is at most that of $\chi^2_1$, which is
`r format(signif(pchisq(lr_nest, df = 1, lower.tail = FALSE), 2))`. With the
random effects on $\theta$ the standard deviation of the nests' effects on
the mean falls from `r fmt(hyper(fit_ow1)$estimate, 3)` to
`r fmt(hyper(fit_ow2)$estimate[1], 3)`: part of what looked like a difference
in the mean calls of the nests is a difference in how much their calls vary.

## Zero inflation {#sec-glmm-zi}

The data frame `Salamanders` of the `glmmTMB` package gives the number of
salamanders found in 23 streams of the Appalachian mountains, some of them
affected by coal mining [@price2016]. Each stream was sampled four times, and
the counts are given for seven species and life stages (`spp`). Many counts
are zero, and some zeros come from a species that is absent from a stream.
The zero-inflated distribution of @sec-distribs-zeros describes this: with
probability $\pi$ the count is a structural zero, and otherwise it is a
Poisson count with mean $\mu$,
$$
P(y = 0) = \pi + (1 - \pi)e^{-\mu}, \qquad
P(y = k) = (1 - \pi)\frac{\mu^k e^{-\mu}}{k!}, \quad k \ge 1
$$
[@lambert1992]. We give each stream a random effect in $\log\mu$ and one in
$\mathrm{logit}\,\pi$:
$$
\log\mu_{ij} = x_{ij}^\top\beta + b_i, \qquad
\mathrm{logit}\,\pi_{ij} = w_{ij}^\top\alpha + c_i ,
$$
with $b_i \sim \mathrm{N}(0, \tau)$ and $c_i \sim \mathrm{N}(0, \kappa)$.

```{r}
#| label: fit-salamanders
data(Salamanders, package = "glmmTMB")
Salamanders$site <- factor(as.character(Salamanders$site))
fit_zi <- statmod(count ~ spp + mined + random(~ 1 | site) |
                    zi ~ mined + random(~ 1 | site),
                  distrib = zero_inflated(poisson_distrib()), data = Salamanders)
summary(fit_zi)
```

In a stream affected by mining (`mined` is `yes`, the reference level) the
probability of a structural zero for a typical stream is
$\mathrm{logit}^{-1}(\hat\alpha_0) = `r fmt(plogis(coef(fit_zi)$zi[[1]]), 3)`$,
and in an unaffected stream it is
$\mathrm{logit}^{-1}(\hat\alpha_0 + \hat\alpha_1) =
`r fmt(plogis(sum(coef(fit_zi)$zi)), 3)`$. Where the species are present,
the unaffected streams also have
$\exp(\hat\beta_{\mathrm{minedno}}) = `r fmt(exp(coef(fit_zi)$mu[["minedno"]]), 2)`$
times the salamanders. In `glmmTMB()` the equation of $\pi$ is the argument
`ziformula`:

```{r}
#| label: tmb-salamanders
tmb_zi <- glmmTMB::glmmTMB(count ~ spp + mined + (1 | site),
                           ziformula = ~ mined + (1 | site),
                           family = poisson, data = Salamanders, REML = TRUE)
glmmTMB::fixef(tmb_zi)
glmmTMB::VarCorr(tmb_zi)
logLik(tmb_zi)
logLik(fit_zi, type = "marginal")
```

### A negative binomial count

The Poisson part may be too narrow. We replace it with a negative binomial and
keep the random effect on $\mu$. The equation of $\pi$ keeps the mining
covariate, but we drop its random effect; the next subsection shows why. To
compare the two distributions of the count, we also fit the Poisson model
with the same equation for $\pi$:

```{r}
#| label: fit-zinb
fit_zinb <- statmod(count ~ spp + mined + random(~ 1 | site) | zi ~ mined,
                    distrib = zero_inflated(negbin2_distrib()), data = Salamanders)
summary(fit_zinb)
tmb_zinb <- glmmTMB::glmmTMB(count ~ spp + mined + (1 | site),
                             ziformula = ~ mined, family = glmmTMB::nbinom2,
                             data = Salamanders, REML = TRUE)
logLik(tmb_zinb)
logLik(fit_zinb, type = "marginal")
fit_zip2 <- statmod(count ~ spp + mined + random(~ 1 | site) | zi ~ mined,
                    distrib = zero_inflated(poisson_distrib()), data = Salamanders)
c(zip = AIC(fit_zip2), zinb = AIC(fit_zinb))
```

The two models in the last line have the same equation for $\pi$, so the
comparison concerns only the distribution of the count. The negative binomial
has the smaller conditional AIC. With it, the
probability of a structural zero in an unaffected stream goes to zero:
$\hat\alpha_0 + \hat\alpha_1 = `r fmt(sum(coef(fit_zinb)$zi), 2)`$, a
probability of `r fmt(plogis(sum(coef(fit_zinb)$zi)), 4)`, with a standard
error of `r fmt(sqrt(diag(vcov(fit_zinb)))[["zi:minedno"]], 1)` for
$\hat\alpha_1$. On the logit scale the probability approaches zero only as
the coefficient goes to $-\infty$, so the likelihood is flat in that
direction and the standard error is large. `glmmTMB()` gives the same
estimate. The negative binomial explains the zeros of the unaffected streams
by its larger variance, and the extra zeros are needed only in the affected
streams.

### A random effect on the zero inflation

In the Poisson model the probability $\pi$ had a random effect on the stream.
We can ask for the same random effect in the negative binomial model. The fit
takes more than a minute, and during it the sparse Cholesky factorization
reports many times that a matrix is not positive definite; we hide these
warnings here:

```{r}
#| label: fit-zinb-re
#| warning: false
fit_zinb_re <- statmod(count ~ spp + mined + random(~ 1 | site) |
                         zi ~ mined + random(~ 1 | site),
                       distrib = zero_inflated(negbin2_distrib()),
                       data = Salamanders)
summary(fit_zinb_re)
statmod_certificate(fit_zinb_re)$state
```

The certificate is `unknown`. The curvature of the REML criterion at the
reported point is not negative definite (the `min eigen` of the `outer` line
of the summary is negative), so the fit cannot confirm that the two standard
deviations sit at a maximum of the criterion, and the summary gives them no
standard error. `glmmTMB()` has
the same difficulty with this model:

```{r}
#| label: tmb-zinb-re
tmb_zinb_re <- glmmTMB::glmmTMB(count ~ spp + mined + (1 | site),
                                ziformula = ~ mined + (1 | site),
                                family = glmmTMB::nbinom2,
                                data = Salamanders, REML = TRUE)
glmmTMB::fixef(tmb_zinb_re)$zi
glmmTMB::VarCorr(tmb_zinb_re)
logLik(tmb_zinb_re)
```

It reports a Hessian that is not positive definite and a false convergence,
and it returns no log-likelihood. The two packages reach similar estimates of
the standard deviation of the random effect on $\pi$
(`r fmt(hyper(fit_zinb_re)$estimate[2], 3)` against
`r fmt(attr(glmmTMB::VarCorr(tmb_zinb_re)$zi[[1]], "stddev"), 3)`), but
different estimates of the intercept $\hat\alpha_0$
(`r fmt(coef(fit_zinb_re)$zi[[1]], 2)` against
`r fmt(glmmTMB::fixef(tmb_zinb_re)$zi[[1]], 2)`) and of the slope
$\hat\alpha_1$ (`r fmt(coef(fit_zinb_re)$zi[[2]], 2)` against
`r fmt(glmmTMB::fixef(tmb_zinb_re)$zi[[2]], 2)`). Neither package can
confirm its estimates. The cause is the one of the previous subsection. In the unaffected
streams $\pi$ is already close to zero, $\hat\alpha_1$ moves towards
$-\infty$, and the likelihood is flat in that direction. A random effect on
$\mathrm{logit}\,\pi$ adds a second direction that the data do not determine
well. For this reason we keep the negative binomial model without it.

## Predictions for a new group {#sec-glmm-predict}

@sec-lmm-predict introduced the arguments `random`, `interval` and
`predictive` of `predict()`. With a link that is not the identity two things
change: the typical group and the average group have different predictions,
and the intervals are not symmetric around the prediction. We write $h$ for
the link of the parameter, $\eta_0 = x^\top\hat\beta$ for the linear
predictor of a typical group, $\mathrm{se}_0$ for its standard error, and
$v = z^\top\Sigma_b z$ for the variance that the random effects of a new
group add to the predictor.

**Two predictions of a parameter.** With $u = z^\top b \sim \mathrm{N}(0, \sqrt{v})$:

| `random` | prediction | log link | logit link |
|---|---|---|---|
| `"zero"` | $h^{-1}(\eta_0)$, the group with $b = 0$ | $e^{\eta_0}$ | $\mathrm{logit}^{-1}(\eta_0)$ |
| `"marginal"` | $\bar\theta = \mathbb{E}_u\{h^{-1}(\eta_0 + u)\}$, the average over the groups | $e^{\eta_0 + v/2}$ | Gauss-Hermite quadrature |

The log column is the moment generating function of $u$ at $t = 1$, as for
$\sigma$ in @sec-lmm-predict. For the logit link the integral has no closed
form. A useful approximation comes from approximating the logistic
distribution function with a normal one [@zeger1988]:
$$
\bar p \approx \mathrm{logit}^{-1}\!\left(\frac{\eta_0}{\sqrt{1 + c^2 v}}\right),
\qquad c = \frac{16\sqrt3}{15\pi} .
$$
The average probability is therefore closer to one half than the probability
of the typical group.

**Three intervals.** The predictor of a new group is
$\eta^* = \eta_0 + e + u$, where $e \sim \mathrm{N}(0, \mathrm{se}_0)$ is
the estimation error of $\eta_0$. Its variance is $s^2 = \mathrm{se}_0^2 + v$.

| `interval` | the interval covers | ends | `se` |
|---|---|---|---|
| `"confidence"` | $h^{-1}(x^\top\beta)$, the parameter of a typical group | $h^{-1}(\eta_0 \pm 1.96\,\mathrm{se}_0)$ | delta method |
| `"group"` | $\theta^* = h^{-1}(\eta^*)$, the parameter of a new group | $h^{-1}(\eta_0 \pm 1.96\, s)$ | $\mathrm{sd}(\theta^*)$ |
| `"prediction"` | one new observation $y$ of a new group | quantiles of $G$, below | $\mathrm{sd}(y)$ |

The inverse link is monotone, so the quantiles of $\eta^*$ carried through it
are the quantiles of $\theta^*$, and the group interval is exact. Its
standard error is the standard deviation of $\theta^*$. For the log link
$\theta^*$ is lognormal, and
$$
\mathrm{sd}(\theta^*) = e^{\eta_0 + s^2/2}\sqrt{e^{s^2} - 1} ;
$$
for any other link `predict()` computes the two moments of $\theta^*$ with
Gauss-Hermite quadrature. The prediction `fit` is $h^{-1}(\eta_0)$ under
`random = "zero"`, which is the median of $\theta^*$, and $\bar\theta$ under
`random = "marginal"`.

**One new observation.** Given the predictors, $y$ has the distribution $F$
of the family, so its distribution function averages $F$ over the
predictors:
$$
G(y) = \int F\{y \mid h^{-1}(\eta)\}\, \phi(\eta; \eta_0, s^2)\, d\eta .
$$
With random effects in several equations the integral is over all their
predictors. For a count the ends are integers:

| quantity | value |
|---|---|
| lower end | the smallest $y$ with $G(y) \ge 0.025$ |
| prediction | the smallest $y$ with $G(y) \ge 0.5$ |
| upper end | the smallest $y$ with $G(y) \ge 0.975$ |

so the probability that the interval holds a new count is at least 95%. For
a binary response $y$ is Bernoulli with $P(y = 1) = \bar p$, and the
interval is always $[0, 1]$.

**The typical child and the average child.** We predict a child of each
treatment group of the `bacteria` data after week 2, and compute
$\bar p$ with `integrate()` and with the approximation:

```{r}
#| label: predict-bacteria
new_child <- data.frame(trt = factor(levels(bacteria$trt), levels(bacteria$trt)),
                        late = 1, ID = "new")
p_zero <- predict(fit_bc, what = "mu", newdata = new_child, random = "zero")
p_marg <- predict(fit_bc, what = "mu", newdata = new_child,
                  random = "marginal")
eta0 <- predict(fit_bc, what = "link:mu", newdata = new_child,
                random = "zero")
tau_bc <- hyper(fit_bc)$estimate
p_int <- vapply(eta0, function(e) {
  integrate(function(u) plogis(e + u) * dnorm(u, sd = tau_bc),
            lower = -Inf, upper = Inf)$value
}, numeric(1))
c16 <- 16 * sqrt(3) / (15 * pi)
p_zeger <- plogis(eta0 / sqrt(1 + c16^2 * tau_bc^2))
data.frame(trt = levels(bacteria$trt), zero = p_zero, marginal = p_marg,
           integrate = p_int, approximation = p_zeger)
```

The column `marginal` agrees with `integrate()`, and the approximation is
within `r fmt(max(abs(p_zeger - p_int)), 3)` of both.

**The log odds ratio of the average child.** The coefficient
$\hat\beta_{\mathrm{drug}} = `r fmt(coef(fit_bc)$mu[["trtdrug"]], 3)`$ is the
log odds ratio of the drug for one child. The log odds ratio between the
average probabilities of the two groups is smaller, by about the factor
$1/\sqrt{1 + c^2 v}$:

```{r}
#| label: attenuation
lor_cond <- coef(fit_bc)$mu[["trtdrug"]]
lor_marg <- qlogis(p_marg[2]) - qlogis(p_marg[1])
c(conditional = lor_cond, marginal = lor_marg, ratio = lor_marg / lor_cond,
  approximation = 1 / sqrt(1 + c16^2 * tau_bc^2))
```

| log odds ratio | compares | estimated by |
|---|---|---|
| conditional | the same child with and without the drug | the mixed model |
| marginal | the proportion of children with the bacterium in a treated and in an untreated population | a model without random effects [@zeger1988] |

@fig-attenuation shows the two curves.

```{r}
#| label: fig-attenuation
#| fig-cap: "Probability of the bacterium against the linear predictor of a typical child, for the typical child and averaged over the children. The points are the three treatment groups after week 2."
#| echo: false
#| fig-height: 3.8
eg <- seq(from = -4, to = 6, length.out = 101)
pm_curve <- vapply(eg, function(e) integrate(function(u)
  plogis(e + u) * dnorm(u, sd = tau_bc), lower = -Inf, upper = Inf)$value, 0)
att_lev <- c("typical child", "average over children",
             "treatment groups, typical", "treatment groups, average")
att_lines <- data.frame(
  eta = rep(eg, times = 2), p = c(plogis(eg), pm_curve),
  curve = factor(rep(att_lev[1:2], each = length(eg)), levels = att_lev))
att_points <- data.frame(
  eta = rep(eta0, times = 2), p = c(p_zero, p_marg),
  curve = factor(rep(att_lev[3:4], each = length(eta0)), levels = att_lev))
ggplot(att_lines, aes(x = eta, y = p, colour = curve)) +
  geom_line(aes(linetype = curve), linewidth = 0.9) +
  geom_point(data = att_points, aes(shape = curve), size = 2.2) +
  scale_colour_manual(values = c(chalk, rust, chalk, rust), limits = att_lev) +
  scale_linetype_manual(values = c("solid", "dashed", "blank", "blank"),
                        limits = att_lev) +
  scale_shape_manual(values = c(NA, NA, 16, 17), limits = att_lev) +
  scale_y_continuous(limits = c(0, 1)) +
  guides(colour = guide_legend(nrow = 2), linetype = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2)) +
  labs(x = expression("linear predictor " * eta[0]),
       y = "probability of the bacterium")
```

**The dispersion of a new nest.** We predict $\theta$ of a new nest of the
`Owls` data, deprived and satiated, with a male parent and four nestlings,
and compute the ends and the standard deviation of the group interval from
the formulas above. On the log scale the standard error is $s$:

```{r}
#| label: predict-owls-group
new_nest <- data.frame(
  FoodTreatment = factor(c("Deprived", "Satiated"),
                         levels(Owls$FoodTreatment)),
  SexParent = factor("Male", levels(Owls$SexParent)),
  logBroodSize = log(4), Nest = "new")
g_theta <- predict(fit_ow2, what = "theta", newdata = new_nest,
                   random = "zero", interval = "group")
g_theta
g_link <- predict(fit_ow2, what = "link:theta", newdata = new_nest,
                  random = "zero", interval = "group")
g_link
s_th <- g_link$se
cbind(lower = exp(g_link$fit - qnorm(0.975) * s_th),
      upper = exp(g_link$fit + qnorm(0.975) * s_th),
      sd = exp(g_link$fit + s_th^2 / 2) * sqrt(exp(s_th^2) - 1))
```

The interval is wide, because $\hat\kappa = `r fmt(hyper(fit_ow2)$estimate[2], 2)`$:
the dispersion of one new nest can be
`r fmt(exp(qnorm(0.975) * s_th[1]), 1)` times larger or smaller than that of
the typical nest. The standard deviation of $\theta^*$ is larger than the
fit, because the lognormal distribution has a long right tail. A first-order
approximation, $e^{\eta_0}\,s$, would give
`r fmt(exp(g_link$fit[1]) * s_th[1], 2)` for the deprived nest, against
`r fmt(g_theta$se[1], 2)`.

**One new count of a new nest.** We check the prediction interval by
simulation: a million new nests, each with its two random effects, and one
count from each. The estimates are held at their values
(`predictive = "plugin"`), so that the simulation and `predict()` describe
the same distribution:

```{r}
#| label: predict-owls-response
p_calls <- predict(fit_ow2, what = "response", newdata = new_nest, se = TRUE,
                   random = "zero", interval = "prediction",
                   predictive = "plugin")
p_calls
eta_mu <- predict(fit_ow2, what = "link:mu", newdata = new_nest,
                  random = "zero")
eta_th <- predict(fit_ow2, what = "link:theta", newdata = new_nest,
                  random = "zero")
tau_ow <- hyper(fit_ow2)$estimate[1]
kappa_ow <- hyper(fit_ow2)$estimate[2]
set.seed(1)
n_sim <- 1e6
b_sim <- rnorm(n_sim, mean = 0, sd = tau_ow)
c_sim <- rnorm(n_sim, mean = 0, sd = kappa_ow)
y_sim <- rnbinom(n_sim, mu = exp(eta_mu[1] + b_sim),
                 size = exp(eta_th[1] + c_sim))
quantile(y_sim, probs = c(0.025, 0.5, 0.975), type = 1)
c(G_below_upper = mean(y_sim <= p_calls$upper[1] - 1),
  G_at_upper = mean(y_sim <= p_calls$upper[1]))
sd(y_sim)
```

The quantiles of the simulated counts are the ends of the interval for the
deprived nest, and the empirical distribution function crosses 0.975 at the
upper end. The standard deviation agrees to within half a per cent, which is
the Monte Carlo error of a count with such a long tail.

**A new stream of the salamanders.** In the zero-inflated Poisson model the
probability of a zero and the counts of a new stream depend on its two
random effects. We predict a new stream affected by mining and one that is
not, for the species `GP`, and check the unaffected one by simulation:

```{r}
#| label: predict-salamanders
new_site <- data.frame(spp = factor("GP", levels(Salamanders$spp)),
                       mined = factor(c("yes", "no"), levels(Salamanders$mined)),
                       site = "new")
p_sal <- predict(fit_zi, what = "response", newdata = new_site, se = TRUE,
                 random = "zero", interval = "prediction",
                 predictive = "plugin")
p_sal
eta_mu_s <- predict(fit_zi, what = "link:mu", newdata = new_site,
                    random = "zero")
eta_zi_s <- predict(fit_zi, what = "link:zi", newdata = new_site,
                    random = "zero")
tau_s <- hyper(fit_zi)$estimate[1]
kappa_s <- hyper(fit_zi)$estimate[2]
set.seed(2)
zero_sim <- runif(n_sim) < plogis(eta_zi_s[2] + rnorm(n_sim, sd = kappa_s))
y_sal <- ifelse(zero_sim, 0,
                rpois(n_sim, exp(eta_mu_s[2] + rnorm(n_sim, sd = tau_s))))
quantile(y_sal, probs = c(0.025, 0.5, 0.975), type = 1)
mean(y_sal == 0)
predict(fit_zi, what = "mean", newdata = new_site, random = "marginal")
mean(y_sal)
```

For an unaffected stream the probability of no salamanders of this species
in one sample is `r fmt(mean(y_sal == 0), 2)`, and the mean count over new
streams is the value of `random = "marginal"` for `what = "mean"`.

```{r}
#| label: gate-08
#| include: false
assert_glmm_ok(
  fits = list(ep = fit_ep, ep_ml = fit_ep_ml, ep_all = fit_ep_all,
              bc = fit_bc, bc_all = fit_bc_all, ow1 = fit_ow1, owf = fit_owf,
              ow2 = fit_ow2, zi = fit_zi, zinb = fit_zinb, zip2 = fit_zip2),
  ref = list(ep = tmb_ep, ep_glmer = glmer_ep, bc = glmer_bc,
             bc_25 = glmer_25, ow1 = tmb_ow1, ow2 = tmb_ow2, zi = tmb_zi,
             zinb = tmb_zinb),
  data = list(epil = epil, bacteria = bacteria, Owls = Owls, Salamanders = Salamanders),
  claims = list(p_zero = p_zero, p_marg = p_marg, p_int = p_int,
                p_zeger = p_zeger, lor_cond = lor_cond, lor_marg = lor_marg,
                lr_food = lr_food, lr_nest = lr_nest, g_theta = g_theta,
                g_link = g_link, p_calls = p_calls, p_sal = p_sal,
                new_child = new_child, new_nest = new_nest,
                new_site = new_site, zinb_re = fit_zinb_re,
                tmb_zinb_re = tmb_zinb_re))
```
