# Generalized additive mixed models {#sec-gamm}

```{r}
#| label: setup-09
#| include: false
source("R/_setup.R")
source("R/gamm-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"; ink <- "grey15"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

A generalized additive mixed model has smooth functions and random effects in
the same predictor [@lin1999; @ruppert2003]:
$$
\eta_{ij} = x_{ij}^\top\beta + \sum_k f_k(t_{ij}) + z_{ij}^\top b_i ,
$$
where $i$ is the group and $j$ the observation. In `statmodels7` the two kinds
of term are built in the same way. The coefficients of a smooth have the
penalty $\lambda\,\beta^\top S\beta$ of @sec-gam, and the random effects have
the penalty $b^\top\Sigma_b^{-1}b$ of @sec-lmm. Both penalties are minus the
logarithm of a gaussian density, so the REML criterion of the previous
chapters integrates both kinds of coefficient and estimates $\lambda$ and
$\Sigma_b$ together. Nothing new is needed to fit the model: a formula with
`s()` and `random()` is enough.

Two other packages fit these models, and they meet the same idea from the two
opposite sides. The package `mgcv` writes a random intercept as a smooth whose
penalty is the identity matrix (`bs = "re"`) [@wood2017]. The package
`glmmTMB` writes a smooth as a random effect: the part of the smooth that the
penalty shrinks becomes a vector of random effects with a common variance, and
the part that the penalty leaves free becomes a fixed effect [@brooks2017]. We
compare with both throughout the chapter. As in @sec-gam we call `mgcv` as
`mgcv::gam()` without attaching it.

## A smooth and a random intercept {#sec-gamm-intercept}

The data frame `ChickWeight` of @sec-gam-by gives the weight, in grams, of 50
chicks from birth to day 21 [@crowder1990]. There we fitted one smooth curve of
time for each diet and ignored that the observations of one chick are
correlated. We now give each chick a random intercept and start with one curve
for all the chicks:
$$
\mu_{ij} = \beta_0 + f(t_{ij}) + b_i, \qquad b_i \sim \mathrm{N}(0, \tau).
$$

```{r}
#| label: fit-ri
ChickWeight$Diet <- factor(ChickWeight$Diet, ordered = FALSE)
fit_ri <- statmod(weight ~ s(Time, bspline_smooth(k = 8)) + random(~ 1 | Chick),
                  distrib = gaussian1_distrib(), data = ChickWeight)
summary(fit_ri)
```

The summary has a block for the smooth and a block for the random intercept.
The smooth spends `r fmt(fit_ri@edf$edf[2], 2)` effective degrees of
freedom out of its 7 coefficients. The standard deviation of the chicks'
intercepts is $\hat\tau = `r fmt(hyper(fit_ri)$estimate[2], 2)`$ grams, about
as large as the residual standard deviation
$\hat\sigma = `r fmt(exp(coef(fit_ri)$sigma[[1]]), 2)`$ grams.

In `mgcv` the random intercept is the smooth `s(Chick, bs = "re")`. Its
penalty is $\lambda_b\, b^\top b$, divided by the residual variance, so the
standard deviation of the random effects is $\sigma/\sqrt{\lambda_b}$:

```{r}
#| label: mgcv-ri
g_ri <- mgcv::gam(weight ~ s(Time, bs = "bs", k = 8) + s(Chick, bs = "re"),
                  data = ChickWeight, method = "REML")
c(statmod = hyper(fit_ri)$estimate[2],
  mgcv = sqrt(g_ri$sig2 / g_ri$sp[2]))
max(abs(fitted(fit_ri) - fitted(g_ri)))
```

The two packages give the same standard deviation and the same fitted values.

### A random slope

The chicks grow at different rates, so their weights spread out with time. A
random slope on time describes this:
$$
\mu_{ij} = \beta_0 + f(t_{ij}) + b_{0i} + b_{1i} t_{ij}, \qquad
(b_{0i}, b_{1i})^\top \sim \mathrm{N}(0, \Sigma_b).
$$

```{r}
#| label: fit-rs
fit_rs <- statmod(weight ~ s(Time, bspline_smooth(k = 8)) + random(~ Time | Chick),
                  distrib = gaussian1_distrib(), data = ChickWeight)
summary(fit_rs)
```

The two models have the same fixed effects, so their restricted
log-likelihoods can be compared. Under the smaller model the variance of the
slopes is zero, on the boundary of its range, so the likelihood ratio has the
distribution $\tfrac12\chi^2_1 + \tfrac12\chi^2_2$ and not $\chi^2_2$
[@self1987]:

```{r}
#| label: lr-rs
l_ri <- as.numeric(logLik(fit_ri, type = "marginal"))
l_rs <- as.numeric(logLik(fit_rs, type = "marginal"))
lr_rs <- 2 * (l_rs - l_ri)
lr_rs
0.5 * pchisq(lr_rs, df = 1, lower.tail = FALSE) +
  0.5 * pchisq(lr_rs, df = 2, lower.tail = FALSE)
```

The random slope is needed. The correlation between the intercepts and the
slopes is strongly negative. The deviation of a chick from the typical curve
is $b_{0i} + b_{1i}t$, and its variance
$\Sigma_{11} + 2t\,\Sigma_{12} + t^2\Sigma_{22}$ is smallest at
$t = -\Sigma_{12}/\Sigma_{22}$. As in @sec-lmm-slope, `param_value()` turns
the hyperparameters of the random slope into $\Sigma_b$:

```{r}
#| label: sigma-rs
h_rs <- hyper(fit_rs)
S_rs <- param_value(log_cholesky(2), eta = h_rs$estimate[h_rs$name != "lambda"])
S_rs
t_min <- -S_rs[1, 2] / S_rs[2, 2]
t_min
```

The correlation is
`r fmt(S_rs[1, 2] / sqrt(S_rs[1, 1] * S_rs[2, 2]), 3)`, and the chicks are
closest to each other `r fmt(t_min, 1)` days after birth and spread out
afterwards.

`glmmTMB()` accepts a smooth in its formula and gives the same estimates:

```{r}
#| label: tmb-rs
tmb_rs <- glmmTMB::glmmTMB(weight ~ s(Time, bs = "bs", k = 8) + (Time | Chick),
                           data = ChickWeight, REML = TRUE)
glmmTMB::VarCorr(tmb_rs)
logLik(tmb_rs)
logLik(fit_rs, type = "marginal")
```

The group `dummy` of `glmmTMB` is the smooth: its standard deviation measures
how much the wiggly part of the curve is shrunk, and it plays the role of
$\lambda$. The two restricted log-likelihoods differ by a constant. REML
integrates every coefficient without a penalty with a flat prior, and the
integral depends on the scale of the columns of these coefficients: if a
column is multiplied by $a$, its coefficient is divided by $a$ and the
integral changes by $\log\lvert a\rvert$. The free part of the smooth is a
straight line in time, and the two packages scale its column differently:

```{r}
#| label: reml-constant
x_sm <- model.matrix(fit_rs)[, "s(Time).lin"]
x_tmb <- glmmTMB::getME(tmb_rs, "X")[, "s(Time)1"]
a_lin <- coef(lm(x_sm ~ x_tmb))[["x_tmb"]]
log(abs(a_lin))
as.numeric(logLik(tmb_rs)) - l_rs
```

The constant is the same for every model with this smooth, so a difference
between two restricted log-likelihoods, like the likelihood ratio above, is
the same in both packages.

@fig-chicks shows the curve of a typical chick, with $b_i = 0$, and the
curves of five chicks.

```{r}
#| label: fig-chicks
#| fig-cap: "Weight against time for five chicks, with each chick's fitted curve and the curve of a typical chick (random intercept and slope)."
#| echo: false
#| fig-height: 4
tg <- seq(from = 0, to = 21, length.out = 100)
pick <- c("1", "13", "16", "18", "35")
cols <- book_palette[1:5]
chick_lev <- c(paste("chick", pick), "typical chick")
in_pick <- ChickWeight$Chick %in% pick
chick_pts <- data.frame(
  Time = ChickWeight$Time[in_pick], weight = ChickWeight$weight[in_pick],
  curve = factor(paste("chick", ChickWeight$Chick[in_pick]), levels = chick_lev))
chick_fit <- do.call(rbind, lapply(seq_along(pick), function(k) {
  nd <- data.frame(Time = tg, Chick = factor(pick[k],
                                             levels = levels(ChickWeight$Chick)))
  data.frame(Time = tg, weight = predict(fit_rs, what = "mu", newdata = nd),
             curve = chick_lev[k])
}))
nd0 <- data.frame(Time = tg, Chick = factor("1", levels = levels(ChickWeight$Chick)))
chick_fit <- rbind(chick_fit, data.frame(
  Time = tg, weight = predict(fit_rs, what = "mu", newdata = nd0, random = "zero"),
  curve = chick_lev[6]))
chick_fit$curve <- factor(chick_fit$curve, levels = chick_lev)
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point(colour = book_grey, size = 0.6, alpha = 0.4) +
  geom_point(data = chick_pts, aes(colour = curve, shape = curve), size = 1.2) +
  geom_line(data = chick_fit,
            aes(colour = curve, linetype = curve, linewidth = curve)) +
  scale_colour_manual(values = c(cols, "black"), limits = chick_lev) +
  scale_shape_manual(values = c(rep(16, 5), NA), limits = chick_lev) +
  scale_linetype_manual(values = c(rep("solid", 5), "dashed"),
                        limits = chick_lev) +
  scale_linewidth_manual(values = c(rep(0.6, 5), 1.2), limits = chick_lev) +
  guides(colour = guide_legend(nrow = 1), shape = guide_legend(nrow = 1),
         linetype = guide_legend(nrow = 1), linewidth = guide_legend(nrow = 1)) +
  labs(x = "day", y = "weight (g)")
```

### The band of the typical curve {#sec-gamm-band}

The curve of a typical chick is an estimate, and `predict()` with
`se = TRUE` gives it a pointwise 95% band. The band depends on the variance
matrix of @sec-beyond-variances. In a model with random effects the three
matrices answer three different questions. The matrix $V_f$ treats the
effects $b_i$ of the 50 chicks as fixed numbers, so it measures how much the
curve would change if we weighed the same 50 chicks again. The matrix $V_b$
treats the $b_i$ as draws from their gaussian distribution, so it measures
how well a sample of 50 chicks determines the curve of the population of
chicks. The matrix $V_u$ adds to $V_b$ the uncertainty of $\lambda$ and of
$\Sigma_b$. We compute the three bands at days 0, 10.5 and 21:

```{r}
#| label: band-types
tg <- seq(from = 0, to = 21, length.out = 101)
nd_t <- data.frame(Time = tg, Chick = factor("1", levels = levels(ChickWeight$Chick)))
types <- c("bayesian", "frequentist", "unconditional")
bands <- lapply(types, function(ty) {
  predict(fit_rs, what = "mu", newdata = nd_t, random = "zero", se = TRUE,
          type = ty)
})
names(bands) <- types
sapply(bands, function(p) p$se[c(1, 51, 101)])
```

The frequentist standard errors stay between
`r fmt(min(bands$frequentist$se), 1)` and
`r fmt(max(bands$frequentist$se), 1)` grams at every age. The
Bayesian ones grow with age, from `r fmt(bands$bayesian$se[1], 1)` grams at
birth to `r fmt(bands$bayesian$se[101], 1)` grams at day 21. The growth comes
from the random slopes: the chicks spread out with age, so the average curve
of 50 chicks is less certain at day 21 than at birth. The variance of the
deviation $b_{0i} + b_{1i}t$ of one chick is $(1, t)\,\Sigma_b\,(1, t)^\top$,
and the mean of 50 such deviations has a standard deviation of
$\sqrt{(1, t)\,\Sigma_b\,(1, t)^\top / 50}$:

```{r}
#| label: band-check
sd_mean <- function(t) sqrt(drop(c(1, t) %*% S_rs %*% c(1, t)) / 50)
rbind(bayesian = bands$bayesian$se[c(1, 51, 101)],
      chicks = sapply(c(0, 10.5, 21), sd_mean))
```

From the first week on, most of the Bayesian standard error is the variation
among chicks. The
frequentist band leaves this variation out, so it is the right band only for
a statement about these 50 chicks. For a statement about chicks in general,
which is what a curve of a typical chick is for, the Bayesian band is the
right one. The unconditional band is wider than the Bayesian band by at most
`r fmt(100 * max(bands$unconditional$se / bands$bayesian$se - 1), 1)` per
cent: the 50 chicks determine $\lambda$ and $\Sigma_b$ well enough that their
uncertainty adds little. @fig-band shows the three bands.

```{r}
#| label: fig-band
#| fig-cap: "Top: the curve of a typical chick with the 95% band from the Bayesian variance (shaded) and from the frequentist variance (dashed). Bottom: the standard error of the curve under the three variance matrices."
#| echo: false
#| fig-height: 6
pb <- bands$bayesian
pan <- factor(c("weight (g)", "standard error (g)"),
              levels = c("weight (g)", "standard error (g)"))
band_obs <- data.frame(Time = ChickWeight$Time, y = ChickWeight$weight,
                       panel = pan[1])
band_rib <- data.frame(Time = tg, lower = pb$lower, upper = pb$upper,
                       panel = pan[1])
band_lines <- rbind(
  data.frame(Time = tg, y = pb$fit, type = "bayesian", piece = "fit"),
  data.frame(Time = tg, y = bands$frequentist$lower, type = "frequentist",
             piece = "lower"),
  data.frame(Time = tg, y = bands$frequentist$upper, type = "frequentist",
             piece = "upper"))
band_lines$panel <- pan[1]
band_se <- data.frame(Time = rep(tg, times = 3),
                      y = c(pb$se, bands$frequentist$se, bands$unconditional$se),
                      type = rep(types, each = length(tg)),
                      piece = "se", panel = pan[2])
band_lines <- rbind(band_lines, band_se)
band_lines$type <- factor(band_lines$type, levels = types)
ggplot(band_lines, aes(x = Time, y = y)) +
  geom_point(data = band_obs, colour = book_grey, size = 0.6, alpha = 0.4) +
  geom_ribbon(data = band_rib, aes(y = NULL, ymin = lower, ymax = upper),
              fill = chalk, alpha = 0.3) +
  geom_line(aes(colour = type, linetype = type, linewidth = type,
                group = interaction(type, piece))) +
  geom_blank(data = data.frame(Time = 0, y = 0, panel = pan[2])) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_colour_manual(values = c(chalk, rust, ink)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
  scale_linewidth_manual(values = c(0.9, 0.6, 0.6)) +
  labs(x = "day", y = NULL) +
  theme(strip.placement = "outside", strip.background = element_blank(),
        strip.text = element_text(size = rel(1)))
```

In `mgcv` the matrices $V_b$ and $V_f$ are the components `Vp` and `Ve` of
a fit. On the model with a random intercept the typical curve, which leaves
out the term `s(Chick)`, has the same two bands in the two packages:

```{r}
#| label: band-mgcv
Xg <- predict(g_ri, newdata = nd_t, type = "lpmatrix")
Xg[, grepl("Chick", colnames(Xg))] <- 0
se_mgcv <- function(V) sqrt(rowSums((Xg %*% V) * Xg))
se_sm <- function(ty) {
  predict(fit_ri, what = "mu", newdata = nd_t, random = "zero", se = TRUE,
          type = ty)$se
}
c(bayesian = max(abs(se_sm("bayesian") / se_mgcv(g_ri$Vp) - 1)),
  frequentist = max(abs(se_sm("frequentist") / se_mgcv(g_ri$Ve) - 1)))
```

## A smooth for each group {#sec-gamm-by}

The chicks received four diets. As in @sec-gam-by, the argument `by = Diet`
of `s()` builds one curve of time for each diet, and the main effect `Diet`
carries the levels of the curves. We keep the random intercept and slope of
each chick:
$$
\mu_{ij} = \beta_0 + \beta_{d(i)} + f_{d(i)}(t_{ij}) + b_{0i} + b_{1i} t_{ij},
$$
where $d(i)$ is the diet of chick $i$. With `by_hyper = "shared"`, the
default, the four curves have one smoothing parameter; with
`by_hyper = "level"` each curve has its own.

```{r}
#| label: fit-by
fit_sh <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8), by = Diet) +
                    random(~ Time | Chick),
                  distrib = gaussian1_distrib(), data = ChickWeight)
fit_lv <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8), by = Diet,
                                    by_hyper = "level") +
                    random(~ Time | Chick),
                  distrib = gaussian1_distrib(), data = ChickWeight)
hyper(fit_sh)[, c("name", "estimate")]
hyper(fit_lv)[, c("name", "estimate")]
```

The smoothing parameter of diet 3 is about a quarter of those of diets 1 and
2, so its curve bends more, and that of diet 4 is in between. The model with one parameter is
the model with four under three equality constraints, and the constrained
values are inside the range of the parameters, so the likelihood ratio has
three degrees of freedom:

```{r}
#| label: lr-by
l_sh <- as.numeric(logLik(fit_sh, type = "marginal"))
l_lv <- as.numeric(logLik(fit_lv, type = "marginal"))
lr_by <- 2 * (l_lv - l_sh)
lr_by
pchisq(lr_by, df = 3, lower.tail = FALSE)
```

The data do not need four smoothing parameters.

In `mgcv` a smooth with a factor `by` has one smoothing parameter for each
level by default, and the argument `id` makes the levels share one. Its random
effects are smooths with an identity penalty, so the intercept and the slope of
a chick cannot be correlated there. With `correlated = FALSE` the two packages
fit the same model:

```{r}
#| label: mgcv-by
fit_shu <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8), by = Diet) +
                     random(~ Time | Chick, correlated = FALSE),
                   distrib = gaussian1_distrib(), data = ChickWeight)
g_shu <- mgcv::gam(weight ~ Diet + s(Time, bs = "bs", k = 8, by = Diet, id = 1) +
                     s(Chick, bs = "re") + s(Chick, Time, bs = "re"),
                   data = ChickWeight, method = "REML")
max(abs(fitted(fit_shu) - fitted(g_shu)))
edf_mu <- function(fit) sum(fit@edf$edf[fit@edf$parameter == "mu"])
c(statmod = edf_mu(fit_shu), mgcv = sum(g_shu$edf))
```

@fig-diets shows the curve of a typical chick of each diet with its 95%
Bayesian band, and the curve of the model with one smoothing parameter for
each diet. The two curves of a diet are close to each other and well inside
the band. The bands of the four diets overlap during the first week, and the
curves separate in the second and third weeks, when diet 3 gives the heaviest
chicks and diet 1 the lightest.

```{r}
#| label: fig-diets
#| fig-cap: "Weight against time for the chicks of each diet. The solid curve is a typical chick of the diet in the model with one smoothing parameter, with its 95% Bayesian band (shaded); the dashed curve is the model with one smoothing parameter for each diet."
#| echo: false
#| fig-height: 6
diet_lev <- paste("diet", 1:4)
diet_mod <- c("one smoothing parameter", "one for each diet")
diet_obs <- data.frame(Time = ChickWeight$Time, weight = ChickWeight$weight,
                       diet = factor(paste("diet", ChickWeight$Diet),
                                     levels = diet_lev))
diet_fit <- do.call(rbind, lapply(1:4, function(d) {
  nd <- data.frame(Time = tg, Diet = factor(d, levels = levels(ChickWeight$Diet)),
                   Chick = factor("1", levels = levels(ChickWeight$Chick)))
  pd <- predict(fit_sh, what = "mu", newdata = nd, random = "zero", se = TRUE)
  pl <- predict(fit_lv, what = "mu", newdata = nd, random = "zero")
  data.frame(Time = rep(tg, times = 2), weight = c(pd$fit, pl),
             lower = c(pd$lower, rep(NA, length(tg))),
             upper = c(pd$upper, rep(NA, length(tg))),
             model = rep(diet_mod, each = length(tg)), diet = diet_lev[d])
}))
diet_fit$model <- factor(diet_fit$model, levels = diet_mod)
diet_fit$diet <- factor(diet_fit$diet, levels = diet_lev)
ggplot(diet_obs, aes(x = Time, y = weight)) +
  geom_point(colour = book_grey, size = 0.6, alpha = 0.5) +
  geom_ribbon(data = diet_fit[diet_fit$model == diet_mod[1], ],
              aes(ymin = lower, ymax = upper), fill = chalk, alpha = 0.3) +
  geom_line(data = diet_fit, aes(colour = model, linetype = model),
            linewidth = 0.9) +
  facet_wrap(~ diet, ncol = 2) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "day", y = "weight (g)")
```

### A curve for each chick {#sec-gamm-chick}

In the models so far a chick differs from the typical chick of its diet by a
straight line, $b_{0i} + b_{1i}t$. @fig-chicks shows that some chicks have
curves of a different shape: some grow fast and then slow down, and others
start slowly. A model with `s(Time, by = Chick)` alone gives each chick a
smooth curve of its own. Each chick received one diet, so the curves of the
chicks then contain the curves of the diets, and the model has nothing left to
compare the diets. We keep the curves of the diets and add to each chick a
smooth deviation from the curve of its diet:
$$
\mu_{ij} = \beta_0 + \beta_{d(i)} + f_{d(i)}(t_{ij}) + b_{0i} + b_{1i} t_{ij}
  + g_i(t_{ij}).
$$
The 50 deviations $g_i$ have one smoothing parameter, so the data of all the
chicks determine how much a chick's curve may bend. This is a hierarchical
additive model in the sense of @pedersen2019, with one curve for each diet in
place of one curve for all the groups. The straight line of $g_i$ is already
in the random intercept and slope, so we drop it with
`null_space = "drop"`; then $g_i$ has no coefficient without a penalty. The
deviations have 200 coefficients, most of them zero in each row of the design,
and `sparse = TRUE` stores their block as a sparse matrix. The argument
`label` gives the coefficients of the deviations the prefix `dev`, because the
levels `1` to `4` of `Chick` would give them the same names as the curves of
the four diets.

```{r}
#| label: fit-gs
fit_gs <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8), by = Diet) +
                    random(~ Time | Chick) +
                    s(Time, bspline_smooth(k = 6, null_space = "drop"), by = Chick,
                      sparse = TRUE, label = "dev"),
                  distrib = gaussian1_distrib(), data = ChickWeight)
summary(fit_gs)
```

```{r}
#| label: sigma-gs
#| include: false
h_gs <- hyper(fit_gs)
S_gs <- param_value(log_cholesky(2),
                    eta = h_gs$estimate[h_gs$name != "lambda"])
```

The summary has one block for each of the four terms of the mean. The
parametric terms `Diet2`, `Diet3` and `Diet4` shift the curves of the diets
up or down. Each curve of a diet also has a level of its own, set by the
constraint that removes the constant from its basis, so these three
coefficients are not the differences between the diets on a given day; those
differences are read from the curves of @fig-diets. The
block of the diet curves shows the one smoothing parameter of the four curves
and the free straight line of each of them, `1.lin` to `4.lin`. As in
@sec-model, each multiplies the time standardized to mean zero and standard
deviation one over all the observations, so it is the slope of the linear
part of the diet's curve per standard deviation of time,
`r fmt(sd(ChickWeight$Time), 2)` days. The slopes are largest for diet 3 and
smallest for diet 1, as the curves of @fig-diets show. The block spends
`r fmt(fit_gs@edf$edf[2], 1)` effective degrees of freedom of its 28
coefficients. The block of the random intercept and slope reports the
standard deviation of the chicks' intercepts,
`r fmt(sqrt(S_gs[1, 1]), 1)` grams, and of their slopes,
`r fmt(sqrt(S_gs[2, 2]), 2)` grams a day, and their correlation,
`r fmt(S_gs[1, 2] / sqrt(S_gs[1, 1] * S_gs[2, 2]), 2)`. The block `dev` has
only one row: the smoothing parameter of the 50 deviations, with its
standard error and interval. Its 200 coefficients are not printed, because
each of them is the coordinate of one chick's deviation in a basis and has
no meaning of its own. The equation of $\sigma$ has only an intercept, on the
log scale, so the residual standard deviation is
$\exp(`r fmt(coef(fit_gs)$sigma[[1]], 3)`) = `r fmt(exp(coef(fit_gs)$sigma[[1]]), 2)`$
grams. The last lines give the effective degrees of freedom of the whole
model, `r fmt(sum(fit_gs@edf$edf), 1)` for 578 observations, and the
conditional AIC.

With straight deviations the residual standard deviation was larger:

```{r}
#| label: sigma-compare
c(straight = exp(coef(fit_sh)$sigma[[1]]), smooth = exp(coef(fit_gs)$sigma[[1]]))
```

The deviations spend `r fmt(fit_gs@edf$edf[4], 1)` effective degrees of
freedom of their 200 coefficients, about
`r fmt(fit_gs@edf$edf[4] / 50, 1)` for each chick. The residual standard
deviation falls from `r fmt(exp(coef(fit_sh)$sigma[[1]]), 1)` to
`r fmt(exp(coef(fit_gs)$sigma[[1]]), 1)` grams, so the residual variance
falls by
`r fmt(100 * (1 - exp(2 * (coef(fit_gs)$sigma[[1]] - coef(fit_sh)$sigma[[1]]))), 0)`
per cent. Most of what the model with straight deviations left as noise is
the shape of the chicks' own curves. The two models have the same coefficients without a penalty, so their
restricted log-likelihoods can be compared. Under the smaller model the
variance $1/\lambda$ of the deviations is zero, on the boundary of its range,
and the likelihood ratio has the distribution
$\tfrac12\chi^2_0 + \tfrac12\chi^2_1$:

```{r}
#| label: lr-gs
l_gs <- as.numeric(logLik(fit_gs, type = "marginal"))
lr_gs <- 2 * (l_gs - l_sh)
lr_gs
0.5 * pchisq(lr_gs, df = 1, lower.tail = FALSE)
```

The curve of a typical chick of a diet is now
$\beta_0 + \beta_d + f_d(t)$, without $b_{0i} + b_{1i}t$ and without $g_i$.
`predict()` with `random = "zero"` sets the random intercept and slope to
zero, but $g_i$ is a smooth and not a `random()` term, so `predict()` keeps it.
We build the curve, and any other part of the predictor, from the columns of
the design and from the same rows and columns of `vcov()`. The function
`part_band()` takes the prefixes of the columns and the rows of the data where
the part is wanted:

```{r}
#| label: part-band
part_band <- function(fit, prefix, rows = NULL, type = "bayesian") {
  X <- as.matrix(model.matrix(fit))
  if (!is.null(rows)) X <- X[rows, , drop = FALSE]
  keep <- Reduce(`|`, lapply(prefix, function(p) startsWith(colnames(X), p)))
  nm <- colnames(X)[keep]
  Xk <- X[, nm, drop = FALSE]
  V <- as.matrix(vcov(fit, type = type)[paste0("mu:", nm), paste0("mu:", nm)])
  data.frame(fit = drop(Xk %*% coef(fit)$mu[nm]),
             se = sqrt(rowSums((Xk %*% V) * Xk)))
}
```

On the model of @sec-gamm-intercept the part `s(Time)` is the contribution of
the smooth, which sums to zero over the data, and it has the value and the
standard error that `mgcv` reports with `type = "terms"`:

```{r}
#| label: part-mgcv
p_ri <- part_band(fit_ri, prefix = "s(Time).")
t_ri <- predict(g_ri, type = "terms", se.fit = TRUE)
c(fit = max(abs(p_ri$fit - t_ri$fit[, "s(Time)"])),
  se = max(abs(p_ri$se - t_ri$se.fit[, "s(Time)"])))
```

The curve of diet 3 is the part with the prefixes `(Intercept)`, `Diet3` and
`s(Time).3.`. We read it at the twelve days of chick 33, which received
diet 3 and was weighed on every day:

```{r}
#| label: diets-gs
r33 <- which(ChickWeight$Chick == "33")
d3 <- c("(Intercept)", "Diet3", "s(Time).3.")
rbind(day = ChickWeight$Time[r33],
      straight = part_band(fit_sh, prefix = d3, rows = r33)$fit,
      smooth = part_band(fit_gs, prefix = d3, rows = r33)$fit)
rbind(straight = part_band(fit_sh, prefix = d3, rows = r33)$se,
      smooth = part_band(fit_gs, prefix = d3, rows = r33)$se)
```

The curve of diet 3 is the same in the two models to within
`r fmt(max(abs(part_band(fit_sh, prefix = d3, rows = r33)$fit - part_band(fit_gs, prefix = d3, rows = r33)$fit)), 0)`
grams. Its band is a little wider when the deviations are smooth, although
the residual standard deviation is three times smaller. As in
@sec-gamm-band, most of the standard error of a typical curve is the variation
among chicks, and with smooth deviations the chicks vary in the shape of their
curves as well as in their intercept and slope.

@fig-gs shows the six chicks whose curves change most between the two models.
With a straight deviation the curve of a chick has the shape of the curve of
its diet. With a smooth deviation it follows the chick's own weights: chicks
33 and 11 stop growing in the last week, and chicks 29 and 7 grow faster than
the curve of their diet in the third week. The straight deviation cannot
describe these shapes, and to come close to the late weights of chick 33 it
puts the weight at birth about 20 grams too high.

```{r}
#| label: fig-gs
#| fig-cap: "Six chicks: their weights, the fitted curve with a smooth deviation from the diet's curve (solid) and with a straight deviation (dashed), and the curve of a typical chick of the diet (dotted)."
#| echo: false
#| fig-height: 5.5
cid <- as.character(ChickWeight$Chick)
full <- names(which(table(cid) == 12))
gap <- tapply((fitted(fit_gs) - fitted(fit_sh))^2, cid, mean)[full]
show6 <- names(sort(gap, decreasing = TRUE))[1:6]
show6 <- show6[order(as.numeric(show6))]
gs_lev <- c("smooth deviation", "straight deviation", "typical chick of the diet")
gs_strip <- character(0)
gs_obs <- list()
gs_fit <- list()
for (ck in show6) {
  rows <- which(cid == ck)
  d <- ChickWeight$Diet[rows][1]
  strip <- paste0("chick ", ck, ", diet ", d)
  gs_strip <- c(gs_strip, strip)
  nd <- data.frame(Time = tg, Diet = d,
                   Chick = factor(ck, levels = levels(ChickWeight$Chick)))
  y_gs <- predict(fit_gs, what = "mu", newdata = nd)
  y_sh <- predict(fit_sh, what = "mu", newdata = nd)
  y_d <- part_band(fit_gs, prefix = c("(Intercept)", paste0("Diet", d),
                                      paste0("s(Time).", d, ".")), rows = rows)$fit
  gs_obs[[ck]] <- data.frame(Time = ChickWeight$Time[rows],
                             weight = ChickWeight$weight[rows], chick = strip)
  gs_fit[[ck]] <- data.frame(
    Time = c(tg, tg, ChickWeight$Time[rows]), weight = c(y_gs, y_sh, y_d),
    curve = rep(gs_lev, times = c(length(tg), length(tg), length(rows))),
    chick = strip)
}
gs_obs <- do.call(rbind, gs_obs)
gs_fit <- do.call(rbind, gs_fit)
gs_obs$chick <- factor(gs_obs$chick, levels = gs_strip)
gs_fit$chick <- factor(gs_fit$chick, levels = gs_strip)
gs_fit$curve <- factor(gs_fit$curve, levels = gs_lev)
ggplot(gs_obs, aes(x = Time, y = weight)) +
  geom_point(colour = book_grey, size = 1.2) +
  geom_line(data = gs_fit, aes(colour = curve, linetype = curve,
                               linewidth = curve)) +
  facet_wrap(~ chick, ncol = 3, scales = "free_y") +
  scale_colour_manual(values = c(chalk, rust, ink)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
  scale_linewidth_manual(values = c(0.9, 0.9, 0.7)) +
  labs(x = "day", y = "weight (g)")
```

## Smooths and random effects in sigma {#sec-gamm-sigma}

The spread of the weights around a chick's curve also grows with age, and some
chicks are more variable than others. We give the equation of $\sigma$ a
smooth of time and a random intercept of its own:
$$
\log\sigma_{ij} = \gamma_0 + g(t_{ij}) + c_i, \qquad
c_i \sim \mathrm{N}(0, \kappa),
$$
independent of the random intercept $b_i$ of the mean.

```{r}
#| label: fit-sigma
fit_sg0 <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8)) +
                     random(~ 1 | Chick) |
                     sigma ~ s(Time, bspline_smooth(k = 8)),
                   distrib = gaussian1_distrib(), data = ChickWeight)
fit_sg <- statmod(weight ~ Diet + s(Time, bspline_smooth(k = 8)) +
                    random(~ 1 | Chick) |
                    sigma ~ s(Time, bspline_smooth(k = 8)) + random(~ 1 | Chick),
                  distrib = gaussian1_distrib(), data = ChickWeight)
summary(fit_sg)
```

The random intercepts of $\log\sigma$ have a standard deviation of
$\hat\kappa = `r fmt(hyper(fit_sg)$estimate[4], 2)`$: the residual standard
deviation of one chick can be
$\exp(1.96\,\hat\kappa) = `r fmt(exp(1.96 * hyper(fit_sg)$estimate[4]), 1)`$
times larger or smaller than that of a typical chick. The two models have the
same fixed effects in the mean, and under the smaller one $\kappa^2$ is zero,
on the boundary of its range, so the likelihood ratio has the distribution
$\tfrac12\chi^2_0 + \tfrac12\chi^2_1$ [@self1987]:

```{r}
#| label: lr-sigma
lr_sg <- 2 * (as.numeric(logLik(fit_sg, type = "marginal")) -
                as.numeric(logLik(fit_sg0, type = "marginal")))
lr_sg
0.5 * pchisq(lr_sg, df = 1, lower.tail = FALSE)
c(without = hyper(fit_sg0)$estimate[2], with = hyper(fit_sg)$estimate[2])
```

The last line gives the standard deviation of the chicks' intercepts in the
mean, without and with the random effects in $\sigma$:
`r fmt(hyper(fit_sg0)$estimate[2], 2)` and
`r fmt(hyper(fit_sg)$estimate[2], 2)` grams. In the model of
@sec-gamm-intercept, with a constant $\sigma$, it was
`r fmt(hyper(fit_ri)$estimate[2], 1)` grams. A chick that grows faster than
the others has its weights close to the typical curve at the start and far
from it at the end. With a constant $\sigma$ the model describes this with a
large random intercept, which shifts the whole curve of the chick. With a
$\sigma$ that grows with time it describes it with a residual spread that
grows, and the random intercept is left with the small differences at birth.

In `glmmTMB()` the equation of $\sigma$ is the argument `dispformula`, which
accepts a smooth and a random effect:

```{r}
#| label: tmb-sigma
tmb_sg <- glmmTMB::glmmTMB(weight ~ Diet + s(Time, bs = "bs", k = 8) + (1 | Chick),
                           dispformula = ~ s(Time, bs = "bs", k = 8) + (1 | Chick),
                           data = ChickWeight, REML = TRUE)
glmmTMB::VarCorr(tmb_sg)
max(abs(fitted(fit_sg) - fitted(tmb_sg)))
```

The standard deviations differ by less than 3 per cent, and the fitted means
differ by up
to `r fmt(max(abs(fitted(fit_sg) - fitted(tmb_sg))), 2)` grams. The difference
comes from the smooth of $\sigma$. A smooth with a penalty on the second
derivative splits into a straight line, which the penalty leaves free, and a
wiggly part, which the penalty shrinks. The split is not unique, because a
straight line can be added to the wiggly part, and `statmodels7` and `mgcv`
(whose smooths `glmmTMB` uses) make different choices. REML integrates the
straight line of the mean, and then the choice changes the criterion only by
the constant of @sec-gamm-intercept. The straight line of $\sigma$ is not
integrated but estimated (@sec-beyond-outer), and then the prior of the wiggly
part, and so the estimates, depend on the choice. The same happens to the
smooth of the mean under maximum likelihood, which estimates its straight
line:

```{r}
#| label: ml-split
fit_ml <- statmod(weight ~ s(Time, bspline_smooth(k = 8)),
                  distrib = gaussian1_distrib(), data = ChickWeight,
                  outer_criterion = ml(marginal = "all"))
g_ml <- mgcv::gam(weight ~ s(Time, bs = "bs", k = 8), data = ChickWeight,
                  method = "ML")
c(statmod = exp(coef(fit_ml)$sigma[[1]]), mgcv = sqrt(g_ml$sig2))
c(statmod = as.numeric(logLik(fit_ml, type = "marginal")), mgcv = -g_ml$gcv.ubre)
```

The two maxima differ: they are maxima of two slightly different models.

@fig-sigma shows the residual standard deviation of a typical chick against
time, with the interval that holds the standard deviation of 95% of the
chicks (`interval = "group"`, @sec-glmm-predict).

```{r}
#| label: fig-sigma
#| fig-cap: "Residual standard deviation of the weight against time: a typical chick (solid) and the interval that holds 95% of the chicks (dashed)."
#| echo: false
#| fig-height: 3.8
nd_s <- data.frame(Time = tg, Diet = factor(1, levels = levels(ChickWeight$Diet)),
                   Chick = "new")
sg <- predict(fit_sg, what = "sigma", newdata = nd_s, random = "zero",
              interval = "group")
sg_lev <- c("typical chick", "95% of the chicks")
sg_df <- data.frame(Time = rep(tg, times = 3),
                    sigma = c(sg$fit, sg$lower, sg$upper),
                    curve = factor(rep(sg_lev[c(1, 2, 2)], each = length(tg)),
                                   levels = sg_lev),
                    piece = rep(c("fit", "lower", "upper"), each = length(tg)))
ggplot(sg_df, aes(x = Time, y = sigma)) +
  geom_line(aes(linetype = curve, linewidth = curve, group = piece),
            colour = chalk) +
  expand_limits(y = 0) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_linewidth_manual(values = c(0.9, 0.5)) +
  labs(x = "day", y = "sigma (g)")
```

## A model for counts {#sec-gamm-counts}

We return to the barn owls of @sec-glmm-negbin: the number of calls of the
nestlings in 27 nests before a parent arrived [@roulin2007]. The parents
arrived between about 21:40 and 5:15, and @zuur2009 found that the number of calls
changes with the time of arrival. We add a smooth of the arrival time to the
negative binomial model of @sec-glmm-negbin:
$$
\log\mu_{ij} = \log n_{ij} + x_{ij}^\top\beta + f(t_{ij}) + b_i, \qquad
b_i \sim \mathrm{N}(0, \tau),
$$
where $n_{ij}$ is the number of nestlings and $t_{ij}$ the arrival time, in
hours after midnight of the previous day (25 is 1:00 in the morning).

```{r}
#| label: fit-owls
data(Owls, package = "glmmTMB")
fit_ow <- statmod(SiblingNegotiation ~ FoodTreatment * SexParent +
                    offset(logBroodSize) + s(ArrivalTime, bspline_smooth(k = 10)) +
                    random(~ 1 | Nest),
                  distrib = negbin2_distrib(), data = Owls)
summary(fit_ow)
```

The family `nb()` of `mgcv` estimates $\theta$ together with the smoothing
parameters, and `glmmTMB()` accepts the smooth with the family `nbinom2`:

```{r}
#| label: ref-owls
g_ow <- mgcv::gam(SiblingNegotiation ~ FoodTreatment * SexParent +
                    offset(logBroodSize) + s(ArrivalTime, bs = "bs", k = 10) +
                    s(Nest, bs = "re"),
                  family = mgcv::nb(), data = Owls, method = "REML")
tmb_ow <- glmmTMB::glmmTMB(SiblingNegotiation ~ FoodTreatment * SexParent +
                             offset(logBroodSize) +
                             s(ArrivalTime, bs = "bs", k = 10) + (1 | Nest),
                           family = glmmTMB::nbinom2, data = Owls, REML = TRUE)
c(statmod = exp(coef(fit_ow)$theta[[1]]),
  mgcv = g_ow$family$getTheta(TRUE), glmmTMB = glmmTMB::sigma(tmb_ow))
max(abs(predict(fit_ow, what = "link:mu") - predict(g_ow, type = "link")))
```

@fig-owls shows the calls per nestling of a typical nest against the arrival
time. The nestlings call most when a parent arrives at about 22:40, and least
when it arrives near midnight and near 3:00. The model has no interaction
between the food treatment and the time, so on the log scale the two curves
differ by a constant, and the food-deprived nestlings call more at every hour.
The bands are widest at the two ends of the night, where few parents
arrived.

```{r}
#| label: fig-owls
#| fig-cap: "Calls per nestling against the arrival time of the parent, for a typical nest with a male parent: food-deprived (solid) and satiated (dashed) nestlings, with 95% Bayesian bands (shaded). The points are the observed calls per nestling."
#| echo: false
#| fig-height: 3.8
ag <- seq(from = min(Owls$ArrivalTime), to = max(Owls$ArrivalTime),
          length.out = 100)
food <- c("Deprived", "Satiated")
food_lev <- c("deprived", "satiated")
owl_fit <- do.call(rbind, lapply(1:2, function(k) {
  nd <- data.frame(ArrivalTime = ag,
                   FoodTreatment = factor(food[k], levels(Owls$FoodTreatment)),
                   SexParent = factor("Male", levels(Owls$SexParent)),
                   logBroodSize = 0, Nest = "new")
  po <- predict(fit_ow, what = "mu", newdata = nd, random = "zero", se = TRUE)
  data.frame(ArrivalTime = ag, fit = po$fit, lower = po$lower,
             upper = po$upper, food = food_lev[k])
}))
owl_fit$food <- factor(owl_fit$food, levels = food_lev)
owl_obs <- data.frame(ArrivalTime = Owls$ArrivalTime,
                      calls = Owls$SiblingNegotiation / Owls$BroodSize)
ggplot(owl_fit, aes(x = ArrivalTime)) +
  geom_point(data = owl_obs, aes(y = calls), colour = book_grey, size = 0.6,
             alpha = 0.5) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = food), alpha = 0.25) +
  geom_line(aes(y = fit, colour = food, linetype = food), linewidth = 0.9) +
  scale_colour_manual(values = c(rust, chalk)) +
  scale_fill_manual(values = c(rust, chalk)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "arrival time (hours)", y = "calls per nestling")
```

### Random effects on the dispersion

As in @sec-glmm-negbin we let $\theta$ depend on the food treatment and on the
nest:

```{r}
#| label: fit-owls-theta
fit_ow2 <- statmod(SiblingNegotiation ~ FoodTreatment * SexParent +
                     offset(logBroodSize) + s(ArrivalTime, bspline_smooth(k = 10)) +
                     random(~ 1 | Nest) |
                     theta ~ FoodTreatment + random(~ 1 | Nest),
                   distrib = negbin2_distrib(), data = Owls)
hyper(fit_ow2)[, c("parameter", "name", "estimate")]
tmb_ow2 <- glmmTMB::glmmTMB(SiblingNegotiation ~ FoodTreatment * SexParent +
                              offset(logBroodSize) +
                              s(ArrivalTime, bs = "bs", k = 10) + (1 | Nest),
                            dispformula = ~ FoodTreatment + (1 | Nest),
                            family = glmmTMB::nbinom2, data = Owls, REML = TRUE)
glmmTMB::VarCorr(tmb_ow2)
```

The two models have the same fixed effects in the mean and the same smooth, so
the constant of @sec-gamm-intercept cancels in the difference of their
restricted log-likelihoods, and the two packages give the same likelihood
ratio:

```{r}
#| label: lr-owls
lr_ow <- 2 * (as.numeric(logLik(fit_ow2, type = "marginal")) -
                as.numeric(logLik(fit_ow, type = "marginal")))
lr_ow_tmb <- 2 * (as.numeric(logLik(tmb_ow2)) - as.numeric(logLik(tmb_ow)))
c(statmod = lr_ow, glmmTMB = lr_ow_tmb)
```

## Shared smoothing parameters {#sec-gamm-id}

The argument `by_hyper` decides whether the curves of one factor share a
smoothing parameter. The argument `id` of `s()`, `te()`, `random()` and the
penalized terms is more general: the hyperparameters of all the terms with the
same `id` are estimated at one value, wherever the terms are in the formula.

### Lagged covariates

A typical use is a covariate that enters the model at several lags. In the
Chicago data of @sec-gam-counts [@peng2008] the deaths of a day may depend on
the temperature of that day and of the two days before. The three curves
describe the same kind of effect, and it is natural to smooth them by the same
amount:
$$
\log\mu_t = x_t^\top\beta + h(t) + f_0(T_t) + f_1(T_{t-1}) + f_2(T_{t-2}),
$$
where $h$ is a smooth of the date, $x_t$ is the day of the week and $T_t$ the
mean temperature of day $t$.

```{r}
#| label: fit-lags
data(chicagoNMMAPS, package = "dlnm")
chicagoNMMAPS <- chicagoNMMAPS[chicagoNMMAPS$year >= 1995 &
                                 chicagoNMMAPS$year <= 2000, ]
chicagoNMMAPS$temp1 <- c(NA, head(chicagoNMMAPS$temp, -1))
chicagoNMMAPS$temp2 <- c(NA, NA, head(chicagoNMMAPS$temp, -2))
chicagoNMMAPS <- chicagoNMMAPS[-(1:2), ]
fit_id <- statmod(death ~ dow + s(time, bspline_smooth(k = 24)) +
                    s(temp, bspline_smooth(k = 8), id = "T") +
                    s(temp1, bspline_smooth(k = 8), id = "T") +
                    s(temp2, bspline_smooth(k = 8), id = "T"),
                  distrib = poisson_distrib(), data = chicagoNMMAPS)
fit_free <- statmod(death ~ dow + s(time, bspline_smooth(k = 24)) +
                      s(temp, bspline_smooth(k = 8)) +
                      s(temp1, bspline_smooth(k = 8)) +
                      s(temp2, bspline_smooth(k = 8)),
                    distrib = poisson_distrib(), data = chicagoNMMAPS)
hyper(fit_id)[, c("term", "estimate")]
hyper(fit_free)[, c("term", "estimate")]
```

The `id` of `mgcv` does the same:

```{r}
#| label: mgcv-lags
g_id <- mgcv::gam(death ~ dow + s(time, bs = "bs", k = 24) +
                    s(temp, bs = "bs", k = 8, id = 1) +
                    s(temp1, bs = "bs", k = 8, id = 1) +
                    s(temp2, bs = "bs", k = 8, id = 1),
                  family = poisson, data = chicagoNMMAPS, method = "REML")
max(abs(predict(fit_id, what = "link:mu") - predict(g_id, type = "link")))
```

The shared parameter is a constraint on the free ones, inside their range, so
we compare the two models with a likelihood ratio on two degrees of freedom:

```{r}
#| label: lr-lags
lr_id <- 2 * (as.numeric(logLik(fit_free, type = "marginal")) -
                as.numeric(logLik(fit_id, type = "marginal")))
lr_id
pchisq(lr_id, df = 2, lower.tail = FALSE)
fit_free@edf[3:5, c("term", "edf")]
```

Here the data reject the shared parameter. With three free parameters the
curve of the same day's temperature is smoothed much more than the other two:
it spends `r fmt(fit_free@edf$edf[3], 1)` effective degrees of freedom of its
7 coefficients, against `r fmt(fit_free@edf$edf[4], 1)` and
`r fmt(fit_free@edf$edf[5], 1)` for the two previous days.

The function `part_band()` of @sec-gamm-chick gives the contribution
$f_k(T)$ of each curve with its band. The contribution sums to zero over the
data, and its band leaves out the uncertainty of the intercept. @fig-lags shows
the three curves of the two models as ratios of the death rate,
$\exp\{f_k(T)\}$, with their 95% bands.

```{r}
#| label: fig-lags
#| fig-cap: "Ratio of the daily death rate to the rate at the average of the smooth, against the temperature of the same day (lag 0) and of the two days before (lags 1 and 2), with one shared smoothing parameter (solid, band shaded) and one for each lag (dashed, band dashed)."
#| echo: false
#| fig-height: 3.6
labs <- c("s(temp)", "s(temp1)", "s(temp2)")
vars <- c("temp", "temp1", "temp2")
lag_lev <- paste("lag", 0:2)
lag_mod <- c("one shared smoothing parameter", "one for each lag")
lo <- function(b) exp(b$fit - 1.96 * b$se)
hi <- function(b) exp(b$fit + 1.96 * b$se)
lag_df <- do.call(rbind, lapply(1:3, function(k) {
  x <- chicagoNMMAPS[[vars[k]]]
  o <- order(x)
  a <- part_band(fit_id, prefix = paste0(labs[k], "."))[o, ]
  f <- part_band(fit_free, prefix = paste0(labs[k], "."))[o, ]
  data.frame(temp = rep(x[o], times = 2),
             ratio = exp(c(a$fit, f$fit)),
             lower = c(lo(a), lo(f)), upper = c(hi(a), hi(f)),
             model = rep(lag_mod, each = length(o)), lag = lag_lev[k])
}))
lag_df$model <- factor(lag_df$model, levels = lag_mod)
lag_df$lag <- factor(lag_df$lag, levels = lag_lev)
lag_free <- lag_df[lag_df$model == lag_mod[2], ]
lag_rug <- data.frame(
  temp = unlist(lapply(vars, function(v) chicagoNMMAPS[[v]]), use.names = FALSE),
  lag = factor(rep(lag_lev, each = nrow(chicagoNMMAPS)), levels = lag_lev))
ggplot(lag_df, aes(x = temp)) +
  geom_hline(yintercept = 1, linetype = "dotted", colour = grey) +
  geom_ribbon(data = lag_df[lag_df$model == lag_mod[1], ],
              aes(ymin = lower, ymax = upper), fill = chalk, alpha = 0.3) +
  geom_line(data = lag_free, aes(y = lower), colour = rust,
            linetype = "dashed", linewidth = 0.4) +
  geom_line(data = lag_free, aes(y = upper), colour = rust,
            linetype = "dashed", linewidth = 0.4) +
  geom_line(aes(y = ratio, colour = model, linetype = model), linewidth = 0.9) +
  geom_rug(data = lag_rug, colour = grey, length = unit(0.02, "npc"),
           alpha = 0.3) +
  facet_wrap(~ lag, nrow = 1) +
  scale_y_log10() +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "temperature (C)", y = "rate ratio")
```

```{r}
#| label: lags-numbers
#| include: false
hot <- which.max(chicagoNMMAPS$temp)
lag_at <- function(fit, k, i) {
  exp(part_band(fit, prefix = paste0(labs[k], "."), rows = i)$fit)
}
hot1 <- which.max(chicagoNMMAPS$temp1)
hot2 <- which.max(chicagoNMMAPS$temp2)
cold2 <- which.min(abs(chicagoNMMAPS$temp2 + 15))
```

The three temperatures are strongly correlated, so each curve is the effect
of the temperature of one day with the temperatures of the other two days
held fixed. On 98 days of 100 the two models give curves that differ by less
than half a per cent in the rate. The shared parameter changes the curves at
their ends, where the data are few, and most of all the curve of the same
day: with its own parameter this curve is smoother, flatter on the coldest
days and less steep on the hottest. This is the difference that the
likelihood ratio detects. It does not change what the curves say about the
effect of temperature on mortality. Heat acts within two days. The highest
mean temperature in the data, `r fmt(chicagoNMMAPS$temp[hot], 1)` degrees, was
reached on `r format(chicagoNMMAPS$date[hot], "%d %B %Y")`, during the
Chicago heat wave of July 1995. At this temperature the rate ratio is
`r fmt(lag_at(fit_free, 2, hot1), 2)` when it is the temperature of the day
before and `r fmt(lag_at(fit_free, 3, hot2), 2)` when it is the temperature of
two days before. Cold acts later: at lag 2 a temperature
of $-15$ degrees multiplies the rate by `r fmt(lag_at(fit_free, 3, cold2), 2)`,
while on the same day the cold lowers it a little.

### Levels that share a parameter

`id` also groups some levels of a factor `by`. Under `by_hyper = "level"` the
hyperparameters of the levels are named `lambda.1`, `lambda.2` and so on, and
`id` takes one label for each of them. In @sec-gamm-by the smoothing
parameter of diet 3 was the smallest, so we let diets 1, 2 and 4 share one
parameter and give diet 3 its own:

```{r}
#| label: fit-id-levels
fit_lid <- statmod(weight ~ Diet +
                     s(Time, bspline_smooth(k = 8), by = Diet, by_hyper = "level",
                       id = c(lambda.1 = "A", lambda.2 = "A", lambda.3 = "B",
                              lambda.4 = "A")) +
                     random(~ Time | Chick),
                   distrib = gaussian1_distrib(), data = ChickWeight)
hyper(fit_lid)[, c("name", "estimate")]
l_lid <- as.numeric(logLik(fit_lid, type = "marginal"))
lr_lid <- 2 * (l_lv - l_lid)
lr_lid
pchisq(lr_lid, df = 2, lower.tail = FALSE)
```

The model with two smoothing parameters is the model with four under two
equality constraints, and the likelihood ratio does not reject them: diets 1,
2 and 4 can share one smoothing parameter, and diet 3 keeps its own.

```{r}
#| label: gate-09
#| include: false
assert_gamm_ok(
  fits = list(ri = fit_ri, rs = fit_rs, sh = fit_sh, lv = fit_lv, shu = fit_shu,
              gs = fit_gs,
              sg0 = fit_sg0, sg = fit_sg, ml = fit_ml, ow = fit_ow, ow2 = fit_ow2,
              id = fit_id, free = fit_free, lid = fit_lid),
  ref = list(g_ri = g_ri, tmb_rs = tmb_rs, g_shu = g_shu, tmb_sg = tmb_sg,
             g_ml = g_ml, g_ow = g_ow, tmb_ow = tmb_ow, tmb_ow2 = tmb_ow2,
             g_id = g_id),
  data = list(ChickWeight = ChickWeight, Owls = Owls,
              chicagoNMMAPS = chicagoNMMAPS),
  claims = list(lr_rs = lr_rs, lr_by = lr_by, lr_sg = lr_sg, lr_ow = lr_ow,
                lr_ow_tmb = lr_ow_tmb, lr_id = lr_id, lr_lid = lr_lid,
                a_lin = a_lin, t_min = t_min, bands = bands, part_ri = p_ri,
                lr_gs = lr_gs,
                diet3 = list(straight = part_band(fit_sh, prefix = d3, rows = r33),
                             smooth = part_band(fit_gs, prefix = d3, rows = r33)),
                lag_ratios = c(lag_at(fit_free, 2, hot1), lag_at(fit_free, 3, hot2),
                               lag_at(fit_free, 3, cold2))))
```
