# Penalized regression {#sec-penalized}

```{r}
#| label: setup-10
#| include: false
source("R/_setup.R")
source("R/penalized-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"; ink <- "grey15"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
code <- function(v) {
  v <- paste0("\x60", v, "\x60")
  if (length(v) < 2) return(v)
  paste(paste(v[-length(v)], collapse = ", "), "and", v[length(v)])
}
```

A penalized regression subtracts a penalty on the coefficients from the
log-likelihood. The penalty shrinks the estimates towards zero. Some penalties
also set coefficients exactly to zero, and so they choose which covariates
stay in the model. In `statmodels7` a penalty is a term of the formula. The
terms `ridge()`, `lasso()`, `enet()`, `scad()` and `mcp()` each contain a
block of covariates, and the penalty acts only on the coefficients of that
block. Each penalty is minus the logarithm of a density for the coefficients,
with its normalizing constant. The penalty of a smooth (@sec-gam) and the
penalty of a random effect (@sec-lmm) have the same form, so one model can
contain all three.

## Ridge, lasso and elastic net {#sec-penalized-basic}

The data frame `UScrime` gives the crime rate in 47 states of the USA in
1960, together with 15 covariates [@ehrlich1973; @venables2002]. The
covariates include the police expenditure in 1960 (`Po1`) and in 1959
(`Po2`), the mean years of schooling (`Ed`), the income inequality (`Ineq`)
and the probability of imprisonment (`Prob`).

```{r}
#| label: crime-data
data(UScrime, package = "MASS")
UScrime$lrate <- log(UScrime$y)
xs <- ~ M + So + Ed + Po1 + Po2 + LF + M.F + Pop + NW + U1 + U2 + GDP +
  Ineq + Prob + Time
X <- model.matrix(xs, data = UScrime)[, -1]
cor(UScrime$Po1, UScrime$Po2)
```

With 47 observations and 15 covariates, the least squares estimates are
unstable. Some covariates are also strongly correlated, for example the two
police expenditures. Let $y_i$ be the logarithm of the crime rate. The model
is
$$
y_i \sim \mathrm{N}(\mu_i, \sigma), \qquad
\mu_i = \beta_0 + \sum_{j=1}^{15} \beta_j x_{ij}.
$$
Let $s_j$ be the standard deviation of covariate $j$, and let
$u_j = s_j \beta_j$. We subtract from the log-likelihood one of three
penalties on the $u_j$:
$$
\begin{aligned}
\text{ridge: } & \tfrac{1}{2}\lambda \textstyle\sum_j u_j^2
  - \tfrac{p}{2}\log\{\lambda/(2\pi)\}, \\
\text{lasso: } & \lambda \textstyle\sum_j \lvert u_j\rvert
  - p\log(\lambda/2), \\
\text{elastic net: } & \lambda \textstyle\sum_j
  \big\{\alpha\lvert u_j\rvert + \tfrac{1}{2}(1-\alpha)u_j^2\big\}
  + p\log Z(\lambda, \alpha).
\end{aligned}
$$
Here $p = 15$, and $Z$ is the normalizing constant of the density
$\exp[-\lambda\{\alpha\lvert u\rvert + (1-\alpha)u^2/2\}]$. The ridge
penalty is minus the log density of a gaussian distribution with mean zero
and precision $\lambda$ [@hoerl1970]. The lasso penalty is minus the log
density of a Laplace distribution with rate $\lambda$ [@tibshirani1996]. The
elastic net mixes the two [@zou2005]. The factor $s_j$ puts the covariates
on a common scale, so that the penalty does not depend on the units of each
covariate. We include this factor by setting `standardize = TRUE` in the
term.

### The ridge penalty

```{r}
#| label: fit-ridge
fit_ridge <- statmod(lrate ~ ridge(xs, standardize = TRUE),
                     distrib = gaussian1_distrib(), data = UScrime)
summary(fit_ridge)
```

The ridge penalty has derivatives of every order, so `statmod()` estimates
its $\lambda$ by REML, in the same way as a smoothing parameter
(@sec-beyond-outer). The block of the ridge term reports
$\hat\lambda = `r fmt(hyper(fit_ridge)$estimate, 1)`$ with a standard error
and an interval. The block spends `r fmt(fit_ridge@edf$edf[2], 2)` effective
degrees of freedom on its 15 coefficients. The ridge shrinks every
coefficient and sets none of them to zero. The standard errors and the tests
of the coefficients use the Bayesian variance of @sec-beyond-variances, as
for the coefficients of a smooth. The summary prints the first ten
coefficients of the block, and `coef(fit_ridge)` gives all of them.

The package `mgcv` fits the same model with the argument `paraPen`, which
puts a penalty on the coefficients of a matrix in the formula. Its penalty is
$\theta\, u^\top u$, added to the residual sum of squares. This is our
penalty multiplied by $2\sigma^2$, so its smoothing parameter is
$\theta = \sigma^2\lambda$:

```{r}
#| label: mgcv-ridge
UScrime$Z <- scale(X)
g_ridge <- mgcv::gam(lrate ~ Z, data = UScrime,
                     paraPen = list(Z = list(diag(15))), method = "REML")
c(statmod = hyper(fit_ridge)$estimate, mgcv = g_ridge$sp / g_ridge$sig2)
max(abs(fitted(fit_ridge) - fitted(g_ridge)))
c(statmod = fit_ridge@criterion, mgcv = -g_ridge$gcv.ubre)
```

The two packages give the same $\lambda$, the same fitted values and the
same REML criterion.

### The lasso penalty

```{r}
#| label: fit-lasso
fit_lasso <- statmod(lrate ~ lasso(xs, standardize = TRUE),
                     distrib = gaussian1_distrib(), data = UScrime)
summary(fit_lasso)
```

```{r}
#| label: lasso-numbers
#| include: false
lam_l <- hyper(fit_lasso)$estimate
sig_l <- exp(coef(fit_lasso)$sigma[[1]])
b_l <- coef(fit_lasso)$mu[-1]
keep_l <- colnames(X)[b_l != 0]
zero_l <- colnames(X)[b_l == 0]
```

The lasso penalty has a kink at zero, where its derivative does not exist.
REML needs the second derivative, so `statmod()` cannot use it here.
Instead, `statmod()` fits the model at a sequence of values of $\lambda$. The
sequence starts at the value that sets every coefficient to zero and goes
down to a small value. `statmod()` then keeps the value with the smallest
BIC (@sec-penalized-hyper). The block reports
$\hat\lambda = `r fmt(lam_l, 2)`$ with the label `[bic]` and without a
standard error, because $\hat\lambda$ is one value of a finite set and not
the root of a derivative. The lasso keeps `r length(keep_l)` covariates and
sets `r length(zero_l)` coefficients exactly to zero: `r code(zero_l)`. It
keeps `Po1` and drops `Po2`, while the ridge shares the effect between the
two. The effective degrees of freedom of the block are the number of
coefficients that are not zero [@zou2007].

The standard errors of the coefficients that are not zero are those of a
model in which we knew that the other coefficients are zero. They ignore the
selection, so they are too small, and the intervals cover less than 95% of
the time. The line `zeros` checks the condition for a coefficient to be
zero. At a zero coefficient, the derivative of the log-likelihood must be at
most $\lambda s_j$ in absolute value, so the ratio on that line must be at
most one.

The package `glmnet` fits the lasso by coordinate descent [@friedman2010]. It
minimizes
$\tfrac{1}{2n}\sum_i (y_i - \mu_i)^2 + \lambda_g \sum_j \tilde s_j
\lvert\beta_j\rvert$. Here $\tilde s_j$ is the standard deviation of
covariate $j$, computed with the divisor $n$. With $\sigma$ at its estimate,
our objective in the coefficients is
$\tfrac{1}{2\sigma^2}\sum_i (y_i - \mu_i)^2 + \lambda\sum_j s_j\lvert
\beta_j\rvert$. If we multiply it by $\sigma^2/n$, we obtain the objective of
`glmnet` with $\lambda_g = \lambda\sigma^2/n \cdot s_j/\tilde s_j =
\lambda\sigma^2/\sqrt{n(n-1)}$:

```{r}
#| label: glmnet-lasso
n <- nrow(UScrime)
lam_g <- hyper(fit_lasso)$estimate * exp(2 * coef(fit_lasso)$sigma[[1]]) /
  sqrt(n * (n - 1))
g_lasso <- glmnet::glmnet(x = X, y = UScrime$lrate, lambda = lam_g,
                          control = list(thresh = 1e-14))
cbind(statmod = coef(fit_lasso)$mu, glmnet = as.numeric(coef(g_lasso)))
```

The two fits agree to
`r format(signif(max(abs(coef(fit_lasso)$mu - as.numeric(coef(g_lasso)))), 1))`.

### The elastic net

```{r}
#| label: fit-enet
fit_enet <- statmod(lrate ~ enet(xs, standardize = TRUE),
                    distrib = gaussian1_distrib(), data = UScrime)
hyper(fit_enet)[, c("name", "estimate", "source")]
```

```{r}
#| label: enet-numbers
#| include: false
h_e <- hyper(fit_enet)
lam_e <- h_e$estimate[h_e$name == "lambda"]
alp_e <- h_e$estimate[h_e$name == "alpha"]
same_e <- identical(coef(fit_enet)$mu[-1] != 0, coef(fit_lasso)$mu[-1] != 0)
n_e <- sum(coef(fit_enet)$mu[-1] != 0)
zero_e <- colnames(X)[coef(fit_enet)$mu[-1] == 0]
edf_e <- fit_enet@edf$edf[2]
```

The elastic net has two hyperparameters, and `statmod()` chooses both of them
by BIC over a grid of values (@sec-penalized-hyper). It chooses
$\hat\lambda = `r fmt(lam_e, 2)`$ and $\hat\alpha = `r fmt(alp_e, 3)`$. With
this value of $\alpha$, most of the penalty comes from its ridge part. The
fit keeps `r n_e` covariates, more than the lasso, and sets only
`r code(zero_e)` to zero. The block spends `r fmt(edf_e, 2)` effective
degrees of freedom on these `r n_e` coefficients. The number is smaller than
`r n_e` because the quadratic part of the penalty also shrinks the
coefficients that are not zero. Let $A$ be the set of these coefficients,
let $H$ be the information of the log-likelihood and let $S$ be the second
derivative of the penalty. The effective degrees of freedom of the block are
$\operatorname{tr}\{(H_{AA} + S_{AA})^{-1} H_{AA}\}$. For the lasso,
$S_{AA} = 0$, and the trace equals the number of coefficients in $A$.

@fig-crime shows the standardized coefficients $s_j\hat\beta_j$ of the three
fits and of least squares.

```{r}
#| label: fig-crime
#| fig-cap: "Standardized coefficients $s_j\\hat\\beta_j$ of the fifteen covariates of the `UScrime` data under least squares, the ridge, the lasso and the elastic net."
#| echo: false
#| fig-height: 5.2
sdx <- apply(X, 2, sd)
ols <- stats::lm(UScrime$lrate ~ X)
std <- cbind(ols = coef(ols)[-1] * sdx, ridge = coef(fit_ridge)$mu[-1] * sdx,
             lasso = coef(fit_lasso)$mu[-1] * sdx,
             enet = coef(fit_enet)$mu[-1] * sdx)
rownames(std) <- colnames(X)
meth <- c(ols = "least squares", ridge = "ridge", lasso = "lasso",
          enet = "elastic net")
std_df <- data.frame(
  covariate = factor(rep(rownames(std), times = ncol(std)),
                     levels = rev(rownames(std))),
  method = factor(rep(meth[colnames(std)], each = nrow(std)), levels = meth),
  value = as.vector(std))
ggplot(std_df, aes(x = value, y = covariate, colour = method,
                   shape = method)) +
  geom_vline(xintercept = 0, linetype = 2, colour = grey) +
  geom_point(size = 2.2, stroke = 0.9) +
  scale_colour_manual(values = c(ink, grey, chalk, rust)) +
  scale_shape_manual(values = c(1, 16, 2, 4)) +
  labs(x = "standardized coefficient", y = NULL)
```

### A matrix as input

The first argument of a penalized term is a one-sided formula, as above, or a
matrix. We add the matrix to the data frame as a single column with `$`,
because `data.frame(X = X)` would split it into 15 columns. The names of its
columns become the names of the coefficients:

```{r}
#| label: fit-matrix
UScrime$X <- X
fit_mat <- statmod(lrate ~ lasso(X, standardize = TRUE),
                   distrib = gaussian1_distrib(), data = UScrime)
max(abs(coef(fit_mat)$mu - coef(fit_lasso)$mu))
```

A matrix is useful when there are many covariates, or when the covariates
come from another computation and have no formula. `statmod()` keeps a
sparse matrix of the package `Matrix` sparse.

### The scale of the covariates

The penalty treats all the coefficients in the same way. Without the factor
$s_j$, the fit would therefore depend on the units of the covariates. The
covariates of `UScrime` have been rescaled to convenient numbers
[@venables2002]. We change the units of two of them: we multiply the
population `Pop` by 1000 and the gross domestic product `GDP` by 10, in a
second formula of the covariates. We then fit the lasso at the same
$\lambda$ with and without `standardize`:

```{r}
#| label: fit-units
xs_units <- ~ M + So + Ed + Po1 + Po2 + LF + M.F + I(1000 * Pop) + NW + U1 +
  U2 + I(10 * GDP) + Ineq + Prob + Time
fit_s1 <- statmod(lrate ~ lasso(xs, standardize = TRUE, lambda = 17),
                  distrib = gaussian1_distrib(), data = UScrime)
fit_s2 <- statmod(lrate ~ lasso(xs_units, standardize = TRUE, lambda = 17),
                  distrib = gaussian1_distrib(), data = UScrime)
fit_r1 <- statmod(lrate ~ lasso(xs, lambda = 17),
                  distrib = gaussian1_distrib(), data = UScrime)
fit_r2 <- statmod(lrate ~ lasso(xs_units, lambda = 17),
                  distrib = gaussian1_distrib(), data = UScrime)
c(standardized = max(abs(fitted(fit_s1) - fitted(fit_s2))),
  raw = max(abs(fitted(fit_r1) - fitted(fit_r2))))
```

With `standardize = TRUE` the two fits are the same. Without it, the fitted
values change by up to `r fmt(max(abs(fitted(fit_r1) - fitted(fit_r2))), 3)`
on the log scale. When we multiply a covariate by $c$, its coefficient is
divided by $c$, and so is the penalty on that coefficient. In the new units
the penalty on `Pop` is therefore a thousand times smaller, and the penalty on
`GDP` ten times smaller. The default of `standardize` is `FALSE`, because a
block of indicators, or a block of covariates on one scale, does not need it.

## SCAD and MCP {#sec-penalized-scad}

The lasso shrinks every coefficient that it keeps, the large ones as much as
the small ones, so these coefficients are biased towards zero. The SCAD
penalty [@fan2001] and the MCP penalty [@zhang2010] grow like the lasso
penalty near zero, but they stop growing beyond a certain size of the
coefficient. They therefore do not shrink a large coefficient. They are
defined by their derivatives for $t \ge 0$:
$$
p'_{\text{SCAD}}(t) = \lambda\Big\{1(t \le \lambda)
  + \frac{(a\lambda - t)_+}{(a - 1)\lambda}\, 1(t > \lambda)\Big\},
\qquad
p'_{\text{MCP}}(t) = \Big(\lambda - \frac{t}{\gamma}\Big)_+ ,
$$
with $a > 2$ and $\gamma > 1$. The derivative of SCAD equals $\lambda$, as
for the lasso, up to $t = \lambda$. It then falls linearly to zero at
$t = a\lambda$, and it is zero beyond $a\lambda$. The derivative of MCP falls
linearly from $\lambda$ at $t = 0$ to zero at $t = \gamma\lambda$. Neither
penalty is minus the log of a density: both are constant for large
coefficients, so their exponential has no finite integral.

The left panel of @fig-shapes shows the three penalties. The right panel
shows their effect in the simplest problem. For one observation $z$, the
estimate $\hat\theta$ minimizes $\tfrac12 (z - \theta)^2 + p(\lvert\theta\rvert)$.
The lasso sets $\hat\theta$ to zero for $\lvert z\rvert \le \lambda$ and
moves every other $z$ towards zero by $\lambda$. SCAD and MCP also set small
values to zero, but they leave $z$ unchanged beyond $a\lambda$ and
$\gamma\lambda$.

```{r}
#| label: fig-shapes
#| fig-cap: "Left: the lasso, SCAD ($a = 3.7$) and MCP ($\\gamma = 3$) penalties at $\\lambda = 1$. Right: the estimate $\\hat\\theta$ that minimizes $\\frac12 (z - \\theta)^2 + p(\\lvert\\theta\\rvert)$ under each penalty, with the identity as a dotted line."
#| echo: false
#| fig-height: 3.6
tt <- seq(from = 0, to = 5, length.out = 400)
pen_lasso <- tt
pen_scad <- ifelse(tt <= 1, tt,
                   ifelse(tt <= 3.7, (2 * 3.7 * tt - tt^2 - 1) / (2 * 2.7), 4.7 / 2))
pen_mcp <- ifelse(tt <= 3, tt - tt^2 / 6, 1.5)
zz <- seq(from = -5, to = 5, length.out = 801)
soft <- sign(zz) * pmax(abs(zz) - 1, 0)
th_scad <- ifelse(abs(zz) <= 2, soft,
                  ifelse(abs(zz) <= 3.7, (2.7 * zz - sign(zz) * 3.7) / 1.7, zz))
th_mcp <- ifelse(abs(zz) <= 3, soft / (1 - 1 / 3), zz)
pan_shp <- c('"penalty"~p(abs(theta))~"against"~abs(theta)',
             'hat(theta)~"against"~z')
pen_names <- c("lasso", "SCAD", "MCP")
shp_df <- rbind(
  data.frame(panel = pan_shp[1], x = tt, y = c(pen_lasso, pen_scad, pen_mcp),
             penalty = rep(pen_names, each = length(tt))),
  data.frame(panel = pan_shp[2], x = zz, y = c(soft, th_scad, th_mcp),
             penalty = rep(pen_names, each = length(zz))))
shp_df$panel <- factor(shp_df$panel, levels = pan_shp)
shp_df$penalty <- factor(shp_df$penalty, levels = pen_names)
id_line <- data.frame(panel = factor(pan_shp[2], levels = pan_shp))
ggplot(shp_df, aes(x = x, y = y)) +
  geom_abline(data = id_line, aes(intercept = 0, slope = 1), linetype = 3,
              colour = grey) +
  geom_line(aes(colour = penalty, linetype = penalty), linewidth = 0.8) +
  facet_wrap(~panel, nrow = 1, scales = "free", labeller = label_parsed) +
  scale_colour_manual(values = c(chalk, rust, ink)) +
  scale_linetype_manual(values = c(1, 2, 4)) +
  labs(x = NULL, y = NULL)
```

The two authors define the shape parameters $a$ and $\gamma$ for the problem
of the right panel, in which the loss $\tfrac12 (z - \theta)^2$ has
curvature one. A regression is different. Fix every coefficient except
$\beta_j$ at its estimate. As a function of $\beta_j$ alone, minus the
log-likelihood is then close to a parabola
$\tfrac12 c_j (\beta_j - z_j)^2$ plus a constant. Here $z_j$ is the value of
$\beta_j$ at the minimum of the parabola, that is, the estimate of $\beta_j$
without the penalty when the other coefficients are fixed. The curvature of
the parabola is
$$
c_j = \sum_{i=1}^{n} w_i (x_{ij} - \bar x_j)^2 ,
$$
where $w_i$ is the information of observation $i$ in the equation that
contains $\beta_j$. For a gaussian mean, $w_i = 1/\sigma^2$. So $c_j$ plays
the role of the curvature one in the problem of the right panel.
`statmodels7` therefore uses the penalty
$c_j\, p_{\lambda/c_j}(\lvert\beta_j\rvert)$, with $c_j$ computed at the
fitted coefficients. Under `standardize = TRUE` the penalty acts on
$s_j\beta_j$, and $c_j$ is divided by $s_j^2$. The derivative of this penalty
at zero is $\lambda$, as for the lasso. The penalty stops growing at
$\lvert\beta_j\rvert = a\lambda/c_j$, and the lasso threshold of the same
problem is $\lambda/c_j$. The ratio between the two points is $a$ for every
coefficient, whatever the units of the covariate and whatever the
distribution. With this scaling the defaults of the two authors, $a = 3.7$
and $\gamma = 3$, have the meaning that they have in their papers.

We fit SCAD and MCP with their default shape parameters:

```{r}
#| label: fit-scad
fit_scad <- statmod(lrate ~ scad(xs, standardize = TRUE),
                    distrib = gaussian1_distrib(), data = UScrime)
summary(fit_scad)
```

```{r}
#| label: fit-mcp
fit_mcp <- statmod(lrate ~ mcp(xs, standardize = TRUE),
                   distrib = gaussian1_distrib(), data = UScrime)
summary(fit_mcp)
```

```{r}
#| label: scad-numbers
#| include: false
on_s <- coef(fit_scad)$mu[-1] != 0
keep_s <- colnames(X)[on_s]
on_m <- coef(fit_mcp)$mu[-1] != 0
keep_m <- colnames(X)[on_m]
only_l <- setdiff(keep_l, keep_s)
only_s <- setdiff(keep_s, keep_l)
lam_mcp <- hyper(fit_mcp)$estimate[1]
c_mcp <- (n - 1) / exp(2 * coef(fit_mcp)$sigma[[1]])
edf_s <- fit_scad@edf$edf[2]
edf_m <- fit_mcp@edf$edf[2]
```

In both summaries the shape parameter has the label `[fixed]`, because we
did not ask `statmod()` to estimate it. Only $\lambda$ is chosen by BIC:
$\hat\lambda = `r fmt(hyper(fit_scad)$estimate[1], 2)`$ for SCAD and
$\hat\lambda = `r fmt(lam_mcp, 2)`$ for MCP. SCAD keeps `r length(keep_s)`
covariates. Compared with the lasso, it drops `r code(only_l)` and adds
`r code(only_s)`. MCP keeps `r length(keep_m)` covariates: `r code(keep_m)`.
The SCAD block spends `r fmt(edf_s, 2)` effective degrees of freedom on its
`r length(keep_s)` coefficients. The MCP block spends `r fmt(edf_m, 2)` on
its `r length(keep_m)` coefficients, which is more than one per
coefficient. We explain the reason below. The estimates of `Po1`, `Ed`,
`Ineq` and `M` are similar in the two fits, and the coefficient of `Prob`
is less certain under MCP than under SCAD.

The effective degrees of freedom of MCP depend on where each coefficient
lies. Under `standardize = TRUE` and a gaussian mean, $c_j/s_j^2$ equals
$(n-1)/\sigma^2$ for every covariate. On the scale of $u_j = s_j\beta_j$,
the MCP penalty therefore stops growing at
$\gamma\hat\lambda\hat\sigma^2/(n-1) = `r fmt(3 * lam_mcp / c_mcp, 4)`$.
Beyond this point the derivative of the penalty is zero. At the fit, the
derivative of the log-likelihood in a coefficient beyond this point must then
be zero too, as in least squares. Inside the point, the derivative of the
log-likelihood equals the derivative of the penalty, which on this scale is
$\lambda - c\lvert u_j\rvert/\gamma$ with $c = (n-1)/\sigma^2$. We compute
both derivatives for the six coefficients that MCP keeps:

```{r}
#| label: mcp-score
u_m <- coef(fit_mcp)$mu[-1] * apply(X, 2, sd)
r_m <- UScrime$lrate - fitted(fit_mcp)
score_u <- crossprod(scale(X, center = TRUE, scale = FALSE), r_m) /
  exp(2 * coef(fit_mcp)$sigma[[1]]) / apply(X, 2, sd)
pen_u <- pmax(lam_mcp - c_mcp * abs(u_m) / 3, 0) * sign(u_m)
round(cbind(u = u_m, score = score_u[, 1], penalty = pen_u)[on_m, ], 4)
```

Four coefficients lie beyond the point. Their score is zero, so MCP does not
shrink them at all. The coefficients of
`r code(keep_m[abs(u_m[on_m]) < 3 * lam_mcp / c_mcp])` lie inside the
point. Their score equals the derivative of the penalty, so MCP shrinks
them. In this region the second derivative of the penalty is negative, so
the trace $\operatorname{tr}\{(H_{AA} + S_{AA})^{-1} H_{AA}\}$ counts each of
these two coefficients as more than one parameter. This is why the MCP block
spends more effective degrees of freedom than it has coefficients. For SCAD
the corresponding point is $a\lambda/c_j$ with $a = 3.7$, further from zero,
and on these data no SCAD coefficient lies in the region where its penalty
bends, so its effective degrees of freedom equal its number of coefficients.

The package `ncvreg` fits SCAD and MCP by coordinate descent along a sequence
of values of $\lambda$ [@breheny2011]. It minimizes
$\tfrac{1}{2n}\sum_i (y_i - \mu_i)^2 + \sum_j p_{\lambda_n}(\lvert
v_j\rvert)$, where $v_j = \tilde s_j\beta_j$ and the columns are standardized
with the divisor $n$. Under `standardize = TRUE` and a gaussian mean,
$c_j = (n-1)/\sigma^2$. If we multiply our objective by $\sigma^2/n$ and
write it in terms of $v_j$, we obtain the objective of `ncvreg` with
$\lambda_n = \lambda\sigma^2/\sqrt{n(n-1)}$ and the same $a$ or $\gamma$:

```{r}
#| label: ncvreg-scad
s_n <- apply(X, 2, sd) * sqrt((n - 1) / n)
Zn <- scale(X, center = TRUE, scale = s_n)
yc <- UScrime$lrate - mean(UScrime$lrate)
lam_s <- hyper(fit_scad)$estimate[1] * exp(2 * coef(fit_scad)$sigma[[1]]) /
  sqrt(n * (n - 1))
ncv_scad <- ncvreg::ncvfit(X = Zn, y = yc, penalty = "SCAD", gamma = 3.7,
                           lambda = lam_s, eps = 1e-12, max.iter = 1e6)
max(abs(ncv_scad$beta / s_n - coef(fit_scad)$mu[-1]))
```

The two SCAD fits agree. SCAD and MCP are not convex, so their objective
can have several local minima. A coordinate descent can reach a different
minimum when it starts from a different point. `statmod()` fits each value of
$\lambda$ starting from the fit at the previous, larger value. We compute the
MCP fit of `ncvreg` in two ways: along a sequence of values of $\lambda$ that
ends at ours, as `statmod()` does, and with `ncvfit()` started from zero.

```{r}
#| label: ncvreg-mcp
lam_m <- hyper(fit_mcp)$estimate[1] * exp(2 * coef(fit_mcp)$sigma[[1]]) /
  sqrt(n * (n - 1))
lam_top <- max(abs(crossprod(Zn, yc))) / n
lam_seq <- exp(seq(from = log(lam_top), to = log(lam_m), length.out = 100))
ncv_path <- ncvreg::ncvreg(X = Zn, y = yc, penalty = "MCP", gamma = 3,
                           lambda = lam_seq, eps = 1e-12, max.iter = 1e6)
ncv_zero <- ncvreg::ncvfit(X = Zn, y = yc, penalty = "MCP", gamma = 3,
                           lambda = lam_m, eps = 1e-12, max.iter = 1e6)
v_mcp <- coef(fit_mcp)$mu[-1] * s_n
round(cbind(statmod = v_mcp, path = ncv_path$beta[-1, 100],
            from_zero = ncv_zero$beta), 4)
```

```{r}
#| label: ncvreg-obj
#| include: false
obj_mcp <- function(v) {
  sum((yc - Zn %*% v)^2) / (2 * n) +
    sum(ifelse(abs(v) <= 3 * lam_m, lam_m * abs(v) - v^2 / 6, 1.5 * lam_m^2))
}
o_ours <- obj_mcp(v_mcp)
o_zero <- obj_mcp(ncv_zero$beta)
```

On these data both computations of `ncvreg` reach our estimates. On other
data the computation from zero can end in a different local minimum. Neither
package searches for a lower minimum than the one that its sequence of
starting points reaches.

### The shape parameter

The shape parameter is a hyperparameter like $\lambda$. If we give it one
number, as the default does, `statmod()` keeps it fixed at that value. If we
give it several numbers, `statmod()` adds them to the grid of the path. If we
give `NULL`, `statmod()` builds a grid of values itself. We give SCAD four
values of $a$:

```{r}
#| label: fit-scad-a
fit_scad_a <- statmod(lrate ~ scad(xs, standardize = TRUE,
                                   a = c(2.5, 3.7, 8, 30)),
                      distrib = gaussian1_distrib(), data = UScrime)
hyper(fit_scad_a)[, c("name", "estimate", "source")]
```

```{r}
#| label: scad-a-numbers
#| include: false
pa <- fit_scad_a@history$outer
best_a <- tapply(pa$criterion, factor(pa$setting, levels = unique(pa$setting)), min)
```

For each of the four values of $a$, the path runs over 25 values of
$\lambda$, so `statmod()` fits the model `r nrow(pa)` times. The smallest BIC
for each value of $a$ is

```{r}
#| label: scad-a-best
round(best_a, 3)
```

The smallest of these is at
$a = `r fmt(hyper(fit_scad_a)$estimate[2], 1)`$, the default. The values 2.5
and 3.7 give nearly the same BIC. The larger values of $a$ make SCAD closer
to the lasso, so they shrink the large coefficients more, and they give a
larger BIC.

## The hyperparameters {#sec-penalized-hyper}

If a penalty has derivatives of every order, as the ridge does, `statmod()`
estimates its hyperparameter by REML, together with the other smooth
hyperparameters of the model (@sec-beyond-outer). If a penalty has a kink,
`statmod()` chooses its hyperparameters along a path. The path starts at the
value of $\lambda$ that sets every coefficient of the block to zero. It goes
down geometrically to a fraction `min_ratio` of this value, `1e-4` by
default, in `n_lambda` steps, 25 by default. At each value of $\lambda$,
`statmod()` fits the model starting from the fit at the previous value. It
then computes the criterion that we set with the argument `sparse_criterion`
of `statmod()`, BIC by default. In the end it keeps the value with the
smallest criterion. The component `history$outer` of the fit contains the
path:

```{r}
#| label: path-lasso
fit_lasso@history$outer[1:10, c("value", "criterion")]
```

At the first value of the path every coefficient of the block is zero, and
the first covariate enters at the second value. @fig-path shows the
coefficients and the BIC along the whole path.

```{r}
#| label: path-coef
#| include: false
# the coefficients and sigma at each value of a path, each fit started from
# the previous one as statmod() does
path_coefs <- function(lams) {
  B <- matrix(NA_real_, length(lams), ncol(X),
              dimnames = list(NULL, colnames(X)))
  sg <- numeric(length(lams))
  prev <- NULL
  for (k in seq_along(lams)) {
    st <- if (is.null(prev)) start_intercepts() else start_from(prev)
    f_k <- statmod(lrate ~ lasso(xs, standardize = TRUE, lambda = lams[k]),
                   distrib = gaussian1_distrib(), data = UScrime, start = st)
    B[k, ] <- coef(f_k)$mu[-1] * apply(X, 2, sd)
    sg[k] <- exp(coef(f_k)$sigma[[1]])
    prev <- f_k
  }
  list(B = B, sigma = sg)
}
lams <- fit_lasso@history$outer$value
pc25 <- path_coefs(lams)
B_path <- pc25$B
sig_path <- pc25$sigma
n_path <- rowSums(abs(B_path) > 1e-6)
lam_eff <- lams * sig_path^2 / sqrt(n * (n - 1))
i_jump <- which(diff(n_path) == max(diff(n_path)))[1]
```

```{r}
#| label: fig-path
#| fig-cap: "The lasso on the `UScrime` data along its path of 25 values of $\\lambda$. Top: the standardized coefficients $s_j\\hat\\beta_j$, with `Po1` and `Prob` in colour. Bottom: BIC. The vertical line marks the chosen value."
#| echo: false
#| fig-height: 6
pan_path <- c("standardized coefficient", "BIC")
on_panel <- function(df, k) {
  df$panel <- factor(pan_path[k], levels = pan_path)
  df
}
path_df <- function(lam, B) {
  nm <- rep(colnames(B), each = nrow(B))
  grp <- ifelse(nm %in% c("Po1", "Prob"), nm, "other covariates")
  on_panel(data.frame(x = log(rep(lam, times = ncol(B))), y = as.vector(B),
                      coefficient = nm,
                      group = factor(grp, levels = c("Po1", "Prob",
                                                     "other covariates"))),
           k = 1)
}
path_plot <- function(cf) {
  ggplot(mapping = aes(x = x, y = y)) +
    geom_hline(data = on_panel(data.frame(y = 0), k = 1),
               aes(yintercept = y), linetype = 3, colour = grey) +
    geom_line(data = cf[cf$group == "other covariates", ],
              aes(group = coefficient, colour = group, linewidth = group)) +
    geom_line(data = cf[cf$group != "other covariates", ],
              aes(group = coefficient, colour = group, linewidth = group)) +
    facet_wrap(~panel, ncol = 1, scales = "free_y", strip.position = "left") +
    scale_colour_manual(values = c(Po1 = chalk, Prob = rust,
                                   "other covariates" = grey),
                        breaks = levels(cf$group)) +
    scale_linewidth_manual(values = c(Po1 = 0.9, Prob = 0.9,
                                      "other covariates" = 0.4),
                           breaks = levels(cf$group)) +
    guides(colour = guide_legend(order = 1),
           linewidth = guide_legend(order = 1)) +
    labs(x = expression(log(lambda)), y = NULL) +
    theme(strip.placement = "outside", strip.background = element_blank(),
          strip.text = element_text(size = rel(1)))
}
bic_25 <- on_panel(data.frame(x = log(lams),
                              y = fit_lasso@history$outer$criterion,
                              path = "25 values"), k = 2)
path_plot(path_df(lams, B_path)) +
  geom_vline(xintercept = log(lam_l), linetype = 2, colour = ink) +
  geom_line(data = bic_25, colour = ink) +
  geom_point(data = bic_25, colour = ink, size = 1.2)
```

Between $\lambda = `r fmt(lams[i_jump], 1)`$ and
$\lambda = `r fmt(lams[i_jump + 1], 1)`$, which are two neighbouring values
of the path, the number of covariates in the model grows from
`r n_path[i_jump]` to `r n_path[i_jump + 1]`. The cause is $\sigma$, which
`statmod()` estimates together with the coefficients. The comparison with
`glmnet` showed that the selection depends on $\lambda\sigma^2$, not on
$\lambda$ alone. When a covariate enters, $\hat\sigma$ falls. Then
$\lambda\hat\sigma^2$ falls more than $\lambda$, and more covariates can
enter. Between the two values, $\lambda$ falls by a factor of
`r fmt(lams[i_jump] / lams[i_jump + 1], 2)`, and $\lambda\hat\sigma^2$ falls
by a factor of `r fmt(lam_eff[i_jump] / lam_eff[i_jump + 1], 2)`. For the same
reason BIC is not a smooth function of $\lambda$. It jumps by
$\log n = `r fmt(log(n), 2)`$ each time a coefficient enters, and the value
that it chooses depends on where the points of the path fall. We fit the
lasso again with a path of 100 values:

```{r}
#| label: fit-lasso-100
fit_lasso_100 <- statmod(lrate ~ lasso(xs, standardize = TRUE, n_lambda = 100),
                         distrib = gaussian1_distrib(), data = UScrime)
c(n25 = hyper(fit_lasso)$estimate, n100 = hyper(fit_lasso_100)$estimate)
c(n25 = BIC(fit_lasso), n100 = BIC(fit_lasso_100))
c(n25 = sum(coef(fit_lasso)$mu[-1] != 0),
  n100 = sum(coef(fit_lasso_100)$mu[-1] != 0))
```

```{r}
#| label: path-coef-100
#| include: false
lams_100 <- fit_lasso_100@history$outer$value
pc100 <- path_coefs(lams_100)
lam_100 <- hyper(fit_lasso_100)$estimate
```

@fig-path-100 shows the coefficients along the path of 100 values and the
BIC of both paths. The path of 100 values visits values between the 25
points, and some of these values have a smaller BIC. The finer path chooses
$\hat\lambda = `r fmt(lam_100, 2)`$ with a smaller BIC and the same number of
covariates. Both fits are reasonable. The difference shows that the chosen
$\lambda$ is uncertain: many values have nearly the same criterion.

```{r}
#| label: fig-path-100
#| fig-cap: "The lasso on the `UScrime` data along a path of 100 values of $\\lambda$. Top: the standardized coefficients, with `Po1` and `Prob` in colour. Bottom: BIC along the path of 100 values (line) and of 25 values (points). The vertical lines mark the values chosen by the two paths."
#| echo: false
#| fig-height: 6
bic_100 <- on_panel(data.frame(x = log(lams_100),
                               y = fit_lasso_100@history$outer$criterion,
                               path = "100 values"), k = 2)
path_plot(path_df(lams_100, pc100$B)) +
  geom_vline(xintercept = log(lam_l), linetype = 2, colour = ink) +
  geom_vline(xintercept = log(lam_100), linetype = 3, colour = ink) +
  geom_line(data = bic_100, aes(linetype = path), colour = ink) +
  geom_point(data = bic_25, aes(shape = path), colour = ink, size = 1.2) +
  scale_linetype_manual(values = 1, guide = guide_legend(order = 2)) +
  scale_shape_manual(values = 16, guide = guide_legend(order = 3))
```

### Held and written values

If we give the hyperparameter one number, `statmod()` holds it fixed at that
value. If we give it several numbers, `statmod()` uses them as the values of
the path:

```{r}
#| label: fit-held
fit_held <- statmod(lrate ~ lasso(xs, standardize = TRUE, lambda = 40),
                    distrib = gaussian1_distrib(), data = UScrime)
hyper(fit_held)[, c("name", "estimate", "held", "source")]
fit_grid <- statmod(lrate ~ lasso(xs, standardize = TRUE,
                                  lambda = c(5, 10, 20, 40, 80)),
                    distrib = gaussian1_distrib(), data = UScrime)
fit_grid@history$outer[, c("value", "criterion")]
hyper(fit_grid)[, c("name", "estimate", "held", "source")]
```

The fixed value has the source `fixed`, and at $\lambda = 40$ the fit keeps
`r sum(coef(fit_held)$mu[-1] != 0)` covariate. `statmod()` visits the
written values from the largest to the smallest and keeps the value with the
smallest BIC, $\lambda = `r fmt(hyper(fit_grid)$estimate, 0)`$.

### Two hyperparameters

The elastic net has two hyperparameters. By default `statmod()` builds a path
of 25 values of $\lambda$ for each of 5 values of $\alpha$, and it computes
the criterion at all 125 pairs (`search = "grid"`). For each $\alpha$, the
path of $\lambda$ starts at the value that sets every coefficient to zero.
This value is $\lambda_{\max}/\alpha$, where $\lambda_{\max}$ is the starting
value of the lasso, so the paths of different values of $\alpha$ cover
different ranges. With `search = "cyclic"`, `statmod()` instead changes one
hyperparameter at a time, with the other fixed, until neither changes:

```{r}
#| label: fit-enet-cyclic
fit_enet_c <- statmod(lrate ~ enet(xs, standardize = TRUE, search = "cyclic"),
                      distrib = gaussian1_distrib(), data = UScrime)
c(grid = nrow(fit_enet@history$outer), cyclic = nrow(fit_enet_c@history$outer))
rbind(grid = hyper(fit_enet)$estimate, cyclic = hyper(fit_enet_c)$estimate)
c(grid = BIC(fit_enet), cyclic = BIC(fit_enet_c))
```

The cyclic search computes the criterion at
`r nrow(fit_enet_c@history$outer)` pairs, against the
`r nrow(fit_enet@history$outer)` pairs of the grid, and it stops at a pair
with a larger BIC. The size of the kink of the elastic net penalty is
$\lambda\alpha$, so the criterion has a valley along the curves where
$\lambda\alpha$ is constant. A search that changes one coordinate at a time
moves along such a valley in small steps. The grid is therefore the safer
default. The arguments `n_lambda` and `n_alpha` set the size of the grid.

## Penalties on sigma and nu {#sec-penalized-sigma}

Every parameter of the distribution can have its own equation, and a
penalized term can enter any of these equations. We simulate 300
observations with eight covariates. The mean depends on the first two
covariates, and the standard deviation depends on the third and the fourth:
$$
y_i \sim \mathrm{N}(\mu_i, \sigma_i), \qquad
\mu_i = 1 + x_{i1} - 0.5\,x_{i2}, \qquad
\log\sigma_i = -0.5 + 0.6\,x_{i3} - 0.3\,x_{i4}.
$$

```{r}
#| label: sim-sigma
set.seed(10)
sim <- as.data.frame(matrix(rnorm(300 * 8), nrow = 300, ncol = 8))
names(sim) <- paste0("x", 1:8)
sim$y <- rnorm(300, mean = 1 + sim$x1 - 0.5 * sim$x2,
               sd = exp(-0.5 + 0.6 * sim$x3 - 0.3 * sim$x4))
xx <- ~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8
```

The covariates are already on one scale, so we do not standardize them. We
fit two models. The first has a lasso on the mean and a constant $\sigma$.
The second has a lasso on the mean and a lasso on $\log\sigma$:

```{r}
#| label: fit-sigma
fit_sig0 <- statmod(y ~ lasso(xx), distrib = gaussian1_distrib(), data = sim)
fit_sig1 <- statmod(y ~ lasso(xx) | sigma ~ lasso(xx),
                  distrib = gaussian1_distrib(), data = sim)
summary(fit_sig1)
```

The summary has a lasso block in each equation, and each block has its own
$\lambda$, chosen along its own path. `statmod()` alternates between the two
paths. It chooses $\lambda$ for the mean with the other $\lambda$ fixed, then
the $\lambda$ of $\log\sigma$ with the first fixed, and it repeats this until
neither value changes. The lasso of $\log\sigma$ keeps `x3` and `x4`, with
coefficients `r fmt(coef(fit_sig1)$sigma[["lasso.x3"]], 3)` and
`r fmt(coef(fit_sig1)$sigma[["lasso.x4"]], 3)` against the true values 0.6
and $-0.3$, and it sets the other six coefficients to zero. The lasso of the
mean keeps only `x1` and `x2`. The model with a constant $\sigma$ also keeps
covariates that are not in the mean:

```{r}
#| label: compare-sigma
round(cbind(constant = coef(fit_sig0)$mu, modelled = coef(fit_sig1)$mu), 3)
c(constant = BIC(fit_sig0), modelled = BIC(fit_sig1))
```

With a constant $\sigma$, the observations with a large variance have the
same weight as the others. To fit them, the lasso keeps
`r code(setdiff(names(which(coef(fit_sig0)$mu[-1] != 0)), c("lasso.x1", "lasso.x2")))`
in the mean. The model of $\sigma$ removes these covariates and lowers BIC by
`r fmt(BIC(fit_sig0) - BIC(fit_sig1), 0)`.

The degrees of freedom $\nu$ of a Student $t$ distribution can also have a
penalized equation. We simulate a response with scale $\exp(-0.5)$ and
$\log\nu_i = 1 + 0.8\,x_{i5}$, so the tails are heavier where $x_5$ is
small, and we put a lasso on $\log\nu$:

```{r}
#| label: fit-nu
sim$yt <- 1 + sim$x1 + exp(-0.5) * rt(300, df = exp(1 + 0.8 * sim$x5))
fit_nu <- statmod(yt ~ x1 | nu ~ lasso(xx),
                  distrib = student_t1_distrib(), data = sim)
round(coef(fit_nu)$nu, 3)
```

The lasso keeps `r code(names(which(coef(fit_nu)$nu[-1] != 0)))`, with
coefficient `r fmt(coef(fit_nu)$nu[coef(fit_nu)$nu != 0][-1], 2)` against the
true value 0.8. Like every coefficient that the lasso keeps, this
coefficient is shrunk towards zero. In addition, $\nu$ is hard to estimate
from 300 observations, because only the tails carry information about it.

## A penalty beside smooths and random effects {#sec-penalized-mixed}

A penalized term is one term of the formula, so a model can contain it
together with a smooth and a random effect. The data frame `Boston` gives the
median value of the houses in 506 areas of Boston, with 13 covariates
[@harrison1978]. The proportion of the population of lower status, `lstat`,
has a strong and curved effect on the logarithm of the value. We give `lstat`
a smooth and put a lasso on the other twelve covariates:
$$
\log y_i = \beta_0 + f(\mathrm{lstat}_i) + \sum_{j=1}^{12} \beta_j x_{ij} +
\varepsilon_i .
$$

```{r}
#| label: fit-boston
data(Boston, package = "MASS")
Boston$lmedv <- log(Boston$medv)
bx <- ~ crim + zn + indus + chas + nox + rm + age + dis + rad + tax +
  ptratio + black
fit_b <- statmod(lmedv ~ s(lstat, bspline_smooth(k = 6)) +
                   lasso(bx, standardize = TRUE, n_lambda = 12),
                 distrib = gaussian1_distrib(), data = Boston)
summary(fit_b)
```

The model has two hyperparameters of two kinds. `statmod()` estimates the
smoothing parameter of `s(lstat)` by REML and chooses the $\lambda$ of the
lasso by BIC along its path. At each value of $\lambda$ on the path,
`statmod()` estimates the smoothing parameter again. The path therefore costs
one REML fit for each of its values, and we use a path of 12 values to keep
the computation short. The lasso keeps
`r sum(coef(fit_b)$mu[grepl("^lasso", names(coef(fit_b)$mu))] != 0)` of the
12 covariates and sets
`r code(sub("lasso.", "", names(which(coef(fit_b)$mu[grepl("^lasso", names(coef(fit_b)$mu))] == 0)), fixed = TRUE))`
to zero. The smooth spends `r fmt(fit_b@edf$edf[2], 2)` effective degrees of
freedom on its 5 coefficients. Its free straight line `lin` is the slope of
the linear part of $f$ per standard deviation of `lstat`.

Suppose we fix the smooth and the intercept at their fitted values. The lasso
coefficients then solve a lasso problem in which the smooth and the intercept
are a fixed offset. `glmnet()` accepts an offset, so we can compare the lasso
block with `glmnet`, using the conversion of @sec-penalized-basic:

```{r}
#| label: glmnet-boston
X_b <- as.matrix(model.matrix(fit_b))
is_l <- startsWith(colnames(X_b), "lasso.")
off_b <- drop(X_b[, !is_l] %*% coef(fit_b)$mu[!is_l])
n_b <- nrow(Boston)
lam_b <- hyper(fit_b)$estimate[2] * exp(2 * coef(fit_b)$sigma[[1]]) /
  sqrt(n_b * (n_b - 1))
g_b <- glmnet::glmnet(x = model.matrix(bx, data = Boston)[, -1],
                      y = Boston$lmedv, offset = off_b, lambda = lam_b,
                      intercept = FALSE, control = list(thresh = 1e-14))
max(abs(as.numeric(coef(g_b))[-1] - coef(fit_b)$mu[is_l]))
```

@fig-boston shows the smooth of `lstat`, with the other covariates at their
medians.

```{r}
#| label: fig-boston
#| fig-cap: "Logarithm of the median value of the houses against the proportion of the population of lower status, with the fitted curve and its 95% band, the other covariates at their medians."
#| echo: false
#| fig-height: 3.8
lg <- seq(from = min(Boston$lstat), to = max(Boston$lstat), length.out = 100)
nd_b <- data.frame(lstat = lg)
for (v in all.vars(bx)) nd_b[[v]] <- median(Boston[[v]])
pb <- predict(fit_b, what = "mu", newdata = nd_b, se = TRUE)
band_b <- data.frame(lstat = lg, fit = pb$fit, lower = pb$lower,
                     upper = pb$upper)
ggplot(Boston, aes(x = lstat, y = lmedv)) +
  geom_point(colour = book_grey, size = 0.6) +
  geom_ribbon(data = band_b, aes(x = lstat, ymin = lower, ymax = upper,
                                 fill = "95% band"),
              inherit.aes = FALSE, alpha = 0.2) +
  geom_line(data = band_b, aes(x = lstat, y = fit, colour = "fitted curve"),
            linewidth = 0.9) +
  scale_colour_manual(values = chalk) +
  scale_fill_manual(values = chalk) +
  labs(x = "lstat (%)", y = "log median value")
```

A random effect works in the same way. The data frame `Cars93`, which we
used in @sec-distribs-skew, gives the list price of 93 cars made by 32
manufacturers. The cars of one manufacturer share a price level that their
characteristics do not explain. We therefore model the logarithm of the
price with a lasso on twelve characteristics of the car and a random
intercept for the manufacturer:
$$
\log y_i = \beta_0 + \sum_{j=1}^{12} \beta_j x_{ij} + b_{m(i)} +
\varepsilon_i , \qquad b_m \sim \mathrm{N}(0, \sigma_b) .
$$

```{r}
#| label: fit-cars
data(Cars93, package = "MASS")
Cars93$lprice <- log(Cars93$Price)
cx <- ~ Horsepower + EngineSize + RPM + MPG.city + Fuel.tank.capacity +
  Passengers + Length + Wheelbase + Width + Turn.circle + Weight + Rev.per.mile
fit_c <- statmod(lprice ~ lasso(cx, standardize = TRUE, n_lambda = 12) +
                   random(~ 1 | Manufacturer),
                 distrib = gaussian1_distrib(), data = Cars93)
summary(fit_c)
```

```{r}
#| label: cars-numbers
#| include: false
fit_c0 <- statmod(lprice ~ lasso(cx, standardize = TRUE, n_lambda = 12),
                  distrib = gaussian1_distrib(), data = Cars93)
names_c <- all.vars(cx)
b_c <- coef(fit_c)$mu[paste0("lasso.", names_c)]
b_c0 <- coef(fit_c0)$mu[paste0("lasso.", names_c)]
drop_c <- names_c[b_c == 0 & b_c0 != 0]
sd_b <- hyper(fit_c)$estimate[2]
```

`statmod()` estimates the standard deviation of the random intercept by
REML, $\hat\sigma_b = `r fmt(sd_b, 3)`$, and chooses the $\lambda$ of the
lasso by BIC along its path. At each value of $\lambda$ it estimates
$\sigma_b$ again. On the log scale, this value of $\hat\sigma_b$ means that
the price level of a manufacturer typically differs from the average by a
factor of about $\exp(\hat\sigma_b) = `r fmt(exp(sd_b), 2)`$. The lasso
keeps `r sum(b_c != 0)` of the 12 characteristics. Without the random
intercept it keeps `r sum(b_c0 != 0)`, and these include `r code(drop_c)`.
Part of what these covariates describe is a difference between
manufacturers, and in the model with the random intercept this difference
goes into the random intercept.

As for the smooth, we fix the intercept and the random effects at their
fitted values. The lasso block then solves a lasso problem with an offset,
and we compare it with `glmnet`:

```{r}
#| label: glmnet-cars
X_c <- as.matrix(model.matrix(fit_c))
is_c <- startsWith(colnames(X_c), "lasso.")
off_c <- drop(X_c[, !is_c] %*% coef(fit_c)$mu[!is_c])
n_c <- nrow(Cars93)
lam_c <- hyper(fit_c)$estimate[1] * exp(2 * coef(fit_c)$sigma[[1]]) /
  sqrt(n_c * (n_c - 1))
g_c <- glmnet::glmnet(x = model.matrix(cx, data = Cars93)[, -1],
                      y = Cars93$lprice, offset = off_c, lambda = lam_c,
                      intercept = FALSE, control = list(thresh = 1e-14))
max(abs(as.numeric(coef(g_c))[-1] - coef(fit_c)$mu[is_c]))
```

## Shared hyperparameters {#sec-penalized-id}

The argument `id` of @sec-gamm-id works for every penalized term. Terms with
the same `id` share one value of the hyperparameter. We split the 15
covariates of `UScrime` into two blocks. The first block contains the
economy and the police expenditure, and the second contains the population
and the justice system. We put a ridge on each block:

```{r}
#| label: fit-id
xa <- ~ Po1 + Po2 + LF + U1 + U2 + GDP + Ineq
xb <- ~ M + So + Ed + M.F + Pop + NW + Prob + Time
fit_id <- statmod(lrate ~ ridge(xa, standardize = TRUE, label = "econ", id = "R") +
                    ridge(xb, standardize = TRUE, label = "pop", id = "R"),
                  distrib = gaussian1_distrib(), data = UScrime)
fit_two <- statmod(lrate ~ ridge(xa, standardize = TRUE, label = "econ") +
                     ridge(xb, standardize = TRUE, label = "pop"),
                   distrib = gaussian1_distrib(), data = UScrime)
hyper(fit_id)[, c("term", "estimate")]
hyper(fit_two)[, c("term", "estimate")]
```

Two ridges with one $\lambda$ are the same model as the ridge of
@sec-penalized-basic on all 15 covariates:

```{r}
#| label: id-same
max(abs(fitted(fit_id) - fitted(fit_ridge)))
c(shared = fit_id@criterion, one = fit_ridge@criterion)
```

With two values of $\lambda$, the block of the police and the economy is
shrunk less than the other block. The model with one $\lambda$ is the model
with two values under one equality constraint, inside the range of the
parameters. The two models also have the same coefficients without a
penalty. We can therefore compare their restricted log-likelihoods:

```{r}
#| label: lr-id
lr_id <- 2 * (as.numeric(logLik(fit_two, type = "marginal")) -
                as.numeric(logLik(fit_id, type = "marginal")))
lr_id
pchisq(lr_id, df = 1, lower.tail = FALSE)
```

The data do not reject one $\lambda$ for the two blocks. For a lasso, one
`id` gives one $\lambda$ on one path. The path starts at the value that sets
the coefficients of all the terms with this `id` to zero. Two blocks with one
`id` therefore visit the same values as one block that contains both, and
they choose the same $\lambda$:

```{r}
#| label: fit-lasso-id
fit_lid <- statmod(lrate ~ lasso(xa, standardize = TRUE, label = "econ", id = "L") +
                     lasso(xb, standardize = TRUE, label = "pop", id = "L"),
                   distrib = gaussian1_distrib(), data = UScrime)
c(one = hyper(fit_lasso)$estimate, shared = hyper(fit_lid)$estimate[1])
max(abs(fit_lasso@history$outer$value - fit_lid@history$outer$value))
```

Two lassos without an `id` have two paths. `statmod()` runs them in turn,
each with the other $\lambda$ fixed, until neither value changes:

```{r}
#| label: fit-lasso-two
fit_l2 <- statmod(lrate ~ lasso(xa, standardize = TRUE, label = "econ") +
                    lasso(xb, standardize = TRUE, label = "pop"),
                  distrib = gaussian1_distrib(), data = UScrime)
hyper(fit_l2)[, c("term", "estimate")]
c(one = BIC(fit_lasso), two = BIC(fit_l2))
```

Terms with one `id` can also be in different equations. On the simulated
data of @sec-penalized-sigma, we give the lasso of the mean and the lasso of
$\log\sigma$ one $\lambda$:

```{r}
#| label: fit-id-sigma
fit_sid <- statmod(y ~ lasso(xx, id = "L") | sigma ~ lasso(xx, id = "L"),
                   distrib = gaussian1_distrib(), data = sim)
hyper(fit_sid)[, c("parameter", "estimate")]
c(shared = BIC(fit_sid), separate = BIC(fit_sig1))
```

The path starts at the larger of the two values that set each block to zero,
so at its first value both blocks are zero. The scores of the two equations
have different sizes, so one $\lambda$ for both equations is a strong
constraint. On these data the model with two values of $\lambda$ has the
smaller BIC.

## Choosing the value of lambda {#sec-penalized-choose}

The argument `sparse_criterion` of `statmod()` sets the criterion of the
path. It can be `bic()`, the default, `aic()`, or `cv()`, which computes a
ten-fold cross-validated deviance. AIC and BIC are $-2\ell + \kappa\tau$,
where $\ell$ is the log-likelihood, $\tau$ is the effective number of
parameters, and $\kappa$ is 2 for AIC and $\log n$ for BIC. We fit the lasso
of @sec-penalized-basic with the other two criteria:

```{r}
#| label: fit-crit
fit_aic <- statmod(lrate ~ lasso(xs, standardize = TRUE),
                   distrib = gaussian1_distrib(), data = UScrime,
                   sparse_criterion = aic())
set.seed(1)
fit_cv <- statmod(lrate ~ lasso(xs, standardize = TRUE),
                  distrib = gaussian1_distrib(), data = UScrime,
                  sparse_criterion = cv())
c(bic = hyper(fit_lasso)$estimate, aic = hyper(fit_aic)$estimate,
  cv = hyper(fit_cv)$estimate)
c(bic = sum(coef(fit_lasso)$mu[-1] != 0), aic = sum(coef(fit_aic)$mu[-1] != 0),
  cv = sum(coef(fit_cv)$mu[-1] != 0))
```

```{r}
#| label: crit-numbers
#| include: false
path_b <- fit_lasso@history$outer
path_a <- fit_aic@history$outer
path_c <- fit_cv@history$outer
ok_c <- is.finite(path_c$criterion)
i_min <- which.min(ifelse(ok_c, path_c$criterion, Inf))
n_within <- sum(ok_c & path_c$criterion <= path_c$criterion[i_min] +
                  path_c$se[i_min])
keep_cv <- colnames(X)[coef(fit_cv)$mu[-1] != 0]
ordinal <- function(i) c("first", "second", "third", "fourth", "fifth",
                         "sixth")[i]
```

BIC adds $\log n = `r fmt(log(n), 2)`$ for each parameter and AIC adds 2. The
minimum of AIC therefore lies at a smaller $\lambda$, and AIC keeps more
covariates. Cross-validation chooses the `r ordinal(i_min)` value of the path, and at
this value the model keeps only `r code(keep_cv)`. @fig-crit shows the three
criteria along the path. With `r n` observations, the cross-validated
deviance has a large standard error, and its minimum is poorly determined:
at `r n_within` of the `r nrow(path_c)` values of $\lambda$, the deviance
lies within one standard error of the minimum. The path also contains the
standard errors. The rule of one standard error keeps the largest $\lambda$
whose deviance lies within one standard error of the minimum, and we can
compute it from the path:

```{r}
#| label: one-se
cv_path <- fit_cv@history$outer
best <- which.min(cv_path$criterion)
within <- cv_path$criterion <= cv_path$criterion[best] + cv_path$se[best]
lam_1se <- max(cv_path$value[within])
lam_1se
```

Here the rule gives the `r ordinal(match(lam_1se, cv_path$value))` value of
the path, $\lambda = `r fmt(lam_1se, 2)`$, which is larger than the minimum
at $\lambda = `r fmt(cv_path$value[best], 2)`$: its deviance lies within one
standard error of the minimum, so the rule prefers the simpler model. The
argument `cv(rule = "1se")` applies this rule during the fit.

```{r}
#| label: fig-crit
#| fig-cap: "The three criteria along the path of values of $\\lambda$ for the lasso on the `UScrime` data. Left: BIC and AIC. Right: the cross-validated deviance with a band of one standard error. The vertical lines mark the chosen values."
#| echo: false
#| fig-height: 3.6
ok_b <- is.finite(path_b$criterion); ok_a <- is.finite(path_a$criterion)
x_c <- log(path_c$value[ok_c]); y_c <- path_c$criterion[ok_c]
s_c <- path_c$se[ok_c]
pan_crit <- c("criterion", "cross-validated deviance")
crit_names <- c("BIC", "AIC", "deviance")
n_crit <- c(sum(ok_b), sum(ok_a), length(x_c))
crit_df <- data.frame(
  panel = factor(rep(pan_crit[c(1, 1, 2)], times = n_crit), levels = pan_crit),
  criterion = factor(rep(crit_names, times = n_crit), levels = crit_names),
  x = c(log(path_b$value[ok_b]), log(path_a$value[ok_a]), x_c),
  y = c(path_b$criterion[ok_b], path_a$criterion[ok_a], y_c))
se_df <- data.frame(panel = factor(pan_crit[2], levels = pan_crit), x = x_c,
                    lower = y_c - s_c, upper = y_c + s_c)
chosen <- data.frame(
  panel = factor(pan_crit[c(1, 1, 2)], levels = pan_crit),
  criterion = factor(crit_names, levels = crit_names),
  x = log(c(hyper(fit_lasso)$estimate, hyper(fit_aic)$estimate,
            hyper(fit_cv)$estimate)))
ggplot(crit_df, aes(x = x, y = y)) +
  geom_ribbon(data = se_df, aes(x = x, ymin = lower, ymax = upper,
                                fill = "one standard error"),
              inherit.aes = FALSE, alpha = 0.2) +
  geom_vline(data = chosen, aes(xintercept = x, colour = criterion),
             linetype = 2, show.legend = FALSE) +
  geom_line(aes(colour = criterion)) +
  geom_point(aes(colour = criterion, shape = criterion), size = 1.2) +
  facet_wrap(~panel, nrow = 1, scales = "free_y", strip.position = "left") +
  scale_colour_manual(values = c(chalk, rust, ink)) +
  scale_shape_manual(values = c(16, 17, 16)) +
  scale_fill_manual(values = ink) +
  guides(colour = guide_legend(order = 1), shape = guide_legend(order = 1),
         fill = guide_legend(order = 2)) +
  labs(x = expression(log(lambda)), y = NULL) +
  theme(strip.placement = "outside", strip.background = element_blank(),
        strip.text = element_text(size = rel(1)))
```

The three criteria answer different questions, and @sec-choosing compares
them on other models. A prediction criterion treats the chosen $\lambda$ as
known. The standard errors of the coefficients that are not zero do the same.

```{r}
#| label: gate-10
#| include: false
assert_penalized_ok(
  fits = list(ridge = fit_ridge, lasso = fit_lasso, enet = fit_enet,
              mat = fit_mat, s1 = fit_s1, s2 = fit_s2, r1 = fit_r1, r2 = fit_r2,
              scad = fit_scad, mcp = fit_mcp, scad_a = fit_scad_a,
              lasso_100 = fit_lasso_100, held = fit_held, grid = fit_grid,
              enet_c = fit_enet_c, sig0 = fit_sig0, sig1 = fit_sig1, nu = fit_nu,
              boston = fit_b, id = fit_id, two = fit_two, l2 = fit_l2,
              lid = fit_lid, sid = fit_sid, aic = fit_aic, cv = fit_cv,
              cars = fit_c, cars0 = fit_c0),
  ref = list(g_ridge = g_ridge, g_lasso = g_lasso, ncv_scad = ncv_scad,
             ncv_path = ncv_path, ncv_zero = ncv_zero, g_b = g_b, g_c = g_c),
  data = list(UScrime = UScrime, X = X, sim = sim,
              Boston = Boston, xs = xs, bx = bx, Cars93 = Cars93, cx = cx),
  claims = list(keep_l = keep_l, same_e = same_e, keep_s = keep_s,
                n_path = n_path, i_jump = i_jump, lam_eff = lam_eff,
                o_ours = o_ours, o_zero = o_zero, lr_id = lr_id,
                n_within = n_within, path_c = path_c, drop_c = drop_c,
                lam_1se = lam_1se,
                edf_s = edf_s, edf_m = edf_m, lam_100 = lam_100,
                lams_100 = lams_100))
```
