# Break-points {#sec-breakpoints}

```{r}
#| label: setup-11
#| include: false
source("R/_setup.R")
source("R/breakpoint-certificates.R")
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
yr <- function(x, d = 1) formatC(x, format = "f", digits = d)
```

A break-point is a value of a covariate at which a regression changes. The
change can be in the slope, in the level, or in both. The position of the
break-point is a parameter of the model and is estimated with the other
coefficients. In `statmodels7` a break-point is a term of the formula, and
three terms cover the three kinds of change: `seg()` changes the slope,
`jump()` changes the level, and `jseg()` changes both. Like every other term,
a break-point term can sit in the equation of any parameter of the
distribution, and each of its coefficients, the position included, can have
covariates and random effects of its own.

The package `segmented` fits changes of slope in linear, generalized linear
and linear mixed models, and changes of level in linear and generalized
linear models [@muggeo2003; @muggeo2014; @fasola2018]. We compare our fits
with it throughout the chapter.

## A change of slope: seg() {#sec-breakpoints-seg}

The data frame `GAGurine` gives the concentration of the chemical GAG in the
urine of 314 children aged from 0 to 17 years [@venables2002]. The
concentration falls steeply in the first years of life and slowly afterwards.
Let $y_i$ be the logarithm of the concentration and $a_i$ the age. A model with
one change of slope at the age $\psi$ is
$$
y_i \sim \mathrm{N}(\mu_i, \sigma), \qquad
\mu_i = \beta_0 + \beta_1 a_i + \gamma\, (a_i - \psi)_+ ,
$$
where $(u)_+$ is $u$ when $u > 0$ and zero otherwise. The slope is $\beta_1$
before $\psi$ and $\beta_1 + \gamma$ after it. The mean is a continuous
function of $\psi$, and it has a derivative in $\psi$ everywhere except at
the observed ages, so $\psi$ is estimated like any other coefficient and has a
standard error [@muggeo2003].

```{r}
#| label: fit-gag
data(GAGurine, package = "MASS")
GAGurine$lGAG <- log(GAGurine$GAG)
fit_gag <- statmod(lGAG ~ seg(Age), distrib = gaussian1_distrib(),
                   data = GAGurine)
summary(fit_gag)
```

The term `seg(Age)` carries the linear effect $\beta_1 a$ itself, so the
formula does not write `Age` again. The summary gives the slope `beta`, the
change of slope `gamma1` and the position `psi1` with its standard error. We
fit the same model with `segmented`:

```{r}
#| label: segmented-gag
lm_gag <- lm(lGAG ~ Age, data = GAGurine)
s_gag <- segmented::segmented(lm_gag, seg.Z = ~ Age)
s_gag$psi
rss_gag <- sum((GAGurine$lGAG - fitted(fit_gag, what = "mu"))^2)
c(statmod = rss_gag, segmented = sum(residuals(s_gag)^2))
```

```{r}
#| label: gag-numbers
#| include: false
ci_gag <- confint(fit_gag)
psi_gag <- ci_gag["mu:seg.psi1", ]
cf_gag <- coef(fit_gag)$mu
```

The two fits give the same position, $\hat\psi = `r fmt(psi_gag$estimate, 4)`$
years, with the same standard error, `r fmt(psi_gag$se, 4)`, and the same
residual sum of squares. The logarithm of the concentration falls by
`r fmt(-cf_gag[["seg.beta"]], 3)` per year before the break-point and by
`r fmt(-(cf_gag[["seg.beta"]] + cf_gag[["seg.gamma1"]]), 3)` per year after
it. The two log-likelihoods differ slightly, because `statmodels7` estimates
$\sigma$ by REML, as `lm()` does, and `logLik()` of `segmented` reads the
maximum likelihood estimate. @fig-gag-seg shows the fitted lines.

```{r}
#| label: fig-gag-seg
#| fig-cap: "Logarithm of the GAG concentration against age, with the fitted model of one break-point. The vertical line marks the estimated position."
#| echo: false
#| fig-height: 3.8
ga2 <- data.frame(Age = seq(min(GAGurine$Age), max(GAGurine$Age),
                            length.out = 400))
ga2$mu <- predict(fit_gag, what = "mu", newdata = ga2)
ggplot(GAGurine, aes(x = Age, y = lGAG)) +
  geom_point(colour = book_grey, size = 0.9) +
  geom_line(data = ga2, aes(y = mu, colour = "fitted mean"), linewidth = 0.9) +
  geom_vline(data = data.frame(psi = psi_gag$estimate),
             aes(xintercept = psi, colour = "fitted mean"), linetype = 3) +
  labs(x = "age (years)", y = "log GAG")
```

### Is there a break-point?

The standard error of $\psi$ says how precisely the position is located if
there is a break-point. Whether there is one at all is a different question.
Its null hypothesis is $\gamma = 0$, and under it the position does not enter
the model: $\psi$ is a parameter that exists only under the alternative, so
the usual tests of $\gamma = 0$ have no fixed reference distribution
[@davies1987]. Davies' approach fixes the position at a value $q$, computes
the statistic of $\gamma = 0$ for that $q$, and reads the supremum of the
resulting process over $q$. The function `statmod_breakpoint_test()` uses
Rao's score statistic
$$
S(q) = U_\gamma(q)^2 \, \big[K(q)^{-1}\big]_{\gamma\gamma},
$$
where $U_\gamma(q)$ is the score of $\gamma$ in the model with the break-point
at $q$, evaluated at the fit of the model without the break-point, and
$K(q)$ is the information of that model. One fit of the model without the
break-point serves every $q$, so the process costs one gradient and one
Hessian per position. For a change of slope $S(q)$ is a smooth function of
$q$, and the p-value of $M = \sup_q S(q)$ has the upper bound of
@davies1987,
$$
P\Big(\sup_q S(q) > M\Big) \le P(\chi^2_1 > M) +
V \, \frac{e^{-M/2}}{\sqrt{2\pi}}, \qquad
V = \sum_j \big|S(q_{j+1})^{1/2} - S(q_j)^{1/2}\big| ,
$$
with the sum over the grid of positions at which the process is read (ten
by default). The test reads the score of whatever model carries the term,
so it works for any family, in the equation of any parameter, and inside a
parameter of `nl()`.

```{r}
#| label: test-gag
test_gag <- statmod_breakpoint_test(fit_gag)
test_gag
segmented::davies.test(lm_gag, seg.Z = ~ Age)
```

Both tests reject the model without a break-point. The function
`davies.test()` of `segmented` reads the likelihood ratio on its own grid
for a linear or generalized linear model; our score process gives the same
conclusion. @fig-gag-test shows the process: it is largest near the age at
which the slope changes most, and it is far above the quantiles of a
$\chi^2_1$ everywhere in the middle of the range.

```{r}
#| label: fig-gag-test
#| fig-cap: "Score statistic of the change of slope with the position held at each of ten ages, for the GAG data. The dashed line is the 0.95 quantile of a chi-squared with one degree of freedom."
#| echo: false
#| fig-height: 3.2
ggplot(test_gag$process, aes(x = psi, y = score)) +
  geom_line(colour = book_palette[1]) +
  geom_point(colour = book_palette[1]) +
  geom_hline(yintercept = qchisq(0.95, df = 1), linetype = 2) +
  labs(x = "position q (years)", y = "score statistic S(q)")
```

### Several break-points

The argument `npsi` sets the number of break-points. With two of them the
model has two changes of slope, $\gamma_1$ at $\psi_1$ and $\gamma_2$ at
$\psi_2$:

```{r}
#| label: fit-gag2
fit_gag2 <- statmod(lGAG ~ seg(Age, npsi = 2), distrib = gaussian1_distrib(),
                    data = GAGurine)
summary(fit_gag2)
s_gag2 <- segmented::segmented(lm_gag, seg.Z = ~ Age, npsi = 2)
s_gag2$psi
rss_gag2 <- sum((GAGurine$lGAG - fitted(fit_gag2, what = "mu"))^2)
c(statmod = rss_gag2, segmented = sum(residuals(s_gag2)^2))
```

```{r}
#| label: gag2-numbers
#| include: false
cf_gag2 <- coef(fit_gag2)$mu
```

The second break-point is the same in the two fits, at
`r fmt(cf_gag2[["seg.psi2"]], 2)` years. The first one is at
`r fmt(cf_gag2[["seg.psi1"]], 2)` years in our fit, which is one of the
observed ages, and at `r fmt(s_gag2$psi[1, "Est."], 3)` years in the fit of
`segmented`. Our residual sum of squares is the smaller of the two. The mean
has no derivative in $\psi$ at an observed value of the covariate, and the
minimum of the residual sum of squares can fall exactly there
(@sec-breakpoints-locate explains how the positions are found).
@fig-gag-two compares the two models.

```{r}
#| label: fig-gag-two
#| fig-cap: "Logarithm of the GAG concentration against age, with the fitted models of one and two break-points. The vertical lines mark the estimated positions."
#| echo: false
#| fig-height: 3.8
ga2_long <- rbind(
  data.frame(Age = ga2$Age, mu = ga2$mu, model = "one break-point"),
  data.frame(Age = ga2$Age, model = "two break-points",
             mu = predict(fit_gag2, what = "mu", newdata = ga2[, "Age", drop = FALSE])))
vl_gag <- data.frame(
  psi = c(psi_gag$estimate, cf_gag2[c("seg.psi1", "seg.psi2")]),
  model = c("one break-point", "two break-points", "two break-points"))
ggplot(GAGurine, aes(x = Age, y = lGAG)) +
  geom_point(colour = book_grey, size = 0.9) +
  geom_line(data = ga2_long, aes(y = mu, colour = model, linetype = model),
            linewidth = 0.9) +
  geom_vline(data = vl_gag, aes(xintercept = psi, colour = model),
             linetype = 3) +
  labs(x = "age (years)", y = "log GAG")
```

The choice between one and two break-points is a choice of the number of
break-points, and @sec-breakpoints-number makes it by an information
criterion.

### A Poisson model

A break-point term goes into any family. The data frame `down` of the package
`segmented` gives the number of babies born with Down syndrome (`cases`) and
the number of births (`births`) for mothers of each age from 17 to 47. We
model the rate of cases per birth on the logarithmic scale, with a change of
slope at the age $\psi$:
$$
y_i \sim \mathrm{Poisson}(\mu_i), \qquad
\log\mu_i = \log b_i + \beta_0 + \beta_1 a_i + \gamma\,(a_i - \psi)_+ .
$$

```{r}
#| label: fit-down
data(down, package = "segmented")
fit_down <- statmod(cases ~ seg(age) + offset(log(births)),
                    distrib = poisson_distrib(), data = down)
summary(fit_down)
glm_down <- glm(cases ~ age + offset(log(births)), family = poisson,
                data = down)
s_down <- segmented::segmented(glm_down, seg.Z = ~ age)
s_down$psi
c(statmod = logLik(fit_down), segmented = logLik(s_down))
```

```{r}
#| label: down-numbers
#| include: false
cf_down <- coef(fit_down)$mu
```

The two fits agree. Before the age of `r fmt(cf_down[["seg.psi1"]], 1)`
years the rate changes by
`r fmt(100 * (exp(cf_down[["seg.beta"]]) - 1), 1)` per cent a year, which is
almost constant, and after it the rate grows by
`r fmt(100 * (exp(cf_down[["seg.beta"]] + cf_down[["seg.gamma1"]]) - 1), 1)`
per cent a year. @fig-down shows the observed rates per 1,000 births with the
fitted rate, which we obtain from `predict()` with `births = 1000` in
`newdata`, so that the offset gives the rate per 1,000 births.

```{r}
#| label: fig-down
#| fig-cap: "Cases of Down syndrome per 1,000 births against the age of the mother, on the logarithmic scale, with the fitted rate. The vertical line marks the estimated break-point."
#| echo: false
#| fig-height: 3.6
ag <- data.frame(age = seq(min(down$age), max(down$age), length.out = 300),
                 births = 1000)
ag$rate <- predict(fit_down, what = "mu", newdata = ag)
ggplot(down, aes(x = age, y = 1000 * cases / births)) +
  geom_point(colour = book_grey) +
  geom_line(data = ag, aes(y = rate, colour = "fitted rate"), linewidth = 0.9) +
  geom_vline(data = data.frame(psi = cf_down[["seg.psi1"]]),
             aes(xintercept = psi, colour = "fitted rate"), linetype = 3) +
  scale_y_log10() +
  labs(x = "age of the mother (years)", y = "cases per 1,000 births")
```

## A change of level: jump() and jseg() {#sec-breakpoints-jump}

The series `Nile` gives the annual flow of the river Nile at Aswan, in units
of $10^8$ cubic meters, from 1871 to 1970 [@cobb1978]. The flow drops at the
end of the nineteenth century. We model the mean as a step function of the
year $t$:
$$
y_t \sim \mathrm{N}(\mu_t, \sigma), \qquad
\mu_t = \beta_0 + \delta\, \mathbb{1}(t > \psi) .
$$

```{r}
#| label: fit-nile
nile <- data.frame(year = 1871:1970, flow = as.numeric(Nile))
fit_nile <- statmod(flow ~ jump(year), distrib = gaussian1_distrib(),
                    data = nile)
summary(fit_nile)
```

The mean is now a step function of $\psi$: it does not change while $\psi$
moves between two consecutive years. So the log-likelihood has no curvature
in $\psi$, and $\psi$ has no standard error. The position is the minimum of
the residual sum of squares over the intervals between consecutive years
(@sec-breakpoints-locate), and it is held there while the other
coefficients are fitted. With the position held, the model is linear in
$\beta_0$ and $\delta$, and their standard errors are those of `lm()` with
the indicator written as a column. They are conditional on the position.
The position counts one degree of freedom.

The function `stepmented()` of `segmented` fits the same model:

```{r}
#| label: stepmented-nile
s_nile <- segmented::stepmented(lm(flow ~ 1, data = nile), ~ year)
s_nile$psi
rss_nile <- sum((nile$flow - fitted(fit_nile, what = "mu"))^2)
c(statmod = rss_nile, stepmented = sum(residuals(s_nile)^2))
```

```{r}
#| label: nile-numbers
#| include: false
cf_nile <- coef(fit_nile)$mu
ci_nile <- confint(fit_nile)
```

Both fits put the step between 1898 and 1899, and they have the same residual
sum of squares. We compare the residual sums of squares and not the
log-likelihoods, because our $\sigma$ is the REML estimate
$\sqrt{\mathrm{RSS}/(n-3)}$ and `stepmented()` reports the maximum likelihood
one. The mean flow is `r fmt(cf_nile[["(Intercept)"]], 0)` before the step
and `r fmt(cf_nile[["(Intercept)"]] + cf_nile[["jump.delta1"]], 0)` after it,
a drop of `r fmt(-cf_nile[["jump.delta1"]], 0)` with a standard error of
`r fmt(ci_nile["mu:jump.delta1", "se"], 1)`. @fig-nile-step shows the series with
the fitted step.

```{r}
#| label: fig-nile-step
#| fig-cap: "Annual flow of the Nile at Aswan, with the fitted step."
#| echo: false
#| fig-height: 3.4
nile$step <- predict(fit_nile, what = "mu", newdata = nile)
ggplot(nile, aes(x = year)) +
  geom_line(aes(y = flow), colour = book_grey) +
  geom_step(aes(y = step, colour = "fitted step"), linewidth = 0.9) +
  labs(x = "year", y = expression(flow ~ (10^8 ~ m^3)))
```

### A bootstrap for the position {#sec-breakpoints-bootstrap}

The parametric bootstrap gives the uncertainty of the position of a step
[@efron1993]. We draw new data sets from the fitted model with `simulate()`,
fit the model again to each of them, and look at the spread of the new
positions:

```{r}
#| label: boot-nile
set.seed(1)
sims <- simulate(fit_nile, nsim = 100)
psi_star <- numeric(100)
for (k in 1:100) {
  nile_k <- nile
  nile_k$flow <- sims[, k]
  fit_k <- statmod(flow ~ jump(year), distrib = gaussian1_distrib(),
                   data = nile_k)
  psi_star[k] <- coef(fit_k)$mu[["jump.psi1"]]
}
quantile(psi_star, probs = c(0.025, 0.975))
mean(floor(psi_star) == 1898)
```

```{r}
#| label: boot-numbers
#| include: false
q_psi <- stats::quantile(psi_star, c(0.025, 0.975))
share <- mean(floor(psi_star) == 1898)
```

In `r fmt(100 * share, 0)` per cent of the simulated data sets the step falls
between 1898 and 1899, as in the original fit, and 95 per cent of the
positions lie between `r yr(q_psi[1])` and `r yr(q_psi[2])`.

### Is there a step?

The test of @sec-breakpoints-seg applies to a change of level too, with one
difference. The score process of a step is constant between consecutive
years and jumps at each of them, and Davies' bound, which adds up those
jumps, is then too wide to be useful. `statmod_breakpoint_test()` therefore
reads the supremum over every interval between consecutive years, and
estimates its distribution under the null hypothesis by a parametric
bootstrap: it draws new responses from the fit without a step, fits that
model again to each, and computes the supremum of each new process. The
p-value is the share of bootstrap suprema at least as large as the observed
one, counting the observed one. We use 99 replicates:

```{r}
#| label: test-nile
test_nile <- statmod_breakpoint_test(fit_nile, n_boot = 99, seed = 1)
test_nile
```

The p-value is `r test_nile$p.value`, the smallest that 99 replicates can
give: none of the bootstrap suprema reaches the observed one.

### A change of level in a Poisson model

A change of level goes into any family, as a change of slope does. The data
frame `coal` of the package `boot` gives the dates of the explosions in
British coal mines that killed ten or more people, from 1851 to 1962
[@jarrett1979; @davison1997]. We count the disasters of each year and model
the count as a Poisson variable whose mean drops once:
$$
y_t \sim \mathrm{Poisson}(\mu_t), \qquad
\log\mu_t = \beta_0 + \delta\,\mathbb{1}(t > \psi) .
$$

```{r}
#| label: fit-coal
data(coal, package = "boot")
coal_y <- data.frame(year = 1851:1962)
coal_y$count <- as.numeric(table(factor(floor(coal$date), levels = 1851:1962)))
fit_coal <- statmod(count ~ jump(year), distrib = poisson_distrib(),
                    data = coal_y)
summary(fit_coal)
s_coal <- segmented::stepmented(glm(count ~ 1, family = poisson, data = coal_y),
                                ~ year)
s_coal$psi
c(statmod = logLik(fit_coal), stepmented = logLik(s_coal))
```

```{r}
#| label: coal-numbers
#| include: false
cf_coal <- coef(fit_coal)$mu
```

The two fits have the same log-likelihood, and both put the change between
1891 and 1892. The expected number of disasters falls from
`r fmt(exp(cf_coal[["(Intercept)"]]), 2)` a year to
`r fmt(exp(cf_coal[["(Intercept)"]] + cf_coal[["jump.delta1"]]), 2)`, a ratio
of `r fmt(exp(cf_coal[["jump.delta1"]]), 2)`. @fig-coal shows the counts with
the fitted mean.

```{r}
#| label: fig-coal
#| fig-cap: "Number of coal-mining disasters in Britain in each year, with the fitted mean of the Poisson model with one change of level."
#| echo: false
#| fig-height: 3.4
coal_y$mu <- predict(fit_coal, what = "mu", newdata = coal_y)
ggplot(coal_y, aes(x = year)) +
  geom_segment(aes(xend = year, y = 0, yend = count), colour = book_grey,
               linewidth = 0.8) +
  geom_step(aes(y = mu, colour = "fitted mean"), linewidth = 0.9) +
  labs(x = "year", y = "disasters")
```

### A change of level and of slope

The term `jseg()` changes both the level and the slope at the same position:
$$
\mu_i = \beta_0 + \beta_1 x_i + \gamma\,(x_i - \psi)_+ + \delta\,\mathbb{1}(x_i > \psi) .
$$
We simulate 200 observations from this model with $\psi = 6$, a jump of 1.5
and a change of slope of $-0.6$, and we fit the three terms:

```{r}
#| label: fit-jseg
set.seed(11)
x <- sort(runif(200, 0, 10))
y <- 1 + 0.3 * x + ifelse(x > 6, 1.5 - 0.6 * (x - 6), 0) +
  rnorm(200, sd = 0.5)
sim <- data.frame(x = x, y = y)
fit_jseg <- statmod(y ~ jseg(x), distrib = gaussian1_distrib(), data = sim)
summary(fit_jseg)
```

The summary gives the slope `beta`, the change of slope `gamma1`, the change
of level `delta1` and the position `psi1`, which, as for `jump()`, has no
standard error. We also fit the model with a change of slope only and the
model with a change of level only. In `y ~ x + jump(x)` the formula writes `x`
for the linear effect, because `jump()` carries no slope of its own.

```{r}
#| label: fit-jseg-others
fit_seg0 <- statmod(y ~ seg(x), distrib = gaussian1_distrib(), data = sim)
summary(fit_seg0)
fit_jump0 <- statmod(y ~ x + jump(x), distrib = gaussian1_distrib(), data = sim)
summary(fit_jump0)
c(jseg = BIC(fit_jseg), seg = BIC(fit_seg0), jump = BIC(fit_jump0))
```

The BIC is smallest for `jseg()`, whose position is
`r fmt(coef(fit_jseg)$mu[["jseg.psi1"]], 2)` against the true 6. @fig-jseg
shows the three fits. The model with a change of slope only joins the two
lines at the break-point and so misses the jump; the model with a change of
level only keeps one slope on both sides.

```{r}
#| label: fig-jseg
#| fig-cap: "Simulated data with a change of level and of slope at 6, and the fitted means of jseg(), seg() and x + jump()."
#| echo: false
#| fig-height: 3.6
xg <- data.frame(x = seq(0, 10, length.out = 500))
jseg_fits <- rbind(
  data.frame(x = xg$x, model = "jseg()",
             mu = predict(fit_jseg, what = "mu", newdata = xg)),
  data.frame(x = xg$x, model = "seg()",
             mu = predict(fit_seg0, what = "mu", newdata = xg)),
  data.frame(x = xg$x, model = "x + jump()",
             mu = predict(fit_jump0, what = "mu", newdata = xg)))
jseg_fits$model <- factor(jseg_fits$model, levels = c("jseg()", "seg()",
                                                      "x + jump()"))
ggplot(sim, aes(x = x, y = y)) +
  geom_point(colour = book_grey, size = 0.9) +
  geom_line(data = jseg_fits, aes(y = mu, colour = model, linetype = model),
            linewidth = 0.9) +
  labs(x = "x", y = "y")
```

### A change of level and of slope in a Gamma model

For a positive response with a variance that grows with the mean we use the
Gamma family with a logarithmic link. We simulate 300 observations whose mean
has a change of level of 0.8 and a change of slope of $-0.15$ at $x = 4$ on
the log scale:
$$
y_i \sim \mathrm{Gamma}(\mu_i, \phi), \qquad
\log\mu_i = \beta_0 + \beta_1 x_i + \gamma\,(x_i - \psi)_+ +
\delta\,\mathbb{1}(x_i > \psi) .
$$

```{r}
#| label: fit-gamma-jseg
set.seed(4)
x_g <- runif(300, 0, 10)
mu_g <- exp(0.5 + 0.1 * x_g + ifelse(x_g > 4, 0.8 - 0.15 * (x_g - 4), 0))
sim_g <- data.frame(x = x_g, y = rgamma(300, shape = 4, rate = 4 / mu_g))
fit_gj <- statmod(y ~ jseg(x), distrib = gamma1_distrib(), data = sim_g)
summary(fit_gj)
fit_gs <- statmod(y ~ seg(x), distrib = gamma1_distrib(), data = sim_g)
summary(fit_gs)
c(jseg = BIC(fit_gj), seg = BIC(fit_gs))
```

```{r}
#| label: gamma-jseg-numbers
#| include: false
cf_gj <- coef(fit_gj)$mu
```

The estimates are close to the values of the simulation: the position is
`r fmt(cf_gj[["jseg.psi1"]], 2)` against 4, the change of level
`r fmt(cf_gj[["jseg.delta1"]], 2)` against 0.8 and the change of slope
`r fmt(cf_gj[["jseg.gamma1"]], 3)` against $-0.15$. The model without the
change of level has the larger BIC. As in the gaussian case, the position has
no standard error, and the other coefficients have standard errors
conditional on it. @fig-gamma-jseg shows the two fits.

```{r}
#| label: fig-gamma-jseg
#| fig-cap: "Simulated Gamma responses with a change of level and of slope of the log mean at 4, with the fitted means of jseg() and seg()."
#| echo: false
#| fig-height: 3.6
gam_fits <- rbind(
  data.frame(x = xg$x, model = "jseg()",
             mu = predict(fit_gj, what = "mu", newdata = xg)),
  data.frame(x = xg$x, model = "seg()",
             mu = predict(fit_gs, what = "mu", newdata = xg)))
ggplot(sim_g, aes(x = x, y = y)) +
  geom_point(colour = book_grey, size = 0.9) +
  geom_line(data = gam_fits, aes(y = mu, colour = model, linetype = model),
            linewidth = 0.9) +
  labs(x = "x", y = "y")
```

## Locating the break-points {#sec-breakpoints-locate}

The position of a break-point enters the model through the columns
$(x - \psi)_+$ and $\mathbb{1}(x > \psi)$, which are not differentiable in
$\psi$ at the observed values of $x$. `statmod()` finds the positions in two
stages. It first runs an iteration that moves the positions together with
the other coefficients: the linearization of @muggeo2003 for a change of
slope, and the working construction of @fasola2018 for a change of level.
It then polishes the positions on the exact profile of the objective, which
this section describes, and refits the other coefficients there.

### The profile over every interval

Hold every position but one, $\psi_k = q$, and write $F$ for the columns that
do not depend on $q$: the intercept, the covariate if the term carries a
slope, the columns of the other break-points and the columns of the other
terms. Write $A(q)$ for the columns that do depend on $q$: $(x - q)_+$ for
`seg()`, $\mathbb{1}(x > q)$ for `jump()`, and both for `jseg()`. By the
theorem of Frisch, Waugh and Lovell, the residual sum of squares of the
response on $F$ and $A(q)$ is the residual sum of squares on $F$ alone,
$\mathrm{RSS}_F$, less what $A(q)$ explains of the residuals $r = M_F y$:
$$
\mathrm{RSS}(q) = \mathrm{RSS}_F - c(q)' S(q)^{-1} c(q), \qquad
c(q) = A(q)' M_F\, y, \qquad S(q) = A(q)' M_F\, A(q),
$$
where $M_F = I - F(F'F)^{-1}F'$. The profile $\mathrm{RSS}(q)$ is what the
polish minimizes.

For a change of level, $A(q)$ changes only when $q$ crosses an observed value
of $x$. So $\mathrm{RSS}(q)$ is constant between consecutive observed values,
and its minimum over all $q$ is found by evaluating it once in each interval.
Every cross product in $c(q)$ and $S(q)$ is a sum over the observations above
$q$, of a fixed quantity or of a fixed quantity minus $q$ times another.
After sorting the observations once, these sums are cumulative sums read
from the top, so all the intervals together cost $O(np^2)$ operations, $p$
being the number of columns of $F$, rather than one regression each. The
position is placed at the midpoint of the best interval.

For a change of slope the profile is not constant inside an interval, but it
has a simple form there. Between the consecutive observed values
$u_j < u_{j+1}$ the column is $(x - q)\, s$ with $s = \mathbb{1}(x > u_j)$
fixed, so with $t = x\, s$ it is $t - q\, s$, and
$$
\mathrm{RSS}(q) = \mathrm{RSS}_F -
\frac{(c_t - q\, c_s)^2}{a_{tt} - 2q\, a_{ts} + q^2 a_{ss}} ,
$$
where $c_t = t' M_F y$, $c_s = s' M_F y$ and the $a$ are the cross products
of $t$ and $s$ after $M_F$. The ratio has one stationary point,
$q^* = -b/g$, where $g$ and $b$ are the coefficients of $t$ and $s$ in the
regression of $r$ on the two columns. The minimum over the interval is at
$q^*$ when $q^*$ falls inside it, and at one of its ends otherwise, which is
an observed value. This is why the first position of the GAG model with two
break-points is exactly one of the observed ages.

With several break-points the polish minimizes the profile over one position
at a time, with the others held, and cycles through the positions until none
of them moves. Each step searches every interval, so a position can move
from one local minimum to another however far apart they are.

In a model whose objective is not a residual sum of squares (another
family, or the equation of a parameter other than a gaussian mean), the
same profile is computed on the working response of a scoring step,
$$
z_i = \eta_i + \frac{\partial \ell_i / \partial \eta_i}{h_i}, \qquad
h_i = -\mathbb{E}\Big(\frac{\partial^2 \ell_i}{\partial \eta_i^2}\Big),
$$
with weights $h_i$, which is the quadratic approximation of the
log-likelihood around the current fit. Since the approximation is not the
objective, a polished position is kept only when the refit there improves
the objective, and the five deepest local minima of the working profile are
also refitted and compared on the objective itself.

When the position has covariates, $\psi_i = w_i' p$, the polish moves the
coefficients $p$ along a line, $p + \lambda v$. The position of observation
$i$ crosses $x_i$ at $\lambda_i = (x_i - \psi_i)/(w_i' v)$, so the profile is
again evaluated once between consecutive crossings. With `psi ~ 0 + g` the
lines move the position of one group at a time, which is a search over every
interval of that group. With a continuous covariate in the formula of the
position, the lines follow a fixed set of directions, and the result is the
best point along those lines, not a guaranteed global minimum.

### Starting values

The data frame `globTempAnom` of the package `segmented` gives the yearly
anomaly of the global temperature from 1850 to 2023. We fit two changes of
slope with the default starting values, and from $\psi = (1880, 1950)$
without the bootstrap restarts of @muggeo2003 (`n_boot = 0`):

```{r}
#| label: fit-glob
data(globTempAnom, package = "segmented")
fit_glob <- statmod(Anomaly ~ seg(Year, npsi = 2),
                    distrib = gaussian1_distrib(), data = globTempAnom)
summary(fit_glob)
fit_glob_0 <- statmod(Anomaly ~ seg(Year, npsi = 2, psi = c(1880, 1950),
                                    n_boot = 0),
                      distrib = gaussian1_distrib(), data = globTempAnom)
rss <- function(fit) {
  sum((globTempAnom$Anomaly - fitted(fit, what = "mu"))^2)
}
lm_glob <- lm(Anomaly ~ Year, data = globTempAnom)
s_glob <- segmented::segmented(lm_glob, seg.Z = ~ Year, npsi = 2)
rbind(default = coef(fit_glob)$mu[c("seg.psi1", "seg.psi2")],
      start = coef(fit_glob_0)$mu[c("seg.psi1", "seg.psi2")],
      segmented = s_glob$psi[, "Est."])
c(default = rss(fit_glob), start = rss(fit_glob_0),
  segmented = sum(residuals(s_glob)^2))
```

The two fits reach the same positions as `segmented`. @fig-glob shows why
the start matters for the iteration and not for the polish: with the first
position held at 1910, the profile in the second position has two minima,
and an iteration that starts from 1950 can stop at the one near 2014, where
the second change of slope rests on a handful of years. The polish evaluates
every interval and moves the position to the deeper minimum.

```{r}
#| label: fig-glob
#| fig-cap: "Residual sum of squares of the model with two changes of slope as a function of the second break-point, with the first held at 1910. The vertical line marks the second break-point of the fit."
#| echo: false
#| fig-height: 3.4
prof_glob <- data.frame(psi2 = seq(1930, 2015, by = 0.5))
prof_glob$rss <- vapply(prof_glob$psi2, function(q) {
  X2 <- cbind(1, globTempAnom$Year, pmax(globTempAnom$Year - 1910, 0),
              pmax(globTempAnom$Year - q, 0))
  sum(lm.fit(X2, globTempAnom$Anomaly)$residuals^2)
}, numeric(1))
ggplot(prof_glob, aes(x = psi2, y = rss)) +
  geom_line(colour = book_palette[4]) +
  geom_vline(xintercept = coef(fit_glob)$mu[["seg.psi2"]],
             colour = book_palette[1], linetype = 2) +
  labs(x = expression(psi[2]), y = "residual sum of squares")
```

The function `seg_start()` returns the starting positions that `statmod()`
computes by default, so they can be inspected before a fit:

```{r}
#| label: seg-start
seg_start(seg(Year, npsi = 2), data = globTempAnom,
          y = globTempAnom$Anomaly)@spec$psi
```

### The number of break-points {#sec-breakpoints-number}

The number of break-points is a choice between models, and we make it with
an information criterion. Each position counts one degree of freedom, as
does each change of slope. We fit from none to five changes of slope to the
temperature anomalies:

```{r}
#| label: fit-glob-npsi
fits_glob <- list(statmod(Anomaly ~ Year, distrib = gaussian1_distrib(),
                          data = globTempAnom))
for (k in 1:5) {
  fits_glob[[k + 1]] <- statmod(Anomaly ~ seg(Year, npsi = k),
                                distrib = gaussian1_distrib(),
                                data = globTempAnom)
}
tab_npsi <- data.frame(
  npsi = 0:5,
  rss = vapply(fits_glob, rss, numeric(1)),
  edf = vapply(fits_glob, function(f) sum(f@edf$edf), numeric(1)),
  AIC = vapply(fits_glob, AIC, numeric(1)),
  BIC = vapply(fits_glob, BIC, numeric(1)))
tab_npsi
```

```{r}
#| label: glob-npsi-numbers
#| include: false
k_bic <- tab_npsi$npsi[which.min(tab_npsi$BIC)]
k_aic <- tab_npsi$npsi[which.min(tab_npsi$AIC)]
fit_bic <- fits_glob[[k_bic + 1]]
psi_bic <- coef(fit_bic)$mu[grepl("psi", names(coef(fit_bic)$mu))]
```

The BIC is smallest with `r k_bic` break-points, at
`r paste(yr(psi_bic, 0), collapse = ", ")`, and the AIC, which penalizes
each degree of freedom less, keeps adding break-points up to `r k_aic`. The
function `selgmented()` of `segmented` makes the choice by the BIC with its
own fits:

```{r}
#| label: selgmented-glob
s_sel <- segmented::selgmented(lm_glob, seg.Z = ~ Year, type = "bic",
                               Kmax = 5, msg = FALSE)
s_sel$psi
s_glob4 <- segmented::segmented(lm_glob, seg.Z = ~ Year, npsi = 4)
c(statmod = rss(fits_glob[[5]]), segmented = sum(residuals(s_glob4)^2))
```

It chooses `r nrow(s_sel$psi)` break-points, because its fit with four
stops at a local minimum whose residual sum of squares is larger than ours,
as the last line shows. The search over every interval of @sec-breakpoints-locate matters
most where the number of break-points grows, because the profile of each
position then has more local minima. @fig-glob-npsi shows the fit chosen by
the BIC.

```{r}
#| label: fig-glob-npsi
#| fig-cap: "Yearly anomaly of the global temperature, with the fit of the number of changes of slope chosen by the BIC. The vertical lines mark the estimated positions."
#| echo: false
#| fig-height: 3.6
yg <- data.frame(Year = seq(min(globTempAnom$Year), max(globTempAnom$Year),
                            length.out = 500))
yg$mu <- predict(fit_bic, what = "mu", newdata = yg)
ggplot(globTempAnom, aes(x = Year, y = Anomaly)) +
  geom_point(colour = book_grey, size = 0.9) +
  geom_line(data = yg, aes(y = mu, colour = "fitted mean"), linewidth = 0.9) +
  geom_vline(xintercept = psi_bic, colour = book_palette[1], linetype = 3) +
  labs(x = "year", y = "anomaly (degrees C)")
```

## Covariates on the coefficients {#sec-breakpoints-by}

Every coefficient of a break-point term can depend on covariates through a
formula with the coefficient on the left. The argument `by = ~ 0 + g` gives
every coefficient one value for each level of the factor `g`, and
`psi ~ 0 + g` gives only the position one value per level. The data frame
`plant` of the package `segmented` gives a measurement `y` of three plants
(`group`) at several times `time`. We fit a model with one change of slope
for each plant, and a model in which the three plants share the slopes but not
the position:

```{r}
#| label: fit-plant
data(plant, package = "segmented")
fit_by <- statmod(y ~ group + seg(time, by = ~ 0 + group),
                  distrib = gaussian1_distrib(), data = plant)
summary(fit_by)
fit_psi <- statmod(y ~ group + seg(time, psi ~ 0 + group),
                   distrib = gaussian1_distrib(), data = plant)
summary(fit_psi)
```

In the summary each developed coefficient has a compartment headed by its
formula, with one value per level of `group`. The function `segreg()` of
`segmented` fits the first model with the term `seg(time, by = group)` of
its own formula language. Our `seg()` has the same name, so we evaluate
`segmented`'s inside a local environment:

```{r}
#| label: segreg-plant
s_plant <- local({
  seg <- segmented::seg
  segmented::segreg(y ~ group + seg(time, by = group), data = plant)
})
s_plant$psi
c(statmod = sum((plant$y - fitted(fit_by, what = "mu"))^2),
  segreg = sum(residuals(s_plant)^2))
```

The two fits have the same positions and the same residual sum of squares.
@fig-plant shows the data of the three plants with the two fits.

```{r}
#| label: fig-plant
#| fig-cap: "The measurement of each plant against time, with the fit that gives every coefficient one value per plant and the fit with common slopes and one position per plant. The vertical lines mark the positions of the first fit."
#| echo: false
#| fig-height: 3.8
#| fig-width: 9
cb_plant <- coef(fit_by)$mu
tg_plant <- expand.grid(time = seq(min(plant$time), max(plant$time),
                                   length.out = 300),
                        group = levels(plant$group))
tg_plant$group <- factor(tg_plant$group, levels = levels(plant$group))
plant_fits <- rbind(
  data.frame(tg_plant, model = "by = ~ 0 + group",
             mu = predict(fit_by, what = "mu", newdata = tg_plant)),
  data.frame(tg_plant, model = "psi ~ 0 + group",
             mu = predict(fit_psi, what = "mu", newdata = tg_plant)))
vl_plant <- data.frame(
  group = factor(levels(plant$group), levels = levels(plant$group)),
  psi = cb_plant[paste0("seg.psi1.group", levels(plant$group))],
  model = "by = ~ 0 + group")
ggplot(plant, aes(x = time, y = y)) +
  geom_point(colour = book_grey) +
  geom_line(data = plant_fits, aes(y = mu, colour = model, linetype = model),
            linewidth = 0.9) +
  geom_vline(data = vl_plant, aes(xintercept = psi, colour = model),
             linetype = 3) +
  facet_wrap(~ group) +
  labs(x = "time", y = "y")
```

The second model is nested in the first: it has one slope and one change of
slope where the first has three of each, so it has four coefficients fewer.
We compare the two with the likelihood ratio test:

```{r}
#| label: lr-plant
l0 <- logLik(fit_psi)
l1 <- logLik(fit_by)
lr <- 2 * (as.numeric(l1) - as.numeric(l0))
lr
pchisq(lr, df = 4, lower.tail = FALSE)
```

The common slopes are rejected. The test treats the positions as known in
both models, which is the usual practice for nested break-point models with
the same number of break-points [@muggeo2003]. The same formulas work in
`jump()` and `jseg()`, whose positions are then located as
@sec-breakpoints-locate describes.

## A random break-point {#sec-breakpoints-random}

In a panel each subject can have its own break-point. The model with a random
position [@muggeo2014] is
$$
\mu_{ij} = \beta_0 + b_i + \beta_1 t_{ij} + \gamma\,(t_{ij} - \psi_i)_+ ,
\qquad \psi_i = \psi + u_i, \quad u_i \sim \mathrm{N}(0, \tau),
$$
with a random intercept $b_i \sim \mathrm{N}(0, \sigma_b)$. We simulate 20
subjects observed at 15 times, with $\psi = 5$ and $\tau = 0.8$, and write the
random position as a formula on `psi`:

```{r}
#| label: fit-panel
set.seed(3)
id <- factor(rep(1:20, each = 15))
t <- rep(seq(0, 10, length.out = 15), 20)
psi_i <- 5 + rnorm(20, sd = 0.8)
b0 <- 2 + rnorm(20, sd = 0.5)
y <- b0[id] + 0.8 * t - 1.2 * pmax(t - psi_i[id], 0) + rnorm(300, sd = 0.4)
panel <- data.frame(id = id, t = t, y = y)
fit_pan <- statmod(y ~ random(~ 1 | id) + seg(t, psi ~ random(~ 1 | id)),
                   distrib = gaussian1_distrib(), data = panel)
summary(fit_pan)
```

The compartment of `psi1` gives the fixed effect of its formula, the
population position, and then its random part, headed by the `random()`
term: the standard deviation $\tau$ of the positions, `sigma`, and a
summary of the predicted deviations of the twenty subjects. @muggeo2014 estimate the same model by
maximum likelihood, linearizing the model in the positions of the subjects
and fitting the resulting linear mixed model with `nlme`, and `segmented`
implements their method in `segmented.lme()`:

```{r}
#| label: segmented-lme
lme_pan <- nlme::lme(y ~ t, random = ~ 1 | id, data = panel)
s_pan <- segmented::segmented.lme(lme_pan, seg.Z = ~ t,
                                  random = list(id = nlme::pdDiag(~ 1 + G0)))
cf_pan <- coef(fit_pan)$mu
h_pan <- hyper(fit_pan)
sd_lme <- as.numeric(nlme::VarCorr(s_pan$lme.fit)[, "StdDev"])
cbind(statmod = c(position = cf_pan[["seg.psi1.(Intercept)"]],
                  gamma = cf_pan[["seg.gamma1"]],
                  sd_position = h_pan$estimate[grepl("seg", h_pan$term)],
                  sd_intercept = h_pan$estimate[!grepl("seg", h_pan$term)]),
      segmented.lme = c(nlme::fixef(s_pan$lme.fit)[c("G0", "U")],
                        sd_lme[2:1]))
```

The two fits give the same population position and change of slope, and
standard deviations that agree within one per cent. The positions of the subjects
are the population position plus each subject's random effect, and we
compare them with those of `segmented.lme()` and with the positions used in
the simulation:

```{r}
#| label: panel-psi
psi_hat <- cf_pan[["seg.psi1.(Intercept)"]] +
  cf_pan[paste0("seg.psi1.random.", 1:20)]
round(cbind(statmod = psi_hat, segmented.lme = s_pan$psi.i,
            true = psi_i), 3)[1:6, ]
cor(psi_hat, psi_i)
```

@fig-panel shows three subjects, the one with the earliest true position,
one in the middle and the one with the latest, each with the curve that the
model fits to it.

```{r}
#| label: fig-panel
#| fig-cap: "Three subjects of the simulated panel, each with the curve that the model fits to it, the estimated position of its break-point (dashed) and the position used in the simulation (dotted)."
#| echo: false
#| fig-height: 3.8
#| fig-width: 9
sel <- sort(c(which.min(psi_i), order(psi_i)[10], which.max(psi_i)))
lab_sel <- paste("subject", sel)
tg_pan <- expand.grid(t = seq(0, 10, length.out = 200), id = sel)
tg_pan$id <- factor(tg_pan$id, levels = levels(panel$id))
tg_pan$mu <- predict(fit_pan, what = "mu", newdata = tg_pan)
tg_pan$subject <- factor(paste("subject", tg_pan$id), levels = lab_sel)
pts_pan <- panel[panel$id %in% sel, ]
pts_pan$subject <- factor(paste("subject", pts_pan$id), levels = lab_sel)
vl_pan <- rbind(
  data.frame(subject = factor(lab_sel, levels = lab_sel),
             psi = psi_hat[sel], what = "estimated position"),
  data.frame(subject = factor(lab_sel, levels = lab_sel),
             psi = psi_i[sel], what = "true position"))
ggplot(pts_pan, aes(x = t, y = y)) +
  geom_point(colour = book_grey) +
  geom_line(data = tg_pan, aes(y = mu, colour = "fitted curve"),
            linewidth = 0.9) +
  geom_vline(data = vl_pan, aes(xintercept = psi, linetype = what)) +
  scale_linetype_manual(values = c(2, 3)) +
  facet_wrap(~ subject) +
  labs(x = "t", y = "y")
```

### Random effects on the other coefficients

The position is not the only coefficient that can vary between subjects.
Each coefficient of the term takes a formula of its own, so the change of
slope can have a random effect too:
$$
\mu_{ij} = \beta_0 + b_i + \beta_1 t_{ij} + \gamma_i\,(t_{ij} - \psi_i)_+ ,
\qquad \psi_i = \psi + u_i, \qquad \gamma_i = \gamma + v_i ,
$$
with $u_i \sim \mathrm{N}(0, \tau_\psi)$ and
$v_i \sim \mathrm{N}(0, \tau_\gamma)$. We simulate a new panel in which the
change of slope varies between subjects with standard deviation 0.3, and we
fit it:

```{r}
#| label: fit-panel-gamma
set.seed(6)
g_i <- -1.2 + rnorm(20, sd = 0.3)
y_g <- b0[id] + 0.8 * t + g_i[id] * pmax(t - psi_i[id], 0) +
  rnorm(300, sd = 0.4)
panel_g <- data.frame(id = id, t = t, y = y_g)
fit_pg <- statmod(y ~ random(~ 1 | id) +
                    seg(t, psi ~ random(~ 1 | id), gamma1 ~ random(~ 1 | id)),
                  distrib = gaussian1_distrib(), data = panel_g)
summary(fit_pg)
```

The block of the term now has two compartments, one for `psi1` and one for
`gamma1`. Each gives the fixed effect of its formula and then the standard
deviation of its random effects. Every `random()` term has its own variance,
so $\tau_\psi$ and $\tau_\gamma$ are estimated separately, by REML, together
with the standard deviation of the random intercept.

To give two random effects the same variance we write the same label in the
argument `id` of the two `random()` terms. Here this means
$\tau_\psi = \tau_\gamma$:

```{r}
#| label: fit-panel-gamma-id
fit_pg_id <- statmod(y ~ random(~ 1 | id) +
                       seg(t, psi ~ random(~ 1 | id, id = "re"),
                           gamma1 ~ random(~ 1 | id, id = "re")),
                     distrib = gaussian1_distrib(), data = panel_g)
summary(fit_pg_id)
hyper(fit_pg_id)
```

The two models differ by one variance parameter, and we compare them with the
likelihood ratio of their REML criteria:

```{r}
#| label: lr-panel-gamma
l0 <- logLik(fit_pg_id, type = "marginal")
l1 <- logLik(fit_pg, type = "marginal")
lr_g <- 2 * (as.numeric(l1) - as.numeric(l0))
lr_g
pchisq(lr_g, df = 1, lower.tail = FALSE)
```

The common variance is rejected. This is what we expect, because the two
standard deviations are in different units: $\tau_\psi$ is a length of time
and $\tau_\gamma$ a change of slope, so they are equal only by accident. A
common variance is natural when the coefficients are on the same scale, for
example the changes of level of two break-points, `delta1` and `delta2`.
The hypothesis $\tau_\psi = \tau_\gamma$ does not put either standard
deviation at zero, so the usual $\chi^2_1$ reference applies.
@fig-panel-gamma shows three subjects with the curves of the model with two
separate variances.

```{r}
#| label: fig-panel-gamma
#| fig-cap: "Three subjects of the panel in which the change of slope varies between subjects (the smallest, a middle and the largest true change of slope), with the curve that the model with random position and random change of slope fits to each of them."
#| echo: false
#| fig-height: 3.8
#| fig-width: 9
sel_g <- sort(c(which.min(g_i), order(g_i)[10], which.max(g_i)))
lab_g <- sprintf("subject %d, change of slope %.2f", sel_g, g_i[sel_g])
tg_pg <- expand.grid(t = seq(0, 10, length.out = 200), id = sel_g)
tg_pg$id <- factor(tg_pg$id, levels = levels(panel_g$id))
tg_pg$mu <- predict(fit_pg, what = "mu", newdata = tg_pg)
tg_pg$subject <- factor(lab_g[match(tg_pg$id, sel_g)], levels = lab_g)
pts_pg <- panel_g[panel_g$id %in% sel_g, ]
pts_pg$subject <- factor(lab_g[match(pts_pg$id, sel_g)], levels = lab_g)
ggplot(pts_pg, aes(x = t, y = y)) +
  geom_point(colour = book_grey) +
  geom_line(data = tg_pg, aes(y = mu, colour = "fitted curve"),
            linewidth = 0.9) +
  facet_wrap(~ subject) +
  labs(x = "t", y = "y")
```

### Heavy-tailed random positions

The random effects need not be gaussian. The argument `distrib` of
`random()` takes any distribution of `distributions7` with location zero, and
a Student $t$ gives positions whose distribution has heavier tails, so that a
few subjects whose break-point lies far from the others do not inflate the
spread of all the rest. We move the break-points of two subjects of the
first panel to 1.5 and 8.8, and fit the model with gaussian random positions
and with $t$ random positions whose degrees of freedom are estimated with the
scale:

```{r}
#| label: fit-panel-t
psi_o <- psi_i
psi_o[c(4, 17)] <- c(1.5, 8.8)
set.seed(7)
y_o <- b0[id] + 0.8 * t - 1.2 * pmax(t - psi_o[id], 0) + rnorm(300, sd = 0.4)
panel_o <- data.frame(id = id, t = t, y = y_o)
fit_og <- statmod(y ~ random(~ 1 | id) + seg(t, psi ~ random(~ 1 | id)),
                  distrib = gaussian1_distrib(), data = panel_o)
t_prior <- fixed(student_t1_distrib(), mu = 0)
fit_ot <- statmod(y ~ random(~ 1 | id) +
                    seg(t, psi ~ random(~ 1 | id, distrib = t_prior)),
                  distrib = gaussian1_distrib(), data = panel_o)
summary(fit_ot)
```

```{r}
#| label: panel-t-numbers
psi_of <- function(fit) {
  cf <- coef(fit)$mu
  cf[["seg.psi1.(Intercept)"]] + cf[paste0("seg.psi1.random.", 1:20)]
}
h_og <- hyper(fit_og)
h_ot <- hyper(fit_ot)
rbind(gaussian = c(scale = h_og$estimate[h_og$name == "sigma" &
                                           grepl("seg", h_og$term)],
                   rmse = sqrt(mean((psi_of(fit_og) - psi_o)^2)),
                   rmse_others = sqrt(mean((psi_of(fit_og) - psi_o)[-c(4, 17)]^2))),
      t = c(scale = h_ot$estimate[h_ot$name == "sigma" &
                                    grepl("seg", h_ot$term)],
            rmse = sqrt(mean((psi_of(fit_ot) - psi_o)^2)),
            rmse_others = sqrt(mean((psi_of(fit_ot) - psi_o)[-c(4, 17)]^2))))
```

```{r}
#| label: panel-t-inline
#| include: false
nu_ot <- h_ot$estimate[h_ot$name == "nu"]
sc_og <- h_og$estimate[h_og$name == "sigma" & grepl("seg", h_og$term)]
sc_ot <- h_ot$estimate[h_ot$name == "sigma" & grepl("seg", h_ot$term)]
```

The compartment of `psi1` now reports the two parameters of the $t$
distribution under its random part, the scale and the degrees of freedom.
The degrees of freedom are estimated at `r fmt(nu_ot, 2)`, which gives tails
about as heavy as those of a Cauchy distribution. The gaussian model
inflates its standard deviation to `r fmt(sc_og, 2)` to cover the two
distant subjects, while the scale of the $t$, `r fmt(sc_ot, 2)`, describes
the spread of the other eighteen. The positions of the subjects are
estimated about equally well by the two models, overall and for the
eighteen subjects that were not moved.

## Smoothed break-points {#sec-breakpoints-smooth}

A change of level has no derivative in its position, and so no standard
error. One way to obtain one replaces the step by a smooth function that
rises from 0 to 1 over a small width around the position. The construction
starts from two identities. With $u = x - \psi$,
$$
(u)_+ = \frac{u + \lvert u\rvert}{2}, \qquad
\mathbb{1}(u > 0) = \frac{1 + \operatorname{sgn} u}{2}
= \frac{1}{2}\Big(1 + \frac{d\lvert u\rvert}{du}\Big) \quad (u \ne 0),
$$
so the truncated line and the step are algebraic sums of $u$, of the
absolute value $\lvert u\rvert$ and of its derivative. The absolute value is
the one non-smooth ingredient. Replacing it by a smooth function $s_h(u)$
with $s_h(u) \to \lvert u\rvert$ as $h \to 0$ gives
$$
(u)_+ \approx \frac{u + s_h(u)}{2}, \qquad
\mathbb{1}(u > 0) \approx \frac{1 + s_h'(u)}{2},
$$
a smoothed truncated line and a smoothed step. Their derivatives in $\psi$
are $-\{1 + s_h'(u)\}/2$ and $-s_h''(u)/2$, and the second is not zero near
the position, so the likelihood has a curvature in $\psi$ and $\psi$ has a
standard error. The argument `smoothed` of `jump()` and `jseg()` chooses the
function, and `statmodels7` has three of them:

* `smooth_probit(h)` takes $s_h(u) = \mathbb{E}\lvert u + hZ\rvert =
  u\{2\Phi(u/h) - 1\} + 2h\,\phi(u/h)$ with $Z$ standard normal, whose
  derivative is $2\Phi(u/h) - 1$, so the step is the gaussian distribution
  function $\Phi(u/h)$. It differs from the sharp step only within a few
  widths of $\psi$, because the difference has gaussian tails.
* `smooth_quintic(h)` uses a polynomial on $[-h, h]$, equal to
  $\lvert u\rvert$ outside it, so the step is 0 below $\psi - h$ and 1 above
  $\psi + h$ exactly. Outside that window the model is the sharp model.
* `smooth_hyperbolic(c)` takes $s_h(u) = \sqrt{u^2 + c}$, so the step is
  $\{1 + u/\sqrt{u^2 + c}\}/2$. It is the simplest of the three, but it
  approaches 0 and 1 only as $1/u^2$, so it moves the fitted mean far from
  $\psi$ as well.

We fit the three smoothed steps to the Nile with their default widths:

```{r}
#| label: fit-nile-smooth
fit_np <- statmod(flow ~ jump(year, smoothed = smooth_probit()),
                  distrib = gaussian1_distrib(), data = nile)
summary(fit_np)
fit_nq <- statmod(flow ~ jump(year, smoothed = smooth_quintic()),
                  distrib = gaussian1_distrib(), data = nile)
summary(fit_nq)
fit_nh <- statmod(flow ~ jump(year, smoothed = smooth_hyperbolic()),
                  distrib = gaussian1_distrib(), data = nile)
summary(fit_nh)
```

We put the estimates of the sharp step of @sec-breakpoints-jump beside those
of the three smoothed models, with their log-likelihoods:

```{r}
#| label: smooth-compare
ci_n <- confint(fit_nile)
ci_np <- confint(fit_np)
ci_nq <- confint(fit_nq)
ci_nh <- confint(fit_nh)
est <- cbind(sharp = ci_n$estimate, probit = ci_np$estimate,
             quintic = ci_nq$estimate, hyperbolic = ci_nh$estimate)
se <- cbind(sharp = ci_n$se, probit = ci_np$se, quintic = ci_nq$se,
            hyperbolic = ci_nh$se)
rownames(est) <- rownames(ci_n)
rownames(se) <- rownames(ci_n)
round(est, 2)
round(se, 2)
c(sharp = logLik(fit_nile), probit = logLik(fit_np), quintic = logLik(fit_nq),
  hyperbolic = logLik(fit_nh))
```

The three smoothed models put the step at
`r yr(ci_np["mu:jump.psi1", "estimate"], 2)`,
`r yr(ci_nq["mu:jump.psi1", "estimate"], 2)` and
`r yr(ci_nh["mu:jump.psi1", "estimate"], 2)`, with standard errors of
`r fmt(ci_np["mu:jump.psi1", "se"], 2)`,
`r fmt(ci_nq["mu:jump.psi1", "se"], 2)` and
`r fmt(ci_nh["mu:jump.psi1", "se"], 2)` years. The sharp step sits in the
same interval, between 1898 and 1899, and has no standard error. The levels
and the drop change by a few units, and their standard errors are a little
larger in the smoothed models, which add the uncertainty of the position. The
log-likelihood of the sharp step is the highest, and the smoothed ones are
lower by less than one unit: the smoothed mean moves from one level to the
other over a year or two, and with yearly data this is almost the same model.
@fig-nile-smooth shows the four fitted means around the break-point.

```{r}
#| label: fig-nile-smooth
#| fig-cap: "Annual flow of the Nile from 1880 to 1920, with the fitted means of the sharp step and of the three smoothed steps."
#| echo: false
#| fig-height: 3.8
yw <- data.frame(year = seq(1880, 1920, length.out = 801))
nile_smooth <- rbind(
  data.frame(year = yw$year, model = "sharp step",
             mu = predict(fit_nile, what = "mu", newdata = yw)),
  data.frame(year = yw$year, model = "probit",
             mu = predict(fit_np, what = "mu", newdata = yw)),
  data.frame(year = yw$year, model = "quintic",
             mu = predict(fit_nq, what = "mu", newdata = yw)),
  data.frame(year = yw$year, model = "hyperbolic",
             mu = predict(fit_nh, what = "mu", newdata = yw)))
nile_smooth$model <- factor(nile_smooth$model,
                            levels = c("sharp step", "probit", "quintic",
                                       "hyperbolic"))
ggplot(nile[nile$year >= 1880 & nile$year <= 1920, ], aes(x = year, y = flow)) +
  geom_point(colour = book_grey) +
  geom_line(data = nile_smooth, aes(y = mu, colour = model, linetype = model),
            linewidth = 0.8) +
  labs(x = "year", y = expression(flow ~ (10^8 ~ m^3)))
```

### The width of the smoothing

The width $h$ is a constant of the model, chosen before the fit and not
estimated. When it is not given, it is computed from the spacing of the
covariate: $h$ is the spacing for the probit, 3.61 times the spacing for the
quintic, so that about five observations fall inside $[-h, h]$, and $c$ is
the square of the spacing for the hyperbolic. The choice matters because the
standard error of $\psi$ belongs to the smoothed model at that width. We fit
the probit step to the Nile with widths from a tenth of a year to eight
years:

```{r}
#| label: nile-width
widths <- c(0.1, 0.25, 0.5, 1, 2, 4, 8)
tab_h <- t(vapply(widths, function(h) {
  f <- statmod(flow ~ jump(year, smoothed = smooth_probit(h)),
               distrib = gaussian1_distrib(), data = nile)
  ci <- confint(f)
  c(h = h, psi = ci["mu:jump.psi1", "estimate"], se_psi = ci["mu:jump.psi1", "se"],
    delta = ci["mu:jump.delta1", "estimate"], se_delta = ci["mu:jump.delta1", "se"],
    logLik = as.numeric(logLik(f)),
    converged = statmod_certificate(f)$state == "converged")
}, numeric(7)))
round(tab_h, 3)
```

As $h$ decreases the smoothed model approaches the sharp step and its
log-likelihood rises towards that of the sharp step, so the likelihood
cannot choose $h$: it would always choose the sharp step. The curvature in
$\psi$ comes from the observations within a few widths of $\psi$. With
$h = 0.1$ there is hardly any observation there, the curvature is close to
zero, the fit does not converge and the standard error of $\psi$ is
meaningless. From $h = 0.5$ on, the transition from one level to the other
spreads over more years as $h$ grows, the fit gets worse, and the standard
error of $\psi$ grows with $h$, from less than a year to several years. A width close to the
spacing of the covariate is the smallest one that gives every position a
curvature from a few observations, and it keeps the model close to the sharp
step; this is the default.

With random positions the width also changes the variance of the positions.
On the panel of @sec-breakpoints-marginal, which has a change of level, the
estimated standard deviation of the positions grows with $h$, and the change
of level is estimated too large when the width is wide:

```{r}
#| label: panel-width
#| echo: false
set.seed(4)
y_j <- b0[id] + 1.5 * (t > psi_i[id]) + rnorm(300, sd = 0.4)
panel_j <- data.frame(id = id, t = t, y = y_j)
tab_hr <- t(vapply(c(0.25, 0.71, 2), function(h) {
  f <- statmod(y ~ random(~ 1 | id) +
                 jump(t, psi ~ random(~ 1 | id), smoothed = smooth_probit(h)),
               distrib = gaussian1_distrib(), data = panel_j)
  hp <- hyper(f)
  c(h = h, position = coef(f)$mu[["jump.psi1.(Intercept)"]],
    sd_position = hp$estimate[grepl("jump", hp$term)],
    delta = coef(f)$mu[["jump.delta1"]])
}, numeric(4)))
round(tab_hr, 3)
```

The positions were drawn with a standard deviation of 0.8 (the sample
standard deviation of the twenty positions is `r fmt(sd(psi_i), 2)`) and the
change of level is 1.5. The width 0.71 is the default, the spacing of the
times.

### Which smoother to use

On the Nile the three smoothers give almost the same fit, because the width
is about one year and the series has one observation a year. They differ in
three properties, and these decide the choice.

* **How far from $\psi$ the fit is changed.** The probit step differs from
  the sharp step by a gaussian tail, so a few widths away from $\psi$ the two
  agree to rounding. The quintic step equals the sharp step exactly outside
  $[\psi - h, \psi + h]$. The hyperbolic step differs from the sharp step by
  a term that decays only as $c/u^2$, so it moves the fitted mean at every
  observation, also far from the break-point.
* **Whether $\psi$ always has a curvature.** The curvature in $\psi$ comes
  from observations inside the window of the smoother. The probit and the
  hyperbolic windows never end, so every observation contributes a little.
  The quintic window ends at $\pm h$, and if no observation falls inside it,
  $\psi$ has no curvature and is not identified; this is why its default
  width is wider, 3.61 spacings, which puts about five observations inside
  the window of a uniform covariate.
* **Whether a criterion over the hyperparameters is smooth.** The probit is
  an analytic function, so a REML criterion with a smoothed break-point, for
  example the standard deviation of random positions, has exact derivatives
  everywhere. The quintic has a fourth derivative that jumps at $\pm h$, and
  the criterion has small jumps in its gradient whenever an observation
  crosses $\psi \pm h$.

So `smooth_probit()` is the default to use: it changes the model only near
$\psi$, it always gives $\psi$ a curvature, and it keeps the REML criterion
smooth. Use `smooth_quintic()` when the model should be the sharp model away
from the break-point, for example when the observations near $\psi$ are few
and the fit elsewhere must not change, and there is no hyperparameter to
estimate by REML. `smooth_hyperbolic()` is the simplest form and has the
cheapest derivatives, but it changes the fit everywhere; it is rarely the
right choice for a break-point.

## The marginal likelihood of a random break-point {#sec-breakpoints-marginal}

The random positions of @sec-breakpoints-random are estimated as penalized
coefficients, one for each subject, and $\tau$ comes from the Laplace
approximation of REML. For a change of level this is not available, because
the indicator gives the log-likelihood no curvature in $\psi_i$. The argument
`marginal = TRUE` integrates each subject's position out of the likelihood
instead:
$$
L_i = \int \prod_j p\{y_{ij} \mid \mu_{ij}(\psi_i)\}\,
\frac{1}{\tau}\phi\!\left(\frac{\psi_i - m}{\tau}\right) d\psi_i .
$$
For `jump()` the integrand is constant between consecutive times, so the
integral is a finite sum over the intervals, each weighted by the normal
probability of the interval. For `seg()` and `jseg()` it is computed by
quadrature. The term then has the parameters $m$, $\tau$ and $\delta$ of the
population, and no coefficient per subject. The panel with a change of level
of the previous section, with $\psi_i$ drawn from $\mathrm{N}(5, 0.8)$ and
$\delta = 1.5$, is fitted with:

```{r}
#| label: fit-marginal
fit_marg <- statmod(y ~ random(~ 1 | id) +
                      jump(t, psi ~ random(~ 1 | id), marginal = TRUE),
                    distrib = gaussian1_distrib(), data = panel_j)
summary(fit_marg)
statmod_certificate(fit_marg)
```

The block of the term has three parameters, and they are the parameters of
the population of positions, not of any one subject. The positions of the
subjects follow $\psi_i \sim \mathrm{N}(m, \tau)$, and:

* `psi1.mean` is $m$, the mean of the positions in the population. It is
  the population position, the same quantity as `psi1.(Intercept)` in the
  model with a random position of @sec-breakpoints-random;
* `psi1.sd` is $\tau$, the standard deviation of the positions between
  subjects, the same quantity as the `sigma` of the random part of that
  model;
* `delta1` is $\delta$, the change of level, common to all subjects.

The number 1 refers to the first break-point; with `npsi = 2` the block also
has `psi2.mean` and `psi2.sd`. The summary has no row for the position of
each subject, because the model has no coefficient for it: each
$\psi_i$ has been integrated out of the likelihood. What the data say about
the position of subject $i$ is its posterior distribution given that
subject's observations and the fitted $m$, $\tau$, $\delta$ and $\sigma$. The
function `statmod_latent()` returns its mean and standard deviation, one row
per subject and break-point: the column `group` names the subject, `psi`
says which break-point the row is about (here always the first), and `mean`
and `sd` are the posterior mean and standard deviation of that subject's
position:

```{r}
#| label: marginal-latent
lat <- statmod_latent(fit_marg)
head(lat)
psi_post <- lat$mean[match(levels(panel_j$id), as.character(lat$group))]
cbind(posterior_mean = psi_post, true = psi_i)[1:6, ]
cor(psi_post, psi_i)
```

The posterior mean of $\psi_i$ is a weighted average of the intervals
between consecutive times, each interval weighted by how well a step there
fits the observations of subject $i$ and by its probability under
$\mathrm{N}(m, \tau)$. A subject whose step is clear in the data has a
small posterior standard deviation; a subject whose step is hidden by noise
has a posterior mean pulled towards $m$ and a larger standard deviation.

The fitted values of the marginal model are the posterior means of the
predictor at the observed times: for subject $i$ at time $t_{ij}$,
$$
\mathbb{E}(\mu_{ij} \mid y_i) = \beta_0 + b_i +
\delta\, P(\psi_i < t_{ij} \mid y_i),
$$
which rises from one level to the other over the times where the posterior
of $\psi_i$ has its mass. @fig-marginal-subjects shows six subjects with
their observations, these fitted values, and the posterior mean and
standard deviation of their positions.

```{r}
#| label: fig-marginal-subjects
#| fig-cap: "Six subjects of the panel with a change of level, with the fitted values of the marginal model (the posterior mean of the predictor), the posterior mean of the position (dashed) with one posterior standard deviation on each side (shaded), and the position used in the simulation (dotted)."
#| echo: false
#| fig-height: 5
#| fig-width: 9
sel_m <- sort(order(psi_i)[c(1, 4, 8, 12, 16, 20)])
lab_m <- paste("subject", sel_m)
pj <- panel_j
pj$fitted <- fitted(fit_marg, what = "mu")
pj <- pj[pj$id %in% sel_m, ]
pj$subject <- factor(paste("subject", pj$id), levels = lab_m)
post_m <- data.frame(subject = factor(lab_m, levels = lab_m),
                     mean = psi_post[sel_m],
                     sd = lat$sd[match(as.character(sel_m), as.character(lat$group))],
                     true = psi_i[sel_m])
ggplot(pj, aes(x = t)) +
  geom_rect(data = post_m, aes(xmin = mean - sd, xmax = mean + sd,
                               ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = book_palette[1], alpha = 0.15) +
  geom_point(aes(y = y), colour = book_grey) +
  geom_line(aes(y = fitted, colour = "fitted values"), linewidth = 0.9) +
  geom_vline(data = post_m, aes(xintercept = mean, linetype = "posterior mean")) +
  geom_vline(data = post_m, aes(xintercept = true, linetype = "true position")) +
  scale_linetype_manual(values = c(2, 3)) +
  facet_wrap(~ subject) +
  labs(x = "t", y = "y")
```

The smoothed step of @sec-breakpoints-smooth gives the other route: with
`smoothed = smooth_probit()` the positions of the subjects are penalized
coefficients again, as for `seg()` in @sec-breakpoints-random.

```{r}
#| label: fit-smooth-random
fit_srand <- statmod(y ~ random(~ 1 | id) +
                       jump(t, psi ~ random(~ 1 | id),
                            smoothed = smooth_probit()),
                     distrib = gaussian1_distrib(), data = panel_j)
summary(fit_srand)
cf_sr <- coef(fit_srand)$mu
h_sr <- hyper(fit_srand)
cf_mg <- coef(fit_marg)$mu
cbind(marginal = c(position = cf_mg[["jump.psi1.mean"]], sd = cf_mg[["jump.psi1.sd"]],
                   delta = cf_mg[["jump.delta1"]]),
      smoothed = c(position = cf_sr[["jump.psi1.(Intercept)"]],
                   sd = h_sr$estimate[grepl("jump", h_sr$term)],
                   delta = cf_sr[["jump.delta1"]]),
      true = c(5, 0.8, 1.5))
```

```{r}
#| label: marginal-numbers
#| include: false
cf_marg <- coef(fit_marg)$mu
ce_marg <- statmod_certificate(fit_marg)
```

The population position is `r fmt(cf_marg[["jump.psi1.mean"]], 2)`, the standard
deviation of the positions `r fmt(cf_marg[["jump.psi1.sd"]], 2)` and the change
of level `r fmt(cf_marg[["jump.delta1"]], 2)`. The criterion of this model
has no exact derivative in the standard deviation of the random intercept.
`statmod()` therefore searches over that one hyperparameter with Brent's
method, and `statmod_certificate()` reads the gradient and the curvature of
the criterion by refitting the model at nearby values, which gives the
`curvature` reported above.

The two routes give similar estimates. The standard deviation of the
smoothed model is larger, because the smoothing spreads each step over a
width $h$, and a step spread in this way looks like extra variation in the
positions (@sec-breakpoints-smooth measured how this grows with $h$). The
marginal fit costs more: the integral is computed for every subject at every
evaluation. @fig-marginal shows the mean of the responses at each time with
the population mean of the marginal model,
$\beta_0 + \delta\,\Phi\{(t - m)/\tau\}$, and the average of the subject
curves of the smoothed model.

```{r}
#| label: fig-marginal
#| fig-cap: "Mean of the twenty responses at each time, with the population mean of the marginal model and the average of the subject curves of the smoothed model."
#| echo: false
#| fig-height: 3.6
tgrid <- seq(0, 10, length.out = 300)
pop_m <- cf_mg[["(Intercept)"]] +
  cf_mg[["jump.delta1"]] * pnorm((tgrid - cf_mg[["jump.psi1.mean"]]) /
                                   cf_mg[["jump.psi1.sd"]])
avg_s <- numeric(length(tgrid))
for (i in levels(panel_j$id)) {
  nd <- data.frame(t = tgrid, id = factor(i, levels = levels(panel_j$id)))
  avg_s <- avg_s + predict(fit_srand, what = "mu", newdata = nd) / 20
}
ybar <- data.frame(t = sort(unique(panel_j$t)),
                   y = as.numeric(tapply(panel_j$y, panel_j$t, mean)))
marg_lines <- rbind(data.frame(t = tgrid, mu = pop_m, model = "marginal model"),
                    data.frame(t = tgrid, mu = avg_s, model = "smoothed model"))
ggplot(ybar, aes(x = t, y = y)) +
  geom_point(colour = book_grey) +
  geom_line(data = marg_lines, aes(y = mu, colour = model, linetype = model),
            linewidth = 0.9) +
  labs(x = "t", y = "mean response")
```

## A break-point in the standard deviation {#sec-breakpoints-sigma}

A break-point term can sit in the equation of $\sigma$. The data frame
`mcycle` gives the acceleration of the head of a crash-test dummy against the
time after impact [@silverman1985]. The variance is small before the impact
and large after it. We keep a smooth for the mean and try the three
break-point terms for $\log\sigma$:

```{r}
#| label: fit-mcycle-sigma
data(mcycle, package = "MASS")
fit_ss <- statmod(accel ~ s(times, bspline_smooth(k = 20)) | sigma ~ seg(times),
                  distrib = gaussian1_distrib(), data = mcycle)
summary(fit_ss)
fit_sj <- statmod(accel ~ s(times, bspline_smooth(k = 20)) | sigma ~ jump(times),
                  distrib = gaussian1_distrib(), data = mcycle)
summary(fit_sj)
fit_sjs <- statmod(accel ~ s(times, bspline_smooth(k = 20)) | sigma ~ jseg(times),
                   distrib = gaussian1_distrib(), data = mcycle)
summary(fit_sjs)
c(seg = AIC(fit_ss), jump = AIC(fit_sj), jseg = AIC(fit_sjs))
```

```{r}
#| label: mcycle-numbers
#| include: false
cf_sjs <- coef(fit_sjs)$sigma
```

The AIC is smallest for `jseg()`: at `r fmt(cf_sjs[["jseg.psi1"]], 1)` ms the
logarithm of $\sigma$ jumps by `r fmt(cf_sjs[["jseg.delta1"]], 2)`, and its
slope changes from `r fmt(cf_sjs[["jseg.beta"]], 3)` to
`r fmt(cf_sjs[["jseg.beta"]] + cf_sjs[["jseg.gamma1"]], 3)` per ms. In the
equation of $\sigma$ the residual sum of squares is not the objective, so the
positions are located on the working response of that equation and then
compared on the REML criterion itself (@sec-breakpoints-locate). The test of
@sec-breakpoints-seg reads the score of the equation of $\sigma$ in the same
way, and it confirms that the standard deviation changes its slope:

```{r}
#| label: test-mcycle
statmod_breakpoint_test(fit_ss)
```

@fig-mcycle shows the data with the fitted mean and the band of two standard
deviations of the `jseg()` model, and the fitted standard deviation of the
three models.

```{r}
#| label: fig-mcycle
#| fig-cap: "Left: acceleration against time, with the fitted mean and the band of plus or minus two fitted standard deviations of the jseg() model. Right: fitted standard deviation of the acceleration against time for the three break-point terms in the equation of sigma."
#| echo: false
#| fig-height: 3.8
#| fig-width: 9
tt <- data.frame(times = seq(min(mcycle$times), max(mcycle$times), length.out = 400))
m_sjs <- predict(fit_sjs, what = "mu", newdata = tt)
sd_sjs <- predict(fit_sjs, what = "sigma", newdata = tt)
p_mean <- ggplot(mcycle, aes(x = times, y = accel)) +
  geom_point(colour = book_grey, size = 0.9) +
  geom_ribbon(data = data.frame(times = tt$times, lo = m_sjs - 2 * sd_sjs,
                                hi = m_sjs + 2 * sd_sjs),
              aes(x = times, ymin = lo, ymax = hi), inherit.aes = FALSE,
              fill = book_palette[1], alpha = 0.15) +
  geom_line(data = data.frame(times = tt$times, mu = m_sjs),
            aes(y = mu), colour = book_palette[1], linewidth = 0.9) +
  labs(x = "time (ms)", y = "acceleration (g)")
sd_lines <- rbind(
  data.frame(times = tt$times, model = "jseg()", sd = sd_sjs),
  data.frame(times = tt$times, model = "seg()",
             sd = predict(fit_ss, what = "sigma", newdata = tt)),
  data.frame(times = tt$times, model = "jump()",
             sd = predict(fit_sj, what = "sigma", newdata = tt)))
sd_lines$model <- factor(sd_lines$model, levels = c("jseg()", "seg()", "jump()"))
p_sd <- ggplot(sd_lines, aes(x = times, y = sd, colour = model,
                             linetype = model)) +
  geom_line(linewidth = 0.9) +
  labs(x = "time (ms)", y = expression(hat(sigma)))
patchwork::wrap_plots(p_mean, p_sd, ncol = 2)
```

```{r}
#| label: gate-11
#| include: false
assert_breakpoint_ok(
  fits = list(gag = fit_gag, gag2 = fit_gag2, down = fit_down, nile = fit_nile,
              jseg = fit_jseg, seg0 = fit_seg0, jump0 = fit_jump0,
              coal = fit_coal, gj = fit_gj, gs = fit_gs,
              glob = fit_glob, glob_0 = fit_glob_0, glob_k = fits_glob,
              by = fit_by, psi = fit_psi, pan = fit_pan, np = fit_np,
              nq = fit_nq, nh = fit_nh, pg = fit_pg, pg_id = fit_pg_id,
              og = fit_og, ot = fit_ot,
              marg = fit_marg, srand = fit_srand, ss = fit_ss,
              sj = fit_sj, sjs = fit_sjs),
  ref = list(s_gag = s_gag, s_gag2 = s_gag2, s_down = s_down, s_nile = s_nile,
             s_coal = s_coal, s_glob = s_glob, s_glob4 = s_glob4,
             s_sel = s_sel, s_plant = s_plant, s_pan = s_pan),
  data = list(GAGurine = GAGurine, nile = nile, globTempAnom = globTempAnom,
              plant = plant, mcycle = mcycle, panel = panel,
              panel_j = panel_j),
  claims = list(lr = lr, rss_gag = rss_gag, rss_gag2 = rss_gag2,
                rss_nile = rss_nile, psi_star = psi_star, share = share,
                lr_g = lr_g, test_gag = test_gag, tab_npsi = tab_npsi,
                k_bic = k_bic, k_aic = k_aic, tab_h = tab_h, tab_hr = tab_hr,
                psi_hat = psi_hat, psi_o = psi_o, nu_ot = nu_ot,
                test_nile = test_nile))
```
