# Nonlinear models {#sec-nonlinear}

```{r}
#| label: setup-12
#| include: false
source("R/_setup.R")
source("R/nonlinear-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"; ink <- "grey15"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

In a nonlinear model a parameter of the distribution, on the scale of its
link, is a nonlinear function of the coefficients. The parameter is often the
mean, but it can be any of them: a standard deviation that decays with time,
or a probability that approaches an asymptote. Many models of this kind come
from a mechanism: the rate of an enzymatic reaction, the growth of a tree, the
concentration of a drug in the blood. Their coefficients have a meaning of
their own, such as a maximum rate or a final size, and the form of the
function is fixed by the mechanism. In `statmodels7` such a function is a term
of the formula, `nl()`, which takes the formula of the curve as a function of
its parameters. Like every other term, `nl()` can sit in the equation of any
parameter of the distribution, and each of its parameters can have a link,
covariates, random effects and penalties of its own. The examples of this
chapter put `nl()` in the equation of the mean.

We compare our fits with `nls()` of the package `stats` and with `nlme()` of
the package `nlme` [@bates1988; @pinheiro2000].

## A nonlinear mean: nl() {#sec-nonlinear-nl}

The data frame `Puromycin` gives the initial rate of an enzymatic reaction,
in counts per minute per minute, against the concentration of the substrate,
in parts per million [@bates1988]. We use the
`r sum(Puromycin$state == "treated")` measurements on cells treated with the
antibiotic puromycin. The Michaelis–Menten model describes the rate as
$$
y_i \sim \mathrm{N}(\mu_i, \sigma), \qquad
\mu_i = \frac{V_m\, c_i}{K + c_i},
$$
where $c_i$ is the concentration. The parameter $V_m$ is the largest rate,
which the reaction approaches as the concentration grows, and $K$ is the
concentration at which the rate reaches half of $V_m$.

```{r}
#| label: nl-fit-mm
puro <- Puromycin[Puromycin$state == "treated", ]
fit_mm <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc)),
                  distrib = gaussian1_distrib(), data = puro)
summary(fit_mm)
```

The first argument of `nl()` is a one-sided formula. The names in it that are
not columns of the data, here `Vm` and `K`, are the parameters of the curve,
and their coefficients are called `nl.Vm` and `nl.K`. The formula of the model
starts with `0` because the curve has no intercept. The derivatives of the
curve in its parameters come from `deriv()`, which differentiates the formula
symbolically.

We give no starting values, and `statmod()` computes them from the data. It
takes the response on the scale of the predictor, which here is the rate
itself because the link of $\mu$ is the identity. The curve is linear in
$V_m$ when $K$ is fixed, so for every value of $K$ the best $V_m$ is a least
squares solution. The parameters in which the curve is not linear, here $K$,
are searched on a grid of about 600 points on the scale of their links. The
grid covers a range built from the quantiles, the standard deviation and the
range of the response and of the covariates. The twelve best points of the
grid are refined by a one-dimensional search, or by the Nelder–Mead method
when there are several such parameters, and the best of them is the start.
A value written in the argument `start` is kept as it is, and only the other
parameters are searched.

The function `nls()` fits the same model by least squares:

```{r}
#| label: nl-nls-mm
nls_mm <- nls(rate ~ Vm * conc / (K + conc), data = puro,
              start = list(Vm = 200, K = 0.05))
summary(nls_mm)
```

We put the estimates, the standard errors and the two estimates of $\sigma$
side by side:

```{r}
#| label: nl-compare-mm
ci_mm <- confint(fit_mm)
cbind(statmod = ci_mm[c("mu:nl.Vm", "mu:nl.K"), "estimate"],
      nls = coef(nls_mm),
      se_statmod = ci_mm[c("mu:nl.Vm", "mu:nl.K"), "se"],
      se_nls = summary(nls_mm)$coefficients[, "Std. Error"])
c(statmod = exp(coef(fit_mm)$sigma[["(Intercept)"]]), nls = sigma(nls_mm))
c(statmod = logLik(fit_mm), nls = logLik(nls_mm))
```

```{r}
#| label: nl-mm-numbers
#| include: false
cf_mm <- coef(fit_mm)$mu
sig_mm <- exp(coef(fit_mm)$sigma[["(Intercept)"]])
```

The estimates and the standard errors agree with those of `nls()` to four
significant digits. The rate approaches $\hat V_m = `r fmt(cf_mm[["nl.Vm"]], 1)`$
counts per minute per minute, and it is half of that at
$\hat K = `r fmt(cf_mm[["nl.K"]], 4)`$ parts per million. The two estimates
of $\sigma$, `r fmt(sig_mm, 5)` and `r fmt(sigma(nls_mm), 5)`, differ in the
fourth decimal place. Both are the restricted maximum
likelihood estimate, which for a gaussian model with $p = 2$ coefficients is
$\{\sum_i (y_i - \hat\mu_i)^2/(n - p)\}^{1/2}$. `nls()` computes it in closed
form, while `statmod()` maximizes the restricted likelihood numerically and
stops when the criterion no longer changes beyond its tolerance, which leaves
the last digits of $\hat\sigma$ approximate. The two log-likelihoods differ
for another reason: ours is evaluated at $\hat\sigma = `r fmt(sig_mm, 3)`$,
and `logLik()` of `nls()` is evaluated at the maximum likelihood estimate
$\{\sum_i (y_i - \hat\mu_i)^2/n\}^{1/2}$.
@fig-nl-mm shows the fitted curve with a 95% confidence band, which we obtain
from `predict()` with `se = TRUE`.

```{r}
#| label: fig-nl-mm
#| fig-cap: "Initial rate against the concentration of the substrate for treated cells, with the fitted Michaelis–Menten curve and its 95% confidence band."
#| echo: false
#| fig-height: 3.8
gc <- data.frame(conc = seq(0, max(puro$conc), length.out = 200))
band_mm <- predict(fit_mm, what = "mu", newdata = gc, se = TRUE)
band_df <- data.frame(conc = gc$conc, fit = band_mm$fit,
                      lower = band_mm$lower, upper = band_mm$upper)
ggplot(puro, aes(x = conc, y = rate)) +
  geom_ribbon(data = band_df, aes(x = conc, ymin = lower, ymax = upper,
                                  fill = "95% band"),
              inherit.aes = FALSE, alpha = 0.2) +
  geom_point(colour = book_grey) +
  geom_line(data = band_df, aes(x = conc, y = fit, colour = "fitted curve"),
            linewidth = 0.9) +
  scale_colour_manual(values = chalk) +
  scale_fill_manual(values = chalk) +
  coord_cartesian(ylim = c(0, 230)) +
  labs(x = "concentration (ppm)", y = "rate (counts/min/min)")
```

## Links on the parameters {#sec-nonlinear-links}

Both parameters of the Michaelis–Menten curve are positive. The argument
`links` of `nl()` gives each parameter a link from `linkfunctions7`, so that
the coefficient is the parameter on the scale of the link. With the
logarithm, the coefficient can take any real value and the parameter stays
positive:

```{r}
#| label: nl-fit-mm-log
fit_mm_log <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc),
                                    links = list(Vm = log_link(),
                                                 K = log_link())),
                      distrib = gaussian1_distrib(), data = puro)
summary(fit_mm_log)
exp(coef(fit_mm_log)$mu)
c(identity = logLik(fit_mm), log = logLik(fit_mm_log))
```

The fitted curve is the same, so the log-likelihood is the same, and the
exponentials of the two coefficients are the estimates of the previous
section. What changes is the scale on which the standard errors and the
intervals are computed. An interval for $\log K$, carried back with the
exponential, is not symmetric about $\hat K$ and contains only positive
values:

```{r}
#| label: nl-ci-mm-log
ci_log <- confint(fit_mm_log)
exp(ci_log["mu:nl.K", c("estimate", "lower", "upper")])
ci_mm["mu:nl.K", c("estimate", "lower", "upper")]
```

A parameter without a link keeps the identity. A link also helps the fit
itself, because the optimizer cannot propose a negative value for a
parameter that must be positive.

### A gamma response

The gaussian model gives every observation the same variance. For a rate it
is often more natural to have a constant coefficient of variation, that is a
standard deviation proportional to the mean. The family `gamma1_distrib()`
has mean $\mu$ and variance $\phi\mu^2$. The curve gives the mean itself, so
we set the link of $\mu$ to the identity. The identity link does not keep
$\mu$ positive by itself, but the curve does: with $V_m > 0$ and $K > 0$, the
mean is positive at every positive concentration. We keep $K$ on the
logarithmic scale, so that it stays positive whatever its coefficient:
$$
y_i \sim \mathrm{Gamma}(\mu_i, \phi), \qquad
\mu_i = \frac{V_m\, c_i}{K + c_i}.
$$

```{r}
#| label: nl-fit-mm-gamma
fit_mm_gam <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc),
                                    links = list(K = log_link())),
                      distrib = gamma1_distrib(link_mu = identity_link()),
                      data = puro, outer_criterion = ml())
summary(fit_mm_gam)
```

This model is also a generalized linear model. The inverse of the mean is
$1/\mu_i = 1/V_m + (K/V_m)(1/c_i)$, which is linear in $1/c_i$, so `glm()`
with the gamma family and the inverse link fits it, and its two coefficients
give $V_m$ and $K$ [@mccullagh1989]:

```{r}
#| label: nl-glm-mm-gamma
glm_mm <- glm(rate ~ I(1 / conc), family = Gamma(link = "inverse"),
              data = puro)
summary(glm_mm)
b_glm <- coef(glm_mm)
cbind(statmod = c(Vm = coef(fit_mm_gam)$mu[["nl.Vm"]],
                  K = exp(coef(fit_mm_gam)$mu[["nl.K"]])),
      glm = c(Vm = 1 / b_glm[[1]], K = b_glm[[2]] / b_glm[[1]]))
c(statmod = logLik(fit_mm_gam), glm = logLik(glm_mm))
```

```{r}
#| label: nl-gamma-numbers
#| include: false
phi_gam <- exp(coef(fit_mm_gam)$phi[["(Intercept)"]])
```

The estimates of $V_m$ and $K$ are the same. We fit the model under `ml()`,
so $\hat\phi = `r fmt(phi_gam, 5)`$ is the maximum likelihood estimate.
`logLik()` of `glm()` evaluates the log-likelihood at the deviance divided by
$n$, `r fmt(deviance(glm_mm) / nrow(puro), 5)`, which is close to it, and the
two log-likelihoods agree to four decimals. @fig-nl-gamma compares the
gaussian and the gamma fits.

```{r}
#| label: fig-nl-gamma
#| fig-cap: "Initial rate against the concentration for treated cells, with the Michaelis–Menten curves fitted with a gaussian response and with a gamma response."
#| echo: false
#| fig-height: 3.8
curves_gam <- data.frame(
  conc = rep(gc$conc, times = 2),
  rate = c(predict(fit_mm, what = "mu", newdata = gc),
           predict(fit_mm_gam, what = "mu", newdata = gc)),
  curve = factor(rep(c("gaussian", "gamma"), each = nrow(gc)),
                 levels = c("gaussian", "gamma")))
ggplot(puro, aes(x = conc, y = rate)) +
  geom_point(colour = book_grey) +
  geom_line(data = curves_gam, aes(colour = curve, linetype = curve),
            linewidth = 0.9) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c(1, 2)) +
  coord_cartesian(ylim = c(0, 230)) +
  labs(x = "concentration (ppm)", y = "rate (counts/min/min)")
```

## Parameters with covariates {#sec-nonlinear-covariates}

The data frame `Puromycin` also has `r sum(Puromycin$state == "untreated")`
measurements on untreated cells. Each parameter of the curve can depend on
covariates through a formula of its own, written as a further argument of
`nl()` with the parameter on the left-hand side. With `Vm ~ state` and
`K ~ state` each parameter has an intercept, the value for treated cells, and
a difference for untreated cells:
$$
\mu_i = \frac{V_{m,i}\, c_i}{K_i + c_i}, \qquad
V_{m,i} = v_0 + v_1 u_i, \qquad K_i = k_0 + k_1 u_i,
$$
where $u_i$ is 1 for untreated cells and 0 otherwise.

```{r}
#| label: nl-fit-state
fit_st <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc), Vm ~ state,
                                K ~ state),
                  distrib = gaussian1_distrib(), data = Puromycin,
                  outer_criterion = ml())
summary(fit_st)
```

We fit the model under `ml()`, because below we compare it with a smaller
model by the likelihood ratio, which needs the maximized log-likelihood of
both. Under `ml()` the estimate of $\sigma$ is the maximum likelihood
estimate. The function `nls()` fits the same model when each parameter is
indexed by the factor:

```{r}
#| label: nl-nls-state
nls_st <- nls(rate ~ Vm[state] * conc / (K[state] + conc), data = Puromycin,
              start = list(Vm = c(200, 160), K = c(0.05, 0.05)))
summary(nls_st)
cf_st <- coef(fit_st)$mu
cbind(statmod = c(cf_st[["nl.Vm.(Intercept)"]],
                  cf_st[["nl.Vm.(Intercept)"]] + cf_st[["nl.Vm.stateuntreated"]],
                  cf_st[["nl.K.(Intercept)"]],
                  cf_st[["nl.K.(Intercept)"]] + cf_st[["nl.K.stateuntreated"]]),
      nls = coef(nls_st))
c(statmod = logLik(fit_st), nls = logLik(nls_st))
```

The two fits agree, and so do the log-likelihoods, which are both maximized
over $\sigma$. The summary suggests that $K$ does not depend on the
treatment. The smaller model has `K` without a formula, so $K$ is the same
for the two states:

```{r}
#| label: nl-fit-state0
fit_st0 <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc), Vm ~ state),
                   distrib = gaussian1_distrib(), data = Puromycin,
                   outer_criterion = ml())
summary(fit_st0)
l0 <- logLik(fit_st0)
l1 <- logLik(fit_st)
lr_st <- 2 * (as.numeric(l1) - as.numeric(l0))
lr_st
pchisq(lr_st, df = 1, lower.tail = FALSE)
```

```{r}
#| label: nl-state-numbers
#| include: false
p_st <- pchisq(lr_st, df = 1, lower.tail = FALSE)
cf_st0 <- coef(fit_st0)$mu
```

The likelihood ratio is `r fmt(lr_st, 3)`, with a p-value of
`r fmt(p_st, 3)`, so the data are compatible with a common $K$. In the
smaller model the treatment raises the largest rate from
`r fmt(cf_st0[["nl.Vm.(Intercept)"]] + cf_st0[["nl.Vm.stateuntreated"]], 1)`
to `r fmt(cf_st0[["nl.Vm.(Intercept)"]], 1)`, and the common $K$ is
`r fmt(cf_st0[["nl.K"]], 4)`. @fig-nl-state shows the two models, one panel
for each state.

```{r}
#| label: fig-nl-state
#| fig-cap: "Initial rate against the concentration, one panel for each state of the cells, with the curves of the model in which both parameters depend on the state and of the model with a common K."
#| echo: false
#| fig-height: 3.6
#| fig-width: 8
gs <- expand.grid(conc = gc$conc, state = levels(Puromycin$state))
gs$state <- factor(gs$state, levels = levels(Puromycin$state))
st_names <- c("Vm and K by state", "common K")
curves_st <- data.frame(
  conc = rep(gs$conc, times = 2), state = rep(gs$state, times = 2),
  rate = c(predict(fit_st, what = "mu", newdata = gs),
           predict(fit_st0, what = "mu", newdata = gs)),
  curve = factor(rep(st_names, each = nrow(gs)), levels = st_names))
ggplot(Puromycin, aes(x = conc, y = rate)) +
  geom_point(colour = book_grey) +
  geom_line(data = curves_st, aes(colour = curve, linetype = curve),
            linewidth = 0.9) +
  facet_wrap(~state, nrow = 1) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c(1, 2)) +
  coord_cartesian(ylim = c(0, 230)) +
  labs(x = "concentration (ppm)", y = "rate (counts/min/min)")
```

The formula of a parameter takes any term of the package, not only factors:
`K ~ x` for a continuous covariate, `K ~ s(x)` for a smooth function of it.
The formula acts on the scale of the parameter's link, so with a logarithmic
link each covariate multiplies the parameter.

## Random effects on the parameters {#sec-nonlinear-random}

In data with groups, a parameter of the curve can vary from group to group.
The data frame `Orange` gives the circumference of the trunk, in millimeters,
of five orange trees at seven ages, in days [@draper1998; @pinheiro2000]. The
logistic growth curve describes the circumference as
$$
\mu_{ij} = \frac{A_i}{1 + \exp\{(x_m - a_{ij})/s\}} ,
$$
where $a_{ij}$ is the age of tree $i$ at its $j$-th measurement. The
parameter $A_i$ is the final circumference of tree $i$, $x_m$ is the age at
which a tree reaches half of it, and $s$ sets how fast it grows. We let each
tree have its own final circumference, $A_i = A + b_i$, with
$b_i \sim \mathrm{N}(0, \tau)$. In the formula of `Asym` we write the
random effect with `random()`:

```{r}
#| label: nl-fit-orange
fit_or <- statmod(circumference ~ 0 +
                    nl(~ Asym / (1 + exp((xmid - age) / scal)),
                       Asym ~ 1 + random(~ 1 | Tree)),
                  distrib = gaussian1_distrib(), data = Orange)
summary(fit_or)
```

The summary has a compartment for `Asym`. It gives the population value $A$,
the `(Intercept)` of its formula, and the standard deviation $\tau$ of the
random effects, which `statmod()` estimates by restricted maximum likelihood
with the other variance parameters. The function `nlme()` fits the same model
[@lindstrom1990; @pinheiro2000]:

```{r}
#| label: nl-nlme-orange
nlme_or <- nlme::nlme(circumference ~ SSlogis(age, Asym, xmid, scal),
                      data = Orange, fixed = Asym + xmid + scal ~ 1,
                      random = Asym ~ 1 | Tree,
                      start = c(Asym = 190, xmid = 700, scal = 350),
                      method = "REML")
summary(nlme_or)
cbind(statmod = c(coef(fit_or)$mu[c("nl.Asym.(Intercept)", "nl.xmid",
                                    "nl.scal")],
                  tau = hyper(fit_or)$estimate,
                  sigma = exp(coef(fit_or)$sigma[["(Intercept)"]])),
      nlme = c(nlme::fixef(nlme_or),
               as.numeric(nlme::VarCorr(nlme_or)[, "StdDev"])))
c(statmod = logLik(fit_or, type = "marginal"), nlme = logLik(nlme_or))
```

```{r}
#| label: nl-orange-numbers
#| include: false
tau_or <- hyper(fit_or)$estimate
tau_nlme <- as.numeric(nlme::VarCorr(nlme_or)[1, "StdDev"])
ll_or <- as.numeric(logLik(fit_or, type = "marginal"))
ll_nlme <- as.numeric(logLik(nlme_or))
```

The fixed parameters are close, and the standard deviation of the trees is
`r fmt(tau_or, 2)` in our fit against `r fmt(tau_nlme, 2)` in that of
`nlme()`. The two packages maximize the same approximation to the restricted
likelihood: the restricted likelihood of the linear mixed model obtained by
replacing the curve with its first-order expansion in the coefficients and in
the random effects [@lindstrom1990]. The algorithm of `nlme()` alternates a
penalized nonlinear least squares step and a step on the linear mixed model,
and it stops at a point where the criterion is
`r fmt(ll_nlme, 3)`. Our estimates give `r fmt(ll_or, 3)`, a larger value.
The criterion changes little with $\tau$, so a small difference in its value
goes with a visible difference in $\tau$. With five trees the data carry
little information about the variance between trees.

The curve is linear in the random effects $b_i$, so for given $x_m$ and $s$
the likelihood integrated over the $b_i$ has a closed form, and maximum
likelihood involves no approximation. We fit the model by maximum
likelihood with both packages:

```{r}
#| label: nl-ml-orange
fit_or_ml <- statmod(circumference ~ 0 +
                       nl(~ Asym / (1 + exp((xmid - age) / scal)),
                          Asym ~ 1 + random(~ 1 | Tree)),
                     distrib = gaussian1_distrib(), data = Orange,
                     outer_criterion = ml())
summary(fit_or_ml)
nlme_or_ml <- nlme::nlme(circumference ~ SSlogis(age, Asym, xmid, scal),
                         data = Orange, fixed = Asym + xmid + scal ~ 1,
                         random = Asym ~ 1 | Tree,
                         start = c(Asym = 190, xmid = 700, scal = 350),
                         method = "ML")
summary(nlme_or_ml)
cbind(statmod = c(coef(fit_or_ml)$mu[c("nl.Asym.(Intercept)", "nl.xmid",
                                       "nl.scal")],
                  tau = hyper(fit_or_ml)$estimate,
                  sigma = exp(coef(fit_or_ml)$sigma[["(Intercept)"]])),
      nlme = c(nlme::fixef(nlme_or_ml),
               as.numeric(nlme::VarCorr(nlme_or_ml)[, "StdDev"])))
c(statmod = logLik(fit_or_ml, type = "marginal"), nlme = logLik(nlme_or_ml))
```

The two maxima differ by less than 0.001, ours being the larger, and the
estimates agree to three significant digits.

The argument `random` of `predict()` sets the random effects aside. With
`random = "zero"` every tree is given $b_i = 0$, which is the curve of a
typical tree, also for a tree that the fit never saw. @fig-nl-orange shows
three trees, the one with the smallest final circumference, one in the
middle and the one with the largest, each with its own curve and with the
curve of the typical tree.

```{r}
#| label: fig-nl-orange
#| fig-cap: "Trunk circumference against age for three orange trees, each with the curve that the model fits to it and with the curve of a typical tree, which has the random effect at zero."
#| echo: false
#| fig-height: 3.6
#| fig-width: 9
b_or <- coef(fit_or)$mu[paste0("nl.Asym.random.", levels(Orange$Tree))]
names(b_or) <- levels(Orange$Tree)
sel_or <- names(sort(b_or))[c(1, 3, 5)]
sel_or <- sel_or[order(as.numeric(sel_or))]
ag <- seq(100, 1600, length.out = 200)
or_names <- c("this tree", "typical tree")
curves_or <- do.call(rbind, lapply(sel_or, function(tr) {
  nd <- data.frame(age = ag, Tree = factor(tr, levels = levels(Orange$Tree)))
  data.frame(age = rep(ag, times = 2), tree = tr,
             circumference = c(predict(fit_or, what = "mu", newdata = nd),
                               predict(fit_or, what = "mu", newdata = nd,
                                       random = "zero")),
             curve = rep(or_names, each = length(ag)))
}))
curves_or$curve <- factor(curves_or$curve, levels = or_names)
obs_or <- data.frame(age = Orange$age, tree = as.character(Orange$Tree),
                     circumference = Orange$circumference)
obs_or <- obs_or[obs_or$tree %in% sel_or, ]
curves_or$panel <- factor(paste("tree", curves_or$tree),
                          levels = paste("tree", sel_or))
obs_or$panel <- factor(paste("tree", obs_or$tree),
                       levels = paste("tree", sel_or))
ggplot(obs_or, aes(x = age, y = circumference)) +
  geom_point(colour = book_grey) +
  geom_line(data = curves_or, aes(colour = curve, linetype = curve),
            linewidth = 0.9) +
  facet_wrap(~panel, nrow = 1) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c(1, 2)) +
  coord_cartesian(xlim = range(ag), ylim = range(Orange$circumference)) +
  labs(x = "age (days)", y = "circumference (mm)")
```

### Random effects on two parameters

The data frame `Theoph` gives the concentration of theophylline in the blood,
in milligrams per liter, of twelve subjects at eleven times after an oral
dose, in hours [@pinheiro2000]. The dose, in milligrams per kilogram of body
weight, varies between subjects. The one-compartment model with first-order
absorption describes the concentration at time $t$ after a dose $D$ as
$$
\mu = \frac{D\, k_e k_a}{C (k_a - k_e)}
      \left(e^{-k_e t} - e^{-k_a t}\right),
$$
where $k_a$ is the rate of absorption, $k_e$ the rate of elimination and $C$
the clearance. The three are positive, so we write the curve in their
logarithms, `lKa`, `lKe` and `lCl`, as `SSfol()` of the package `stats` does.
We give the absorption and the clearance a random effect for each subject,
with two separate variances:

```{r}
#| label: nl-fit-theoph
fit_th <- statmod(conc ~ 0 +
                    nl(~ Dose * exp(lKe + lKa - lCl) *
                         (exp(-exp(lKe) * Time) - exp(-exp(lKa) * Time)) /
                         (exp(lKa) - exp(lKe)),
                       lKa ~ 1 + random(~ 1 | Subject),
                       lCl ~ 1 + random(~ 1 | Subject),
                       start = list(lKe = -2.5, lKa = 0.5, lCl = -3)),
                  distrib = gaussian1_distrib(), data = Theoph)
summary(fit_th)
```

Here we give starting values through the argument `start`, on the scale of
the parameters. The random effects enter the curve through the exponential
of `lKa` and `lCl`, so the curve is not linear in them. We fit the same model
with `nlme()`:

```{r}
#| label: nl-nlme-theoph
nlme_th <- nlme::nlme(conc ~ SSfol(Dose, Time, lKe, lKa, lCl),
                      data = Theoph, fixed = lKe + lKa + lCl ~ 1,
                      random = nlme::pdDiag(lKa + lCl ~ 1),
                      groups = ~ Subject,
                      start = c(lKe = -2.5, lKa = 0.5, lCl = -3),
                      method = "REML")
summary(nlme_th)
h_th <- hyper(fit_th)
cbind(statmod = c(coef(fit_th)$mu[c("nl.lKe", "nl.lKa.(Intercept)",
                                    "nl.lCl.(Intercept)")],
                  tau_lKa = h_th$estimate[1], tau_lCl = h_th$estimate[2],
                  sigma = exp(coef(fit_th)$sigma[["(Intercept)"]])),
      nlme = c(nlme::fixef(nlme_th),
               as.numeric(nlme::VarCorr(nlme_th)[, "StdDev"])))
c(statmod = logLik(fit_th, type = "marginal"), nlme = logLik(nlme_th))
```

```{r}
#| label: nl-theoph-numbers
#| include: false
ll_th <- as.numeric(logLik(fit_th, type = "marginal"))
ll_th_nlme <- as.numeric(logLik(nlme_th))
```

The comparison is the one of the orange trees: the estimates are close, and
our restricted criterion, `r fmt(ll_th, 3)`, is larger than the one at the
estimates of `nlme()`, `r fmt(ll_th_nlme, 3)`.

When the curve is not linear in the random effects, the curve of the typical
subject, with the effects at zero, is not the average curve over the
subjects. With `random = "marginal"`, `predict()` averages the curve over
the distribution of the random effects, with a Gauss–Hermite rule on a grid
of values of the two effects. @fig-nl-theoph shows three subjects, each
with its own curve, the curve of the typical subject and the average curve.

```{r}
#| label: fig-nl-theoph
#| fig-cap: "Concentration of theophylline against time for three subjects, each with the curve that the model fits to it, the curve of a typical subject (random effects at zero) and the curve averaged over the distribution of the random effects, all at the dose of that subject."
#| echo: false
#| fig-height: 3.6
#| fig-width: 9
cl_th <- coef(fit_th)$mu[paste0("nl.lCl.random.", levels(Theoph$Subject))]
names(cl_th) <- levels(Theoph$Subject)
sel_th <- names(sort(cl_th))[c(1, 6, 12)]
sel_th <- sel_th[order(as.numeric(sel_th))]
tt <- seq(0, 25, length.out = 200)
th_names <- c("this subject", "typical subject", "average")
curves_th <- do.call(rbind, lapply(sel_th, function(s) {
  rows <- Theoph$Subject == s
  nd <- data.frame(Time = tt, Dose = Theoph$Dose[rows][1],
                   Subject = factor(s, levels = levels(Theoph$Subject)))
  data.frame(Time = rep(tt, times = 3), subject = s,
             conc = c(predict(fit_th, what = "mu", newdata = nd),
                      predict(fit_th, what = "mu", newdata = nd,
                              random = "zero"),
                      predict(fit_th, what = "mu", newdata = nd,
                              random = "marginal")),
             curve = rep(th_names, each = length(tt)))
}))
curves_th$curve <- factor(curves_th$curve, levels = th_names)
obs_th <- data.frame(Time = Theoph$Time, subject = as.character(Theoph$Subject),
                     conc = Theoph$conc)
obs_th <- obs_th[obs_th$subject %in% sel_th, ]
curves_th$panel <- factor(paste("subject", curves_th$subject),
                          levels = paste("subject", sel_th))
obs_th$panel <- factor(paste("subject", obs_th$subject),
                       levels = paste("subject", sel_th))
ggplot(obs_th, aes(x = Time, y = conc)) +
  geom_point(colour = book_grey) +
  geom_line(data = curves_th, aes(colour = curve, linetype = curve,
                                  linewidth = curve)) +
  facet_wrap(~panel, nrow = 1) +
  scale_colour_manual(values = c(chalk, rust, ink)) +
  scale_linetype_manual(values = c(1, 2, 3)) +
  scale_linewidth_manual(values = c(0.9, 0.9, 0.7)) +
  coord_cartesian(ylim = c(0, 12)) +
  labs(x = "time (hours)", y = "concentration (mg/L)")
```

## A lasso on the levels of a parameter {#sec-nonlinear-lasso}

A random effect shrinks every group towards the population value, and every
group keeps a departure from it. When most groups share one value of a
parameter and only a few differ, a lasso on the levels can set some of the
departures exactly to zero. We simulate the initial rates of eight enzymes, each
measured twice at six concentrations, with a common $K = 0.07$ and with
$V_m = 200$ for six of them, 260 for enzyme C and 150 for enzyme F:

```{r}
#| label: nl-sim-lasso
set.seed(1)
enzyme <- factor(rep(LETTERS[1:8], each = 12))
conc <- rep(c(0.02, 0.06, 0.11, 0.22, 0.56, 1.10), 16)
vm_true <- c(A = 200, B = 200, C = 260, D = 200, E = 200, F = 150, G = 200,
             H = 200)
rate <- vm_true[enzyme] * conc / (0.07 + conc) + rnorm(96, sd = 10)
enz <- data.frame(enzyme = enzyme, conc = conc, rate = rate)
```

In the formula of `Vm` we write an intercept and a lasso on the levels of
`enzyme`. The term `lasso(~ enzyme)` has one coefficient for every level,
and the intercept is the value of the levels whose departure is zero:

```{r}
#| label: nl-fit-lasso
fit_las <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc),
                                 Vm ~ 1 + lasso(~ enzyme)),
                   distrib = gaussian1_distrib(), data = enz)
summary(fit_las)
```

The penalty is $\lambda\sum_g \lvert d_g\rvert$ on the departures $d_g$, and
$\lambda$ is chosen along a path by the BIC, as in @sec-penalized. We fit the
same model with a random effect in place of the lasso:

```{r}
#| label: nl-fit-lasso-random
fit_ran <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc),
                                 Vm ~ 1 + random(~ 1 | enzyme)),
                   distrib = gaussian1_distrib(), data = enz)
summary(fit_ran)
d_las <- coef(fit_las)$mu[paste0("nl.Vm.lasso.enzyme", LETTERS[1:8])]
d_ran <- coef(fit_ran)$mu[paste0("nl.Vm.random.", LETTERS[1:8])]
round(cbind(true = vm_true - 200, lasso = d_las, random = d_ran), 2)
```

```{r}
#| label: nl-lasso-numbers
#| include: false
zero_las <- LETTERS[1:8][d_las == 0]
kept_las <- LETTERS[1:8][d_las != 0]
```

The lasso sets exactly to zero the departures of `r length(zero_las)`
enzymes, `r paste(zero_las, collapse = ", ")`, and keeps those of
`r paste(kept_las, collapse = ", ")`. The departures of C and F are close to
the simulated ones, and those of the other enzymes that it keeps are small.
The random effect gives every enzyme a departure. @fig-nl-lasso shows the eight enzymes, one panel
each, with the curves of the two models.

```{r}
#| label: fig-nl-lasso
#| fig-cap: "Initial rate against the concentration for the eight simulated enzymes, one panel each, with the curves of the model with a lasso on the levels of Vm and of the model with a random effect on Vm."
#| echo: false
#| fig-height: 5
#| fig-width: 9
las_names <- c("lasso", "random effect")
curves_las <- do.call(rbind, lapply(LETTERS[1:8], function(e) {
  ne <- data.frame(conc = gc$conc[gc$conc <= 1.1],
                   enzyme = factor(e, levels = LETTERS[1:8]))
  data.frame(conc = rep(ne$conc, times = 2), panel = paste("enzyme", e),
             rate = c(predict(fit_las, what = "mu", newdata = ne),
                      predict(fit_ran, what = "mu", newdata = ne)),
             curve = rep(las_names, each = nrow(ne)))
}))
curves_las$curve <- factor(curves_las$curve, levels = las_names)
curves_las$panel <- factor(curves_las$panel,
                           levels = paste("enzyme", LETTERS[1:8]))
obs_las <- data.frame(conc = enz$conc, rate = enz$rate,
                      panel = factor(paste("enzyme", enz$enzyme),
                                     levels = paste("enzyme", LETTERS[1:8])))
ggplot(obs_las, aes(x = conc, y = rate)) +
  geom_point(colour = book_grey) +
  geom_line(data = curves_las, aes(colour = curve, linetype = curve),
            linewidth = 0.9) +
  facet_wrap(~panel, nrow = 2) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c(1, 2)) +
  coord_cartesian(ylim = c(0, 270)) +
  labs(x = "concentration", y = "rate")
```

## A break-point inside a parameter {#sec-nonlinear-breakpoint}

The formula of a parameter can also contain a break-point term from
@sec-breakpoints, so that the parameter changes at an estimated value of a
covariate. We simulate a sequence of 30 daily experiments. In each one a
count decays over eight times $x$, and from day 18 the decay is faster:
$$
y \sim \mathrm{Poisson}(\mu), \qquad \mu = a\, e^{-r x}, \qquad
\log r = r_0 + \delta\, \mathbb{1}(\text{day} > \psi).
$$

```{r}
#| label: nl-sim-decay
set.seed(2)
decay <- data.frame(day = rep(1:30, each = 8),
                    x = rep(seq(0.25, 4, length.out = 8), 30))
r_true <- exp(log(0.6) + 0.5 * (decay$day > 17))
decay$y <- rpois(240, 80 * exp(-r_true * decay$x))
```

Inside `nl()` the break-point term must be smoothed, because its block has to
be a Jacobian like the rest of the term. We take the smoothed step of
`smooth_probit()` (@sec-breakpoints-smooth). The counts are Poisson with the
identity link, since the curve gives the mean itself, and both $a$ and $r$
are on the logarithmic scale:

```{r}
#| label: nl-fit-decay
fit_dec <- statmod(y ~ 0 + nl(~ a * exp(-r * x),
                              r ~ jump(day, smoothed = smooth_probit()),
                              links = list(a = log_link(), r = log_link()),
                              start = list(a = 50, r = 1)),
                   distrib = poisson_distrib(link_mu = identity_link()),
                   data = decay)
summary(fit_dec)
statmod_certificate(fit_dec)$state
```

```{r}
#| label: nl-decay-numbers
#| include: false
cf_dec <- coef(fit_dec)$mu
```

The position of the change is $\hat\psi = `r fmt(cf_dec[["nl.r.jump.psi1"]], 2)`$,
between day 17 and day 18, and it has a standard error because the step is
smooth. The logarithm of the rate rises by
`r fmt(cf_dec[["nl.r.jump.delta1"]], 3)`, against 0.5 in the simulation. We
compare the model with the one without a change:

```{r}
#| label: nl-fit-decay0
fit_dec0 <- statmod(y ~ 0 + nl(~ a * exp(-r * x),
                               links = list(a = log_link(), r = log_link()),
                               start = list(a = 50, r = 1)),
                    distrib = poisson_distrib(link_mu = identity_link()),
                    data = decay)
summary(fit_dec0)
c(no_change = AIC(fit_dec0), change = AIC(fit_dec))
```

The AIC prefers the model with the change. The likelihood ratio between the
two models has no $\chi^2$ reference, because the position of the change is
not defined when there is no change [@davies1987]. @fig-nl-decay shows three
days, one before the change, the first day after it and one later, and the
rate of decay estimated for each day. We compute the rate from two
predictions, because $\mu(1)/\mu(0) = e^{-r}$.

```{r}
#| label: fig-nl-decay
#| fig-cap: "Left: the counts of three days with the fitted curves. Right: the rate of decay estimated for each day, with the rate used in the simulation."
#| echo: false
#| fig-height: 3.6
#| fig-width: 10
xs <- seq(0, 4, length.out = 100)
days <- seq(1, 30, length.out = 300)
m0 <- predict(fit_dec, what = "mu", newdata = data.frame(day = days, x = 0))
m1 <- predict(fit_dec, what = "mu", newdata = data.frame(day = days, x = 1))
dec_days <- c(10, 18, 28)
pan_dec <- c(paste("count against x, day", dec_days),
             "rate of decay r against day")
dec_names <- c("fitted", "simulated")
obs_dec <- decay[decay$day %in% dec_days, c("day", "x", "y")]
obs_dec$panel <- factor(paste("count against x, day", obs_dec$day),
                        levels = pan_dec)
curves_dec <- rbind(
  do.call(rbind, lapply(dec_days, function(dd) {
    data.frame(panel = paste("count against x, day", dd), x = xs,
               y = predict(fit_dec, what = "mu",
                           newdata = data.frame(day = dd, x = xs)),
               curve = "fitted")
  })),
  data.frame(panel = pan_dec[4], x = days, y = log(m0 / m1), curve = "fitted"))
curves_dec$panel <- factor(curves_dec$panel, levels = pan_dec)
curves_dec$curve <- factor(curves_dec$curve, levels = dec_names)
sim_dec <- data.frame(panel = factor(pan_dec[4], levels = pan_dec), x = days,
                      y = exp(log(0.6) + 0.5 * (days > 17)),
                      curve = factor("simulated", levels = dec_names))
lim_dec <- data.frame(panel = factor(rep(pan_dec, each = 2), levels = pan_dec),
                      x = NA_real_, y = c(rep(c(0, 90), times = 3), 0.5, 1.1))
ggplot(obs_dec, aes(x = x, y = y)) +
  geom_blank(data = lim_dec) +
  geom_point(colour = book_grey) +
  geom_line(data = curves_dec, aes(colour = curve, linetype = curve),
            linewidth = 0.9) +
  geom_step(data = sim_dec, aes(colour = curve, linetype = curve),
            linewidth = 0.7) +
  facet_wrap(~panel, nrow = 1, scales = "free") +
  scale_colour_manual(values = c(chalk, rust), drop = FALSE) +
  scale_linetype_manual(values = c(1, 2), drop = FALSE) +
  labs(x = NULL, y = NULL)
```

The sharp term `jump(day)` is rejected inside `nl()`, with a message that
names the argument `smoothed`. Its block is a working linearization rather
than a Jacobian, and it cannot sit inside the Jacobian of another term.

## A function written in R {#sec-nonlinear-function}

When the curve cannot be written as a formula, the first argument of `nl()`
can be an R function of two arguments, the covariates `x` and a list `theta`
of the parameters, vectorized in both. The argument `params` names the
parameters, and the argument `x` gives the expression, evaluated in the data,
that is passed to the function. A function can be anything that R computes,
for example a model function of another package. The function `SSfol()` of
`stats` computes the one-compartment curve of the previous section, and we
use it as it is:

```{r}
#| label: nl-fit-function
fol <- function(x, theta) {
  as.numeric(SSfol(x[, "Dose"], x[, "Time"], theta$lKe, theta$lKa,
                   theta$lCl))
}
fit_fn <- statmod(conc ~ 0 + nl(fol, params = c("lKe", "lKa", "lCl"),
                                x = cbind(Dose = Dose, Time = Time),
                                start = list(lKe = -2.5, lKa = 0.5,
                                             lCl = -3)),
                  distrib = gaussian1_distrib(), data = Theoph)
summary(fit_fn)
```

This model has no random effects: all subjects share the three parameters.
Inside a function the place where each parameter enters is not visible to
`statmod()`, so a parameter of a function cannot have a formula of its own.
The same model written as a formula, and the fit of `nls()`, give the same
estimates:

```{r}
#| label: nl-fit-function-compare
fit_fm <- statmod(conc ~ 0 +
                    nl(~ Dose * exp(lKe + lKa - lCl) *
                         (exp(-exp(lKe) * Time) - exp(-exp(lKa) * Time)) /
                         (exp(lKa) - exp(lKe)),
                       start = list(lKe = -2.5, lKa = 0.5, lCl = -3)),
                  distrib = gaussian1_distrib(), data = Theoph)
nls_fn <- nls(conc ~ SSfol(Dose, Time, lKe, lKa, lCl), data = Theoph)
cbind(fn = coef(fit_fn)$mu, formula = coef(fit_fm)$mu, nls = coef(nls_fn))
```

The derivatives of a function in its parameters are computed by finite
differences. Where they are known, the argument
`gradient` supplies the first ones, as a function of `theta` and of `data`,
the variables of the data, that returns a named list with one vector per
parameter. `SSfol()` returns its gradient as an attribute when its
parameters are given as names of variables, so we copy them into three
variables and read the gradient from there:

```{r}
#| label: nl-fit-function-gradient
fol_grad <- function(theta, data) {
  lKe <- theta$lKe
  lKa <- theta$lKa
  lCl <- theta$lCl
  g <- attr(SSfol(data$Dose, data$Time, lKe, lKa, lCl), "gradient")
  list(lKe = g[, "lKe"], lKa = g[, "lKa"], lCl = g[, "lCl"])
}
fit_fg <- statmod(conc ~ 0 + nl(fol, params = c("lKe", "lKa", "lCl"),
                                x = cbind(Dose = Dose, Time = Time),
                                gradient = fol_grad,
                                start = list(lKe = -2.5, lKa = 0.5,
                                             lCl = -3)),
                  distrib = gaussian1_distrib(), data = Theoph)
coef(fit_fg)$mu
```

The estimates are the same. With the exact gradient the design block of the
term is exact, and the higher derivatives are finite differences of the
gradient rather than of the function itself. @fig-nl-function shows the fitted
curve for three subjects with different doses.

```{r}
#| label: fig-nl-function
#| fig-cap: "Concentration of theophylline against time for three subjects with different doses, with the curve of the model in which all subjects share the three parameters, at the dose of each subject."
#| echo: false
#| fig-height: 3.6
#| fig-width: 9
dose_s <- tapply(Theoph$Dose, Theoph$Subject, function(v) v[1])
sel_fn <- names(sort(dose_s))[c(1, 6, 12)]
sel_fn <- sel_fn[order(as.numeric(sel_fn))]
pan_fn <- sprintf("subject %s, dose %.2f", sel_fn, dose_s[sel_fn])
names(pan_fn) <- sel_fn
curves_fn <- do.call(rbind, lapply(sel_fn, function(s) {
  nd <- data.frame(Time = tt, Dose = dose_s[[s]])
  data.frame(Time = tt, panel = pan_fn[[s]],
             conc = predict(fit_fn, what = "mu", newdata = nd))
}))
curves_fn$panel <- factor(curves_fn$panel, levels = pan_fn)
obs_fn <- data.frame(Time = Theoph$Time, conc = Theoph$conc,
                     subject = as.character(Theoph$Subject))
obs_fn <- obs_fn[obs_fn$subject %in% sel_fn, ]
obs_fn$panel <- factor(unname(pan_fn[obs_fn$subject]), levels = pan_fn)
ggplot(obs_fn, aes(x = Time, y = conc)) +
  geom_point(colour = book_grey) +
  geom_line(data = curves_fn, aes(colour = "fitted curve"), linewidth = 0.9) +
  facet_wrap(~panel, nrow = 1) +
  scale_colour_manual(values = chalk) +
  coord_cartesian(ylim = c(0, 12)) +
  labs(x = "time (hours)", y = "concentration (mg/L)")
```

```{r}
#| label: gate-12
#| include: false
assert_nonlinear_ok(
  fits = list(mm = fit_mm, mm_log = fit_mm_log, mm_gam = fit_mm_gam,
              st = fit_st, st0 = fit_st0, or = fit_or, or_ml = fit_or_ml,
              th = fit_th, las = fit_las, ran = fit_ran, dec = fit_dec,
              dec0 = fit_dec0, fn = fit_fn, fm = fit_fm, fg = fit_fg),
  ref = list(nls_mm = nls_mm, glm_mm = glm_mm, nls_st = nls_st,
             nlme_or = nlme_or, nlme_or_ml = nlme_or_ml, nlme_th = nlme_th,
             nls_fn = nls_fn),
  data = list(puro = puro, Puromycin = Puromycin, orange = Orange,
              theoph = Theoph, enz = enz, decay = decay),
  claims = list(lr_st = lr_st, vm_true = vm_true, zero_las = zero_las))
```
