# Time series {#sec-timeseries}

```{r}
#| label: setup-13
#| include: false
source("R/_setup.R")
source("R/timeseries-certificates.R")
chalk <- "#3D6B4C"; rust <- "#9C3E11"; grey <- "grey55"; ink <- "grey15"
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

In a time series the distribution of an observation depends on the
observations before it. `statmodels7` has two model terms for this kind of
dependence. The term `gas()` adds to the predictor of one parameter of the
distribution a level $f_t$, and it updates the level at each time with the
score of the previous observation [@creal2013; @harvey2013]. The term
`regime()` makes the model a hidden Markov model [@hamilton1989;
@zucchini2016]. At each time the series is in one of $k$ states, which we do
not observe. The state at time $t$ depends only on the state at time
$t - 1$, through a matrix of transition probabilities, and each state adds
its own level to the predictor. In both cases the contribution of the term at
one time depends on the data at earlier times, so it is computed by a
recursion and it cannot be written as a block of a design matrix. We call
such a term structural. A model can contain one structural term, in the
equation of any parameter of the distribution.

## The score-driven filter {#sec-timeseries-gas}

The term `gas()` adds a level $f_t$ to the predictor of one parameter of the
distribution. If $\theta_t$ is that parameter and $g$ its link,
$$
g(\theta_t) = \eta_t = x_t^\top \beta + f_t, \qquad
f_t = \omega + \sum_{i=1}^{p} \kappa_i\, u_{t-i} + \sum_{j=1}^{q} \xi_j f_{t-j},
$$
where $x_t^\top \beta$ is the rest of the predictor. The recursion is driven
by the scaled score of the previous observations [@creal2013; @harvey2013],
$$
u_t = s_t\, \mathcal{I}_t^{-d}, \qquad
s_t = \frac{\partial \log p(y_t)}{\partial \eta_t}, \qquad
\mathcal{I}_t = \mathrm{E}\left[-\frac{\partial^2 \log p(y_t)}{\partial \eta_t^2}\right].
$$
The score $s_t$ is the derivative of the log-density of observation $t$ with
respect to the predictor. An observation that the current predictor fits
badly has a large score, and the recursion moves the predictor of the next
times in the direction that would have fitted it better. The quantity
$\mathcal{I}_t$ is the expected information of the predictor, which is also
the variance of $s_t$. The argument `scaling` of `gas()` sets the power $d$,
and three values are common:

- with $d = 0$, the default, the score is used as it is;
- with $d = 1/2$ the score is divided by its standard deviation, so $u_t$
  has variance one;
- with $d = 1$ the score is multiplied by the inverse of its information,
  which is the step of Newton's method for one observation.

For every $d$ the scaled score has mean zero, so $f_t$ fluctuates around
$\omega/(1 - \sum_j \xi_j)$, and the recursion starts there. If
$\mathcal{I}_t$ is the same at every time, the power $d$ multiplies $u_t$ by
a constant, the loadings absorb that constant, and the model does not depend
on $d$. If $\mathcal{I}_t$ changes over time, each $d$ gives a different
model.

The intercept of the predictor and the level $\omega$ cannot both be
estimated. Both move the predictor by a constant: with an intercept
$\beta_0$ the predictor fluctuates around $\beta_0 + \omega/(1 - \sum_j\xi_j)$,
and adding $c$ to $\beta_0$ while subtracting $c\,(1 - \sum_j\xi_j)$ from
$\omega$ leaves every $\eta_t$ unchanged. When the formula has an intercept,
`statmod()` therefore holds $\omega$ at zero and estimates $\beta_0$, which
is then the level around which the predictor fluctuates. Written without an
intercept, as `y ~ 0 + gas(...)`, the same model has $\omega$ free instead,
and the two spellings are related by
$$
\beta_0 = \frac{\omega}{1 - \sum_j \xi_j}, \qquad
\omega = \beta_0\,\Bigl(1 - \sum_j \xi_j\Bigr).
$$

Several models of the time series literature are score-driven models with a
particular family, parameter, link and power $d$, and `gas()` fits them by
naming those four. @tbl-ts-known lists the ones of this chapter, and the
following sections fit each of them and compare the fit with the package
that implements the model. Where the information is constant, every $d$
gives the same model, and the table says "any".

| model | family | parameter, link | $d$ | section |
|---|---|---|---|---|
| ARMA(1, 1) | `gaussian1_distrib()` | mean, identity | any | @sec-timeseries-gas |
| GARCH(1, 1) [@bollerslev1986] | `gaussian2_distrib()` | variance, identity | 1 | @sec-timeseries-volatility |
| Beta-t-EGARCH [@harvey2013] | `student_t1_distrib()` | scale, log | any | @sec-timeseries-volatility |
| INGARCH(1, 1) [@ferland2006] | `poisson_distrib()` | mean, identity | 1 | @sec-timeseries-counts |
| ACD(1, 1) [@engle1998] | `exponential_distrib()` | mean, identity | 1 | @sec-timeseries-families |
| logarithmic ACD, second form [@bauwens2000] | `exponential_distrib()` | mean, log | any | @sec-timeseries-families |

: Known models that `gas()` fits exactly. {#tbl-ts-known}

### The working parametrization

The loadings and the persistences are constrained. A loading is positive, so
that an observation the predictor underestimates raises the predictor of
the next times, and the persistences must keep the recursion stationary, so
that $f_t$ fluctuates around its level instead of drifting away from it.
`gas()` does not search the parameters in these constrained sets. It
searches unconstrained working parameters, from which the parameters of the
model follow, so that every point the search visits is an admissible model.

With one lag of each kind, $p = q = 1$, the recursion is
$f_t = \omega + \kappa\, u_{t-1} + \xi f_{t-1}$, and it is stationary when
$|\xi| < 1$. The working parameters are
$$
a = \log \kappa, \qquad z = \operatorname{atanh} \xi
= \frac{1}{2}\log\frac{1 + \xi}{1 - \xi},
$$
so $\kappa = e^{a}$ is positive and $\xi = \tanh z$ lies in $(-1, 1)$ for
every real $a$ and $z$. The level $\omega$ needs no constraint and is its
own working parameter. The summary reports $\kappa$ and $\xi$, and builds
their intervals on the working scale and maps them back, so the interval of
$\kappa$ is positive and that of $\xi$ lies inside $(-1, 1)$.

With several lags each loading has its own working parameter,
$a_i = \log \kappa_i$ for $i = 1, \dots, p$. The persistences need more
care. The recursion is stationary when the roots of
$1 - \xi_1 z - \dots - \xi_q z^q$ lie outside the unit circle, and this
region is not a box: for $q = 2$ it is the triangle $|\xi_2| < 1$,
$\xi_2 + \xi_1 < 1$, $\xi_2 - \xi_1 < 1$. The recursion is therefore
parametrized by its partial autocorrelations $r_1, \dots, r_q$, which range
over the box $(-1, 1)^q$, and every point of the box gives a stationary
recursion, each stationary recursion being given by exactly one point
[@barndorffnielsen1973; @monahan1984]. Each partial autocorrelation has the
working parameter $z_k = \operatorname{atanh} r_k$, and the persistences
follow from the Durbin-Levinson recursion,
$$
\xi^{(k)}_k = r_k, \qquad
\xi^{(k)}_j = \xi^{(k-1)}_j - r_k\, \xi^{(k-1)}_{k-j}, \quad j = 1, \dots, k - 1,
$$
for $k = 1, \dots, q$, with $\xi_j = \xi^{(q)}_j$. For $q = 2$ this gives
$\xi_1 = r_1 (1 - r_2)$ and $\xi_2 = r_2$. The same recursion gives
$$
1 - \sum_{j=1}^{q} \xi_j = \prod_{k=1}^{q} (1 - r_k),
$$
so the level $\omega / (1 - \sum_j \xi_j)$ around which $f_t$ fluctuates is
finite at every point of the box. With one lag the persistence and the
partial autocorrelation coincide. With several, the summary reports the
persistences $\xi_j$, with standard errors from the Jacobian of the
Durbin-Levinson recursion, and builds their intervals on the scale of the
persistences, because the region they range over is not a box.

### The mean of a gaussian response

The series `Nile` gives the annual flow of the river Nile at Aswan from 1871
to 1970, in $10^8$ cubic meters [@cobb1978]. We put the filter in the
equation of the mean of a gaussian response:
$$
y_t \sim \mathrm{N}(\mu_t, \sigma), \qquad
\mu_t = \beta_0 + f_t, \qquad
f_t = \kappa\, u_{t-1} + \xi f_{t-1}.
$$
The score with respect to $\mu_t$ and its information are
$$
s_t = \frac{y_t - \mu_t}{\sigma^2}, \qquad \mathcal{I}_t = \frac{1}{\sigma^2},
$$
so $u_t = (y_t - \mu_t)\,\sigma^{2(d - 1)}$. An observation above its mean
raises the mean of the next year, and the parameter $\xi$ sets how long the
effect lasts. The information does not change over time, so $d$ only sets
the units of $\kappa$. We fit the model with the default $d = 0$:

```{r}
#| label: ts-fit-nile
nile <- data.frame(year = 1871:1970, flow = as.numeric(Nile))
fit_nile <- statmod(flow ~ gas(p = 1, q = 1, time = year),
                    distrib = gaussian1_distrib(), data = nile)
summary(fit_nile)
```

The argument `p` of `gas()` is the number of lagged scores and `q` the number
of lagged levels, and `time` gives the order of the observations. The
summary marks $\omega$ as held at zero. With $d = 1$ the fit is the same, and only the loading changes, by
the factor $\sigma^2$:

```{r}
#| label: ts-fit-nile-d1
fit_nile_1 <- statmod(flow ~ gas(p = 1, q = 1, time = year, scaling = 1),
                      distrib = gaussian1_distrib(), data = nile)
c(d_0 = logLik(fit_nile), d_1 = logLik(fit_nile_1))
c(kappa_d0_over_s2 = coef(fit_nile)$mu[["gas.kappa1"]] /
    exp(2 * coef(fit_nile)$sigma[["(Intercept)"]]),
  kappa_d1 = coef(fit_nile_1)$mu[["gas.kappa1"]])
```

```{r}
#| label: ts-nile-numbers
#| include: false
cf_nile <- coef(fit_nile)$mu
s2_nile <- exp(2 * coef(fit_nile)$sigma[["(Intercept)"]])
a_nile <- cf_nile[["gas.kappa1"]] / s2_nile
b_nile <- cf_nile[["gas.xi1"]]
```

With $d = 1$ the scaled score is the residual $y_t - \mu_t$, and the loading
$\hat\kappa = `r fmt(a_nile, 3)`$ is the fraction of the last residual added
to the mean. With $\hat\xi = `r fmt(b_nile, 3)`$, a flow that exceeds its
mean by 100 raises the mean of the next year by about
`r fmt(100 * a_nile, 0)`, and a third of that remains after
`r fmt(log(1/3) / log(b_nile), 0)` years.

This model is a model of the autoregressive moving average family
[@brockwell1991]. Writing $e_t = y_t - \mu_t$ and substituting the recursion
with $d = 1$ gives
$$
y_t - \beta_0 = \xi\,(y_{t-1} - \beta_0) + e_t + (\kappa - \xi)\, e_{t-1},
$$
an ARMA(1, 1) with autoregressive coefficient $\xi$ and moving average
coefficient $\kappa - \xi$. The function `arima()` of `stats` fits it by
maximum likelihood:

```{r}
#| label: ts-arima-nile
arima_nile <- arima(nile$flow, order = c(1, 0, 1), method = "ML")
arima_nile
cbind(statmod = c(ar1 = b_nile, ma1 = a_nile - b_nile,
                  intercept = cf_nile[["(Intercept)"]]),
      arima = coef(arima_nile))
c(statmod = logLik(fit_nile), arima = logLik(arima_nile))
```

The estimates are close but not equal, and so are the log-likelihoods. The
two functions maximize different likelihoods of the same model. Our
recursion starts at the level $f_1 = 0$, so the first observation has
variance $\sigma^2$. The likelihood of `arima()` gives the first observation
the variance of the stationary process, which is larger. Both
log-likelihoods are maximized over all the parameters, $\sigma$ included.

A second lagged level, `q = 2`, gives the recursion
$f_t = \kappa u_{t-1} + \xi_1 f_{t-1} + \xi_2 f_{t-2}$. We compare the two
models by the likelihood ratio:

```{r}
#| label: ts-fit-nile-q2
fit_nile2 <- statmod(flow ~ gas(p = 1, q = 2, time = year),
                     distrib = gaussian1_distrib(), data = nile)
summary(fit_nile2)
l0 <- logLik(fit_nile)
l1 <- logLik(fit_nile2)
lr_nile <- 2 * (as.numeric(l1) - as.numeric(l0))
lr_nile
pchisq(lr_nile, df = 1, lower.tail = FALSE)
```

```{r}
#| label: ts-nile-q2-numbers
#| include: false
p_nile <- pchisq(lr_nile, df = 1, lower.tail = FALSE)
```

The smaller model has $\xi_2 = 0$, a point inside the range of the
partial autocorrelation, so the ratio has its usual $\chi^2_1$ reference. It
is `r fmt(lr_nile, 3)`, with a p-value of `r fmt(p_nile, 3)`, and one lag is
enough. With two lags the coefficients $\xi_1$ and $\xi_2$ are functions
of both partial autocorrelations, and the region where the recursion is
stationary is not a product of two intervals, so their intervals are built on
the scale of the coefficients and can leave $(-1, 1)$, as the interval of
$\xi_1$ does here. Lagged scores are added in the same way with `p`, each
with a loading of its own. @fig-ts-nile shows the flows with the mean of the
first model and its 95% confidence band, which `predict()` computes with
`se = TRUE`.

```{r}
#| label: fig-ts-nile
#| fig-cap: "Annual flow of the Nile with the mean of the score-driven model and its 95% confidence band."
#| echo: false
#| fig-height: 3.8
band_nile <- predict(fit_nile, what = "mu", se = TRUE)
band_df <- data.frame(year = nile$year, fit = band_nile$fit,
                      lower = band_nile$lower, upper = band_nile$upper)
ggplot(nile, aes(x = year, y = flow)) +
  geom_ribbon(data = band_df, aes(x = year, ymin = lower, ymax = upper,
                                  fill = "95% band"),
              inherit.aes = FALSE, alpha = 0.2) +
  geom_point(colour = book_grey, size = 1.2) +
  geom_line(data = band_df, aes(x = year, y = fit,
                                colour = "mean of the model"),
            linewidth = 0.9) +
  scale_colour_manual(values = chalk, guide = guide_legend(order = 1)) +
  scale_fill_manual(values = chalk, guide = guide_legend(order = 2)) +
  labs(x = "year", y = expression(flow ~ (10^8 ~ m^3)))
```

## Volatility {#sec-timeseries-volatility}

The data set `EuStockMarkets` gives the daily closing prices of four
European stock indices from 1991 to 1998. We use the German index DAX and its
daily returns in percent, $r_t = 100\,(\log P_t - \log P_{t-1})$. Returns go
through quiet periods and turbulent periods, so their variance changes over
time, and we put the filter in the equation of the variance or of the scale.

### GARCH(1, 1)

The best-known model of a changing variance is the GARCH(1, 1) model
[@bollerslev1986], which writes the variance as
$$
\sigma_t^2 = \omega_G + \alpha_G\,(r_{t-1} - \mu)^2 + \beta_G\,\sigma_{t-1}^2 .
$$
The package `rugarch` fits it, and we print its table of estimates
[@galanos2026]:

```{r}
#| label: ts-garch-dax
dax <- data.frame(day = 1:1859,
                  r = 100 * diff(log(as.numeric(EuStockMarkets[, "DAX"]))))
spec_n <- rugarch::ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
  distribution.model = "norm")
garch_n <- rugarch::ugarchfit(spec = spec_n, data = dax$r, solver = "hybrid")
garch_n@fit$matcoef
```

The gaussian distribution with the variance $\sigma_t^2$ as its parameter is
`gaussian2_distrib()`. On the identity link of the variance the score and
its information are
$$
s_t = \frac{(r_t - \mu)^2 - \sigma_t^2}{2\sigma_t^4}, \qquad
\mathcal{I}_t = \frac{1}{2\sigma_t^4},
$$
and the information changes over time. With $d = 1$ the recursion is driven
by $u_t = (r_t - \mu)^2 - \sigma_t^2$. We write the model with an intercept
$\gamma_0$:
$$
r_t \sim \mathrm{N}(\mu, \sigma_t^2), \qquad
\sigma_t^2 = \gamma_0 + f_t, \qquad
f_t = \kappa\left[(r_{t-1} - \mu)^2 - \sigma_{t-1}^2\right] + \xi f_{t-1}.
$$
Substituting $f_t = \sigma_t^2 - \gamma_0$ gives
$$
\sigma_t^2 = \gamma_0\,(1 - \xi) + \kappa\,(r_{t-1} - \mu)^2 +
(\xi - \kappa)\,\sigma_{t-1}^2,
$$
the GARCH(1, 1) recursion with $\omega_G = \gamma_0 (1 - \xi)$,
$\alpha_G = \kappa$ and $\beta_G = \xi - \kappa$:

```{r}
#| label: ts-fit-garch
fit_garch <- statmod(r ~ 1 | sigma2 ~ gas(p = 1, q = 1, time = day, scaling = 1),
                     distrib = gaussian2_distrib(link_sigma2 = identity_link()),
                     data = dax)
summary(fit_garch)
cf_g <- coef(fit_garch)$sigma2
cbind(statmod = c(mu = coef(fit_garch)$mu[["(Intercept)"]],
                  omega = cf_g[["(Intercept)"]] * (1 - cf_g[["gas.xi1"]]),
                  alpha1 = cf_g[["gas.kappa1"]],
                  beta1 = cf_g[["gas.xi1"]] - cf_g[["gas.kappa1"]]),
      rugarch = rugarch::coef(garch_n))
c(statmod = logLik(fit_garch), rugarch = rugarch::likelihood(garch_n))
```

```{r}
#| label: ts-garch-numbers
#| include: false
d_garch <- abs(as.numeric(logLik(fit_garch)) - rugarch::likelihood(garch_n))
```

The estimates agree to two decimals, and the log-likelihoods differ by
`r fmt(d_garch, 3)`: `rugarch` starts the recursion at the mean of the
squared residuals, and we start it at $\gamma_0$. The identity link does not
keep $\sigma_t^2$ positive by itself. At these estimates $\omega_G$,
$\alpha_G$ and $\beta_G$ are positive, so every $\sigma_t^2$ is positive, as
in the GARCH model.

### The logarithm of the scale

The variance can also be filtered on the logarithmic scale, which keeps it
positive whatever the loadings are. With the gaussian distribution of scale
$\sigma_t$,
$$
r_t \sim \mathrm{N}(\mu, \sigma_t), \qquad
\log \sigma_t = \gamma_0 + f_t, \qquad
f_t = \kappa\, u_{t-1} + \xi f_{t-1},
$$
the score with respect to $\log \sigma_t$ and its information are
$$
s_t = z_t^2 - 1, \qquad \mathcal{I}_t = 2, \qquad z_t = \frac{r_t - \mu}{\sigma_t}.
$$
The information is constant, so $d$ does not change the model. A return that
is large compared with $\sigma_t$ has a positive score and raises the scale
of the next day; a small one lowers it.

```{r}
#| label: ts-fit-dax
fit_dax <- statmod(r ~ 1 | sigma ~ gas(p = 1, q = 1, time = day),
                   distrib = gaussian1_distrib(), data = dax)
summary(fit_dax)
c(log_scale = logLik(fit_dax), garch = logLik(fit_garch))
```

```{r}
#| label: ts-dax-numbers
#| include: false
sd_gas <- predict(fit_dax, what = "sigma")
sd_garch <- sqrt(predict(fit_garch, what = "sigma2"))
d_n <- as.numeric(logLik(fit_garch)) - as.numeric(logLik(fit_dax))
```

Both models have four parameters. With the gaussian distribution the GARCH
model has the larger log-likelihood, by `r fmt(d_n, 1)`. The two fitted
standard deviations follow the same quiet and turbulent periods (their
correlation over the days is `r fmt(cor(sd_gas, sd_garch), 2)`), and
@fig-ts-dax shows them over the returns.

```{r}
#| label: fig-ts-dax
#| fig-cap: "Absolute daily returns of the DAX, with the standard deviation of the score-driven model on the logarithm of the scale and of the GARCH(1, 1) model, both with a gaussian distribution."
#| echo: false
#| fig-height: 3.8
yr <- as.numeric(time(EuStockMarkets))[-1]
sd_names <- c("logarithm of the scale", "GARCH(1, 1)")
curves_dax <- data.frame(
  year = rep(yr, times = 2), sd = c(sd_gas, sd_garch),
  curve = factor(rep(sd_names, each = length(yr)), levels = sd_names))
ggplot(data.frame(year = yr, r = abs(dax$r)), aes(x = year)) +
  geom_linerange(aes(ymin = 0, ymax = r), colour = book_grey,
                 linewidth = 0.3) +
  geom_line(data = curves_dax, aes(y = sd, colour = curve, linetype = curve),
            linewidth = 0.6) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c(1, 2)) +
  scale_x_continuous(breaks = seq(ceiling(min(yr)), floor(max(yr)))) +
  labs(x = "year", y = "absolute return (%)")
```

### A heavy-tailed distribution

The recursion uses the score of whatever distribution the model carries, so
the same term gives a different model with each family. Daily returns have
heavier tails than the gaussian distribution. With the Student $t$
distribution of location $\mu$, scale $\sigma_t$ and $\nu$ degrees of
freedom, the score with respect to $\log \sigma_t$ and its information are
$$
s_t = \frac{(\nu + 1)\, z_t^2}{\nu + z_t^2} - 1 , \qquad
\mathcal{I}_t = \frac{2\nu}{\nu + 3}.
$$
The score tends to $\nu$ as $z_t$ grows. A single very large return therefore
moves the scale by a bounded amount, while under the gaussian distribution
its effect grows with $z_t^2$ (@fig-ts-score). The information is again
constant, so $d$ does not change the model. This model, a Student $t$
response whose log-scale is driven by its own score, is the Beta-t-EGARCH
model of @harvey2013, so `gas()` fits it with the family and nothing else.

```{r}
#| label: ts-fit-dax-t
fit_dax_t <- statmod(r ~ 1 | sigma ~ gas(p = 1, q = 1, time = day),
                     distrib = student_t1_distrib(), data = dax)
summary(fit_dax_t)
c(gaussian = AIC(fit_dax), t = AIC(fit_dax_t))
```

```{r}
#| label: ts-dax-t-numbers
#| include: false
nu_dax <- exp(coef(fit_dax_t)$nu[["(Intercept)"]])
```

The $t$ distribution has $\hat\nu = `r fmt(nu_dax, 2)`$ and a much lower AIC.
The gaussian distribution is the limit $\nu \to \infty$, a point outside the
range of $\nu$, so the likelihood ratio between the two has no $\chi^2$
reference and we compare them by the AIC. The GARCH(1, 1) model with the
$t$ distribution has five parameters, as ours does:

```{r}
#| label: ts-garch-dax-t
spec_t <- rugarch::ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
  distribution.model = "std")
garch_t <- rugarch::ugarchfit(spec = spec_t, data = dax$r, solver = "hybrid")
garch_t@fit$matcoef
c(statmod = logLik(fit_dax_t), rugarch = rugarch::likelihood(garch_t))
```

```{r}
#| label: ts-dax-t-compare
#| include: false
d_t <- as.numeric(logLik(fit_dax_t)) - rugarch::likelihood(garch_t)
```

With the $t$ distribution the order is reversed: the score-driven model has
the larger log-likelihood, by `r fmt(d_t, 1)`. The degrees of freedom of the
two fits are close (`r fmt(nu_dax, 2)` here,
`r fmt(rugarch::coef(garch_t)[["shape"]], 2)` in the GARCH model). In the
GARCH model a large return enters the variance as a squared residual whatever
the distribution, while in the score-driven model the distribution decides
how much it counts.

@fig-ts-dax-t compares the standard deviations of the two score-driven
models on the logarithm of the scale. The scale $\sigma_t$ of the $t$
distribution is not its standard deviation, which is
$\sigma_t\sqrt{\nu/(\nu - 2)}$, so we plot the latter for both models. After
the large returns of August 1991 the gaussian model raises its standard
deviation more and keeps it high longer, because each of those returns
enters through $z_t^2$; the $t$ model bounds their effect.

```{r}
#| label: fig-ts-dax-t
#| fig-cap: "Standard deviation of the DAX returns under the score-driven model with a gaussian distribution and with a Student t distribution, over the absolute returns. For the t distribution the standard deviation is the scale times the square root of nu/(nu - 2)."
#| echo: false
#| fig-height: 3.8
sd_t <- predict(fit_dax_t, what = "sigma") * sqrt(nu_dax / (nu_dax - 2))
sdt_names <- c("gaussian", "Student t")
curves_dax_t <- data.frame(
  year = rep(yr, times = 2), sd = c(sd_gas, sd_t),
  curve = factor(rep(sdt_names, each = length(yr)), levels = sdt_names))
ggplot(data.frame(year = yr, r = abs(dax$r)), aes(x = year)) +
  geom_linerange(aes(ymin = 0, ymax = r), colour = book_grey,
                 linewidth = 0.3) +
  geom_line(data = curves_dax_t, aes(y = sd, colour = curve, linetype = curve),
            linewidth = 0.6) +
  scale_colour_manual(values = c(chalk, rust)) +
  scale_linetype_manual(values = c(1, 2)) +
  scale_x_continuous(breaks = seq(ceiling(min(yr)), floor(max(yr)))) +
  labs(x = "year", y = "absolute return (%)")
```

```{r}
#| label: fig-ts-score
#| fig-cap: "The score with respect to the log-scale as a function of the standardized observation, for the gaussian distribution and for the Student t distribution with the degrees of freedom estimated on the DAX returns."
#| echo: false
#| fig-height: 3.6
zz <- seq(-6, 6, length.out = 400)
score_names <- c("gaussian", sprintf("Student t, nu = %.2f", nu_dax),
                 "limit of the t score")
curves_score <- data.frame(
  z = rep(zz, times = 3),
  score = c(zz^2 - 1, (nu_dax + 1) * zz^2 / (nu_dax + zz^2) - 1,
            rep(nu_dax, times = length(zz))),
  curve = factor(rep(score_names, each = length(zz)), levels = score_names))
ggplot(curves_score, aes(x = z, y = score)) +
  geom_line(aes(colour = curve, linetype = curve, linewidth = curve)) +
  scale_colour_manual(values = c(chalk, rust, grey)) +
  scale_linetype_manual(values = c(1, 2, 3)) +
  scale_linewidth_manual(values = c(0.9, 0.9, 0.5)) +
  coord_cartesian(ylim = c(-1, 12)) +
  labs(x = "standardized observation z", y = "score")
```

## Counts {#sec-timeseries-counts}

The series `discoveries` gives the number of great inventions and scientific
discoveries in each year from 1860 to 1959. We put the filter in the
equation of the mean of a Poisson response.

### INGARCH(1, 1)

On the identity link the score with respect to $\mu_t$ is
$(y_t - \mu_t)/\mu_t$ and its information is $1/\mu_t$, so with $d = 1$ the
recursion is driven by $u_t = y_t - \mu_t$. With an intercept $\gamma_0$, as
for the GARCH model,
$$
y_t \sim \mathrm{Poisson}(\mu_t), \qquad
\mu_t = \gamma_0\,(1 - \xi) + \kappa\, y_{t-1} + (\xi - \kappa)\,\mu_{t-1},
$$
which is the INGARCH(1, 1) model of counts [@ferland2006; @fokianos2009].
The package `tscount` fits it with the identity link [@liboschik2017]. There
the coefficient of the past count is named `beta_1` and that of the past
mean `alpha_1`:

```{r}
#| label: ts-fit-ingarch
disc <- data.frame(year = 1860:1959, n = as.numeric(discoveries))
fit_ingarch <- statmod(n ~ gas(p = 1, q = 1, time = year, scaling = 1),
                       distrib = poisson_distrib(link_mu = identity_link()),
                       data = disc)
summary(fit_ingarch)
tsc_id <- tscount::tsglm(discoveries, model = list(past_obs = 1, past_mean = 1),
                         link = "identity", distr = "poisson")
cf_i <- coef(fit_ingarch)$mu
cbind(statmod = c(intercept = cf_i[["(Intercept)"]] * (1 - cf_i[["gas.xi1"]]),
                  beta_1 = cf_i[["gas.kappa1"]],
                  alpha_1 = cf_i[["gas.xi1"]] - cf_i[["gas.kappa1"]]),
      tscount = coef(tsc_id))
c(statmod = logLik(fit_ingarch), tscount = logLik(tsc_id))
```

The estimates agree to two decimals, and the log-likelihoods differ by less
than 0.001, because the two recursions start from different values.

### The logarithmic link

On the logarithmic link the score with respect to $\log \mu_t$ is
$y_t - \mu_t$ and its information is $\mu_t$, so
$$
y_t \sim \mathrm{Poisson}(\mu_t), \qquad
\log \mu_t = \beta_0 + f_t, \qquad
f_t = \kappa\,\frac{y_{t-1} - \mu_{t-1}}{\mu_{t-1}^{d}} + \xi f_{t-1}.
$$
The information changes over time, so each $d$ gives a different model. A
year with more discoveries than expected raises the expected number of the
next year: with $d = 0$ in proportion to the excess $y_t - \mu_t$, with
$d = 1$ in proportion to the relative excess $(y_t - \mu_t)/\mu_t$. We fit
the three models:

```{r}
#| label: ts-fit-disc
fit_disc <- statmod(n ~ gas(p = 1, q = 1, time = year),
                    distrib = poisson_distrib(), data = disc)
fit_disc_h <- statmod(n ~ gas(p = 1, q = 1, time = year, scaling = 0.5),
                      distrib = poisson_distrib(), data = disc)
fit_disc_1 <- statmod(n ~ gas(p = 1, q = 1, time = year, scaling = 1),
                      distrib = poisson_distrib(), data = disc)
summary(fit_disc_1)
c(d_0 = logLik(fit_disc), d_half = logLik(fit_disc_h), d_1 = logLik(fit_disc_1))
```

```{r}
#| label: ts-disc-scaling-numbers
#| include: false
d_sc <- as.numeric(logLik(fit_disc_1)) - as.numeric(logLik(fit_disc))
```

The three models have the same number of parameters, so we compare their
log-likelihoods directly. The model with $d = 1$ has the largest, by
`r fmt(d_sc, 2)` over the model with the score as it is. We compare the
score-driven models with the constant ones by the AIC, since the model
without dynamics is the limit $\kappa \to 0$, which is the edge of the range
of a loading on the logarithmic scale:

```{r}
#| label: ts-fit-disc-static
fit_disc0 <- statmod(n ~ 1, distrib = poisson_distrib(), data = disc)
fit_disc_nb <- statmod(n ~ 1, distrib = negbin2_distrib(), data = disc)
c(poisson = AIC(fit_disc0), negbin = AIC(fit_disc_nb),
  score_driven = AIC(fit_disc))
```

The score-driven Poisson model has a lower AIC than both the constant
Poisson model and the constant negative binomial model: part of the
overdispersion of the counts is a mean that changes over time. The rest of
it can be given to the distribution. With the negative binomial of mean
$\mu_t$ and size $\theta$ the score with respect to $\log\mu_t$ is
$\theta(y_t - \mu_t)/(\theta + \mu_t)$, which grows more slowly than the
Poisson score for large counts:

```{r}
#| label: ts-fit-disc-nb
fit_disc_gnb <- statmod(n ~ gas(p = 1, q = 1, time = year),
                        distrib = negbin2_distrib(), data = disc)
summary(fit_disc_gnb)
c(poisson = AIC(fit_disc), negbin = AIC(fit_disc_gnb))
```

The log-linear model of `tscount` writes
$\log\mu_t = \beta_0 + \beta_1 \log(y_{t-1} + 1) + \alpha_1 \log\mu_{t-1}$,
so the past count enters through its logarithm, where the score-driven model
uses the difference $y_{t-1} - \mu_{t-1}$. It is not a score-driven model,
but it has the same number of parameters, and we compare the maximized
log-likelihoods for both distributions:

```{r}
#| label: ts-tscount-disc
tsc_p <- tscount::tsglm(discoveries, model = list(past_obs = 1, past_mean = 1),
                        link = "log", distr = "poisson")
summary(tsc_p)
tsc_nb <- tscount::tsglm(discoveries, model = list(past_obs = 1, past_mean = 1),
                         link = "log", distr = "nbinom")
rbind(poisson = c(statmod = logLik(fit_disc), tscount = logLik(tsc_p)),
      negbin = c(statmod = logLik(fit_disc_gnb), tscount = logLik(tsc_nb)))
```

```{r}
#| label: ts-tscount-numbers
#| include: false
d_tsc_p <- as.numeric(logLik(fit_disc)) - as.numeric(logLik(tsc_p))
d_tsc_nb <- as.numeric(logLik(fit_disc_gnb)) - as.numeric(logLik(tsc_nb))
```

On these counts the score-driven models have the larger log-likelihood, by
`r fmt(d_tsc_p, 2)` with the Poisson distribution and by
`r fmt(d_tsc_nb, 2)` with the negative binomial. @fig-ts-disc shows the
counts with the means of the Poisson and negative binomial score-driven
models.

```{r}
#| label: fig-ts-disc
#| fig-cap: "Number of great discoveries in each year, with the means of the score-driven Poisson and negative binomial models and the mean of the constant model."
#| echo: false
#| fig-height: 3.6
disc_names <- c("Poisson", "negative binomial", "constant mean")
curves_disc <- data.frame(
  year = rep(disc$year, times = 3),
  n = c(predict(fit_disc, what = "mu"), predict(fit_disc_gnb, what = "mu"),
        rep(mean(disc$n), nrow(disc))),
  curve = factor(rep(disc_names, each = nrow(disc)), levels = disc_names))
ggplot(disc, aes(x = year, y = n)) +
  geom_point(colour = book_grey, size = 1.2) +
  geom_line(data = curves_disc, aes(colour = curve, linetype = curve),
            linewidth = 0.8) +
  scale_colour_manual(values = c(chalk, rust, ink)) +
  scale_linetype_manual(values = c(1, 2, 3)) +
  labs(x = "year", y = "discoveries")
```

## Durations, positive measurements and proportions {#sec-timeseries-families}

### Durations

The time between two trades of a stock is a positive quantity whose mean
changes during the day and from one trade to the next. Models of
autoregressive conditional duration describe the mean duration by a
recursion [@engle1998]. The package `ACDm` carries a series of durations
between trades, already adjusted for the daily pattern so that their mean is
close to one [@belfrage2025], and we use its first 1000.

The exponential distribution with its mean $\psi_t$ on the identity link has
the score $(x_t - \psi_t)/\psi_t^2$ and the information $1/\psi_t^2$, so with
$d = 1$ the recursion is driven by $x_t - \psi_t$, and with an intercept
$\gamma_0$
$$
x_t \sim \mathrm{Exp}(\psi_t), \qquad
\psi_t = \gamma_0\,(1 - \xi) + \kappa\, x_{t-1} + (\xi - \kappa)\,\psi_{t-1}
$$
is the ACD(1, 1) model of @engle1998, which `ACDm` fits as the model `"ACD"`:

```{r}
#| label: ts-fit-acd-linear
data(adjDurData, package = "ACDm")
dur <- data.frame(i = 1:1000, x = adjDurData$adjDur[1:1000])
fit_acd_lin <- statmod(x ~ gas(p = 1, q = 1, time = i, scaling = 1),
                       distrib = exponential_distrib(link_mu = identity_link()),
                       data = dur)
summary(fit_acd_lin)
acd_lin <- ACDm::acdFit(durations = dur$x, model = "ACD", dist = "exponential",
                        order = c(1, 1), output = FALSE)
cf_l <- coef(fit_acd_lin)$mu
cbind(statmod = c(omega = cf_l[["(Intercept)"]] * (1 - cf_l[["gas.xi1"]]),
                  alpha1 = cf_l[["gas.kappa1"]],
                  beta1 = cf_l[["gas.xi1"]] - cf_l[["gas.kappa1"]]),
      ACDm = acd_lin$mPar)
c(statmod = logLik(fit_acd_lin),
  ACDm = acd_lin$goodnessOfFit["LogLikelihood", "value"])
```

```{r}
#| label: ts-acd-linear-numbers
#| include: false
d_acl <- abs(as.numeric(logLik(fit_acd_lin)) -
               acd_lin$goodnessOfFit["LogLikelihood", "value"])
```

The estimates agree to two decimals, and the log-likelihoods differ by
`r fmt(d_acl, 3)`, because the two recursions start from different values.

On the logarithmic link the score with respect to $\log\psi_t$ is
$x_t/\psi_t - 1$ and its information is one, so $d$ does not change the
model:
$$
x_t \sim \mathrm{Exp}(\psi_t), \qquad
\log\psi_t = \beta_0 + f_t, \qquad
f_t = \kappa\left(\frac{x_{t-1}}{\psi_{t-1}} - 1\right) + \xi f_{t-1}.
$$
This is the logarithmic ACD model of @bauwens2000 in its second form,
$\log\psi_t = \omega_A + \kappa\, x_{t-1}/\psi_{t-1} + \xi\log\psi_{t-1}$,
with $\omega_A = \beta_0 (1 - \xi) - \kappa$.

```{r}
#| label: ts-fit-acd
fit_acd <- statmod(x ~ gas(p = 1, q = 1, time = i),
                   distrib = exponential_distrib(), data = dur)
summary(fit_acd)
acd <- ACDm::acdFit(durations = dur$x, model = "LACD2", dist = "exponential",
                    order = c(1, 1), output = FALSE)
cf_acd <- coef(fit_acd)$mu
cbind(statmod = c(omega = cf_acd[["(Intercept)"]] * (1 - cf_acd[["gas.xi1"]]) -
                    cf_acd[["gas.kappa1"]],
                  alpha1 = cf_acd[["gas.kappa1"]], beta1 = cf_acd[["gas.xi1"]]),
      ACDm = acd$mPar)
c(statmod = logLik(fit_acd),
  ACDm = acd$goodnessOfFit["LogLikelihood", "value"])
```

```{r}
#| label: ts-acd-numbers
#| include: false
d_acd <- as.numeric(logLik(fit_acd)) -
  acd$goodnessOfFit["LogLikelihood", "value"]
```

The estimates agree to three decimals, and the log-likelihoods differ by
`r fmt(abs(d_acd), 3)`, because the two recursions start from different
values.

### Positive measurements and proportions

The family can be any distribution of `distributions7`. The series `lh` gives
the level of luteinizing hormone in blood samples taken every ten minutes
from a woman, 48 positive measurements [@diggle1990]. A gamma response with
mean $\mu_t$ and dispersion $\phi$ has the score
$(y_t/\mu_t - 1)/\phi$ with respect to $\log\mu_t$ and the information
$1/\phi$, which is constant, so $d$ does not change the model:

```{r}
#| label: ts-fit-lh
lhd <- data.frame(t = 1:48, y = as.numeric(lh))
fit_lh <- statmod(y ~ gas(p = 1, q = 1, time = t),
                  distrib = gamma1_distrib(), data = lhd)
summary(fit_lh)
fit_lh0 <- statmod(y ~ 1, distrib = gamma1_distrib(), data = lhd)
c(constant = AIC(fit_lh0), score_driven = AIC(fit_lh))
```

A proportion has a beta distribution. The series `Seatbelts` gives the
monthly numbers of car drivers, front-seat passengers and rear-seat
passengers killed or seriously injured in Great Britain from 1969 to 1984,
and a law made front seat belts compulsory from February 1983
[@harvey1986]. We model the share of front-seat passengers among all the
passengers, $y_t$, with a beta distribution of mean $\mu_t$ and precision
$\phi$, the law as a covariate of the mean and a score-driven filter beside
it:
$$
y_t \sim \mathrm{Beta}(\mu_t, \phi), \qquad
\mathrm{logit}\,\mu_t = \beta_0 + \beta_1\,\mathrm{law}_t + f_t .
$$
On the logit link the information of the beta distribution changes with
$\mu_t$, so each $d$ gives a different model; we use the default $d = 0$.

```{r}
#| label: ts-fit-seatbelts
belts <- data.frame(month = 1:192, law = Seatbelts[, "law"],
                    share = Seatbelts[, "front"] /
                      (Seatbelts[, "front"] + Seatbelts[, "rear"]))
fit_belt <- statmod(share ~ law + gas(p = 1, q = 1, time = month),
                    distrib = beta1_distrib(), data = belts)
summary(fit_belt)
fit_belt0 <- statmod(share ~ law, distrib = beta1_distrib(), data = belts)
c(static = AIC(fit_belt0), score_driven = AIC(fit_belt))
```

```{r}
#| label: ts-belt-numbers
#| include: false
cf_belt <- coef(fit_belt)$mu
```

With the filter beside it the law lowers the logit of the front-seat share
by `r fmt(-cf_belt[["law"]], 3)`. @fig-ts-families shows the fitted means of
the three series.

```{r}
#| label: fig-ts-families
#| fig-cap: "Three score-driven models with other families: the first 200 trade durations with the mean of the logarithmic ACD model, the hormone levels with the mean of the gamma model, and the share of front-seat passengers with the mean of the beta model. The dashed line in the last panel marks the seat-belt law."
#| echo: false
#| fig-height: 3.4
#| fig-width: 9
fam_obs <- rbind(
  data.frame(x = 1:200, y = dur$x[1:200], panel = "durations (exponential)"),
  data.frame(x = lhd$t, y = lhd$y, panel = "hormone (gamma)"),
  data.frame(x = 1969 + (belts$month - 1) / 12, y = belts$share,
             panel = "front-seat share (beta)"))
fam_fit <- rbind(
  data.frame(x = 1:200, y = predict(fit_acd, what = "mu")[1:200],
             panel = "durations (exponential)"),
  data.frame(x = lhd$t, y = predict(fit_lh, what = "mu"),
             panel = "hormone (gamma)"),
  data.frame(x = 1969 + (belts$month - 1) / 12,
             y = predict(fit_belt, what = "mu"),
             panel = "front-seat share (beta)"))
lev_fam <- c("durations (exponential)", "hormone (gamma)",
             "front-seat share (beta)")
fam_obs$panel <- factor(fam_obs$panel, levels = lev_fam)
fam_fit$panel <- factor(fam_fit$panel, levels = lev_fam)
ggplot(fam_obs, aes(x = x, y = y)) +
  geom_point(aes(colour = "observations"), size = 0.9) +
  geom_line(data = fam_fit, aes(colour = "mean of the model"),
            linewidth = 0.8) +
  geom_vline(data = data.frame(x = 1983 + 1 / 12,
                               panel = factor("front-seat share (beta)",
                                              levels = lev_fam)),
             aes(xintercept = x), linetype = 2, colour = ink) +
  facet_wrap(~panel, nrow = 1, scales = "free") +
  scale_colour_manual(values = c("observations" = book_grey,
                                 "mean of the model" = chalk)) +
  labs(x = NULL, y = NULL)
```

## Panels of short series {#sec-timeseries-panels}

The data frame `Ovary` of the package `nlme` gives the number of ovarian
follicles larger than 10 mm in eleven mares, counted on 25 to 31 days of one
cycle [@pinheiro2000]. The variable `Time` is scaled so that ovulations fall
at 0 and 1, and the mean follows the cycle, which we describe with
$\sin(2\pi t)$ and $\cos(2\pi t)$. Around that mean the counts of one mare
are correlated from one day to the next. We treat the counts as gaussian, as
@pinheiro2000 do, and we give every mare a score-driven filter:
$$
y_{it} \sim \mathrm{N}(\mu_{it}, \sigma), \qquad
\mu_{it} = \beta_0 + \beta_1 \sin(2\pi t) + \beta_2 \cos(2\pi t) + f_{it},
$$
$$
f_{it} = \kappa\, s_{i,t-1} + \xi f_{i,t-1}.
$$
The argument `by` of `gas()` names the grouping: the recursion runs over the
series of each mare separately, from its own starting level, and the
parameters are shared.

```{r}
#| label: ts-fit-ovary
data(Ovary, package = "nlme")
fit_ov <- statmod(follicles ~ sin(2 * pi * Time) + cos(2 * pi * Time) +
                    gas(p = 1, q = 1, by = Mare, time = Time),
                  distrib = gaussian1_distrib(), data = Ovary)
summary(fit_ov)
```

As in @sec-timeseries-gas, a gaussian filter on the mean is an ARMA(1, 1),
now one for each mare. The function `gls()` of `nlme` fits the same mean with
ARMA(1, 1) errors within each mare:

```{r}
#| label: ts-gls-ovary
gls_ov <- nlme::gls(follicles ~ sin(2 * pi * Time) + cos(2 * pi * Time),
                    data = Ovary, method = "ML",
                    correlation = nlme::corARMA(p = 1, q = 1, form = ~ 1 | Mare))
summary(gls_ov)
cf_ov <- coef(fit_ov)$mu
s2_ov <- exp(2 * coef(fit_ov)$sigma[["(Intercept)"]])
arma_gls <- coef(gls_ov$modelStruct$corStruct, unconstrained = FALSE)
cbind(statmod = c(cf_ov[c("(Intercept)", "sin(2 * pi * Time)",
                          "cos(2 * pi * Time)")],
                  ar1 = cf_ov[["gas.xi1"]],
                  ma1 = cf_ov[["gas.kappa1"]] / s2_ov - cf_ov[["gas.xi1"]]),
      gls = c(coef(gls_ov), arma_gls))
c(statmod = logLik(fit_ov), gls = logLik(gls_ov))
```

The estimates are close, and the log-likelihoods differ for the reason given
for the Nile: `gls()` gives the first count of each mare the variance of the
stationary process, while our filter starts every mare at the level
$f_{i1} = 0$.

### A level for each mare

The mares differ in their average number of follicles. In a score-driven
model the level around which $f_{it}$ fluctuates is $\omega/(1-\xi)$, so a
mare-specific $\omega_i$ gives each mare a level of its own. A parameter of
`gas()` takes a formula, as the parameters of `nl()` do in
@sec-nonlinear-covariates, and with `omega ~ 1 + random(~ 1 | Mare)` it is
$\omega_i = \omega + b_i$ with $b_i \sim \mathrm{N}(0, \tau)$. The formula of
the mean has no intercept, because the intercept of `omega` plays its role:

```{r}
#| label: ts-fit-ovary-omega
fit_ov_om <- statmod(follicles ~ 0 + sin(2 * pi * Time) + cos(2 * pi * Time) +
                       gas(p = 1, q = 1, omega ~ 1 + random(~ 1 | Mare),
                           by = Mare, time = Time),
                     distrib = gaussian1_distrib(), data = Ovary)
summary(fit_ov_om)
```

```{r}
#| label: ts-ovary-omega-numbers
#| include: false
cf_om <- coef(fit_ov_om)$mu
tau_om <- hyper(fit_ov_om)$estimate
lev_sd <- tau_om / (1 - cf_om[["gas.xi1"]])
```

The standard deviation of the $\omega_i$ is $\hat\tau = `r fmt(tau_om, 3)`$,
which is a standard deviation of
$\hat\tau/(1 - \hat\xi) = `r fmt(lev_sd, 2)`$ follicles between the levels
of the mares. As for any random effect, $\tau$ is estimated by restricted
maximum likelihood.

The model with shared parameters is this model with $\tau = 0$. We compare
the two by the likelihood ratio, which needs the maximized likelihood of
both, so we fit the second under `ml()`:

```{r}
#| label: ts-lr-ovary
fit_ov_ml <- statmod(follicles ~ 0 + sin(2 * pi * Time) + cos(2 * pi * Time) +
                       gas(p = 1, q = 1, omega ~ 1 + random(~ 1 | Mare),
                           by = Mare, time = Time),
                     distrib = gaussian1_distrib(), data = Ovary,
                     outer_criterion = ml())
l0 <- logLik(fit_ov)
l1 <- logLik(fit_ov_ml, type = "marginal")
lr_ov <- 2 * (as.numeric(l1) - as.numeric(l0))
lr_ov
0.5 * pchisq(lr_ov, df = 1, lower.tail = FALSE)
```

```{r}
#| label: ts-lr-ovary-numbers
#| include: false
p_ov <- 0.5 * pchisq(lr_ov, df = 1, lower.tail = FALSE)
```

The value $\tau = 0$ is the edge of the range of a standard deviation, so the
ratio has the reference distribution of @self1987: half the time it is zero
and half the time a $\chi^2_1$, which halves the p-value of the $\chi^2_1$.
The ratio is `r fmt(lr_ov, 2)`, with a p-value of `r fmt(p_ov, 3)`. Each
mare's filter already moves towards the counts of that mare, so a level for
each mare adds little.

The same family of means can be written with a random intercept in the
formula of the mean and $\omega$ held at zero, as in @sec-timeseries-gas:
$\mu_{it} = \beta_0 + c_i + \beta_1\sin(2\pi t) + \beta_2\cos(2\pi t) + f_{it}$
with $c_i \sim \mathrm{N}(0, \tau_c)$. For a given $\xi$, a mare's
intercept $c_i$ and her level $\omega_i$ move the mean in the same way when
$\omega_i - \omega = c_i\,(1 - \xi)$:

```{r}
#| label: ts-fit-ovary-intercept
fit_ov_int <- statmod(follicles ~ sin(2 * pi * Time) + cos(2 * pi * Time) +
                        random(~ 1 | Mare) +
                        gas(p = 1, q = 1, by = Mare, time = Time),
                      distrib = gaussian1_distrib(), data = Ovary,
                      outer_criterion = ml())
summary(fit_ov_int)
c(intercept = logLik(fit_ov_int, type = "marginal"),
  level = logLik(fit_ov_ml, type = "marginal"))
```

```{r}
#| label: ts-ovary-intercept-numbers
#| include: false
tau_int <- hyper(fit_ov_int)$estimate
lev_ml <- hyper(fit_ov_ml)$estimate / (1 - coef(fit_ov_ml)$mu[["gas.xi1"]])
```

The two spellings are not the same model. The distribution of the random
effects is placed on different quantities: on $c_i$ the standard deviation
of a mare's level is $\tau_c$ whatever $\xi$ is, while on $\omega_i$ it is
$\tau/(1 - \xi)$, which changes with $\xi$. Since $\xi$ is estimated
together with the effects, the two criteria differ. Here the random
intercept goes to $\hat\tau_c = `r format(signif(tau_int, 2))`$, which is the
model with shared parameters, while the level of each mare has a standard
deviation of `r fmt(lev_ml, 2)` follicles under `ml()`, with the larger
maximized likelihood.

### A loading for each mare

The loading can vary between mares in the same way. It is estimated on the
logarithmic scale, and the formula of `kappa1` acts on that scale, so
$\kappa_i = \exp(a + c_i)$ with $c_i \sim \mathrm{N}(0, \tau_\kappa)$:

```{r}
#| label: ts-fit-ovary-alpha
fit_ov_al <- statmod(follicles ~ 0 + sin(2 * pi * Time) + cos(2 * pi * Time) +
                       gas(p = 1, q = 1, omega ~ 1 + random(~ 1 | Mare),
                           kappa1 ~ 1 + random(~ 1 | Mare),
                           by = Mare, time = Time),
                     distrib = gaussian1_distrib(), data = Ovary)
summary(fit_ov_al)
```

```{r}
#| label: ts-ovary-alpha-numbers
#| include: false
h_al <- hyper(fit_ov_al)
cf_al <- coef(fit_ov_al)$mu
al_mare <- exp(cf_al[["gas.kappa1.(Intercept)"]] +
                 cf_al[paste0("gas.kappa1.random.", levels(Ovary$Mare))])
```

The loadings of the mares range from `r fmt(min(al_mare), 2)` to
`r fmt(max(al_mare), 2)`. @fig-ts-ovary shows three mares, the one with the
lowest level, one in the middle and the one with the highest, with the means
of the three models.

```{r}
#| label: fig-ts-ovary
#| fig-cap: "Number of large follicles over one cycle for three mares, with the means of the model with shared parameters, of the model with a level for each mare and of the model with a level and a loading for each mare."
#| echo: false
#| fig-height: 3.6
#| fig-width: 9
b_om <- cf_om[paste0("gas.omega.random.", levels(Ovary$Mare))]
names(b_om) <- levels(Ovary$Mare)
sel_ov <- names(sort(b_om))[c(1, 6, 11)]
sel_ov <- sel_ov[order(as.numeric(sel_ov))]
m_sh <- predict(fit_ov, what = "mu", newdata = Ovary)
m_om <- predict(fit_ov_om, what = "mu", newdata = Ovary)
m_al <- predict(fit_ov_al, what = "mu", newdata = Ovary)
rows_ov <- as.character(Ovary$Mare) %in% sel_ov
obs_ov <- data.frame(Time = Ovary$Time[rows_ov],
                     follicles = Ovary$follicles[rows_ov],
                     panel = factor(paste("mare", Ovary$Mare[rows_ov]),
                                    levels = paste("mare", sel_ov)))
ov_names <- c("shared", "level by mare", "level and loading")
curves_ov <- data.frame(
  Time = rep(obs_ov$Time, times = 3), panel = rep(obs_ov$panel, times = 3),
  follicles = c(m_sh[rows_ov], m_om[rows_ov], m_al[rows_ov]),
  curve = factor(rep(ov_names, each = nrow(obs_ov)), levels = ov_names))
ggplot(obs_ov, aes(x = Time, y = follicles)) +
  geom_point(colour = book_grey) +
  geom_line(data = curves_ov, aes(colour = curve, linetype = curve,
                                  linewidth = curve)) +
  facet_wrap(~panel, nrow = 1) +
  scale_colour_manual(values = c(ink, chalk, rust)) +
  scale_linetype_manual(values = c(3, 1, 2)) +
  scale_linewidth_manual(values = c(0.6, 0.9, 0.7)) +
  coord_cartesian(ylim = range(Ovary$follicles)) +
  labs(x = "time in the cycle", y = "follicles")
```

The argument `random` of `predict()` cannot set these random effects aside.
The level of the filter is a recursion read at the predictor, so removing a
random effect would change the whole path of the recursion. Called with
`random = "zero"` on a model with a structural term, `predict()` signals an
error.

## Forecasts beyond the series {#sec-timeseries-forecast}

`predict()` continues the filter past the last observation when `newdata`
gives times that the fit did not see. The score has mean zero, so beyond the
data the expected score is zero, and the forecast follows the recursion
without the score:
$$
f_{n+1} = \omega + \kappa\, s_n + \xi f_n, \qquad
f_{n+h} = \omega + \xi f_{n+h-1} \quad (h \ge 2).
$$
The forecast therefore approaches the level $\omega/(1 - \xi)$ around which
the series fluctuates, at the rate $\xi$. In the Nile model $\omega$ is held
at zero and the level is the intercept.

```{r}
#| label: ts-forecast-nile
future <- data.frame(year = 1971:1990)
fc_nile <- predict(fit_nile, what = "mu", newdata = future)
head(fc_nile, n = 5)
```

With `se = TRUE`, `predict()` reports a standard error and an interval for
each forecast. They are the uncertainty of the estimated parameters, carried
by the delta method through the continued recursion, whose derivative in the
parameters is computed exactly. A forecast also carries the uncertainty of
the future scores, which this standard error leaves out, and `predict()`
issues a warning that states this:

```{r}
#| label: ts-forecast-se
#| warning: true
fc_nile_se <- predict(fit_nile, what = "mu", newdata = future, se = TRUE)
head(fc_nile_se, n = 5)
```

```{r}
#| label: ts-forecast-se-numbers
#| include: false
se_fc <- fc_nile_se$se
```

The standard error is `r fmt(se_fc[1], 1)` one year ahead, rises to
`r fmt(max(se_fc), 1)` `r which.max(se_fc)` years ahead and then stays near
`r fmt(se_fc[20], 0)`. Far ahead the forecast approaches the level of the
series, the intercept, so its standard error approaches that of the
intercept, `r fmt(confint(fit_nile)["mu:(Intercept)", "se"], 1)`.

In a panel the forecast continues the series of one group. The rows of
`newdata` give the group and times after its last observation, and the
formula computes the terms of the cycle from `Time`:

```{r}
#| label: ts-forecast-ovary
last_1 <- max(Ovary$Time[Ovary$Mare == "1"])
new_1 <- data.frame(Mare = "1", Time = last_1 + 0.045 * (1:6))
fc_ov <- predict(fit_ov_om, what = "mu", newdata = new_1)
fc_ov
```

@fig-ts-forecast shows the three forecasts.

```{r}
#| label: fig-ts-forecast
#| fig-cap: "Forecasts of the score-driven models beyond the last observation: the Nile flow, the number of discoveries and the follicles of mare 1, with the 95% band that the uncertainty of the parameters alone gives. The dotted line is the level around which each series fluctuates."
#| echo: false
#| fig-height: 3.6
#| fig-width: 9
fc_disc <- predict(fit_disc, what = "mu",
                   newdata = data.frame(year = 1960:1979))
r1 <- which(Ovary$Mare == "1")
r1 <- r1[order(Ovary$Time[r1])]
cf1 <- coef(fit_ov_om)$mu
w1 <- cf1[["gas.omega.(Intercept)"]] + cf1[["gas.omega.random.1"]]
lev_ov <- cf1[["sin(2 * pi * Time)"]] * sin(2 * pi * new_1$Time) +
  cf1[["cos(2 * pi * Time)"]] * cos(2 * pi * new_1$Time) +
  w1 / (1 - cf1[["gas.xi1"]])
pan_fc <- c("Nile flow against year", "discoveries against year",
            "follicles of mare 1 against time in the cycle")
fc_names <- c("mean of the model", "forecast", "level")
fc_part <- function(k, x, y, curve) {
  data.frame(panel = factor(pan_fc[k], levels = pan_fc), x = x, y = y,
             curve = factor(curve, levels = fc_names))
}
obs_fc <- rbind(fc_part(1, nile$year, nile$flow, NA),
                fc_part(2, disc$year, disc$n, NA),
                fc_part(3, Ovary$Time[r1], Ovary$follicles[r1], NA))
curves_fc <- rbind(
  fc_part(1, nile$year, predict(fit_nile, what = "mu"), fc_names[1]),
  fc_part(1, future$year, fc_nile, fc_names[2]),
  fc_part(1, c(1871, 1990), rep(cf_nile[["(Intercept)"]], 2), fc_names[3]),
  fc_part(2, disc$year, predict(fit_disc, what = "mu"), fc_names[1]),
  fc_part(2, 1960:1979, fc_disc, fc_names[2]),
  fc_part(2, c(1860, 1979), rep(exp(coef(fit_disc)$mu[["(Intercept)"]]), 2),
          fc_names[3]),
  fc_part(3, Ovary$Time[r1], m_om[r1], fc_names[1]),
  fc_part(3, new_1$Time, fc_ov, fc_names[2]),
  fc_part(3, new_1$Time, lev_ov, fc_names[3]))
band_disc <- predict(fit_disc, what = "mu",
                     newdata = data.frame(year = 1960:1979), se = TRUE)
band_ov <- predict(fit_ov_om, what = "mu", newdata = new_1, se = TRUE)
bands_fc <- rbind(
  data.frame(panel = factor(pan_fc[1], levels = pan_fc), x = future$year,
             lower = fc_nile_se$lower, upper = fc_nile_se$upper),
  data.frame(panel = factor(pan_fc[2], levels = pan_fc), x = 1960:1979,
             lower = band_disc$lower, upper = band_disc$upper),
  data.frame(panel = factor(pan_fc[3], levels = pan_fc), x = new_1$Time,
             lower = band_ov$lower, upper = band_ov$upper))
ggplot(obs_fc, aes(x = x, y = y)) +
  geom_ribbon(data = bands_fc, aes(x = x, ymin = lower, ymax = upper,
                                   fill = "95% band of the parameters"),
              inherit.aes = FALSE, alpha = 0.2) +
  geom_point(colour = book_grey, size = 1) +
  geom_line(data = curves_fc, aes(colour = curve, linetype = curve,
                                  linewidth = curve)) +
  facet_wrap(~panel, nrow = 1, scales = "free") +
  scale_fill_manual(values = rust) +
  scale_colour_manual(values = c(chalk, rust, ink)) +
  scale_linetype_manual(values = c(1, 1, 3)) +
  scale_linewidth_manual(values = c(0.7, 0.9, 0.5)) +
  labs(x = NULL, y = NULL)
```

## Hidden regimes: regime() {#sec-timeseries-regime}

The data frame `geyser` of the package `MASS` records
`r nrow(MASS::geyser)` successive eruptions of the Old Faithful geyser in
Yellowstone National Park [@azzalini1990; @venables2002]. The variable
`waiting` is the time in minutes between the start of one eruption and the
start of the next. The waiting times form two groups, one around 55 minutes
and one around 80 minutes, and a short waiting time is almost never followed
by another short one. A hidden Markov model with two states describes this
[@hamilton1989; @zucchini2016]:
$$
y_t \sim \mathrm{N}(\mu_{S_t}, \sigma), \qquad
P(S_t = j \mid S_{t-1} = i) = p_{ij}.
$$
The state $S_t$ is not observed. It follows a Markov chain with transition
matrix $P = (p_{ij})$, and each state has its own mean. In econometrics the
same model is called a Markov-switching or regime-switching model. The
likelihood is a sum over all paths of the chain, and the forward recursion
computes it in one pass over the series, starting from the stationary
distribution of the chain.

```{r}
#| label: ts-fit-geyser
data(geyser, package = "MASS")
fit_gey <- statmod(waiting ~ regime(k = 2), distrib = gaussian1_distrib(),
                   data = geyser)
print(summary(fit_gey), notes = TRUE)
```

The term `regime(k = 2)` adds the level of the current state to the
equation of `mu`. The levels are written as a first level and positive gaps,
so the second state is always the one with the larger mean. The intercept of
the equation plays the role of the first level, which is held at zero, as
the level of `gas()` is. The summary gives the gap and the four transition
probabilities $p_{ij}$, named `p1.1`, `p1.2`, `p2.1` and `p2.2`. They are
estimated through the logarithms of the ratios within each row, and their
intervals are built on the logit scale and mapped back.

```{r}
#| label: ts-geyser-numbers
#| include: false
cg <- coef(fit_gey)$mu
mu_gey <- cg[["(Intercept)"]] + c(0, cg[["regime.gap2"]])
```

The two means are `r fmt(mu_gey[1], 1)` and `r fmt(mu_gey[2], 1)` minutes.
After a short waiting time the next one is long with probability
`r fmt(cg[["regime.p1.2"]], 3)`: $\hat p_{11}$ is at the edge of its range,
where the logarithm of its ratio runs to $-\infty$. There the derivative of
the probability with respect to that coordinate goes to zero, and the delta
method would report a standard error that vanishes exactly where the data
stop locating the probability. The summary therefore reports the first row
of $P$ without a standard error and an interval, and a note below the
summary gives the reason.
After a long waiting time the next one is short with probability
`r fmt(cg[["regime.p2.1"]], 2)`.

The function `statmod_latent()` gives the probability of each state at each
time given the whole series, which the forward recursion and a backward
recursion compute together:

```{r}
#| label: ts-latent-geyser
st_gey <- statmod_latent(fit_gey)
head(st_gey, n = 4)
```

The package `depmixS4` fits hidden Markov models [@visser2010]. Its forward
recursion, run at our estimates with the initial probabilities set to the
stationary distribution of the chain, gives the same log-likelihood and the
same probabilities of the states. Its own fit is a different model: each
state has its own standard deviation, and the probabilities of the first
state are estimated as two further parameters. We start its EM algorithm
from our estimates, without the random starts it uses by default.

```{r}
#| label: ts-depmix-geyser
cf_gey <- coef(fit_gey)$mu
means_gey <- cf_gey[["(Intercept)"]] + c(0, cf_gey[["regime.gap2"]])
P_gey <- matrix(cf_gey[c("regime.p1.1", "regime.p1.2", "regime.p2.1",
                         "regime.p2.2")], nrow = 2, byrow = TRUE)
delta_gey <- c(P_gey[2, 1], P_gey[1, 2]) / (P_gey[1, 2] + P_gey[2, 1])
sd_gey <- exp(coef(fit_gey)$sigma[["(Intercept)"]])
hmm_gey <- depmixS4::depmix(waiting ~ 1, data = geyser, nstates = 2,
                            family = gaussian())
hmm_gey <- depmixS4::setpars(hmm_gey, c(delta_gey, P_gey[1, ], P_gey[2, ],
                                        means_gey[1], sd_gey,
                                        means_gey[2], sd_gey))
fb_gey <- depmixS4::forwardbackward(hmm_gey)
c(statmod = logLik(fit_gey), depmixS4 = fb_gey$logLike)
max(abs(as.matrix(st_gey) - fb_gey$gamma))
hmm_fit <- depmixS4::fit(hmm_gey, verbose = FALSE,
                         emcontrol = depmixS4::em.control(random.start = FALSE))
depmixS4::getpars(hmm_fit)
ll_hmm <- depmixS4::forwardbackward(hmm_fit)$logLike
bic_hmm <- -2 * ll_hmm + log(nrow(geyser)) * depmixS4::freepars(hmm_fit)
c(statmod = BIC(fit_gey), depmixS4 = bic_hmm)
```

```{r}
#| label: ts-depmix-numbers
#| include: false
gap_fb <- max(abs(as.matrix(st_gey) - fb_gey$gamma))
```

The two forward recursions agree, and the probabilities of the states differ
by at most `r format(signif(gap_fb, 1))`. With a standard deviation for
each state the fit of `depmixS4` has a lower BIC. In `statmodels7` a regime
term sits in the equation of one parameter, so the standard deviation of this
model is common to the two states.

@fig-ts-geyser shows the first 100 waiting times with the mean of the model,
which `predict()` computes as the average of the two means weighted by the
probabilities of the states. The background shows the most probable state of
each eruption, and the lower panel the probability of the long state.

```{r}
#| label: ts-state-runs
#| include: false
# consecutive runs of the most probable state, as rectangles for a background
state_runs <- function(x, prob) {
  s <- max.col(as.matrix(prob), ties.method = "first")
  r <- rle(s)
  end <- cumsum(r$lengths)
  start <- end - r$lengths + 1
  step <- if (length(x) > 1) median(diff(x)) / 2 else 0.5
  data.frame(xmin = x[start] - step, xmax = x[end] + step,
             state = factor(paste("state", r$values)))
}
```

```{r}
#| label: fig-ts-geyser
#| fig-cap: "Top: the first 100 waiting times of Old Faithful with the mean of the two-state model; the background shows the most probable state of each eruption. Bottom: the probability that each waiting time belongs to the long state, given the whole series."
#| echo: false
#| fig-height: 5
ii <- 1:100
pan_gey <- c("waiting (min)", "P(long state)")
on_gey <- function(df, k) {
  df$panel <- factor(pan_gey[k], levels = pan_gey)
  df
}
runs_gey <- on_gey(state_runs(ii, st_gey[ii, ]), k = 1)
ggplot(mapping = aes(x = x, y = y)) +
  geom_rect(data = runs_gey, aes(xmin = xmin, xmax = xmax, ymin = -Inf,
                                 ymax = Inf, fill = state),
            inherit.aes = FALSE, alpha = 0.18) +
  geom_blank(data = on_gey(data.frame(x = NA_real_, y = c(0, 1)), k = 2)) +
  geom_point(data = on_gey(data.frame(x = ii, y = geyser$waiting[ii]), k = 1),
             colour = book_grey, size = 1.2) +
  geom_line(data = on_gey(data.frame(x = ii,
                                     y = predict(fit_gey, what = "mu")[ii]),
                          k = 1),
            aes(colour = "mean of the model"), linewidth = 0.7) +
  geom_linerange(data = on_gey(data.frame(x = ii, y = st_gey$state2[ii]),
                               k = 2),
                 aes(ymin = 0, ymax = y), colour = rust, linewidth = 0.8) +
  facet_wrap(~panel, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_colour_manual(values = chalk) +
  scale_fill_manual(values = c("state 1" = "#5B7DB1", "state 2" = rust),
                    breaks = c("state 1", "state 2"),
                    labels = c("short state", "long state")) +
  labs(x = "eruption", y = NULL) +
  theme(strip.placement = "outside", strip.background = element_blank(),
        strip.text = element_text(size = rel(1)))
```

### More states, and more than one start

The likelihood of a regime model often has several local maxima, and the
starting point decides which one the fit reaches. The first start puts the
levels at quantiles of the response. The argument `n_start` of `regime()`
adds further starts around it, and the fit keeps the best. With three
states:

```{r}
#| label: ts-fit-geyser3
fit_gey3 <- statmod(waiting ~ regime(k = 3, n_start = 3),
                    distrib = gaussian1_distrib(), data = geyser)
summary(fit_gey3)
c(two = BIC(fit_gey), three = BIC(fit_gey3))
```

```{r}
#| label: ts-geyser3-numbers
#| include: false
cg3 <- coef(fit_gey3)$mu
mu_gey3 <- cg3[["(Intercept)"]] + c(0, cg3[["regime.gap2"]],
                                    cg3[["regime.gap2"]] + cg3[["regime.gap3"]])
```

The three means are `r fmt(mu_gey3[1], 1)`, `r fmt(mu_gey3[2], 1)` and
`r fmt(mu_gey3[3], 1)` minutes, and the BIC prefers three states. The model
with two states is the one with three in which two states coincide, a point
where the transition probabilities of the third state are not defined, so
the likelihood ratio between them has no $\chi^2$ reference
[@davies1987]. The criteria do not have this problem. @fig-ts-geyser3 shows
the first 100 waiting times with the mean of the three-state model and the
most probable state of each eruption.

```{r}
#| label: fig-ts-geyser3
#| fig-cap: "The first 100 waiting times of Old Faithful with the mean of the three-state model. The background shows the most probable state of each eruption."
#| echo: false
#| fig-height: 3.4
st_gey3 <- statmod_latent(fit_gey3)
runs_gey3 <- state_runs(ii, st_gey3[ii, ])
ggplot(data.frame(x = ii, y = geyser$waiting[ii]), aes(x = x, y = y)) +
  geom_rect(data = runs_gey3, aes(xmin = xmin, xmax = xmax, ymin = -Inf,
                                  ymax = Inf, fill = state),
            inherit.aes = FALSE, alpha = 0.18) +
  geom_point(colour = book_grey, size = 1.2) +
  geom_line(data = data.frame(x = ii, y = predict(fit_gey3, what = "mu")[ii]),
            aes(colour = "mean of the model"), linewidth = 0.7) +
  scale_colour_manual(values = chalk) +
  scale_fill_manual(values = c("state 1" = "#5B7DB1", "state 2" = "#C49A2C",
                               "state 3" = rust),
                    breaks = c("state 1", "state 2", "state 3"),
                    labels = sprintf("state %d, mean %.1f", 1:3, mu_gey3)) +
  labs(x = "eruption", y = "waiting (min)")
```

## Regimes with covariates, and on the scale {#sec-timeseries-regime-more}

The other terms of an equation can sit beside `regime()`. The waiting time
after an eruption depends on how long the eruption lasted. We add the
duration of the previous eruption, `dlag`, to the equation of the mean:
$$
y_t \sim \mathrm{N}(\mu_t, \sigma), \qquad
\mu_t = \beta_0 + \beta_1\, d_{t-1} + \lambda_{S_t}.
$$
The first eruption has no previous one, so we fit the model to the other
`r nrow(geyser) - 1` waiting times, and the model without states to the same
ones.

```{r}
#| label: ts-fit-geyser-dlag
geyser$dlag <- c(NA, head(geyser$duration, n = -1))
fit_gey_d <- statmod(waiting ~ dlag + regime(k = 2),
                     distrib = gaussian1_distrib(), data = geyser[-1, ])
summary(fit_gey_d)
fit_gey_d0 <- statmod(waiting ~ dlag, distrib = gaussian1_distrib(),
                      data = geyser[-1, ])
c(static = BIC(fit_gey_d0), regime = BIC(fit_gey_d))
```

```{r}
#| label: ts-geyser-dlag-numbers
#| include: false
cgd <- coef(fit_gey_d)$mu
```

A covariate shifts the mean of every state by the same amount, here
`r fmt(cgd[["dlag"]], 2)` minutes for each minute of the previous eruption.
The gap between the states falls from `r fmt(cg[["regime.gap2"]], 1)` to
`r fmt(cgd[["regime.gap2"]], 1)` minutes, because the duration of the
previous eruption explains most of the difference between short and long
waiting times, and the BIC still prefers the model with the states. The
transition probabilities of `regime()` are constant: they do not depend on
covariates, and a covariate can enter the model only through the equations
of the parameters.

The term can also sit in the equation of another parameter. On the DAX
returns, a chain on the scale gives a model with a quiet state and a
turbulent state:
$$
r_t \sim \mathrm{N}(\mu, \sigma_t), \qquad
\log \sigma_t = \gamma_0 + \lambda_{S_t}.
$$

```{r}
#| label: ts-fit-dax-regime
fit_dax_r <- statmod(r ~ 1 | sigma ~ regime(k = 2), distrib = gaussian1_distrib(),
                     data = dax)
summary(fit_dax_r)
c(score_driven = AIC(fit_dax), regime = AIC(fit_dax_r))
```

```{r}
#| label: ts-dax-regime-numbers
#| include: false
cdr <- coef(fit_dax_r)$sigma
sd_dr <- exp(cdr[["(Intercept)"]] + c(0, cdr[["regime.gap2"]]))
st_dax <- statmod_latent(fit_dax_r)
```

The two states have standard deviations of `r fmt(sd_dr[1], 2)` and
`r fmt(sd_dr[2], 2)` percent. Each state lasts long: the probability of
staying in the quiet state is `r fmt(cdr[["regime.p1.1"]], 3)` and that of
staying in the turbulent state `r fmt(cdr[["regime.p2.2"]], 3)`. With the
gaussian distribution this model has a lower AIC than the score-driven model
of @sec-timeseries-volatility. @fig-ts-dax-regime shows the fitted standard
deviation over the absolute returns, with the most probable state of each day
in the background, and the probability of the turbulent state. `predict()`
gives the standard deviation as the exponential of the predictor averaged
over the probabilities of the states.

```{r}
#| label: fig-ts-dax-regime
#| fig-cap: "Top: absolute daily returns of the DAX with the standard deviation of the two-state model on the scale; the background shows the most probable state of each day. Bottom: the probability of the turbulent state, given the whole series."
#| echo: false
#| fig-height: 5
pan_reg <- c("absolute return (%)", "P(turbulent state)")
reg_ret <- data.frame(x = yr, hi = abs(dax$r),
                      panel = factor(pan_reg[1], levels = pan_reg))
reg_sd <- data.frame(x = yr, y = predict(fit_dax_r, what = "sigma"),
                     panel = factor(pan_reg[1], levels = pan_reg))
reg_prob <- data.frame(x = yr, y = st_dax$state2,
                       panel = factor(pan_reg[2], levels = pan_reg))
reg_lim <- data.frame(x = NA_real_, y = c(0, 1),
                      panel = factor(pan_reg[2], levels = pan_reg))
runs_dax <- state_runs(yr, st_dax)
ggplot(mapping = aes(x = x)) +
  geom_rect(data = runs_dax, aes(xmin = xmin, xmax = xmax, ymin = -Inf,
                                 ymax = Inf, fill = state),
            inherit.aes = FALSE, alpha = 0.18) +
  geom_blank(data = reg_lim, aes(y = y)) +
  geom_linerange(data = reg_ret, aes(ymin = 0, ymax = hi),
                 colour = book_grey, linewidth = 0.3) +
  geom_line(data = reg_sd, aes(y = y, colour = "standard deviation"),
            linewidth = 0.6) +
  geom_line(data = reg_prob, aes(y = y), colour = rust, linewidth = 0.5) +
  facet_wrap(~panel, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_colour_manual(values = chalk) +
  scale_fill_manual(values = c("state 1" = "#5B7DB1", "state 2" = rust),
                    breaks = c("state 1", "state 2"),
                    labels = c("quiet state", "turbulent state")) +
  scale_x_continuous(breaks = seq(ceiling(min(yr)), floor(max(yr)))) +
  labs(x = "year", y = NULL) +
  theme(strip.placement = "outside", strip.background = element_blank(),
        strip.text = element_text(size = rel(1)))
```

```{r}
#| label: gate-13
#| include: false
assert_timeseries_ok(
  fits = list(nile = fit_nile, nile2 = fit_nile2, dax = fit_dax,
              dax_t = fit_dax_t, disc = fit_disc, disc0 = fit_disc0,
              disc_nb = fit_disc_nb, disc_gnb = fit_disc_gnb, acd = fit_acd,
              disc_h = fit_disc_h, disc_1 = fit_disc_1, garch = fit_garch,
              ingarch = fit_ingarch, acd_lin = fit_acd_lin, nile_1 = fit_nile_1,
              lh = fit_lh, lh0 = fit_lh0, belt = fit_belt, belt0 = fit_belt0,
              ov = fit_ov, ov_om = fit_ov_om, ov_ml = fit_ov_ml,
              ov_int = fit_ov_int, ov_al = fit_ov_al, gey = fit_gey,
              gey3 = fit_gey3, gey_d = fit_gey_d, gey_d0 = fit_gey_d0,
              dax_r = fit_dax_r),
  ref = list(arima_nile = arima_nile, garch_n = garch_n, garch_t = garch_t,
             tsc_p = tsc_p, tsc_nb = tsc_nb, acd = acd, tsc_id = tsc_id,
             acd_lin = acd_lin, gls_ov = gls_ov,
             fb_gey = fb_gey, hmm_fit = hmm_fit),
  data = list(nile = nile, dax = dax, disc = disc, dur = dur, lhd = lhd,
              belts = belts, ov = Ovary, geyser = geyser, gey = geyser[-1, ]),
  claims = list(lr_nile = lr_nile, fc_nile = fc_nile, se_fc = se_fc,
                lr_ov = lr_ov, fc_ov = fc_ov, new_1 = new_1,
                tau_int = tau_int, lev_ml = lev_ml, st_gey = st_gey,
                gap_fb = gap_fb, bic_hmm = bic_hmm, st_dax = st_dax))
```
