# Choosing {#sec-choosing}

```{r}
#| label: setup-14
#| include: false
source("R/_setup.R")
source("R/choosing-certificates.R")
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x, d = 1) paste0(fmt(100 * x, d), "%")
sci <- function(x, d = 1) {
  e <- floor(log10(abs(x)))
  sprintf("%.*f \\times 10^{%d}", d, x / 10^e, e)
}
data(species, package = "gamlss.data")
data(rent, package = "gamlss.data")
data(abdom, package = "gamlss.data")
data(film90, package = "gamlss.data")
data(mcycle, package = "MASS")
data(GAGurine, package = "MASS")
```

In the previous chapters we fitted each model with a distribution, a
criterion for the hyperparameters and an optimizer that we chose for it.
This chapter explains how we make these choices. Each section works through
a data set of its own: we choose a distribution for counts and for positive
measurements, we compare the criteria that estimate the hyperparameters,
we read the certificate of a fit and decide what to do when it does not
converge, and we close with the models that `statmodels7` cannot fit. Most
of the data sets come from the `gamlss.data` package, which accompanies the
book of @stasinopoulos2017 on distributional regression.

## The distribution {#sec-choosing-distribution}

The type of the response narrows the choice of the distribution: a count
needs a distribution on the integers, a positive measurement a distribution
on the positive half-line, a proportion a distribution on the unit interval.
Within one type, the families differ mainly in how the variance and the
shape depend on the mean. We compare them in three ways. The quantile
residuals of @dunn1996 are standard normal when the distribution is right,
so their quantile plot shows a wrong family directly. The Pearson ratio,
the sum of the squared Pearson residuals divided by the residual degrees of
freedom, is close to one when the variance function is right. The AIC
compares the families on the scale of the likelihood.

### Counts

The data frame `species` gives the number of fish species `fish` in 70
lakes and the area of each lake [@stasinopoulos2017]. The number of species
grows with the logarithm of the area, and we fit four count distributions
with the same equation for the mean: the Poisson distribution, the two
negative binomial distributions of @sec-glm-overdispersion, and the
Poisson-inverse gaussian distribution with the dispersion $\sigma$
(`pig1_distrib()`).

```{r}
#| label: fit-species
species$log_lake <- log(species$lake)
sp_pois <- statmod(fish ~ log_lake, distrib = poisson_distrib(), data = species,
                   outer_criterion = ml())
sp_nb2 <- statmod(fish ~ log_lake, distrib = negbin2_distrib(), data = species,
                  outer_criterion = ml())
sp_nb1 <- statmod(fish ~ log_lake, distrib = negbin1_distrib(), data = species,
                  outer_criterion = ml())
sp_pig <- statmod(fish ~ log_lake, distrib = pig1_distrib(), data = species,
                  outer_criterion = ml())
summary(sp_nb2)
```

The criterion is `ml()` in all four fits. With the default `reml()`,
`statmod()` estimates the coefficient of a dispersion, here the intercept of
$\log\theta$, by maximizing the restricted likelihood, as `lm()` estimates
the residual variance with $n - p$ in the denominator. The AIC needs the
maximum likelihood estimate:

```{r}
#| label: species-default
sp_nb2_reml <- statmod(fish ~ log_lake, distrib = negbin2_distrib(),
                       data = species)
c(reml = exp(coef(sp_nb2_reml)$theta[[1]]), ml = exp(coef(sp_nb2)$theta[[1]]),
  glm.nb = MASS::glm.nb(fish ~ log_lake, data = species)$theta)
```

The fit with `ml()` gives the estimate of `glm.nb()` of the `MASS` package
[@venables2002]. We compute the Pearson ratio and the AIC of the four
families:

```{r}
#| label: species-table
pearson_ratio <- function(fit) {
  r <- residuals(fit, type = "pearson")
  sum(r^2) / (nobs(fit) - sum(fit@edf$edf))
}
sp_table <- data.frame(
  family = c("poisson", "negbin2", "negbin1", "pig1"),
  logLik = c(logLik(sp_pois), logLik(sp_nb2), logLik(sp_nb1), logLik(sp_pig)),
  AIC = c(AIC(sp_pois), AIC(sp_nb2), AIC(sp_nb1), AIC(sp_pig)),
  pearson = c(pearson_ratio(sp_pois), pearson_ratio(sp_nb2),
              pearson_ratio(sp_nb1), pearson_ratio(sp_pig)))
sp_table
```

The Pearson ratio of the Poisson model is
`r fmt(sp_table$pearson[1], 1)`: the counts vary far more than a Poisson
distribution allows. The three distributions with a dispersion bring the
ratio close to one. The AIC orders them, and the Poisson-inverse gaussian
distribution has the smallest. The package `gamlss` [@rigby2005] fits the
same distributions under the names `NBI` and `PIG`:

```{r}
#| label: species-gamlss
gl_nbi <- gamlss::gamlss(fish ~ log_lake, family = gamlss.dist::NBI(),
                         data = species, trace = FALSE)
gl_pig <- gamlss::gamlss(fish ~ log_lake, family = gamlss.dist::PIG(),
                         data = species, trace = FALSE)
rbind(negbin2 = c(statmod = logLik(sp_nb2), gamlss = logLik(gl_nbi)),
      pig1 = c(statmod = logLik(sp_pig), gamlss = logLik(gl_pig)))
```

The quantile residuals show the difference between the Poisson model and
the others (@fig-species-resid). We draw them with a fixed seed, because
the quantile residual of a count is randomized within the jump of the
distribution function:

```{r}
#| label: fig-species-resid
#| fig-cap: "Quantile residuals of the `species` models against gaussian quantiles. Left: Poisson. Right: negative binomial with variance $\\mu + \\mu^2/\\theta$. The line is the identity."
#| fig-height: 3.6
qr_pois <- residuals(sp_pois, seed = 1)
qr_nb2 <- residuals(sp_nb2, seed = 1)
qq <- data.frame(residual = c(qr_pois, qr_nb2),
                 model = rep(c("poisson", "negbin2"), each = nrow(species)))
qq$model <- factor(qq$model, levels = c("poisson", "negbin2"))
ggplot(qq, aes(sample = residual)) +
  stat_qq(colour = book_grey, size = 0.9) +
  geom_abline(intercept = 0, slope = 1, colour = book_palette[1]) +
  facet_wrap(facets = vars(model), scales = "free_y") +
  labs(x = "gaussian quantile", y = "quantile residual")
```

The residuals of the negative binomial model lie on the identity line, and
their standard deviation is `r fmt(sd(qr_nb2), 3)`. Those of the Poisson
model have a standard deviation of `r fmt(sd(qr_pois), 2)`.

A distributional model can also give the dispersion an equation. The
dispersion of the negative binomial model may change with the size of the
lake:

```{r}
#| label: species-theta
sp_nb2_th <- statmod(fish ~ log_lake | theta ~ log_lake,
                     distrib = negbin2_distrib(), data = species,
                     outer_criterion = ml())
coef(sp_nb2_th)$theta
c(constant = AIC(sp_nb2), modelled = AIC(sp_nb2_th))
```

The AIC falls by `r fmt(AIC(sp_nb2) - AIC(sp_nb2_th), 2)`, more than the
AIC of the Poisson-inverse gaussian model falls below that of the negative
binomial one. Choosing a distribution and choosing its equations are one
choice, and we compare all the candidates on the same AIC.

### Positive measurements

The data frame `rent` gives the monthly net rent `R`, in German marks, of
`r nrow(rent)` flats in Munich in 1993, with the floor space `Fl` in square
metres, the year of construction `A` and the location `loc`, in three
classes [@rigby2005]. The rent grows with the floor space and with the year
in two curves, which we model with smooth terms. We fit four distributions
for positive responses: the gamma, the lognormal, the inverse gaussian and
the generalized gamma, which has a third parameter for the shape.

```{r}
#| label: fit-rent
rent_gamma <- statmod(R ~ s(Fl) + s(A) + loc, distrib = gamma1_distrib(),
                      data = rent)
rent_lnorm <- statmod(R ~ s(Fl) + s(A) + loc, distrib = lognormal1_distrib(),
                      data = rent)
rent_ig <- statmod(R ~ s(Fl) + s(A) + loc, distrib = invgauss1_distrib(),
                   data = rent)
rent_gg <- statmod(R ~ s(Fl) + s(A) + loc, distrib = gengamma1_distrib(),
                   data = rent)
summary(rent_gamma)
```

The models have smooth terms, so we compare them with the conditional AIC
of @sec-beyond-likelihood, which `summary()` prints:

```{r}
#| label: rent-table
caic <- function(fit) summary(fit)@aic
rent_table <- data.frame(
  family = c("gamma", "lognormal", "inverse gaussian", "generalized gamma"),
  logLik = c(logLik(rent_gamma), logLik(rent_lnorm), logLik(rent_ig),
             logLik(rent_gg)),
  cAIC = c(caic(rent_gamma), caic(rent_lnorm), caic(rent_ig), caic(rent_gg)))
rent_table
```

The gamma and the generalized gamma distributions have nearly the same
cAIC, `r fmt(rent_table$cAIC[1], 1)` and `r fmt(rent_table$cAIC[4], 1)`,
and the lognormal and the inverse gaussian distributions are worse by
`r fmt(rent_table$cAIC[2] - rent_table$cAIC[1], 0)` and
`r fmt(rent_table$cAIC[3] - rent_table$cAIC[1], 0)`. The quantile residuals
show why (@fig-rent-resid): the lognormal and the inverse gaussian
distributions are skewed to the right more than the rents are, so the
residuals of the largest rents fall below the identity line.

```{r}
#| label: fig-rent-resid
#| fig-cap: "Quantile residuals of the four `rent` models against gaussian quantiles. The line is the identity."
#| fig-height: 5
rq <- data.frame(
  residual = c(residuals(rent_gamma), residuals(rent_lnorm),
               residuals(rent_ig), residuals(rent_gg)),
  model = rep(rent_table$family, each = nrow(rent)))
rq$model <- factor(rq$model, levels = rent_table$family)
ggplot(rq, aes(sample = residual)) +
  stat_qq(colour = book_grey, size = 0.5) +
  geom_abline(intercept = 0, slope = 1, colour = book_palette[1]) +
  facet_wrap(facets = vars(model), ncol = 2) +
  labs(x = "gaussian quantile", y = "quantile residual")
```

The gamma model with the log link is also a generalized additive model, and
`mgcv` fits it with the same REML criterion [@wood2017]:

```{r}
#| label: rent-mgcv
g_rent <- mgcv::gam(R ~ s(Fl) + s(A) + loc, family = Gamma(link = "log"),
                    data = rent, method = "REML")
c(statmod = as.numeric(logLik(rent_gamma)), mgcv = as.numeric(logLik(g_rent)))
```

The two log-likelihoods agree to `r fmt(abs(as.numeric(logLik(rent_gamma)) - as.numeric(logLik(g_rent))), 2)`;
the bases of the smooth terms of the two packages differ, so the fitted
curves are not identical. The dispersion $\phi$ of the gamma distribution
can have an equation as well. We let it depend on the floor space and the
location:

```{r}
#| label: rent-phi
rent_phi <- statmod(R ~ s(Fl) + s(A) + loc | phi ~ s(Fl) + loc,
                    distrib = gamma1_distrib(), data = rent)
c(constant = caic(rent_gamma), modelled = caic(rent_phi))
```

The cAIC falls by `r fmt(caic(rent_gamma) - caic(rent_phi), 1)`, much more
than it separates the gamma from the generalized gamma distribution. For
these data the equation of the dispersion matters more than the family.

## The hyperparameters {#sec-choosing-hyper}

A penalty has hyperparameters, for example the smoothing parameter
$\lambda$ of a smooth term or the variance of a random effect.
`statmodels7` estimates each one in one of two ways, according to the
penalty.

A penalty with derivatives of every order, such as the penalty of a smooth
term, a ridge or a random effect, has its hyperparameters estimated by
maximizing a criterion. The argument `outer_criterion` of `statmod()`
selects it: `reml()`, the default, `ml()`, or the prediction criteria
`aic()` and `bic()`. The first two are the marginal likelihoods of
@sec-beyond-likelihood. The prediction criteria compute
$-2\ell + \kappa\tau$ with $\kappa = 2$ for AIC and $\kappa = \log n$ for
BIC, where $\ell$ is the conditional log-likelihood and $\tau$ is the
effective number of parameters.

A penalty with a kink at zero, such as the lasso, the elastic net, SCAD and
MCP, has no second derivative at zero, and the marginal criteria need it.
For these penalties `statmod()` fits the model along a decreasing sequence
of values of $\lambda$ and keeps the value with the smallest prediction
criterion. We select this criterion with the argument `sparse_criterion`:
`bic()`, the default, `aic()` or `cv()`.

### Four criteria for two smooth terms

The data frame `abdom` gives the abdominal circumference `y`, in
millimetres, of `r nrow(abdom)` fetuses and their gestational age `x`, in
weeks [@stasinopoulos2017]. Both the mean and the spread of the
circumference grow with the age. We give both parameters of a gaussian
distribution a smooth term and estimate the two smoothing parameters with
each of the four criteria:

```{r}
#| label: fit-abdom
f_ab <- y ~ s(x) | sigma ~ s(x)
ab_reml <- statmod(f_ab, distrib = gaussian1_distrib(), data = abdom)
ab_ml <- statmod(f_ab, distrib = gaussian1_distrib(), data = abdom,
                 outer_criterion = ml())
ab_aic <- statmod(f_ab, distrib = gaussian1_distrib(), data = abdom,
                  outer_criterion = aic())
ab_bic <- statmod(f_ab, distrib = gaussian1_distrib(), data = abdom,
                  outer_criterion = bic())
summary(ab_reml)
```

```{r}
#| label: abdom-table
ab_lambda <- function(fit) hyper(fit)$estimate
ab_table <- rbind(
  reml = c(ab_lambda(ab_reml), edf = sum(ab_reml@edf$edf), logLik = logLik(ab_reml)),
  ml = c(ab_lambda(ab_ml), edf = sum(ab_ml@edf$edf), logLik = logLik(ab_ml)),
  aic = c(ab_lambda(ab_aic), edf = sum(ab_aic@edf$edf), logLik = logLik(ab_aic)),
  bic = c(ab_lambda(ab_bic), edf = sum(ab_bic@edf$edf), logLik = logLik(ab_bic)))
colnames(ab_table)[1:2] <- c("lambda_mu", "lambda_sigma")
ab_table
```

The four criteria agree on the spread: the smoothing parameter of $\sigma$
is above $`r sci(min(ab_table[, "lambda_sigma"]))`$ in all of them, so the
smooth term of $\sigma$ reduces to its straight line, and the standard
deviation grows exponentially with the age. They differ on the mean. REML,
ML and AIC give the curve of the mean about
`r fmt(ab_table["reml", "edf"] - 4, 1)` effective degrees of freedom beyond
its straight line, and BIC, whose penalty grows with $\log n$, gives it
`r fmt(ab_table["bic", "edf"] - 4, 1)`. @fig-abdom shows the two curves with
the bands of two standard deviations.

```{r}
#| label: fig-abdom
#| fig-cap: "Abdominal circumference against gestational age. The lines are the fitted mean and the mean plus and minus two standard deviations, with the smoothing parameters estimated by REML and by BIC."
#| fig-height: 4
xg <- data.frame(x = seq(from = min(abdom$x), to = max(abdom$x), length.out = 200))
band <- function(fit, label) {
  m <- predict(fit, what = "mu", newdata = xg)
  s <- predict(fit, what = "sigma", newdata = xg)
  data.frame(x = xg$x, mean = m, lower = m - 2 * s, upper = m + 2 * s,
             criterion = label)
}
bands_ab <- rbind(band(ab_reml, "REML"), band(ab_bic, "BIC"))
ggplot(abdom, aes(x = x, y = y)) +
  geom_point(colour = book_grey, size = 0.6, alpha = 0.6) +
  geom_line(data = bands_ab, aes(y = mean, colour = criterion, linetype = criterion)) +
  geom_line(data = bands_ab, aes(y = lower, colour = criterion, linetype = criterion)) +
  geom_line(data = bands_ab, aes(y = upper, colour = criterion, linetype = criterion)) +
  labs(x = "gestational age (weeks)", y = "abdominal circumference (mm)")
```

The package `mgcv` fits the same model with its family `gaulss()`, by REML:

```{r}
#| label: abdom-mgcv
g_ab <- mgcv::gam(list(y ~ s(x), ~ s(x)), family = mgcv::gaulss(),
                  data = abdom, method = "REML")
c(statmod = as.numeric(logLik(ab_reml)), mgcv = as.numeric(logLik(g_ab)))
```

The effective number of parameters $\tau$ treats the hyperparameters as
known, although the fit estimated them from the same data. @greven2010
showed that the conditional AIC of @vaida2005, computed with this $\tau$,
favors the larger of two models. @woodpya2016 proposed to add the
uncertainty of the smoothing parameters to $\tau$, through the
unconditional variance $V_u$ of @sec-beyond-variances:
$$
\tau_u = \operatorname{tr}(V_u H) = \tau + \operatorname{tr}\{(V_u - V_b) H\} .
$$
`summary(fit, correct = TRUE)` computes the cAIC with $\tau_u$:

```{r}
#| label: abdom-correct
s_naive <- summary(ab_reml)
s_corr <- summary(ab_reml, correct = TRUE)
c(naive = s_naive@df, corrected = s_corr@df)
c(naive = s_naive@aic, corrected = s_corr@aic)
```

The correction adds `r fmt(s_corr@df - s_naive@df, 2)` to the effective
number of parameters. We measure what it does to a choice between two models
with a simulation. We draw 100 data sets with 20 groups of 5 observations
from $y_{ij} = 1 + 0.5 x_{ij} + \varepsilon_{ij}$, which has no group
effect, fit the model with and without a random intercept, and count how
often each version of the cAIC is smaller for the model with the random
intercept. We repeat the simulation with a group effect of standard
deviation 0.3.

```{r}
#| label: sim-greven
caic_of <- function(fit, correct) summary(fit, correct = correct)@aic
pick <- function(sd_group, seed) {
  set.seed(seed)
  t(replicate(100, {
    d <- data.frame(g = factor(rep(1:20, each = 5)), x = rnorm(100))
    d$y <- 1 + 0.5 * d$x + rnorm(20, 0, sd_group)[d$g] + rnorm(100)
    a <- statmod(y ~ x, distrib = gaussian1_distrib(), data = d)
    b <- statmod(y ~ x + random(~ 1 | g), distrib = gaussian1_distrib(),
                 data = d)
    c(naive = caic_of(b, FALSE) < caic_of(a, FALSE),
      corrected = caic_of(b, TRUE) < caic_of(a, TRUE))
  }))
}
none <- colMeans(pick(0, 11))
some <- colMeans(pick(0.3, 12))
rbind(`no group effect` = none, `group sd 0.3` = some)
```

Without a group effect, the naive cAIC is smaller for the model with the
random intercept in `r pct(none[["naive"]], 0)` of the data sets, and the
corrected cAIC in `r pct(none[["corrected"]], 0)`. The correction has a
cost: with a group effect the naive cAIC is smaller for the right model in
`r pct(some[["naive"]], 0)` of the data sets and the corrected one in
`r pct(some[["corrected"]], 0)`.

### A lasso chosen by BIC and by cross-validation

The flats of `rent` have three binary attributes: a bathroom of their own
(`B`), central heating (`H`) and a full kitchen (`L`). We put these and the
location under a lasso, beside the two smooth terms, and choose the
$\lambda$ of the lasso by BIC and by ten-fold cross-validation. The
smoothing parameters are estimated by REML at each point of the sequence of
$\lambda$, as @sec-penalized-hyper describes.

```{r}
#| label: fit-rent-lasso
rent_bic <- statmod(R ~ s(Fl) + s(A) + lasso(~ B + H + L + loc),
                    distrib = gamma1_distrib(), data = rent)
set.seed(4)
rent_cv <- statmod(R ~ s(Fl) + s(A) + lasso(~ B + H + L + loc),
                   distrib = gamma1_distrib(), data = rent,
                   sparse_criterion = cv())
hyper(rent_bic)
hyper(rent_cv)
```

```{r}
#| label: rent-lasso-coef
lasso_coef <- function(fit) {
  cf <- coef(fit)$mu
  cf[startsWith(names(cf), "lasso.")]
}
rbind(bic = lasso_coef(rent_bic), cv = lasso_coef(rent_cv))
```

```{r}
#| label: rent-lasso-numbers
#| include: false
lam_bic <- hyper(rent_bic)$estimate[3]
lam_cv <- hyper(rent_cv)$estimate[3]
```

BIC chooses $\lambda = `r fmt(lam_bic, 3)`$ and cross-validation
$\lambda = `r fmt(lam_cv, 3)`$. The two values differ by a factor of
`r fmt(lam_cv / lam_bic, 0)`, and the coefficients hardly differ: with
`r nrow(rent)` flats the penalty of either value is small against the
log-likelihood, and the lasso sets none of the attributes to zero. The
column `B1` has no estimate. The lasso codes all the levels of the first
factor of its formula, the two columns of `B` add up to the intercept, and
the fit reports the second column as not identified. @sec-penalized-choose
compares the three criteria of a lasso on the crime data. Cross-validation refits the model on each fold, so it
costs about ten times as much as BIC.

### The Student t and its information

The data frame `film90` gives, for `r nrow(film90)` American films of
1990, the logarithm of the box office revenue `lborev1` against the
logarithm of the number of screens in the opening week `lboopen`
[@stasinopoulos2017]. The revenue has outliers in both directions, and we
compare a gaussian and a Student $t$ distribution, each with smooth terms
in the location and in the scale:

```{r}
#| label: fit-film
f_film <- lborev1 ~ s(lboopen) | sigma ~ s(lboopen)
film_g <- statmod(f_film, distrib = gaussian1_distrib(), data = film90)
film_t <- statmod(f_film, distrib = student_t1_distrib(), data = film90)
summary(film_t)
c(gaussian = caic(film_g), t = caic(film_t))
```

The $t$ distribution has $\hat\nu = `r fmt(exp(coef(film_t)$nu[[1]]), 1)`$
degrees of freedom and a cAIC lower by
`r fmt(caic(film_g) - caic(film_t), 0)`. The REML criterion of these fits
uses the observed information in its determinant. The argument `hessian`
of `reml()` and `ml()` replaces it with the expected information, which for
a $t$ distribution is positive definite at every point, while the observed
information of the location is not positive at an observation far in the
tails:

```{r}
#| label: fit-film-expected
film_te <- statmod(f_film, distrib = student_t1_distrib(), data = film90,
                   outer_criterion = reml(hessian = "expected"))
c(observed = exp(coef(film_t)$nu[[1]]), expected = exp(coef(film_te)$nu[[1]]))
c(observed = as.numeric(logLik(film_t)), expected = as.numeric(logLik(film_te)))
```

The two informations give the same fit here. The two criteria are not the
same function, so their values differ, and we compare the fits by their
estimates and their log-likelihoods. @fig-film shows the fitted median and
the 5% and 95% quantiles of the $t$ model against those of the gaussian
model.

```{r}
#| label: fig-film
#| fig-cap: "Logarithm of the box office revenue against the logarithm of the opening screens. The lines are the fitted median and the 5% and 95% quantiles of the Student t model and of the gaussian model."
#| fig-height: 4
og <- data.frame(lboopen = seq(from = min(film90$lboopen),
                               to = max(film90$lboopen), length.out = 200))
quant <- function(fit, dist, label) {
  th <- list(mu = predict(fit, what = "mu", newdata = og),
             sigma = predict(fit, what = "sigma", newdata = og))
  if (label == "t") th$nu <- predict(fit, what = "nu", newdata = og)
  rbind(data.frame(lboopen = og$lboopen, quantile = "5%", model = label,
                   value = distrib_quantile(dist, p = 0.05, theta = th)),
        data.frame(lboopen = og$lboopen, quantile = "50%", model = label,
                   value = distrib_quantile(dist, p = 0.5, theta = th)),
        data.frame(lboopen = og$lboopen, quantile = "95%", model = label,
                   value = distrib_quantile(dist, p = 0.95, theta = th)))
}
qf <- rbind(quant(film_t, student_t1_distrib(), "t"),
            quant(film_g, gaussian1_distrib(), "gaussian"))
qf$group <- paste(qf$model, qf$quantile)
ggplot(film90, aes(x = lboopen, y = lborev1)) +
  geom_point(colour = book_grey, size = 0.4, alpha = 0.4) +
  geom_line(data = qf, aes(y = value, colour = model, linetype = model,
                           group = group)) +
  labs(x = "log opening screens", y = "log box office revenue")
```

## When a fit does not converge {#sec-choosing-convergence}

An optimizer reports whether its own stopping rule was met. This is a
statement about the search, and it does not tell us whether the returned
point is the maximum that the model defines. The function
`statmod_certificate()` of @sec-model-certificate checks this from two
readings at the returned point. The first is about the coefficients: with
$g$ the gradient of the penalized log-likelihood and $K$ its curvature, the
quantity $\tfrac{1}{2} g^\top K^{-1} g$ is the increase that one Newton step
would still give, in log-likelihood units. The second is the same quantity
for the criterion of the hyperparameters. The certificate returns one of
four states: `converged` when the second reading is below its tolerance,
`boundary` when it is below the tolerance with at least one hyperparameter
at the edge of its range, `not converged` when the criterion would still
increase, and `unknown` when there is no criterion with derivatives to
check.

### The four states

We fit five models that end in the four states. The first is the
motorcycle model of @sec-model with a smooth standard deviation. In the
second the response is pure noise. The third and the fourth have a lasso,
on the ships of @sec-glm-counts and on the crime data of
@sec-penalized-choose. The fifth is the fourth with the intercept of
$\log\sigma$ estimated by the REML criterion.

```{r}
#| label: fit-states
fit1 <- statmod(accel ~ s(times, bspline_smooth(k = 20)) |
                  sigma ~ s(times, bspline_smooth(k = 10)),
                distrib = gaussian1_distrib(), data = mcycle)
set.seed(3)
noise <- data.frame(x = runif(200), y = rnorm(200))
fit_n <- statmod(y ~ s(x), distrib = gaussian1_distrib(), data = noise)
data(ships, package = "MASS")
ships <- ships[ships$service > 0, ]
ships$year <- factor(ships$year)
ships$period <- factor(ships$period)
fit_s <- statmod(incidents ~ offset(log(service)) + lasso(~ type + year + period),
                 distrib = poisson_distrib(), data = ships)
data(UScrime, package = "MASS")
UScrime$lrate <- log(UScrime$y)
f_crime <- lrate ~ lasso(~ M + So + Ed + Po1 + Po2 + LF + M.F + Pop + NW + U1 +
                           U2 + GDP + Ineq + Prob + Time, standardize = TRUE)
fit_l <- statmod(f_crime, distrib = gaussian1_distrib(), data = UScrime)
fit_l2 <- statmod(f_crime, distrib = gaussian1_distrib(), data = UScrime,
                  outer_criterion = reml(marginal = "sigma"))
statmod_certificate(fit1)$state
statmod_certificate(fit_n)$state
statmod_certificate(fit_s)$state
statmod_certificate(fit_l)$state
statmod_certificate(fit_l2)$state
```

```{r}
#| label: states-numbers
#| include: false
cert1 <- statmod_certificate(fit1)
lam_n <- hyper(fit_n)$estimate[1]
```

The motorcycle model is `converged`: the increase that remains is
$`r sci(cert1$decrement)`$ for the REML criterion and
$`r sci(cert1$mode_error)`$ for the penalized log-likelihood. With a
response of pure noise the best smooth function is a straight line, REML
takes the smoothing parameter to $`r sci(lam_n)`$, where the criterion is
flat, and the state is `boundary`. The two models with a lasso have one
hyperparameter with a kink, chosen along a sequence of values, so there is
no criterion with derivatives and the state is `unknown`. When we name
`sigma` in the argument `marginal` of `reml()`, the criterion estimates the
intercept of $\log\sigma$ at the value of $\lambda$ that the sequence
chose, the criterion has a gradient in that estimate, and the state is
`converged`.

### A kink in the log-density

The Laplace distribution has the log-density
$-\log(2\sigma) - \lvert y - \mu\rvert/\sigma$, which has no derivative in
$\mu$ at $\mu = y$, and its maximum likelihood estimate of $\mu$ is the
least absolute deviations fit. We fit it to the `stackloss` data of
@brownlee1965 and compare the result with the exact fit of the `quantreg`
package [@koenker2005]:

```{r}
#| label: fit-lad
lad <- statmod(stack.loss ~ Air.Flow + Water.Temp + Acid.Conc.,
               distrib = laplace_distrib(), data = stackloss)
rq_fit <- quantreg::rq(stack.loss ~ Air.Flow + Water.Temp + Acid.Conc.,
                       data = stackloss)
rbind(statmodels7 = coef(lad)$mu, quantreg = coef(rq_fit))
statmod_certificate(lad)$state
```

```{r}
#| label: lad-numbers
#| include: false
gap_lad <- max(abs(coef(lad)$mu - coef(rq_fit)))
obj_lad <- sum(abs(stackloss$stack.loss - predict(lad, what = "mu", newdata = stackloss)))
obj_rq <- sum(abs(stats::resid(rq_fit)))
```

The two sets of coefficients differ by at most `r fmt(gap_lad, 3)`, and the
sums of absolute residuals are `r fmt(obj_lad, 4)` for `statmodels7` and
`r fmt(obj_rq, 4)` for the exact fit. The inner fit stops when the gradient
of the log-likelihood is small, and at a kink the gradient does not become
small, so the fit stops close to the minimum but not at it. It reports
`converged = `r lad@converged``, and the certificate cannot read a
curvature at the kink, so its state is `unknown`. For such a distribution
the coefficients of `quantreg` are the reference.

### A search that needs other arguments

The Student $t$ model of @sec-distribs-heavy with an equation for the
degrees of freedom is a harder problem:

```{r}
#| label: gag-formula
f_nu <- GAG ~ Age | sigma ~ Age | nu ~ Age
```

With the default arguments the outer search does not converge on these
data: it runs for about a minute and a half and stops at a point where the
gradient of the criterion cannot be computed. We show the same kind of
failure in a fraction of the time by giving Newton's method, which the
search uses with the observed information, a budget of three iterations:

```{r}
#| label: fit-gag-budget
gag_b <- statmod(f_nu, distrib = student_t1_distrib(), data = GAGurine,
                 outer_optimizer = optimizers7::newton(maxit = 3))
cert_b <- statmod_certificate(gag_b)
cert_b$state
cert_b$reason
```

The certificate says that the criterion is still rising. The first thing to
change is the information in the criterion. The expected information of the
$t$ distribution is positive definite everywhere:

```{r}
#| label: fit-gag-expected
gag_e <- statmod(f_nu, distrib = student_t1_distrib(), data = GAGurine,
                 outer_criterion = reml(hessian = "expected"))
statmod_certificate(gag_e)$state
c(criterion = gag_e@criterion, logLik = as.numeric(logLik(gag_e)))
coef(gag_e)$nu
```

This fit is certified, and the degrees of freedom fall with the age. A
certified point is a maximum of the criterion, but not necessarily the
highest one. We check this by starting the search from another point: the
fit with the hyperparameters held at their starting values
(`outer_criterion = NULL`), which estimates every coefficient at the joint
mode, gives a start through `start_from()`:

```{r}
#| label: fit-gag-joint
gag_0 <- statmod(f_nu, distrib = student_t1_distrib(), data = GAGurine,
                 outer_criterion = NULL)
gag_je <- statmod(f_nu, distrib = student_t1_distrib(), data = GAGurine,
                  outer_criterion = reml(hessian = "expected"),
                  start = start_from(gag_0))
gag_jo <- statmod(f_nu, distrib = student_t1_distrib(), data = GAGurine,
                  start = start_from(gag_0))
statmod_certificate(gag_je)$state
statmod_certificate(gag_jo)$state
c(expected = gag_je@criterion, observed = gag_jo@criterion)
coef(gag_je)$nu
```

```{r}
#| label: gag-numbers
#| include: false
gap_gag <- gag_je@criterion - gag_e@criterion
```

From this start the search with the expected information reaches a
criterion higher by `r fmt(gap_gag, 2)` than the first fit, and the search
with the observed information converges as well, with the default
arguments. The two maxima describe different models (@fig-gag-nu): at the
first the degrees of freedom fall from about
`r fmt(exp(sum(coef(gag_e)$nu * c(1, 0))), 0)` at birth, at the second they
rise with the age from below one. The criterion prefers the second. The
general lesson is that a certificate is local: when a model is hard, we fit
it from more than one start, compare the criteria of fits that use the same
information, and keep the highest.

```{r}
#| label: fig-gag-nu
#| fig-cap: "Degrees of freedom of the Student t against age, at the two maxima of the REML criterion with the expected information."
#| fig-height: 3.6
ag <- data.frame(Age = seq(from = 0, to = 17, length.out = 100))
nu_df <- rbind(
  data.frame(Age = ag$Age, nu = predict(gag_e, what = "nu", newdata = ag),
             maximum = "default start"),
  data.frame(Age = ag$Age, nu = predict(gag_je, what = "nu", newdata = ag),
             maximum = "start from the joint mode"))
ggplot(nu_df, aes(x = Age, y = nu, colour = maximum, linetype = maximum)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "age (years)", y = expression(nu))
```

## What statmodels7 does not do {#sec-choosing-limits}

Some models cannot be fitted with `statmodels7` at present, and for each of
them `statmod()` stops with a message that names the cause.

A response with several components, such as a pair of measurements on the
same subject, needs a multivariate distribution. `distributions7` has four
multivariate families, and `distributions7::fit_distrib()` fits them with
constant parameters, but `statmod()` does not let their parameters depend
on covariates:

```{r}
#| label: limit-mv
#| error: true
set.seed(3)
d_mv <- data.frame(x = runif(50))
d_mv$Y <- cbind(rnorm(50), rnorm(50))
statmod(Y ~ x, distrib = mvgaussian1_distrib(2), data = d_mv)
```

A censored response, in which some values are known only to lie above or
below a limit, is marked by `cens()`, but `statmod()` has no censored
likelihood:

```{r}
#| label: limit-cens
#| error: true
statmod(cens(accel, accel < -100) ~ times, distrib = gaussian1_distrib(),
        data = mcycle)
```

A model can contain at most one structural term, that is one `gas()`, one
`regime()` or one marginal break-point term. Two score-driven terms, one in
the mean and one in the scale, would form one joint recursion, which
`statmodels7` does not implement:

```{r}
#| label: limit-structural
#| error: true
data(EuStockMarkets)
dax <- data.frame(r = 100 * diff(log(EuStockMarkets[, "DAX"])))
statmod(r ~ gas(p = 1, q = 1) | sigma ~ gas(p = 1, q = 1),
        distrib = gaussian1_distrib(), data = dax)
```

The hidden Markov model of `regime()` has one more restriction: its
transition probabilities are constant, so covariates can shift the mean of
every state, as in `waiting ~ duration + regime(k = 2)`, but they cannot
change the probability of moving from one state to another.

A distribution with a kink in a parameter, such as the Laplace distribution
above, cannot be combined with a smooth term or a random effect in that
parameter, because the REML criterion reads the second derivative of the
log-density:

```{r}
#| label: limit-laplace
#| error: true
statmod(accel ~ s(times), distrib = laplace_distrib(), data = mcycle)
```

A discontinuous break-point term cannot develop its break-point with a
random effect, because its coefficients are the development multiplied by
the change of level, and the penalty would act on that product. The
smoothed constructions of @sec-breakpoints-smooth do not have this
restriction:

```{r}
#| label: limit-jump
#| error: true
set.seed(5)
d_j <- data.frame(id = factor(rep(1:12, each = 25)),
                  x = rep(seq(from = 0, to = 10, length.out = 25), 12))
d_j$y <- 1 + 2 * (d_j$x > 5) + rnorm(300, sd = 0.3)
statmod(y ~ jump(x, psi ~ random(~ 1 | id)), distrib = gaussian1_distrib(),
        data = d_j)
```

The wrapper `fixed()` holds a parameter at one value, and a value for each
observation is not accepted:

```{r}
#| label: limit-fixed
#| error: true
fixed(gaussian1_distrib(), sigma = runif(10) + 1)
```

A term `nl()` written as an R function cannot give its parameters equations
of their own, because only a formula says where a parameter enters the
model:

```{r}
#| label: limit-nl
#| error: true
statmod(rate ~ 0 + nl(function(conc, Vm, K) Vm * conc / (K + conc), K ~ state,
                      params = c("Vm", "K"), start = list(Vm = 200, K = 0.1)),
        distrib = gaussian1_distrib(), data = Puromycin)
```

A moment of the distribution, such as the mean of a gamma response with a
dispersion, has a prediction but no standard error:

```{r}
#| label: limit-moment
#| error: true
gag_gamma <- statmod(GAG ~ Age, distrib = gamma1_distrib(), data = GAGurine)
predict(gag_gamma, what = "mean", newdata = data.frame(Age = 5), se = TRUE)
```

Cross-validation chooses only the hyperparameter of a penalty with a kink.
For a smooth term the criteria of the first part of
@sec-choosing-hyper estimate the smoothing parameter without refitting:

```{r}
#| label: limit-cv
#| error: true
statmod(accel ~ s(times), distrib = gaussian1_distrib(), data = mcycle,
        outer_criterion = cv())
```

Finally, the intervals of `confint()` and `predict()` use the gaussian
limit of the estimators, with the quantile 1.96 for a 95% interval. With a
few observations these intervals are too narrow. We measure this for the
half-saturation concentration $K$ of the Michaelis–Menten model of
@sec-beyond-tests, with `r nrow(Puromycin[Puromycin$state == "treated", ])`
observations. We draw 200 data sets from the fitted model and count how
often the 95% Wald interval of `statmodels7` and the interval of
`confint()` for `nls()`, which uses the $t$ distribution, contain the true
$K$.

```{r}
#| label: sim-puro
puro <- Puromycin[Puromycin$state == "treated", ]
fit_p <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc)),
                 distrib = gaussian1_distrib(), data = puro)
th_p <- coef(fit_p)$mu
s_p <- exp(coef(fit_p)$sigma[[1]])
set.seed(13)
cover_p <- t(replicate(200, {
  d <- puro
  d$rate <- th_p[["nl.Vm"]] * d$conc / (th_p[["nl.K"]] + d$conc) +
    rnorm(nrow(d), 0, s_p)
  f <- statmod(rate ~ 0 + nl(~ Vm * conc / (K + conc)),
               distrib = gaussian1_distrib(), data = d)
  ci <- confint(f, readable = FALSE)
  k <- ci[ci$coefficient == "nl.K", ]
  n <- suppressMessages(confint(nls(rate ~ Vm * conc / (K + conc), data = d,
                                    start = list(Vm = 200, K = 0.05))))["K", ]
  c(wald = k$lower <= th_p[["nl.K"]] && th_p[["nl.K"]] <= k$upper,
    nls = n[[1]] <= th_p[["nl.K"]] && th_p[["nl.K"]] <= n[[2]])
}))
colMeans(cover_p)
```

```{r}
#| label: puro-numbers
#| include: false
cv_p <- colMeans(cover_p)
```

The Wald intervals contain the true $K$ in `r pct(cv_p[["wald"]])` of the
data sets and the intervals of `nls()` in `r pct(cv_p[["nls"]])`, against a
nominal 95%. With 200 data sets the standard error of each proportion is
about 1.5%. For small samples we can use the parametric bootstrap of
@sec-breakpoints-bootstrap instead of the gaussian limit.

```{r}
#| label: gate-14
#| include: false
assert_choosing_ok(
  fits = list(sp_pois = sp_pois, sp_nb2 = sp_nb2, sp_nb1 = sp_nb1, sp_pig = sp_pig,
              sp_nb2_reml = sp_nb2_reml, sp_nb2_th = sp_nb2_th,
              rent_gamma = rent_gamma, rent_lnorm = rent_lnorm, rent_ig = rent_ig,
              rent_gg = rent_gg, rent_phi = rent_phi,
              ab_reml = ab_reml, ab_ml = ab_ml, ab_aic = ab_aic, ab_bic = ab_bic,
              rent_bic = rent_bic, rent_cv = rent_cv,
              film_g = film_g, film_t = film_t, film_te = film_te,
              fit1 = fit1, fit_n = fit_n, fit_s = fit_s, fit_l = fit_l,
              fit_l2 = fit_l2, lad = lad, gag_b = gag_b, gag_e = gag_e,
              gag_je = gag_je, gag_jo = gag_jo, fit_p = fit_p),
  ref = list(gl_nbi = gl_nbi, gl_pig = gl_pig, g_rent = g_rent, g_ab = g_ab,
             rq_fit = rq_fit),
  data = list(species = species, rent = rent, stackloss = stackloss),
  claims = list(sp_table = sp_table, rent_table = rent_table, ab_table = ab_table,
                s_naive = s_naive, s_corr = s_corr, none = none, some = some,
                lam_bic = lam_bic, lam_cv = lam_cv, qr_pois = qr_pois,
                qr_nb2 = qr_nb2, gap_lad = gap_lad, obj_lad = obj_lad,
                obj_rq = obj_rq, cert_b = cert_b, gap_gag = gap_gag,
                cv_p = cv_p))
```
