# Writing a distribution and a link {#sec-own}

```{r}
#| label: setup-15
#| include: false
source("R/_setup.R")
source("R/own-certificates.R")
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
sci <- function(x, d = 1) {
  e <- floor(log10(abs(x)))
  sprintf("%.*f \\times 10^{%d}", d, x / 10^e, e)
}
data(rent, package = "gamlss.data")
data(menarche, package = "MASS")
elapsed <- function(expr) {
  t0 <- proc.time()[[3]]
  force(expr)
  proc.time()[[3]] - t0
}
```

The distributions and the links of the previous chapters are objects of
`distributions7` and `linkfunctions7`. A distribution or a link that these
packages do not provide is written by the user as an object of the same
kind, and `statmod()` then uses it as it uses the built-in ones: every
parameter can have an equation with smooth terms, random effects and
penalties. @sec-distribs-own and @sec-glm-links showed the shortest case, a
density and a pair of functions. This chapter shows the complete case. We
write a discrete distribution and a continuous one, first with the density
alone and then with the cdf, the quantile function and the score in closed
form, and we write a link whose shape is set by a parameter.

## The structure of a distribution {#sec-own-structure}

A distribution is an S7 class that inherits from `continuous_distrib` or
from `discrete_distrib`, with a method for `distrib_pdf()`. The method
receives the response `y`, a list `theta` with one element for each
parameter, and the argument `log`; it returns the density or its logarithm
at each observation. A constructor then creates the object with the names
of the parameters, their bounds and their links.

Every other function of a distribution has a numerical fallback, which
`distributions7` uses when the class has no method of its own. The cdf of a
continuous distribution is a quadrature of the density, and the cdf of a
discrete one is a sum over the support. The quantile function inverts the
cdf, by root finding for a continuous distribution and by a search in the
table of cumulative probabilities for a discrete one. The random numbers
of a discrete distribution are quantiles of uniform numbers, and those of a
continuous distribution come from a ratio-of-uniforms generator. The mean and
the variance are integrals or sums of the density. The score and the
Hessian of the log-likelihood in the parameters are finite differences of
the log-density, and the derivatives of order three and four are computed
by one tensor stencil applied to the highest order that the class
implements itself: the log-density if the class has only the density, the
score if it has the score. A difference is never taken of another
difference in the same parameter, because the error of a nested difference
grows with each level.

A method written by the user replaces the fallback for that function and,
through it, every quantity computed from it. A method must accept `...`,
because `distributions7` passes arguments such as the number of threads
through it. A method for a derivative in the parameters must also accept
`scale = c("parameter", "link")`: it returns the derivative in the
parameters, and the generic converts it to the scale of the link. The
function `statmod()` checks these signatures before the fit and signals an
error that names the method.

## A discrete distribution: the generalized Poisson {#sec-own-genpois}

The generalized Poisson distribution of @consul1973 has the probability
function
$$
P(Y = y) = \frac{\theta\,(\theta + \lambda y)^{y - 1}\, e^{-\theta - \lambda y}}{y!},
\qquad y = 0, 1, 2, \ldots,
$$
with $\theta > 0$ and $0 \le \lambda < 1$. Its mean is $\theta/(1 - \lambda)$
and its variance is $\theta/(1 - \lambda)^3$. For regression we use the
parametrization of @consul1992 in the mean $\mu$ and a dispersion $\phi$,
with
$$
\theta = \frac{\mu}{\sqrt{\phi}}, \qquad \lambda = 1 - \frac{1}{\sqrt{\phi}},
$$
so that the mean is $\mu$ and the variance is $\phi\,\mu$. The variance
is proportional to the mean, as in the quasi-Poisson model, and
$\phi = 1$ gives the Poisson distribution. The parametrization requires
$\phi \ge 1$, the overdispersed case; we write the density for every
$\phi > 0$ and set it to zero where $\theta + \lambda y \le 0$. The class,
its density and its constructor are:

```{r}
#| label: genpois-class
GenPois <- S7::new_class("GenPois", parent = discrete_distrib)
S7::method(distrib_pdf, GenPois) <- function(distrib, y, theta, log = FALSE,
                                             ...) {
  mu <- theta[[1]]
  phi <- theta[[2]]
  s <- sqrt(phi)
  th <- mu / s
  la <- 1 - 1 / s
  a <- th + la * y
  ld <- ifelse(a > 0, log(th) + (y - 1) * log(a) - a - lgamma(y + 1), -Inf)
  if (log) ld else exp(ld)
}
genpois_distrib <- function(link_mu = log_link(), link_phi = log_link()) {
  GenPois(distrib_name = "generalized Poisson", dimension = "univariate",
          bounds = c(0, Inf), params = c("mu", "phi"),
          params_interpretation = c(mu = "mean", phi = "dispersion"),
          n_params = 2,
          params_bounds = list(mu = c(0, Inf), phi = c(0, Inf)),
          link_params = list(mu = link_mu, phi = link_phi))
}
gp <- genpois_distrib()
c(mean = mean(gp, theta = list(mu = 4, phi = 2.5)),
  variance = variance(gp, theta = list(mu = 4, phi = 2.5)))
```

The mean and the variance come from sums over the support, and they are
$\mu = 4$ and $\phi\mu = 10$. The function `check_distrib()` tests the
fallbacks against each other at one value of the parameters:

```{r}
#| label: genpois-check
chk_gp <- check_distrib(gp, theta = list(mu = 4, phi = 2.5))
```

The `r nrow(chk_gp)` checks pass, and the checks of the derivatives in the
parameters are not run, for the reason given in @sec-distribs-own.

### The model for the warp breaks

The data frame `warpbreaks` gives the number of breaks of the warp on a loom
for 54 pieces of yarn, of two wools (`A` and `B`) and three tensions (`L`,
`M` and `H`). We fit the generalized Poisson model with the interaction of
the two factors for the mean and the wool for the dispersion, and we
compare it with `glmmTMB()` [@brooks2017], whose family `genpois()` is the
same distribution in the same parametrization. The criterion is `ml()`,
which estimates the coefficients of $\phi$ by maximum likelihood, as
`glmmTMB()` does:

```{r}
#| label: fit-genpois
t_gp_density <- elapsed(
  fit_gp <- statmod(breaks ~ wool * tension | phi ~ wool, distrib = gp,
                    data = warpbreaks, outer_criterion = ml()))
summary(fit_gp)
tmb_gp <- glmmTMB::glmmTMB(breaks ~ wool * tension, dispformula = ~ wool,
                           family = glmmTMB::genpois(), data = warpbreaks)
rbind(statmod = c(coef(fit_gp)$mu, coef(fit_gp)$phi),
      glmmTMB = c(glmmTMB::fixef(tmb_gp)$cond, glmmTMB::fixef(tmb_gp)$disp))
c(statmod = as.numeric(logLik(fit_gp)), glmmTMB = as.numeric(logLik(tmb_gp)))
```

The two fits give the same coefficients and the same log-likelihood. The
dispersion of wool `A` is $\exp(`r fmt(coef(fit_gp)$phi[[1]], 3)`) =
`r fmt(exp(coef(fit_gp)$phi[[1]]), 2)`$, so its variance is about
`r fmt(exp(coef(fit_gp)$phi[[1]]), 1)` times its mean, and the dispersion
of wool `B` is smaller. @fig-genpois shows the fitted means with an
interval of two standard deviations, $\mu \pm 2\sqrt{\phi\mu}$.

```{r}
#| label: fig-genpois
#| fig-cap: "Number of warp breaks for each wool and tension. The lines join the fitted means of the generalized Poisson model, and the bars span two standard deviations on either side."
#| echo: false
#| fig-height: 3.8
cells <- unique(warpbreaks[, c("wool", "tension")])
cells$mu <- predict(fit_gp, what = "mu", newdata = cells)
cells$phi <- predict(fit_gp, what = "phi", newdata = cells)
op <- par(mar = c(4, 4, 1, 1))
xpos <- as.integer(warpbreaks$tension) + ifelse(warpbreaks$wool == "A", -0.12, 0.12)
plot(xpos, warpbreaks$breaks, pch = ifelse(warpbreaks$wool == "A", 1, 2),
     col = "grey45", xaxt = "n", xlab = "tension", ylab = "breaks",
     xlim = c(0.6, 3.4), ylim = c(0, max(warpbreaks$breaks)))
axis(1, at = 1:3, labels = levels(warpbreaks$tension))
for (w in c("A", "B")) {
  cw <- cells[cells$wool == w, ]
  cw <- cw[order(cw$tension), ]
  xx <- as.integer(cw$tension) + if (w == "A") -0.12 else 0.12
  sdv <- sqrt(cw$phi * cw$mu)
  segments(xx, cw$mu - 2 * sdv, xx, cw$mu + 2 * sdv, lwd = 1.5)
  lines(xx, cw$mu, lwd = 2, lty = if (w == "A") 1 else 2)
}
legend("topright", legend = c("wool A", "wool B"), pch = 1:2,
       lty = 1:2, bty = "n")
par(op)
```

The negative binomial distribution of @sec-glm-overdispersion has a
variance that grows with the square of the mean. The AIC compares the two
assumptions on these data:

```{r}
#| label: genpois-negbin
fit_gp_nb <- statmod(breaks ~ wool * tension | theta ~ wool,
                     distrib = negbin2_distrib(), data = warpbreaks,
                     outer_criterion = ml())
c(generalized_poisson = AIC(fit_gp), negative_binomial = AIC(fit_gp_nb))
```

The difference is `r fmt(abs(AIC(fit_gp) - AIC(fit_gp_nb)), 1)`, too
small to prefer one variance function to the other with 54 observations.

### The score in closed form

The fit above computes the score by finite differences of the
log-density. The score in closed form is short. With
$a = \theta + \lambda y$, the log-probability is
$\log\theta + (y - 1)\log a - a - \log y!$, and its derivatives in
$\theta$ and $\lambda$ are
$$
\frac{\partial \ell}{\partial \theta} = \frac{1}{\theta} + \frac{y - 1}{a} - 1,
\qquad
\frac{\partial \ell}{\partial \lambda} = \frac{y\,(y - 1)}{a} - y .
$$
The chain rule through $\theta = \mu/\sqrt\phi$ and
$\lambda = 1 - 1/\sqrt\phi$ gives the score in $(\mu, \phi)$:
$$
\frac{\partial \ell}{\partial \mu} = \frac{1}{\sqrt\phi}\,\frac{\partial \ell}{\partial \theta},
\qquad
\frac{\partial \ell}{\partial \phi} = \frac{1}{2\phi\sqrt\phi}
\left(\frac{\partial \ell}{\partial \lambda} - \mu\,\frac{\partial \ell}{\partial \theta}\right).
$$
We register it as the method of `distrib_gradient()`. The method returns a
list with one vector for each parameter, on the scale of the parameters:

```{r}
#| label: genpois-score
S7::method(distrib_gradient, GenPois) <- function(distrib, y, theta,
                                                  scale = c("parameter", "link"),
                                                  ...) {
  mu <- theta[[1]]
  phi <- theta[[2]]
  s <- sqrt(phi)
  th <- mu / s
  la <- 1 - 1 / s
  a <- th + la * y
  d_th <- 1 / th + (y - 1) / a - 1
  d_la <- y * (y - 1) / a - y
  list(mu = d_th / s, phi = (d_la - mu * d_th) / (2 * phi * s))
}
score_gap <- max(abs(
  unlist(distrib_gradient(gp, y = 0:12, theta = list(mu = 4, phi = 2.5))) -
  unlist(numerical_gradient(gp, y = 0:12, theta = list(mu = 4, phi = 2.5)))))
score_gap
```

The closed form and the finite differences agree to
$`r sci(score_gap, 0)`$. With the score registered, the Hessian
is one difference of the score, and the derivatives of order three and four
are one stencil on the score, which is more accurate than a stencil on the
log-density. The same model, fitted again, reaches the same point in less
time:

```{r}
#| label: fit-genpois-score
t_gp_score <- elapsed(
  fit_gp_s <- statmod(breaks ~ wool * tension | phi ~ wool, distrib = gp,
                      data = warpbreaks, outer_criterion = ml()))
c(logLik_density = as.numeric(logLik(fit_gp)),
  logLik_score = as.numeric(logLik(fit_gp_s)),
  seconds_density = t_gp_density, seconds_score = t_gp_score)
```

### A random effect

A distribution written by the user takes random effects as a built-in one
does. We treat the three tensions as a sample of groups, with a random
intercept in the equation of $\mu$, and compare the fit with `glmmTMB()`:

```{r}
#| label: fit-genpois-random
fit_gp_r <- statmod(breaks ~ wool + random(~ 1 | tension) | phi ~ wool,
                    distrib = gp, data = warpbreaks, outer_criterion = ml())
summary(fit_gp_r)
fit_gp_ra <- statmod(breaks ~ wool + random(~ 1 | tension) | phi ~ wool,
                     distrib = gp, data = warpbreaks,
                     outer_criterion = ml(marginal = "all"))
tmb_gp_r <- glmmTMB::glmmTMB(breaks ~ wool + (1 | tension),
                             dispformula = ~ wool,
                             family = glmmTMB::genpois(), data = warpbreaks)
data.frame(
  fit = c("ml()", "ml(marginal = \"all\")", "glmmTMB"),
  intercept = c(coef(fit_gp_r)$mu[[1]], coef(fit_gp_ra)$mu[[1]],
                glmmTMB::fixef(tmb_gp_r)$cond[[1]]),
  sd = c(hyper(fit_gp_r)$estimate, hyper(fit_gp_ra)$estimate,
         attr(glmmTMB::VarCorr(tmb_gp_r)$cond$tension, "stddev")),
  logLik = c(logLik(fit_gp_r, type = "marginal"),
             logLik(fit_gp_ra, type = "marginal"), logLik(tmb_gp_r)))
```

The criterion `ml()` holds the fixed effects of $\mu$ at the joint mode of
the fixed and the random effects, and `glmmTMB()` chooses them to maximize
the Laplace approximation of the likelihood. For a count response the two
differ slightly, as @sec-glmm-poisson explains, and `ml(marginal = "all")`
reproduces `glmmTMB()`.

## A continuous distribution: the Burr XII {#sec-own-burr}

The Burr XII distribution [@burr1942] with scale $\mu$ and shapes $c$ and $k$
has the density
$$
f(y) = \frac{c\,k}{\mu}\left(\frac{y}{\mu}\right)^{c - 1}
\left\{1 + \left(\frac{y}{\mu}\right)^{c}\right\}^{-k - 1}, \qquad y > 0 .
$$
The shape $c$ governs the lower tail and the product $ck$ the upper tail,
which decays as $y^{-ck - 1}$; the log-logistic distribution of
@sec-distribs-own is the case $k = 1$. It is also the generalized beta
distribution of the second kind of @mcdonald1984 with its third parameter
equal to one, which is how `gamlss` provides it (the family `GB2()` with
$\nu = 1$). We write the class with the density alone:

```{r}
#| label: burr-class
Burr <- S7::new_class("Burr", parent = continuous_distrib)
S7::method(distrib_pdf, Burr) <- function(distrib, y, theta, log = FALSE, ...) {
  mu <- theta[[1]]
  cc <- theta[[2]]
  k <- theta[[3]]
  z <- y / mu
  ld <- log(cc) + log(k) - log(mu) + (cc - 1) * log(z) - (k + 1) * log1p(z^cc)
  if (log) ld else exp(ld)
}
burr_distrib <- function(link_mu = log_link(), link_c = log_link(),
                         link_k = log_link()) {
  Burr(distrib_name = "Burr XII", dimension = "univariate",
       bounds = c(0, Inf), params = c("mu", "c", "k"),
       params_interpretation = c(mu = "scale", c = "shape", k = "shape"),
       n_params = 3,
       params_bounds = list(mu = c(0, Inf), c = c(0, Inf), k = c(0, Inf)),
       link_params = list(mu = link_mu, c = link_c, k = link_k))
}
bd <- burr_distrib()
chk_burr <- check_distrib(bd, theta = list(mu = 800, c = 3, k = 1.5))
```

The data frame `rent` of the `gamlss.data` package gives the monthly rent
`R` of 1969 flats in Munich, with the floor space `Fl` in square meters,
the year of construction `A` and the location `loc` in three classes
[@stasinopoulos2017]. A first model has a linear effect of the floor space
and of the location on the scale, the location on $c$, and $k$ constant. We
fit it with `ml()` and compare it with `gamlss()`:

```{r}
#| label: fit-burr-linear
fit_burr_lin <- statmod(R ~ Fl + loc | c ~ loc, distrib = bd, data = rent,
                        outer_criterion = ml())
summary(fit_burr_lin)
gl_burr <- gamlss::gamlss(R ~ Fl + loc, sigma.formula = ~ loc,
                          nu.formula = ~ 1, tau.formula = ~ 1,
                          family = gamlss.dist::GB2(), data = rent,
                          nu.start = 1, nu.fix = TRUE, n.cyc = 200,
                          trace = FALSE)
c(statmod = as.numeric(logLik(fit_burr_lin)),
  gamlss = as.numeric(logLik(gl_burr)))
```

The log-likelihood of `statmod()` is
`r fmt(as.numeric(logLik(fit_burr_lin)) - as.numeric(logLik(gl_burr)), 2)`
higher. The algorithm of `gamlss()` updates one parameter at a time and
stops when the global deviance changes by less than its tolerance, here
before the maximum; its estimate of $\log k$ is
`r fmt(coef(gl_burr, what = "tau")[[1]], 2)` against
`r fmt(coef(fit_burr_lin)$k[[1]], 2)`.

### Smooth effects

The rent grows with the floor space and with the year of construction, and
neither effect is linear. We give both a smooth term in the equation of the
scale, and we compare the Burr XII model with a gamma model that has the
same equation for the mean and the location for the dispersion:

```{r}
#| label: fit-burr-smooth
t_burr_density <- elapsed(
  fit_burr <- statmod(R ~ s(Fl) + s(A) + loc | c ~ loc, distrib = bd,
                      data = rent))
summary(fit_burr)
fit_rent_gamma <- statmod(R ~ s(Fl) + s(A) + loc | phi ~ loc,
                          distrib = gamma1_distrib(), data = rent)
c(burr = summary(fit_burr)@aic, gamma = summary(fit_rent_gamma)@aic)
```

The two cAIC are close, and the Burr XII model is slightly better. The
quantile residuals of @sec-model-resid need the cdf, which for this class
is a quadrature of the density at each observation:

```{r}
#| label: burr-resid-density
t_resid_density <- elapsed(r_burr <- residuals(fit_burr))
c(sd = sd(r_burr), seconds = t_resid_density)
```

### The cdf, the quantile function and the score

The Burr XII distribution has a cdf and a quantile function in closed form,
$$
F(y) = 1 - \left\{1 + \left(\frac{y}{\mu}\right)^{c}\right\}^{-k},
\qquad
F^{-1}(p) = \mu\left\{(1 - p)^{-1/k} - 1\right\}^{1/c},
$$
and with $z = y/\mu$ and $r = z^c/(1 + z^c)$ the score is
$$
\frac{\partial \ell}{\partial \mu} = \frac{c\,\{(k + 1)\,r - 1\}}{\mu},
\qquad
\frac{\partial \ell}{\partial c} = \frac{1}{c} + \log z\,\{1 - (k + 1)\,r\},
\qquad
\frac{\partial \ell}{\partial k} = \frac{1}{k} - \log(1 + z^c).
$$
We register the three methods. The cdf is computed from the logarithm of
the survival function, $-k\log(1 + z^c)$, with `expm1()`, so that it keeps
its precision where $F$ is small:

```{r}
#| label: burr-closed
S7::method(distrib_cdf, Burr) <- function(distrib, q, theta, lower.tail = TRUE,
                                          log.p = FALSE, ...) {
  mu <- theta[[1]]
  cc <- theta[[2]]
  k <- theta[[3]]
  log_s <- -k * log1p((pmax(q, 0) / mu)^cc)
  p <- if (lower.tail) -expm1(log_s) else exp(log_s)
  if (log.p) log(p) else p
}
S7::method(distrib_quantile, Burr) <- function(distrib, p, theta,
                                               lower.tail = TRUE,
                                               log.p = FALSE, ...) {
  mu <- theta[[1]]
  cc <- theta[[2]]
  k <- theta[[3]]
  if (log.p) p <- exp(p)
  if (!lower.tail) p <- 1 - p
  mu * ((1 - p)^(-1 / k) - 1)^(1 / cc)
}
S7::method(distrib_gradient, Burr) <- function(distrib, y, theta,
                                               scale = c("parameter", "link"),
                                               ...) {
  mu <- theta[[1]]
  cc <- theta[[2]]
  k <- theta[[3]]
  z <- y / mu
  lz <- log(z)
  r <- z^cc / (1 + z^cc)
  list(mu = cc * ((k + 1) * r - 1) / mu,
       c = 1 / cc + lz * (1 - (k + 1) * r),
       k = 1 / k - log1p(z^cc))
}
chk_burr_closed <- check_distrib(bd, theta = list(mu = 800, c = 3, k = 1.5))
```

The table now has a row for the score, because the score is the class's
own method and finite differences of the log-density are an independent
check of it. We fit the same model again and compute the residuals with the
closed cdf:

```{r}
#| label: fit-burr-closed
t_burr_closed <- elapsed(
  fit_burr_c <- statmod(R ~ s(Fl) + s(A) + loc | c ~ loc, distrib = bd,
                        data = rent))
t_resid_closed <- elapsed(r_burr_c <- residuals(fit_burr_c))
data.frame(
  methods = c("density", "density, cdf, quantile, score"),
  logLik = c(logLik(fit_burr), logLik(fit_burr_c)),
  fit_seconds = c(t_burr_density, t_burr_closed),
  residual_seconds = c(t_resid_density, t_resid_closed))
```

The fit reaches the same point, and the residuals, which needed one
quadrature for each of the 1969 observations, now cost one evaluation of a
formula. @fig-burr-qq compares the quantile residuals of the Burr XII model
and of the gamma model.

```{r}
#| label: fig-burr-qq
#| fig-cap: "Normal quantile plots of the quantile residuals of the Burr XII model (left) and of the gamma model (right) for the Munich rents."
#| echo: false
#| fig-height: 3.6
r_gamma <- residuals(fit_rent_gamma)
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
qqnorm(r_burr_c, main = "Burr XII", pch = 20, col = "grey40", cex = 0.5)
abline(0, 1)
qqnorm(r_gamma, main = "gamma", pch = 20, col = "grey40", cex = 0.5)
abline(0, 1)
par(op)
```

## A link with a parameter: Aranda-Ordaz {#sec-own-link}

A link is an S7 class that inherits from `link`, with the two methods
`linkfun()` and `linkinv()` (@sec-glm-links). The property `link_params`
holds the parameters of the link, so one class describes a whole family of
links. The asymmetric family of @arandaordaz1981 for a probability $\mu$ is
$$
\eta = \log\frac{(1 - \mu)^{-\alpha} - 1}{\alpha},
\qquad
\mu = 1 - (1 + \alpha e^{\eta})^{-1/\alpha},
\qquad \alpha > 0 .
$$
With $\alpha = 1$ it is the logit link, and as $\alpha$ tends to zero it
tends to the complementary log-log link. The class reads $\alpha$ from
`link_params`, and a constructor sets it:

```{r}
#| label: ao-class
ArandaOrdaz <- S7::new_class("ArandaOrdaz", parent = link)
S7::method(linkfun, ArandaOrdaz) <- function(x, theta) {
  a <- x@link_params$alpha
  log(((1 - theta)^(-a) - 1) / a)
}
S7::method(linkinv, ArandaOrdaz) <- function(x, eta) {
  a <- x@link_params$alpha
  1 - (1 + a * exp(eta))^(-1 / a)
}
ao_link <- function(alpha) {
  ArandaOrdaz(link_name = sprintf("aranda-ordaz(%g)", alpha),
              link_bounds = c(0, 1), link_params = list(alpha = alpha))
}
link_fallback_orders(ao_link(0.5))
chk_ao <- check_link(ao_link(0.5))
```

The derivatives of both directions are numerical, and `check_link()` marks
them so. The data frame `menarche` of the `MASS` package gives, for 25
groups of girls of increasing mean age, the number of girls who had reached
menarche. The parameter $\alpha$ is not estimated by `statmod()`, because a
link is fixed during a fit. We profile the log-likelihood over $\alpha$ and
maximize it with `optimize()`:

```{r}
#| label: ao-profile
ao_loglik <- function(alpha) {
  f <- statmod(cbind(Menarche, Total - Menarche) ~ Age,
               distrib = binomial_distrib(link_mu = ao_link(alpha)),
               data = menarche)
  as.numeric(logLik(f))
}
ao_opt <- optimize(ao_loglik, interval = c(0.05, 5), maximum = TRUE)
alpha_hat <- ao_opt$maximum
fit_ao <- statmod(cbind(Menarche, Total - Menarche) ~ Age,
                  distrib = binomial_distrib(link_mu = ao_link(alpha_hat)),
                  data = menarche)
summary(fit_ao)
```

The estimate is $\hat\alpha = `r fmt(alpha_hat, 3)`$. The function `glm()`
accepts a link written by the user as an object of class `link-glm`, with
the link, its inverse and the derivative of the inverse. We write it for
the same $\alpha$, and we fit the logit and the complementary log-log links
for comparison:

```{r}
#| label: ao-glm
ao_glm <- function(a) {
  structure(list(
    linkfun = function(mu) log(((1 - mu)^(-a) - 1) / a),
    linkinv = function(eta) 1 - (1 + a * exp(eta))^(-1 / a),
    mu.eta = function(eta) exp(eta) * (1 + a * exp(eta))^(-1 / a - 1),
    valideta = function(eta) TRUE, name = "aranda-ordaz"), class = "link-glm")
}
glm_ao <- glm(cbind(Menarche, Total - Menarche) ~ Age,
              family = binomial(link = ao_glm(alpha_hat)), data = menarche)
rbind(statmod = coef(fit_ao)$mu, glm = coef(glm_ao))
fit_logit <- statmod(cbind(Menarche, Total - Menarche) ~ Age,
                     distrib = binomial_distrib(link_mu = logit_link()),
                     data = menarche)
fit_cloglog <- statmod(cbind(Menarche, Total - Menarche) ~ Age,
                       distrib = binomial_distrib(link_mu = cloglog_link()),
                       data = menarche)
c(aranda_ordaz = as.numeric(logLik(fit_ao)), glm = as.numeric(logLik(glm_ao)),
  logit = as.numeric(logLik(fit_logit)),
  cloglog = as.numeric(logLik(fit_cloglog)))
```

The two fits with the Aranda-Ordaz link agree. The estimate of $\alpha$
lies above one, on the other side of the logit link from the complementary
log-log link, and the log-likelihood rises by
`r fmt(as.numeric(logLik(fit_ao)) - as.numeric(logLik(fit_logit)), 2)`
over the logit link, which is the special case $\alpha = 1$. The
likelihood ratio statistic for $\alpha = 1$ is therefore
`r fmt(2 * (as.numeric(logLik(fit_ao)) - as.numeric(logLik(fit_logit))), 2)`
on one degree of freedom. @fig-ao shows the three fitted curves.

```{r}
#| label: fig-ao
#| fig-cap: "Proportion of girls who had reached menarche against the mean age of each group, with the fitted probability under the Aranda-Ordaz link at the estimated alpha (solid), the logit link (dashed) and the complementary log-log link (dotted)."
#| echo: false
#| fig-height: 3.8
ages <- data.frame(Age = seq(min(menarche$Age), max(menarche$Age),
                             length.out = 200), Total = 1)
op <- par(mar = c(4, 4, 1, 1))
plot(menarche$Age, menarche$Menarche / menarche$Total, pch = 1,
     col = "grey40", xlab = "mean age (years)",
     ylab = "proportion reached menarche")
lines(ages$Age, predict(fit_ao, what = "mu", newdata = ages), lwd = 2)
lines(ages$Age, predict(fit_logit, what = "mu", newdata = ages), lwd = 1.5,
      lty = 2)
lines(ages$Age, predict(fit_cloglog, what = "mu", newdata = ages), lwd = 1.5,
      lty = 3)
legend("bottomright", lty = 1:3, lwd = c(2, 1.5, 1.5), bty = "n",
       legend = c(sprintf("Aranda-Ordaz, alpha = %.2f", alpha_hat), "logit",
                  "complementary log-log"))
par(op)
```

### Analytic derivatives of a link

A method for one of the eight derivative generics replaces its fallback, as
a method of a distribution does. The derivative of the inverse link is
$$
\frac{d\mu}{d\eta} = e^{\eta}\,(1 + \alpha e^{\eta})^{-1/\alpha - 1}.
$$
We register it as the method of `dlinkinv()`. The fallbacks of the higher
orders of the inverse link then take one central difference of this
analytic first derivative, and `check_link()` compares it with a finite
difference of `linkinv()`:

```{r}
#| label: ao-dlinkinv
S7::method(dlinkinv, ArandaOrdaz) <- function(x, eta) {
  a <- x@link_params$alpha
  exp(eta) * (1 + a * exp(eta))^(-1 / a - 1)
}
link_fallback_orders(ao_link(0.5))
chk_ao_d <- check_link(ao_link(0.5))
```

The first order of the inverse link is now analytic and passes the check,
and the four orders above it are reported as numerical.

```{r}
#| label: gate-15
#| include: false
assert_own_ok(
  fits = list(gp = fit_gp, gp_s = fit_gp_s, gp_nb = fit_gp_nb,
              gp_r = fit_gp_r, gp_ra = fit_gp_ra,
              burr_lin = fit_burr_lin, burr = fit_burr, burr_c = fit_burr_c,
              rent_gamma = fit_rent_gamma, ao = fit_ao, logit = fit_logit,
              cloglog = fit_cloglog),
  ref = list(tmb_gp = tmb_gp, tmb_gp_r = tmb_gp_r, gl_burr = gl_burr,
             glm_ao = glm_ao),
  data = list(warpbreaks = warpbreaks, rent = rent, menarche = menarche),
  claims = list(gp = gp, bd = bd, chk_gp = chk_gp, chk_burr = chk_burr,
                chk_burr_closed = chk_burr_closed, chk_ao = chk_ao,
                chk_ao_d = chk_ao_d, score_gap = score_gap,
                alpha_hat = alpha_hat, ao_loglik = ao_loglik,
                r_burr = r_burr, r_burr_c = r_burr_c,
                t = c(gp_density = t_gp_density, gp_score = t_gp_score,
                      burr_density = t_burr_density,
                      burr_closed = t_burr_closed,
                      resid_density = t_resid_density,
                      resid_closed = t_resid_closed)))
```
