# Writing a model term {#sec-own-terms}

```{r}
#| label: setup-16
#| include: false
source("R/_setup.R")
source("R/own-terms-certificates.R")
source("R/own-optim-certificates.R")
fmt <-function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
sci <- function(x, d = 1) {
  e <- floor(log10(abs(x)))
  sprintf("%.*f \\times 10^{%d}", d, x / 10^e, e)
}
elapsed <- function(expr) {
  t0 <- proc.time()[[3]]
  force(expr)
  proc.time()[[3]] - t0
}
```

Chapter 14 showed how `statmod()` uses the terms of `modelterms7`: a linear
term, a smooth, a random effect, a break-point and a few others, each written
in the formula and each contributing a block of the design. A term that the
package does not provide is written by the user as an S7 object of the same
kind, and the formula then accepts it. This chapter writes two. The first is
a seasonal term with a fixed block, which uses one of the penalties of
`penalties7`. The second is an asymptotic growth term whose block moves with
its own coefficients, which is the case the package handles through a
refresh step at each iteration.

## The structure of a model term {#sec-own-term-structure}

A term of the additive branch is an S7 class that inherits from
`additive_term`. The class adds four properties to the common ones of every
term: the design block `X`, the names of its coefficients `coef_names`, a
list `blueprint` that keeps what is needed to rebuild the block at new data,
and the penalty `penalty`, which is `NULL` for an unpenalized term. The
label, the hyperparameters and the variables of the formula are common to all
terms.

Four methods make the term work in a fit. `term_build()` receives the term and
the data, evaluates the variables and returns the term with `X`, `coef_names`,
`blueprint` and `penalty` filled in. `term_predict()` receives the term and
new data and returns the block at those data; `predict()` and the residuals
use it. A term whose block does not depend on the coefficients needs nothing
more. A term whose block does depend on them also has `term_refresh()`, which
receives the term and a vector of coefficients and returns the term with the
block recomputed at those values, `term_value()`, which returns the fitted
contribution to the predictor, and `term_coef_start()`, which gives starting
values for the coefficients. Section @sec-own-asym shows these three.

The function `check_term()` runs the methods on a data frame and reports six
checks. It builds the block, names its columns, counts its parameters, checks
the penalty when there is one, reproduces the block at the training data, and
gives the matching rows of the block for a subset of the data. A check that
fails names the method at fault, so a term that passes is one that `statmod()`
can fit.

## A fixed block: a seasonal term {#sec-own-season}

A seasonal cycle of period $P$ is a sum of harmonics. With $\omega = 2\pi/P$
and $k$ harmonics, the block has the columns $\cos(j\omega t)$ and
$\sin(j\omega t)$ for $j = 1, \ldots, k$. We write a term `season(x, period,
k)` whose block is this matrix, with the column $x$ taken as the time within
the period.

```{r}
#| label: season-class
SeasonTerm <- S7::new_class("SeasonTerm", parent = additive_term,
                            properties = list(expr = S7::class_any,
                                              spec = S7::class_list))
season <- function(x, period, k = 3) {
  xe <- substitute(x)
  SeasonTerm(label = sprintf("season(%s)", deparse(xe)), expr = xe,
             spec = list(period = period, k = k), hyper = list(), X = NULL,
             coef_names = character(0), blueprint = list(), penalty = NULL)
}
season_X <- function(x, period, k) {
  w <- outer(2 * pi * x / period, seq_len(k))
  X <- cbind(cos(w), sin(w))
  colnames(X) <- c(paste0("cos", seq_len(k)), paste0("sin", seq_len(k)))
  X
}
S7::method(term_build, SeasonTerm) <- function(term, data, ...) {
  x <- as.numeric(eval(term@expr, data, baseenv()))
  sp <- term@spec
  X <- season_X(x, sp$period, sp$k)
  S7::set_props(term, X = X, coef_names = colnames(X), blueprint = sp,
                penalty = penalties7::quadratic_penalty(
                  diag(rep(seq_len(sp$k)^4, 2))))
}
S7::method(term_predict, SeasonTerm) <- function(term, newdata, ...) {
  x <- as.numeric(eval(term@expr, newdata, baseenv()))
  season_X(x, term@blueprint$period, term@blueprint$k)
}
```

The penalty is the quadratic penalty with the diagonal matrix whose entries
are $j^4$, each repeated for the cosine and the sine of harmonic $j$. The
reason lies in the derivative of the harmonic. The second derivative of
$a\cos(j\omega t) + b\sin(j\omega t)$ has amplitude $j^2\omega^2$, so the
integral of its square over one period is
$$
\int_0^P f''(t)^2\,dt = \frac{P}{2}\,\omega^4 \sum_{j=1}^{k} j^4\,(a_j^2 + b_j^2).
$$
The penalty $\lambda\,\beta^\top\!\mathrm{diag}(j^4)\,\beta$ is therefore a
roughness penalty on the integrated squared second derivative, which is the
penalty of the Fourier smoother `fourier_smooth(order = 2)` in `basis7`. The
constant $\omega^4$ is absorbed by the smoothing parameter.

We check the term on the monthly temperature of Nottingham, a series of 240
observations from 1920 to 1939 that comes with R.

```{r}
#| label: season-data
nt <- data.frame(temp = as.numeric(nottem), time = as.numeric(time(nottem)))
nt$month <- (seq_len(nrow(nt)) - 1) %% 12 + 1
chk_season <- check_term(season(month, period = 12, k = 4), nt, verbose = FALSE)
chk_season
```

Every check passes. The model has a linear trend in time and the seasonal
term for the mean, with four harmonics and the penalty above:

```{r}
#| label: fit-season
f_season <- statmod(temp ~ time + season(month, period = 12, k = 4),
                    distrib = gaussian1_distrib(), data = nt)
summary(f_season)
```

The same model written with the smoother of `basis7` reaches the same
solution. The smoother is `fourier_smooth()` with nine basis functions, which
are the constant and the four harmonics, and with the second-derivative
penalty on the interval of one period:

```{r}
#| label: fit-season-basis
f_season_b <- statmod(temp ~ time + s(month, smoother = basis7::fourier_smooth(
                                k = 9, order = 2, lower = 0, upper = 12)),
                      distrib = gaussian1_distrib(), data = nt)
c(season = as.numeric(logLik(f_season)), basis7 = as.numeric(logLik(f_season_b)),
  max_fitted_difference = max(abs(fitted(f_season) - fitted(f_season_b))))
```

The two fits have the same effective degrees of freedom and the same
criterion. Their fitted values differ by less than $10^{-5}$ in the units of
the series, which is the agreement one expects from two routes to the same
penalized problem that differ in how the basis is built.

The term works in the equation of the scale as well. A seasonal change of the
spread of the temperature is one more equation of the same kind:

```{r}
#| label: fit-season-sigma
f_season_sd <- statmod(temp ~ time + season(month, period = 12, k = 4) |
                         sigma ~ season(month, period = 12, k = 2),
                       distrib = gaussian1_distrib(), data = nt)
c(logLik = as.numeric(logLik(f_season_sd)),
  state = statmod_certificate(f_season_sd)$state)
```

The fitted values follow the series closely:

```{r}
#| label: fig-season
#| fig-cap: "Monthly temperature at Nottingham (points) and the fitted values of the seasonal model (line)."
plot(nt$time, nt$temp, pch = 16, cex = 0.5, col = "grey40",
     xlab = "year", ylab = "temperature")
lines(nt$time, fitted(f_season), lwd = 2)
```

## A moving block: an asymptotic growth term {#sec-own-asym}

The asymptotic curve $f(x) = \mathrm{Asym}\,\{1 - \exp(-e^{\mathrm{lrc}}\,x)\}$
has two coefficients, the asymptote and the log of the rate. Its block is the
Jacobian of $f$ with respect to them: the column of $\mathrm{Asym}$ is
$1 - e^{-e^{\mathrm{lrc}}x}$, and the column of $\mathrm{lrc}$ is
$\mathrm{Asym}\,x\,e^{\mathrm{lrc}}e^{-e^{\mathrm{lrc}}x}$. The block depends
on the coefficients, so it has to be recomputed whenever they move. The term
keeps the coefficients in its blueprint, and `term_refresh()` puts the new
values in and rebuilds the block.

```{r}
#| label: asym-class
AsymTerm <- S7::new_class("AsymTerm", parent = additive_term,
                          properties = list(expr = S7::class_any,
                                            start = S7::class_numeric))
asym <- function(x, start = c(Asym = 60, lrc = -2)) {
  xe <- substitute(x)
  AsymTerm(label = sprintf("asym(%s)", deparse(xe)), expr = xe, start = start,
           hyper = list(), X = NULL, coef_names = character(0),
           blueprint = list(), penalty = NULL)
}
asym_parts <- function(x, cf) {
  r <- exp(cf[2]); e <- exp(-r * x)
  list(value = cf[1] * (1 - e),
       J = cbind(Asym = 1 - e, lrc = cf[1] * x * r * e))
}
S7::method(term_build, AsymTerm) <- function(term, data, ...) {
  x <- as.numeric(eval(term@expr, data, baseenv()))
  cf <- as.numeric(term@start)
  S7::set_props(term, X = asym_parts(x, cf)$J, coef_names = c("Asym", "lrc"),
                blueprint = list(x = x, coef = cf), penalty = NULL)
}
S7::method(term_refresh, AsymTerm) <- function(term, coef, ...) {
  bp <- term@blueprint
  bp$coef <- as.numeric(coef)
  S7::set_props(term, X = asym_parts(bp$x, bp$coef)$J, blueprint = bp)
}
S7::method(term_value, AsymTerm) <- function(term, coef = NULL, newdata = NULL, ...) {
  bp <- term@blueprint
  cf <- if (is.null(coef)) bp$coef else as.numeric(coef)
  x <- if (is.null(newdata)) bp$x else as.numeric(eval(term@expr, newdata, baseenv()))
  asym_parts(x, cf)$value
}
S7::method(term_coef_start, AsymTerm) <- function(term, target = NULL, ...) {
  term@blueprint$coef
}
S7::method(term_predict, AsymTerm) <- function(term, newdata, ...) {
  asym_parts(as.numeric(eval(term@expr, newdata, baseenv())),
             term@blueprint$coef)$J
}
```

The data are the Loblolly pines of the `datasets` package: height in feet of
fourteen trees, measured at several ages, the trees being distinguished by
the factor `Seed`. We check the term and then fit the curve to the pooled
data.

```{r}
#| label: asym-check
lob <- as.data.frame(Loblolly)
chk_asym <- check_term(asym(age), lob, verbose = FALSE)
chk_asym
```

```{r}
#| label: fit-asym
f_asym <- statmod(height ~ 0 + asym(age), distrib = gaussian1_distrib(),
                  data = lob)
summary(f_asym)
n_asym <- nls(height ~ Asym * (1 - exp(-exp(lrc) * age)), data = lob,
              start = list(Asym = 60, lrc = -2))
rbind(statmod = coef(f_asym)$mu, nls = coef(n_asym))
```

The estimates agree with `nls()` to a relative difference of
$`r sci(max(abs(unname(coef(f_asym)$mu) - unname(coef(n_asym)))) / unname(coef(n_asym))[1])`$.
The two log-likelihoods
differ by a constant. The fit of `statmod()` is the restricted likelihood, and
the difference from the likelihood of `nls()` is its dispersion term, which
for $n = 84$ observations and $p = 2$ coefficients is
$42\log(84/82) - 1 = 0.0121$. The difference between the two fits is
`r fmt(as.numeric(logLik(f_asym)) - as.numeric(logLik(n_asym)), 4)`.
The same curve written with the package's nonlinear term `nl()` gives the same
log-likelihood, which is a check that the refresh of the block is right:

```{r}
#| label: fit-asym-nl
f_asym_nl <- statmod(height ~ 0 + nl(~ Asym * (1 - exp(-exp(lrc) * age)),
                                     start = list(Asym = 60, lrc = -2)),
                     distrib = gaussian1_distrib(), data = lob)
c(asym = as.numeric(logLik(f_asym)), nl = as.numeric(logLik(f_asym_nl)))
```

The term takes part in the other equations as any term does. The spread of
the height can grow with the age, and a random intercept for each tree
accounts for the repeated measures:

```{r}
#| label: fit-asym-sigma
f_asym_sd <- statmod(height ~ 0 + asym(age) | sigma ~ age,
                     distrib = gaussian1_distrib(), data = lob)
f_asym_re <- statmod(height ~ 0 + asym(age) + random(~ 1 | Seed),
                     distrib = gaussian1_distrib(), data = lob)
data.frame(model = c("sigma ~ age", "random(~ 1 | Seed)"),
           logLik = c(as.numeric(logLik(f_asym_sd)),
                      as.numeric(logLik(f_asym_re))),
           state = c(statmod_certificate(f_asym_sd)$state,
                     statmod_certificate(f_asym_re)$state))
```

The predicted curve, with its standard errors, comes from the same term:

```{r}
#| label: asym-predict
pa <- predict(f_asym, what = "mu",
              newdata = data.frame(age = c(5, 15, 25)), se = TRUE)
pa
```

```{r}
#| label: fig-asym
#| fig-cap: "Height of the Loblolly pines against age (points), with the asymptotic curve fitted by asym() (line)."
plot(lob$age, lob$height, pch = 16, cex = 0.5, col = "grey40",
     xlab = "age (years)", ylab = "height (feet)")
ag <- seq(min(lob$age), max(lob$age), length.out = 100)
lines(ag, predict(f_asym, what = "mu", newdata = data.frame(age = ag),
                  se = TRUE)$fit, lwd = 2)
```

## The structure of an optimizer and its stopping rule {#sec-own-optim}

An optimizer of `optimizers7` is an S7 object that inherits from `optimizer`.
It holds a criterion, which is its stopping rule, a budget of iterations and of
evaluations, and the settings for the trace. Its method `minimize()` receives
the objective `fn`, the starting point `par`, the gradient `gr` and the Hessian
`he`, and it returns an `optimizer_result` with the point, the value, the
number of iterations, the counts of evaluations and whether the criterion was
met. The method calls `crit_met()` after each accepted step, and the criterion
decides what finished means. The optimizer also declares what it needs, through
`optimizer_provides()`, and whether it can handle bounds, through
`optimizer_bounded()`, so that `check_optimizer()` tests it against what it
declares.

The criterion of the chapter is the relative gradient. The run stops when the
largest component of the gradient is below the tolerance times $\max(1, |f|)$:

```{r}
#| label: crit-relgrad
CritRelGrad <- S7::new_class("CritRelGrad", parent = criterion,
                             properties = list(tol = S7::class_numeric))
crit_relgrad <- function(tol = 1e-8) {
  CritRelGrad(label = sprintf("relative gradient < %g", tol), tol = tol)
}
S7::method(crit_met, CritRelGrad) <- function(criterion, state) {
  !is.null(state$gradient) &&
    max(abs(state$gradient)) < criterion@tol * max(1, abs(state$f_new))
}
S7::method(crit_needs, CritRelGrad) <- function(criterion) "gradient"
```

### The trust-region subproblem {#sec-own-tr-sub}

The method is a Newton method with a trust region @conn2000. At the current
point $x$, with gradient $g$ and Hessian $H$, it replaces the objective by the
quadratic model $m(p) = g^\top p + \tfrac12 p^\top H p$ and takes the step $p$
that minimizes $m$ inside a ball of radius $\Delta$:
$$
\min_p\; m(p) \quad \text{subject to} \quad \|p\| \le \Delta .
$$
Write $H = V\Lambda V^\top$ with $\Lambda = \mathrm{diag}(\lambda_1, \ldots,
\lambda_n)$, and let $\tilde g = V^\top g$. The solution has the form
$$
p(\mu) = -(H + \mu I)^{-1} g ,
$$
for a multiplier $\mu \ge \max(0, -\lambda_{\min})$ such that $\|p(\mu)\| =
\Delta$. If the Newton step $-H^{-1}g$ already lies inside the ball, then
$\mu = 0$ and the step is the Newton step. The multiplier is the root of the
secular equation $\|p(\mu)\| = \Delta$, which the method finds with a
bracketing root search in $\mu$.

The case that needs care is the hard case. When $\tilde g$ has no component
along the eigenvector of $\lambda_{\min}$ and $\lambda_{\min} \le 0$, the norm
$\|p(\mu)\|$ does not reach $\Delta$ as $\mu$ falls to its lower bound. The
step then takes $p = p_\perp + \tau v_{\min}$, where $p_\perp$ is the step on
the other eigenvectors and $\tau$ is chosen to reach the boundary. Between the
two signs of $\tau$ the method keeps the one with the smaller model.

The step is correct when it satisfies the first-order conditions of the
problem. These are
$$
(H + \mu I)\,p = -g, \qquad \mu \ge 0, \qquad \mu\,(\Delta - \|p\|) = 0,
\qquad H + \mu I \succeq 0,
$$
and they are sufficient for $p$ to be a global minimizer of $m$ over the
ball @conn2000. The gate of this chapter checks them on random problems.

```{r}
#| label: tr-step
tr_step <- function(g, H, Delta) {
  e <- eigen(H, symmetric = TRUE)
  lam <- e$values
  V <- e$vectors
  gt <- drop(crossprod(V, g))
  lmin <- min(lam)
  if (lmin > 0) {
    p0 <- -gt / lam
    if (sqrt(sum(p0^2)) <= Delta) {
      p <- drop(V %*% p0)
      attr(p, "mu") <- 0
      return(p)
    }
  }
  lo <- max(0, -lmin)
  scale <- max(1, abs(lmin), sqrt(sum(gt^2)))
  norm_mu <- function(mu) sqrt(sum((gt / (lam + mu))^2))
  mu_lo <- lo + 1e-12 * scale
  if (norm_mu(mu_lo) <= Delta) {
    # hard case: the most negative eigenvector carries no gradient
    i_min <- which.min(lam)
    keep <- seq_along(lam)[-i_min]
    p_perp <- -gt[keep] / (lam[keep] + lo)
    tau <- sqrt(max(0, Delta^2 - sum(p_perp^2)))
    cand <- lapply(c(1, -1), function(s) {
      pc <- numeric(length(lam))
      pc[keep] <- p_perp
      pc[i_min] <- s * tau
      pc
    })
    model <- vapply(cand, function(pc) {
      sum(gt * pc) + 0.5 * sum(lam * pc^2)
    }, numeric(1))
    p <- drop(V %*% cand[[which.min(model)]])
    attr(p, "mu") <- lo
    return(p)
  }
  mu <- stats::uniroot(function(mu) norm_mu(mu) - Delta,
                       c(mu_lo, lo + sqrt(sum(g^2)) / Delta),
                       tol = 1e-14 * scale, extendInt = "no")$root
  p <- drop(V %*% (-gt / (lam + mu)))
  attr(p, "mu") <- mu
  p
}
```

### The method and its radius {#sec-own-tr-method}

After the step, the method compares the decrease of the objective with the
decrease the model predicts. The ratio is
$$
\rho = \frac{f(x) - f(x + p)}{m(0) - m(p)} .
$$
The step is accepted when $\rho > 0$. The radius follows the classical rules of
@nocedal2006: it is divided by four when $\rho < 1/4$, and it is doubled, up to
a maximum, when $\rho > 3/4$ and the step reaches the boundary. A step that
the model did not predict well therefore shortens the radius, and a step that
the model predicted exactly lengthens it.

A second stop is needed. Near the minimum the predicted decrease falls below
the rounding of the objective, so that $f(x + p)$ and $f(x)$ are equal in
floating point, every step is rejected, and the radius falls to zero. When the
predicted decrease is below $10^3\,\varepsilon\,\max(1, |f|)$, the method stops
and says so. Its `criterion_met` then names the resolution of the objective,
and the result is not marked as converged.

```{r}
#| label: tr-class
TrustRegion <- S7::new_class("TrustRegion", parent = optimizer,
                             properties = list(radius = S7::class_numeric,
                                               radius_max = S7::class_numeric))
trust_region <- function(criterion = crit_relgrad(1e-8), radius = 1,
                         radius_max = 1e4, maxit = 200, max_eval = 2000,
                         verbose = FALSE, refresh = 10, keep_trace = FALSE) {
  TrustRegion(name = "trust-region Newton", criterion = criterion,
              maxit = maxit, max_eval = max_eval, verbose = verbose,
              refresh = refresh, keep_trace = keep_trace,
              radius = radius, radius_max = radius_max)
}
S7::method(optimizer_provides, TrustRegion) <- function(optimizer) "gradient"
# No box: the radius is the only constraint, so the method answers FALSE and
# refuses finite bounds, as prox_grad() does.
S7::method(optimizer_bounded, TrustRegion) <- function(optimizer) FALSE
S7::method(minimize, TrustRegion) <- function(optimizer, fn, par, gr = NULL,
                                              he = NULL, lower = -Inf,
                                              upper = Inf, ...) {
  if (is.null(gr) || is.null(he)) {
    stop("trust_region() needs a gradient and a Hessian.", call. = FALSE)
  }
  if (any(is.finite(lower)) || any(is.finite(upper))) {
    stop("trust_region() has no box; its radius is the only constraint.",
         call. = FALSE)
  }
  check_criterion(optimizer)
  o <- optimizer
  x <- as.numeric(par)
  f <- fn(x)
  nf <- 1L
  g <- gr(x)
  H <- he(x)
  ng <- 1L
  nh <- 1L
  Delta <- o@radius
  conv <- FALSE
  it <- 0L
  reason <- "stopped without converging"
  rows <- list()
  if (crit_met(o@criterion, list(gradient = g, f_new = f))) conv <- TRUE
  for (k in seq_len(if (conv) 0L else o@maxit)) {
    it <- k
    H <- (H + t(H)) / 2
    p <- tr_step(g, H, Delta)
    mu <- attr(p, "mu")
    pn <- sqrt(sum(p^2))
    pred <- -(sum(g * p) + 0.5 * sum(p * drop(H %*% p)))
    xn <- x + p
    fx <- fn(xn)
    nf <- nf + 1L
    rho <- if (is.finite(fx) && pred > 0) (f - fx) / pred else -Inf
    accepted <- rho > 0
    radius_in <- Delta
    if (rho < 0.25) {
      Delta <- Delta / 4
    } else if (rho > 0.75 && pn >= 0.99 * Delta) {
      Delta <- min(2 * Delta, o@radius_max)
    }
    if (o@keep_trace) {
      rows[[it]] <- data.frame(iteration = it,
                               value = if (accepted) fx else f,
                               step = pn, radius = radius_in, rho = rho,
                               accepted = accepted, mu = mu)
    }
    if (accepted) {
      g <- gr(xn)
      ng <- ng + 1L
      H <- he(xn)
      nh <- nh + 1L
      state <- list(iter = it, f_new = fx, f_old = f, x_new = xn,
                    x_old = x, gradient = g)
      x <- xn
      f <- fx
      if (crit_met(o@criterion, state)) {
        conv <- TRUE
        reason <- o@criterion@label
        break
      }
    }
    if (isTRUE(pred <= 1e3 * .Machine$double.eps * max(1, abs(f)))) {
      reason <- "resolution of the objective reached"
      break
    }
    if (nf >= o@max_eval || Delta < 1e-12) {
      reason <- "budget or radius exhausted"
      break
    }
  }
  trace <- if (o@keep_trace && length(rows)) do.call(rbind, rows) else NULL
  optimizer_result(par = x, value = f, gradient = g,
                   counts = c(f = nf, g = ng, h = nh), iterations = it,
                   converged = conv,
                   criterion_met = reason,
                   message = "", trace = trace, optimizer = o,
                   elapsed = NA_real_)
}
```

The method has to pass the same battery as the optimizers of the package. The
battery runs on the sphere, Rosenbrock and other test functions, and it checks
the value, the counts, the trace and the rejection of a bad criterion.

```{r}
#| label: tr-check
chk_tr <- check_optimizer(trust_region(), verbose = FALSE)
chk_tr$checks
```

## Using the method in a fit {#sec-own-optim-fit}

A gamma distribution fitted to the GAG concentrations of the `MASS` package
(`GAGurine`) is the smallest test. We fit it with the Fisher scoring that
`fit_distrib()` uses by default and with the trust-region method:

```{r}
#| label: fit-gamma-tr
data(GAGurine, package = "MASS")
fd_default <- fit_distrib(gamma1_distrib(), GAGurine$GAG)
fd_tr <- fit_distrib(gamma1_distrib(), GAGurine$GAG,
                     method = trust_region())
c(default = fd_default@loglik, trust_region = fd_tr@loglik)
```

Inside `statmod()` the method is an inner optimizer, which fits the
coefficients for given hyperparameters. The fit of the acceleration of the
motorcycle crash data (`mcycle`) has a smooth in the mean and in the scale, and
it runs twenty-six inner fits. To count them, we keep the results of the runs
with a wrapper that forwards every call to the optimizer it holds:

```{r}
#| label: run-log
RunLog <- S7::new_class("RunLog", parent = optimizer,
                        properties = list(inner = S7::class_any,
                                          log = S7::class_environment))
keep_runs <- function(o, log) {
  RunLog(name = o@name, criterion = o@criterion, maxit = o@maxit,
         max_eval = o@max_eval, verbose = o@verbose, refresh = o@refresh,
         keep_trace = o@keep_trace, inner = o, log = log)
}
S7::method(optimizer_provides, RunLog) <- function(optimizer) {
  optimizer_provides(optimizer@inner)
}
S7::method(optimizer_bounded, RunLog) <- function(optimizer) {
  optimizer_bounded(optimizer@inner)
}
S7::method(minimize, RunLog) <- function(optimizer, fn, par, gr = NULL,
                                         he = NULL, lower = -Inf,
                                         upper = Inf, ...) {
  r <- optimizers7::minimize(optimizer@inner, fn, par, gr = gr, he = he,
                             lower = lower, upper = upper, ...)
  n <- length(optimizer@log$runs) + 1L
  optimizer@log$runs[[n]] <- list(iterations = r@iterations,
                                  counts = r@counts,
                                  converged = r@converged,
                                  value = r@value, trace = r@trace)
  r
}
runs_summary <- function(log) {
  rs <- log$runs
  data.frame(runs = length(rs),
             iterations = sum(vapply(rs, function(r) r$iterations, 0)),
             not_converged = sum(!vapply(rs, function(r) r$converged, TRUE)))
}
```

```{r}
#| label: fit-mcycle-inner
data(mcycle, package = "MASS")
inner_log <- new.env()
inner_log$runs <- list()
f_inner <- statmod(accel ~ s(times) | sigma ~ s(times),
                   distrib = gaussian1_distrib(), data = mcycle,
                   inner_optimizer = keep_runs(trust_region(), inner_log))
summary(f_inner)
runs_summary(inner_log)
```

Every inner run meets the rule, and the log-likelihood is the one that the
default method finds. The check in the gate compares the two.

## The same method as an outer optimizer {#sec-own-optim-outer}

The outer optimizer searches over the smoothing parameters, and each of its
evaluations is a complete inner fit. We compare the trust-region method with
`newton()` in two forms, one with the default unlimited step and one with the
step limited to a norm of five, which is the `max_length` argument of
`newton()`:

```{r}
#| label: outer-runs
outer_runs <- list(
  trust_region = trust_region(criterion = crit_relgrad(1e-8), keep_trace = TRUE),
  newton_unlimited = newton(criterion = crit_relgrad(1e-8)),
  newton_limited = newton(criterion = crit_relgrad(1e-8), max_length = 5)
)
outer_fits <- list()
outer_logs <- list()
outer_time <- c()
for (nm in names(outer_runs)) {
  lg <- new.env()
  lg$runs <- list()
  outer_logs[[nm]] <- lg
  o <- keep_runs(outer_runs[[nm]], lg)
  outer_time[nm] <- elapsed(outer_fits[[nm]] <- statmod(
    accel ~ s(times) | sigma ~ s(times), distrib = gaussian1_distrib(),
    data = mcycle, outer_optimizer = o))
}
outer_table <- data.frame(
  method = names(outer_runs),
  seconds = round(as.numeric(outer_time), 2),
  iterations = vapply(outer_logs, function(lg) sum(vapply(lg$runs, function(r) r$iterations, 0)), 0),
  log_likelihood = vapply(outer_fits, function(f) as.numeric(logLik(f)), 0),
  row.names = NULL
)
outer_table
```

The three methods reach the same hyperparameters and the same log-likelihood.
The trust-region method takes `r outer_table$iterations[1]` iterations against
`r outer_table$iterations[2]` for the unlimited Newton method and
`r outer_table$iterations[3]` for the limited one. The run times are of the same
order, so the iteration count is the quantity that separates the methods here.

The trace of the trust region shows the radius and the ratio $\rho$ at each
iteration. The first step is cut by the initial radius, and the radius then
grows as the model proves reliable:

```{r}
#| label: fig-tr-trace
#| fig-cap: "Radius of the trust region (line) and ratio of actual to predicted decrease (points) along the outer search on mcycle."
tr_trace <- outer_logs$trust_region$runs[[length(outer_logs$trust_region$runs)]]$trace
par(mfrow = c(1, 2))
plot(tr_trace$iteration, tr_trace$radius, type = "b", pch = 16,
     xlab = "iteration", ylab = "radius")
plot(tr_trace$iteration, tr_trace$rho, type = "b", pch = 16,
     xlab = "iteration", ylab = expression(rho))
abline(h = c(0.25, 0.75), lty = 3)
par(mfrow = c(1, 1))
```

The starting radius of one is a constant that the reader chooses. The outer
search gives the same hyperparameters with the other two starting radii as
well, and the iteration counts show how little the choice matters:

```{r}
#| label: outer-radius
r0_counts <- vapply(c(0.1, 1, 10), function(r0) {
  lg <- new.env()
  lg$runs <- list()
  invisible(statmod(accel ~ s(times) | sigma ~ s(times),
                    distrib = gaussian1_distrib(), data = mcycle,
                    outer_optimizer = keep_runs(
                      trust_region(criterion = crit_relgrad(1e-8), radius = r0),
                      lg)))
  sum(vapply(lg$runs, function(r) r$iterations, 0))
}, 0)
names(r0_counts) <- c("r0 = 0.1", "r0 = 1", "r0 = 10")
r0_counts
```

```{r}
#| label: gate-16b
#| include: false
assert_own_optim_ok(
  fits = list(gamma_default = fd_default, gamma_tr = fd_tr,
              inner = f_inner, outer = outer_fits),
  claims = list(chk_tr = chk_tr, inner_log = inner_log, outer_table = outer_table,
                tr_trace = tr_trace, r0_counts = r0_counts),
  data = list(mcycle = mcycle))
```

```{r}
#| label: gate-16
#| include: false
assert_own_terms_ok(
  fits = list(season = f_season, season_b = f_season_b,
              season_sd = f_season_sd, asym = f_asym, asym_nl = f_asym_nl,
              asym_sd = f_asym_sd, asym_re = f_asym_re),
  ref = list(nls = n_asym, season_pred = fitted(f_season)),
  data = list(nottem = nt, Loblolly = lob),
  claims = list(chk_season = chk_season, chk_asym = chk_asym,
                season_pred_grid = predict(f_season, what = "mu",
                                           newdata = nt, se = TRUE),
                pa = pa))
```
