## Distributions {#sec-distributions}

```{r}
#| label: setup-03
#| include: false
source("R/_setup.R")
source("R/link-formulas.R")
source("R/distribution-catalog.R")
```

This chapter develops the likelihood theory that the toolkit implements and
then presents the univariate families that `distributions7` ships. The
theory comes first because it is shared: the score, the two informations, the
Bartlett identities and the change of scale are defined once, at the level of
the base class, and every distribution, including a user-defined one, inherits
them. The catalog follows, with particular attention to the places where the
parametrization chosen here differs from the one used by base R, because those
are the places where misreadings arise.

### The distrib class {#sec-distrib-object}

A distribution is an S7 object. It carries its name, its dimensionality, the
bounds of its support, the names and interpretations of its parameters, their
individual domains, and the link function attached to each parameter. It also
carries a per-parameter flag recording whether the log-likelihood is
differentiable in that parameter, whose use is described in @sec-nonregular. Two
subclasses, `continuous_distrib` and `discrete_distrib`, provide the defaults
appropriate to each kind.

```{r}
#| label: object-anatomy
d <- gamma2_distrib()
d@params
d@params_bounds
d@link_params$sigma2
```

Parameters travel through the API as a named list called `theta`, and every
generic normalizes this list before doing anything else. The normalization does three things. It reorders the entries by name, so that
`list(sigma2 = 2, mu = 3)` and `list(mu = 3, sigma2 = 2)` mean the same
thing. It strips names from the *values*, because a parameter that has passed
through a link function comes back carrying its own name, which would
otherwise leak into the results and label a density `"sigma2"`. And it
validates every value against the parameter's domain.

The domains are treated as **open** intervals. A Gaussian requires $\sigma > 0$, not $\sigma \geq 0$, and a Bernoulli
requires $0 < \mu < 1$. At the closed boundary the log-likelihood and its
derivatives do not exist, and accepting a boundary value would replace an
informative error now with an unexplained `NaN` several calls later.

```{r}
#| label: bounds-error
#| error: true
distrib_pdf(gaussian1_distrib(), 1, list(mu = 0, sigma = 0))
```

The generics fall into four groups:

| Group | Generics |
|---|---|
| Probability | `distrib_pdf`, `distrib_cdf`, `distrib_quantile`, `distrib_rng`, `distrib_atoms` |
| Parameter derivatives | `distrib_gradient`, `distrib_hessian`, `distrib_expected_hessian`, `distrib_deriv3`, `distrib_deriv4` |
| Response derivatives | `distrib_grad_y`, `distrib_hess_y` |
| Moments | `expectation`, `mean`, `variance`, `std_dev`, `skewness`, `kurtosis`, `moment` |

: The generics {.striped}

Of all of these, only `distrib_pdf` has no default; everything else can be
derived from the density alone, as the Rayleigh example of @sec-extension
showed.

### Score and information {#sec-likelihood}

Fix a distribution with density or mass function $f(y;\theta)$, where
$\theta = (\theta_1, \dots, \theta_p)^\top$, and write
$\ell(\theta; y) = \log f(y;\theta)$ for the log-likelihood of a single
observation. Derivatives with respect to the parameters are written with
parenthesized superscripts,

$$
\ell^{(i)} = \frac{\partial \ell}{\partial \theta_i}, \qquad
\ell^{(ij)} = \frac{\partial^{2} \ell}{\partial \theta_i \partial \theta_j}, \qquad
\ell^{(ijk)} = \frac{\partial^{3} \ell}{\partial \theta_i \partial \theta_j \partial \theta_k},
$$

and so on. This notation keeps the subscript position free for its usual meanings in
likelihood work: $\ell_i$ for the contribution of the $i$-th observation, or
$\ell_Y$ for the log-likelihood of a transformed model in the next section.
The parenthesized superscript unambiguously marks a derivative, as $f^{(k)}$
does for a function of one variable. The two usages
agree: when $\ell$ depends on a single parameter $\theta$, the $k$-th
derivative can be written either as $\ell^{(k)}$ with an integer order or as
$\ell^{(\theta\cdots\theta)}$ with the label repeated, and both forms appear
below, each where it is clearer.

The vector $(\ell^{(1)}, \dots, \ell^{(p)})^\top$ is the **score**; the
matrix $-[\ell^{(ij)}]$ is the **observed information**; and the **expected
information**, or Fisher information, is
$\mathcal{I}(\theta) = -\mathbb{E}[\ell^{(ij)}]$, the expectation being taken over
$Y \sim f(\cdot\,;\theta)$ at the same parameter value. The classical account of
these quantities and of the asymptotics that rest on them is @coxhinkley1974,
with @severini2000 and @pace1997 as fuller modern treatments.

Both sign conventions circulate in the literature, and a sign error is
invisible inside a symmetric matrix, so the convention used here needs to be
stated explicitly. The package returns *derivatives*,
not informations: `distrib_hessian()` gives $[\ell^{(ij)}]$, and
`distrib_expected_hessian()` gives $\mathbb{E}[\ell^{(ij)}]$, each with the sign
of the derivative, so the informations are the negatives of what the functions
return.

Since mixed partial derivatives are symmetric, only one representative of each
multi-index is stored, and components are returned as named lists keyed in a
canonical order — `"mu_sigma2"` stands for
$\ell^{(\mu\sigma^2)} = \ell^{(\sigma^2\mu)}$:

```{r}
#| label: derivatives-demo
d <- gamma2_distrib()
th <- list(mu = 3, sigma2 = 2)
y <- c(1, 2.5, 7)

str(distrib_gradient(d, y, th))
str(distrib_hessian(d, y, th))
```

Everything in this section and the next two assumes the model is *regular* at
$\theta$: the support of $f$ does not depend on $\theta$, the log-likelihood is
four times differentiable, and derivatives may be exchanged with the integral
sign at the orders involved. Every family in the catalog satisfies these
conditions everywhere except the Laplace,
whose location parameter sits at a kink of the log-likelihood, and
@sec-nonregular is devoted to what happens then.

### The Bartlett identities {#sec-bartlett}

The information matrix equality
$\mathbb{E}[\ell^{(ij)}] = -\mathbb{E}[\ell^{(i)} \ell^{(j)}]$ is the second
member of an infinite family of identities, one for each order of
differentiation, due to @bartlett1953a and @bartlett1953b and treated
systematically in @mccullagh1987. Taken together, the family allows an
expected derivative of any order to be computed from expectations of products
of lower-order ones. `distributions7` implements the family in full
generality, so this section derives it.

The starting point is the fact that a density integrates to one for every
value of the parameter,

$$\int f(y;\theta)\, dy = 1 \qquad \text{for all } \theta \in \Theta.$$

The right-hand side is constant in $\theta$, so every derivative of the
left-hand side must vanish, and regularity lets us take those derivatives under
the integral sign:

$$\int \frac{\partial^{k} f}{\partial\theta_{i_1}\cdots\partial\theta_{i_k}}\,dy = 0.
$$ {#eq-density-derivs-vanish}

Everything now depends on expressing the derivatives of $f$ through the
derivatives of $\ell = \log f$, and the expression has a precise
combinatorial structure.

::: {#lem-exp-deriv}
## Derivatives of $e^{\ell}$

For every $k \geq 1$,

$$
\frac{\partial^{k} e^{\ell}}{\partial\theta_{i_1}\cdots\partial\theta_{i_k}}
= e^{\ell} \sum_{\pi \in \Pi_k} \prod_{B \in \pi} \ell^{(B)} ,
$$

where $\Pi_k$ is the set of all partitions of the index set
$\{i_1, \dots, i_k\}$ into non-empty blocks, and $\ell^{(B)}$ is the derivative of
$\ell$ with respect to the parameters indexed by the block $B$.
:::

The formula is the Faà di Bruno expansion of the exponential, and the reason
the partitions appear is visible in a single differentiation. Differentiate the expression at order $k$ with respect to a new index. The
derivative falls either on the leading factor $e^{\ell}$, which contributes a
new block containing that index alone, or on one of the factors
$\ell^{(B)}$, which enlarges the block $B$. These are exactly the two ways of
extending a partition of a $k$-element set to a partition of a
$(k{+}1)$-element one, and each new partition arises exactly once. The
combinatorial identity behind this is the recursion for the complete Bell
polynomials, treated in @comtet1974 and @stanley2012.

Substituting the lemma into @eq-density-derivs-vanish, and noticing that $e^{\ell} = f$ is exactly
the density needed to turn the integral into an expectation, we arrive at the
result.

::: {#thm-bartlett}
## Bartlett identities

Under the regularity conditions above, for every $k \geq 1$ and every choice of
indices,

$$\sum_{\pi \,\in\, \Pi_k} \mathbb{E}\!\left[\;\prod_{B \in \pi} \ell^{(B)} \right] = 0.$$
:::

To use the identity, single out the partition consisting of one big block: its
contribution is $\mathbb{E}[\ell^{(i_1 \cdots i_k)}]$, the expected derivative of
top order, and moving everything else to the other side,

$$
\mathbb{E}\!\left[\ell^{(i_1\cdots i_k)}\right]
= -\sum_{\substack{\pi \in \Pi_k \\ |\pi| > 1}}
   \mathbb{E}\!\left[\prod_{B\in\pi}\ell^{(B)}\right].
$$ {#eq-bartlett-top-order}

The first three cases, written out, are:

$$
\begin{aligned}
k = 1: \quad & \mathbb{E}[\ell^{(i)}] = 0, \\[4pt]
k = 2: \quad & \mathbb{E}[\ell^{(ij)}] = -\mathbb{E}[\ell^{(i)}\ell^{(j)}], \\[4pt]
k = 3: \quad & \mathbb{E}[\ell^{(ijk)}] = -\Big(\mathbb{E}[\ell^{(ij)}\ell^{(k)}]
   + \mathbb{E}[\ell^{(ik)}\ell^{(j)}] + \mathbb{E}[\ell^{(jk)}\ell^{(i)}]
   + \mathbb{E}[\ell^{(i)}\ell^{(j)}\ell^{(k)}]\Big).
\end{aligned}
$$

The case $k = 1$ is the zero-mean property of the score. The case $k = 2$ is
the information matrix equality, and read from right to left it shows that the
outer-product-of-gradients estimator of the information is the second Bartlett
identity itself. The implementation follows the
mathematics directly: it enumerates the set partitions of the index set and
sums products of quantities the distribution already provides. The number of
terms is the Bell number $B_k$ — one, two, five and fifteen for
$k = 1, \dots, 4$ — so the computation stays small at every order the toolkit
supports.

#### Approximating expected derivatives {#sec-approx}

When a distribution has no closed-form expected derivative, the expectation has
to be approximated, and the derivative generics accept an `approx` argument
choosing among three strategies. The strategy `"bartlett"` (also spelled
`"opg"`) applies @eq-bartlett-top-order; at second order it needs nothing but the score, which
makes it the cheapest option, and it is the default there. The strategy
`"integrate"` computes $\mathbb{E}[\partial^k \ell]$ literally, by quadrature
for continuous distributions and by summation for discrete ones; when the
observed derivative is available in closed form this is usually the most
accurate route, and it is the default at orders three and four. The strategy
`"mc"` simulates from the distribution and averages the observed derivative
over the draws; it is robust when quadrature struggles, at the price of Monte
Carlo noise.

The first strategy differs from the other two in more than cost.
`"integrate"` and `"mc"` estimate the expected derivative itself, whereas
`"bartlett"` estimates the quantity that the identity says the expected
derivative equals *under regularity*. When regularity fails the two quantities
differ, and the quantity that retains the meaning of an information is the
Bartlett one; @sec-nonregular shows this on a concrete example.

The quadrature behind `"integrate"` splits its integral at three quantiles of
the distribution. The choice of three knots rather than more, and the failure
mode a finer subdivision introduces at extreme parameter values, are described
in @sec-fallbacks-caveat.

### Derivatives on the link scale {#sec-link-scale}

The parameters of a distribution are constrained, and the linear predictors of
a model are not, so every derivative generic accepts an argument
`scale = c("parameter", "link")`. With `scale = "link"` the derivatives are
taken with respect to

$$\eta_i = g_i(\theta_i), \qquad \theta_i = h_i(\eta_i), \qquad h_i = g_i^{-1},$$

where $g_i$ is the link that the distribution carries for its $i$-th
parameter. The essential structural fact is that each parameter has its own
link and the map is applied coordinate by coordinate, so the Jacobian of
$\theta$ with respect to $\eta$ is **diagonal**. The diagonality makes the
general multivariate chain rule collapse into a product of univariate ones,
and the derivation below proceeds in stages, starting with a single parameter.

#### One parameter {#sec-faa-univariate}

Repeated differentiation of the composition $\ell(h(\eta))$ is governed by
Faà di Bruno's formula, which organizes the higher-order chain rule through
the *partial Bell polynomials* $B_{m,j}$.

::: {#thm-faa}
## Faà di Bruno

If $\ell$ is $m$ times differentiable in $\theta$ and $h$ is $m$ times
differentiable in $\eta$, with $\theta = h(\eta)$, then

$$
\frac{\partial^{m}}{\partial\eta^{m}}\,\ell(h(\eta))
= \sum_{j=1}^{m} \ell^{(j)}(\theta)\;
  B_{m,j}\!\left(h', h'', \dots, h^{(m-j+1)}\right).
$$
:::

For the orders the toolkit uses, the Bell polynomials are small enough to list in
full,

$$
\begin{array}{llll}
B_{1,1} = h' \\
B_{2,1} = h'' & B_{2,2} = (h')^{2} \\
B_{3,1} = h''' & B_{3,2} = 3h'h'' & B_{3,3} = (h')^{3} \\
B_{4,1} = h'''' & B_{4,2} = 4h'h''' + 3(h'')^{2} & B_{4,3} = 6(h')^{2}h'' & B_{4,4} = (h')^{4}
\end{array}
$$

and substituting them into the theorem gives the four formulas that the package
applies:

$$
\begin{aligned}
\frac{\partial \ell}{\partial \eta} &= \ell^{(\theta)}\,h', \\[4pt]
\frac{\partial^{2} \ell}{\partial \eta^{2}} &= \ell^{(\theta\theta)}\,(h')^{2} + \ell^{(\theta)}\,h'', \\[4pt]
\frac{\partial^{3} \ell}{\partial \eta^{3}} &= \ell^{(\theta\theta\theta)}\,(h')^{3} + 3\,\ell^{(\theta\theta)}\,h'h'' + \ell^{(\theta)}\,h''', \\[4pt]
\frac{\partial^{4} \ell}{\partial \eta^{4}} &= \ell^{(\theta\theta\theta\theta)}\,(h')^{4} + 6\,\ell^{(\theta\theta\theta)}\,(h')^{2}h'' + \ell^{(\theta\theta)}\left(4h'h''' + 3(h'')^{2}\right) + \ell^{(\theta)}\,h''''.
\end{aligned}
$$

The first two lines can be checked by hand: the first is the chain rule, and
the second follows from differentiating it as a product. All four show the
property that makes the construction generic: each link-scale derivative is a
*linear combination* of the parameter-scale derivatives of order up to $m$,
with coefficients that depend only on the link. The transformation never needs
to know which distribution it is transforming, so it can be written once, in
the body of the generic, for all distributions at once.

#### Several parameters {#sec-faa-multivariate}

With several parameters, a mixed derivative such as
$\partial^4 \ell / \partial\eta_a^2\,\partial\eta_b^2$ must be handled, and here the
diagonality of the Jacobian becomes essential.

::: {#prp-faa-diag}
## Faà di Bruno with a diagonal Jacobian

Let the multi-index involve the distinct parameters $p_1, \dots, p_r$ with
multiplicities $m_1, \dots, m_r$, and let $k = \sum_t m_t$. Then

$$
\frac{\partial^{k} \ell}{\partial \eta_{p_1}^{m_1} \cdots \partial \eta_{p_r}^{m_r}}
= \sum_{j_1=1}^{m_1} \cdots \sum_{j_r=1}^{m_r}
   \ell^{(p_1^{j_1} \cdots p_r^{j_r})}
   \prod_{t=1}^{r} B_{m_t, j_t}\!\left(h_{p_t}', h_{p_t}'', \dots\right),
$$

where $\ell^{(p_1^{j_1}\cdots p_r^{j_r})}$ denotes the parameter-scale
derivative taken $j_t$ times with respect to $\theta_{p_t}$.
:::

The diagonality of the Jacobian is what makes this factorization possible.
Each $\theta_{p_t}$ depends on its own linear predictor and on nothing else,
so the operator $\partial^{m_t}/\partial\eta_{p_t}^{m_t}$ sees $\ell$ as a
function of the single variable $\theta_{p_t}$, and @thm-faa applies in that
variable alone. The Bell polynomial it produces involves only derivatives of
$h_{p_t}$, so it is constant under differentiation with respect to any other
predictor, and the operators can be applied one after another in any order. Each round multiplies in its own
factor and deepens the sum by one index. Without diagonality the general
multivariate version of the formula would be needed instead, and it is
considerably heavier; @constantine1996 gives it.

Two special cases cover most of what happens in practice. A purely mixed second derivative across two different parameters
involves $m_1 = m_2 = 1$, each inner sum has a single term, and the formula
collapses to a rescaling,

$$
\frac{\partial^{2}\ell}{\partial\eta_a \partial\eta_b} = \ell^{(ab)}\,h_a'\,h_b'
\qquad (a \neq b).
$$

And the number of terms in the general sum is $\prod_t m_t$ — at most eight for
the orders the toolkit supports — so the computation remains trivial even at
fourth order.

#### The expected Hessian on the link scale {#sec-link-scale-expected}

For expected second derivatives the first-order term drops out on its own.

::: {#cor-congruence}
## Congruence

$$\mathbb{E}\!\left[\frac{\partial^{2}\ell}{\partial\eta_a\partial\eta_b}\right]
= h_a'\,h_b'\;\mathbb{E}\!\left[\ell^{(ab)}\right],$$

or in matrix form
$\mathbb{E}[H_\eta] = \operatorname{diag}(h')\,\mathbb{E}[H_\theta]\,\operatorname{diag}(h')$.
:::

The first-order term is what disappears. For $a = b$ the univariate formula
gives $\partial^2\ell/\partial\eta_a^2 = \ell^{(aa)}(h_a')^2 + \ell^{(a)} h_a''$,
and taking expectations kills the second piece because the first Bartlett
identity says the score has mean zero; for $a \neq b$ there was never a
first-order term to begin with.

The corollary has a practical consequence. A congruence transform of a positive
semi-definite matrix is positive semi-definite, so the link-scale Fisher
information inherits the definiteness of the parameter-scale one
automatically. Had one instead transformed the *observed* Hessian and then
averaged numerically, the $\ell^{(a)} h_a''$ term would have survived and that
guarantee would have been lost.

#### Implementation {#sec-link-scale-impl}

The change of scale is implemented once, in the body of each derivative
generic, between the dispatch and the return: a method always computes
parameter-scale values, and the generic transforms them if `scale = "link"` was
requested.

``` r
distrib_gradient <- S7::new_generic(
  "distrib_gradient", "distrib",
  function(distrib, y, theta, scale = c("parameter", "link"), ...) {
    ...
    res <- S7::S7_dispatch()                    # parameter scale, always
    if (scale == "link") to_link_scale(...) else res
  })
```

This placement has one visible consequence for users who write their own
derivative methods. S7 requires a method's formal arguments to include the
generic's named arguments, so such a method must declare
`scale = c("parameter", "link"), ...` in its signature even though it never
touches `scale` itself. Omitting the declaration makes dispatch fail.

### Derivatives with respect to the response {#sec-response-derivatives}

Besides derivatives in the parameters, continuous distributions provide
$\partial\ell/\partial y$ and $\partial^{2}\ell/\partial y^{2}$ through
`distrib_grad_y()` and `distrib_hess_y()`; for a discrete variable there is
nothing to differentiate, so discrete distributions have no such methods. These
quantities feed quantile residuals and smoothness criteria further up the
toolkit. The default implementation is a central finite difference of the
log-density with a step that shrinks near the boundary of the support, so that
$y \pm \delta$ never leaves it.

```{r}
#| label: grad-y-demo
d <- gamma2_distrib()
th <- list(mu = 3, sigma2 = 2)
distrib_grad_y(d, c(1, 3, 6), th)
# closed form: with shape a = mu^2/sigma2 and rate b = mu/sigma2,
# dl/dy = (a - 1)/y - b
(3^2 / 2 - 1) / c(1, 3, 6) - 3 / 2
```

### Derivatives of the distribution function {#sec-cdf-derivatives}

Everything so far has differentiated the *density*. There is a second family of
derivatives that a modeling framework needs and that is almost never provided:
those of the **distribution function** with respect to the parameters. The
reason they matter is censoring. An observation known only to be at most $q$
contributes $\log F(q;\theta)$ to the log-likelihood rather than
$\log f(y;\theta)$; one known only to exceed $q$ contributes
$\log\big(1 - F(q;\theta)\big)$; and an interval-censored one contributes
$\log\big(F(b;\theta) - F(a;\theta)\big)$. Without $\partial F/\partial\theta$
none of these can be fitted by any method that uses a score, which rules out
survival models, tobit models and every problem with a detection limit; the
likelihoods themselves are standard, and @klein2003 sets them out in full.

The mathematics is a single exchange of derivative and integral. The region of
integration is $(-\infty, q]$, which is fixed and does not depend on $\theta$
— the same condition that makes truncation a regular problem in
@thm-truncated — so

$$
\frac{\partial F(q;\theta)}{\partial\theta_i}
= \int_{-\infty}^{q}\frac{\partial f}{\partial\theta_i}\,\mathrm{d}y
= \int_{-\infty}^{q} f\,\ell^{(i)}\,\mathrm{d}y
= F(q)\;\mathbb{E}\!\left[\ell^{(i)} \,\middle|\, Y \leq q\right],
$$

using $\partial_i f = f\,\ell^{(i)}$ in the middle step and dividing by $F(q)$
in the last to recognize the conditional density. So the gradient of the *log*
distribution function is a **partial mean of the score**:
$\partial_i \log F(q) = \mathbb{E}[\ell^{(i)} \mid Y \le q]$.

The second order follows the same way, with @lem-fij supplying the integrand:

$$
\frac{\partial^{2} F(q)}{\partial\theta_i\partial\theta_j}
= F(q)\;\mathbb{E}\!\left[\ell^{(ij)} + \ell^{(i)}\ell^{(j)} \,\middle|\, Y \leq q\right],
$$

and the log scale from $\partial_{ij}\log P = \partial_{ij}P/P - (\partial_iP/P)(\partial_jP/P)$.

The implementation treats the two kinds of distribution differently.
For a **discrete** distribution the expectation is a finite sum over the support
up to $q$, so the identity is not an approximation at all and can be evaluated
as it stands. For a **continuous** one it is an integral over a semi-infinite
region, and evaluating it by quadrature is both slower and less accurate than
differencing the distribution function, which for every family in the
catalog is an analytic expression. The package uses the exact sum for
discrete families and the difference for continuous ones.

Some families need neither. For a location-scale family, $F(y) = G(z)$ with
$z = (y-\mu)/\sigma$, and differentiating the composition gives the whole set in
terms of quantities the distribution already provides:

$$
\frac{\partial F}{\partial\mu} = -f(y), \qquad
\frac{\partial F}{\partial\sigma} = -z\,f(y), \qquad
\frac{\partial^{2} F}{\partial\mu^{2}} = f(y)\,\frac{\partial \ell}{\partial y},
$$

and similarly for the remaining second derivatives. The first of these has a
direct interpretation: the sensitivity of the distribution function to a shift
in location is minus the density at that point, because shifting the
distribution to the right by $\mathrm{d}\mu$ removes exactly
$f(q)\,\mathrm{d}\mu$ of mass from below $q$.

```{r}
#| label: cdf-derivatives
d <- gaussian1_distrib()
th <- list(mu = 1.2, sigma = 1.7)
q <- c(0, 1.2, 3)

distrib_grad_cdf(d, q, th, log = FALSE)$mu
-dnorm(q, 1.2, 1.7)
```

Putting it to work on a censored sample takes one line per kind of observation:

```{r}
#| label: censored-score
set.seed(1)
y <- distrib_rng(d, 200, list(mu = 1, sigma = 2))
cens <- y > 2.5          # everything above 2.5 is only known to be above it
y[cens] <- 2.5

# the score in mu: ordinary observations contribute the usual score, censored
# ones the derivative of log(1 - F)
sum(distrib_gradient(d, y[!cens], th)$mu) +
  sum(distrib_grad_cdf(d, y[cens], th, lower.tail = FALSE, log = TRUE)$mu)
```

An interval-censored observation is assembled from the unlogged derivatives at
the two endpoints, $\partial\log(F(b)-F(a)) = (\partial F(b) - \partial F(a))/(F(b)-F(a))$,
which is why the generic offers `log = FALSE` as well.

### One name per parametrization {#sec-numbering}

Several families are useful in more than one parametrization. A Gaussian may
carry its scale, its variance or its precision as the second parameter; a
negative binomial may have its variance quadratic in the mean or linear in
it; a von Mises may carry its concentration or its mean resultant length. In
a modeling framework the parametrization decides what a linear predictor
acts on, so each of these is a genuinely different model, and the package
gives each one a name of its own: a numeral appended to the family, as in
`gaussian1` for the scale form, `gaussian2` for the variance and `gaussian3`
for the precision. Where the literature already numbers the forms the
numbering follows it -- `negbin1` and `negbin2` are the NB1 and NB2 of
@cameron1986, and `weibull1`/`weibull3` follow the `WEI`/`WEI3` naming of
@rigby2005, which is why there is no `weibull2`. Each numbered family is a
family in its own right, with exact derivatives on its own scale.

```{r}
#| label: numbered-table
#| echo: false
knitr::kable(numbered_families_table(), align = "ll", row.names = FALSE,
             caption = "Families with several parametrizations")
```

Most of these are produced mechanically by the `reparametrize()` wrapper of
@sec-reparametrize, validated against a hand-written twin where one exists.
Two are written by hand for reasons the general mechanism cannot deliver:
the centered skew normal, whose value lies in a cancellation that a generic
chain rule would compute as a difference of large numbers, and the von Mises
in its mean resultant length, whose map has no elementary inverse and is
differentiated by the inverse function rule of @sec-numerical-special.

### Moments in closed form {#sec-closed-moments}

`mean()`, `variance()`, `skewness()` and `kurtosis()` are implemented
analytically for every family whose moment has a closed form, and fall back
on the numerical expectation otherwise; a moment that does not exist, like a
Student t's kurtosis below five degrees of freedom, is reported as `NaN`
rather than as the artifact of a divergent quadrature. The numerical route
itself is the batched quadrature of @sec-numerical-quadrature: a vector of
parameter values costs one adaptive pass, not one per value.

### The catalog {#sec-distrib-catalog}

The table below is generated from the distribution objects themselves, and
a detailed entry follows for every univariate family: constructor, support,
density in the package's parametrization, moments, and the property specific
to the family. Each printed density is compared with `distrib_pdf()` while
the book renders.

```{r}
#| label: catalog-table
#| echo: false
knitr::kable(catalog_table(), align = "lllll", row.names = FALSE,
             caption = "The univariate families")
```

A reader who wants any of these families in
more depth than a catalog entry can give will find them in @johnson1994,
@johnson1995 and @johnson2005. Several of the parametrizations differ from the ones used by the
corresponding `d*` function in base R: the Gamma is parametrized by mean and
variance rather than shape and rate, the Beta by mean and precision, the
negative binomial in its NB2 form. In each such case the entry states the map
between the two explicitly. Every univariate family implements its
observed derivatives to fourth order in closed form -- the core families as
compiled Rcpp kernels, the newer ones in vectorized R where a port was
measured to buy nothing -- except the skew t's degrees-of-freedom
components, which have no elementary form and come from one verified
stencil. The continuous-integration matrix builds the compiled kernels with
several compilers on several platforms.

```{r}
#| label: consistency-numbering
#| include: false
assert_numbering_ok()
```

```{r}
#| label: catalog-entries
#| results: asis
#| echo: false
for (id in names(DISTRIBS)) render_distrib_entry(id, DISTRIBS[[id]])
```

### Non-regular models {#sec-nonregular}

The Laplace distribution, with density
$f(y) = (2b)^{-1} e^{-|y-\mu|/b}$, is the one non-regular family in the
catalog: its log-likelihood has a kink at $y = \mu$. Families of this kind
are treated in @lehmann1998, where the information of a non-regular model is
defined as the variance of the score rather than as an expected second
derivative, which is the definition that survives here. Away from the kink
one computes

$$\ell^{(\mu)} = \frac{\operatorname{sign}(y-\mu)}{b}, \qquad \ell^{(\mu\mu)} = 0,$$

so the observed information in $\mu$ vanishes identically — the log-likelihood
is piecewise linear in the location — and yet the Fisher information certainly
exists. The second Bartlett identity delivers it directly: since
$\operatorname{sign}(y-\mu)^2 = 1$ almost surely,

$$\mathcal{I}_{\mu\mu} = \mathbb{E}\left[(\ell^{(\mu)})^{2}\right]
= \mathbb{E}\left[\frac{\operatorname{sign}(y-\mu)^{2}}{b^{2}}\right]
= \frac{1}{b^{2}}.$$

The distribution the package ships is not affected by any of this in daily
use. `laplace_distrib()` implements the expected Hessian in closed form,
returning the score-variance value $-1/b^{2}$ just computed, and, as always
when an analytical method exists, the `approx` argument is simply ignored:

```{r}
#| label: laplace-nonregular
lap <- laplace_distrib()
th <- list(mu = 0, b = 2)
c(bartlett  = distrib_expected_hessian(lap, 0, th, approx = "bartlett")$mu_mu,
  integrate = distrib_expected_hessian(lap, 0, th, approx = "integrate")$mu_mu,
  exact     = -1 / th$b^2)
```

The three strategies of @sec-approx give different answers only on a
distribution that carries no closed form for the expectation, which is the
situation of a user-defined non-regular family. The following bare Laplace
supplies only the density, the score and the observed Hessian:

```{r}
#| label: laplace-bare
BareLaplace <- S7::new_class("BareLaplace", parent = continuous_distrib, package = NULL)
S7::method(distrib_pdf, BareLaplace) <- function(distrib, y, theta, log = FALSE) {
  ld <- -log(2 * theta[[2]]) - abs(y - theta[[1]]) / theta[[2]]
  if (log) ld else exp(ld)
}
S7::method(distrib_gradient, BareLaplace) <- function(distrib, y, theta,
                                                      scale = c("parameter", "link"), ...) {
  r <- y - theta[[1]]; b <- theta[[2]]
  list(mu = sign(r) / b, b = (abs(r) / b - 1) / b)
}
S7::method(distrib_hessian, BareLaplace) <- function(distrib, y, theta,
                                                     scale = c("parameter", "link"), ...) {
  r <- y - theta[[1]]; b <- theta[[2]]; n <- length(y)
  list(mu_mu = rep(0, n), b_b = (b - 2 * abs(r)) / b^3, mu_b = -sign(r) / b^2)
}
bare <- BareLaplace(
  distrib_name = "bare laplace", dimension = "univariate", bounds = c(-Inf, Inf),
  params = c("mu", "b"), params_interpretation = c(mu = "location", b = "scale"),
  n_params = 2, params_bounds = list(mu = c(-Inf, Inf), b = c(0, Inf)),
  link_params = list(mu = identity_link(), b = log_link()),
  params_smooth = c(mu = FALSE, b = TRUE)
)

set.seed(1)
c(bartlett  = distrib_expected_hessian(bare, 0, th, approx = "bartlett")$mu_mu,
  integrate = distrib_expected_hessian(bare, 0, th, approx = "integrate")$mu_mu,
  mc        = distrib_expected_hessian(bare, 0, th, approx = "mc")$mu_mu,
  exact     = -1 / th$b^2)
```

On this distribution the strategies disagree, and the pattern of the
disagreement identifies the mechanism. The Bartlett route recovers $-1/b^{2}$
exactly, because it works from the score alone and never touches the second
derivative. The other two both return zero, for the same reason. Each averages the
observed $\ell^{(\mu\mu)}$, by quadrature in one case and over simulated
draws in the other, and that function is identically zero away from the
single point $y = \mu$, which the integral cannot see and the draws almost
surely never hit. Averaging the second derivative, however the average is taken,
produces the expected *derivative*, which for this family really is zero. What
fails is the identification of that quantity with the information, and only
the strategy built on the score remains valid when it fails.

The base class records which parameters are differentiable in the
`params_smooth` property, and this single flag propagates through the toolkit:
`fit_distrib()` works on the Laplace precisely because Fisher scoring uses the
expected information, which exists, rather than the observed one, which is
degenerate.

```{r}
#| label: params-smooth
param_smoothness(laplace_distrib())
```

The flag has a second use, in validation. Comparing an analytical derivative
against a finite difference is invalid at a kink: a central difference
straddling $y = \mu$ converges to nothing, so the *reference* is wrong, not
the value under test. When a distribution declares a non-smooth parameter, the validation machinery
recomputes each finite-difference reference with the step halved and discards
the observations where the two estimates disagree, judging the disagreement
relative to the estimates' own magnitude. Observations near the kink are
thereby excluded from the comparison. For a smooth log-likelihood nothing is
ever discarded, and the check keeps its full strength.

```{r}
#| label: consistency-distributions
#| include: false
# Internal consistency gate: fails the render if any density printed in this
# chapter disagrees with the package, if a Bartlett identity fails, if the
# link-scale machinery disagrees with finite differences in eta, or if
# check_distrib() fails anywhere in the catalog.
assert_distributions_ok()
```
