## Transformations of distributions {#sec-transformations}

```{r}
#| label: setup-04
#| include: false
source("R/_setup.R")
source("R/link-formulas.R")
source("R/distribution-catalogue.R")
source("R/transformation-certificates.R")
```

The chapters so far have dealt with fixed objects: a Gamma score is a Gamma
score, written once and used everywhere. The present chapter is about something
of a different nature — **operators on distributions**. The constructor
`zero_inflated(D)` is not itself a distribution; it is a rule that turns any
suitable $D$ into a new one, and the score of the result is assembled at run
time out of $D$'s score, $D$'s Hessian and $D$'s density at zero, whatever $D$
happens to be. The formulas such an operator uses must therefore be derived in
full generality, once, for an arbitrary parent, and that is what this chapter
does: for the change of variables, for zero-inflation, for the hurdle, for the
zero-adjusted continuous family, and for truncation, every density, score,
Hessian, expected Hessian and moment is derived from the definition with the
steps written out.

A small amount of shared notation keeps the derivations light. The parent
distribution has density or mass function $f(y;\theta)$ with log-density
$\ell(y) = \log f(y;\theta)$, and we write

$$
\ell^{(i)}(y) = \frac{\partial \ell(y)}{\partial\theta_i}, \qquad
\ell^{(ij)}(y) = \frac{\partial^{2} \ell(y)}{\partial\theta_i\partial\theta_j}, \qquad
f_0 = f(0;\theta)
$$

for its score, its Hessian and its mass at zero, following the convention of
@sec-likelihood throughout: a parenthesized superscript on $\ell$ is a
derivative with respect to the parameters it names. One elementary identity
will be needed again and again, so we record it once, with its one-line proof.

::: {#lem-fij}
## Second derivative of a density through its log

$$\frac{1}{f}\,\partial_{ij} f = \ell^{(ij)} + \ell^{(i)} \ell^{(j)},$$

where $\partial_{ij}$ abbreviates $\partial^{2}/\partial\theta_i\partial\theta_j$.
:::

::: {.proof}
From $\ell = \log f$ we get $\ell^{(i)} = (\partial_i f)/f$, that is
$\partial_i f = f\,\ell^{(i)}$. Differentiating this product with respect to
$\theta_j$ gives

$$\partial_{ij} f = (\partial_j f)\,\ell^{(i)} + f\,\partial_j \ell^{(i)}
= f\,\ell^{(j)}\ell^{(i)} + f\,\ell^{(ij)},$$

and dividing by $f$ finishes the proof. $\square$
:::

### Change of variables {#sec-transformation}

The oldest of the operations is the classical one: given a random variable $X$
with a known continuous distribution, describe the distribution of $Y = g(X)$.

::: {#def-transformer}
## Transformer

A **transformer** is a map $g : S \to T$, defined on the support $S$ of the
parent, which is a strictly monotone, continuously differentiable bijection
with $g' \neq 0$ throughout. In the package it is described by $g$, by its
inverse $g^{-1}$, by the absolute Jacobian of the inverse
$\lvert J(y)\rvert = \lvert\,\mathrm{d}g^{-1}/\mathrm{d}y\,\rvert$, by its effect on the support
bounds, and by a flag recording whether it is decreasing. Crucially, **a
transformer does not depend on $\theta$**; this apparently innocent remark will
do most of the work of the section.
:::

#### The density {#sec-transform-density}

::: {#thm-cov}
## Change of variables

If $Y = g(X)$ with $g$ a transformer, then

$$f_Y(y;\theta) = f_X\!\left(g^{-1}(y);\theta\right)\,\left\lvert J(y)\right\rvert,
\qquad y \in T .$$
:::

::: {.proof}
Suppose first that $g$ is increasing. Then the event $g(X) \leq y$ is the same
as the event $X \leq g^{-1}(y)$, so the distribution functions are related by

$$F_Y(y) = P\!\left(X \leq g^{-1}(y)\right) = F_X\!\left(g^{-1}(y)\right),$$

and differentiating in $y$ with the chain rule gives
$f_Y(y) = f_X(g^{-1}(y))\cdot\mathrm{d}g^{-1}/\mathrm{d}y$. For an increasing
$g$ the derivative of $g^{-1}$ is positive, so it coincides with its absolute
value and the theorem holds.

If instead $g$ is decreasing, the inequality flips: $g(X) \leq y$ exactly when
$X \geq g^{-1}(y)$, so $F_Y(y) = 1 - F_X(g^{-1}(y))$, and differentiation now
produces a leading minus sign, $f_Y(y) = -f_X(g^{-1}(y))\cdot\mathrm{d}g^{-1}/\mathrm{d}y$.
But in this case $\mathrm{d}g^{-1}/\mathrm{d}y$ is negative, so the minus sign
restores the absolute value and both cases end with the same formula. $\square$
:::

From the same two computations one reads off the cdf, and by inversion the
quantile function and the generator: in the increasing case
$F_Y = F_X \circ g^{-1}$ and $Q_Y(p) = g(Q_X(p))$, while in the decreasing case
the tails swap, $F_Y = 1 - F_X \circ g^{-1}$ and $Q_Y(p) = g(Q_X(1-p))$.
Generating a draw is simply a matter of drawing $X$ from the parent and
applying $g$, which is exact and costs a single evaluation. The decreasing case
is not hypothetical — the reciprocal transformer and the power transformer with
a negative exponent are both decreasing — and a forgotten tail swap produces a
quantile function that is wrong by $p \mapsto 1-p$, an error invisible at the
median and largest in the tails, where one looks last.

#### The derivatives do not change {#sec-transform-invariance}

We now come to the fact that makes this wrapper essentially free.

::: {#thm-transform-invariance}
## Invariance of the likelihood derivatives

Let $Y = g(X)$ with $g$ a transformer. Then for every order $k \geq 1$ and
every multi-index,

$$
\frac{\partial^{k}\,\ell_Y(\theta;y)}{\partial\theta_{i_1}\cdots\partial\theta_{i_k}}
= \left.\frac{\partial^{k}\,\ell_X(\theta;x)}{\partial\theta_{i_1}\cdots\partial\theta_{i_k}}
\right|_{x = g^{-1}(y)},
$$

and the expected derivatives of the two models coincide exactly:
$\mathbb{E}_Y\!\left[\partial^{k}\ell_Y\right] = \mathbb{E}_X\!\left[\partial^{k}\ell_X\right]$.
:::

::: {.proof}
Taking the logarithm of @thm-cov,

$$\ell_Y(\theta;y) = \ell_X\!\left(\theta;\, g^{-1}(y)\right) + \log\lvert J(y)\rvert .$$

Look at the two terms through the eyes of $\partial/\partial\theta$. The
Jacobian term depends on $y$ alone — this is where the requirement that a
transformer not depend on $\theta$ pays off — so it is a constant for the
purposes of differentiation in $\theta$ and disappears at the first derivative,
never to return. The first term is $\ell_X$ evaluated at the point
$x = g^{-1}(y)$, and since this point also does not move when $\theta$ does,
differentiating in $\theta$ simply passes through the substitution. That proves
the first claim at every order simultaneously.

For the second claim, observe that $g^{-1}(Y) = X$ holds with probability one,
so for any function $u$ we have
$\mathbb{E}_Y\!\left[u(g^{-1}(Y))\right] = \mathbb{E}_X[u(X)]$. Applying this
with $u = \partial^{k}\ell_X$, and using the first claim to recognise the left
side as $\mathbb{E}_Y[\partial^k \ell_Y]$, completes the proof. $\square$
:::

The theorem has two practical consequences, and both amount to *not* doing work
that a less careful implementation might do. First, no simulation is ever
needed for the expected information of a transformed distribution: the parent's
expected Hessian already *is* the answer, exactly, and the implementation is a
single line. Second, the link-scale machinery of @sec-link-scale is unaffected,
because the links attach to the parameters and the parameters have not changed.

The Jacobian, having dropped out of the $\theta$-derivatives, has not left the
theory altogether: it survives in the density itself and in the derivatives
with respect to $y$. This is why a transformer also carries
$\partial_y \log\lvert J\rvert$ and $\partial^2_y \log\lvert J\rvert$, which feed the response
derivatives of @sec-response-derivatives.

#### The twelve transformers {#sec-transformer-catalogue}

```{r}
#| label: transformer-table
#| echo: false
tt <- do.call(rbind, lapply(TRANSFORMERS, function(tr) data.frame(
  transformer = sprintf("`%s`", tr$ctor),
  `$g(x)$` = sprintf("$%s$", tr$g),
  `$g^{-1}(y)$` = sprintf("$%s$", tr$ginv),
  `$\\lvert J(y)\\rvert$` = sprintf("$%s$", tr$jac),
  `domain` = sprintf("$%s$", tr$domain),
  check.names = FALSE, stringsAsFactors = FALSE)))
knitr::kable(tt, align = "lllll", row.names = FALSE,
             caption = "The twelve transformers")
```

Each transformer declares which supports it is willing to act on, and the
constructor refuses an incompatible pairing at once, rather than letting it
surface later as a `NaN`:

```{r}
#| label: transformer-validity
#| error: true
transformation(gaussian_distrib(), log_transform())
```

As a closing illustration, the lognormal distribution can be reached along two
entirely different roads — as a catalogue entry with its own density, or as the
exponential of a Gaussian through @thm-cov — and the two agree to machine
precision:

```{r}
#| label: lognormal-identity
y <- seq(0.05, 12, length.out = 40)
a <- distrib_pdf(transformation(gaussian_distrib(), exp_transform()),
                 y, list(mu = 0.5, sigma = sqrt(1.3)), log = TRUE)
b <- distrib_pdf(lognormal_distrib(), y, list(mu = 0.5, sigma2 = 1.3), log = TRUE)
max(abs(a - b))
```

### Zero-inflation {#sec-zero-inflation}

Count data very often contain more zeros than any standard count distribution
can produce, and the classical remedy is to let the zeros come from two
sources: the count process itself, and a separate mechanism that switches the
count off entirely.

::: {#def-zi}
## Zero-inflated distribution

Let $D$ be a discrete distribution with $0$ in its support and mass function
$f(y;\theta)$, and let $\zeta \in (0,1)$. The **zero-inflated** version of $D$
is the mixture

$$
P(Y = y;\theta,\zeta) = \zeta\,\mathbb{I}(y=0) + (1-\zeta) f(y;\theta)
= \begin{cases}
\zeta + (1-\zeta) f_0 & y = 0,\\[2pt]
(1-\zeta) f(y;\theta) & y > 0 .
\end{cases}
$$
:::

Summing over the support gives $\zeta + (1-\zeta)\sum_y f(y) = 1$, so the
mixture is a proper distribution. Its interpretation should be kept clearly in
mind, because it drives every formula below: an observed zero may be
*structural*, produced by the switching mechanism with probability $\zeta$, or
it may be an ordinary *sampling* zero from the count process, and no individual
zero can be attributed to either source — only their proportions are
identified. Two derived quantities will recur constantly. The first is the
total mass at zero,

$$L_0 = \zeta + (1-\zeta)f_0,$$

and the second is the conditional probability that an observed zero came from
the count process,

$$w_0 = \frac{(1-\zeta)f_0}{L_0} \in (0,1).$$

Readers familiar with the EM algorithm will recognise $w_0$ as the E-step
weight of this model, and its appearance throughout the derivatives below is no
coincidence.

Before the derivatives, the distribution functions, which follow from the
mixture structure directly. The cdf mixes the parent's cdf with a step at zero,

$$F_Y(q) = (1-\zeta)F(q;\theta) + \zeta\,\mathbb{I}(q \geq 0),$$

and inverting it splits the probability axis at $L_0 = F_Y(0)$: any
$p \leq L_0$ lands in the mass at zero, while for $p > L_0$ we solve
$\zeta + (1-\zeta)F(q) = p$ for $F(q)$ and apply the parent's quantile
function, obtaining

$$
Q_Y(p) = \begin{cases}
0 & p \leq L_0,\\[4pt]
Q\!\left(\dfrac{p - \zeta}{1-\zeta};\theta\right) & p > L_0 .
\end{cases}
$$

Generation is even simpler: draw from the parent, then replace a
$\mathrm{Bernoulli}(\zeta)$ fraction of the draws by structural zeros.

#### The score {#sec-zi-score}

::: {#prp-zi-score}
## Zero-inflated score

$$
\frac{\partial \ell_Y}{\partial \theta_i} =
\begin{cases} w_0\, \ell^{(i)}(0) & y = 0 \\ \ell^{(i)}(y) & y > 0\end{cases}
\qquad\qquad
\frac{\partial \ell_Y}{\partial \zeta} =
\begin{cases} \dfrac{1 - f_0}{L_0} & y = 0 \\[6pt] -\dfrac{1}{1-\zeta} & y > 0 .\end{cases}
$$
:::

::: {.proof}
Consider first an observation $y > 0$, where the log-likelihood is
$\log(1-\zeta) + \ell(y)$. The two terms separate cleanly: the first involves
only $\zeta$ and differentiates to $-1/(1-\zeta)$, the second involves only
$\theta$ and differentiates to $\ell^{(i)}(y)$.

At $y = 0$ the log-likelihood is $\log L_0$, and both parameters enter through
$L_0$. For $\theta_i$, using $\partial_i f_0 = f_0\,\ell^{(i)}(0)$,

$$\frac{\partial \ell_Y}{\partial\theta_i}
= \frac{(1-\zeta)\,\partial_i f_0}{L_0}
= \frac{(1-\zeta) f_0}{L_0}\cdot \ell^{(i)}(0) = w_0\,\ell^{(i)}(0),$$

while for $\zeta$ the derivative of $L_0 = \zeta + (1-\zeta)f_0$ is simply
$1 - f_0$, giving $(1-f_0)/L_0$. $\square$
:::

The formula is worth reading in words: away from zero, the score is the
parent's, untouched; at the zeros, the parent's score is *down-weighted* by
$w_0$, the probability that the zero actually came from the count process. An
observation that is almost certainly structural contributes almost nothing to
the estimation of $\theta$, which is exactly as it should be.

#### The observed Hessian {#sec-zi-hessian}

::: {#prp-zi-hessian}
## Zero-inflated observed Hessian

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j} =
\begin{cases}
w_0 \ell^{(ij)}(0) + w_0(1-w_0)\,\ell^{(i)}(0)\ell^{(j)}(0) & y = 0\\[4pt]
\ell^{(ij)}(y) & y > 0
\end{cases}
$$

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\zeta} =
\begin{cases} -\dfrac{f_0\,\ell^{(i)}(0)}{L_0^{2}} & y = 0 \\[6pt] 0 & y > 0\end{cases}
\qquad\quad
\frac{\partial^{2}\ell_Y}{\partial\zeta^{2}} =
\begin{cases} -\dfrac{(1-f_0)^{2}}{L_0^{2}} & y = 0 \\[6pt] -\dfrac{1}{(1-\zeta)^{2}} & y > 0 .\end{cases}
$$
:::

::: {.proof}
We work block by block, and the only case requiring care is $y = 0$, since for
$y > 0$ the separated form $\log(1-\zeta) + \ell(y)$ immediately gives
$\ell^{(ij)}(y)$ for the $\theta\theta$ block, zero for the mixed block, and
$-1/(1-\zeta)^2$ for the $\zeta\zeta$ block.

*The $\theta\theta$ block at $y = 0$.* We differentiate the score
$\partial_i \ell_Y = (1-\zeta)\,\partial_i f_0 / L_0$ with respect to
$\theta_j$, applying the quotient rule and remembering that
$\partial_j L_0 = (1-\zeta)\,\partial_j f_0$:

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j}
= \frac{(1-\zeta)\,\partial_{ij} f_0}{L_0}
- \frac{(1-\zeta)^{2}\,\partial_i f_0\,\partial_j f_0}{L_0^{2}} .
$$

Each of the two pieces can now be rewritten in terms of $w_0$. For the first,
multiply and divide by $f_0$ and invoke @lem-fij:

$$\frac{(1-\zeta)\,\partial_{ij} f_0}{L_0}
= w_0\,\frac{\partial_{ij} f_0}{f_0}
= w_0\left(\ell^{(ij)}(0) + \ell^{(i)}(0)\ell^{(j)}(0)\right).$$

For the second, note that $(1-\zeta)\,\partial_i f_0 / L_0 = w_0\,\ell^{(i)}(0)$, so
the piece equals $w_0^{2}\,\ell^{(i)}(0)\ell^{(j)}(0)$. Subtracting, the pure product terms
combine into $w_0(1 - w_0)\,\ell^{(i)}(0)\ell^{(j)}(0)$, and the stated formula emerges.

*The mixed block at $y = 0$.* Differentiate the same score, this time with
respect to $\zeta$. Both the factor $(1-\zeta)$ and the denominator $L_0$
depend on $\zeta$, with $\partial_\zeta L_0 = 1 - f_0$, so

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\zeta}
= -\frac{\partial_i f_0}{L_0}
- \frac{(1-\zeta)\,\partial_i f_0\,(1-f_0)}{L_0^{2}}
= -\frac{\partial_i f_0}{L_0^{2}}\Big[L_0 + (1-\zeta)(1-f_0)\Big].
$$

At this point something pleasant happens: the bracket collapses. Expanding it,

$$L_0 + (1-\zeta)(1-f_0)
= \zeta + (1-\zeta)f_0 + (1-\zeta) - (1-\zeta)f_0
= \zeta + (1-\zeta) = 1,$$

so the whole block reduces to $-\partial_i f_0 / L_0^{2} = -f_0\,\ell^{(i)}(0)/L_0^{2}$.
The simplification is not cosmetic — it removes a numerical cancellation that
the unsimplified expression would have to perform at run time.

*The $\zeta\zeta$ block at $y = 0$.* Differentiating $(1-f_0)/L_0$ in $\zeta$
once more, only the denominator moves, and
$\partial_\zeta (1/L_0) = -(1-f_0)/L_0^{2}$ gives $-(1-f_0)^2/L_0^2$. $\square$
:::

#### The expected Hessian {#sec-zi-expected}

To average the observed Hessian we split the expectation according to where the
mass sits: the point $\{0\}$ carries probability $L_0$, and each $y > 0$
carries $(1-\zeta)f(y)$, so that for any function $u$,

$$\mathbb{E}[u(Y)] = L_0\,u(0) + (1-\zeta)\sum_{y>0} f(y)\,u(y).$$

::: {#prp-zi-expected}
## Zero-inflated expected Hessian

$$
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j}\right]
= L_0\Big[w_0\ell^{(ij)}(0) + w_0(1-w_0)\ell^{(i)}(0)\ell^{(j)}(0)\Big]
+ (1-\zeta)\Big(\mathbb{E}_f[\ell^{(ij)}] - f_0 \ell^{(ij)}(0)\Big),
$$

$$
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\zeta}\right]
= -\frac{f_0\,\ell^{(i)}(0)}{L_0},
\qquad
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\zeta^{2}}\right]
= -\left[\frac{(1-f_0)^{2}}{L_0} + \frac{1-L_0}{(1-\zeta)^{2}}\right].
$$
:::

::: {.proof}
For the $\theta\theta$ block, the zero branch of @prp-zi-hessian is a constant,
so it contributes $L_0$ times its value. The positive branch contributes
$(1-\zeta)\sum_{y>0} f(y) \ell^{(ij)}(y)$; since summing over the *whole* support
would give $\mathbb{E}_f[\ell^{(ij)}]$, removing the $y = 0$ term leaves
$\mathbb{E}_f[\ell^{(ij)}] - f_0 \ell^{(ij)}(0)$, and the first formula follows. The
mixed block vanishes on the positive branch, so only the zero branch survives,
weighted by $L_0$, which cancels one power of the denominator:
$L_0 \cdot (-f_0 \ell^{(i)}(0)/L_0^2) = -f_0 \ell^{(i)}(0)/L_0$. The $\zeta\zeta$ block is
the weighted average of its two branches with weights $L_0$ and $1-L_0$.

As a cross-check, the second Bartlett identity of @thm-bartlett requires this
last quantity to equal $-\mathbb{E}[(\partial_\zeta \ell_Y)^2]$; computing that
expectation from the two branches of the score gives
$-\big[L_0\,((1-f_0)/L_0)^2 + (1-L_0)/(1-\zeta)^2\big]$, which is the same
expression. $\square$
:::

#### Moments {#sec-zi-moments}

Since the extra component sits entirely at zero, it contributes nothing to any
moment of positive order, and for every $k \geq 1$,

$$\mathbb{E}[Y^{k}] = (1-\zeta)\,\mathbb{E}_f[Y^{k}].$$

If the parent has mean $\mu$ and variance $\sigma^{2}$, this gives
$\mathbb{E}[Y] = (1-\zeta)\mu$ at once, and for the variance we compute

$$\operatorname{Var}(Y) = (1-\zeta)\left(\sigma^{2}+\mu^{2}\right) - (1-\zeta)^{2}\mu^{2}
= (1-\zeta)\sigma^{2} + \zeta(1-\zeta)\mu^{2}.$$

For a Poisson parent, where $\sigma^2 = \mu$, the variance
$(1-\zeta)\mu + \zeta(1-\zeta)\mu^2$ strictly exceeds the mean $(1-\zeta)\mu$:
a zero-inflated Poisson is always overdispersed, which is the second reason,
after the zeros themselves, that the model gets reached for.

### Zero-adjustment: the discrete hurdle {#sec-hurdle}

Zero-inflation keeps the parent's zeros and adds more. The hurdle model takes
the opposite view: it treats the zeros as coming from a mechanism of their own,
removes the parent's mass at zero entirely, and puts a free parameter in its
place.

::: {#def-hurdle}
## Hurdle distribution

Let $D$ be a discrete distribution with $0$ in its support and $f_0 < 1$, and
let $\pi \in (0,1)$. The **zero-adjusted** (hurdle) version of $D$ is

$$
P(Y=y;\theta,\pi) =
\begin{cases}
\pi & y = 0,\\[4pt]
(1-\pi)\dfrac{f(y;\theta)}{1-f_0} & y > 0 .
\end{cases}
$$
:::

The division by $1 - f_0$ is the truncation, and it is the entire difference
between this definition and @def-zi. Since $\sum_{y>0} f(y) = 1 - f_0$ exactly,
the positive part sums to $1-\pi$ and the whole object is a proper
distribution. Two structural consequences are visible before any derivative is
taken. First, $P(Y=0) = \pi$ is completely free in $(0,1)$: where
zero-inflation can only push the mass at zero *above* $f_0$, the hurdle can
place it anywhere, so it also fits data with *fewer* zeros than the parent
would produce. Second, $\theta$ has changed its meaning — it now parametrises
the zero-truncated law rather than the original count process — and this
matters whenever the estimates are interpreted rather than merely used.

The deepest property of the model, from which everything else follows, is that
its log-likelihood **separates**. Writing it out,

$$
\ell_Y(\theta,\pi;y) =
\underbrace{\mathbb{I}(y=0)\log\pi + \mathbb{I}(y>0)\log(1-\pi)}_{\text{a Bernoulli likelihood in } \pi}
\;+\; \mathbb{I}(y>0)\underbrace{\Big[\ell(y) - \log(1-f_0)\Big]}_{\text{a truncated likelihood in } \theta},
$$

and the two groups share no parameter. All mixed $\theta$–$\pi$ derivatives
therefore vanish identically, at every order, observed and expected, and the
two halves of the model could be fitted separately with identical results.

#### The score {#sec-hurdle-score}

::: {#prp-hurdle-score}
## Hurdle score

$$
\frac{\partial\ell_Y}{\partial\theta_i} =
\begin{cases} 0 & y = 0 \\[4pt]
\ell^{(i)}(y) + \dfrac{f_0}{1-f_0}\,\ell^{(i)}(0) & y > 0\end{cases}
\qquad\qquad
\frac{\partial\ell_Y}{\partial\pi} =
\begin{cases} \dfrac{1}{\pi} & y = 0 \\[6pt] -\dfrac{1}{1-\pi} & y > 0 .\end{cases}
$$
:::

::: {.proof}
At $y = 0$ the log-likelihood is $\log\pi$, which does not contain $\theta$, so
the $\theta$-score vanishes and the $\pi$-score is $1/\pi$. At $y > 0$ it is
$\log(1-\pi) + \ell(y) - \log(1-f_0)$; the $\pi$-derivative of the first term
is $-1/(1-\pi)$, the $\theta_i$-derivative of the second is $\ell^{(i)}(y)$, and for
the third,

$$-\frac{\partial}{\partial\theta_i}\log(1-f_0)
= \frac{\partial_i f_0}{1-f_0} = \frac{f_0}{1-f_0}\,\ell^{(i)}(0),$$

again using $\partial_i f_0 = f_0\,\ell^{(i)}(0)$. $\square$
:::

The extra term is the **truncation correction**, and it has a transparent
reading: because the law has been renormalised over the positive values,
raising the parent's probability of a zero now *lowers* the probability of
every positive observation, and the score must account for that.

#### The Hessians {#sec-hurdle-hessian}

::: {#prp-hurdle-hessian}
## Hurdle Hessians

Abbreviate

$$
C_{ij} = \frac{(1-f_0)\,\partial_{ij}f_0 + \partial_i f_0\,\partial_j f_0}{(1-f_0)^{2}},
\qquad\text{where}\quad
\partial_i f_0 = f_0\,\ell^{(i)}(0), \quad
\partial_{ij} f_0 = f_0\left(\ell^{(ij)}(0) + \ell^{(i)}(0)\ell^{(j)}(0)\right).
$$

Then the observed Hessian is

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j} =
\begin{cases} 0 & y=0 \\ \ell^{(ij)}(y) + C_{ij} & y>0\end{cases}
\qquad
\frac{\partial^{2}\ell_Y}{\partial\pi^{2}} =
\begin{cases} -\dfrac{1}{\pi^{2}} & y=0\\[6pt] -\dfrac{1}{(1-\pi)^{2}} & y>0,\end{cases}
$$

with the mixed block identically zero, and in expectation

$$
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j}\right]
= (1-\pi)\left[\frac{\mathbb{E}_f[\ell^{(ij)}] - f_0 \ell^{(ij)}(0)}{1-f_0} + C_{ij}\right],
\qquad
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\pi^{2}}\right] = -\frac{1}{\pi(1-\pi)} .
$$
:::

::: {.proof}
For the observed $\theta\theta$ block at $y > 0$, differentiate the score of
@prp-hurdle-score with respect to $\theta_j$. The term $\ell^{(i)}(y)$ gives
$\ell^{(ij)}(y)$. The correction term is the quotient
$\partial_i f_0 / (1-f_0)$, and by the quotient rule — keeping in mind that
$\partial_j(1-f_0) = -\partial_j f_0$, which turns the usual minus into a plus —

$$\frac{\partial}{\partial\theta_j}\,\frac{\partial_i f_0}{1-f_0}
= \frac{(1-f_0)\,\partial_{ij}f_0 + \partial_i f_0\,\partial_j f_0}{(1-f_0)^{2}}
= C_{ij},$$

where the expressions for $\partial_i f_0$ and $\partial_{ij} f_0$ come from
the definition of the score and from @lem-fij respectively. At $y = 0$ the
log-likelihood is free of $\theta$, and the $\pi$ blocks follow by
differentiating $1/\pi$ and $-1/(1-\pi)$ once more.

For the expectation of the $\theta\theta$ block, the zero branch contributes
nothing, and each positive $y$ carries mass $(1-\pi)f(y)/(1-f_0)$, so

$$\mathbb{E}\left[\partial^{2}_{ij}\ell_Y\right]
= \frac{1-\pi}{1-f_0}\sum_{y>0} f(y)\Big(\ell^{(ij)}(y) + C_{ij}\Big).$$

The conditional weights $f(y)/(1-f_0)$ sum to one over the positive values, so
the constant $C_{ij}$ passes through with factor $(1-\pi)$, and the sum
$\sum_{y>0} f(y)\ell^{(ij)}(y)$ is $\mathbb{E}_f[\ell^{(ij)}] - f_0 \ell^{(ij)}(0)$, as in the
zero-inflated case. Finally, averaging the two branches of the $\pi\pi$ block
with weights $\pi$ and $1-\pi$,

$$-\frac{\pi}{\pi^{2}} - \frac{1-\pi}{(1-\pi)^{2}}
= -\frac{1}{\pi} - \frac{1}{1-\pi} = -\frac{1}{\pi(1-\pi)},$$

which is the Fisher information of a single Bernoulli trial — exactly what the
separation of the likelihood promised. $\square$
:::

For the moments, the same argument as before — the zero contributes nothing to
positive powers — gives
$\mathbb{E}[Y^{k}] = (1-\pi)\,\mathbb{E}_f[Y^{k}]/(1-f_0)$ for every
$k \geq 1$, and in particular $\mathbb{E}[Y] = (1-\pi)\mu/(1-f_0)$.

### Zero-adjustment for continuous distributions {#sec-za-continuous}

When the parent is continuous the situation simplifies in one way and deepens
in another. It simplifies because $P(W = 0) = 0$ already, so there is nothing
to truncate: adjustment amounts to placing a point mass at zero next to the
density. It deepens because the resulting object is neither continuous nor
discrete but **mixed**, and mixed distributions require a moment of care about
what "density" even means.

::: {#def-za-continuous}
## Zero-adjusted continuous distribution

Let $W$ be continuous with density $f_W(\cdot\,;\theta)$, and let
$\pi \in (0,1)$. Define $Y$ by

$$f_Y(0) = \pi, \qquad f_Y(y) = (1-\pi)f_W(y;\theta) \quad (y \neq 0),$$

where $f_Y$ is a density with respect to the measure $\nu = \delta_0 + \lambda$
— a point mass at zero plus Lebesgue measure.
:::

Integrating against $\nu$ picks up the atom separately from the density:

$$\int f_Y \,\mathrm{d}\nu = f_Y(0) + \int f_Y\,\mathrm{d}\lambda
= \pi + (1-\pi)\cdot 1 = 1,$$

so the object is a proper distribution — but note that its density integrates,
in the ordinary Lebesgue sense, to $1 - \pi$ rather than one. This is the
family of the zero-adjusted Gamma, inverse Gaussian and lognormal used for
semicontinuous data, and of the zero-adjusted Beta used for proportions with
observed zeros.

Because part of the mass is invisible to ordinary integration, the package
gives mixed distributions a way to declare their atoms: the generic
`distrib_atoms()` returns the locations and probabilities of the point masses,
the default returning none — the right answer for every ordinary distribution,
since a continuous one has no atoms and a discrete one consists of nothing else
— and only genuinely mixed objects registering more. Every routine that
integrates a density, `check_distrib()` included, consults it.

```{r}
#| label: atoms
zag <- zero_adjusted(gamma_distrib())
distrib_atoms(zag, list(mu = 3, sigma2 = 2, za = 0.3))
distrib_atoms(gamma_distrib(), list(mu = 3, sigma2 = 2))
```

The cdf jumps by $\pi$ at zero,

$$F_Y(q) = (1-\pi)F_W(q;\theta) + \pi\,\mathbb{I}(q \geq 0),$$

and inverting it requires three regions rather than two, because nothing forces
the parent's support to lie above zero — a zero-adjusted Gaussian is a
perfectly sensible "spike at zero" model whose quantile function must be able
to return negative values. Below the jump, $Q_Y(p) = Q_W(p/(1-\pi))$; inside
the jump, that is for $(1-\pi)F_W(0) \leq p \leq (1-\pi)F_W(0) + \pi$, the
quantile is zero; above it, $Q_Y(p) = Q_W((p-\pi)/(1-\pi))$.

The likelihood separates even more completely than the hurdle's, since there is
no truncation constant at all:

$$\ell_Y = \mathbb{I}(y=0)\log\pi
+ \mathbb{I}(y\neq 0)\Big[\log(1-\pi) + \ell_W(y)\Big].$$

Reading off the branches, the score in $\theta_i$ is $\ell^{(i)}(y)$ off the atom and
zero on it; the score in $\pi$ is $1/\pi$ on the atom and $-1/(1-\pi)$ off it;
the $\theta\theta$ block of the Hessian is $\ell^{(ij)}(y)$ off the atom; the mixed
block vanishes; and since $P(Y \neq 0) = 1-\pi$ while conditionally on
$Y \neq 0$ the law of $Y$ is exactly that of $W$,

$$
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j}\right]
= (1-\pi)\,\mathbb{E}_W[\ell^{(ij)}],
\qquad
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\pi^{2}}\right] = -\frac{1}{\pi(1-\pi)} .
$$

There is no analogue of the $C_{ij}$ correction, because there was no
truncation constant to differentiate — a contrast with the hurdle worth
noticing.

Expectations decompose in the way the measure $\nu$ suggests: for any $u$ with
$\mathbb{E}_W|u(W)| < \infty$,

$$\mathbb{E}[u(Y)] = \pi\,u(0) + (1-\pi)\,\mathbb{E}_W[u(W)],$$

and this identity is registered as the `expectation()` method of the class, so
that means, variances and all higher moments account for the atom
automatically. In particular, with $\mu$ and $\sigma^2$ the parent's mean and
variance, $\mathbb{E}[Y] = (1-\pi)\mu$ and
$\operatorname{Var}(Y) = (1-\pi)\sigma^{2} + \pi(1-\pi)\mu^{2}$ — the same
formulas as for zero-inflation, which is no accident, as the next section
explains.

One last point concerns derivatives in the response. Away from zero the factor
$1-\pi$ is constant in $y$, so $\partial_y \ell_Y = \partial_y \ell_W$ exactly;
at $y = 0$, however, the log-density jumps between $\log((1-\pi)f_W)$ and
$\log\pi$, no derivative exists, and the methods return `NaN` rather than the
meaningless number a finite difference across the jump would produce.

```{r}
#| label: response-at-atom
th <- list(mu = 3, sigma2 = 2, za = 0.3)
distrib_grad_y(zag, c(0, 1, 4), th)
distrib_grad_y(gamma_distrib(), c(1, 4), list(mu = 3, sigma2 = 2))
```

### Truncation {#sec-truncation}

The last of the operations restricts a distribution to an interval and
renormalises what remains. It differs from the zero wrappers in a way that
shapes its entire theory: **truncation adds no parameter**. The endpoints are
known constants, on the same footing as a binomial's number of trials. What
truncation adds instead is a normalising constant that depends on $\theta$, and
differentiating that constant is all the work there is.

::: {#def-truncated}
## Truncated distribution

Let $D$ have density or mass function $f(\cdot\,;\theta)$ and support $S$, and
let $L < U$ be truncation points with $T = S \cap [L, U]$ non-empty.
Writing

$$Z(\theta) = P(L \leq Y \leq U) = F(U;\theta) - F(L^-;\theta), \qquad
F(L^-) = F(L) - P(Y = L),$$

the **truncated** version of $D$ is

$$f_T(y;\theta) = \frac{f(y;\theta)}{Z(\theta)} \ \ (y \in T),
\qquad 0 \text{ otherwise.}$$
:::

Both endpoints are included in the truncated support. For a continuous parent
this is a distinction without a difference, since $P(Y = L) = 0$; for a
discrete parent it is the whole point — truncating a Poisson below at
$L = 1$ produces the zero-truncated Poisson on $\{1, 2, \dots\}$, and the
mass sitting exactly at the included endpoint must be added back into $Z$,
whence the correction $F(L^-) = F(L) - f(L)$.

There is a subtlety here that repays attention. The correction "subtract the
mass at $L$" is naturally read as belonging to the discrete case, but what
it really belongs to is the case where the parent places an **atom** at $L$
— and @sec-za-continuous has just manufactured continuous distributions that do
exactly that. The cdf of a zero-adjusted Gamma already contains the point mass
at zero, so for that parent $F(0) \neq F(0^-)$ even though the object is
formally continuous, and a truncation that retained the atom but used $F(0)$ in
$Z$ would lose precisely that mass from the normalising constant. The
implementation therefore obtains the mass at $L$ from `distrib_atoms()`,
uniformly for every kind of parent.

The distribution functions follow the same pattern as before. For $q \in T$,

$$F_T(q) = \frac{F(q;\theta) - F(L^-;\theta)}{Z(\theta)},$$

clamped to $[0,1]$ outside, and solving $F_T(q) = p$ for $q$ gives the quantile
function by delegation to the parent,

$$Q_T(p) = Q\!\left(F(L^-;\theta) + p\,Z(\theta)\right),$$

a relation that holds verbatim for the generalized inverse of a lattice cdf, so
the discrete case needs no separate treatment. Generation is inverse transform
through this formula, $Y = Q(F(L^-) + VZ)$ with $V$ uniform on $(0,1)$ — exact, and
guaranteed to finish in one pass no matter how small $Z$ is, which is where
rejection sampling would stall.

#### Derivatives {#sec-trunc-derivatives}

::: {#thm-truncated}
## Derivatives of a truncated log-likelihood

Let $\ell_T = \ell - \log Z$, and define, with all expectations taken under the
**truncated** distribution itself,

$$m_i = \mathbb{E}_T[\ell^{(i)}], \qquad
M_{ij} = \mathbb{E}_T\!\left[\ell^{(ij)} + \ell^{(i)}\ell^{(j)}\right].$$

Then

$$
\frac{\partial \ell_T}{\partial\theta_i} = \ell^{(i)}(y) - m_i,
\qquad
\frac{\partial^{2}\ell_T}{\partial\theta_i\partial\theta_j}
= \ell^{(ij)}(y) - M_{ij} + m_im_j,
\qquad
\mathbb{E}\!\left[\frac{\partial^{2}\ell_T}{\partial\theta_i\partial\theta_j}\right]
= -\operatorname{Cov}_T\!\left(\ell^{(i)}, \ell^{(j)}\right).
$$
:::

::: {.proof}
Everything reduces to differentiating $\log Z$. The domain of the integral
defining $Z$ is the fixed set $T$ — the endpoints are constants, and the
parent's support does not move with $\theta$ — and this is what licenses
differentiation under the integral sign; it is also why truncation at fixed
points is a regular problem, where truncation at a $\theta$-dependent point
would not be.

Differentiating once, and using $\partial_i f = f\,\ell^{(i)}$,

$$\frac{\partial Z}{\partial\theta_i}
= \int_T \partial_i f
= \int_T f\,\ell^{(i)}
= Z \int_T \frac{f}{Z}\,\ell^{(i)}
= Z\,\mathbb{E}_T[\ell^{(i)}] = Z\,m_i,$$

where the second-to-last step divides and multiplies by $Z$ so as to recognise
$f/Z = f_T$ inside the integral. Hence $\partial_i \log Z = m_i$, and the score
of the truncated model is the parent's score minus its truncated mean.

Differentiating $Z$ a second time, the same manoeuvre together with @lem-fij
gives

$$\frac{\partial^{2} Z}{\partial\theta_i\partial\theta_j}
= \int_T \partial_{ij} f
= \int_T f\left(\ell^{(ij)} + \ell^{(i)}\ell^{(j)}\right)
= Z\,M_{ij},$$

and then, by the quotient rule applied to $\partial_i \log Z = \partial_i Z/Z$,

$$\frac{\partial^{2}\log Z}{\partial\theta_i\partial\theta_j}
= \frac{\partial_{ij}Z}{Z} - \frac{\partial_i Z\,\partial_j Z}{Z^{2}}
= M_{ij} - m_im_j,$$

which yields the observed Hessian as stated.

For the expected Hessian, take $\mathbb{E}_T$ of the observed one. The terms
$M_{ij}$ and $m_im_j$ do not involve $y$, so they pass through the expectation
unchanged, and

$$\mathbb{E}_T\!\left[\partial_{ij}\ell_T\right]
= \mathbb{E}_T[\ell^{(ij)}] - M_{ij} + m_im_j .$$

Substituting the definition
$M_{ij} = \mathbb{E}_T[\ell^{(ij)}] + \mathbb{E}_T[\ell^{(i)}\ell^{(j)}]$, the two occurrences of
$\mathbb{E}_T[\ell^{(ij)}]$ cancel — a cancellation worth savouring, since it
removes the parent's expected Hessian from the formula entirely — and what
remains is

$$-\left(\mathbb{E}_T[\ell^{(i)}\ell^{(j)}] - m_im_j\right) = -\operatorname{Cov}_T(\ell^{(i)}, \ell^{(j)}). \qquad\square$$
:::

The three formulas invite three readings. The score of the truncated model is
the parent's score *recentred* at its mean over the retained region, so the
first Bartlett identity holds by construction. The expected information is the
*covariance matrix of the parent's score* under truncation — which is the
second Bartlett identity making an appearance from the other direction, and
which guarantees positive semi-definiteness on sight. And none of the
ingredients $m_i$, $M_{ij}$ has a closed form for a general parent: they are
computed by quadrature or summation through `expectation()`, which makes the
derivatives of a truncated distribution markedly more expensive than the
parent's own, with third and fourth orders falling back to finite differences
of the analytical Hessian.

Inside $[L, U]$ the constant $\log Z$ does not depend on $y$, so the
response derivatives are the parent's, unchanged; outside, the log-density is
$-\infty$ and they are `NaN`.

#### Which truncations make sense {#sec-trunc-validity}

The constructor `truncated(distrib, lower, upper)` accepts either endpoint or
both, and it examines what it is given, because many syntactically valid calls
are statistically meaningless. The guiding example: what should
`truncated(gamma_distrib(), lower = -2)` do? The Gamma lives on $(0,\infty)$,
so truncating at $-2$ removes no mass whatsoever and the "truncated"
distribution would be the Gamma itself. A user who writes such a call has
misunderstood something, and returning the parent silently would bury the
misunderstanding; the constructor refuses instead, and explains.

```{r}
#| label: trunc-refusals
#| error: true
truncated(gamma_distrib(), lower = -2)
```

The same reasoning rejects a non-integer endpoint for a discrete parent, which
is ambiguous between the two neighbouring lattice points:

```{r}
#| label: trunc-refusals-2
#| error: true
truncated(poisson_distrib(), lower = 1.5)
```

A discrete truncation must also leave enough support to identify the
parameters. A distribution on $k$ points has $k-1$ free probabilities, so a
parent with $p$ parameters needs at least $p+1$ retained points — and since
truncation adds no parameter of its own, this is the only identifiability
condition it ever raises:

```{r}
#| label: trunc-refusals-3
#| error: true
truncated(negbin_distrib(), 3, 4)
```

One further refusal connects this section to the previous ones.

::: {#prp-trunc-zero}
## Truncating away the zero of a zero wrapper destroys its parameter

Let $D_\zeta$ be the zero-inflated version of $D$ and suppose
$0 \notin [L, U]$. Then the truncated law of $D_\zeta$ does not depend on
$\zeta$ at all.
:::

::: {.proof}
For $y \in [L, U]$, which excludes zero, the mass function of $D_\zeta$ is
$(1-\zeta)f(y)$, and by the same token the retained mass is
$Z = (1-\zeta)\big(F(U) - F(L^-)\big)$, where $F$ is the parent's cdf. In
the ratio $f_T = (1-\zeta)f(y)/Z$ the factor $(1-\zeta)$ cancels, leaving
$f(y)/(F(U) - F(L^-))$, which is free of $\zeta$. The score with respect to
$\zeta$ is therefore identically zero. The argument for the two zero-adjusted
forms is the same with $(1-\pi)$ in place of $(1-\zeta)$. $\square$
:::

A parameter with an identically zero score is invisible to the likelihood, and
an optimiser asked to estimate it will wander along a flat ridge; the
constructor refuses the construction, while allowing truncations that keep zero
in the interval, in which case the atom passes through, rescaled by $1/Z$:

```{r}
#| label: trunc-zero-wrapper
#| error: true
truncated(zero_inflated(poisson_distrib()), lower = 1)
```

```{r}
#| label: trunc-zero-wrapper-ok
tz <- truncated(zero_adjusted(gamma_distrib()), upper = 5)
th <- list(mu = 3, sigma2 = 2, za = 0.3)
distrib_atoms(tz, th)
0.3 / distrib_cdf(zero_adjusted(gamma_distrib()), 5, th)
```

Finally, truncating an already truncated distribution is permitted and is
collapsed into a single object over the intersection of the two intervals —
nesting would be mathematically harmless but would pay the quadrature cost of
$m$ and $M$ twice.

```{r}
#| label: trunc-nested
d <- truncated(truncated(gaussian_distrib(), -1, 3), upper = 2)
c(d@lower, d@upper)
```

To close the section, the two truncated families with textbook closed forms,
recovered by the general machinery:

```{r}
#| label: trunc-closed-forms
# zero-truncated Poisson: P(Y = y) = f(y)/(1 - f(0)), E[Y] = mu/(1 - exp(-mu))
ztp <- truncated(poisson_distrib(), lower = 1)
distrib_pdf(ztp, 1:4, list(mu = 2))
dpois(1:4, 2) / (1 - dpois(0, 2))
c(mean(ztp, list(mu = 2)), 2 / (1 - exp(-2)))

# truncated Gaussian on [-1, 2]: E[Y] = mu + sigma (phi(a) - phi(b))/Z
tn <- truncated(gaussian_distrib(), lower = -1, upper = 2)
a <- (-1 - 0.5) / 1.5; b <- (2 - 0.5) / 1.5; Z <- pnorm(b) - pnorm(a)
c(mean(tn, list(mu = 0.5, sigma = 1.5)), 0.5 + 1.5 * (dnorm(a) - dnorm(b)) / Z)
```

### The higher orders, all at once {#sec-higher-orders}

Everything derived so far has stopped at the second derivative, which is where
a fitting routine stops. The toolkit promises four orders, though, and the
promise is worth keeping for the wrappers as well as for the distributions they
wrap. What makes that cheap is a structural observation: **every wrapper's
log-likelihood has the same shape**. Setting the branches side by side,

$$
\begin{aligned}
\text{zero-inflated, } y = 0:\quad & \ell_Y = \log L_0, \qquad L_0 = \zeta + (1-\zeta)f_0,\\
\text{hurdle, } y > 0:\quad & \ell_Y = \ell(y) - \log(1-f_0) + \log(1-\pi),\\
\text{zero-adjusted continuous, } y \neq 0:\quad & \ell_Y = \ell_W(y) + \log(1-\pi),\\
\text{truncated}:\quad & \ell_T = \ell(y) - \log Z,
\end{aligned}
$$

each is the parent's log-density plus, or instead of, **the logarithm of some
quantity that depends on $\theta$**. So one derivation covers all four, provided
we can differentiate a logarithm to arbitrary order — and we can, using nothing
that has not already appeared in this book.

#### Two facts, both sums over partitions {#sec-two-facts}

The first is @lem-exp-deriv, read in the direction opposite to the one
@sec-bartlett used it. There it turned $\partial^k f$ into something whose
expectation vanishes; here it simply says how to get the derivatives of a
density from those of its logarithm:

$$\frac{\partial^{I} f}{f} \;=\; \sum_{\pi \in \Pi_I} \prod_{B \in \pi} \ell^{(B)},$$

the complete Bell polynomial in the parent's derivatives, where $\Pi_I$ runs
over the set partitions of the multi-index $I$.

The second is its inverse, and is the classical relation between moments and
cumulants.

::: {#prp-log-deriv}
## Derivatives of a logarithm

For a positive function $L$ of the parameters and a multi-index $I$,

$$\partial^{I} \log L \;=\; \sum_{\pi \in \Pi_I} (-1)^{|\pi|-1}\,(|\pi|-1)!
   \prod_{B \in \pi} \frac{\partial^{B} L}{L}.$$
:::

::: {.proof}
Both sides are determined by their values on the algebra of partitions, so it
suffices to check that the two expansions invert one another. Substituting the
first fact into the second with $L = f$ gives, on the right, a sum over pairs
$(\pi, \sigma)$ with $\sigma$ refining $\pi$, weighted by
$(-1)^{|\pi|-1}(|\pi|-1)!$. For a fixed $\sigma$ the inner sum over the $\pi$
it refines is $\sum_{\pi \geq \sigma}(-1)^{|\pi|-1}(|\pi|-1)!$, which is the
Möbius function of the partition lattice and vanishes unless $\sigma$ is the
single-block partition. What survives is $\ell^{(I)}$, which is what
$\partial^I \log f$ should be. $\square$
:::

Only the *ratios* $\partial^B L / L$ are ever needed, never $L$'s derivatives
on their own — which matters because for the truncated wrapper those ratios are
exactly the truncated expectations already met in @thm-truncated:

$$\frac{\partial^{B} Z}{Z} = \mathbb{E}_T\!\left[\frac{\partial^{B}f}{f}\right].$$

At second order this is $M_{ij}$, and @prp-log-deriv reduces to
$\partial_{ij}\log Z = M_{ij} - m_i m_j$ — the formula derived by hand earlier.
The same collapse happens everywhere: putting $|I| = 1$ or $2$ into the two
facts reproduces every score and Hessian in this chapter, including the
$w_0(1-w_0)$ of @prp-zi-hessian and the truncation correction $C_{ij}$ of
@prp-hurdle-hessian. The general machinery and the hand derivations are two
routes to the same formulas, and each is a check on the other.

#### What each wrapper needs {#sec-higher-per-wrapper}

With the two facts in hand, each wrapper is a matter of naming its $L$ and its
ratios.

For a **transformed** distribution there is nothing to do at all: by
@thm-transform-invariance the derivatives are the parent's at $x = g^{-1}(y)$,
at every order.

For **zero-inflation** at $y = 0$ we have $L = L_0$, which is *affine* in
$\zeta$ — a fact that does most of the work, since it kills every block
containing two or more $\zeta$'s. Writing $A$ for the $\theta$-part of a block
and counting its $\zeta$'s,

$$
\frac{\partial^{B}L_0}{L_0} =
\begin{cases}
w_0 \cdot \dfrac{\partial^{A} f_0}{f_0} & \text{no } \zeta,\\[8pt]
-f_0\,\dfrac{\partial^{A} f_0}{f_0}\Big/L_0 & \text{one } \zeta,\ A \neq \emptyset,\\[8pt]
(1-f_0)/L_0 & \text{one } \zeta,\ A = \emptyset,\\[4pt]
0 & \text{two or more } \zeta.
\end{cases}
$$

Above zero the likelihood separates, so mixed components vanish and the pure
ones are the parent's and $\partial^k\log(1-\zeta) = -(k-1)!/(1-\zeta)^k$.

For the **hurdle** the separation is complete at every order, so all mixed
components are identically zero; the $\theta$ part above zero is the parent's
derivative minus that of $\log(1-f_0)$, whose ratios are
$-f_0(\partial^B f_0/f_0)/(1-f_0)$. The **zero-adjusted continuous** case is the
same without the truncation constant. And **truncation** is $\ell$ minus
$\log Z$ with the expectations above.

As a small closed form to end on, the pure $\zeta$ derivatives at zero: since
$\ell_Y = \log\big(f_0 + \zeta(1-f_0)\big)$ is a logarithm of an affine
function,

$$\frac{\partial^{k}\ell_Y}{\partial\zeta^{k}}\bigg|_{y=0}
= (-1)^{k-1}(k-1)!\left(\frac{1-f_0}{L_0}\right)^{k}.$$

```{r}
#| label: zi-pure-zeta
d <- zero_inflated(poisson_distrib())
f0 <- dpois(0, 3); L0 <- 0.25 + 0.75 * f0; r <- (1 - f0) / L0
c(package = distrib_deriv3(d, 0, list(mu = 3, zi = 0.25))[["zi_zi_zi"]],
  closed  = 2 * r^3)
```

#### What it costs {#sec-higher-cost}

For the zero wrappers, nothing to speak of: the ratios are algebraic in $f_0$
and the parent's derivatives, and the partition sums have at most fifteen terms
at fourth order. For truncation each *distinct block* costs one quadrature or
summation, so a fourth-order component with two parameters needs a handful —
memoised across the partition sum, since a partition of four indices asks for
the same small blocks repeatedly. It is the same trade the second order already
made, and the reason the closed forms of a future `distrib_grad_cdf` would pay
for themselves.

### Choosing among the constructions {#sec-choosing}

The three families answer different questions, and the differences are easiest
to see side by side.

| | Zero-inflation | Zero-adjustment (hurdle) | Truncation |
|---|---|---|---|
| Effect on $P(Y=0)$ | *adds*: $\zeta + (1-\zeta)f_0 > f_0$ | *replaces*: $\pi$, free in $(0,1)$ | removes it, if $0 \notin [L,U]$ |
| Parameters added | one, $\zeta$ | one, $\pi$ | none |
| Fewer zeros than the parent? | impossible | possible | only none at all |
| Meaning of $\theta$ | the original count process | the zero-truncated law | the restricted law |
| Likelihood | does not separate | binary $\times$ truncated | $\ell - \log Z(\theta)$ |
| Discrete parent required | yes | no | no |

: The three constructions {.striped}

The choice between the first two is a choice about the zeros themselves. When a
zero is observable evidence of a distinct decision — no claim was filed, no
purchase was made, no rain fell — the zeros and the positives genuinely come
from two mechanisms, and the hurdle, whose likelihood separates along exactly
that line, is the natural model. When the two kinds of zero cannot be told
apart, and $\theta$ is meant to keep describing the count process that the
ordinary observations come from, zero-inflation is the natural model instead.
The two are not nested, and where both fit similarly they still differ in what
their $\theta$ *means*.

Truncation answers a different question altogether. It is not a model for
surplus or deficient zeros but for **unobservable regions**: a detection limit,
a register that only records positive counts, a measuring instrument with a
ceiling. Part of the sample space was never available, and the model says so.

One boundary between the families is worth marking explicitly, because the
package enforces it. For a continuous parent, "zero-inflation" and
"zero-adjustment" would denote the same object — with $P(W=0)=0$ the mixture
$\zeta\delta_0 + (1-\zeta)f_W$ *is* @def-za-continuous with $\zeta$ in the role
of $\pi$ — and offering both names would only invite the belief that they
differ. The constructor therefore accepts a continuous parent only under the
name `zero_adjusted()`, and points there when asked otherwise:

```{r}
#| label: zi-continuous-refused
#| error: true
zero_inflated(gamma_distrib())
```

### Identifiability {#sec-identifiability}

The zero wrappers add a parameter, and adding a parameter to a family does not
always produce an identified family. What makes this dangerous is that nothing
at run time betrays the failure: the mass function still sums to one, every
numerical check still passes, and an optimiser still converges — to an
arbitrary point of a flat ridge, with standard errors to match. The only place
the failure can be caught is the constructor, so the constructors carry the
mathematics needed to catch it. This section derives that mathematics.

#### Wrappers do not stack {#sec-no-stacking}

::: {#prp-no-stack-1}
## Zero-truncation annihilates zero-inflation

Let $D_\zeta$ be the zero-inflated version of $D$. Then the zero-truncated law
of $D_\zeta$ coincides with the zero-truncated law of $D$ itself, whatever the
value of $\zeta$.
:::

::: {.proof}
For $y > 0$, the mass function of $D_\zeta$ is $(1-\zeta)f(y)$ and its total
mass on the positive integers is
$1 - L_0 = 1 - \zeta - (1-\zeta)f_0 = (1-\zeta)(1-f_0)$. Dividing,

$$
\frac{(1-\zeta)f(y)}{(1-\zeta)(1-f_0)} = \frac{f(y)}{1-f_0},
$$

and $\zeta$ has cancelled. $\square$
:::

The consequence is immediate: in `zero_adjusted(zero_inflated(D))` the
parameter $\zeta$ appears nowhere in the resulting mass function, its score is
identically zero, and it cannot be estimated. The reverse composition fails
similarly — in `zero_inflated(zero_adjusted(D))` the pair $(\zeta, \pi)$ enters
the law only through the combination $c = \zeta + (1-\zeta)\pi$, since the mass
at zero is $c$ and the positive part carries the factor
$(1-\zeta)(1-\pi) = 1 - c$, so every pair on the curve
$\zeta + (1-\zeta)\pi = c$ produces exactly the same distribution. Both
constructions are therefore refused:

```{r}
#| label: stacking-demo
try(zero_adjusted(zero_inflated(poisson_distrib())))
try(zero_inflated(zero_adjusted(poisson_distrib())))
```

That these refusals guard against something real, and not a theoretical
scruple, can be seen by building the forbidden object by hand, bypassing the
constructor, and watching the parameter vanish:

```{r}
#| label: stacking-evidence
#| echo: false
zip <- zero_inflated(poisson_distrib())
stacked <- distributions7:::ZeroAdjustedDiscreteDistrib(
  parent_distrib = zip, distrib_name = "hand-built hurdle over a ZIP",
  dimension = "univariate", bounds = c(0, Inf),
  params = c("mu", "zi", "za"),
  params_interpretation = c(mu = "mean", zi = "structural zero", za = "prob. of zero"),
  n_params = 3,
  params_bounds = list(mu = c(0, Inf), zi = c(0, 1), za = c(0, 1)),
  link_params = list(mu = log_link(), zi = logit_link(), za = logit_link()))
```

```{r}
#| label: stacking-evidence-show
distrib_pdf(stacked, 0:5, list(mu = 3, zi = 0.20, za = 0.4))
distrib_pdf(stacked, 0:5, list(mu = 3, zi = 0.70, za = 0.4))
distrib_gradient(stacked, 0:5, list(mu = 3, zi = 0.20, za = 0.4))$zi
```

#### Counting the support {#sec-counting-rule}

::: {#prp-counting}
## The counting rule

A distribution supported on $k$ points is an element of the
$(k-1)$-dimensional probability simplex, so a family with $q$ free parameters
can be identified only if $q \leq k-1$. Both zero wrappers have $q = p+1$
parameters, where $p$ is the number of the parent's; hence at least

$$k \geq p + 2$$

support points are necessary for either wrapper to be identified.
:::

The condition is necessary but not sufficient — a badly parametrised family can
fail to be identified with any amount of support — yet it is exactly what rules
out the case that arises in practice: the Bernoulli, with $k = 2$ and $p = 1$.
Zero-inflating a Bernoulli spends two parameters describing the single free
cell of $\{0, 1\}$. Zero-adjusting one is even more striking: the truncation
concentrates everything on $\{1\}$, so
$f(1)/(1-f_0) = \mu/(1-(1-\mu)) = 1$ whatever $\mu$ is, and the parent's
parameter drops out of the model completely, leaving
$P(Y=0) = \pi,\ P(Y=1) = 1-\pi$.

```{r}
#| label: bernoulli-refused
#| error: true
zero_inflated(bernoulli_distrib())
```

```{r}
#| label: bernoulli-refused-2
#| error: true
zero_adjusted(binomial_distrib(size = 1))
```

The same rule admits a binomial with two trials, whose three support points are
enough for the two parameters:

```{r}
#| label: binomial-two-ok
zero_inflated(binomial_distrib(size = 2))@params
```

A caveat to end on: the counting rule protects against structural
non-identification, the kind that no amount of data can cure. Weak
identification of the statistical kind — a Poisson mean so large that $f_0$
underflows, a $\zeta$ so close to zero that the data barely see it — remains
possible in any mixture model, and there it is the standard errors from
`fit_distrib()` that tell the story.

```{r}
#| label: consistency-transformations
#| include: false
# Internal consistency gate: fails the render if any density, score, Hessian,
# expected Hessian, moment or closed form derived in this chapter disagrees
# with the implementation, or if check_distrib() fails on any wrapper.
assert_transformations_ok()
```
