## Transformations of distributions {#sec-transformations}

```{r}
#| label: setup-04
#| include: false
source("R/_setup.R")
source("R/link-formulas.R")
source("R/distribution-catalog.R")
source("R/transformation-certificates.R")
```

The chapters so far have dealt with fixed objects: a Gamma score is a Gamma
score, written once and used everywhere. The present section concerns **operators on distributions**. The constructor `zero_inflated(D)` is not itself a distribution. It is a rule
that turns any suitable $D$ into a new distribution, and the score of the
result is assembled at run time from $D$'s score, $D$'s Hessian and $D$'s
density at zero, whatever $D$ happens to be. The formulas such an operator uses must therefore be derived in full
generality, once, for an arbitrary parent, and that is what this section
does. For the change of variables, zero-inflation, the hurdle, the
zero-adjusted continuous family and truncation, every density, score,
Hessian, expected Hessian and moment is stated for an arbitrary parent.

A small amount of shared notation keeps the derivations light. The parent
distribution has density or mass function $f(y;\theta)$ with log-density
$\ell(y) = \log f(y;\theta)$, and we write

$$
\ell^{(i)}(y) = \frac{\partial \ell(y)}{\partial\theta_i}, \qquad
\ell^{(ij)}(y) = \frac{\partial^{2} \ell(y)}{\partial\theta_i\partial\theta_j}, \qquad
f_0 = f(0;\theta)
$$

for its score, its Hessian and its mass at zero, following the convention of
@sec-likelihood throughout: a parenthesized superscript on $\ell$ is a
derivative with respect to the parameters it names. One elementary identity will
be needed again and again, so we record it once.

::: {#lem-fij}
## Second derivative of a density through its log

$$\frac{1}{f}\,\partial_{ij} f = \ell^{(ij)} + \ell^{(i)} \ell^{(j)},$$

where $\partial_{ij}$ abbreviates $\partial^{2}/\partial\theta_i\partial\theta_j$.
:::

It follows in one line from $\partial_i f = f\,\ell^{(i)}$, differentiated once
more as a product. In the language of @sec-bartlett it is the order-two case of
the partition expansion of $\partial^k f / f$, whose two partitions of a
two-element set are the single block $\{ij\}$ and the pair of singletons.

### Change of variables {#sec-transformation}

The oldest of the operations is the classical one: given a random variable $X$
with a known continuous distribution, describe the distribution of $Y = g(X)$.

::: {#def-transformer}
## Transformer

A **transformer** is a map $g : S \to T$, defined on the support $S$ of the
parent, which is a strictly monotone, continuously differentiable bijection
with $g' \neq 0$ throughout. In the package it is described by $g$, by its
inverse $g^{-1}$, by the absolute Jacobian of the inverse
$\lvert J(y)\rvert = \lvert\,\mathrm{d}g^{-1}/\mathrm{d}y\,\rvert$, by its effect on the support
bounds, and by a flag recording whether it is decreasing. Crucially, **a transformer does not depend on $\theta$**; most of the results
below rest on this property.
:::

#### The density {#sec-transform-density}

::: {#thm-cov}
## Change of variables

If $Y = g(X)$ with $g$ a transformer, then

$$f_Y(y;\theta) = f_X\!\left(g^{-1}(y);\theta\right)\,\left\lvert J(y)\right\rvert,
\qquad y \in T .$$
:::

This is the classical change-of-variables formula, and the absolute value is
what reconciles its two cases. For increasing $g$ the event $g(X) \le y$ is the
event $X \le g^{-1}(y)$, so $F_Y = F_X \circ g^{-1}$ and differentiation gives
the density with a positive Jacobian. For decreasing $g$ the inequality flips,
$F_Y = 1 - F_X \circ g^{-1}$, and differentiation produces a leading minus sign
that exactly cancels the now-negative Jacobian. A textbook statement is in
@casellaberger2002 or @johnson1994.

The same two computations give the cdf, and by inversion the quantile
function and the generator. In the increasing case
$F_Y = F_X \circ g^{-1}$ and $Q_Y(p) = g(Q_X(p))$. In the decreasing case the
tails swap: $F_Y = 1 - F_X \circ g^{-1}$ and $Q_Y(p) = g(Q_X(1-p))$.
Generating a draw is simply a matter of drawing $X$ from the parent and
applying $g$, which is exact and costs a single evaluation. The decreasing case occurs in practice: the reciprocal transformer and the
power transformer with a negative exponent are both decreasing. A forgotten
tail swap produces a quantile function that is wrong by $p \mapsto 1-p$, an
error that is invisible at the median and largest in the tails.

#### Invariance of the parameter derivatives {#sec-transform-invariance}

The next result is the reason this wrapper costs essentially nothing.

::: {#thm-transform-invariance}
## Invariance of the likelihood derivatives

Let $Y = g(X)$ with $g$ a transformer. Then for every order $k \geq 1$ and
every multi-index,

$$
\frac{\partial^{k}\,\ell_Y(\theta;y)}{\partial\theta_{i_1}\cdots\partial\theta_{i_k}}
= \left.\frac{\partial^{k}\,\ell_X(\theta;x)}{\partial\theta_{i_1}\cdots\partial\theta_{i_k}}
\right|_{x = g^{-1}(y)},
$$

and the expected derivatives of the two models coincide exactly:
$\mathbb{E}_Y\!\left[\partial^{k}\ell_Y\right] = \mathbb{E}_X\!\left[\partial^{k}\ell_X\right]$.
:::

Everything here follows from the log-density of the transformed variable,

$$\ell_Y(\theta;y) = \ell_X\!\left(\theta;\, g^{-1}(y)\right) + \log\lvert J(y)\rvert ,$$

considered as a function of $\theta$. The Jacobian term depends on $y$ alone; this is where the requirement that a
transformer not depend on $\theta$ is used. It is therefore a constant for
the purpose of differentiating in $\theta$, and it disappears at the first
derivative. The point $x = g^{-1}(y)$ at which the first term is evaluated
also does not move when $\theta$ does, so differentiation passes through the
substitution. This proves the claim at every order at
once. The statement about expectations is then the change-of-variables
formula for expectations applied to $g^{-1}(Y) = X$, which holds with probability
one.

The theorem has two practical consequences, and both save work. First, no simulation is ever
needed for the expected information of a transformed distribution: the parent's
expected Hessian already *is* the answer, exactly, and the implementation is a
single line. Second, the link-scale machinery of @sec-link-scale is unaffected,
because the links attach to the parameters and the parameters have not changed.

The Jacobian, having dropped out of the $\theta$-derivatives, has not left the
theory altogether: it survives in the density itself and in the derivatives
with respect to $y$. This is why a transformer also carries
$\partial_y \log\lvert J\rvert$ and $\partial^2_y \log\lvert J\rvert$, which feed the response
derivatives of @sec-response-derivatives.

#### The twelve transformers {#sec-transformer-catalog}

```{r}
#| label: transformer-table
#| echo: false
tt <- do.call(rbind, lapply(TRANSFORMERS, function(tr) data.frame(
  transformer = sprintf("`%s`", tr$ctor),
  `$g(x)$` = sprintf("$%s$", tr$g),
  `$g^{-1}(y)$` = sprintf("$%s$", tr$ginv),
  `$\\lvert J(y)\\rvert$` = sprintf("$%s$", tr$jac),
  `domain` = sprintf("$%s$", tr$domain),
  check.names = FALSE, stringsAsFactors = FALSE)))
knitr::kable(tt, align = "lllll", row.names = FALSE,
             caption = "The twelve transformers")
```

Two of these deserve a word of attribution, since they were designed for exactly
the purpose the wrapper puts them to: the power transformation of @box1964, and
the extension of @yeo2000 which accepts negative values as well. Each transformer
declares which supports it is willing to act on, and the constructor refuses an
incompatible pairing at once, rather than letting it
surface later as a `NaN`:

```{r}
#| label: transformer-validity
#| error: true
transformation(gaussian1_distrib(), log_transform())
```

As a closing illustration, the lognormal distribution can be reached along two
entirely different roads — as a catalog entry with its own density, or as the
exponential of a Gaussian through @thm-cov — and the two agree to machine
precision:

```{r}
#| label: lognormal-identity
y <- seq(0.05, 12, length.out = 40)
a <- distrib_pdf(transformation(gaussian1_distrib(), exp_transform()),
                 y, list(mu = 0.5, sigma = sqrt(1.3)), log = TRUE)
b <- distrib_pdf(lognormal1_distrib(), y, list(mu = 0.5, sigma2 = 1.3), log = TRUE)
max(abs(a - b))
```

#### Families that need no code of their own {#sec-transform-reach}

The lognormal identity above is one instance of something more general. A
number of families that a table of distributions lists separately are the
image of one already implemented, and the wrappers reach them without any
derivation being repeated: the score, the information and the higher orders
are the parent's, evaluated at $g^{-1}(y)$, because $g$ carries no parameter.

The inverse of a Gamma is an inverse Gamma. Writing the Gamma in
$(\mu, \sigma^2)$ as this package does, the shape and rate are
$\alpha = \mu^2/\sigma^2$ and $\beta = \mu/\sigma^2$, and the density of
$1/Y$ is the standard $\beta^{\alpha}y^{-\alpha-1}e^{-\beta/y}/\Gamma(\alpha)$.

```{r}
#| label: inverse-gamma
a <- 4; b <- 2
ig <- transformation(gamma2_distrib(), inverse_transform(),
                     new_name = "inverse gamma")
y <- c(0.3, 0.8, 1.5)
max(abs(distrib_pdf(ig, y, list(mu = a / b, sigma2 = a / b^2)) -
        b^a / gamma(a) * y^(-a - 1) * exp(-b / y)))
```

The `new_name` argument is what lets the result print as what it is rather
than as the recipe that produced it; nothing else about the object depends on
it.

Holding a parameter reaches a second group. The Rayleigh distribution is the
Weibull with shape two, and since that shape is a *value* the parameter can be
held at, `fixed()` produces it exactly. Writing $s$ for the Rayleigh scale, the
Weibull scale is $s\sqrt{2}$.

```{r}
#| label: rayleigh
s <- 1.7
ray <- fixed(weibull1_distrib(), sigma = 2)
y <- c(0.3, 1, 2, 4)
max(abs(distrib_pdf(ray, y, list(mu = s * sqrt(2))) -
        (y / s^2) * exp(-y^2 / (2 * s^2))))
```

The distinction between the two groups is worth stating, because it decides
whether a family needs writing at all. `fixed()` can hold a parameter at a
number; it cannot impose a *relation between* parameters. The exponential is
the Gamma with unit shape, and unit shape means $\sigma^2 = \mu^2$, which ties
two parameters together rather than pinning one — so the exponential is a
family of its own here, as is the chi-squared, where the relation is
$\sigma^2 = 2\mu$. The exponential *is* reachable as the Weibull with shape
one, and `exponential_distrib()` agrees with that wrapper to machine
precision, which makes the pair a test of both.

The same reasoning places the other standard names. The log-logistic is
$\exp$ of a logistic, the Pareto is $\exp$ of an exponential, and the folded
normal is `folded(gaussian1_distrib())` with the half-normal its location held
at zero — the last through a wrapper rather than a transformer, the absolute
value having two preimages and therefore no inverse to carry a density
through.

### Zero-inflation {#sec-zero-inflation}

Count data very often contain more zeros than any standard count distribution
can produce, and the classical remedy, introduced by @lambert1992, is to let the
zeros come from two sources: the count process itself, and a separate mechanism
that switches the count off entirely.

::: {#def-zi}
## Zero-inflated distribution

Let $D$ be a discrete distribution with $0$ in its support and mass function
$f(y;\theta)$, and let $\zeta \in (0,1)$. The **zero-inflated** version of $D$
is the mixture

$$
P(Y = y;\theta,\zeta) = \zeta\,\mathbb{I}(y=0) + (1-\zeta) f(y;\theta)
= \begin{cases}
\zeta + (1-\zeta) f_0 & y = 0,\\[2pt]
(1-\zeta) f(y;\theta) & y > 0 .
\end{cases}
$$
:::

Summing over the support gives $\zeta + (1-\zeta)\sum_y f(y) = 1$, so the
mixture is a proper distribution. Its interpretation determines every formula
below. An observed zero may be *structural*, produced by the switching
mechanism with probability $\zeta$, or an ordinary *sampling* zero from the
count process. No individual zero can be attributed to either source;
only their proportions are identified. Two derived quantities recur
throughout. The first is the total mass at zero,

$$L_0 = \zeta + (1-\zeta)f_0,$$

and the second is the conditional probability that an observed zero came from
the count process,

$$w_0 = \frac{(1-\zeta)f_0}{L_0} \in (0,1).$$

The quantity $w_0$ is the E-step weight that the EM algorithm attaches to this
model, and it appears throughout the derivatives below.

Before the derivatives, the distribution functions, which follow from the
mixture structure directly. The cdf mixes the parent's cdf with a step at zero,

$$F_Y(q) = (1-\zeta)F(q;\theta) + \zeta\,\mathbb{I}(q \geq 0),$$

and inverting it splits the probability axis at $L_0 = F_Y(0)$: any
$p \leq L_0$ lands in the mass at zero, while for $p > L_0$ we solve
$\zeta + (1-\zeta)F(q) = p$ for $F(q)$ and apply the parent's quantile
function, obtaining

$$
Q_Y(p) = \begin{cases}
0 & p \leq L_0,\\[4pt]
Q\!\left(\dfrac{p - \zeta}{1-\zeta};\theta\right) & p > L_0 .
\end{cases}
$$

Generation is even simpler: draw from the parent, then replace a
$\mathrm{Bernoulli}(\zeta)$ fraction of the draws by structural zeros.

#### The score {#sec-zi-score}

::: {#prp-zi-score}
## Zero-inflated score

$$
\frac{\partial \ell_Y}{\partial \theta_i} =
\begin{cases} w_0\, \ell^{(i)}(0) & y = 0 \\ \ell^{(i)}(y) & y > 0\end{cases}
\qquad\qquad
\frac{\partial \ell_Y}{\partial \zeta} =
\begin{cases} \dfrac{1 - f_0}{L_0} & y = 0 \\[6pt] -\dfrac{1}{1-\zeta} & y > 0 .\end{cases}
$$
:::

Both lines come from differentiating the two branches of the log-likelihood.
Away from zero the log-likelihood is $\log(1-\zeta) + \ell(y)$, in which the
two parameters separate completely. At zero it is $\log L_0$, where they do
not, and the weight $w_0$ appears because
$\partial_i f_0 = f_0 \ell^{(i)}(0)$ leaves the ratio $(1-\zeta)f_0/L_0$ in
front of the parent's score.

In words: away from zero, the score is the
parent's, untouched; at the zeros, the parent's score is *down-weighted* by
$w_0$, the probability that the zero actually came from the count process. An
observation that is almost certainly structural contributes almost nothing to
the estimation of $\theta$, which is the correct weighting.

#### The observed Hessian {#sec-zi-hessian}

::: {#prp-zi-hessian}
## Zero-inflated observed Hessian

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j} =
\begin{cases}
w_0 \ell^{(ij)}(0) + w_0(1-w_0)\,\ell^{(i)}(0)\ell^{(j)}(0) & y = 0\\[4pt]
\ell^{(ij)}(y) & y > 0
\end{cases}
$$

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\zeta} =
\begin{cases} -\dfrac{f_0\,\ell^{(i)}(0)}{L_0^{2}} & y = 0 \\[6pt] 0 & y > 0\end{cases}
\qquad\quad
\frac{\partial^{2}\ell_Y}{\partial\zeta^{2}} =
\begin{cases} -\dfrac{(1-f_0)^{2}}{L_0^{2}} & y = 0 \\[6pt] -\dfrac{1}{(1-\zeta)^{2}} & y > 0 .\end{cases}
$$
:::

The positive branch needs no comment, since there the log-likelihood separates
and each block is the parent's. At zero the three blocks come from differentiating the score of
@prp-zi-score once more, using @lem-fij for the second derivative of $f_0$.
The pure product terms combine into $w_0(1-w_0)$, the variance of the
Bernoulli indicator that the zero came from the count process, which is also
the weight the EM algorithm attaches to it.

One step in that calculation matters for the implementation. The mixed block arrives in the form

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\zeta}
= -\frac{\partial_i f_0}{L_0^{2}}\Big[L_0 + (1-\zeta)(1-f_0)\Big],
$$

and the bracket is exactly $1$, since
$\zeta + (1-\zeta)f_0 + (1-\zeta) - (1-\zeta)f_0 = 1$. That is not a cosmetic
simplification: the unsimplified expression would have to perform the
cancellation numerically at every evaluation.

#### The expected Hessian {#sec-zi-expected}

To average the observed Hessian we split the expectation according to where the
mass sits: the point $\{0\}$ carries probability $L_0$, and each $y > 0$
carries $(1-\zeta)f(y)$, so that for any function $u$,

$$\mathbb{E}[u(Y)] = L_0\,u(0) + (1-\zeta)\sum_{y>0} f(y)\,u(y).$$

::: {#prp-zi-expected}
## Zero-inflated expected Hessian

$$
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j}\right]
= L_0\Big[w_0\ell^{(ij)}(0) + w_0(1-w_0)\ell^{(i)}(0)\ell^{(j)}(0)\Big]
+ (1-\zeta)\Big(\mathbb{E}_f[\ell^{(ij)}] - f_0 \ell^{(ij)}(0)\Big),
$$

$$
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\zeta}\right]
= -\frac{f_0\,\ell^{(i)}(0)}{L_0},
\qquad
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\zeta^{2}}\right]
= -\left[\frac{(1-f_0)^{2}}{L_0} + \frac{1-L_0}{(1-\zeta)^{2}}\right].
$$
:::

Each block is the two branches of @prp-zi-hessian averaged against the mixture,
whose mass is $L_0$ at the point zero and $(1-\zeta)f(y)$ at each $y > 0$. The
only subtlety is bookkeeping: the positive branch sums the parent's Hessian over
the support *without* its zero term, which is why $\mathbb{E}_f[\ell^{(ij)}]$
appears with $f_0\,\ell^{(ij)}(0)$ subtracted from it. The $\zeta\zeta$ block can
be checked independently against the second Bartlett identity of @thm-bartlett,
which requires it to equal $-\mathbb{E}[(\partial_\zeta \ell_Y)^2]$, and it does.

#### Moments {#sec-zi-moments}

Since the extra component sits entirely at zero, it contributes nothing to any
moment of positive order, and for every $k \geq 1$,

$$\mathbb{E}[Y^{k}] = (1-\zeta)\,\mathbb{E}_f[Y^{k}].$$

If the parent has mean $\mu$ and variance $\sigma^{2}$, this gives
$\mathbb{E}[Y] = (1-\zeta)\mu$ at once, and for the variance we compute

$$\operatorname{Var}(Y) = (1-\zeta)\left(\sigma^{2}+\mu^{2}\right) - (1-\zeta)^{2}\mu^{2}
= (1-\zeta)\sigma^{2} + \zeta(1-\zeta)\mu^{2}.$$

For a Poisson parent, where $\sigma^2 = \mu$, the variance
$(1-\zeta)\mu + \zeta(1-\zeta)\mu^2$ strictly exceeds the mean $(1-\zeta)\mu$:
a zero-inflated Poisson is always overdispersed, which is the second reason,
after the zeros themselves, that the model gets reached for.

### Zero-adjustment of a discrete distribution {#sec-hurdle}

Zero-inflation keeps the parent's zeros and adds more. The hurdle model, which
goes back to @mullahy1986 and was developed further by @heilbron1994, takes
the opposite view. It treats the zeros as coming from a mechanism of their
own, removes the parent's mass at zero entirely, and puts a free parameter in
its place.

::: {#def-hurdle}
## Hurdle distribution

Let $D$ be a discrete distribution with $0$ in its support and $f_0 < 1$, and
let $\pi \in (0,1)$. The **zero-adjusted** (hurdle) version of $D$ is

$$
P(Y=y;\theta,\pi) =
\begin{cases}
\pi & y = 0,\\[4pt]
(1-\pi)\dfrac{f(y;\theta)}{1-f_0} & y > 0 .
\end{cases}
$$
:::

The division by $1 - f_0$ is the truncation, and it is the entire difference
between this definition and @def-zi. Since $\sum_{y>0} f(y) = 1 - f_0$ exactly,
the positive part sums to $1-\pi$ and the whole object is a proper
distribution. Two structural consequences are visible before any derivative is
taken. First, $P(Y=0) = \pi$ is completely free in $(0,1)$: where
zero-inflation can only push the mass at zero *above* $f_0$, the hurdle can
place it anywhere, so it also fits data with *fewer* zeros than the parent
would produce. Second, $\theta$ has changed its meaning — it now parametrizes
the zero-truncated law rather than the original count process — and this
matters whenever the estimates are interpreted rather than merely used.

The central property of the model, from which everything else follows, is
that its log-likelihood **separates**. Written out,

$$
\ell_Y(\theta,\pi;y) =
\underbrace{\mathbb{I}(y=0)\log\pi + \mathbb{I}(y>0)\log(1-\pi)}_{\text{a Bernoulli likelihood in } \pi}
\;+\; \mathbb{I}(y>0)\underbrace{\Big[\ell(y) - \log(1-f_0)\Big]}_{\text{a truncated likelihood in } \theta},
$$

and the two groups share no parameter. All mixed $\theta$–$\pi$ derivatives
therefore vanish identically, at every order, observed and expected, and the
two halves of the model could be fitted separately with identical results.

#### The score {#sec-hurdle-score}

::: {#prp-hurdle-score}
## Hurdle score

$$
\frac{\partial\ell_Y}{\partial\theta_i} =
\begin{cases} 0 & y = 0 \\[4pt]
\ell^{(i)}(y) + \dfrac{f_0}{1-f_0}\,\ell^{(i)}(0) & y > 0\end{cases}
\qquad\qquad
\frac{\partial\ell_Y}{\partial\pi} =
\begin{cases} \dfrac{1}{\pi} & y = 0 \\[6pt] -\dfrac{1}{1-\pi} & y > 0 .\end{cases}
$$
:::

Both branches follow from the separated form of the log-likelihood above. The
only term that is not immediate is the truncation constant, whose derivative is

$$-\frac{\partial}{\partial\theta_i}\log(1-f_0)
= \frac{\partial_i f_0}{1-f_0} = \frac{f_0}{1-f_0}\,\ell^{(i)}(0),$$

again using $\partial_i f_0 = f_0\,\ell^{(i)}(0)$.

The extra term is the **truncation correction**, and it has a transparent
reading. Because the law has been renormalized over the positive values,
raising the parent's probability of a zero now *lowers* the probability of
every positive observation, and the score must account for that.

#### The Hessians {#sec-hurdle-hessian}

::: {#prp-hurdle-hessian}
## Hurdle Hessians

Abbreviate

$$
C_{ij} = \frac{(1-f_0)\,\partial_{ij}f_0 + \partial_i f_0\,\partial_j f_0}{(1-f_0)^{2}},
\qquad\text{where}\quad
\partial_i f_0 = f_0\,\ell^{(i)}(0), \quad
\partial_{ij} f_0 = f_0\left(\ell^{(ij)}(0) + \ell^{(i)}(0)\ell^{(j)}(0)\right).
$$

Then the observed Hessian is

$$
\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j} =
\begin{cases} 0 & y=0 \\ \ell^{(ij)}(y) + C_{ij} & y>0\end{cases}
\qquad
\frac{\partial^{2}\ell_Y}{\partial\pi^{2}} =
\begin{cases} -\dfrac{1}{\pi^{2}} & y=0\\[6pt] -\dfrac{1}{(1-\pi)^{2}} & y>0,\end{cases}
$$

with the mixed block identically zero, and in expectation

$$
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j}\right]
= (1-\pi)\left[\frac{\mathbb{E}_f[\ell^{(ij)}] - f_0 \ell^{(ij)}(0)}{1-f_0} + C_{ij}\right],
\qquad
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\pi^{2}}\right] = -\frac{1}{\pi(1-\pi)} .
$$
:::

The quantity $C_{ij}$ is the second derivative of the truncation constant,
obtained by differentiating $\partial_i f_0/(1-f_0)$ once more; the quotient rule
turns its usual minus into a plus, because $\partial_j(1-f_0) = -\partial_j f_0$,
and @lem-fij supplies $\partial_{ij} f_0$. In expectation the zero branch
contributes nothing, each positive $y$ carries mass $(1-\pi)f(y)/(1-f_0)$, and
since those conditional weights sum to one the constant $C_{ij}$ passes straight
through. The $\pi\pi$ block averages to $-1/\pi(1-\pi)$, which is the Fisher
information of a single Bernoulli trial -- exactly what the separation of the
likelihood promised.

For the moments, the same argument as before — the zero contributes nothing to
positive powers — gives
$\mathbb{E}[Y^{k}] = (1-\pi)\,\mathbb{E}_f[Y^{k}]/(1-f_0)$ for every
$k \geq 1$, and in particular $\mathbb{E}[Y] = (1-\pi)\mu/(1-f_0)$.

### Zero-adjustment for continuous distributions {#sec-za-continuous}

When the parent is continuous the situation simplifies in one way and deepens
in another. It simplifies because $P(W = 0) = 0$ already, so there is nothing
to truncate: adjustment amounts to placing a point mass at zero next to the
density. It deepens because the resulting object is neither continuous nor
discrete but **mixed**, and mixed distributions require a moment of care about
what "density" even means.

::: {#def-za-continuous}
## Zero-adjusted continuous distribution

Let $W$ be continuous with density $f_W(\cdot\,;\theta)$, and let
$\pi \in (0,1)$. Define $Y$ by

$$f_Y(0) = \pi, \qquad f_Y(y) = (1-\pi)f_W(y;\theta) \quad (y \neq 0),$$

where $f_Y$ is a density with respect to the measure $\nu = \delta_0 + \lambda$
— a point mass at zero plus Lebesgue measure.
:::

Integrating against $\nu$ picks up the atom separately from the density:

$$\int f_Y \,\mathrm{d}\nu = f_Y(0) + \int f_Y\,\mathrm{d}\lambda
= \pi + (1-\pi)\cdot 1 = 1,$$

so the object is a proper distribution — but note that its density integrates,
in the ordinary Lebesgue sense, to $1 - \pi$ rather than one. This is the
family of the zero-adjusted Gamma, inverse Gaussian and lognormal used for
semicontinuous data, and of the zero-adjusted Beta used for proportions with
observed zeros; @stasinopoulos2017 collects the regression forms of all of
them.

Because part of the mass is invisible to ordinary integration, the package
gives mixed distributions a way to declare their atoms. The generic
`distrib_atoms()` returns the locations and probabilities of the point
masses. The default returns none, which is correct for every ordinary
distribution, since a continuous one has no atoms and a discrete one consists
of nothing else; only genuinely mixed objects register more. Every routine that
integrates a density, `check_distrib()` included, consults it.

```{r}
#| label: atoms
zag <- zero_adjusted(gamma2_distrib())
distrib_atoms(zag, list(mu = 3, sigma2 = 2, za = 0.3))
distrib_atoms(gamma2_distrib(), list(mu = 3, sigma2 = 2))
```

The cdf jumps by $\pi$ at zero,

$$F_Y(q) = (1-\pi)F_W(q;\theta) + \pi\,\mathbb{I}(q \geq 0),$$

and inverting it requires three regions rather than two, because nothing forces
the parent's support to lie above zero — a zero-adjusted Gaussian is a
perfectly sensible "spike at zero" model whose quantile function must be able
to return negative values. Below the jump, $Q_Y(p) = Q_W(p/(1-\pi))$; inside
the jump, that is for $(1-\pi)F_W(0) \leq p \leq (1-\pi)F_W(0) + \pi$, the
quantile is zero; above it, $Q_Y(p) = Q_W((p-\pi)/(1-\pi))$.

The likelihood separates even more completely than the hurdle's, since there is
no truncation constant at all:

$$\ell_Y = \mathbb{I}(y=0)\log\pi
+ \mathbb{I}(y\neq 0)\Big[\log(1-\pi) + \ell_W(y)\Big].$$

Reading off the branches, the score in $\theta_i$ is $\ell^{(i)}(y)$ off the
atom and zero on it, while the score in $\pi$ is $1/\pi$ on the atom and
$-1/(1-\pi)$ off it. The $\theta\theta$ block of the Hessian is
$\ell^{(ij)}(y)$ off the atom, and the mixed block vanishes. Since
$P(Y \neq 0) = 1-\pi$, and conditionally on $Y \neq 0$ the law of $Y$ is
exactly that of $W$,

$$
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\theta_i\partial\theta_j}\right]
= (1-\pi)\,\mathbb{E}_W[\ell^{(ij)}],
\qquad
\mathbb{E}\left[\frac{\partial^{2}\ell_Y}{\partial\pi^{2}}\right] = -\frac{1}{\pi(1-\pi)} .
$$

There is no analogue of the $C_{ij}$ correction, because there is no
truncation constant to differentiate; this is the main difference from the
discrete hurdle.

Expectations decompose in the way the measure $\nu$ suggests: for any $u$ with
$\mathbb{E}_W|u(W)| < \infty$,

$$\mathbb{E}[u(Y)] = \pi\,u(0) + (1-\pi)\,\mathbb{E}_W[u(W)],$$

and this identity is registered as the `expectation()` method of the class, so
that means, variances and all higher moments account for the atom
automatically. In particular, with $\mu$ and $\sigma^2$ the parent's mean and
variance, $\mathbb{E}[Y] = (1-\pi)\mu$ and
$\operatorname{Var}(Y) = (1-\pi)\sigma^{2} + \pi(1-\pi)\mu^{2}$ — the same
formulas as for zero-inflation, since for a continuous parent the two
constructions define the same mixture, a point @sec-choosing returns to.

One last point concerns derivatives in the response. Away from zero the factor $1-\pi$ is constant in $y$, so
$\partial_y \ell_Y = \partial_y \ell_W$ exactly. At $y = 0$ the log-density
jumps between $\log((1-\pi)f_W)$ and $\log\pi$, so no derivative exists, and
the methods return `NaN` rather than the meaningless number a finite
difference across the jump would produce.

```{r}
#| label: response-at-atom
th <- list(mu = 3, sigma2 = 2, za = 0.3)
distrib_grad_y(zag, c(0, 1, 4), th)
distrib_grad_y(gamma2_distrib(), c(1, 4), list(mu = 3, sigma2 = 2))
```

### Truncation {#sec-truncation}

The last of the operations restricts a distribution to an interval and
renormalizes what remains. It differs from the zero wrappers in a way that
shapes its entire theory: **truncation adds no parameter**. The endpoints are
known constants, on the same footing as a binomial's number of trials. What
truncation adds instead is a normalizing constant that depends on $\theta$, and
differentiating that constant is all the work there is.

::: {#def-truncated}
## Truncated distribution

Let $D$ have density or mass function $f(\cdot\,;\theta)$ and support $S$, and
let $L < U$ be truncation points with $T = S \cap [L, U]$ non-empty.
Writing

$$Z(\theta) = P(L \leq Y \leq U) = F(U;\theta) - F(L^-;\theta), \qquad
F(L^-) = F(L) - P(Y = L),$$

the **truncated** version of $D$ is

$$f_T(y;\theta) = \frac{f(y;\theta)}{Z(\theta)} \ \ (y \in T),
\qquad 0 \text{ otherwise.}$$
:::

Both endpoints are included in the truncated support. For a continuous parent the inclusion makes no difference, since
$P(Y = L) = 0$. For a discrete parent it is essential: truncating a Poisson
below at $L = 1$ produces the zero-truncated Poisson on $\{1, 2, \dots\}$,
and the mass sitting exactly at the included endpoint must be added back into
$Z$, which gives the correction $F(L^-) = F(L) - f(L)$.

One subtlety requires attention. The correction "subtract the mass at $L$" is
naturally read as belonging to the discrete case. It really belongs to the
case where the parent places an **atom** at $L$, and @sec-za-continuous has
just manufactured continuous distributions that do exactly that. The cdf of a zero-adjusted Gamma already contains the point mass at zero, so
for that parent $F(0) \neq F(0^-)$ even though the object is formally
continuous. A truncation that retained the atom but used $F(0)$ in $Z$ would
lose exactly that mass from the normalizing constant. The
implementation therefore obtains the mass at $L$ from `distrib_atoms()`,
uniformly for every kind of parent.

The distribution functions follow the same pattern as before. For $q \in T$,

$$F_T(q) = \frac{F(q;\theta) - F(L^-;\theta)}{Z(\theta)},$$

clamped to $[0,1]$ outside, and solving $F_T(q) = p$ for $q$ gives the quantile
function by delegation to the parent,

$$Q_T(p) = Q\!\left(F(L^-;\theta) + p\,Z(\theta)\right),$$

a relation that holds verbatim for the generalized inverse of a discrete cdf, so
the discrete case needs no separate treatment. Generation is inverse transform
through this formula, $Y = Q(F(L^-) + VZ)$ with $V$ uniform on $(0,1)$ — exact, and
guaranteed to finish in one pass no matter how small $Z$ is, which is where
rejection sampling would stall.

#### Derivatives {#sec-trunc-derivatives}

::: {#thm-truncated}
## Derivatives of a truncated log-likelihood

Let $\ell_T = \ell - \log Z$, and define, with all expectations taken under the
**truncated** distribution itself,

$$m_i = \mathbb{E}_T[\ell^{(i)}], \qquad
M_{ij} = \mathbb{E}_T\!\left[\ell^{(ij)} + \ell^{(i)}\ell^{(j)}\right].$$

Then

$$
\frac{\partial \ell_T}{\partial\theta_i} = \ell^{(i)}(y) - m_i,
\qquad
\frac{\partial^{2}\ell_T}{\partial\theta_i\partial\theta_j}
= \ell^{(ij)}(y) - M_{ij} + m_im_j,
\qquad
\mathbb{E}\!\left[\frac{\partial^{2}\ell_T}{\partial\theta_i\partial\theta_j}\right]
= -\operatorname{Cov}_T\!\left(\ell^{(i)}, \ell^{(j)}\right).
$$
:::

Everything here reduces to differentiating $\log Z$, and one condition makes
that legitimate: the region of integration is the fixed set $T$. The
endpoints are constants and the parent's support does not move with
$\theta$, which permits differentiation under the integral sign. The same
condition explains why truncation at fixed points is a regular problem, while
truncation at a $\theta$-dependent point would not be. Differentiating once and using
$\partial_i f = f\,\ell^{(i)}$ gives $\partial_i Z = Z\,m_i$, so that
$\partial_i \log Z = m_i$; differentiating a second time with @lem-fij gives
$\partial_{ij} Z = Z\,M_{ij}$, and the quotient rule turns that into
$\partial_{ij}\log Z = M_{ij} - m_i m_j$. The expected Hessian then follows by
averaging the observed one under $\mathbb{E}_T$, at which point the two
occurrences of $\mathbb{E}_T[\ell^{(ij)}]$ cancel; the cancellation removes
the parent's expected Hessian from the formula altogether and leaves a
covariance. Truncated likelihoods are treated at book
length in @klein2003 and @johnson1994.

Two remarks follow from these formulas. The score of the truncated model is
the parent's score *recentered* at its mean over the retained region, so the
first Bartlett identity holds by construction. The expected information is the
*covariance matrix of the parent's score* under truncation, which is the
second Bartlett identity in the truncated model and guarantees positive
semi-definiteness. None of the ingredients $m_i$, $M_{ij}$ has a closed form for a general
parent. They are computed by quadrature or summation through
`expectation()`, so the derivatives of a truncated distribution cost markedly
more than the parent's own, and the third and fourth orders fall back to
finite differences of the analytical Hessian.

Inside $[L, U]$ the constant $\log Z$ does not depend on $y$, so the
response derivatives are the parent's, unchanged; outside, the log-density is
$-\infty$ and they are `NaN`.

#### Admissible truncations {#sec-trunc-validity}

The constructor `truncated(distrib, lower, upper)` accepts either endpoint or
both, and it examines what it is given, because many syntactically valid calls
are statistically meaningless. The guiding example: what should
`truncated(gamma2_distrib(), lower = -2)` do? The Gamma lives on $(0,\infty)$,
so truncating at $-2$ removes no mass whatsoever and the "truncated"
distribution would be the Gamma itself. A user who writes such a call has
misunderstood something, and returning the parent silently would bury the
misunderstanding; the constructor refuses instead, and explains.

```{r}
#| label: trunc-refusals
#| error: true
truncated(gamma2_distrib(), lower = -2)
```

The same reasoning rejects a non-integer endpoint for a discrete parent, which
is ambiguous between the two neighboring support points:

```{r}
#| label: trunc-refusals-2
#| error: true
truncated(poisson_distrib(), lower = 1.5)
```

A discrete truncation must also leave enough support to identify the
parameters. A distribution on $k$ points has $k-1$ free probabilities, so a
parent with $p$ parameters needs at least $p+1$ retained points — and since
truncation adds no parameter of its own, this is the only identifiability
condition it ever raises:

```{r}
#| label: trunc-refusals-3
#| error: true
truncated(negbin2_distrib(), 3, 4)
```

One further refusal connects this section to the previous ones.

::: {#prp-trunc-zero}
## Truncating away the zero of a zero wrapper destroys its parameter

Let $D_\zeta$ be the zero-inflated version of $D$ and suppose
$0 \notin [L, U]$. Then the truncated law of $D_\zeta$ does not depend on
$\zeta$ at all.
:::

The reason is a cancellation. For $y \in [L, U]$, which excludes zero, the mass function of $D_\zeta$ is
$(1-\zeta)f(y)$, and the retained mass is
$Z = (1-\zeta)\big(F(U) - F(L^-)\big)$, with the same factor in front. In the
ratio $f_T = (1-\zeta)f(y)/Z$ the two factors cancel, leaving
$f(y)/(F(U) - F(L^-))$, which does not depend on $\zeta$. The argument for the two zero-adjusted forms is the
same with $(1-\pi)$ in place of $(1-\zeta)$.

A parameter with an identically zero score is invisible to the likelihood,
and an optimizer asked to estimate it will wander along a flat ridge. The
constructor therefore refuses the construction. Truncations that keep zero in
the interval remain allowed, and in that case the atom passes through,
rescaled by $1/Z$:

```{r}
#| label: trunc-zero-wrapper
#| error: true
truncated(zero_inflated(poisson_distrib()), lower = 1)
```

```{r}
#| label: trunc-zero-wrapper-ok
tz <- truncated(zero_adjusted(gamma2_distrib()), upper = 5)
th <- list(mu = 3, sigma2 = 2, za = 0.3)
distrib_atoms(tz, th)
0.3 / distrib_cdf(zero_adjusted(gamma2_distrib()), 5, th)
```

Finally, truncating an already truncated distribution is permitted and is
collapsed into a single object over the intersection of the two intervals —
nesting would be mathematically harmless but would pay the quadrature cost of
$m$ and $M$ twice.

```{r}
#| label: trunc-nested
d <- truncated(truncated(gaussian1_distrib(), -1, 3), upper = 2)
c(d@lower, d@upper)
```

To close the section, the two truncated families with textbook closed forms,
recovered by the general machinery:

```{r}
#| label: trunc-closed-forms
# zero-truncated Poisson: P(Y = y) = f(y)/(1 - f(0)), E[Y] = mu/(1 - exp(-mu))
ztp <- truncated(poisson_distrib(), lower = 1)
distrib_pdf(ztp, 1:4, list(mu = 2))
dpois(1:4, 2) / (1 - dpois(0, 2))
c(mean(ztp, list(mu = 2)), 2 / (1 - exp(-2)))

# truncated Gaussian on [-1, 2]: E[Y] = mu + sigma (phi(a) - phi(b))/Z
tn <- truncated(gaussian1_distrib(), lower = -1, upper = 2)
a <- (-1 - 0.5) / 1.5; b <- (2 - 0.5) / 1.5; Z <- pnorm(b) - pnorm(a)
c(mean(tn, list(mu = 0.5, sigma = 1.5)), 0.5 + 1.5 * (dnorm(a) - dnorm(b)) / Z)
```

### Third and fourth derivatives of the wrappers {#sec-higher-orders}

Everything derived so far has stopped at the second derivative, which is where
a fitting routine stops. The toolkit promises four orders, though, and the
four orders apply to the wrappers as well as to the distributions they
wrap. What makes that cheap is a structural observation: **every wrapper's
log-likelihood has the same shape**. Setting the branches side by side,

$$
\begin{aligned}
\text{zero-inflated, } y = 0:\quad & \ell_Y = \log L_0, \qquad L_0 = \zeta + (1-\zeta)f_0,\\
\text{hurdle, } y > 0:\quad & \ell_Y = \ell(y) - \log(1-f_0) + \log(1-\pi),\\
\text{zero-adjusted continuous, } y \neq 0:\quad & \ell_Y = \ell_W(y) + \log(1-\pi),\\
\text{truncated}:\quad & \ell_T = \ell(y) - \log Z,
\end{aligned}
$$

each is the parent's log-density plus, or instead of, **the logarithm of some
quantity that depends on $\theta$**. So one derivation covers all four, provided
a logarithm can be differentiated to arbitrary order, and it can, using nothing
that has not already appeared in this book.

#### The two partition expansions {#sec-two-facts}

The first is @lem-exp-deriv, read in the direction opposite to the one taken in
@sec-bartlett. There it turned $\partial^k f$ into something whose
expectation vanishes; here it simply says how to get the derivatives of a
density from those of its logarithm:

$$\frac{\partial^{I} f}{f} \;=\; \sum_{\pi \in \Pi_I} \prod_{B \in \pi} \ell^{(B)},$$

the complete Bell polynomial in the parent's derivatives, where $\Pi_I$ runs
over the set partitions of the multi-index $I$.

The second is its inverse, and is the classical relation between moments and
cumulants.

::: {#prp-log-deriv}
## Derivatives of a logarithm

For a positive function $L$ of the parameters and a multi-index $I$,

$$\partial^{I} \log L \;=\; \sum_{\pi \in \Pi_I} (-1)^{|\pi|-1}\,(|\pi|-1)!
   \prod_{B \in \pi} \frac{\partial^{B} L}{L}.$$
:::

This is the classical moment-to-cumulant relation. The two expansions invert
one another because substituting one into the other produces a sum over pairs
of partitions, one refining the other; the inner sum is the Möbius function
of the partition lattice, which vanishes except on the single-block
partition. @stanley2012 develops the partition
lattice and its Möbius function, and @comtet1974 gives the two expansions
side by side.

Only the *ratios* $\partial^B L / L$ are ever needed, never $L$'s derivatives
on their own — which matters because for the truncated wrapper those ratios are
exactly the truncated expectations already met in @thm-truncated:

$$\frac{\partial^{B} Z}{Z} = \mathbb{E}_T\!\left[\frac{\partial^{B}f}{f}\right].$$

At second order this is $M_{ij}$, and @prp-log-deriv reduces to
$\partial_{ij}\log Z = M_{ij} - m_i m_j$ — the formula derived by hand earlier.
The same collapse happens everywhere: putting $|I| = 1$ or $2$ into the two
expansions reproduces every score and Hessian in this chapter, including the
$w_0(1-w_0)$ of @prp-zi-hessian and the truncation correction $C_{ij}$ of
@prp-hurdle-hessian. The general machinery and the hand derivations are two
routes to the same formulas, and each is a check on the other.

#### Application to each wrapper {#sec-higher-per-wrapper}

With the two expansions in hand, each wrapper is a matter of naming its $L$ and
its ratios.

For a **transformed** distribution there is nothing to do at all: by
@thm-transform-invariance the derivatives are the parent's at $x = g^{-1}(y)$,
at every order.

For **zero-inflation** at $y = 0$ we have $L = L_0$, which is *affine* in
$\zeta$ — a fact that does most of the work, since it kills every block
containing two or more $\zeta$'s. Writing $A$ for the $\theta$-part of a block
and counting its $\zeta$'s,

$$
\frac{\partial^{B}L_0}{L_0} =
\begin{cases}
w_0 \cdot \dfrac{\partial^{A} f_0}{f_0} & \text{no } \zeta,\\[8pt]
-f_0\,\dfrac{\partial^{A} f_0}{f_0}\Big/L_0 & \text{one } \zeta,\ A \neq \emptyset,\\[8pt]
(1-f_0)/L_0 & \text{one } \zeta,\ A = \emptyset,\\[4pt]
0 & \text{two or more } \zeta.
\end{cases}
$$

Above zero the likelihood separates, so mixed components vanish and the pure
ones are the parent's and $\partial^k\log(1-\zeta) = -(k-1)!/(1-\zeta)^k$.

For the **hurdle** the separation is complete at every order, so all mixed
components are identically zero; the $\theta$ part above zero is the parent's
derivative minus that of $\log(1-f_0)$, whose ratios are
$-f_0(\partial^B f_0/f_0)/(1-f_0)$. The **zero-adjusted continuous** case is the
same without the truncation constant. And **truncation** is $\ell$ minus
$\log Z$ with the expectations above.

As a small closed form to end on, the pure $\zeta$ derivatives at zero: since
$\ell_Y = \log\big(f_0 + \zeta(1-f_0)\big)$ is a logarithm of an affine
function,

$$\frac{\partial^{k}\ell_Y}{\partial\zeta^{k}}\bigg|_{y=0}
= (-1)^{k-1}(k-1)!\left(\frac{1-f_0}{L_0}\right)^{k}.$$

```{r}
#| label: zi-pure-zeta
d <- zero_inflated(poisson_distrib())
f0 <- dpois(0, 3); L0 <- 0.25 + 0.75 * f0; r <- (1 - f0) / L0
c(package = distrib_deriv3(d, 0, list(mu = 3, zi = 0.25))[["zi_zi_zi"]],
  closed  = 2 * r^3)
```

#### Computational cost {#sec-higher-cost}

For the zero wrappers, nothing to speak of: the ratios are algebraic in $f_0$
and the parent's derivatives, and the partition sums have at most fifteen terms
at fourth order. For truncation each *distinct block* costs one quadrature or
summation, so a fourth-order component with two parameters needs a handful —
memoized across the partition sum, since a partition of four indices asks for
the same small blocks repeatedly. It is the same trade the second order already
made, and the reason the closed forms of a future `distrib_grad_cdf` would pay
for themselves.

### Reparametrization {#sec-reparametrize}

The wrappers above change the law. A reparametrization changes only the
coordinates the law is written in: a map $\theta = \theta(\psi)$ carries new
parameters onto the parent's, the density is unchanged, and every
log-likelihood derivative in $\psi$ follows from the parent's by the
multivariate Faà di Bruno formula [@constantine1996]. For a multi-index $I$
on the new parameters,
$$
\ell^{(I)}(\psi) \;=\; \sum_{\pi \vdash I}\;
\sum_{j_1, \dots, j_{\lvert\pi\rvert}}
\ell^{(j_1 \cdots j_{\lvert\pi\rvert})}\bigl(\theta(\psi)\bigr)
\prod_{B \in \pi} \partial^{B} \theta_{j_B}(\psi),
$$
the outer sum over the partitions of the positions of $I$, the inner over
assignments of parent components to blocks: the order of the parent
derivative equals the number of blocks. This is the same formula behind the
link scale of @sec-link-scale; the difference is that a link acts on one
parameter at a time, so its Jacobian is diagonal and most terms vanish,
while a reparametrization may mix parameters -- a mean-precision Gaussian
reaches both $\mu$ and $\sigma$ of the parent -- and the full partition sum
is needed. Under the expectation at a fixed $\psi$ the one-block term
carries the parent's score alone and drops by the first Bartlett identity,
which is what makes the expected information in $\psi$ a congruence of the
parent's.

`reparametrize()` implements the formula for any parent and any map. The
derivatives of the map are written out and handed over as `map_derivs`, a
function returning, for each parent parameter, the non-zero partials keyed
by the tuple of new-parameter positions -- which is how every shipped
second parametrization carries its map. When none are supplied, each needed
partial comes from a single finite-difference stencil applied to the
analytic map, accurate enough for exploratory use but not the route a
family fitted in earnest should take.

```{r}
#| label: reparam-twin
g2 <- reparametrize(
  gaussian1_distrib(),
  map = function(psi) list(mu = psi$mu, sigma = sqrt(psi$sigma2)),
  params = c("mu", "sigma2"),
  bounds = list(mu = c(-Inf, Inf), sigma2 = c(0, Inf)),
  links = list(mu = identity_link(), sigma2 = log_link()),
  map_derivs = function(psi) {
    v <- psi[[2]]
    list(
      list("1" = rep_len(1, length(v))),
      list("2" = 0.5 / sqrt(v), "2,2" = -0.25 / v^1.5,
           "2,2,2" = 0.375 / v^2.5, "2,2,2,2" = -0.9375 / v^3.5)
    )
  }
)

# against the hand-written mean-variance Gaussian: two implementations of
# one object, compared component by component -- by name, since the two may
# enumerate the components in different orders
y <- c(-0.5, 0.8, 2.2)
th <- list(mu = 1.2, sigma2 = 3.5)
ha <- distrib_hessian(g2, y, th)
hb <- distrib_hessian(gaussian2_distrib(), y, th)
max(vapply(names(hb), function(nm) max(abs(ha[[nm]] - hb[[nm]])), numeric(1)))
```

Most of the numbered parametrizations of @sec-numbering are exactly this
wrapper with the family's own map; the hand-written ones among them serve as
the wrapper's independent reference, and agree with it to machine precision.

```{r}
#| label: consistency-reparam
#| include: false
assert_reparametrize_ok()
```

### Choice of construction {#sec-choosing}

The three families answer different questions, and the differences are easiest
to see side by side.

| | Zero-inflation | Zero-adjustment (hurdle) | Truncation |
|---|---|---|---|
| Effect on $P(Y=0)$ | *adds*: $\zeta + (1-\zeta)f_0 > f_0$ | *replaces*: $\pi$, free in $(0,1)$ | removes it, if $0 \notin [L,U]$ |
| Parameters added | one, $\zeta$ | one, $\pi$ | none |
| Fewer zeros than the parent? | impossible | possible | only none at all |
| Meaning of $\theta$ | the original count process | the zero-truncated law | the restricted law |
| Likelihood | does not separate | binary $\times$ truncated | $\ell - \log Z(\theta)$ |
| Discrete parent required | yes | no | no |

: The three constructions {.striped}

The choice between the first two is a choice about the zeros themselves. When a zero records a distinct decision — no claim was filed, no purchase was
made, no rain fell — the zeros and the positives come from two different
mechanisms. The hurdle, whose likelihood separates along exactly that line,
is then the natural model. When the two kinds of zero cannot be told
apart, and $\theta$ is meant to keep describing the count process that the
ordinary observations come from, zero-inflation is the natural model instead.
The two are not nested, and where both fit similarly they still differ in what
their $\theta$ *means*.

Truncation answers a different question altogether. It is not a model for
surplus or deficient zeros but for **unobservable regions**: a detection limit,
a register that only records positive counts, a measuring instrument with a
ceiling. Part of the sample space was never available, and the model says so.

One boundary between the families has to be marked explicitly, because the
package enforces it. For a continuous parent, "zero-inflation" and
"zero-adjustment" would denote the same object — with $P(W=0)=0$ the mixture
$\zeta\delta_0 + (1-\zeta)f_W$ *is* @def-za-continuous with $\zeta$ in the role
of $\pi$ — and offering both names would only invite the belief that they
differ. The constructor therefore accepts a continuous parent only under the
name `zero_adjusted()`, and points there when asked otherwise:

```{r}
#| label: zi-continuous-refused
#| error: true
zero_inflated(gamma2_distrib())
```

### Identifiability {#sec-identifiability}

The zero wrappers add a parameter, and adding a parameter to a family does not
always produce an identified family. What makes this dangerous is that nothing
at run time betrays the failure: the mass function still sums to one, every
numerical check still passes, and an optimizer still converges — to an
arbitrary point of a flat ridge, with standard errors to match. The only place
the failure can be caught is the constructor, so the constructors carry the
mathematics needed to catch it, and this section sets out what that mathematics
is.

#### Compositions of the zero wrappers {#sec-no-stacking}

::: {#prp-no-stack-1}
## Zero-truncation annihilates zero-inflation

Let $D_\zeta$ be the zero-inflated version of $D$. Then the zero-truncated law
of $D_\zeta$ coincides with the zero-truncated law of $D$ itself, whatever the
value of $\zeta$.
:::

Again the parameter cancels. For $y > 0$ the mass function of $D_\zeta$ is
$(1-\zeta)f(y)$ and its total mass on the positive integers is
$1 - L_0 = (1-\zeta)(1-f_0)$, so the ratio that defines the zero-truncated law
is $f(y)/(1-f_0)$, in which $\zeta$ no longer appears.

The consequence is immediate: in `zero_adjusted(zero_inflated(D))` the
parameter $\zeta$ appears nowhere in the resulting mass function, its score is
identically zero, and it cannot be estimated. The reverse composition fails in a similar way. In
`zero_inflated(zero_adjusted(D))` the pair $(\zeta, \pi)$ enters the law only
through the combination $c = \zeta + (1-\zeta)\pi$: the mass at zero is $c$
and the positive part carries the factor $(1-\zeta)(1-\pi) = 1 - c$. Every
pair on the curve $\zeta + (1-\zeta)\pi = c$ therefore produces exactly the
same distribution. Both
constructions are therefore refused:

```{r}
#| label: stacking-demo
try(zero_adjusted(zero_inflated(poisson_distrib())))
try(zero_inflated(zero_adjusted(poisson_distrib())))
```

That these refusals guard against something real, and not a theoretical
scruple, can be seen by building the forbidden object by hand, bypassing the
constructor, and watching the parameter vanish:

```{r}
#| label: stacking-evidence
#| echo: false
zip <- zero_inflated(poisson_distrib())
stacked <- distributions7:::ZeroAdjustedDiscreteDistrib(
  parent_distrib = zip, distrib_name = "hand-built hurdle over a ZIP",
  dimension = "univariate", bounds = c(0, Inf),
  params = c("mu", "zi", "za"),
  params_interpretation = c(mu = "mean", zi = "structural zero", za = "prob. of zero"),
  n_params = 3,
  params_bounds = list(mu = c(0, Inf), zi = c(0, 1), za = c(0, 1)),
  link_params = list(mu = log_link(), zi = logit_link(), za = logit_link()))
```

```{r}
#| label: stacking-evidence-show
distrib_pdf(stacked, 0:5, list(mu = 3, zi = 0.20, za = 0.4))
distrib_pdf(stacked, 0:5, list(mu = 3, zi = 0.70, za = 0.4))
distrib_gradient(stacked, 0:5, list(mu = 3, zi = 0.20, za = 0.4))$zi
```

#### A support-counting condition {#sec-counting-rule}

::: {#prp-counting}
## The counting rule

A distribution supported on $k$ points is an element of the
$(k-1)$-dimensional probability simplex, so a family with $q$ free parameters
can be identified only if $q \leq k-1$. Both zero wrappers have $q = p+1$
parameters, where $p$ is the number of the parent's; hence at least

$$k \geq p + 2$$

support points are necessary for either wrapper to be identified.
:::

The condition is necessary but not sufficient, since a badly parametrized
family can fail to be identified with any amount of support. It is, however,
exactly what rules out the case that arises in practice: the Bernoulli, with
$k = 2$ and $p = 1$.
Zero-inflating a Bernoulli spends two parameters describing the single free
cell of $\{0, 1\}$. Zero-adjusting one fails even more completely: the truncation
concentrates everything on $\{1\}$, so
$f(1)/(1-f_0) = \mu/(1-(1-\mu)) = 1$ whatever $\mu$ is, and the parent's
parameter drops out of the model completely, leaving
$P(Y=0) = \pi,\ P(Y=1) = 1-\pi$.

```{r}
#| label: bernoulli-refused
#| error: true
zero_inflated(bernoulli_distrib())
```

```{r}
#| label: bernoulli-refused-2
#| error: true
zero_adjusted(binomial_distrib(size = 1))
```

The same rule admits a binomial with two trials, whose three support points are
enough for the two parameters:

```{r}
#| label: binomial-two-ok
zero_inflated(binomial_distrib(size = 2))@params
```

A caveat to end on: the counting rule protects against structural
non-identification, the kind that no amount of data can cure. Weak identification of the statistical kind remains possible in any mixture
model — a Poisson mean so large that $f_0$ underflows, a $\zeta$ so close to
zero that the data barely see it. In those cases the standard errors from
`fit_distrib()` are the diagnostic.

```{r}
#| label: consistency-transformations
#| include: false
# Internal consistency gate: fails the render if any density, score, Hessian,
# expected Hessian, moment or closed form derived in this chapter disagrees
# with the implementation, or if check_distrib() fails on any wrapper.
assert_transformations_ok()
```
