## Fitting a distribution {#sec-fitting}

```{r}
#| label: setup-05
#| include: false
source("R/_setup.R")
source("R/fit-certificates.R")
```

The preceding sections built objects and derived their properties: links with
four derivatives in both directions, distributions with scores, informations
and derivatives to fourth order, and operators that turn one distribution into
another. None of that estimated anything. This section closes the gap with the
simplest problem that exercises the machinery end to end: given a sample
$y_1, \dots, y_n$, independent and identically distributed from a member of
one of the families in the catalogue, estimate the parameters by maximum
likelihood and report their uncertainty. There is no regression here — no
covariates and no design matrix — and deliberately so: `fit_distrib()` is the
smallest complete consumer of the toolkit, and every component it uses will be
used in the same way by a regression front end.

### Maximising on the unconstrained scale {#sec-fit-scale}

The total log-likelihood of the sample is

$$\ell(\theta) = \sum_{i=1}^{n} \log f(y_i; \theta),$$

a function of the $p$-vector $\theta$, each of whose components is confined to
its own open interval $\Theta_j$. Maximising over a product of open intervals
is an awkward problem for a general optimiser: every proposal must be checked
for admissibility, a step that overshoots must be shortened rather than taken,
and near a boundary the objective is unbounded or undefined, so the optimiser
spends its effort on the constraint instead of on the likelihood.

The toolkit avoids the problem by never optimising in $\theta$. Each parameter
carries a link, and the estimation is carried out in

$$\eta = \big(g_1(\theta_1), \dots, g_p(\theta_p)\big) \in \mathbb{R}^{p},$$

which is unconstrained by construction. The reparameterisation costs nothing,
because maximum likelihood is invariant under reparameterisation: if
$\hat\eta$ maximises $\ell(h(\eta))$, then $h(\hat\eta)$ maximises
$\ell(\theta)$, the two problems having the same stationary points wherever
$h$ is a diffeomorphism, which @def-link guarantees because $h'$ never
vanishes. The gain is that admissibility becomes automatic: whatever the
optimiser proposes, $h(\eta)$ lies inside the domain, because the inverse link
is a bijection onto $\Theta$. There is no rejection step and no boundary to
guard.

The change of scale requires derivatives with respect to $\eta$, and these
were computed in @sec-link-scale. `fit_distrib()` obtains

$$
U(\eta) = \sum_{i=1}^{n} \frac{\partial \ell_i}{\partial \eta}, \qquad
H(\eta) = \sum_{i=1}^{n} \frac{\partial^{2} \ell_i}{\partial \eta \,\partial \eta^\top},
$$

with the sums over observations justified by independence, since the total
log-likelihood is the sum of the individual contributions.

### Fisher scoring {#sec-fit-scoring}

The three optimisers offered by `fit_distrib()` share one structure:
approximate the log-likelihood near the current point by a quadratic and step
to the maximum of that quadratic. Writing $I$ for the matrix that plays the
role of the negative curvature, the update is

$$\eta^{(t+1)} = \eta^{(t)} + I^{-1} U\big(\eta^{(t)}\big),$$

and the methods differ only in the choice of $I$. **Newton–Raphson** uses the
observed information $-H(\eta)$; **Fisher scoring** uses the expected
information $-\mathbb{E}[H(\eta)]$; **BFGS** uses neither, and builds an
approximation from successive gradients. The first two are treated together in
@green1984, whose presentation of scoring as iteratively reweighted least
squares is the one on which @mccullagh1989 builds the generalized linear
model; @nocedal2006 is the reference for the third, and @sec-optimizers7
treats all of them in detail.

The default is Fisher scoring, for a reason supplied by @cor-congruence. On
the link scale the expected information is

$$\mathbb{E}[I_\eta] = \operatorname{diag}(h')\,\mathbb{E}[I_\theta]\,\operatorname{diag}(h'),$$

a congruence transform, and a congruence transform of a positive definite
matrix is positive definite. If the model is identified on the parameter
scale, the matrix Fisher scoring inverts is therefore positive definite on the
link scale as well, and the step it computes is guaranteed to be an ascent
direction. The observed Hessian carries no such guarantee: away from the
optimum, $-H$ can fail to be positive definite, and the Newton step can then
point downhill. This comparison concerns the behaviour of the iterations, not
the quality of the estimators; at the optimum the two informations agree in
expectation.

An ascent direction fixes only the direction, so the step length is chosen by
halving: the candidate $\eta + s\,\Delta$ is tried with $s = 1$, and $s$ is
halved until the log-likelihood does not decrease. Because the direction is an
ascent direction, a sufficiently small $s$ always succeeds, so the iteration
increases the log-likelihood at every step, and a step that would enter a
region where the density overflows is simply not taken.

Convergence is declared when the score is numerically zero or when the
log-likelihood has stopped moving, whichever occurs first:

$$\max_j \lvert U_j \rvert < \texttt{tol}
\qquad\text{or}\qquad
\lvert \Delta \ell \rvert < \texttt{tol}\,(\lvert \ell \rvert + \texttt{tol}).$$

The second criterion is measured relative to the magnitude of the
log-likelihood, because an absolute threshold has entirely different meanings
for a sample of ten and for a sample of ten million.

::: {.callout-note}
## Fitting a non-regular model

@sec-nonregular showed that the Laplace location parameter sits at a kink of
the log-likelihood: the observed second derivative $\ell^{(\mu\mu)}$ is zero
almost everywhere, while the information, defined as the variance of the
score, is $1/b^{2}$. The consequence for estimation is direct. Newton–Raphson
has nothing to invert, because the observed information in $\mu$ vanishes
identically, while Fisher scoring inverts $1/b^{2}$ and proceeds normally. The
default method works here for the same reason the expected information is the
right definition of information for a non-regular family.
:::

### Starting values {#sec-fit-starts}

A likelihood need not be concave, and a fit started at a poor point can
converge to a local maximum, to a boundary, or not at all. Two mechanisms
handle this automatically.

When no starting value is supplied, several are drawn at random from the
admissible region with `generate_random_theta()` and tried in turn, keeping
the first that converges. If the chosen method fails from a given start —
by exceeding its iteration budget, by producing a non-finite step, or by
raising an error, which a numerically approximated expected information can do
when its quadrature diverges at an unfavourable parameter value — the fit
falls back to BFGS from the same start before moving to the next. BFGS is
slower and uses no second-order information, but it makes fewer demands on the
objective, and a slow answer is preferable to none.

Only when every start has failed by every route does `fit_distrib()` stop, and
it then reports the failure and asks for a starting value instead of returning
an unusable result.

### The delta method and confidence intervals {#sec-fit-delta}

The optimiser returns $\hat\eta$ and the information $I_{\hat\eta}$, so the
asymptotic variance on the link scale is the inverse,
$V_\eta = I_{\hat\eta}^{-1}$. The user needs the uncertainty of
$\hat\theta = h(\hat\eta)$, and the diagonal Jacobian provides the conversion.
The delta method, in the form given by @vandervaart1998, yields

$$V_\theta = \operatorname{diag}(h')\,V_\eta\,\operatorname{diag}(h'),$$

with $h'$ evaluated componentwise at $\hat\eta$. This is the same congruence
that @cor-congruence applied to the information, now applied to its inverse;
indeed the two statements are equivalent, since inverting both sides of
$I_\eta = \operatorname{diag}(h')^{-1} I_\theta \operatorname{diag}(h')^{-1}$
gives exactly the display above.

Confidence intervals are built on the unconstrained scale, where the normal
approximation is most accurate, and mapped back afterwards:

$$
\big[\hat\eta_j - z_{1-\alpha/2}\,\mathrm{se}(\hat\eta_j),\;
      \hat\eta_j + z_{1-\alpha/2}\,\mathrm{se}(\hat\eta_j)\big]
\;\longmapsto\;
h_j\big(\,\cdot\,\big).
$$

This order of operations produces two properties. The mapped interval cannot
leave the parameter's domain, because $h_j$ maps into $\Theta_j$ and nowhere
else: a variance never receives a negative lower limit, and a probability
never receives an upper limit above one, regardless of the sample size. And
the interval is not symmetric about $\hat\theta_j$, correctly so, because
symmetry on the $\theta$ scale is an artefact of a linear approximation rather
than a property of the likelihood.

One implementation detail follows from the fact that @def-link does not
require links to be increasing. A decreasing link — the upper-bounded
$g(\theta) = \log(u - \theta)$ is the example in the catalogue — sends the
lower end of the $\eta$-interval to the upper end of the $\theta$-interval, so
the two mapped endpoints are sorted before being reported. Code that assumed
an increasing link would silently return an interval with its ends reversed.

### A complete example {#sec-fit-example}

```{r}
#| label: fit-demo
set.seed(20260730)
d <- gamma_distrib()
y <- distrib_rng(d, 500, list(mu = 4, sigma2 = 6))

fit <- fit_distrib(d, y)
fit
```

The estimates are close to the values the sample was generated from, and the
interval for `sigma2` is visibly asymmetric — longer to the right, as expected
for a variance estimated on a log scale. All components of the fit are
available individually, on either scale:

```{r}
#| label: fit-pieces
coef(fit)
vcov(fit)
fit@eta                 # the estimates on the link scale
fit@ci                  # intervals on the parameter scale
```

The relation between the two variance matrices is the diagonal Jacobian
congruence of @sec-fit-delta, and it can be verified directly:

```{r}
#| label: fit-delta-check
J <- vapply(seq_along(d@params), function(i) {
  linkfunctions7::dlinkinv(d@link_params[[d@params[i]]], fit@eta[i])
}, numeric(1))
diag(J) %*% fit@vcov_eta %*% diag(J)
```

The maximised log-likelihood, the information criteria, the number of
iterations and the method actually used — which may differ from the one
requested when the fallback was taken — are all carried on the fitted object,
and the `simulate()` and `plot()` methods compare the fitted distribution
against the data it came from.

```{r}
#| label: fit-plot
#| fig-cap: "The fitted Gamma against the sample it was estimated from."
plot(fit)
```

`simulate()` draws new samples of the original size from the fitted
distribution, which is the ingredient a parametric bootstrap needs: under the
fitted model, a statistic computed on the simulated samples has the same
distribution that it has under repeated sampling.

```{r}
#| label: fit-simulate
sims <- simulate(fit, nsim = 200, seed = 1)
boot_means <- vapply(sims, mean, numeric(1))
c(observed = mean(y), boot_se = sd(boot_means))
```

### Scope {#sec-fit-scope}

Three extensions lie outside this section.

There are no covariates. Every observation shares one $\theta$, so the link
scale is a $p$-vector rather than a set of linear predictors. Adding
covariates changes the bookkeeping and not the mathematics: the chain rule
acquires one more factor, $\eta = X\beta$, whose Jacobian is $X$ rather than
$\operatorname{diag}(h')$, and everything in @sec-fit-delta goes through with
the two composed. This belongs to the model layer.

Censoring is not handled, although @sec-cdf-derivatives supplied the
derivatives a censored likelihood needs. Assembling them requires a status
vector alongside the response and an interface for stating which observations
are censored and how — a front end that does not yet exist.

Finally, the optimisers are the three described above, and `fit_distrib()`
does not accept a fourth. Removing that restriction is the subject of
@sec-optimizers7, which develops optimisation as a layer of its own. The
further goal of allowing a Bayesian treatment in place of maximum likelihood
belongs to the model layer.

```{r}
#| label: consistency-fitting
#| include: false
# Internal consistency gate: fails the render if the score does not vanish at
# the reported optimum, if the delta-method variance disagrees with the
# link-scale one, if an interval leaves its parameter's domain, or if the three
# methods disagree about where the maximum is.
assert_fit_ok()
```
