## Fitting a distribution {#sec-fitting}

```{r}
#| label: setup-05
#| include: false
source("R/_setup.R")
source("R/fit-certificates.R")
```

The preceding sections built objects and derived their properties: links with
four derivatives in both directions, distributions with scores, informations
and derivatives to fourth order, and operators that turn one distribution into
another. None of that estimated anything. This section closes the gap with the simplest problem that exercises the
machinery end to end. Given a sample $y_1, \dots, y_n$, independent and
identically distributed from a member of one of the families in the
catalog, the task is to estimate the parameters by maximum likelihood and
to report their uncertainty. There is no regression here, no covariates and
no design matrix: `fit_distrib()` is the smallest complete consumer of the
toolkit, and every component it uses will be used in the same way by a
regression front end.

### Maximizing on the unconstrained scale {#sec-fit-scale}

The total log-likelihood of the sample is

$$\ell(\theta) = \sum_{i=1}^{n} \log f(y_i; \theta),$$

a function of the $p$-vector $\theta$, each of whose components is confined to
its own open interval $\Theta_j$. Maximizing over a product of open intervals is an awkward problem for a
general optimizer. Every proposal must be checked for admissibility, and a
step that overshoots must be shortened rather than taken. Near a boundary the
objective is unbounded or undefined, so the optimizer spends its effort on
the constraint instead of on the likelihood.

The toolkit avoids the problem by never optimizing in $\theta$. Each parameter
carries a link, and the estimation is carried out in

$$\eta = \big(g_1(\theta_1), \dots, g_p(\theta_p)\big) \in \mathbb{R}^{p},$$

which is unconstrained by construction. The reparametrization costs nothing,
because maximum likelihood is invariant under reparametrization: if
$\hat\eta$ maximizes $\ell(h(\eta))$, then $h(\hat\eta)$ maximizes
$\ell(\theta)$, the two problems having the same stationary points wherever
$h$ is a diffeomorphism, which @def-link guarantees because $h'$ never
vanishes. The gain is that admissibility becomes automatic: whatever the
optimizer proposes, $h(\eta)$ lies inside the domain, because the inverse link
is a bijection onto $\Theta$. There is no rejection step and no boundary to
guard.

The change of scale requires derivatives with respect to $\eta$, and these
were computed in @sec-link-scale. `fit_distrib()` obtains

$$
U(\eta) = \sum_{i=1}^{n} \frac{\partial \ell_i}{\partial \eta}, \qquad
H(\eta) = \sum_{i=1}^{n} \frac{\partial^{2} \ell_i}{\partial \eta \,\partial \eta^\top},
$$

where the sums over observations follow from the independence of the sample.

### Fisher scoring {#sec-fit-scoring}

The optimization itself is not written here. `fit_distrib()` assembles the
objective, the score and the information on the link scale and hands them to
`optimizers7`, whose methods are the subject of @sec-optimizers7. What this
section describes is the choice of matrix to invert, which belongs to the
likelihood rather than to the algorithm.

The three named strategies share one structure: approximate the
log-likelihood near the current point by a quadratic and step to the maximum
of that quadratic. Writing $I$ for the matrix that plays the role of the
negative curvature, the update is

$$\eta^{(t+1)} = \eta^{(t)} + I^{-1} U\big(\eta^{(t)}\big),$$

and the methods differ only in the choice of $I$. **Newton–Raphson** uses the
observed information $-H(\eta)$; **Fisher scoring** uses the expected
information $-\mathbb{E}[H(\eta)]$; **BFGS** uses neither, and builds an
approximation from successive gradients. The first two are the same algorithm
of @sec-newton with a different second-order argument, so Fisher scoring
requires no separate implementation --- which is why it is not an optimizer
but an object saying which matrix to use, `fisher_scoring()`, handed to the
same argument that would otherwise carry `optimizers7::newton()` or
`optimizers7::lbfgs()`. What that object does carry, and an optimizer could
not, is how the expected information is to be obtained when the family has no
closed form for it. They are treated together in @green1984,
whose presentation of scoring as iteratively reweighted least squares is the
one on which @mccullagh1989 builds the generalized linear model, while
@nocedal2006 is the reference for the third.

The default is Fisher scoring, for a reason supplied by @cor-congruence. On
the link scale the expected information is

$$\mathbb{E}[I_\eta] = \operatorname{diag}(h')\,\mathbb{E}[I_\theta]\,\operatorname{diag}(h'),$$

a congruence transform, and a congruence transform of a positive definite
matrix is positive definite. If the model is identified on the parameter
scale, the matrix Fisher scoring inverts is therefore positive definite on the
link scale as well, and the step it computes is an ascent direction without
any repair. The observed Hessian carries no such guarantee: away from the
optimum, $-H$ can fail to be positive definite, and the raw Newton step can
then point downhill, which is why @sec-newton floors the eigenvalues of the
matrix it inverts. This comparison concerns the behavior of the iterations,
not the quality of the estimators; at the optimum the two informations agree
in expectation.

An ascent direction fixes only the direction, so the step length is chosen by
the line search of @sec-descent-armijo: the candidate $\eta + \alpha\,\Delta$ is tried
with $\alpha = 1$ and $\alpha$ is reduced until the decrease in the objective is a fixed
fraction of the one the linear model predicts. Because the direction is an
ascent direction, a sufficiently small $\alpha$ succeeds in exact arithmetic, so the
iteration increases the log-likelihood at every step, and a step that would
enter a region where the density overflows is simply not taken.

The quantity actually handed to the optimizer is not the log-likelihood but
the log-likelihood per observation, $\ell(\eta)/n$, with the score and the
information divided by $n$ as well. The maximizer is unchanged, and so is
every Newton step, since the factor cancels between the inverse Hessian and
the gradient. What changes is the meaning of a threshold. A score is a sum of
$n$ terms, so a fixed bound on $\max_j \lvert U_j \rvert$ asks a sample of ten
million to reach an accuracy per observation ten million times finer than it
asks of a sample of ten; on the averaged scale the same number means the same
thing at every sample size. Convergence is therefore declared when

$$\max_j \left\lvert \frac{U_j(\eta)}{n} \right\rvert < \texttt{tol},$$

which is the stopping rule `crit_grad()` of @sec-optimizers7, an object the
caller can replace rather than a comparison written inside the iteration. The
tolerance is the rule's own, and the iteration limit the optimizer's, so both
are set on the method rather than beside it: the fit contributes the objective
and not the terms on which the search over it ends. The
log-likelihood, the information and the standard errors are all recomputed
unscaled at the reported optimum, so nothing the fit returns is measured per
observation.

How small a threshold can usefully be asked for is settled by the objective
rather than by the algorithm. Near the maximum the increase a step can produce
is about $\lVert U/n \rVert^{2}/(2\lambda)$ for a curvature $\lambda$, and once
that quantity falls below the rounding of the objective itself, of order
$\varepsilon \lvert \ell/n \rvert$, the line search can no longer tell an
improvement from noise and refuses every step. The smallest score a run can
reach is therefore near
$$\sqrt{2 \lambda \varepsilon \lvert \ell(\eta)/n \rvert},$$
which for a log-likelihood per observation of order one is around $10^{-8}$.
The default tolerance is $10^{-6}$, two orders above that limit, and asking for
much less would demand accuracy the arithmetic cannot certify. Nothing
statistical is given up: a score of $10^{-6}$ per observation places the
estimate a small fraction of a standard error from the maximum, since a
standard error is itself of order $n^{-1/2}$.

::: {.callout-note}
## Fitting a non-regular model

@sec-nonregular showed that the Laplace location parameter sits at a kink of
the log-likelihood: the observed second derivative $\ell^{(\mu\mu)}$ is zero
almost everywhere, while the information, defined as the variance of the
score, is $1/b^{2}$. The consequence for estimation is direct. Newton–Raphson
has nothing to invert, because the observed information in $\mu$ vanishes
identically, while Fisher scoring inverts $1/b^{2}$ and proceeds normally. The
default method works here for the same reason the expected information is the
right definition of information for a non-regular family.
:::

### Starting values {#sec-fit-starts}

A likelihood need not be concave, and a fit started at a poor point can
converge to a local maximum, to a boundary, or not at all. How far the start
is from the answer is not a detail of the implementation but a property of the
problem, and it governs the amount of work the iteration has to do before it
arrives, when it arrives at all.

The starting value comes from a distribution's own `distrib_start()`, which
receives the data. A family that knows an estimator of its parameters returns
it, and the fit then begins where a moment estimator or an exact maximum
likelihood estimator already is; a family that says nothing gets several
random draws from the admissible region, tried in turn. The multivariate
gaussian is the case where the difference is total: its maximum likelihood
estimator for an unstructured covariance is the sample mean and the sample
second moment, so it starts at the answer and confirms it in a single
iteration.

```{r}
#| label: fit-start
d4 <- mvgaussian_distrib(4)
y4 <- as.matrix(iris[, 1:4])

# what the family proposes, against what the data say
start <- distrib_start(d4, y4)[[1]]
rbind(proposed = mv_location(d4, start), sample = colMeans(y4))

# and the fit it leads to
f4 <- fit_distrib(d4, y4)
c(iterations = f4@iterations, converged = f4@converged)
```

The chosen method can still fail from a given start: by exceeding its
iteration budget, by producing a non-finite step, or by raising an error,
which a numerically approximated expected information can do when its
quadrature diverges at an unfavorable parameter value. In that case the fit
falls back to BFGS from the same start before moving to the next. BFGS is
slower and uses no second-order information, but it makes fewer demands on
the objective, so it can produce a result where the chosen method produced
none.

Across the starts the fit keeps the best run rather than the last: a converged
run in preference to one that stopped without meeting its rule, and among runs
of the same status the one with the higher likelihood. Only when every start
has failed by every route does `fit_distrib()` stop; the error message then
requests a starting value instead of an unusable result being returned.

The fallback belongs to the two named second-order strategies. An optimizer
supplied explicitly is used as given and is never replaced, since substituting
a different algorithm for the one the caller chose would report a fit obtained
by a method other than the one requested.

### The delta method and confidence intervals {#sec-fit-delta}

The optimizer returns $\hat\eta$ and the information $I_{\hat\eta}$, so the
asymptotic variance on the link scale is the inverse,
$V_\eta = I_{\hat\eta}^{-1}$. The user needs the uncertainty of
$\hat\theta = h(\hat\eta)$, and the diagonal Jacobian provides the conversion.
The delta method, in the form given by @vandervaart1998, yields

$$V_\theta = \operatorname{diag}(h')\,V_\eta\,\operatorname{diag}(h'),$$

with $h'$ evaluated componentwise at $\hat\eta$. This is the same congruence
that @cor-congruence applied to the information, now applied to its inverse;
indeed the two statements are equivalent, since inverting both sides of
$I_\eta = \operatorname{diag}(h')^{-1} I_\theta \operatorname{diag}(h')^{-1}$
gives exactly the display above.

Confidence intervals are built on the unconstrained scale, where the normal
approximation is most accurate, and mapped back afterwards:

$$
\big[\hat\eta_j - z_{1-\alpha/2}\,\mathrm{se}(\hat\eta_j),\;
      \hat\eta_j + z_{1-\alpha/2}\,\mathrm{se}(\hat\eta_j)\big]
\;\longmapsto\;
h_j\big(\,\cdot\,\big).
$$

This order of operations produces two properties. The mapped interval cannot
leave the parameter's domain, because $h_j$ maps into $\Theta_j$ and nowhere
else: a variance never receives a negative lower limit, and a probability
never receives an upper limit above one, regardless of the sample size. And
the interval is not symmetric about $\hat\theta_j$, since symmetry on the
$\theta$ scale is an artifact of a linear approximation and not a property of
the likelihood.

One implementation detail follows from the fact that @def-link does not
require links to be increasing. A decreasing link — the upper-bounded
$g(\theta) = \log(u - \theta)$ is the example in the catalog — sends the
lower end of the $\eta$-interval to the upper end of the $\theta$-interval, so
the two mapped endpoints are sorted before being reported. Code that assumed
an increasing link would silently return an interval with its ends reversed.

### A complete example {#sec-fit-example}

```{r}
#| label: fit-demo
set.seed(20260730)
d <- gamma2_distrib()
y <- distrib_rng(d, 500, list(mu = 4, sigma2 = 6))

fit <- fit_distrib(d, y)
fit
```

The estimates are close to the values the sample was generated from, and the
interval for `sigma2` is visibly asymmetric — longer to the right, as expected
for a variance estimated on a log scale. All components of the fit are
available individually, on either scale:

```{r}
#| label: fit-pieces
coef(fit)
vcov(fit)
coef(fit, scale = "link")
confint(fit, scale = "link")   # symmetric here
confint(fit)                   # its image under the inverse link
```

Because the optimization is delegated rather than written in, `method` accepts
any optimizer of @sec-optimizers7 in place of one of the three named
strategies, and the object carries its own stopping rule with it:

```{r}
#| label: fit-optimizer
library(optimizers7)

fits <- list(
  fisher = fit_distrib(d, y),
  lbfgs  = fit_distrib(d, y, method = lbfgs(criterion = crit_grad(1e-7))),
  nelder = fit_distrib(d, y, method = nelder_mead(maxit = 5000))
)
vapply(fits, function(f) as.numeric(logLik(f)), numeric(1))
vapply(fits, function(f) f@method, character(1))
```

The three agree on the maximized log-likelihood; the choice of algorithm
affects the path to the optimum and not its location.

The relation between the two variance matrices is the diagonal Jacobian
congruence of @sec-fit-delta, and it can be verified directly:

```{r}
#| label: fit-delta-check
J <- vapply(seq_along(d@params), function(i) {
  linkfunctions7::dlinkinv(d@link_params[[d@params[i]]], fit@eta[i])
}, numeric(1))
diag(J) %*% fit@vcov_eta %*% diag(J)
```

The maximized log-likelihood, the information criteria, the number of
iterations and the method actually used are all carried on the fitted object.
The method recorded may differ from the one requested when the fallback was
taken. The `simulate()` and `plot()` methods compare the fitted distribution
against the data it came from.

```{r}
#| label: fit-plot
#| fig-cap: "The fitted Gamma against the sample it was estimated from."
plot(fit)
```

`simulate()` draws new samples of the original size from the fitted
distribution, which is the ingredient a parametric bootstrap needs: under the
fitted model, a statistic computed on the simulated samples has the same
distribution that it has under repeated sampling.

```{r}
#| label: fit-simulate
sims <- simulate(fit, nsim = 200, seed = 1)
boot_means <- vapply(sims, mean, numeric(1))
c(observed = mean(y), boot_se = sd(boot_means))
```

### Scope {#sec-fit-scope}

Three extensions lie outside this section.

There are no covariates. Every observation shares one $\theta$, so the link
scale is a $p$-vector rather than a set of linear predictors. Adding
covariates changes the bookkeeping and not the mathematics: the chain rule
acquires one more factor, $\eta = X\beta$, whose Jacobian is $X$ rather than
$\operatorname{diag}(h')$, and everything in @sec-fit-delta goes through with
the two composed. This belongs to the model layer.

Censoring is not handled, although @sec-cdf-derivatives supplied the
derivatives a censored likelihood needs. Assembling them requires a status
vector alongside the response and an interface for stating which observations
are censored and how — a front end that does not yet exist.

The remaining goal, a Bayesian treatment in place of maximum likelihood,
belongs to the model layer.

```{r}
#| label: consistency-fitting
#| include: false
# Internal consistency gate: fails the render if the score does not vanish at
# the reported optimum, if the delta-method variance disagrees with the
# link-scale one, if an interval leaves its parameter's domain, or if the three
# methods disagree about where the maximum is.
assert_fit_ok()
```
