## Fitting a distribution {#sec-fitting}

```{r}
#| label: setup-05
#| include: false
source("R/_setup.R")
source("R/fit-certificates.R")
```

The three chapters so far have built objects and derived what can be computed
from them: links carrying four derivatives in both directions, distributions
carrying scores, informations and derivatives to fourth order, and operators
that turn one distribution into another while keeping all of it. Nothing so far
has *estimated* anything. This chapter closes that gap. It is shorter than the
others, and deliberately so: almost everything it needs has already been
derived, and the pleasure of the exercise is watching the earlier machinery fall
into place with very little new mathematics.

The task is the simplest one in which the machinery is exercised end to end.
Given a sample $y_1, \dots, y_n$ assumed independent and identically
distributed from a member of one of the families in the catalogue, estimate the
parameter by maximum likelihood, and report an uncertainty for it. There is no
regression here — no covariates, no design matrix — and that is the point:
`fit_distrib()` is the smallest complete consumer of the toolkit, and every
component it uses is a component that a regression front end will use in the
same way.

### Maximising on the unconstrained scale {#sec-fit-scale}

Write the total log-likelihood of the sample as

$$\ell(\theta) = \sum_{i=1}^{n} \log f(y_i; \theta),$$

a function of the $p$-vector $\theta$, each of whose components is confined to
its own open interval $\Theta_j$. Maximising a function over a product of open
intervals is an awkward problem to hand to a general optimiser. Every proposal
must be checked for admissibility; a step that overshoots must be shortened
rather than taken; and near a boundary the objective is unbounded or undefined,
so the optimiser spends its effort fighting the constraint rather than climbing
the likelihood.

The toolkit disposes of this by never optimising in $\theta$ at all. Each
parameter carries a link, and the estimation is carried out in

$$\eta = \big(g_1(\theta_1), \dots, g_p(\theta_p)\big) \in \mathbb{R}^{p},$$

which is unconstrained by construction. The reparameterisation costs nothing,
because maximum likelihood is invariant under a reparameterisation: if
$\hat\eta$ maximises $\ell(h(\eta))$ then $h(\hat\eta)$ maximises $\ell(\theta)$,
the two problems having the same stationary points wherever $h$ is a
diffeomorphism — which @def-link guarantees, $h'$ never vanishing. The gain is
that admissibility becomes automatic. Whatever the optimiser proposes,
$h(\eta)$ lands inside the domain, because that is what the inverse link is: a
bijection onto $\Theta$. There is no rejection step and no boundary to guard.

What the change of scale does cost is derivatives, and those have already been
paid for. The score and the Hessian with respect to $\eta$ are exactly what
@sec-link-scale computed, and `fit_distrib()` obtains them by asking for them:

$$
U(\eta) = \sum_{i=1}^{n} \frac{\partial \ell_i}{\partial \eta}, \qquad
H(\eta) = \sum_{i=1}^{n} \frac{\partial^{2} \ell_i}{\partial \eta \,\partial \eta^\top},
$$

each summed over observations because the sample is independent and the total
log-likelihood is the sum of the individual ones.

### Fisher scoring, and why it is the default {#sec-fit-scoring}

All three optimisers offered are variations on one idea: approximate the
log-likelihood near the current point by a quadratic and step to the maximum of
that quadratic. Writing $I$ for the matrix that plays the role of the negative
curvature, the update is

$$\eta^{(t+1)} = \eta^{(t)} + I^{-1} U\big(\eta^{(t)}\big),$$

and the three methods differ only in what they use for $I$. **Newton–Raphson**
uses the observed information $-H(\eta)$; **Fisher scoring** uses the expected
information $-\mathbb{E}[H(\eta)]$; **BFGS** uses neither, building an
approximation from successive gradients instead.

The default is Fisher scoring, for a reason that @cor-congruence supplied two
chapters ago and that is worth restating because it is easy to mistake for a
mere convenience. On the link scale the expected information is

$$\mathbb{E}[I_\eta] = \operatorname{diag}(h')\,\mathbb{E}[I_\theta]\,\operatorname{diag}(h'),$$

a congruence transform. A congruence transform of a positive definite matrix is
positive definite, so if the model is identified on the parameter scale then the
matrix Fisher scoring inverts is positive definite on the link scale as well,
and the step it computes is guaranteed to be an ascent direction. The observed
Hessian carries no such guarantee: away from the optimum $-H$ can fail to be
positive definite, and the Newton step can then point downhill. This is not a
statement about which estimator is better — at the optimum the two informations
agree in expectation — but about which iteration is better behaved on the way
there.

Ascent being a direction rather than a distance, the step itself is halved until
it earns its keep: the candidate $\eta + s\,\Delta$ is tried with $s = 1$, and
$s$ is halved until the log-likelihood does not decrease. Since the direction is
an ascent direction, some sufficiently small $s$ always succeeds, so the
iteration is monotone in the log-likelihood by construction, and a step that
would leap into a region where the density overflows is simply not taken.

Convergence is declared when the score is numerically zero or the
log-likelihood has stopped moving, whichever comes first,

$$\max_j \lvert U_j \rvert < \texttt{tol}
\qquad\text{or}\qquad
\lvert \Delta \ell \rvert < \texttt{tol}\,(\lvert \ell \rvert + \texttt{tol}),$$

the second criterion measured relative to the log-likelihood's own magnitude,
since an absolute threshold means something quite different for a sample of ten
and one of ten million.

::: {.callout-note}
## The non-regular case, revisited

@sec-nonregular showed that the Laplace location sits at a kink: its observed
second derivative $\ell^{(\mu\mu)}$ is zero almost everywhere, while the
information, defined as the variance of the score, is $1/b^{2}$. The consequence
for estimation is immediate and rather striking. Newton–Raphson on a Laplace has
nothing to invert — the observed information in $\mu$ vanishes identically —
whereas Fisher scoring has $1/b^{2}$ and proceeds without noticing anything
unusual. That the default method is the one that works here is not a
coincidence: the same property that makes the expected information the right
object for a non-regular family makes it the right matrix to invert.
:::

### Starting values and what happens when they fail {#sec-fit-starts}

A likelihood need not be concave, and a fit started in the wrong place can
converge to a local maximum, to a boundary, or not at all. Two devices handle
this without troubling the user.

When no starting value is supplied, several are drawn at random from the
admissible region with `generate_random_theta()` and tried in turn, the first
that converges being kept. And if the chosen method fails from a given start —
whether by exceeding the iteration budget, by producing a step that is not
finite, or by raising an error outright, which a numerically approximated
expected information can do when its quadrature diverges at an awkward parameter
value — the fit falls back to BFGS from that same start before moving on to the
next. BFGS is slower and uses no second-order information, but it is unfussy,
and a slow answer is better than none.

Only when every start has failed by every route does `fit_distrib()` give up,
and it then says so and asks for a starting value rather than returning
something unusable.

### From the link scale back to the parameter scale {#sec-fit-delta}

The optimiser returns $\hat\eta$ and the information $I_{\hat\eta}$, from which
the asymptotic variance on the link scale is the usual inverse,
$V_\eta = I_{\hat\eta}^{-1}$. What the user wants, though, is an uncertainty for
$\hat\theta = h(\hat\eta)$, and here the diagonal Jacobian does its last piece
of work. The delta method gives

$$V_\theta = \operatorname{diag}(h')\,V_\eta\,\operatorname{diag}(h'),$$

with $h'$ evaluated at $\hat\eta$ componentwise — the same congruence that
@cor-congruence applied to the information, now applied to its inverse. The two
statements are the same statement: if $I_\eta = \operatorname{diag}(h')^{-1} I_\theta \operatorname{diag}(h')^{-1}$
then inverting both sides gives exactly the display above. It is a small
pleasure that the transformation appearing in the derivative calculus of
@sec-link-scale and the one appearing in the delta method are not two facts to
remember but one.

Intervals are built where the normal approximation is most nearly true — on the
unconstrained scale — and mapped back afterwards:

$$
\big[\hat\eta_j - z_{1-\alpha/2}\,\mathrm{se}(\hat\eta_j),\;
      \hat\eta_j + z_{1-\alpha/2}\,\mathrm{se}(\hat\eta_j)\big]
\;\longmapsto\;
h_j\big(\,\cdot\,\big).
$$

Two properties follow, and both are the reason for doing it in this order. The
mapped interval **cannot leave the parameter's domain**, since $h_j$ maps onto
$\Theta_j$ and nowhere else: a variance never acquires a negative lower limit, a
probability never an upper limit above one, however small the sample. And the
interval is **not symmetric** about $\hat\theta_j$, which is as it should be —
symmetry on the $\theta$ scale is an artefact of a linear approximation, not a
property of the likelihood.

One detail of the implementation is a direct consequence of @def-link having
declined to require that links increase. A decreasing link — the upper-bounded
$g(\theta) = \log(u - \theta)$ is the catalogue's example — sends the lower end
of the $\eta$-interval to the *upper* end of the $\theta$-interval, so the two
mapped endpoints are sorted before being reported. Code that assumed an
increasing link would return an interval with its ends the wrong way round, and
would do so silently.

### A fit end to end {#sec-fit-example}

```{r}
#| label: fit-demo
set.seed(20260730)
d <- gamma_distrib()
y <- distrib_rng(d, 500, list(mu = 4, sigma2 = 6))

fit <- fit_distrib(d, y)
fit
```

The estimates sit near the values the sample was generated from, and the
interval for `sigma2` is visibly asymmetric — longer to the right, as a variance
estimated on a log scale should be. The pieces of the fit are all available
individually, on either scale:

```{r}
#| label: fit-pieces
coef(fit)
vcov(fit)
fit@eta                 # the estimates on the link scale
fit@ci                  # intervals on the parameter scale
```

That the two variance matrices are related by the diagonal Jacobian, exactly as
@sec-fit-delta claims, can be seen directly:

```{r}
#| label: fit-delta-check
J <- vapply(seq_along(d@params), function(i) {
  linkfunctions7::dlinkinv(d@link_params[[d@params[i]]], fit@eta[i])
}, numeric(1))
diag(J) %*% fit@vcov_eta %*% diag(J)
```

The maximised log-likelihood, the information criteria, the number of iterations
and the method actually used — which may not be the one requested, if the
fallback was taken — are all carried on the fitted object, and `simulate()` and
`plot()` methods compare the fitted distribution against the data it came from.

```{r}
#| label: fit-plot
#| fig-cap: "The fitted Gamma against the sample it was estimated from."
plot(fit)
```

### What is deliberately absent {#sec-fit-scope}

It is worth being explicit about the boundary of this chapter, because the
apparatus assembled here is capable of a good deal more than it is asked to do.

There are no covariates. Every observation shares one $\theta$, so the link
scale is a $p$-vector rather than a set of linear predictors, and the score is a
$p$-vector rather than a $\mathrm{X}^\top$-weighted sum. Adding covariates
changes the bookkeeping and not the mathematics: the chain rule acquires one
more link in the chain, $\eta = X\beta$, whose Jacobian is $X$ rather than
$\operatorname{diag}(h')$, and everything in @sec-fit-delta goes through with
the two composed. That is the subject of the model layer, not of this one.

Nor is censoring handled here, although @sec-cdf-derivatives supplied the
derivatives that a censored likelihood needs. Assembling them into a likelihood
requires a status vector alongside the response, and an interface for saying
which observations are censored and how — a front end that does not yet exist.

Finally, the optimisers are the three above, and `fit_distrib()` does not let
the user supply a fourth. That restriction is the subject of @sec-optimizers7,
which develops optimisation as its own layer: what makes a step admissible, how
curvature is obtained or approximated, what to do when the objective has a kink
and no gradient becomes small, and how a box constraint is made to vanish rather
than be enforced. The toolkit's further ambition, that the user should be able
to choose a Bayesian treatment in place of maximum likelihood, belongs to the
model layer and to neither of these chapters.

```{r}
#| label: consistency-fitting
#| include: false
# Internal consistency gate: fails the render if the score does not vanish at
# the reported optimum, if the delta-method variance disagrees with the
# link-scale one, if an interval leaves its parameter's domain, or if the three
# methods disagree about where the maximum is.
assert_fit_ok()
```
