## Numerical fallbacks {#sec-fallbacks}

```{r}
#| label: setup-03d
#| include: false
source("R/_setup.R")
```

The contract of @sec-distrib-object is unusually generous: a new distribution
must supply its density and nothing else. Everything the preceding sections
have described — the distribution function, the quantile function, random
generation, derivatives of every order in the parameters and in the response —
is provided by default implementations registered on the base classes, which any
subclass may override but none is obliged to.

That generosity is what makes the extension story credible. A package that only
worked for the fourteen families it ships with would have solved nothing; the
test is whether a user's own distribution is a first-class citizen, and it is one
exactly to the extent that the fallbacks are good. This section describes them,
because two contain ideas worth reading in their own right, and because knowing
where a fallback is weak is part of knowing when to replace it.

### Distribution function, quantiles and derivatives {#sec-fallbacks-easy}

Several fallbacks are what one would guess. The distribution function is a
quadrature of the density from the lower end of the support for a continuous
family, and a partial sum for a discrete support one. The quantile function inverts it by
root-finding. Derivatives with respect to the parameters are central finite
differences, each order differencing the analytical implementation of the order
below when one exists, so that a distribution supplying a closed-form score gets
a Hessian one differencing away rather than two.

The steps of those differences are the one place where care is needed, and the
reason is the openness of the parameter domains. A scale parameter is positive,
not non-negative, so a step chosen from the magnitude of $\sigma$ alone will,
for small $\sigma$, place $\sigma - \delta$ on the wrong side of zero, and the
log-density comes back `NaN` for a reason that looks like a defect in the
density. Each step is therefore shrunk to stay within the domain, and the same
is done in the response direction so that $y \pm \delta$ never leaves the
support — the Gamma near zero being the case that forces it.

### Random generation by ratio of uniforms {#sec-fallbacks-grou}

Random generation is where the obvious fallback is unusable. Inverse transform
sampling needs $Q(V)$ for $V$ uniform, and if the quantile function is itself
the numerical fallback then each draw costs a root-finding call, each iteration
of which costs a quadrature. A sample of a thousand becomes a noticeable wait.

The package instead uses a **generalized ratio-of-uniforms** scheme, which needs
the density alone. The idea is a change of perspective: instead of sampling $Y$
directly, sample uniformly from a two-dimensional region built from the density
and read $Y$ off the ratio of the coordinates.

::: {#thm-grou}
## Generalized ratio-of-uniforms

Let $\kappa(y) \geq 0$ be integrable, let $r > 0$, and let

$$
\mathcal{A}_r = \left\{(u, v) \;:\; 0 < u \leq
   \kappa\!\left(\tfrac{v}{u^{r}}\right)^{1/(r+1)} \right\}.
$$

If $(U, V)$ is uniform on $\mathcal{A}_r$, then $Y = V/U^{r}$ has density
proportional to $\kappa$.
:::

The mechanism is a single change of variables. Writing $v = y\,u^{r}$, the joint
density of $(U, Y)$ is proportional to $u^{r}$ over the region
$\{0 < u \le \kappa(y)^{1/(r+1)}\}$, and integrating $u$ out leaves
$\kappa(y)/(r+1)$, which is proportional to $\kappa$ and free of $u$. The
method is due to @kinderman1977 in its original form and to @wakefield1991 in
the generalised one used here; @devroye1986 remains the standard reference for
non-uniform generation as a whole.

The practical content is that $\mathcal{A}_r$ is contained in a rectangle whose
sides can be computed from $\kappa$, so one samples uniformly from the rectangle
and rejects the points outside $\mathcal{A}_r$. Only the density is needed, and
only up to a constant — which is fortunate, since a normalising constant is
exactly what a user-defined density may not have.

Two devices make the scheme reliable rather than merely correct, and both were
put in after the naive version failed.

The kernel is **recentred at the mode** before the bounding rectangle is
computed. Without this a density concentrated far from the origin — a Gaussian at
$\mu = 1000$, say — produces a rectangle whose acceptance probability is
essentially zero, because the region $\mathcal{A}_r$ is a sliver in a corner of
it. The mode is located by evaluating the log-density on a grid over a
compactified support and keeping the two cells around the largest value, which
for a unimodal density provably still brackets the mode and shrinks it by a
factor of 64 per pass.

::: {.callout-note}
## Locating the mode

The obvious way to locate the mode is a golden-section search on the
compactified scale, and it is not accurate enough. Its tolerance is expressed in
the compactified variable $t$, and the derivative of the compactification can be
enormous: under the tangent map $\mathrm{d}y/\mathrm{d}t$ is of order $y^{2}$, so
a default tolerance of about $10^{-4}$ in $t$ became an error of 125 standard
deviations in $y$ for a density centred at 1000. The grid refinement stops on the
width of the bracket measured in $y$, which is the quantity that matters.
:::

The kernel is also **normalised to a maximum of one**, so that the rectangle does
not inherit an arbitrary scale from an unnormalised density.

So equipped, the method copes with far more than one expects of it: bimodal
densities, a Student $t$ with half a degree of freedom, a Pareto with infinite
mean. There is exactly one shape it cannot handle, and it is worth naming
precisely, because it is also the one the next subsection removes.

### Densities unbounded at an endpoint {#sec-fallbacks-divergence}

If $\kappa$ is unbounded then no bounding rectangle exists and the scheme has
nothing to sample from. This is not an exotic case: a Gamma with shape below one
diverges at zero, as does a Beta with either parameter below one.

The cure is a change of variable that flattens the singularity. Suppose that
near a finite edge $a$ of the support

$$f(y) \sim \lvert y - a \rvert^{\alpha - 1}, \qquad 0 < \alpha < 1,$$

so that the density diverges as the edge is approached. Put
$X = \lvert Y - a \rvert^{1/\lambda}$ for some $\lambda > 0$. The density of $X$
carries the Jacobian of the map, and a short computation gives an exponent
$\lambda\alpha - 1$ in place of $\alpha - 1$: the transformed density behaves
like $x^{\lambda\alpha - 1}$ near the edge, which is **bounded** as soon as
$\lambda\alpha > 1$, and in fact vanishes there. One samples $X$ by the scheme
above and maps back.

What makes this practical rather than merely possible is that $\alpha$ never has
to be searched for. Taking logarithms of the assumed behaviour,

$$\log f(a + \delta) \approx \text{const} + (\alpha - 1)\log\delta,$$

so walking towards the edge in decades — $\delta$, $\delta/10$, $\delta/100$ —
raises the log-density by $(1-\alpha)\log 10$ at every step when the density
diverges, and by an amount that dies away when it does not. **The probe that
detects the divergence therefore also measures it**, to about four decimals,
from the same handful of density evaluations. For a Gamma of shape $0.4$ the
transformation takes a draw from 27 milliseconds to under a microsecond, and the
draws remain exact — nothing here is an approximation, only a reparameterisation.

A density that diverges at **both** ends of a finite interval needs slightly
more, since a single power flattens one edge while steepening the other. What
serves is a map behaving like a different power at each end,

$$T(u) = \frac{u^{p}}{u^{p} + (1-u)^{q}}, \qquad Y = a + (b-a)\,T(U),$$

which increases from $0$ to $1$, behaves like $u^{p}$ on the left and like
$1 - (1-u)^{q}$ on the right, and has a closed-form derivative. The density of
$U$ then carries the exponents $p\alpha - 1$ and $q\beta - 1$ at the two ends and
is bounded as soon as $p\alpha > 1$ and $q\beta > 1$. Both exponents come from
the same probe as before, run once at each edge.

```{r}
#| label: grou-demo
# a shape below one: the density is unbounded at zero
d <- gamma_distrib()
th <- list(mu = 0.4, sigma2 = 1)
set.seed(7)
x <- distrib_rng(d, 5000, th)

c(mean = mean(x), theoretical = 0.4)
c(var = var(x), theoretical = 1)
```

### Discrete distributions {#sec-fallbacks-discrete}

For a discrete family none of the above is required, and it is worth saying why
rather than leaving it implicit. The distribution function of a discrete variable
is a step function, so its generalized inverse is exact: there is no root to
find, only a table to look up. The fallbacks therefore build one cumulative
table over the support — grown geometrically until it covers the probability
asked of it — and read the distribution function, the quantile function and the
random generator off it.

The only thing that ever went wrong here was performance, and not of the
algorithm: the lookup was originally one R-level call per draw, which cost
$12.6\,\mu s$ against $0.1\,\mu s$ once vectorised. The requirement the table
imposes is a finite lower bound on the support, which every standard count
distribution has.

### Limitations {#sec-fallbacks-caveat}

Two warnings belong with all of this.

The first concerns validation and is easy to miss. A distribution relying on the
numerical derivatives will pass the derivative checks of `check_distrib()`
*trivially*, because the analytical value and the numerical reference are then
the same computation. Those checks become informative exactly when a closed form
has been supplied — which is the moment one wants them, but it means a clean
report on a density-only distribution says less than it appears to.

The second concerns quadrature, and is the reason `expectation()` is deliberately
conservative.

::: {.callout-note}
## Quadrature knots

`expectation()` splits its integral at quantiles of the distribution, and uses
three knots rather than nine. More knots would extend the range of densities it
can integrate at all — but they would also convert loud failures into silent
wrong answers. At Gamma shape $0.02$ a nine-knot rule returns $-54.9$ for a
quantity that is exactly zero, while the three-knot rule reports that it could
not compute the integral. In the ordinary range three knots are also the more
accurate choice, every additional panel contributing its own error. An error is
worth more than a plausible number.
:::
