## Numerical fallbacks {#sec-fallbacks}

```{r}
#| label: setup-03d
#| include: false
source("R/_setup.R")
```

The contract of @sec-distrib-object requires a new distribution to supply its
density and nothing else. Everything else that the preceding sections described — the distribution
function, the quantile function, random generation, derivatives of every
order in the parameters and in the response — is provided by default
implementations registered on the base classes. Any subclass may override
them; none is obliged to.

These defaults are what makes user-defined distributions practical: a
distribution written by a user is a first-class citizen exactly to the extent
that the fallbacks are accurate and fast. This section describes how each
fallback works and where its limits are, since the limits of a fallback
determine when a closed form is worth supplying.

### Distribution function, quantiles and derivatives {#sec-fallbacks-easy}

Several fallbacks are direct implementations of their definitions. The
distribution function is computed by quadrature of the density from the lower
end of the support for a continuous family, and by partial summation for a
discrete one; both engines are the batched ones of
@sec-numerical-quadrature, so a vector of quantiles costs one adaptive pass
rather than one per element. The quantile function inverts the distribution
function by root-finding. Derivatives with respect to the parameters are central finite
differences built on the stencil library of @sec-numerical-stencils, and
each order differences the analytical implementation of the
order below when one exists, so a distribution supplying a closed-form score
obtains its Hessian from one differencing rather than two.

The step sizes of these differences require care because the parameter
domains are open. A scale parameter is positive, not non-negative, so a step
chosen only from the magnitude of $\sigma$ will, for small $\sigma$, place
$\sigma - \delta$ below zero, and the log-density then returns `NaN` for a
reason that looks like a defect in the density itself. Each step is therefore
shrunk to stay within the domain, and the same is done in the response
direction, so that $y \pm \delta$ never leaves the support; the Gamma near
zero is the case that requires it.

### Random generation by ratio of uniforms {#sec-fallbacks-grou}

For random generation the direct fallback is too slow to use. Inverse
transform sampling computes $Q(V)$ for uniform $V$, and when the quantile
function is itself the numerical fallback, each draw costs a root-finding
call, and each iteration of the root-finder costs a quadrature; a sample of a
thousand becomes a noticeable wait.

The package instead uses a **generalized ratio-of-uniforms** scheme, which
needs only the density. Instead of sampling $Y$ directly, the scheme samples
uniformly from a two-dimensional region built from the density and obtains $Y$
as the ratio of the coordinates.

::: {#thm-grou}
## Generalized ratio-of-uniforms

Let $\kappa(y) \geq 0$ be integrable, let $r > 0$, and let

$$
\mathcal{A}_r = \left\{(u, v) \;:\; 0 < u \leq
   \kappa\!\left(\tfrac{v}{u^{r}}\right)^{1/(r+1)} \right\}.
$$

If $(U, V)$ is uniform on $\mathcal{A}_r$, then $Y = V/U^{r}$ has density
proportional to $\kappa$.
:::

The result follows from a single change of variables. Writing $v = y\,u^{r}$,
the joint density of $(U, Y)$ is proportional to $u^{r}$ over the region
$\{0 < u \le \kappa(y)^{1/(r+1)}\}$, and integrating $u$ out leaves
$\kappa(y)/(r+1)$, which is proportional to $\kappa$ and free of $u$. The
method is due to @kinderman1977 in its original form and to @wakefield1991 in
the generalized form used here; @devroye1986 is the standard reference for
non-uniform random generation in general.

In practice, $\mathcal{A}_r$ is contained in a rectangle whose sides can be
computed from $\kappa$, so the sampler draws uniformly from the rectangle and
rejects the points outside $\mathcal{A}_r$. Only the density is needed, and
only up to a constant, which matters because a user-defined density often has
no normalizing constant.

Two additional devices are needed for the scheme to work reliably.

First, the kernel is **recentered at the mode** before the bounding rectangle
is computed. Without recentering, a density concentrated far from the origin —
a Gaussian at $\mu = 1000$, for example — produces a rectangle in which the
region $\mathcal{A}_r$ occupies a negligible corner, and the acceptance
probability is essentially zero. The mode is located by evaluating the
log-density on a grid over a compactified support and keeping the two cells
around the largest value. For a unimodal density this bracket always contains
the mode, and it shrinks by a factor of 64 at each pass.

::: {.callout-note}
## Locating the mode

A golden-section search on the compactified scale is the obvious alternative
for locating the mode, and it is not accurate enough. Its tolerance is expressed in the compactified variable $t$, while the
derivative of the compactification grows with the location of the mode. Under
the tangent map, $\mathrm{d}y/\mathrm{d}t$ is of order $y^{2}$, so a fixed
tolerance in $t$ corresponds to an error in $y$ that grows quadratically with
the mode. The grid refinement instead stops on the width of the bracket
measured in $y$.
:::

Second, the kernel is **normalized to a maximum of one**, so that the
rectangle does not inherit an arbitrary scale from an unnormalized density.

With these two devices the method handles bimodal densities, a Student $t$
with half a degree of freedom, and a Pareto with infinite mean. One class of
densities remains outside its reach, and the next subsection removes it.

### Densities unbounded at an endpoint {#sec-fallbacks-divergence}

If $\kappa$ is unbounded, no bounding rectangle exists and the scheme has
nothing to sample from. The case is common: a Gamma with shape below one
diverges at zero, and so does a Beta with either parameter below one.

The remedy is a change of variable that removes the singularity. Suppose that
near a finite edge $a$ of the support

$$f(y) \sim \lvert y - a \rvert^{\alpha - 1}, \qquad 0 < \alpha < 1,$$

so that the density diverges as the edge is approached. Set
$X = \lvert Y - a \rvert^{1/\lambda}$ for some $\lambda > 0$. The density of $X$ carries the Jacobian of the map, and a short computation
replaces the exponent $\alpha - 1$ with $\lambda\alpha - 1$. The transformed
density therefore behaves like $x^{\lambda\alpha - 1}$ near the edge, which
is bounded as soon as $\lambda\alpha > 1$, and in fact vanishes there. The sampler draws $X$ by
the scheme above and maps back.

The exponent $\alpha$ does not need to be searched for. Taking logarithms of
the assumed behavior,

$$\log f(a + \delta) \approx \text{const} + (\alpha - 1)\log\delta,$$

so approaching the edge in decades — $\delta$, $\delta/10$, $\delta/100$ —
raises the log-density by $(1-\alpha)\log 10$ at each step when the density
diverges, and by a vanishing amount when it does not. The same probe that
detects the divergence therefore also measures its exponent, to about four
decimal places, from a handful of density evaluations. The transformation
reduces the cost of a draw by several orders of magnitude for strongly
divergent shapes, and the draws remain exact, because the construction is a
reparametrization and not an approximation.

A density that diverges at **both** ends of a finite interval requires a map
that behaves like a different power at each end,

$$T(u) = \frac{u^{p}}{u^{p} + (1-u)^{q}}, \qquad Y = a + (b-a)\,T(U),$$

which increases from $0$ to $1$, behaves like $u^{p}$ on the left and like
$1 - (1-u)^{q}$ on the right, and has a closed-form derivative. The density of
$U$ then carries the exponents $p\alpha - 1$ and $q\beta - 1$ at the two ends
and is bounded as soon as $p\alpha > 1$ and $q\beta > 1$. Both exponents come
from the same probe as before, run once at each edge.

```{r}
#| label: grou-demo
# a shape below one: the density is unbounded at zero
d <- gamma2_distrib()
th <- list(mu = 0.4, sigma2 = 1)
set.seed(7)
x <- distrib_rng(d, 5000, th)

c(mean = mean(x), theoretical = 0.4)
c(var = var(x), theoretical = 1)
```

### Discrete distributions {#sec-fallbacks-discrete}

A discrete family needs none of the machinery above, for a simple reason: the
distribution function of a discrete variable is a step function, so its
generalized inverse is exact — there is no root to find, only a table to look
up. The fallbacks build one cumulative table over the support, grown
geometrically until it covers the probability requested, and read the
distribution function, the quantile function and the random generator off the
table. The lookup is vectorized, and the only requirement the table imposes is
a finite lower bound on the support, which every standard count distribution
has.

### Limitations {#sec-fallbacks-caveat}

Two warnings apply to everything in this section.

The first concerns validation. A distribution that relies on the numerical
derivatives passes the derivative checks of `check_distrib()` trivially,
because the analytical value and the numerical reference are then the same
computation. Those checks become informative only when a closed form has been
supplied, so a clean report on a density-only distribution covers fewer
properties than the same report on a fully implemented one.

The second concerns the number of quadrature knots.

::: {.callout-note}
## Quadrature knots

`expectation()` subdivides its integral at three quantiles of the
distribution. A finer subdivision would integrate a wider class of densities,
but the additional panels contribute additional quadrature error in ordinary
cases and, at extreme parameter values, can produce a finite result where the
integral is numerically intractable: at Gamma shape $0.02$, a nine-knot rule
yields $-54.9$ for an integral whose value is $0$, while the three-knot rule
signals failure. The coarser rule is therefore used, and failure is signaled
rather than approximated.
:::
