## Several dimensions {#sec-multivariate}

```{r}
#| label: setup-03e
#| include: false
source("R/_setup.R")
source("R/multivariate-certificates.R")
```

The families of @sec-distributions all place their mass on the real line, and
an observation is a number. This section describes what changes when an
observation is a vector of $p$ numbers, and what does not. The likelihood
theory of @sec-likelihood is unaffected: the score, the observed information,
the Bartlett identities and the change of scale are statements about a
parametric family and say nothing about the dimension of its support. What
changes is the parametrization, because a multivariate family carries a matrix
where a univariate one carries a variance, and a matrix is constrained.

Four families are presented here. Two are elliptical, the gaussian and the
Student $t$, and are the ones the toolkit needs first; @kotz2004 is the
reference for the multivariate $t$ and @lange1989 for its use as a robust
replacement for the gaussian. Two are not, the Dirichlet and the multinomial,
and they are what shows which parts of the construction depend on ellipticity
and which do not.

### The response and the parameters {#sec-mv-parameters}

A multivariate distribution places its mass on $\mathbb{R}^p$, and a sample of
$n$ observations is the $n \times p$ matrix $Y$ whose $i$th row is $y_i$. The
number of observations is the row count rather than the length, which is the
only structural change the response requires.

The parameters need more care. A $p$-variate gaussian has a mean vector
$\mu \in \mathbb{R}^p$ and a covariance matrix $\Sigma$, and $\Sigma$ is not a
free object: it must be symmetric and positive definite. A parameter of this
kind is what @sec-parameters7 is about. All this section needs from it is the
contract: a **parameter** is a map

$$
\eta \;\longmapsto\; M(\eta), \qquad \eta \in \mathbb{R}^{d},
$$ {#eq-mv-parameter}

from an unconstrained vector of length $d$ to a symmetric matrix in the
constrained set, supplying the derivatives
$A_k = \partial M / \partial \eta_k$ and
$A_{kl} = \partial^2 M / \partial \eta_k \partial \eta_l$, the
log-determinant and its derivatives, and the solves $M^{-1}B$. The map is
onto the set and one-to-one, so $\eta$ ranges over the whole of
$\mathbb{R}^{d}$ with no constraint of its own.

The parameters of a multivariate distribution are then the $p$ components of
the mean together with the $d$ free values of the parameter, and every one of
them is a **scalar**. This is what lets the rest of the package apply
unchanged. A parameter vector $\theta$ is a named list of numbers, the
derivative components are keyed by parameter names exactly as in
@sec-likelihood, and the enumeration that produces `hess_names()` produces
the same strings whether the parameters are $(\mu, \sigma)$ or
$(\mu_1, \mu_2, \log L_{11}, \log L_{22}, L_{21})$. None of `align_theta()`,
the link scale of @sec-link-scale, or the fitting routine of @sec-fitting
requires a special case for the multivariate family.

Every one of these scalars is unconstrained, so every link is the identity.
The constraint that would otherwise need a link has been moved into the
parameter, where it belongs: it is a constraint on the matrix as a whole and
not on any one entry, and a scalar link cannot express it. A multivariate
gaussian therefore has a link scale that coincides with its parameter scale.
A multivariate $t$ does not, because its degrees of freedom are a genuine
positive scalar and carry a log link.

There is one restriction, and it comes from what a parameter means here. In a
univariate family a parameter may vary from observation to observation, which
is what a regression does. Here it may not: the parameter's free values
describe one matrix for the whole sample. A model in which the matrix varies
by observation belongs to the layer that owns the regression, not to the
distribution, and a parameter of length greater than one is rejected rather
than truncated to its first entry.

The names of the free values carry a prefix saying which matrix they describe,
`sigma_` for a covariance and `omega_` for a precision. The name a parameter
gives a free value records how the matrix is built and not which matrix it is,
so the same parameter on the two sides of a model would otherwise produce
identical parameter names for two different models. The distribution applies
the prefix rather than the parameter, since the parameter does not know which
side it has been handed to.

### The gaussian {#sec-mvgaussian}

The $p$-variate gaussian has density

$$
f(y) \;=\; (2\pi)^{-p/2} \, |\Sigma|^{-1/2}
  \exp\!\left\{ -\tfrac{1}{2} (y-\mu)^\top \Sigma^{-1} (y-\mu) \right\},
$$ {#eq-mvgaussian-density}

so that with $r = y - \mu$ and $w = \Sigma^{-1} r$ the log-density is
$\ell = -\tfrac{p}{2}\log(2\pi) - \tfrac{1}{2}\log|\Sigma|
- \tfrac{1}{2} r^\top w$.

Differentiating with respect to the mean gives $\partial \ell/\partial \mu = w$,
one component per coordinate. For the matrix directions, the derivative of the
quadratic form follows from $\partial_k \Sigma^{-1} = -\Sigma^{-1} A_k
\Sigma^{-1}$, which is the derivative of an inverse, so that
$\partial_k (r^\top \Sigma^{-1} r) = -w^\top A_k w$. Together with the
parameter's own $\partial_k \log|\Sigma|$,

$$
\ell^{(\mu_a)} = w_a, \qquad
\ell^{(\eta_k)} = -\tfrac{1}{2}\,\partial_k \log|\Sigma|
   \;+\; \tfrac{1}{2}\, w^\top A_k w .
$$ {#eq-mvgaussian-score}

Both pieces on the right come from the parameter, and neither depends on which
family of matrices it belongs to. The distribution is written once and works
with any parameter that satisfies the contract.

The second derivatives follow by differentiating again. The mean block is
$-\Sigma^{-1}$, constant in $y$; the mixed block is
$\ell^{(\mu_a \eta_k)} = -(\Sigma^{-1} A_k w)_a$; and the matrix block is

$$
\ell^{(\eta_k \eta_l)} = -\tfrac{1}{2}\,\partial_{kl} \log|\Sigma|
  + \tfrac{1}{2}\, w^\top A_{kl} w
  - w^\top A_l \Sigma^{-1} A_k w .
$$ {#eq-mvgaussian-hessian}

The expected information simplifies more than the observed Hessian suggests.
Taking expectations uses only $\mathbb{E}[w] = 0$ and
$\mathbb{E}[w w^\top] = \Sigma^{-1}$. The first kills the mixed block exactly,
so the mean and the matrix are orthogonal; the second turns each quadratic
form into a trace. Both second-derivative terms of the matrix block then
collapse, and

$$
\mathbb{E}\!\left[\ell^{(\mu_a \mu_b)}\right] = -(\Sigma^{-1})_{ab},
\qquad
\mathbb{E}\!\left[\ell^{(\mu_a \eta_k)}\right] = 0,
\qquad
\mathbb{E}\!\left[\ell^{(\eta_k \eta_l)}\right]
  = -\tfrac{1}{2}\,\mathrm{tr}\!\left(\Sigma^{-1} A_k \Sigma^{-1} A_l\right).
$$ {#eq-mvgaussian-information}

The second derivatives $A_{kl}$ have disappeared, and so has the second
derivative of the log-determinant. The information needs only the first
derivatives of the parameter, which is what makes Fisher scoring cheaper here
than a Newton step, and it is what @sec-fit-scoring already predicted in
one dimension.

```{r}
#| label: mvgaussian-demo
d <- mvgaussian_distrib(2)
d@params

theta <- list(mu1 = 0.5, mu2 = -0.3, sigma_log_L1 = 0.1, sigma_log_L2 = -0.2, sigma_L2.1 = 0.4)
mv_sigma(d, theta)
```

A gaussian may equally be written in the precision $\Omega = \Sigma^{-1}$, and
which of the two a model parametrizes is a modeling decision rather than a
notational one: a parameter imposed on $\Sigma$ is a different family from the
same parameter imposed on $\Omega$, since the inverse of a compound-symmetry
matrix is compound symmetry while the inverse of an AR(1) is not AR(1). The
constructor takes the parameter on either side, and everything above still
applies, because $\partial \Sigma / \partial \eta_k = -\Sigma A_k \Sigma$ when
$A_k$ is a derivative of the precision. The two parametrizations describe the
same law when the matrices are matched.

```{r}
#| label: mvgaussian-precision
do <- mvgaussian_distrib(2, omega = parameters7::log_cholesky(2))
eta_o <- parameters7::param_free(do@param, solve(mv_sigma(d, theta)))
theta_o <- as.list(stats::setNames(c(0.5, -0.3, eta_o), do@params))

y <- rbind(c(0, 0), c(1, -1), c(-0.5, 0.8))
max(abs(distrib_pdf(do, y, theta_o, log = TRUE) -
        distrib_pdf(d, y, theta, log = TRUE)))
```

### The Student t {#sec-mvstudent-t}

The multivariate $t$ replaces the exponential tail with a polynomial one. With
$q = (y - \mu)^\top \Sigma^{-1} (y - \mu)$ its density is

$$
f(y) \;=\;
\frac{\Gamma\!\left(\frac{\nu+p}{2}\right)}
     {\Gamma\!\left(\frac{\nu}{2}\right) (\nu\pi)^{p/2} |\Sigma|^{1/2}}
\left( 1 + \frac{q}{\nu} \right)^{-(\nu+p)/2}.
$$ {#eq-mvt-density}

Here $\Sigma$ is the **scale** matrix and not the covariance. The covariance is
$\nu\Sigma/(\nu-2)$ where it exists, and for $\nu \le 2$ it does not exist at
all while the distribution is perfectly well defined; the mean exists for
$\nu > 1$. Keeping the two apart is what allows a heavy-tailed family to be
described and fitted at degrees of freedom where its second moment is
infinite, so the package returns the scale matrix from `mv_sigma()` and the
covariance from `variance()`.

The distribution is the scale mixture

$$
Y \mid V \sim N_p\!\left(\mu,\; \tfrac{\nu}{V}\,\Sigma \right),
\qquad V \sim \chi^2_\nu,
$$ {#eq-mvt-mixture}

which is how the package generates from it and where the shape of its score
comes from. Writing

$$
c \;=\; \frac{\nu + p}{\nu + q}
$$ {#eq-mvt-weight}

for the weight an observation receives, the derivatives with respect to the
mean and the parameter are those of the gaussian, @eq-mvgaussian-score, with
every term involving the data multiplied by $c$:

$$
\ell^{(\mu_a)} = c \, w_a, \qquad
\ell^{(\eta_k)} = -\tfrac{1}{2}\,\partial_k \log|\Sigma|
   \;+\; \tfrac{1}{2}\, c \, w^\top A_k w .
$$ {#eq-mvt-score}

The weight is what makes the family resistant. An observation with a large
Mahalanobis distance has a large $q$, hence a small $c$, and contributes
proportionally less to the estimate of both the location and the scale; as
$\nu \to \infty$ the weight tends to $1$ uniformly and the gaussian is
recovered. A gaussian fitted to data containing a few distant rows has no such
mechanism and inflates $\Sigma$ until it covers them.

The degrees of freedom contribute one further component,

$$
\ell^{(\nu)} = \tfrac{1}{2}\left[
  \psi\!\left(\tfrac{\nu+p}{2}\right) - \psi\!\left(\tfrac{\nu}{2}\right)
  - \frac{p}{\nu} - \log\!\left(1 + \frac{q}{\nu}\right)
  + \frac{(\nu+p)\,q}{\nu(\nu+q)} \right],
$$ {#eq-mvt-score-nu}

with $\psi$ the digamma function. The first three terms come from the
normalizing constant and do not involve the data; the last two are the
derivative of the kernel. Since $\nu > 0$ it carries a log link, and the
derivative on the link scale is $\nu\,\ell^{(\nu)}$ by the chain rule of
@sec-link-scale, with no change to the rest.

The expected information of the $t$ has no closed form in the package, so it
is approximated by one of the strategies of @sec-approx. How the expectation
is approximated is a property of Fisher scoring rather than of the fit, so the
choice travels with the method: `fisher_scoring(approx = , nsim = )` carries
it, and it is accepted only for a family that has no closed form, since for
one that has, the argument would be silently ignored and the fit would report
a method it did not use.

```{r}
#| label: mvt-demo
dt <- mvstudent_t_distrib(2)
theta_t <- list(mu1 = 0.5, mu2 = -0.3, sigma_log_L1 = 0.1, sigma_log_L2 = -0.2,
                sigma_L2.1 = 0.4, nu = 4)

# the scale matrix is what the parametrization carries; the covariance is a
# moment, and at nu = 4 it is twice as large
mv_sigma(dt, theta_t)
variance(dt, theta_t)
```

### The simplex and the counts {#sec-mv-simplex}

Two families of the package are not elliptical. The Dirichlet places its mass
on the simplex, a set of dimension $p-1$ inside $\mathbb{R}^p$, and the
multinomial places its mass on a finite set of points. Neither has a location
and a scale to separate, and in both the matrix parameter of
@sec-mv-parameters is replaced by a simplex parameter, which is the same idea
under a different constraint: a map from an unconstrained vector onto the set
the family requires. Everything in @sec-mv-parameters that made the parameters
scalars applies unchanged, so the derivatives, the link scale and the fitting
routine need no case of their own; what does change is every construction that
assumed the density is taken with respect to Lebesgue measure on the whole of
$\mathbb{R}^p$.

The Dirichlet is written in a mean vector $\mu$ on the simplex and a
concentration $\varphi > 0$, from which the usual shapes are
$\alpha = \varphi\mu$. Its density is

$$
f(y) = \frac{\Gamma(\varphi)}{\prod_j \Gamma(\alpha_j)}
       \prod_j y_j^{\alpha_j - 1},
$$ {#eq-dirichlet}

taken with respect to Lebesgue measure on the simplex. The mean
parametrization is what makes the family usable in a regression, since $\mu$
is the quantity a model would put a linear predictor on, and $\varphi$ then
governs the dispersion around it at a fixed mean. Writing $\eta$ for the free
vector of the simplex parameter, $A = \partial\mu/\partial\eta$ for its
Jacobian and $g_j = \log y_j - \psi(\alpha_j)$, the score is

$$
\ell^{(\eta_k)} = \varphi \sum_j g_j A_{jk},
\qquad
\ell^{(\varphi)} = \psi(\varphi) + \sum_j g_j \mu_j .
$$ {#eq-dirichlet-score}

The expected information is closed form, and the reason is one identity used
twice. Differentiating $\sum_j \mu_j = 1$ once shows that every column of $A$
sums to zero, and differentiating it again shows that every second-derivative
vector of the simplex parameter does too. Under expectation
$\mathbb{E}[g_j] = 0$, so every term carrying the data disappears and what is
left is

$$
\mathbb{E}\!\left[\ell^{(\eta_k\eta_l)}\right]
  = -\varphi^2 \sum_j \psi'(\alpha_j) A_{jk} A_{jl},
\qquad
\mathbb{E}\!\left[\ell^{(\varphi\varphi)}\right]
  = \psi'(\varphi) - \sum_j \psi'(\alpha_j)\mu_j^2 .
$$ {#eq-dirichlet-information}

The covariance of the Dirichlet is
$\operatorname{Cov}(Y_i, Y_j) = (\delta_{ij}\mu_i - \mu_i\mu_j)/(\varphi+1)$,
singular because the coordinates sum to a constant. That singularity is what
makes the support lower-dimensional, and it decides how the density can be
integrated numerically. `check_distrib()` confirms that a multivariate density
integrates to one by importance sampling, and for an elliptical family the
proposal it draws from is a gaussian with the family's own mean and an
inflated covariance. Such a proposal spreads its mass over a set of dimension
$p$ and places none of it on the simplex, so the estimate it produces is zero
to within rounding. The proposal must place its mass on the support of the
family, and the uniform distribution on the simplex (the Dirichlet with every
shape equal to one, of constant density $\Gamma(p)$) is the simplest one that
does. A family declares its own proposal through `mv_reference_draw()`.

The multinomial is the discrete case, with size $n$ and probabilities $\pi$
carried by a simplex parameter, and its mass function is
$P(Y = y) = n!\prod_j \pi_j^{y_j}/\prod_j y_j!$ on the vectors of non-negative
integers summing to $n$. Its score and Hessian are the same two zero sums at
work again, with $y_j/\pi_j$ in place of $g_j$, and its expected information
follows from $\mathbb{E}[y_j] = n \pi_j$, the second-derivative term vanishing
as before:

$$
\mathbb{E}\!\left[\ell^{(\eta_k\eta_l)}\right]
  = -n \sum_j \frac{A_{jk} A_{jl}}{\pi_j} .
$$ {#eq-multinomial-information}

Being discrete with a finite support, the multinomial can enumerate it. The
support is the set of weak compositions of $n$ into $p$ parts, of which there
are $\binom{n+p-1}{p-1}$, and `mv_support()` returns them as a matrix. Every
expectation the family needs is then a sum over that matrix rather than a
quadrature or a sample, and is therefore exact up to rounding. Both quantities
a validator would otherwise approximate are computed this way: the mass
function sums to one, and @eq-multinomial-information is also the sum of the
observed Hessian weighted by that mass. The enumeration is practical only for
a moderate size, growing as it does in both $n$ and $p$, and a family too
large for it keeps the approximations of @sec-approx.

Both families have closed-form marginals, a property that is not a
consequence of ellipticity. A single coordinate of a Dirichlet is
$\mathrm{Beta}(\alpha_j, \varphi - \alpha_j)$, which in the
mean-and-precision parametrization of @sec-distributions is mean $\mu_j$ and
precision $\varphi$, so the concentration is shared by every marginal. A
single coordinate of a multinomial is $\mathrm{Binomial}(n, \pi_j)$, the other
outcomes collapsing into a single failure. A group of coordinates is again
Dirichlet or multinomial, but only after the remaining mass is collapsed into
a coordinate of its own, which is a different object, so that case is refused.

```{r}
#| label: dirichlet-demo
dd <- dirichlet_distrib(4)
theta_d <- list(mean_alr1 = 0.4, mean_alr2 = -0.3, mean_alr3 = 0.8, phi = 12)

# the mean is a parameter of the family, and the covariance is singular
mv_location(dd, theta_d)
signif(eigen(mv_sigma(dd, theta_d), only.values = TRUE)$values, 3)

# a marginal is a beta with the same concentration
unlist(mv_marginal(dd, theta_d, which = 2)$theta)
```

```{r}
#| label: multinomial-demo
dn <- multinomial_distrib(3, size = 5)
theta_n <- list(probs_alr1 = 0.3, probs_alr2 = -0.2)

# the support is finite, so the normalization is a sum and not a sample
supp <- mv_support(dn, theta_n)
nrow(supp)
sum(distrib_pdf(dn, supp, theta_n))
```
### Marginals and the pairs plot {#sec-mv-marginal}

Integrating a density over the coordinates one is not interested in has no
closed form for most families. It has one for the elliptical families, where
the marginal belongs to the same family with the mean subsetted and the matrix
reduced to the corresponding block. For the gaussian this is the classical
result; for the $t$ the marginal is a $t$ on the subset with the **same**
degrees of freedom, which follows from @eq-mvt-mixture, since conditioning on
$V$ leaves a gaussian whose marginal is a gaussian with the same mixing
variable. It also has one for the two families of @sec-mv-simplex, where the
marginal changes family rather than staying in it.

This is what makes a picture of a multivariate density possible, because every
panel of a pairs plot displays a marginal. The diagonal shows the univariate
marginal density of each coordinate, the lower triangle the contours of each
bivariate marginal, and, when the plot is drawn from a fit, the observations
and their kernel density estimate are drawn in the same frame. The kernel
estimate is the comparison that does not assume the model.

Beyond three coordinates the panel matrix stops being readable, so a larger
distribution is rejected with a message suggesting a choice of coordinates.
A family with no closed-form marginal is rejected outright, since a
numerical marginal would be a different object under the same name.

```{r}
#| label: mvt-pairs
#| fig-width: 7
#| fig-height: 6.4
#| fig-cap: "A two-dimensional Student t with four degrees of freedom: the marginal densities on the diagonal and the contours of the bivariate density below it."
plot(dt, theta_t)
```

Two quantities of the univariate case have no multivariate counterpart and are
refused rather than approximated. A distribution function on $\mathbb{R}^p$
would be an orthant probability, which is not a one-dimensional integral and
does not reduce to one; a quantile function would need an ordering of
$\mathbb{R}^p$, and there is none. Everything that depends on either is
therefore unavailable here, which is why a multivariate distribution does not
inherit the defaults of `continuous_distrib` but sits on a class of its own.

### Fitting {#sec-mv-fitting}

A fit proceeds exactly as in @sec-fitting. The objective is the sum of the log
densities over the rows of $Y$, the optimization happens on the link scale,
and the standard errors come from the inverse of the information by the delta
method. For the gaussian the maximum likelihood estimator is known in closed
form, the sample mean and the sample second moment about it, which makes it
the natural check on the whole apparatus.

```{r}
#| label: mvgaussian-fit
set.seed(11)
y_big <- distrib_rng(d, 2000, theta)
fit <- fit_distrib(d, y_big)

mu_hat <- colMeans(y_big)
s_hat <- crossprod(sweep(y_big, 2L, mu_hat)) / nrow(y_big)

c(
  loglik_gap = as.numeric(logLik(fit)) -
    sum(distrib_pdf(d, y_big, as.list(stats::setNames(
      c(mu_hat, parameters7::param_free(d@param, s_hat)), d@params
    )), log = TRUE)),
  mean_gap = max(abs(mv_location(d, coef(fit)) - mu_hat)),
  cov_gap = max(abs(mv_sigma(d, coef(fit)) - s_hat))
)
```

### What a fit reports {#sec-mv-reporting}

The free values are coordinates chosen so that an optimizer can move without
constraint, and they are not the quantities a reader of a fit examines. The
logarithm of a diagonal entry of a Cholesky factor has an estimate and a
standard error, and neither answers a question about the data. The quantities
of interest for a fitted multivariate gaussian are the standard deviations
and the correlations, the decomposition $\Sigma = D R D$ with $D$ the
diagonal matrix of standard deviations.

Carrying the uncertainty onto them is the delta method, which
@sec-fit-delta already applies to a univariate parameter. Writing
$g$ for the map from the parameters to the quantities of interest,

$$
\widehat{\operatorname{Var}}\{g(\hat\theta)\}
  = J \, \widehat{\operatorname{Var}}(\hat\theta) \, J^\top,
\qquad
J = \frac{\partial g}{\partial \theta},
$$ {#eq-mv-delta}

and the Jacobian is closed form. With $s_j = \sqrt{\Sigma_{jj}}$ and
$\rho_{jk} = \Sigma_{jk}/(s_j s_k)$,

$$
\frac{\partial s_j}{\partial \eta_k} = \frac{A_k[j,j]}{2 s_j},
\qquad
\frac{\partial \rho_{jk}}{\partial \eta_l}
  = \frac{A_l[j,k]}{s_j s_k}
    - \frac{\rho_{jk}}{2}
      \left(\frac{A_l[j,j]}{\Sigma_{jj}}
           + \frac{A_l[k,k]}{\Sigma_{kk}}\right),
$$ {#eq-mv-sdcor-jacobian}

with $A_k = \partial\Sigma/\partial\eta_k$ the parameter's own derivative,
which @eq-mv-parameter already requires it to supply. Nothing new has to be
computed.

The interval each quantity gets is built on the scale that keeps it in the set
it belongs to and then mapped back, which is the discipline of
@sec-fit-scale applied to a derived quantity rather than to a parameter: a
standard deviation is intervalled on the log scale and a correlation on
Fisher's $z = \operatorname{artanh}\rho$. An interval built on the raw scale
would put a correlation above one whenever the estimate is large and the
sample small.

```{r}
#| label: mvgaussian-fit-print
fit
```

A gaussian written in the precision reports the same standard deviations and
correlations, since those are properties of the law and not of the
parametrization, and adds the two readings that belong to $\Omega$ itself. The
first is the **conditional variance**

$$
\frac{1}{\Omega_{jj}} \;=\; \operatorname{Var}(Y_j \mid Y_{-j}),
$$ {#eq-mv-conditional-variance}

the variance that remains after regressing that coordinate on all the others,
so that its ratio to the marginal variance is $1 - R_j^2$ for that regression.
The variance is the quantity with that reading and its square root is not,
which is why it is the variance that is reported. The second is the **partial
correlation**

$$
-\frac{\Omega_{jk}}{\sqrt{\Omega_{jj}\Omega_{kk}}}
  \;=\; \operatorname{Cor}(Y_j, Y_k \mid Y_{-jk}),
$$ {#eq-mv-partial-correlation}

the correlation of two coordinates given all the others, which is zero exactly
where the precision has a zero --- and that is the reason for parametrizing a
model in $\Omega$ at all, since a zero there is a conditional independence.
In two dimensions there is nothing to condition on and the partial correlation
is the correlation, so it is not printed twice.

Both parametrizations describe the same law, so a fit of one and a fit of the
other must reach the same maximized likelihood and the same information
criteria. Two runs that disagree on this invariance have not both converged.

A structured matrix has quantities of its own, and they are not recoverable
from the standard deviations and correlations. The covariance of a stationary
autoregression of order $q$ is described by a marginal variance and $q$
partial autocorrelations, and the coefficients $\phi$ that a reader of an
autoregression expects appear nowhere in $\Sigma$ at all. A parameter
therefore declares the quantities it is about, together with the Jacobian of
the map from its free vector, and the fit carries the variance matrix onto
them by @eq-mv-delta exactly as it does for the decomposition. For the
autoregression that Jacobian costs nothing: the Levinson-Durbin recursion of
@sec-parameter-arp already propagates every derivative in compiled code, so
the derivatives of the coefficients with respect to every free value are
read off the first-order block of arrays the family has computed anyway.

Each quantity names the scale its interval is built on, by the same rule as
before: a variance on the logarithm, a partial autocorrelation on Fisher's
$z$, and a coefficient on the identity. The coefficients take the identity
scale because the stationary region is not a box, so no scalar
transformation expresses the constraint they are jointly under.

```{r}
#| label: mv-structure-block
d_ar <- mvgaussian_distrib(6, sigma = parameters7::autoregressive(6, order = 2))
theta_ar <- as.list(stats::setNames(
  c(rep(0, 6), log(3), atanh(0.7), atanh(-0.3)), d_ar@params
))
set.seed(15)
tab_ar <- mv_summary(fit_distrib(d_ar, distrib_rng(d_ar, 500, theta_ar)))
round(tab_ar[attr(tab_ar, "block") == "Autoregressive structure", ], 4)
```

The lag-one correlation of an autoregression is its first partial
autocorrelation, so `pacf1` and `cor_v1_v2` are the same number reached two
ways, while the coefficients are new. Note also that `phi2` and `pacf2` share
an estimate and a standard error, since the last coefficient of an order-$q$
autoregression is the last partial autocorrelation, and yet their intervals
differ: they are built on different scales.

For the Student $t$ the matrix is the scale matrix, so the diagonal quantities
are not standard deviations of the response and are named `scale_sd` to say so.
The correlations need no such qualification: the covariance is
$\nu\Sigma/(\nu-2)$, and a positive multiple of a matrix has the same
correlations.

```{r}
#| label: mvt-fit-summary
set.seed(12)
y_t <- distrib_rng(dt, 800, theta_t)
mv_summary(fit_distrib(dt, y_t, method = optimizers7::newton()))
```

```{r}
#| label: consistency-03e
#| include: false
assert_multivariate_ok()
```
