## Several dimensions {#sec-multivariate}

```{r}
#| label: setup-03e
#| include: false
source("R/_setup.R")
source("R/multivariate-certificates.R")
```

The families of @sec-distributions all place their mass on the real line, and
an observation is a number. This section describes what changes when an
observation is a vector of $p$ numbers, and what does not. The likelihood
theory of @sec-likelihood is unaffected: the score, the observed information,
the Bartlett identities and the change of scale are statements about a
parametric family and say nothing about the dimension of its support. What
changes is the parametrisation, because a multivariate family carries a matrix
where a univariate one carries a variance, and a matrix is constrained.

The two families presented here are the elliptical ones the toolkit needs
first. @kotz2004 is the reference for the multivariate $t$ and @lange1989 for
its use as a robust replacement for the gaussian.

### The response and the parameters {#sec-mv-parameters}

A multivariate distribution places its mass on $\mathbb{R}^p$, and a sample of
$n$ observations is the $n \times p$ matrix $Y$ whose $i$th row is $y_i$. The
number of observations is the row count rather than the length, which is the
only structural change the response requires.

The parameters need more care. A $p$-variate gaussian has a mean vector
$\mu \in \mathbb{R}^p$ and a covariance matrix $\Sigma$, and $\Sigma$ is not a
free object: it must be symmetric and positive definite. A parameter of this
kind is what @sec-covstructs7 is about. All this section needs from it is the
contract: a **structure** is a map

$$
\eta \;\longmapsto\; M(\eta), \qquad \eta \in \mathbb{R}^{d},
$$ {#eq-mv-structure}

from an unconstrained vector of length $d$ to a symmetric matrix in the
constrained set, supplying the derivatives
$A_k = \partial M / \partial \eta_k$ and
$A_{kl} = \partial^2 M / \partial \eta_k \partial \eta_l$, the
log-determinant and its derivatives, and the solves $M^{-1}B$. The map is
onto the set and one-to-one, so $\eta$ ranges over the whole of
$\mathbb{R}^{d}$ with no constraint of its own.

The parameters of a multivariate distribution are then the $p$ components of
the mean together with the $d$ free values of the structure, and every one of
them is a **scalar**. This is what lets the rest of the package apply
unchanged. A parameter vector $\theta$ is a named list of numbers, the
derivative components are keyed by parameter names exactly as in
@sec-likelihood, and the enumeration that produces `hess_names()` produces
the same strings whether the parameters are $(\mu, \sigma)$ or
$(\mu_1, \mu_2, \log L_{11}, \log L_{22}, L_{21})$. Nothing in `align_theta()`,
in the link scale of @sec-link-scale, or in the fitting routine of
@sec-fitting has to be told that the family is multivariate.

Every one of these scalars is unconstrained, so every link is the identity.
The constraint that would otherwise need a link has been moved into the
structure, where it belongs: it is a constraint on the matrix as a whole and
not on any one entry, and a scalar link cannot express it. A multivariate
gaussian therefore has a link scale that coincides with its parameter scale.
A multivariate $t$ does not, because its degrees of freedom are a genuine
positive scalar and carry a log link.

There is one restriction, and it comes from what a parameter means here. In a
univariate family a parameter may vary from observation to observation, which
is what a regression does. Here it may not: the structure's free values
describe one matrix for the whole sample. A model in which the matrix varies
by observation belongs to the layer that owns the regression, not to the
distribution, and the constructor refuses a parameter of length greater than
one rather than silently using its first entry.

The names of the free values carry a prefix saying which matrix they describe,
`sigma_` for a covariance and `omega_` for a precision. The name a structure
gives a free value records how the matrix is built and not which matrix it is,
so the same structure on the two sides of a model would otherwise produce
identical parameter names for two different models. The distribution applies
the prefix rather than the structure, since the structure does not know which
side it has been handed to.

### The gaussian {#sec-mvgaussian}

The $p$-variate gaussian has density

$$
f(y) \;=\; (2\pi)^{-p/2} \, |\Sigma|^{-1/2}
  \exp\!\left\{ -\tfrac{1}{2} (y-\mu)^\top \Sigma^{-1} (y-\mu) \right\},
$$ {#eq-mvgaussian-density}

so that with $r = y - \mu$ and $w = \Sigma^{-1} r$ the log-density is
$\ell = -\tfrac{p}{2}\log(2\pi) - \tfrac{1}{2}\log|\Sigma|
- \tfrac{1}{2} r^\top w$.

Differentiating with respect to the mean gives $\partial \ell/\partial \mu = w$,
one component per coordinate. For the matrix directions, the derivative of the
quadratic form follows from $\partial_k \Sigma^{-1} = -\Sigma^{-1} A_k
\Sigma^{-1}$, which is the derivative of an inverse, so that
$\partial_k (r^\top \Sigma^{-1} r) = -w^\top A_k w$. Together with the
structure's own $\partial_k \log|\Sigma|$,

$$
\ell^{(\mu_a)} = w_a, \qquad
\ell^{(\eta_k)} = -\tfrac{1}{2}\,\partial_k \log|\Sigma|
   \;+\; \tfrac{1}{2}\, w^\top A_k w .
$$ {#eq-mvgaussian-score}

Both pieces on the right come from the structure, and neither depends on which
family of matrices it belongs to. The distribution is written once and works
with any structure that satisfies the contract.

The second derivatives follow by differentiating again. The mean block is
$-\Sigma^{-1}$, constant in $y$; the mixed block is
$\ell^{(\mu_a \eta_k)} = -(\Sigma^{-1} A_k w)_a$; and the matrix block is

$$
\ell^{(\eta_k \eta_l)} = -\tfrac{1}{2}\,\partial_{kl} \log|\Sigma|
  + \tfrac{1}{2}\, w^\top A_{kl} w
  - w^\top A_l \Sigma^{-1} A_k w .
$$ {#eq-mvgaussian-hessian}

The expected information is where the multivariate case is more pleasant than
it looks. Taking expectations uses only $\mathbb{E}[w] = 0$ and
$\mathbb{E}[w w^\top] = \Sigma^{-1}$. The first kills the mixed block exactly,
so the mean and the matrix are orthogonal; the second turns each quadratic
form into a trace. Both second-derivative terms of the matrix block then
collapse, and

$$
\mathbb{E}\!\left[\ell^{(\mu_a \mu_b)}\right] = -(\Sigma^{-1})_{ab},
\qquad
\mathbb{E}\!\left[\ell^{(\mu_a \eta_k)}\right] = 0,
\qquad
\mathbb{E}\!\left[\ell^{(\eta_k \eta_l)}\right]
  = -\tfrac{1}{2}\,\mathrm{tr}\!\left(\Sigma^{-1} A_k \Sigma^{-1} A_l\right).
$$ {#eq-mvgaussian-information}

The second derivatives $A_{kl}$ have disappeared, and so has the second
derivative of the log-determinant. The information needs only the first
derivatives of the structure, which is what makes Fisher scoring cheaper here
than a Newton step, and it is what @sec-fit-scoring already predicted in
one dimension.

```{r}
#| label: mvgaussian-demo
d <- mvgaussian_distrib(2)
d@params

theta <- list(mu1 = 0.5, mu2 = -0.3, sigma_log_L1 = 0.1, sigma_log_L2 = -0.2, sigma_L2.1 = 0.4)
mv_sigma(d, theta)
```

A gaussian may equally be written in the precision $\Omega = \Sigma^{-1}$, and
which of the two a model parametrises is a modelling decision rather than a
notational one: a structure imposed on $\Sigma$ is a different family from the
same structure imposed on $\Omega$, since the inverse of a compound-symmetry
matrix is compound symmetry while the inverse of an AR(1) is not AR(1). The
constructor takes the structure on either side, and everything above still
applies, because $\partial \Sigma / \partial \eta_k = -\Sigma A_k \Sigma$ when
$A_k$ is a derivative of the precision. The two parametrisations describe the
same law when the matrices are matched.

```{r}
#| label: mvgaussian-precision
do <- mvgaussian_distrib(2, struct_omega = covstructs7::log_cholesky(2))
eta_o <- covstructs7::struct_free(do@struct, solve(mv_sigma(d, theta)))
theta_o <- as.list(stats::setNames(c(0.5, -0.3, eta_o), do@params))

y <- rbind(c(0, 0), c(1, -1), c(-0.5, 0.8))
max(abs(distrib_pdf(do, y, theta_o, log = TRUE) -
        distrib_pdf(d, y, theta, log = TRUE)))
```

### The Student t {#sec-mvstudent-t}

The multivariate $t$ replaces the exponential tail with a polynomial one. With
$q = (y - \mu)^\top \Sigma^{-1} (y - \mu)$ its density is

$$
f(y) \;=\;
\frac{\Gamma\!\left(\frac{\nu+p}{2}\right)}
     {\Gamma\!\left(\frac{\nu}{2}\right) (\nu\pi)^{p/2} |\Sigma|^{1/2}}
\left( 1 + \frac{q}{\nu} \right)^{-(\nu+p)/2}.
$$ {#eq-mvt-density}

Here $\Sigma$ is the **scale** matrix and not the covariance. The covariance is
$\nu\Sigma/(\nu-2)$ where it exists, and for $\nu \le 2$ it does not exist at
all while the distribution is perfectly well defined; the mean exists for
$\nu > 1$. Keeping the two apart is what allows a heavy-tailed family to be
described and fitted at degrees of freedom where its second moment is
infinite, so the package returns the scale matrix from `mv_sigma()` and the
covariance from `variance()`.

The distribution is the scale mixture

$$
Y \mid V \sim N_p\!\left(\mu,\; \tfrac{\nu}{V}\,\Sigma \right),
\qquad V \sim \chi^2_\nu,
$$ {#eq-mvt-mixture}

which is how the package generates from it and where the shape of its score
comes from. Writing

$$
c \;=\; \frac{\nu + p}{\nu + q}
$$ {#eq-mvt-weight}

for the weight an observation receives, the derivatives with respect to the
mean and the structure are those of the gaussian, @eq-mvgaussian-score, with
every term involving the data multiplied by $c$:

$$
\ell^{(\mu_a)} = c \, w_a, \qquad
\ell^{(\eta_k)} = -\tfrac{1}{2}\,\partial_k \log|\Sigma|
   \;+\; \tfrac{1}{2}\, c \, w^\top A_k w .
$$ {#eq-mvt-score}

The weight is what makes the family resistant. An observation with a large
Mahalanobis distance has a large $q$, hence a small $c$, and contributes
proportionally less to the estimate of both the location and the scale; as
$\nu \to \infty$ the weight tends to $1$ uniformly and the gaussian is
recovered. A gaussian fitted to data containing a few distant rows has no such
mechanism and inflates $\Sigma$ until it covers them.

The degrees of freedom contribute one further component,

$$
\ell^{(\nu)} = \tfrac{1}{2}\left[
  \psi\!\left(\tfrac{\nu+p}{2}\right) - \psi\!\left(\tfrac{\nu}{2}\right)
  - \frac{p}{\nu} - \log\!\left(1 + \frac{q}{\nu}\right)
  + \frac{(\nu+p)\,q}{\nu(\nu+q)} \right],
$$ {#eq-mvt-score-nu}

with $\psi$ the digamma function. The first three terms come from the
normalising constant and do not involve the data; the last two are the
derivative of the kernel. Since $\nu > 0$ it carries a log link, and the
derivative on the link scale is $\nu\,\ell^{(\nu)}$ by the chain rule of
@sec-link-scale, with no change to the rest.

The expected information of the $t$ has no closed form in the package, so it
is approximated by one of the strategies of @sec-approx, and
`fit_distrib()` accepts the choice through its `approx` argument. It accepts
it only for a family that has no closed form, since for one that has, the
argument would be silently ignored and the fit would report a method it did
not use.

```{r}
#| label: mvt-demo
dt <- mvstudent_t_distrib(2)
theta_t <- list(mu1 = 0.5, mu2 = -0.3, sigma_log_L1 = 0.1, sigma_log_L2 = -0.2,
                sigma_L2.1 = 0.4, nu = 4)

# the scale matrix is what the parametrisation carries; the covariance is a
# moment, and at nu = 4 it is twice as large
mv_sigma(dt, theta_t)
variance(dt, theta_t)
```

### Marginals and what a picture shows {#sec-mv-marginal}

Integrating a density over the coordinates one is not interested in has no
closed form for most families. It has one for the elliptical families, where
the marginal belongs to the same family with the mean subsetted and the matrix
reduced to the corresponding block. For the gaussian this is the classical
result; for the $t$ the marginal is a $t$ on the subset with the **same**
degrees of freedom, which follows from @eq-mvt-mixture, since conditioning on
$V$ leaves a gaussian whose marginal is a gaussian with the same mixing
variable.

This is what makes a picture of a multivariate density possible, because every
panel of a pairs plot displays a marginal. The diagonal shows the univariate
marginal density of each coordinate, the lower triangle the contours of each
bivariate marginal, and, when the plot is drawn from a fit, the observations
and their kernel density estimate are drawn in the same frame. The kernel
estimate is the comparison that does not assume the model.

Beyond three coordinates the panel matrix stops being readable, so a larger
distribution is refused with the suggestion of choosing coordinates rather
than being drawn illegibly. A family with no closed-form marginal is refused
outright: a numerical marginal would be a different object wearing the same
name.

```{r}
#| label: mvt-pairs
#| fig-width: 7
#| fig-height: 6.4
#| fig-cap: "A two-dimensional Student t with four degrees of freedom: the marginal densities on the diagonal and the contours of the bivariate density below it."
plot(dt, theta_t)
```

Two quantities of the univariate case have no multivariate counterpart and are
refused rather than approximated. A distribution function on $\mathbb{R}^p$
would be an orthant probability, which is not a one-dimensional integral and
does not reduce to one; a quantile function would need an ordering of
$\mathbb{R}^p$, and there is none. Everything that depends on either is
therefore unavailable here, which is why a multivariate distribution does not
inherit the defaults of `continuous_distrib` but sits on a class of its own.

### Fitting {#sec-mv-fitting}

A fit proceeds exactly as in @sec-fitting. The objective is the sum of the log
densities over the rows of $Y$, the optimisation happens on the link scale,
and the standard errors come from the inverse of the information by the delta
method. For the gaussian the maximum likelihood estimator is known in closed
form, the sample mean and the sample second moment about it, which makes it
the natural check on the whole apparatus.

```{r}
#| label: mvgaussian-fit
set.seed(11)
y_big <- distrib_rng(d, 2000, theta)
fit <- fit_distrib(d, y_big)

mu_hat <- colMeans(y_big)
s_hat <- crossprod(sweep(y_big, 2L, mu_hat)) / nrow(y_big)

c(
  loglik_gap = as.numeric(logLik(fit)) -
    sum(distrib_pdf(d, y_big, as.list(stats::setNames(
      c(mu_hat, covstructs7::struct_free(d@struct, s_hat)), d@params
    )), log = TRUE)),
  mean_gap = max(abs(mv_location(d, coef(fit)) - mu_hat)),
  cov_gap = max(abs(mv_sigma(d, coef(fit)) - s_hat))
)
```

### What a fit reports {#sec-mv-reporting}

The free values are coordinates chosen so that an optimiser can move without
constraint, and they are not quantities anybody reads. The logarithm of a
diagonal entry of a Cholesky factor has an estimate and a standard error, and
neither answers a question about the data. What a reader of a fitted
multivariate gaussian wants is the standard deviations and the correlations,
which are the decomposition $\Sigma = D R D$ with $D$ the diagonal matrix of
standard deviations.

Carrying the uncertainty onto them is the delta method, which
@sec-fit-delta already applies to a univariate parameter. Writing
$g$ for the map from the parameters to the quantities of interest,

$$
\widehat{\operatorname{Var}}\{g(\hat\theta)\}
  = J \, \widehat{\operatorname{Var}}(\hat\theta) \, J^\top,
\qquad
J = \frac{\partial g}{\partial \theta},
$$ {#eq-mv-delta}

and the Jacobian is closed form. With $s_j = \sqrt{\Sigma_{jj}}$ and
$\rho_{jk} = \Sigma_{jk}/(s_j s_k)$,

$$
\frac{\partial s_j}{\partial \eta_k} = \frac{A_k[j,j]}{2 s_j},
\qquad
\frac{\partial \rho_{jk}}{\partial \eta_l}
  = \frac{A_l[j,k]}{s_j s_k}
    - \frac{\rho_{jk}}{2}
      \left(\frac{A_l[j,j]}{\Sigma_{jj}}
           + \frac{A_l[k,k]}{\Sigma_{kk}}\right),
$$ {#eq-mv-sdcor-jacobian}

with $A_k = \partial\Sigma/\partial\eta_k$ the structure's own derivative,
which @eq-mv-structure already requires it to supply. Nothing new has to be
computed.

The interval each quantity gets is built on the scale that keeps it in the set
it belongs to and then mapped back, which is the discipline of
@sec-fit-scale applied to a derived quantity rather than to a parameter: a
standard deviation is intervalled on the log scale and a correlation on
Fisher's $z = \operatorname{artanh}\rho$. An interval built on the raw scale
would put a correlation above one whenever the estimate is large and the
sample small.

```{r}
#| label: mvgaussian-fit-print
fit
```

A gaussian written in the precision reports the same standard deviations and
correlations, since those are properties of the law and not of the
parametrisation, and adds the two readings that belong to $\Omega$ itself. The
first is the **conditional variance**

$$
\frac{1}{\Omega_{jj}} \;=\; \operatorname{Var}(Y_j \mid Y_{-j}),
$$ {#eq-mv-conditional-variance}

the variance that remains after regressing that coordinate on all the others,
so that its ratio to the marginal variance is $1 - R_j^2$ for that regression.
The variance is the quantity with that reading and its square root is not,
which is why it is the variance that is reported. The second is the **partial
correlation**

$$
-\frac{\Omega_{jk}}{\sqrt{\Omega_{jj}\Omega_{kk}}}
  \;=\; \operatorname{Cor}(Y_j, Y_k \mid Y_{-jk}),
$$ {#eq-mv-partial-correlation}

the correlation of two coordinates given all the others, which is zero exactly
where the precision has a zero --- and that is the reason for parametrising a
model in $\Omega$ at all, since a zero there is a conditional independence.
In two dimensions there is nothing to condition on and the partial correlation
is the correlation, so it is not printed twice.

Both parametrisations describe the same law, so a fit of one and a fit of the
other reach the same maximised likelihood and the same information criteria.
That is an invariance worth checking rather than assuming: two runs that
disagree on it have not both converged.

For the Student $t$ the matrix is the scale matrix, so the diagonal quantities
are not standard deviations of the response and are named `scale_sd` to say so.
The correlations need no such qualification: the covariance is
$\nu\Sigma/(\nu-2)$, and a positive multiple of a matrix has the same
correlations.

```{r}
#| label: mvt-fit-summary
set.seed(12)
y_t <- distrib_rng(dt, 800, theta_t)
mv_summary(fit_distrib(dt, y_t, method = optimizers7::newton(), tol = 1e-8))
```

```{r}
#| label: consistency-03e
#| include: false
assert_multivariate_ok()
```
