## In practice {#sec-distrib-practice}

```{r}
#| label: setup-03f
#| include: false
source("R/_setup.R")
```

This section shows the package in use: the census of the shipped families,
the full set of questions one distribution object answers, the wrappers, and
a censored fit assembled from the pieces of @sec-cdf-derivatives. Every
block is executed against the packages loaded from the working tree.

### The families that ship {#sec-distrib-census}

The catalog of @sec-distrib-catalog lists every univariate family with its
density; the count below is computed from the namespace, so it cannot fall
out of date:

```{r}
#| label: dp-census
ctors <- grep("_distrib$", getNamespaceExports("distributions7"), value = TRUE)
skip <- c("check_distrib", "fit_distrib", "continuous_distrib",
          "discrete_distrib", "multivariate_distrib", "mvgaussian_distrib",
          "mvstudent_t_distrib", "dirichlet_distrib", "multinomial_distrib")
c(univariate = length(setdiff(ctors, skip)),
  multivariate = 4L)
```

### One object, every question {#sec-distrib-tour}

A single distribution answers the probability questions, the moment
questions and the derivative questions of @sec-distributions. On the
mean-variance Gamma:

```{r}
#| label: dp-tour-prob
d <- gamma2_distrib()
th <- list(mu = 3, sigma2 = 2)

q <- distrib_quantile(d, c(0.1, 0.5, 0.9), th)
distrib_cdf(d, q, th)                       # the round trip
set.seed(1)
y <- distrib_rng(d, 5, th)

c(mean = mean(d, th), variance = variance(d, th),
  skewness = skewness(d, th), kurtosis = kurtosis(d, th))
```

The derivatives come in the four directions the chapter derived: with
respect to the parameters on both scales, with respect to the response, and
mixed:

```{r}
#| label: dp-tour-deriv
str(distrib_gradient(d, y, th))                    # parameter scale
str(distrib_gradient(d, y, th, scale = "link"))    # d/d eta, eta = g(theta)
str(distrib_hess_y(d, y, th))                      # d2 l / dy2
str(distrib_cross_y(d, y, th))                     # d2 l / dy dtheta
names(distrib_deriv4(d, y[1], th))                 # the fourth-order components
```

### The wrappers at work {#sec-distrib-wrappers-practice}

Each wrapper of @sec-transformations returns an ordinary distribution, and
the machinery above applies to it unchanged:

```{r}
#| label: dp-wrappers
# hold a parameter at a value
ray <- fixed(weibull1_distrib(), sigma = 2)
ray@params

# restrict to an interval and renormalize
tg <- truncated(gamma2_distrib(), lower = 1, upper = 6)
distrib_cdf(tg, c(1, 6), th)

# absolute value, and a change of variables with its own name
mean(folded(gaussian1_distrib()), list(mu = 0, sigma = 1)) - sqrt(2 / pi)
ig <- transformation(gamma2_distrib(), inverse_transform(),
                     new_name = "inverse gamma")
ig@distrib_name

# a point mass at zero next to a density
zag <- zero_adjusted(gamma2_distrib())
distrib_atoms(zag, c(th, za = 0.25))
```

### A censored fit from the pieces {#sec-distrib-censored-practice}

No front end for censored likelihoods ships yet, and none is needed to fit
one: the score of an observation censored at the ceiling $c$ is the
derivative of $\log\{1 - F(c)\}$, which `distrib_grad_cdf()` returns. The
sample below is Weibull with every value above $c = 5.5$ recorded as $5.5$:

```{r}
#| label: dp-censored
library(optimizers7)
set.seed(7)
dw <- weibull1_distrib()
th_true <- list(mu = 4, sigma = 1.6)
y0 <- distrib_rng(dw, 400, th_true)
cl <- 5.5
cens <- y0 > cl
y <- pmin(y0, cl)
mean(cens)                                  # the censored fraction

nll <- function(eta) {
  th <- list(mu = exp(eta[1]), sigma = exp(eta[2]))
  -(sum(distrib_pdf(dw, y[!cens], th, log = TRUE)) +
      sum(log1p(-distrib_cdf(dw, rep(cl, sum(cens)), th))))
}
gr <- function(eta) {
  th <- list(mu = exp(eta[1]), sigma = exp(eta[2]))
  g <- distrib_gradient(dw, y[!cens], th, scale = "link")
  gc <- distrib_grad_cdf(dw, rep(cl, sum(cens)), th,
                         lower.tail = FALSE, log = TRUE)
  -(c(sum(g$mu), sum(g$sigma)) +
      c(sum(gc$mu) * th$mu, sum(gc$sigma) * th$sigma))
}

fit_c <- minimize(lbfgs(), nll, par = c(log(mean(y)), 0), gr = gr)
naive <- fit_distrib(dw, y)                  # censoring ignored

rbind(true = unlist(th_true),
      censored_likelihood = exp(fit_c@par),
      censoring_ignored = unlist(coef(naive)))
```

The censored likelihood recovers the parameters; treating the ceiling
values as exact biases the scale downward, since the largest observations
have been replaced by smaller ones.

### The fit and its extractors {#sec-distrib-fit-practice}

`fit_distrib()` returns an object whose extractors answer on either scale,
and whose `method` argument accepts any optimizer of @sec-optimizers7:

```{r}
#| label: dp-fit
set.seed(3)
yg <- distrib_rng(d, 500, th)
f1 <- fit_distrib(d, yg)
f2 <- fit_distrib(d, yg, method = lbfgs(criterion = crit_grad(1e-7)))

c(logLik(f1) - logLik(f2))                   # the same maximum
coef(f1)
coef(f1, scale = "link")
confint(f1)                                  # mapped through the inverse link
confint(f1, level = 0.99, scale = "link")

sims <- simulate(f1, nsim = 3, seed = 1)
vapply(sims, mean, numeric(1))
```
