## Descent, and what a line search must guarantee {#sec-descent}

```{r}
#| label: setup-04a
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Take a differentiable $f : \mathbb{R}^{p} \to \mathbb{R}$ to be minimised, and
suppose we are standing at a point $x$ where the gradient $g = \nabla f(x)$ is
not zero. The whole of this section concerns one iteration: choosing a direction
$d$, choosing how far along it to go, and arriving somewhere better. It sounds
like a small question, and the reason it is not is that "somewhere better" can
be satisfied infinitely often without ever getting anywhere.

### Directions that go downhill {#sec-descent-directions}

The first-order expansion of $f$ along $d$ is

$$f(x + s d) = f(x) + s\, g^\top d + o(s), \qquad s \to 0^{+},$$

so the sign of $g^\top d$ decides everything about small steps. If
$g^\top d < 0$ then $f(x + sd) < f(x)$ for all sufficiently small $s > 0$, and
$d$ is called a **descent direction**; if $g^\top d > 0$ the same expansion says
every small step uphill; and if $g^\top d = 0$ the first-order term tells us
nothing at all.

::: {#def-descent}
## Descent direction

A vector $d \in \mathbb{R}^{p}$ is a descent direction for $f$ at $x$ when
$g^\top d < 0$, where $g = \nabla f(x)$.
:::

The simplest choice is $d = -g$, for which $g^\top d = -\lVert g \rVert^{2} < 0$
whenever $g \neq 0$; this is the steepest descent direction, in the sense that
it minimises $g^\top d$ over all $d$ of a given Euclidean length. Every method in
@sec-curvature instead takes

$$d = -B g$$

for some symmetric positive definite $B$, and the reason that always works is
immediate:

$$g^\top d = -g^\top B g < 0 \quad \text{for } g \neq 0,$$

which is precisely the definition of positive definiteness. This is worth
stating separately, because it is the single fact that connects the two halves
of this chapter: **the requirement that a curvature model be positive definite
is not a preference, it is what makes the direction it produces usable at all.**
When Newton's method repairs an indefinite Hessian in @sec-newton, it is
defending this inequality and nothing else.

### Why "the objective decreased" is not enough {#sec-descent-armijo}

Having a descent direction, we must choose a step length $s > 0$. The naive
rule — accept $s$ if $f(x + sd) < f(x)$ — is not sufficient, and the reason is
worth seeing concretely rather than in the abstract.

Consider minimising $f(x) = x^{2}$ from $x_0 = 1$, and suppose the step lengths
are chosen so that the iterates are

$$x_k = (-1)^{k}\left(1 + 2^{-k}\right).$$

Each iterate is closer to the origin than the last, so $f$ decreases at every
step; and yet $\lvert x_k \rvert \to 1$, so the sequence converges to nothing
useful and the gradient never approaches zero. The steps were too *long*
relative to the decrease they achieved: the method kept overshooting through the
minimum and being rewarded for it. A symmetric failure occurs when the steps are
too short, $x_k = 1 + 2^{-k}$ say, where $f$ again decreases at every step and
again converges away from the minimiser.

The repair for the first failure is to demand a decrease **proportional to what
the slope promised**. Along $d$ the model predicts a decrease of
$-s\, g^\top d$ per unit of nothing; asking for a fixed fraction of it gives:

::: {#def-armijo}
## Armijo, or sufficient decrease

For $c_1 \in (0, 1)$, a step length $s$ satisfies the Armijo condition when

$$f(x + s d) \le f(x) + c_1 s\, g^\top d. \tag{A}$$
:::

Since $g^\top d < 0$ the right-hand side is strictly below $f(x)$, so (A)
implies decrease; but it implies more, because the amount demanded grows with
$s$. A long step must earn its length. The condition is satisfiable: the
expansion above gives

$$\frac{f(x + sd) - f(x)}{s} \longrightarrow g^\top d < c_1 g^\top d
\quad \text{as } s \to 0^{+},$$

the inequality holding because $c_1 < 1$ and $g^\top d$ is negative, so (A)
holds for all sufficiently small $s$. That is what licenses **backtracking**:
start from a trial length, and if (A) fails, shrink and try again. The procedure
terminates, and the package's `armijo()` is exactly this.

What backtracking does not address is the second failure, steps that are too
short. Nothing in (A) forbids a sequence of ever-tinier accepted steps. To rule
that out we need a condition that a *small* step violates, and the natural one
looks at the slope rather than the value.

::: {#def-wolfe}
## The Wolfe conditions

For $0 < c_1 < c_2 < 1$, a step length $s$ satisfies the **weak Wolfe**
conditions when (A) holds together with

$$\nabla f(x + sd)^\top d \ge c_2\, g^\top d, \tag{W}$$

and the **strong Wolfe** conditions when (W) is strengthened to

$$\bigl\lvert \nabla f(x + sd)^\top d \bigr\rvert
  \le c_2 \bigl\lvert g^\top d \bigr\rvert. \tag{W$'$}$$
:::

Condition (W) says the slope along $d$ must have risen — become less steeply
negative — by a definite factor. At $s = 0$ the slope is $g^\top d$, and (W) is
violated there because $c_2 < 1$; so (W) forces the step away from zero, which
is exactly what (A) fails to do. The two conditions pull in opposite directions
and a step satisfying both exists whenever $f$ is bounded below along $d$: as
$s$ grows, either $f$ eventually violates (A), and at the first such $s$ a
mean-value argument produces a point where both hold, or $f$ decreases for ever
and the problem is unbounded.

The strong form (W$'$) additionally forbids the slope from becoming largely
positive, which keeps the accepted point nearer a local minimum of $s \mapsto
f(x + sd)$. It costs nothing extra, since a search that can test (W) has already
evaluated the gradient there, and the package's `wolfe()` implements it by the
usual two-phase procedure: **bracket** an interval known to contain an
acceptable point by expanding until (A) fails or the slope turns positive, then
**zoom** into it by interpolation until both conditions hold.

### The theorem that makes it worth the trouble {#sec-descent-zoutendijk}

The point of the Wolfe conditions is not tidiness. They buy a convergence
statement that sufficient decrease alone cannot.

::: {#thm-zoutendijk}
## Zoutendijk

Let $f$ be bounded below and continuously differentiable on an open set
containing the level set $\{x : f(x) \le f(x_0)\}$, with $\nabla f$ Lipschitz
there. If each iterate is $x_{k+1} = x_k + s_k d_k$ with $d_k$ a descent
direction and $s_k$ satisfying the Wolfe conditions, then

$$\sum_{k \ge 0} \cos^{2}\!\theta_k \, \lVert g_k \rVert^{2} < \infty,
\qquad \text{where} \quad
\cos\theta_k = \frac{-g_k^\top d_k}{\lVert g_k \rVert\, \lVert d_k \rVert}.$$
:::

::: {.proof}
From (W), $(g_{k+1} - g_k)^\top d_k \ge (c_2 - 1) g_k^\top d_k$, and Lipschitz
continuity of the gradient with constant $L$ gives
$(g_{k+1} - g_k)^\top d_k \le L s_k \lVert d_k \rVert^{2}$. Combining,

$$s_k \ge \frac{c_2 - 1}{L} \cdot \frac{g_k^\top d_k}{\lVert d_k \rVert^{2}},$$

which is positive because $c_2 < 1$ and $g_k^\top d_k < 0$. Substituting this
lower bound on $s_k$ into (A),

$$f_{k+1} \le f_k - \frac{c_1 (1 - c_2)}{L}
  \frac{(g_k^\top d_k)^{2}}{\lVert d_k \rVert^{2}}
= f_k - \frac{c_1 (1 - c_2)}{L} \cos^{2}\!\theta_k \lVert g_k \rVert^{2}.$$

Summing over $k$ and using that $f$ is bounded below makes the total decrease
finite, which is the claim.
:::

The consequence is the one to remember. A convergent series has terms tending to
zero, so $\cos^{2}\theta_k \lVert g_k \rVert^{2} \to 0$; hence **if the
directions never become orthogonal to the gradient** — if $\cos\theta_k \ge
\delta > 0$ for all $k$ — then $\lVert g_k \rVert \to 0$. The line search
supplies half of the guarantee and the direction supplies the other half, and
neither is sufficient alone. This is why the package makes the line search an
object carried by the optimiser rather than a fixed piece of each algorithm: the
theorem is about the pair.

For $d_k = -B_k g_k$ the angle condition becomes a statement about the condition
number of $B_k$,

$$\cos\theta_k = \frac{g_k^\top B_k g_k}{\lVert g_k \rVert \,\lVert B_k g_k \rVert}
\ \ge\ \frac{\lambda_{\min}(B_k)}{\lambda_{\max}(B_k)},$$

so a curvature model whose eigenvalues stay in a fixed positive band inherits
global convergence. That bound is the reason @sec-newton floors the eigenvalues
of a repaired Hessian rather than merely making them positive, and the reason
@sec-bfgs takes the curvature condition seriously enough to skip an update that
would violate it.

### What each search costs {#sec-descent-cost}

The two searches differ in what they must evaluate. Backtracking needs only
values of $f$: it computes $g^\top d$ once, at the current point, and then tries
$f(x + sd)$ for a shrinking sequence of $s$. A Wolfe search must test (W), which
mentions $\nabla f(x + sd)$, so it evaluates the gradient at every trial point.

That is not a defect but a trade. The gradient it evaluates at the accepted
point is the one the next iteration needs anyway, so the marginal cost is the
gradients spent at *rejected* trials; and in exchange the method gets
@thm-zoutendijk and, more practically, a step length that satisfies the
curvature condition — which is exactly the condition @sec-bfgs needs in order to
keep its approximation positive definite. A quasi-Newton method paired with a
Wolfe search never has to skip an update. Paired with backtracking it sometimes
does.

```{r}
#| label: line-search-cost
#| echo: false
f  <- function(p) 100 * (p[2] - p[1]^2)^2 + (1 - p[1])^2
gr <- function(p) c(-400 * p[1] * (p[2] - p[1]^2) - 2 * (1 - p[1]),
                    200 * (p[2] - p[1]^2))

rows <- lapply(list(armijo = optimizers7::armijo(),
                    wolfe  = optimizers7::wolfe()), function(ls) {
  r <- optimizers7::minimize(optimizers7::bfgs(line_search = ls),
                             f, c(-1.2, 1), gr = gr)
  data.frame(iterations = r@iterations,
             `f evaluations` = as.integer(r@counts[["f"]]),
             `gradients` = as.integer(r@counts[["g"]]),
             `gradients per iteration` =
               round(r@counts[["g"]] / r@iterations, 2),
             check.names = FALSE)
})
knitr::kable(
  cbind(`line search` = names(rows), do.call(rbind, rows)),
  row.names = FALSE, align = "lrrrr",
  caption = "BFGS on Rosenbrock from the usual starting point, under the two line searches."
)
```

Read the last column. A backtracking search spends almost exactly one gradient
per iteration — the one it needs to form the next direction, and no more, since
the trials it rejects cost only values. A Wolfe search spends about a third
again, the surplus being the gradients evaluated at trial points it then
declines.

The comparison is not, however, one of cost against nothing. The Wolfe run needs
noticeably fewer iterations, because a step length satisfying the curvature
condition is a better step length, so the total gradient count differs by rather
less than the per-iteration figure suggests. What is bought with the remainder is
@thm-zoutendijk, which backtracking alone does not supply, and the guarantee in
@sec-bfgs that the curvature approximation never has to reject an update.
