## Descent directions and line searches {#sec-descent}

```{r}
#| label: setup-04a
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Let $f : \mathbb{R}^{p} \to \mathbb{R}$ be a differentiable function to be
minimised, and let $x$ be a point where the gradient $g = \nabla f(x)$ is not
zero. This section studies a single iteration: how to choose a direction $d$,
and how to choose the step length along it. Both choices need care, because a
sequence of steps can decrease $f$ at every iteration and still fail to
converge to a minimiser; the conditions introduced below exist to rule out
exactly that failure.

### Descent directions {#sec-descent-directions}

The first-order expansion of $f$ along $d$ is

$$f(x + s d) = f(x) + s\, g^\top d + o(s), \qquad s \to 0^{+},$$

so for small steps the sign of $g^\top d$ determines the behaviour of $f$. If
$g^\top d < 0$, then $f(x + sd) < f(x)$ for all sufficiently small $s > 0$; if
$g^\top d > 0$, every small step increases $f$; if $g^\top d = 0$, the
first-order term gives no information.

::: {#def-descent}
## Descent direction

A vector $d \in \mathbb{R}^{p}$ is a descent direction for $f$ at $x$ when
$g^\top d < 0$, where $g = \nabla f(x)$.
:::

The simplest descent direction is $d = -g$, for which
$g^\top d = -\lVert g \rVert^{2} < 0$ whenever $g \neq 0$; among all
directions of a given Euclidean length it minimises $g^\top d$, and for this
reason it is called the steepest descent direction. Most of the methods in
@sec-curvature instead use

$$d = -B g$$

for some symmetric positive definite matrix $B$. Such a direction is always a
descent direction, because

$$g^\top d = -g^\top B g < 0 \quad \text{for } g \neq 0$$

by the definition of positive definiteness. This fact connects the two halves of the chapter. A curvature model must be
positive definite because otherwise the direction it produces may not be a
descent direction at all, and when Newton's method repairs an indefinite
Hessian in @sec-newton, it does so to preserve exactly this inequality.

Conjugate gradients, in @sec-cg-bb, does not have the form $-Bg$ for any
stored matrix, so its direction is not guaranteed to be a descent direction in
advance. The implementation therefore tests $g^\top d < 0$ at each step and
replaces the direction by $-g$ when the test fails.

### The sufficient-decrease and Wolfe conditions {#sec-descent-armijo}

Given a descent direction, a step length $s > 0$ must be chosen. Accepting any
$s$ with $f(x + sd) < f(x)$ is not sufficient, and a concrete example shows
why.

Consider minimising $f(x) = x^{2}$ from $x_0 = 1$, with step lengths chosen so
that the iterates are

$$x_k = (-1)^{k}\left(1 + 2^{-k}\right).$$

Each iterate is closer to the origin than the last, so $f$ decreases at every
step, and yet $\lvert x_k \rvert \to 1$: the sequence converges to a point
that is not a minimiser, and the gradient never approaches zero. The steps
were too long relative to the decrease they achieved, since each one
overshoots the minimum. The opposite failure occurs when the steps are too
short: with $x_k = 1 + 2^{-k}$, $f$ again decreases at every step and again
converges away from the minimiser.

The first failure is prevented by requiring a decrease proportional to the one
the local slope predicts. Along $d$ the first-order model predicts a decrease
of $-s\, g^\top d$; requiring a fixed fraction of it gives the following
condition.

::: {#def-armijo}
## Sufficient decrease

For $c_1 \in (0, 1)$, a step length $s$ satisfies the Armijo condition when

$$f(x + s d) \le f(x) + c_1 s\, g^\top d. \tag{A}$$
:::

Since $g^\top d < 0$, the right-hand side is strictly below $f(x)$, so (A)
implies a decrease; and because the required decrease grows with $s$, a long
step is accepted only when it achieves a proportionally large decrease. The
condition is always satisfiable: from the first-order expansion,

$$\frac{f(x + sd) - f(x)}{s} \longrightarrow g^\top d < c_1 g^\top d
\quad \text{as } s \to 0^{+},$$

where the inequality holds because $c_1 < 1$ and $g^\top d$ is negative, so
(A) holds for all sufficiently small $s$. This justifies **backtracking**:
start from a trial length, and if (A) fails, shrink the step and try again.
The procedure terminates, and the package's `armijo()` implements exactly
this. The condition is due to @armijo1966.

Backtracking does not address the second failure, steps that are too short:
nothing in (A) forbids a sequence of ever smaller accepted steps. Ruling that
out requires a condition that a small step violates, and the natural quantity
to examine is the slope rather than the value.

::: {#def-wolfe}
## The Wolfe conditions

For $0 < c_1 < c_2 < 1$, a step length $s$ satisfies the **weak Wolfe**
conditions when (A) holds together with

$$\nabla f(x + sd)^\top d \ge c_2\, g^\top d, \tag{W}$$

and the **strong Wolfe** conditions when (W) is strengthened to

$$\bigl\lvert \nabla f(x + sd)^\top d \bigr\rvert
  \le c_2 \bigl\lvert g^\top d \bigr\rvert. \tag{W$'$}$$
:::

Condition (W) requires the slope along $d$ to have risen — to have become less
steeply negative — by a definite factor. At $s = 0$ the slope is $g^\top d$
and (W) is violated because $c_2 < 1$, so (W) forces the step away from zero,
which is precisely what (A) cannot do. The two conditions act in opposite directions, and a step satisfying both
exists whenever $f$ is bounded below along $d$. As $s$ grows, either $f$
eventually violates (A), in which case a mean-value argument at the first
such $s$ produces a point where both conditions hold, or $f$ decreases
forever and the problem is unbounded.

Both forms are due to @wolfe1969, and @morethuente1994 gives the algorithm
that has become the standard way of finding a step that satisfies them. The
strong form (W$'$) additionally forbids the slope from becoming strongly
positive, which keeps the accepted point near a local minimum of
$s \mapsto f(x + sd)$. It costs nothing extra, because a search able to test
(W) has already evaluated the gradient at the trial point. The package's
`wolfe()` implements the strong form by the usual two-phase procedure: it
**brackets** an interval known to contain an acceptable point, expanding until
(A) fails or the slope turns positive, and then **zooms** into the interval by
interpolation until both conditions hold.

### Zoutendijk's theorem {#sec-descent-zoutendijk}

The Wolfe conditions matter because they yield a global convergence result
that sufficient decrease alone does not.

::: {#thm-zoutendijk}
## Zoutendijk

Let $f$ be bounded below and continuously differentiable on an open set
containing the level set $\{x : f(x) \le f(x_0)\}$, with $\nabla f$ Lipschitz
there. If each iterate is $x_{k+1} = x_k + s_k d_k$ with $d_k$ a descent
direction and $s_k$ satisfying the Wolfe conditions, then

$$\sum_{k \ge 0} \cos^{2}\!\theta_k \, \lVert g_k \rVert^{2} < \infty,
\qquad \text{where} \quad
\cos\theta_k = \frac{-g_k^\top d_k}{\lVert g_k \rVert\, \lVert d_k \rVert}.$$
:::

The argument is due to @zoutendijk1970 and is given in full in @nocedal2006.
Its structure is simple. The curvature condition (W) bounds the step length
from below, and the Lipschitz continuity of the gradient bounds it from
above. Substituting the lower bound into the sufficient-decrease condition
(A) shows that each iteration decreases $f$ by at least a fixed multiple of
$\cos^{2}\!\theta_k \lVert g_k \rVert^{2}$. Since $f$ is bounded below, the
total decrease is finite, and the series converges.

The consequence used throughout the chapter is the following. A convergent
series has terms tending to zero, so
$\cos^{2}\theta_k \lVert g_k \rVert^{2} \to 0$. If the directions never
become orthogonal to the gradient, that is, if
$\cos\theta_k \ge \delta > 0$ for all $k$, then $\lVert g_k \rVert \to 0$.
The line search supplies one half of the guarantee and the direction supplies
the other, and neither is sufficient alone. This is also why the package makes
the line search an object carried by the optimiser rather than a fixed part of
each algorithm: the theorem is a statement about the pair.

For $d_k = -B_k g_k$ the angle condition becomes a statement about the
condition number of $B_k$,

$$\cos\theta_k = \frac{g_k^\top B_k g_k}{\lVert g_k \rVert \,\lVert B_k g_k \rVert}
\ \ge\ \frac{\lambda_{\min}(B_k)}{\lambda_{\max}(B_k)},$$

so a curvature model whose eigenvalues stay within a fixed positive band
inherits global convergence. This bound explains two implementation choices
that appear later: @sec-newton floors the eigenvalues of a repaired Hessian
instead of merely making them positive, and @sec-bfgs skips an update that
would violate the curvature condition.

### Cost of the two searches {#sec-descent-cost}

The two searches differ in what they evaluate. Backtracking needs only values
of $f$: it computes $g^\top d$ once, at the current point, and then evaluates
$f(x + sd)$ for a shrinking sequence of $s$. A Wolfe search must test (W),
which involves $\nabla f(x + sd)$, so it evaluates the gradient at every trial
point.

The extra cost buys something specific. The gradient evaluated at the accepted
point is needed by the next iteration in any case, so the additional cost is
only the gradients spent at rejected trials. In exchange the method obtains @thm-zoutendijk and a step length that
satisfies the curvature condition, which is what @sec-bfgs needs to keep its
approximation positive definite. A quasi-Newton method paired with a Wolfe
search never has to skip an update; one paired with backtracking sometimes
does.

```{r}
#| label: line-search-cost
#| echo: false
f  <- function(p) 100 * (p[2] - p[1]^2)^2 + (1 - p[1])^2
gr <- function(p) c(-400 * p[1] * (p[2] - p[1]^2) - 2 * (1 - p[1]),
                    200 * (p[2] - p[1]^2))

rows <- lapply(list(armijo = optimizers7::armijo(),
                    wolfe  = optimizers7::wolfe()), function(ls) {
  r <- optimizers7::minimize(optimizers7::bfgs(line_search = ls),
                             f, c(-1.2, 1), gr = gr)
  data.frame(iterations = r@iterations,
             `f evaluations` = as.integer(r@counts[["f"]]),
             `gradients` = as.integer(r@counts[["g"]]),
             `gradients per iteration` =
               round(r@counts[["g"]] / r@iterations, 2),
             check.names = FALSE)
})
knitr::kable(
  cbind(`line search` = names(rows), do.call(rbind, rows)),
  row.names = FALSE, align = "lrrrr",
  caption = "BFGS on Rosenbrock from the usual starting point, under the two line searches."
)
```

The last column summarises the difference. A backtracking search spends almost
exactly one gradient per iteration — the one it needs to form the next
direction — because its rejected trials cost only function values. A Wolfe
search spends about a third more, the surplus being the gradients evaluated at
trial points it then rejects.

The comparison is not purely a cost. The Wolfe run needs noticeably fewer iterations, because a step length
satisfying the curvature condition is a better step length, so the total
gradient counts differ by less than the per-iteration figures suggest. The
difference that remains pays for @thm-zoutendijk, which backtracking alone
does not provide, and for the guarantee in @sec-bfgs that the curvature
approximation never has to reject an update.
