## Second-order and quasi-Newton methods {#sec-curvature}

```{r}
#| label: setup-04b
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Every method in this section chooses its direction in the same way: it builds
a quadratic model of $f$ around the current point and steps towards the
minimizer of that model. Writing $g = \nabla f(x)$ and letting $M$ denote the
matrix that plays the role of the curvature, the model is

$$m(d) = f(x) + g^\top d + \tfrac{1}{2} d^\top M d,$$

and setting $\nabla m(d) = g + M d = 0$ gives

$$d = -M^{-1} g.$$

The methods differ in what they use for $M$. This section presents them in
order of how much they store: the exact Hessian, then an approximation to its
inverse of the same size, then the last few pairs of vectors that
approximation was built from, then a single vector, then a single number. The
methods remain effective even near the bottom of the list.

### Newton's method {#sec-newton}

The most direct choice is the true Hessian, $M = \nabla^{2} f(x)$, which makes
$m$ the second-order Taylor expansion and the step exact for a quadratic
objective. Near a strict local minimum the resulting iteration converges
quadratically, which no other method in this chapter achieves, and this is the
reason `newton()` exists.

Away from a minimum the Hessian may be indefinite or singular, and the step is
then wrong in a qualitative sense, not just inaccurate. If $H$ has a negative eigenvalue $\lambda < 0$ with eigenvector $v$, the
model $m$ is unbounded below along $v$, and its stationary point is a saddle
rather than a minimum. Moreover, the direction $d = -H^{-1} g$ can satisfy
$g^\top d > 0$, which by @def-descent points uphill. The line search then finds no acceptable step and
the run stops for a reason unrelated to the actual problem.

$H$ must therefore be replaced by a positive definite matrix near it. Two
standard repairs are available, both described in @nocedal2006 and
@dennis1983.

The **eigenvalue repair** takes the symmetric decomposition
$H = V \Lambda V^\top$ and replaces $\Lambda$ by $\tilde\Lambda$ with

$$\tilde\lambda_i = \max(\lambda_i, \varepsilon),$$

for a floor $\varepsilon > 0$, giving $\tilde H = V \tilde\Lambda V^\top$.
This repair is minimal in a precise sense: it changes only the problematic
eigenvalues and leaves the others exactly as they were, so the curvature that
the Hessian estimated correctly is preserved. It is the default. The floor
serves a definite purpose: with $\varepsilon = 0$ the repaired matrix could be
singular, and with $\varepsilon$ very small its condition number could destroy
the angle bound at the end of @sec-descent-zoutendijk. The floor keeps
$\cos\theta_k$ away from zero.

The **ridge repair** takes $\tilde H = H + \tau I$, with $\tau$ the smallest
multiple of the identity that makes the result positive definite, found by
attempting a Cholesky factorization and doubling $\tau$ on failure. It is
cheaper, because it needs no eigendecomposition, but less precise: adding
$\tau I$ shifts every eigenvalue, so the directions in which the Hessian was
informative are distorted together with the ones in which it was not.

Both repairs guarantee the property that matters: the matrix actually inverted
is positive definite, so the direction is a descent direction, and the line
search can find an acceptable step.

::: {.callout-note}
## Numerical Hessians

`newton()` accepts an objective with no analytic Hessian and obtains one by
differencing the gradient once. When the gradient is analytic this is a single
numerical differentiation and the result is accurate. When the gradient is itself obtained by finite differences, the Hessian
becomes a difference of differences and the errors compound. This is the only
place in the toolkit where two layers of numerical differentiation are
composed, and it cannot be avoided: a Hessian computed from function values
alone has no other form. For this reason an objective that supplies neither derivative is better
served by `bfgs()`, which never forms a Hessian.
:::

### The BFGS update {#sec-bfgs}

The Hessian is expensive: $p(p+1)/2$ second derivatives per iteration, plus a
factorization costing $O(p^{3})$. Quasi-Newton methods avoid this cost by
building an approximation from information they already have, namely the
gradients computed at successive iterates.

The available information is a **secant pair**. After moving from $x$ to
$x_{+} = x + \alpha d$, write

$$s = x_{+} - x, \qquad y = \nabla f(x_{+}) - \nabla f(x).$$

For a quadratic with Hessian $H$ these satisfy $y = H s$ exactly, and for a
general $f$ the mean value theorem gives $y = \bar{H} s$ for an average of
Hessians along the segment. In both cases the pair describes the curvature in
one direction, and it is natural to require the approximation to reproduce it.

::: {#def-secant}
## The secant equation

An approximation $B_{+}$ to the Hessian satisfies the secant equation when
$B_{+} s = y$; equivalently, an approximation $H_{+}$ to its inverse satisfies

$$H_{+}\, y = s.$$
:::

These are $p$ equations for the $p(p+1)/2$ unknowns of a symmetric matrix, so
they do not determine $H_{+}$ by themselves. Two further requirements almost
do. The first is that $H_{+}$ be symmetric and positive definite, which
@sec-descent showed is necessary for the direction to be usable. The second is
that $H_{+}$ be as close as possible to the current $H$, because the pair
carries information about one direction only and should not overwrite what is
known about the others.

Making "close" precise in a particular weighted Frobenius norm yields a unique
answer, the BFGS update:

::: {#thm-bfgs}
## The BFGS update

With $\rho = 1 / (y^\top s)$, the update

$$H_{+} = \left(I - \rho\, s y^\top\right) H \left(I - \rho\, y s^\top\right)
          + \rho\, s s^\top$$

satisfies the secant equation, is symmetric, and is positive definite whenever
$H$ is and $y^\top s > 0$.
:::

The three claims are short calculations, set out in @nocedal2006 together with
the variational characterization that singles this update out; the update was
found independently by four authors in 1970, of whom @broyden1970 is cited
here. Applying $H_{+}$ to $y$ and using $\rho\,y^\top s = 1$ collapses the
middle factor to zero and leaves $s$; symmetry is visible in the formula,
since both terms are symmetric; and for positive definiteness one writes
$w = v - \rho\,y(s^\top v)$ and obtains

$$v^\top H_{+} v = w^\top H w + \rho\,(s^\top v)^{2},$$

whose two terms are non-negative and cannot vanish together, because
$s^\top v = 0$ would force $w = v \neq 0$.

The last calculation is where the condition $y^\top s > 0$ enters, and the
condition has a direct meaning. The quantity

$$y^\top s = \bigl(\nabla f(x_{+}) - \nabla f(x)\bigr)^\top (x_{+} - x)$$

is positive exactly when the function curves upward along the step, that is,
when the observed curvature is consistent with a minimum. On a non-convex
objective it can be negative, and then no positive definite matrix satisfies
the secant equation for that pair: the pair reports negative curvature, and a
positive definite matrix cannot reproduce it.

When this happens the package **skips** the pair: whenever
$y^\top s \le \varepsilon \lVert s \rVert \lVert y \rVert$, the approximation
is left unchanged and the skip is recorded in the trace. The alternative,
damping the pair towards one that does satisfy the condition, would inject
curvature that was never observed, so the pair is discarded instead.

A Wolfe search removes the problem entirely. Condition (W) states
$\nabla f(x_{+})^\top d \ge c_2 \nabla f(x)^\top d$; subtracting
$\nabla f(x)^\top d$ from both sides and multiplying by $\alpha > 0$ gives

$$y^\top s \ \ge\ (c_2 - 1)\, \alpha\, \nabla f(x)^\top d \ > \ 0,$$

because $c_2 < 1$ and the directional derivative is negative. A quasi-Newton
method paired with a Wolfe search therefore never skips an update. This is the
practical reason `bfgs()` defaults to the Wolfe search, while `gd()`, which
has no approximation to protect, defaults to backtracking.

This advantage is a guarantee rather than an observed difference. On every
smooth problem in the standard collection, and on a deliberately non-convex
one as well, backtracking also skips no updates, because the curvature
condition is simply never violated there. What the Wolfe search removes is the
possibility of a skipped update, and the difference appears only on objectives
less well behaved than any in this book.

### Limited-memory BFGS {#sec-lbfgs}

BFGS stores a $p \times p$ matrix, which is acceptable for the handful of
parameters of a single distribution and impossible for the thousands of a
model with covariates. L-BFGS instead keeps the last $m$ secant pairs,
typically with $m$ between 5 and 20, and never forms the matrix; the method is
due to @nocedal1980, with the convergence analysis and large-scale evidence in
@liu1989.

The method rests on the observation that the BFGS update in @thm-bfgs is a sum
of outer products, so applying $H_k$ to a vector requires only inner products
with the stored pairs. Unrolling the recursion gives the **two-loop
recursion**: initialize $q = g$, run backwards through the pairs computing

$$\alpha_i = \rho_i\, s_i^\top q, \qquad q \leftarrow q - \alpha_i y_i,$$

scale once by $\gamma = (s_{k}^\top y_{k}) / (y_{k}^\top y_{k})$ using the
most recent pair, then run forwards computing

$$\beta_i = \rho_i\, y_i^\top q, \qquad q \leftarrow q + (\alpha_i - \beta_i) s_i,$$

after which $q = H_k g$. The cost is $O(mp)$ in time and $O(mp)$ in storage,
against $O(p^{2})$ for both in BFGS.

The scaling $\gamma$ is the one part of the recursion that is a choice rather
than an identity. It sets the initial approximation at each iteration to
$\gamma I$, where $\gamma$ is the Rayleigh quotient of the most recent pair,
an estimate of the inverse curvature along the direction most recently
traveled. Without this scaling the recursion would start from $I$, whose
scale has no relation to the problem, and the first step of each iteration
would be badly sized.

The cost of limited memory is that older curvature information is discarded.
On a problem whose Hessian is nearly constant this costs almost nothing; on
one where the curvature varies sharply, discarding can even help, because
stale pairs describe a region the iterate has already left.

```{r}
#| label: memory-cost
#| echo: false
set.seed(4)
p <- 60
A <- crossprod(matrix(rnorm(p * p), p)) + diag(p)
b <- rnorm(p)
fq  <- function(x) as.numeric(0.5 * t(x) %*% A %*% x - sum(b * x))
gq  <- function(x) as.numeric(A %*% x - b)
x0 <- rep(0, p)

rows <- list(
  BFGS = optimizers7::minimize(optimizers7::bfgs(), fq, x0, gr = gq),
  `L-BFGS (m = 5)` = optimizers7::minimize(optimizers7::lbfgs(memory = 5), fq, x0, gr = gq),
  `L-BFGS (m = 20)` = optimizers7::minimize(optimizers7::lbfgs(memory = 20), fq, x0, gr = gq)
)
knitr::kable(
  data.frame(method = names(rows),
             iterations = vapply(rows, function(r) r@iterations, numeric(1)),
             `stored numbers` = c(p * p, 2 * 5 * p, 2 * 20 * p),
             `gap from the optimum` =
               formatC(vapply(rows, function(r) r@value, numeric(1)) -
                         fq(solve(A, b)), format = "e", digits = 1),
             check.names = FALSE),
  row.names = FALSE, align = "lrrr",
  caption = "A quadratic in 60 variables. L-BFGS reaches the same answer while storing a fraction as much."
)
```

### Conjugate gradients and Barzilai-Borwein {#sec-cg-bb}

Two further methods store even less, and between them they cover the case a
statistical model most often presents: a parameter vector too long for any
$p \times p$ matrix.

**Conjugate gradients** keeps one vector. Its direction is

$$d_k = -g_k + \beta_k d_{k-1},$$

which is not of the form $-M^{-1}g$, and does not need to be. Recall from
@sec-descent-directions why steepest descent is slow: under an exact line
search consecutive directions are orthogonal, because the search stops where
$g_{k+1}^\top d_k = 0$, so on a narrow valley the iterates cross the floor
repeatedly instead of traveling along it. The remedy is to combine each new
gradient with the previous direction, and the entire method is the choice of
how much weight $\beta_k$ to give the previous direction.

On a quadratic with Hessian $H$ and an exact line search, the classical
choices of $\beta$ make the directions *conjugate*, meaning
$d_i^\top H d_j = 0$ for $i \neq j$. A set of $p$ conjugate directions spans
the space, and the one-dimensional minimizations along them are independent of
one another, so the method terminates in exactly $p$ steps without ever
forming $H$. The linear case is @hestenes1952; the nonlinear case begins with
@fletcher1964, and @hager2006 surveys the later developments.

The four formulas offered here agree on a quadratic and differ everywhere
else:

$$\beta^{FR} = \frac{g_k^\top g_k}{g_{k-1}^\top g_{k-1}}, \quad
  \beta^{PR} = \frac{g_k^\top y_k}{g_{k-1}^\top g_{k-1}}, \quad
  \beta^{HS} = \frac{g_k^\top y_k}{d_{k-1}^\top y_k}, \quad
  \beta^{DY} = \frac{g_k^\top g_k}{d_{k-1}^\top y_k},$$

with $y_k = g_k - g_{k-1}$ as before, due respectively to @fletcher1964,
@polak1969, @hestenes1952 and @dai1999. Fletcher–Reeves has the cleanest
convergence theory and a known practical fault: after a poor step it can stall
for many iterations. Polak–Ribière recovers immediately, because a small $y$ sends $\beta$
towards zero and the method back to steepest descent. Its known theoretical
non-convergence is repaired by clamping $\beta$ at zero; since $\beta = 0$
restarts the method at steepest descent, the clamp is recorded in the trace
as a restart.

**Barzilai–Borwein** keeps one number. The direction is $d = -\alpha g$ with

$$\alpha^{BB1} = \frac{s^\top s}{s^\top y}, \qquad
  \alpha^{BB2} = \frac{s^\top y}{y^\top y},$$

the two Rayleigh quotients of the secant pair. Each is a scalar estimate of
the inverse curvature along the direction just traveled, so the method is
@sec-bfgs with everything discarded except one number. On a quadratic, where
the curvature is constant, the estimate is exact and the second step lands on
the answer. The method is due to @barzilai1988, and @raydan1997 established
that it is practical on large problems.

```{r}
#| label: memory-ladder
#| echo: false
f  <- function(p) 100 * (p[2] - p[1]^2)^2 + (1 - p[1])^2
gr <- function(p) c(-400 * p[1] * (p[2] - p[1]^2) - 2 * (1 - p[1]),
                    200 * (p[2] - p[1]^2))
os <- list(
  `newton`  = optimizers7::newton(),
  `bfgs`    = optimizers7::bfgs(),
  `lbfgs`   = optimizers7::lbfgs(),
  `cg`      = optimizers7::cg(maxit = 5000),
  `bb`      = optimizers7::bb(maxit = 5000),
  `gd`      = optimizers7::gd(maxit = 20000)
)
store <- c("p(p+1)/2", "p^2", "2mp", "2p", "1", "0")
rows <- lapply(os, function(o)
  optimizers7::minimize(o, f, c(-1.2, 1), gr = gr))
knitr::kable(
  data.frame(method = names(os),
             `curvature stored` = store,
             iterations = vapply(rows, function(r) r@iterations, numeric(1)),
             converged = vapply(rows, function(r) r@converged, logical(1)),
             check.names = FALSE),
  row.names = FALSE, align = "lrrl",
  caption = "Rosenbrock from the usual start, by how much curvature the method keeps."
)
```

The table shows that the amount of stored curvature matters less than its
presence. Keeping the exact Hessian, a full approximation to its inverse, twenty pairs
of vectors, one vector or one number changes the iteration count only
moderately. Keeping nothing at all is the one configuration that fails:
gradient descent is the only row that does not converge, and conjugate
gradients repairs it with a single remembered direction.

Every run can record its own path. With `keep_trace = TRUE` the result
carries one row per iteration: the objective value, the gradient norm, the
step taken and any safeguard that fired. Plotting the recorded values shows
the difference between the rows of the table as trajectories rather than as
two numbers:

```{r}
#| label: trace-plot
#| fig-height: 4
#| fig-cap: "The objective along the run, Rosenbrock from the usual start. BFGS descends in a handful of iterations; gradient descent is still traveling at 500."
rb <- optimizers7::minimize(optimizers7::bfgs(keep_trace = TRUE),
                            f, c(-1.2, 1), gr = gr)
rg <- optimizers7::minimize(optimizers7::gd(maxit = 500, keep_trace = TRUE),
                            f, c(-1.2, 1), gr = gr)
plot(rg@trace$iteration, log10(rg@trace$value), type = "l",
     xlab = "iteration", ylab = "log10 objective")
lines(rb@trace$iteration, log10(rb@trace$value), lty = 2)
legend("topright", c("gradient descent", "BFGS"), lty = c(1, 2), bty = "n")
```

For Barzilai–Borwein the iteration count in the table depends on the line
search as much as on the stored number. The method makes progress through
steps that temporarily increase the objective in order to be aligned with the
curvature later, and the sufficient-decrease condition of @def-armijo rejects
exactly those steps. The standard remedy, due to @grippo1986, is a nonmonotone
line search, which requires improvement over the worst of the last several
values instead of over the current one; @birgin2000 pairs it with
Barzilai-Borwein steps and is the reference for the combination. `bb()` uses
it by default, and with it
the method reaches the minimum of Rosenbrock in 68 iterations and 77 objective
evaluations, against 82 and 186 under plain backtracking.
