## Curvature, and three ways to obtain it {#sec-curvature}

```{r}
#| label: setup-04b
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Every method in this section chooses its direction the same way: build a
quadratic model of $f$ around the current point, and step to the minimiser of
that model. Writing $g = \nabla f(x)$ and letting $M$ stand for whatever matrix
plays the part of the curvature, the model is

$$m(d) = f(x) + g^\top d + \tfrac{1}{2} d^\top M d,$$

and setting $\nabla m(d) = g + M d = 0$ gives

$$d = -M^{-1} g.$$

The three methods differ only in what they use for $M$, and the differences are
about cost and about what happens when the true curvature is unhelpful.

### Newton's method, and the repair it needs {#sec-newton}

The obvious choice is the true Hessian, $M = \nabla^{2} f(x)$, which makes
$m$ the second-order Taylor expansion and the step exact for a quadratic. Near a
strict local minimum this converges quadratically, which no other method here
achieves, and it is why `newton()` exists.

Away from such a point the Hessian may be indefinite or singular, and then the
step is not merely inaccurate but wrong in kind. If $H$ has a negative
eigenvalue $\lambda < 0$ with eigenvector $v$, the model $m$ is unbounded below
along $v$ and its stationary point is a saddle, not a minimum; and worse, taking
$d = -H^{-1} g$ can produce a direction with $g^\top d > 0$, which by
@def-descent points uphill. The line search then finds no acceptable step and
the run stops for a reason that has nothing to do with the problem.

So $H$ must be replaced by a positive definite matrix near it. Two repairs are
standard and both are available.

The **eigenvalue repair** takes the symmetric decomposition $H = V \Lambda
V^\top$ and replaces $\Lambda$ by $\tilde\Lambda$ with

$$\tilde\lambda_i = \max(\lambda_i, \varepsilon),$$

for a floor $\varepsilon > 0$, giving $\tilde H = V \tilde\Lambda V^\top$. This
is minimal in a precise sense — it changes the eigenvalues that are a problem
and leaves the others exactly alone, so the curvature the Hessian got right is
preserved — and it is the default. Note that the floor is not cosmetic: taking
$\varepsilon = 0$ would leave a singular matrix, and taking it very small leaves
a matrix whose condition number destroys the angle bound at the end of
@sec-descent-zoutendijk. The floor is what keeps $\cos\theta_k$ away from zero.

The **ridge repair** instead takes $\tilde H = H + \tau I$ with $\tau$ the
smallest multiple of the identity making the result positive definite, found by
attempting a Cholesky factorisation and doubling $\tau$ on failure. It is
cheaper, since it needs no eigendecomposition, but it is blunter: adding $\tau I$
shifts *every* eigenvalue, so directions in which the Hessian was informative are
distorted along with the ones in which it was not.

Both repairs share the property that matters. Whatever $H$ was, the matrix
actually inverted is positive definite, so the direction is a descent direction,
so the line search has something to find.

::: {.callout-note}
## A Hessian obtained by differencing

`newton()` accepts an objective with no analytic Hessian and differences the
gradient once to obtain one. When the gradient is itself analytic that is a
single numerical differentiation and the result is good. When the gradient is
*also* differenced, the Hessian is a difference of differences, and the error
compounds — which is the one place in the toolkit where two layers of numerical
differentiation are composed. It is unavoidable, since a Hessian from values
alone must be, but it is why an objective supplying neither derivative is better
served by `bfgs()`, which never forms a Hessian at all.
:::

### The quasi-Newton idea {#sec-bfgs}

The Hessian is expensive: $p(p+1)/2$ second derivatives per iteration, and a
factorisation costing $O(p^{3})$. The quasi-Newton methods refuse to pay it and
build an approximation from what they already have — the gradients computed at
successive iterates.

The information available is a **secant pair**. Having moved from $x$ to
$x_{+} = x + s d$, write

$$s = x_{+} - x, \qquad y = \nabla f(x_{+}) - \nabla f(x).$$

For a quadratic with Hessian $H$ these satisfy $y = H s$ exactly, and for a
general $f$ the mean value theorem gives $y = \bar{H} s$ for some average of
Hessians along the segment. Either way the pair says what the curvature does in
one direction, and it is natural to ask an approximation to reproduce it.

::: {#def-secant}
## The secant equation

An approximation $B_{+}$ to the Hessian satisfies the secant equation when
$B_{+} s = y$; equivalently, an approximation $H_{+}$ to its inverse satisfies

$$H_{+}\, y = s.$$
:::

This is $p$ equations for the $p(p+1)/2$ unknowns of a symmetric matrix, so it
does not determine $H_{+}$. Two further requirements do, very nearly. The first
is that $H_{+}$ be **symmetric and positive definite**, which @sec-descent
showed is what makes the direction usable. The second is that it be **as close
as possible to $H$**, on the grounds that the pair carries information about one
direction and should not be allowed to overwrite what is known about the others.

Making "close" precise in a particular weighted Frobenius norm yields a unique
answer, and it is the BFGS update:

::: {#thm-bfgs}
## The BFGS update

With $\rho = 1 / (y^\top s)$, the update

$$H_{+} = \left(I - \rho\, s y^\top\right) H \left(I - \rho\, y s^\top\right)
          + \rho\, s s^\top$$

satisfies the secant equation, is symmetric, and is positive definite whenever
$H$ is and $y^\top s > 0$.
:::

::: {.proof}
For the secant equation, apply $H_{+}$ to $y$ and use $\rho\, y^\top s = 1$:

$$H_{+} y = \left(I - \rho\, s y^\top\right) H
            \underbrace{\left(y - \rho\, y (s^\top y)\right)}_{= \, y - y \, = \, 0}
            + \rho\, s (s^\top y) = s .$$

Symmetry is immediate from the form, both terms being symmetric. For positive
definiteness take any $v \neq 0$ and write $w = v - \rho\, y (s^\top v)$, so that
$\left(I - \rho\, y s^\top\right) v = w$ and

$$v^\top H_{+} v = w^\top H w + \rho\, (s^\top v)^{2}.$$

Both terms are non-negative, the first because $H$ is positive definite and the
second because $\rho > 0$. They vanish together only if $w = 0$ and
$s^\top v = 0$; but $s^\top v = 0$ makes $w = v \neq 0$, a contradiction. Hence
$v^\top H_{+} v > 0$.
:::

The proof shows exactly where $y^\top s > 0$ enters, and it is not a technical
convenience. The quantity

$$y^\top s = \bigl(\nabla f(x_{+}) - \nabla f(x)\bigr)^\top (x_{+} - x)$$

is positive precisely when the function is curving upward along the step — when
the observed curvature is consistent with a minimum. On a non-convex objective
it can be negative, and then no positive definite matrix satisfies the secant
equation for that pair: the pair is asserting negative curvature, and asking a
positive definite matrix to reproduce it is asking for a contradiction.

Two responses are possible, and the package takes the conservative one: when
$y^\top s \le \varepsilon \lVert s \rVert \lVert y \rVert$ the pair is
**skipped**, the approximation left as it was, and the skip recorded in the
trace. The alternative, damping the pair towards one that does satisfy the
condition, injects curvature that was not observed. Skipping loses information;
damping invents it, and this toolkit prefers to lose it.

Note also that a Wolfe search makes the question moot. Condition (W) says
$\nabla f(x_{+})^\top d \ge c_2 \nabla f(x)^\top d$, and multiplying by $s > 0$
and rearranging gives

$$y^\top s \ \ge\ (c_2 - 1)\, s\, \nabla f(x)^\top d \ > \ 0,$$

since $c_2 < 1$ and the directional derivative is negative. **A quasi-Newton
method paired with a Wolfe search never skips an update.** That is the practical
argument for the more expensive search, and it is why `bfgs()` defaults to it
while `gradient_descent()`, which has no approximation to protect, defaults to
backtracking.

The argument should be stated at its true strength, which is that of a
guarantee rather than of an observed advantage. On every smooth problem in the
standard collection, and on a deliberately non-convex one besides, backtracking
skips no updates either: the curvature condition simply is not violated there.
What the Wolfe search removes is not a frequent nuisance but the possibility of
one, and the difference shows on objectives less well behaved than any that
appear in this book.

### Limited memory {#sec-lbfgs}

BFGS stores a $p \times p$ matrix, which is fine for the handful of parameters a
distribution has and impossible for the thousands a model with covariates may
have. L-BFGS keeps instead the last $m$ secant pairs, typically $m$ between 5
and 20, and never forms the matrix.

The observation that makes this possible is that the BFGS update in
@thm-bfgs is a sum of terms each of which is an outer product. Applying $H_k$ to
a vector therefore requires only inner products with the stored pairs, and
unrolling the recursion gives the **two-loop recursion**: initialise $q = g$,
run backwards through the pairs computing

$$\alpha_i = \rho_i\, s_i^\top q, \qquad q \leftarrow q - \alpha_i y_i,$$

scale once by $\gamma = (s_{k}^\top y_{k}) / (y_{k}^\top y_{k})$ using the most
recent pair, then run forwards computing

$$\beta_i = \rho_i\, y_i^\top q, \qquad q \leftarrow q + (\alpha_i - \beta_i) s_i,$$

after which $q = H_k g$. The cost is $O(mp)$ in time and $O(mp)$ in storage,
against $O(p^{2})$ for both in BFGS.

The scaling $\gamma$ deserves a word, because it is the one part of the
recursion that is a choice rather than an identity. It sets the initial
approximation at each iteration to $\gamma I$, and $\gamma$ is the Rayleigh
quotient of the most recent pair — an estimate of the inverse curvature along the
direction most recently travelled. Without it the recursion starts from $I$,
whose scale has nothing to do with the problem, and the first step of each
iteration is badly sized.

The price of limited memory is that older curvature is discarded. On a problem
whose Hessian is nearly constant that costs almost nothing; on one where the
curvature varies sharply, forgetting is sometimes an advantage, since stale
pairs describe a region the iterate has left.

```{r}
#| label: memory-cost
#| echo: false
set.seed(4)
p <- 60
A <- crossprod(matrix(rnorm(p * p), p)) + diag(p)
b <- rnorm(p)
fq  <- function(x) as.numeric(0.5 * t(x) %*% A %*% x - sum(b * x))
gq  <- function(x) as.numeric(A %*% x - b)
x0 <- rep(0, p)

rows <- list(
  BFGS = optimizers7::minimize(optimizers7::bfgs(), fq, x0, gr = gq),
  `L-BFGS (m = 5)` = optimizers7::minimize(optimizers7::lbfgs(memory = 5), fq, x0, gr = gq),
  `L-BFGS (m = 20)` = optimizers7::minimize(optimizers7::lbfgs(memory = 20), fq, x0, gr = gq)
)
knitr::kable(
  data.frame(method = names(rows),
             iterations = vapply(rows, function(r) r@iterations, numeric(1)),
             `stored numbers` = c(p * p, 2 * 5 * p, 2 * 20 * p),
             `gap from the optimum` =
               formatC(vapply(rows, function(r) r@value, numeric(1)) -
                         fq(solve(A, b)), format = "e", digits = 1),
             check.names = FALSE),
  row.names = FALSE, align = "lrrr",
  caption = "A quadratic in 60 variables. L-BFGS reaches the same answer while storing a fraction as much."
)
```
