## Second-order and quasi-Newton methods {#sec-curvature}

```{r}
#| label: setup-04b
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Every method in this section chooses its direction the same way: build a
quadratic model of $f$ around the current point, and step to the minimiser of
that model. Writing $g = \nabla f(x)$ and letting $M$ stand for whatever matrix
plays the part of the curvature, the model is

$$m(d) = f(x) + g^\top d + \tfrac{1}{2} d^\top M d,$$

and setting $\nabla m(d) = g + M d = 0$ gives

$$d = -M^{-1} g.$$

The methods differ in what they use for $M$, and this section takes them in
order of how much they store: the exact Hessian, then an approximation to its
inverse of the same size, then the last few pairs of vectors that approximation
was built from, then a single vector, then a single number. What is remarkable is
how far down that list one can go before the method stops working.

### Newton's method {#sec-newton}

The obvious choice is the true Hessian, $M = \nabla^{2} f(x)$, which makes
$m$ the second-order Taylor expansion and the step exact for a quadratic. Near a
strict local minimum this converges quadratically, which no other method here
achieves, and it is why `newton()` exists.

Away from such a point the Hessian may be indefinite or singular, and then the
step is not merely inaccurate but wrong in kind. If $H$ has a negative
eigenvalue $\lambda < 0$ with eigenvector $v$, the model $m$ is unbounded below
along $v$ and its stationary point is a saddle, not a minimum; and worse, taking
$d = -H^{-1} g$ can produce a direction with $g^\top d > 0$, which by
@def-descent points uphill. The line search then finds no acceptable step and
the run stops for a reason that has nothing to do with the problem.

So $H$ must be replaced by a positive definite matrix near it. Two repairs are
standard, both are available here, and both are described in @nocedal2006 and
@dennis1983.

The **eigenvalue repair** takes the symmetric decomposition $H = V \Lambda
V^\top$ and replaces $\Lambda$ by $\tilde\Lambda$ with

$$\tilde\lambda_i = \max(\lambda_i, \varepsilon),$$

for a floor $\varepsilon > 0$, giving $\tilde H = V \tilde\Lambda V^\top$. This
is minimal in a precise sense — it changes the eigenvalues that are a problem
and leaves the others exactly alone, so the curvature the Hessian got right is
preserved — and it is the default. Note that the floor is not cosmetic: taking
$\varepsilon = 0$ would leave a singular matrix, and taking it very small leaves
a matrix whose condition number destroys the angle bound at the end of
@sec-descent-zoutendijk. The floor is what keeps $\cos\theta_k$ away from zero.

The **ridge repair** instead takes $\tilde H = H + \tau I$ with $\tau$ the
smallest multiple of the identity making the result positive definite, found by
attempting a Cholesky factorisation and doubling $\tau$ on failure. It is
cheaper, since it needs no eigendecomposition, but it is blunter: adding $\tau I$
shifts *every* eigenvalue, so directions in which the Hessian was informative are
distorted along with the ones in which it was not.

Both repairs share the property that matters. Whatever $H$ was, the matrix
actually inverted is positive definite, so the direction is a descent direction,
so the line search has something to find.

::: {.callout-note}
## Numerical Hessians

`newton()` accepts an objective with no analytic Hessian and differences the
gradient once to obtain one. When the gradient is itself analytic that is a
single numerical differentiation and the result is good. When the gradient is
*also* differenced, the Hessian is a difference of differences, and the error
compounds — which is the one place in the toolkit where two layers of numerical
differentiation are composed. It is unavoidable, since a Hessian from values
alone must be, but it is why an objective supplying neither derivative is better
served by `bfgs()`, which never forms a Hessian at all.
:::

### The BFGS update {#sec-bfgs}

The Hessian is expensive: $p(p+1)/2$ second derivatives per iteration, and a
factorisation costing $O(p^{3})$. The quasi-Newton methods refuse to pay it and
build an approximation from what they already have — the gradients computed at
successive iterates.

The information available is a **secant pair**. Having moved from $x$ to
$x_{+} = x + s d$, write

$$s = x_{+} - x, \qquad y = \nabla f(x_{+}) - \nabla f(x).$$

For a quadratic with Hessian $H$ these satisfy $y = H s$ exactly, and for a
general $f$ the mean value theorem gives $y = \bar{H} s$ for some average of
Hessians along the segment. Either way the pair says what the curvature does in
one direction, and it is natural to ask an approximation to reproduce it.

::: {#def-secant}
## The secant equation

An approximation $B_{+}$ to the Hessian satisfies the secant equation when
$B_{+} s = y$; equivalently, an approximation $H_{+}$ to its inverse satisfies

$$H_{+}\, y = s.$$
:::

This is $p$ equations for the $p(p+1)/2$ unknowns of a symmetric matrix, so it
does not determine $H_{+}$. Two further requirements do, very nearly. The first
is that $H_{+}$ be **symmetric and positive definite**, which @sec-descent
showed is what makes the direction usable. The second is that it be **as close
as possible to $H$**, on the grounds that the pair carries information about one
direction and should not be allowed to overwrite what is known about the others.

Making "close" precise in a particular weighted Frobenius norm yields a unique
answer, and it is the BFGS update:

::: {#thm-bfgs}
## The BFGS update

With $\rho = 1 / (y^\top s)$, the update

$$H_{+} = \left(I - \rho\, s y^\top\right) H \left(I - \rho\, y s^\top\right)
          + \rho\, s s^\top$$

satisfies the secant equation, is symmetric, and is positive definite whenever
$H$ is and $y^\top s > 0$.
:::

The three claims are all short calculations, set out in @nocedal2006 along with
the variational characterisation that singles this update out; the update itself
was arrived at independently by four authors in 1970, of whom @broyden1970 is the
one cited here. Applying $H_{+}$ to $y$ and using $\rho\,y^\top s = 1$ collapses
the middle factor to zero and leaves $s$; symmetry is visible in the form, both
terms being symmetric; and for positive definiteness one writes
$w = v - \rho\,y(s^\top v)$ and finds

$$v^\top H_{+} v = w^\top H w + \rho\,(s^\top v)^{2},$$

whose two terms are non-negative and cannot vanish together, since $s^\top v = 0$
would force $w = v \neq 0$.

That last calculation is where the condition $y^\top s > 0$ enters, and it is
not a technical convenience. The quantity

$$y^\top s = \bigl(\nabla f(x_{+}) - \nabla f(x)\bigr)^\top (x_{+} - x)$$

is positive precisely when the function is curving upward along the step — when
the observed curvature is consistent with a minimum. On a non-convex objective
it can be negative, and then no positive definite matrix satisfies the secant
equation for that pair: the pair is asserting negative curvature, and asking a
positive definite matrix to reproduce it is asking for a contradiction.

Two responses are possible, and the package takes the conservative one: when
$y^\top s \le \varepsilon \lVert s \rVert \lVert y \rVert$ the pair is
**skipped**, the approximation left as it was, and the skip recorded in the
trace. The alternative, damping the pair towards one that does satisfy the
condition, injects curvature that was not observed. Skipping loses information;
damping invents it, and this toolkit prefers to lose it.

Note also that a Wolfe search makes the question moot. Condition (W) says
$\nabla f(x_{+})^\top d \ge c_2 \nabla f(x)^\top d$, and multiplying by $s > 0$
and rearranging gives

$$y^\top s \ \ge\ (c_2 - 1)\, s\, \nabla f(x)^\top d \ > \ 0,$$

since $c_2 < 1$ and the directional derivative is negative. **A quasi-Newton
method paired with a Wolfe search never skips an update.** That is the practical
argument for the more expensive search, and it is why `bfgs()` defaults to it
while `gd()`, which has no approximation to protect, defaults to
backtracking.

The argument should be stated at its true strength, which is that of a
guarantee rather than of an observed advantage. On every smooth problem in the
standard collection, and on a deliberately non-convex one besides, backtracking
skips no updates either: the curvature condition simply is not violated there.
What the Wolfe search removes is not a frequent nuisance but the possibility of
one, and the difference shows on objectives less well behaved than any that
appear in this book.

### Limited-memory BFGS {#sec-lbfgs}

BFGS stores a $p \times p$ matrix, which is fine for the handful of parameters a
distribution has and impossible for the thousands a model with covariates may
have. L-BFGS keeps instead the last $m$ secant pairs, typically $m$ between 5
and 20, and never forms the matrix; it is @nocedal1980, with the convergence
analysis and the large-scale evidence in @liu1989.

The observation that makes this possible is that the BFGS update in
@thm-bfgs is a sum of terms each of which is an outer product. Applying $H_k$ to
a vector therefore requires only inner products with the stored pairs, and
unrolling the recursion gives the **two-loop recursion**: initialise $q = g$,
run backwards through the pairs computing

$$\alpha_i = \rho_i\, s_i^\top q, \qquad q \leftarrow q - \alpha_i y_i,$$

scale once by $\gamma = (s_{k}^\top y_{k}) / (y_{k}^\top y_{k})$ using the most
recent pair, then run forwards computing

$$\beta_i = \rho_i\, y_i^\top q, \qquad q \leftarrow q + (\alpha_i - \beta_i) s_i,$$

after which $q = H_k g$. The cost is $O(mp)$ in time and $O(mp)$ in storage,
against $O(p^{2})$ for both in BFGS.

The scaling $\gamma$ deserves a word, because it is the one part of the
recursion that is a choice rather than an identity. It sets the initial
approximation at each iteration to $\gamma I$, and $\gamma$ is the Rayleigh
quotient of the most recent pair — an estimate of the inverse curvature along the
direction most recently travelled. Without it the recursion starts from $I$,
whose scale has nothing to do with the problem, and the first step of each
iteration is badly sized.

The price of limited memory is that older curvature is discarded. On a problem
whose Hessian is nearly constant that costs almost nothing; on one where the
curvature varies sharply, forgetting is sometimes an advantage, since stale
pairs describe a region the iterate has left.

```{r}
#| label: memory-cost
#| echo: false
set.seed(4)
p <- 60
A <- crossprod(matrix(rnorm(p * p), p)) + diag(p)
b <- rnorm(p)
fq  <- function(x) as.numeric(0.5 * t(x) %*% A %*% x - sum(b * x))
gq  <- function(x) as.numeric(A %*% x - b)
x0 <- rep(0, p)

rows <- list(
  BFGS = optimizers7::minimize(optimizers7::bfgs(), fq, x0, gr = gq),
  `L-BFGS (m = 5)` = optimizers7::minimize(optimizers7::lbfgs(memory = 5), fq, x0, gr = gq),
  `L-BFGS (m = 20)` = optimizers7::minimize(optimizers7::lbfgs(memory = 20), fq, x0, gr = gq)
)
knitr::kable(
  data.frame(method = names(rows),
             iterations = vapply(rows, function(r) r@iterations, numeric(1)),
             `stored numbers` = c(p * p, 2 * 5 * p, 2 * 20 * p),
             `gap from the optimum` =
               formatC(vapply(rows, function(r) r@value, numeric(1)) -
                         fq(solve(A, b)), format = "e", digits = 1),
             check.names = FALSE),
  row.names = FALSE, align = "lrrr",
  caption = "A quadratic in 60 variables. L-BFGS reaches the same answer while storing a fraction as much."
)
```

### Conjugate gradients and Barzilai-Borwein {#sec-cg-bb}

Two methods go further down still, and they are worth the page because between
them they cover the case a statistical model most often presents, in which the
parameter vector is too long for a $p \times p$ anything.

**Conjugate gradients** keeps one vector. Its direction is

$$d_k = -g_k + \beta_k d_{k-1},$$

which does not look like $-M^{-1}g$ at all, and it does not have to. Recall from
@sec-descent-directions why steepest descent is poor: under an exact line search
consecutive directions are orthogonal, since the search stops where
$g_{k+1}^\top d_k = 0$, so on a narrow valley the iterate crosses the floor
rather than travelling along it. The remedy is to bend each direction back
towards the last, and the whole of the method is the choice of how far.

On a quadratic with Hessian $H$ and an exact line search, the classical choices
of $\beta$ make the directions *conjugate*, meaning $d_i^\top H d_j = 0$ for
$i \neq j$, and a set of $p$ conjugate directions spans the space with the
minimisations along them independent of one another. The method therefore
terminates in $p$ steps exactly, having formed $H$ nowhere. The linear case is
@hestenes1952; the nonlinear one begins with @fletcher1964, and @hager2006
surveys what has happened since.

The four formulas offered here agree on a quadratic and differ everywhere else:

$$\beta^{FR} = \frac{g_k^\top g_k}{g_{k-1}^\top g_{k-1}}, \quad
  \beta^{PR} = \frac{g_k^\top y_k}{g_{k-1}^\top g_{k-1}}, \quad
  \beta^{HS} = \frac{g_k^\top y_k}{d_{k-1}^\top y_k}, \quad
  \beta^{DY} = \frac{g_k^\top g_k}{d_{k-1}^\top y_k},$$

with $y_k = g_k - g_{k-1}$ as always, and with the four due respectively to
@fletcher1964, @polak1969, @hestenes1952 and @dai1999. Fletcher–Reeves has the
cleanest theory and the practical fault of stalling for many iterations after a
poor step; Polak–Ribière recovers immediately, because a small $y$ sends $\beta$
towards zero and the method back to steepest descent, and its known failure to
converge is repaired by clamping $\beta$ at zero, a clamp which is not a fudge
since $\beta = 0$ *is* a restart.

**Barzilai–Borwein** keeps one number. Take $d = -\alpha g$ with

$$\alpha^{BB1} = \frac{s^\top s}{s^\top y}, \qquad
  \alpha^{BB2} = \frac{s^\top y}{y^\top y},$$

the two Rayleigh quotients of the secant pair. Each is a scalar estimate of the
inverse curvature along the direction just travelled, so this is @sec-bfgs with
everything discarded but one number — and on a quadratic, where the curvature is
constant, that number is exactly right and the second step lands on the answer.
The method is @barzilai1988, and @raydan1997 is what established that it is
practical on large problems rather than merely elegant.

```{r}
#| label: memory-ladder
#| echo: false
f  <- function(p) 100 * (p[2] - p[1]^2)^2 + (1 - p[1])^2
gr <- function(p) c(-400 * p[1] * (p[2] - p[1]^2) - 2 * (1 - p[1]),
                    200 * (p[2] - p[1]^2))
os <- list(
  `newton`  = optimizers7::newton(),
  `bfgs`    = optimizers7::bfgs(),
  `lbfgs`   = optimizers7::lbfgs(),
  `cg`      = optimizers7::cg(maxit = 5000),
  `bb`      = optimizers7::bb(maxit = 5000),
  `gd`      = optimizers7::gd(maxit = 20000)
)
store <- c("p(p+1)/2", "p^2", "2mp", "2p", "1", "0")
rows <- lapply(os, function(o)
  optimizers7::minimize(o, f, c(-1.2, 1), gr = gr))
knitr::kable(
  data.frame(method = names(os),
             `curvature stored` = store,
             iterations = vapply(rows, function(r) r@iterations, numeric(1)),
             converged = vapply(rows, function(r) r@converged, logical(1)),
             check.names = FALSE),
  row.names = FALSE, align = "lrrl",
  caption = "Rosenbrock from the usual start, by how much curvature the method keeps."
)
```

Read it downwards. Keeping the exact Hessian, or a full approximation to its
inverse, or twenty pairs of vectors, or one vector, or one number makes only a
modest difference to the iteration count on this problem — and keeping nothing
at all makes all the difference in the world. It is not the *amount* of
curvature information that matters so much as whether there is any: gradient
descent is the only row that fails, and conjugate gradients repairs it with a
single remembered direction.

Barzilai–Borwein deserves a closing remark, because the number in its row
depends on something other than the number it stores. Its efficiency comes from
steps that make the objective *worse* now in order to be aligned with the
curvature later, and the sufficient-decrease condition of @def-armijo forbids
exactly those. The classical remedy, due to @grippo1986, is a nonmonotone line
search, which asks for improvement against the worst of the last several values
rather than against the present one; it is what `bb()` uses by default, and
under it the method reaches Rosenbrock in 68 iterations and 77 objective
evaluations against 82 and 186 under backtracking. One stored number, and a line
search that lets it be used.
