## Non-smooth objectives {#sec-nonsmooth}

```{r}
#| label: setup-04c
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Everything so far assumed that $f$ is differentiable and that the gradient
becomes small near a solution. The second part of this assumption is the one
that fails first. For an objective with a kink at its minimizer the gradient
does not become small: it does not exist at the minimizer, and the one-sided
derivatives on either side do not approach zero. On such an objective the
methods of @sec-curvature reach the correct point and then report failure,
because the quantity their stopping rule tests never becomes small.

The example is a standard one. Minimizing

$$f(\mu) = \sum_{i=1}^{n} \lvert y_i - \mu \rvert$$

is the estimation of a median, and it is the location part of a Laplace
likelihood; the toolkit met exactly this objective in @sec-fitting. The
function is piecewise linear with a kink at every observation, and its
minimizer is one of the observations.

### The subdifferential {#sec-subdifferential}

For a convex $f$ the derivative is replaced by a set.

::: {#def-subdifferential}
## Subgradient and subdifferential

For convex $f$, a vector $v$ is a **subgradient** at $x$ when

$$f(z) \ \ge\ f(x) + v^\top (z - x) \qquad \text{for all } z,$$

and the set of all such $v$ is the **subdifferential** $\partial f(x)$.
:::

Geometrically, a subgradient is the slope of an affine function that touches
$f$ at $x$ and lies below it everywhere, so the subdifferential collects every
supporting hyperplane at that point. Where $f$ is differentiable the
subdifferential contains exactly one element, $\partial f(x) = \{\nabla
f(x)\}$; at a kink it contains many.

The optimality condition generalizes directly, and follows in one line from
the definition: setting $v = 0$ in @def-subdifferential turns the subgradient
inequality into $f(z) \ge f(x)$ for every $z$, which is minimality, and the
argument reads the same in both directions.

::: {#prp-fermat}
## Fermat's rule

A convex $f$ attains its minimum at $x^{\star}$ if and only if
$0 \in \partial f(x^{\star})$.
:::

The subdifferential, Fermat's rule and the other results this section uses are
developed in @rockafellar1970 and, in the form closest to the algorithms here,
in @hiriarturruty1993.

Applied to the absolute-deviation objective, these definitions give the
following picture. Away from the data points,
$f'(\mu) = \#\{y_i < \mu\} - \#\{y_i > \mu\}$; at an observation
$\mu = y_{(k)}$ the subdifferential is the interval between the two one-sided
slopes. Fermat's rule is satisfied at the median, where the counts on either
side balance. The individual subgradients, however, do not become small there. Each term
$\lvert y_i - \mu \rvert$ contributes $\pm 1$ according to the side on which
$y_i$ falls, so any particular subgradient — the value a `gr` function
returns — is an integer. Away from the median its size is comparable to $n$;
at the median itself it is still at least one.

This is the central difficulty: the quantity a gradient-based stopping rule
tests does not vanish at the solution, so no tolerance both fires there and
stays quiet elsewhere. The method reaches the minimizer, the line search finds
no acceptable step because every direction leads uphill from the kink, and the
run stops with a failure report while standing on the answer.

```{r}
#| label: kink-failure
#| echo: false
set.seed(11)
y <- rnorm(101)
fk <- function(p) sum(abs(y - p))
gk <- function(p) -sum(sign(y - p))

b <- optimizers7::minimize(optimizers7::bfgs(), fk, par = 0, gr = gk)
u <- optimizers7::minimize(optimizers7::bundle(), fk, par = 0, gr = gk)

knitr::kable(
  data.frame(
    method = c("bfgs", "bundle"),
    `point reached` = formatC(c(b@par, u@par), format = "f", digits = 8),
    `distance to the median` =
      formatC(abs(c(b@par, u@par) - median(y)), format = "e", digits = 1),
    `a subgradient there` = formatC(gk(c(b@par, u@par)), format = "f", digits = 1),
    converged = c(b@converged, u@converged),
    check.names = FALSE),
  row.names = FALSE, align = "lrrrl",
  caption = "Both methods reach the median; only the bundle method reports convergence."
)
```

Both methods reach the same point to the last printed digit, and at that point
a subgradient has size one for both; only one of the two reports success. The
difference is not in the arithmetic but in the quantity being tested, which is
the subject of the next two sections.

### The cutting-plane model {#sec-cutting}

Since no single subgradient vanishes at the solution, the bundle method stops
testing single subgradients. It keeps a collection of them, gathered at the
points already visited, and works with the model that the collection defines.

Given points $z_1, \dots, z_k$ with values $f_j = f(z_j)$ and subgradients
$v_j \in \partial f(z_j)$, @def-subdifferential states that each affine
function $\ell_j(x) = f_j + v_j^\top (x - z_j)$ lies below $f$ everywhere.
Their pointwise maximum therefore also lies below $f$, and it becomes a better
model as more planes are collected:

$$\check{f}(x) \ =\ \max_{1 \le j \le k}\ \bigl\{ f_j + v_j^\top (x - z_j) \bigr\}
\ \le\ f(x).$$

This is the **cutting-plane model**, and its useful property is that it is
piecewise linear: a kink, which is two linearizations meeting, is exactly the
shape it represents naturally, while a quadratic model must either smooth the
kink away or place spurious curvature across it.

Minimizing $\check{f}$ alone would not be safe. A model built from finitely
many planes can be unbounded below, and even when it is bounded its minimizer
may lie far outside the region the collected planes describe. A proximal term
is therefore added, anchoring the step to the current point $\hat{x}$, called
the **stability center**:

$$d^{\star} \ =\ \operatorname*{arg\,min}_{d}\ \left\{
  \max_{j}\ \bigl( -\alpha_j + v_j^\top d \bigr)
  + \frac{1}{2t} \lVert d \rVert^{2} \right\},
\tag{P}$$

where $t > 0$ controls how far the model is trusted and

$$\alpha_j \ =\ f(\hat{x}) - \ell_j(\hat{x}) \ \ge\ 0$$

is the **linearization error** of plane $j$, the amount by which the plane
underestimates $f$ at the current point. It is non-negative by
@def-subdifferential, and it gives an old plane, collected far away, less
influence than a recent one.

### The dual of the bundle subproblem {#sec-bundle-dual}

Problem (P) is a quadratic program in $d$ with a piecewise-linear term. It
is solved through its dual, which is a small problem with a simple feasible
set.

::: {#thm-bundle-dual}
## The bundle subproblem and its dual

Let $V$ be the matrix with columns $v_1, \dots, v_k$ and $\alpha$ the vector of
linearization errors. Then the solution of (P) is $d^{\star} = -t\, V\lambda^{\star}$,
where $\lambda^{\star}$ solves

$$\min_{\lambda \in \Delta_k}\ \ \frac{t}{2} \lVert V \lambda \rVert^{2}
   + \alpha^\top \lambda,
\qquad \Delta_k = \Bigl\{ \lambda \ge 0 : \textstyle\sum_j \lambda_j = 1 \Bigr\},$$

and the optimal value of (P) is

$$v \ =\ -\Bigl( t \lVert V\lambda^{\star} \rVert^{2}
                 + \alpha^\top \lambda^{\star} \Bigr) \ \le\ 0 .$$
:::

The dual is derived in the usual way. Write (P) with an epigraph variable $u$, so that the objective becomes
$u + \lVert d \rVert^{2}/(2t)$ subject to $-\alpha_j + v_j^\top d \le u$.
Stationarity of the Lagrangian in $u$ forces $\sum_j \lambda_j = 1$, which is
the origin of the simplex constraint, and stationarity in $d$ gives
$d = -t V\lambda$. Substituting both back yields the
stated minimization over $\Delta_k$, and strong duality holds because the
problem is convex with a strictly feasible point. The construction is due to
@lemarechal1975 and @kiwiel1990; @bonnans2006 presents it in the form used
here.

Write $\bar{v} = V\lambda^{\star}$ for the **aggregate subgradient** and
$\bar\alpha = \alpha^\top \lambda^{\star}$ for the aggregate error. The
aggregate is a convex combination of collected subgradients, and by convexity
of the subdifferential it is itself a subgradient of the model at $\hat{x}$.
Both $\lVert \bar{v} \rVert$ and $\bar\alpha$ are non-negative, and

$$\lVert \bar{v} \rVert^{2} + \bar\alpha \ =\ 0
\iff \bar{v} = 0 \ \text{ and } \ \bar\alpha = 0,$$

which states that zero lies in the convex hull of the collected subgradients
with no linearization error. This is the computable counterpart of
$0 \in \partial f(\hat{x})$ in @prp-fermat, it is the correct quantity to
test, and it is the quantity that `crit_stationary()` reads.

::: {.callout-important}
## The stopping quantity

The value $-v$ from @thm-bundle-dual also vanishes at a solution, and reads
naturally as the improvement the model still expects, so it looks like a
reasonable stopping quantity. It is not, because it carries a factor of $t$,
and $t$ is reduced whenever the model over-promises. A run in difficulty can
therefore drive $-v$ below any fixed tolerance by shrinking its trust
parameter rather than by approaching a minimum, and would then report
convergence at a point its own model still describes as steep. Dropping the
factor of $t$ leaves $\lVert \bar{v} \rVert^{2} + \bar\alpha$, which shrinking
$t$ cannot reduce.
:::

### Serious steps, null steps and aggregation {#sec-bundle-steps}

Given $d^{\star}$, the trial point is $z = \hat{x} + d^{\star}$, and the
decision whether to move there follows the same principle as an Armijo
condition: the step must deliver a fixed fraction of the decrease the model
promised,

$$f(z) \ \le\ f(\hat{x}) + m\, v, \qquad m \in (0, 1).$$

If the condition holds, the center moves; this is a **serious step**. If it
does not, the center stays, but the iteration is still useful, because the
trial point contributes its own subgradient to the collection and the model is
strictly better at the next iteration. This is a **null step**, and a run with
many null steps is refining its representation of a kink, not failing.

Without intervention the collection grows without bound. It is capped by
**aggregation** rather than by discarding: when the bundle is full, the oldest
planes are replaced by the single affine function that the dual solution
defines,

$$\ell_{\text{agg}}(x) = f(\hat{x}) - \bar\alpha + \bar{v}^\top (x - \hat{x}),$$

which lies below $f$ because it is a convex combination of functions that do,
and which summarizes the contribution of the discarded planes. Deleting old
planes outright would lose that information and can stall the method; keeping
the aggregate is what makes a bounded bundle safe.

### The proximal gradient {#sec-prox-grad}

The bundle method needs nothing of the objective but a value and one
subgradient. Where more is known the work can be divided instead. Many
objectives in this toolkit split as

$$F(x) = f(x) + \rho(x),$$

with $f$ smooth and $\rho$ the part carrying the kink -- a penalized
likelihood is exactly that. A proximal gradient method takes a gradient
step on $f$ and then applies the **proximal operator** of $\rho$,

$$\operatorname{prox}_{t\rho}(v) \;=\;
  \arg\min_u \left\{ \rho(u) + \frac{1}{2t}\lVert u - v\rVert^2 \right\},$$

which for the penalties of @sec-penalty-branches is available in closed
form: the soft threshold for the lasso, a threshold followed by a shrinkage
for the elastic net, and the piecewise operators of SCAD and MCP over their
convex regions. The kink is never differentiated; it is solved exactly, once
per step.

`prox_grad()` runs that iteration, accelerated and with an adaptive restart.
Two measured facts decide how it is used. Acceleration is for
ill-conditioned problems: at a condition number of 3 the plain iteration
wins narrowly, at 55 it is 4153 iterations against 126, and at 480 the plain
one does not converge in fifty thousand while the accelerated one takes 334.
And the stationarity measure must be read at the **iterate** and not at the
extrapolated point, which under acceleration is a different place: read at
the wrong one the mapping never vanishes and a run circles the answer
reporting failure, while read at the iterate it agrees with the
Karush-Kuhn-Tucker violation to the digit.

One thing this method does not take is a box. Its constraint belongs inside
the operator, where it composes with the term already there, so bounds
beside the objective would be a second and conflicting route to the same
thing; `optimizer_bounded()` reports that, and the validator of
@sec-optim-check-practice consults it instead of testing bounds on every
method alike.

### Direct search methods {#sec-direct-search}

A second way of handling a non-smooth objective is to use no derivatives at
all. This costs many more evaluations and yields a weaker guarantee, but it
requires nothing of the objective beyond a value, which is sometimes all that
is available.

The concept that carries the theory is the **positive spanning set**; the
theory is collected in @conn2009 and surveyed in @kolda2003.

::: {#def-spanning}
## Positive spanning set

A finite set $D \subset \mathbb{R}^{p}$ positively spans $\mathbb{R}^{p}$ when
every $w \in \mathbb{R}^{p}$ can be written $w = \sum_{d \in D} c_d\, d$ with all
$c_d \ge 0$.
:::

The relevance is the following. If $D$ positively spans the space and $f$ is differentiable at $x$ with
$\nabla f(x) \neq 0$, then some $d \in D$ satisfies
$\nabla f(x)^\top d < 0$. Otherwise $-\nabla f(x)$, a non-negative
combination of the elements of $D$, would give
$-\lVert \nabla f(x) \rVert^{2} \ge 0$, a contradiction. A poll that fails in
every direction of $D$ is therefore evidence about the point rather than about
the directions, and this justifies shrinking the poll radius: the limit points
of a run whose radius goes to zero are stationary. `compass()` polls the $2p$
signed coordinate axes, which positively span the space.

For an $f$ that is only Lipschitz the argument fails, and the way it fails is
informative. A kink whose ridge runs diagonally is descended by no coordinate
direction, so the poll fails at a point that is not stationary, the radius
shrinks, and the run stops there. The repair, which is the idea behind the mesh adaptive direct search of
@audet2006, is to let the set of directions used over the whole run become
dense in the sphere, so that no descent direction can be missed forever.
Drawing a fresh random orthonormal basis at each poll achieves this with
probability one.

```{r}
#| label: diagonal-kink
#| echo: false
kf <- function(p) abs(p[1] + p[2]) + 0.1 * sum(p^2)
kg <- function(p) rep(sign(p[1] + p[2]), 2) + 0.2 * p
set.seed(2)
rs <- list(
  `compass, coordinate` = optimizers7::minimize(
    optimizers7::compass(directions = "coordinate"), kf, c(1, 0.5)),
  `compass, mads` = optimizers7::minimize(
    optimizers7::compass(directions = "mads"), kf, c(1, 0.5)),
  `nelder-mead` = optimizers7::minimize(optimizers7::nelder_mead(), kf, c(1, 0.5)),
  `bundle` = optimizers7::minimize(optimizers7::bundle(), kf, c(1, 0.5), gr = kg)
)
knitr::kable(
  data.frame(method = names(rs),
             `value reached` =
               formatC(vapply(rs, function(r) r@value, numeric(1)),
                       format = "e", digits = 2),
             `evaluations` =
               as.integer(vapply(rs, function(r) r@counts[["f"]], numeric(1))),
             check.names = FALSE),
  row.names = FALSE, align = "lrr",
  # \lvert and \lVert rather than bare pipes: a | inside math in a markdown
  # table splits the cell, and kable escapes it to an entity that MathJax then
  # prints verbatim.
  caption = paste("The minimum of $\\lvert p_1 + p_2 \\rvert +",
                  "0.1 \\lVert p \\rVert^{2}$ is zero, at the origin.",
                  "The ridge runs diagonally, so no coordinate direction",
                  "descends it.")
)
```

The table contains three separate findings.

The first two rows illustrate the argument just given. A fixed set of directions stops on the ridge, at a point that is not
stationary, having exhausted its poll without finding the descent direction
that exists. Random directions do an order of magnitude better and still do
not arrive, because a finite number of directions from a fixed center remains
a poor instrument for a kink.

The third row shows Nelder–Mead solving the problem to machine precision,
better than either poll and better than the bundle method. The reason is that
a simplex is not a poll: it has no fixed center and no fixed set of offsets,
it deforms, and on a problem whose descent direction lies along a ridge it
deforms into a long thin simplex aligned with that ridge. This adaptivity is
exactly what a fixed pattern lacks. What Nelder–Mead lacks in exchange is a
guarantee: the positive-spanning argument does not apply to it, the
counterexample of @mckinnon1998 shows that it can converge to a point that is
not a minimizer, and the positive results of @lagarias1998 cover only one and
two dimensions. It also needs 343 evaluations here.

The last row shows the bundle method arriving, with a guarantee, in a seventh
of the evaluations. It is the only method in the table that uses the available
derivative information without requiring any single derivative to vanish.

### Simulated annealing {#sec-annealing}

Every method so far walks downhill, and on a surface with many local minima
that is the wrong thing to do from the wrong starting point. A method
allowed to walk uphill sometimes can leave one.

Simulated annealing proposes a step, accepts it outright where it improves
the objective, and otherwise accepts it with probability
$\exp(-\Delta f / T)$ [@kirkpatrick1983], so that at a high temperature $T$
almost anything is accepted and at a low one almost nothing is. The
temperature falls on a schedule, and with it the proposal's own step sizes,
which adapt so that about half the proposals are accepted in each
coordinate [@corana1987].

```{r}
#| label: op-sa
rastrigin <- function(p) sum(p^2) - 10 * sum(cos(2 * pi * p)) + 20
rast_gr <- function(p) 2 * p + 20 * pi * sin(2 * pi * p)

set.seed(5)
runs <- list(sa = minimize(sa(), rastrigin, c(3.4, -2.7)),
             bfgs = minimize(bfgs(), rastrigin, c(3.4, -2.7), gr = rast_gr))
data.frame(value = sapply(runs, function(r) round(r@value, 4)),
           par1 = sapply(runs, function(r) round(r@par[1], 4)),
           par2 = sapply(runs, function(r) round(r@par[2], 4)))
```

The minimum is at the origin and is worth zero. The descent method stops at
the nearest local minimum, which it reaches quickly and correctly; the
annealing run crosses the ridges between the wells and ends inside the
global one, close to the origin but not at it. That difference is the
honest description of what the method is for. It locates a basin rather
than a point, and a descent method started there finishes the job in a few
iterations, which is why its place in this toolkit is a starting value
rather than an answer. Neither is the better method in general, and the
second is far the more expensive: what separates them is whether the
surface has one basin or many.

The stopping rule has two halves that must be read together: the value the
walk ends each temperature level at has not moved over the last few levels,
**and** it sits at the best point found. Reading the best-so-far alone stops
the run far too early, since the best stops improving long before the walk
settles.

A global search over a model's coefficients is asked once, before the
fitting begins, and @sec-statmod-start is where that is set out.
```{r}
#| label: consistency-optimizers
#| include: false
# Internal consistency gate: fails the render if an accepted step violates the
# condition claimed for it, if the BFGS update does not satisfy the secant
# equation, if the two-loop recursion disagrees with the explicit inverse, if
# the bound transform's chain rule disagrees with a numerical derivative, if the
# aggregate subgradient fails to vanish where an ordinary one does not, or if
# any method fails its own contract.
assert_optimizers_ok()
```
