## Objectives that are not differentiable {#sec-nonsmooth}

```{r}
#| label: setup-04c
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Everything so far has assumed that $f$ is differentiable and that the gradient
becomes small near a solution. Both halves of that assumption are used, and the
second is the one that fails first. For an objective with a kink at its
minimiser the gradient does not become small — it does not exist there, and the
one-sided derivatives on either side do not approach zero. The methods of
@sec-curvature then do something worse than converging slowly: they walk to the
right answer and report that they have failed.

The example is not contrived. Minimising

$$f(\mu) = \sum_{i=1}^{n} \lvert y_i - \mu \rvert$$

is estimating a median, and it is the location part of a Laplace likelihood; the
toolkit met exactly this in @sec-fitting. The function is piecewise linear with a
kink at every observation, and its minimiser is one of them.

### The subdifferential {#sec-subdifferential}

For a convex $f$ the derivative is replaced by a set.

::: {#def-subdifferential}
## Subgradient and subdifferential

For convex $f$, a vector $v$ is a **subgradient** at $x$ when

$$f(z) \ \ge\ f(x) + v^\top (z - x) \qquad \text{for all } z,$$

and the set of all such $v$ is the **subdifferential** $\partial f(x)$.
:::

Geometrically a subgradient is the slope of an affine function that touches $f$
at $x$ and lies below it everywhere, so the subdifferential collects every
supporting hyperplane at that point. Where $f$ is differentiable there is
exactly one, $\partial f(x) = \{\nabla f(x)\}$; at a kink there are many.

The optimality condition generalises exactly as one would hope, and it is one
line from the definition: putting $v = 0$ into @def-subdifferential turns the
subgradient inequality into $f(z) \ge f(x)$ for every $z$, which is minimality
itself, read in either direction.

::: {#prp-fermat}
## Fermat's rule

A convex $f$ attains its minimum at $x^{\star}$ if and only if
$0 \in \partial f(x^{\star})$.
:::

The subdifferential, Fermat's rule and everything else this section rests on are
developed in @rockafellar1970 and, in the form closest to the algorithms here,
in @hiriarturruty1993.

Now apply this to the absolute-deviation objective. Away from the data
$f'(\mu) = \#\{y_i < \mu\} - \#\{y_i > \mu\}$, and at $\mu = y_{(k)}$ the
subdifferential is the interval between the two one-sided slopes. Fermat's rule
is satisfied at the median, where the counts on either side balance. But notice
what the *individual* subgradients look like there: each $\lvert y_i - \mu
\rvert$ contributes $\pm 1$ depending on which side $y_i$ falls, and any
particular subgradient you happen to evaluate — the one a `gr` function returns —
is an integer of size comparable to $n$ away from the median and of size at
least one at it. It never becomes small.

This is the whole difficulty in one sentence. **The quantity a gradient-based
stopping rule tests does not vanish at the solution**, so no tolerance can be
chosen that both fires there and does not fire everywhere else. The method
arrives, the line search finds no acceptable step because every direction leads
uphill from a kink, and the run stops reporting failure while standing on the
answer.

```{r}
#| label: kink-failure
#| echo: false
set.seed(11)
y <- rnorm(101)
fk <- function(p) sum(abs(y - p))
gk <- function(p) -sum(sign(y - p))

b <- optimizers7::minimize(optimizers7::bfgs(), fk, par = 0, gr = gk)
u <- optimizers7::minimize(optimizers7::bundle(), fk, par = 0, gr = gk)

knitr::kable(
  data.frame(
    method = c("bfgs", "bundle"),
    `point reached` = formatC(c(b@par, u@par), format = "f", digits = 8),
    `distance to the median` =
      formatC(abs(c(b@par, u@par) - median(y)), format = "e", digits = 1),
    `a subgradient there` = formatC(gk(c(b@par, u@par)), format = "f", digits = 1),
    converged = c(b@converged, u@converged),
    check.names = FALSE),
  row.names = FALSE, align = "lrrrl",
  caption = "Both methods find the median. Only one of them knows it."
)
```

The table is the argument. Both methods reach the same point, to the last digit
printed; a subgradient there has size one for both; and one of them reports
success. The difference is not in the arithmetic but in what is being tested.

### Cutting planes {#sec-cutting}

Since no single subgradient vanishes, the repair is to stop asking one to. The
bundle method keeps a collection of them, from points already visited, and works
with the model they define.

Given points $z_1, \dots, z_k$ with values $f_j = f(z_j)$ and subgradients
$v_j \in \partial f(z_j)$, @def-subdifferential says each affine function
$\ell_j(x) = f_j + v_j^\top (x - z_j)$ lies below $f$ everywhere. So does their
pointwise maximum, and that maximum is a better model as more are collected:

$$\check{f}(x) \ =\ \max_{1 \le j \le k}\ \bigl\{ f_j + v_j^\top (x - z_j) \bigr\}
\ \le\ f(x).$$

This is the **cutting-plane model**, and the point of it is that it is
piecewise linear. A kink is exactly what it represents naturally — two
linearisations meeting — where a quadratic model must either smooth it away or
place spurious curvature across it.

Minimising $\check{f}$ alone is unsafe, because a model built from finitely many
planes can be unbounded below, and even when it is not, its minimiser may lie far
outside the region where the collected planes say anything. So a proximal term is
added, anchoring the step to the current point $\hat{x}$, called the **stability
centre**:

$$d^{\star} \ =\ \operatorname*{arg\,min}_{d}\ \left\{
  \max_{j}\ \bigl( -\alpha_j + v_j^\top d \bigr)
  + \frac{1}{2t} \lVert d \rVert^{2} \right\},
\tag{P}$$

where $t > 0$ controls how far the model is trusted and

$$\alpha_j \ =\ f(\hat{x}) - \ell_j(\hat{x}) \ \ge\ 0$$

is the **linearisation error**, measuring how badly plane $j$ misses the current
point. It is non-negative by @def-subdifferential, and it is what makes an old
plane, collected far away, count for less than a recent one.

### The dual problem and the stopping quantity {#sec-bundle-dual}

Problem (P) is a quadratic programme in $d$ with a piecewise-linear term, and it
is solved through its dual, which is small and has a pleasant feasible set.

::: {#thm-bundle-dual}
## The bundle subproblem and its dual

Let $V$ be the matrix with columns $v_1, \dots, v_k$ and $\alpha$ the vector of
linearisation errors. Then the solution of (P) is $d^{\star} = -t\, V\lambda^{\star}$,
where $\lambda^{\star}$ solves

$$\min_{\lambda \in \Delta_k}\ \ \frac{t}{2} \lVert V \lambda \rVert^{2}
   + \alpha^\top \lambda,
\qquad \Delta_k = \Bigl\{ \lambda \ge 0 : \textstyle\sum_j \lambda_j = 1 \Bigr\},$$

and the optimal value of (P) is

$$v \ =\ -\Bigl( t \lVert V\lambda^{\star} \rVert^{2}
                 + \alpha^\top \lambda^{\star} \Bigr) \ \le\ 0 .$$
:::

The route to the dual is the usual one. Writing (P) with an epigraph variable
$u$, so that the objective becomes $u + \lVert d \rVert^{2}/(2t)$ subject to
$-\alpha_j + v_j^\top d \le u$, stationarity of the Lagrangian in $u$ forces
$\sum_j \lambda_j = 1$, which is where the simplex comes from, and stationarity
in $d$ gives $d = -t V\lambda$. Substituting both back leaves the stated
minimisation over $\Delta_k$, and strong duality holds because the problem is
convex with a strictly feasible point. The construction is @lemarechal1975 and
@kiwiel1990; @bonnans2006 gives it in the form used here.

Write $\bar{v} = V\lambda^{\star}$ for the **aggregate subgradient** and
$\bar\alpha = \alpha^\top \lambda^{\star}$ for the aggregate error. The aggregate
is a convex combination of collected subgradients, and by convexity of the
subdifferential it is itself a subgradient of the model at $\hat{x}$. Both
$\lVert \bar{v} \rVert$ and $\bar\alpha$ are non-negative, and

$$\lVert \bar{v} \rVert^{2} + \bar\alpha \ =\ 0
\iff \bar{v} = 0 \ \text{ and } \ \bar\alpha = 0,$$

which says that zero lies in the convex hull of the collected subgradients with
no linearisation error — the computable statement standing in for
$0 \in \partial f(\hat{x})$ in @prp-fermat. **That** is the quantity to test, and
it is the one `crit_stationary()` reads.

::: {.callout-important}
## Why the predicted decrease is not what the stopping rule tests

It is tempting to stop on $-v$ instead, the value from @thm-bundle-dual, since it
also vanishes at a solution and has the appealing reading of "how much better the
model believes it can do". It carries a factor of $t$, and $t$ is reduced
whenever the model over-promises. A run in difficulty therefore drives $-v$ below
any fixed tolerance **by shrinking its trust region**, not by approaching a
minimum, and reports convergence standing somewhere its own model still calls
steep. Dropping the factor leaves $\lVert \bar{v} \rVert^{2} + \bar\alpha$, which no
amount of shrinking can flatter.
:::

### Serious steps, null steps, and bounded memory {#sec-bundle-steps}

With $d^{\star}$ in hand, the trial point is $z = \hat{x} + d^{\star}$, and
whether to move there is decided by the same bargain an Armijo condition strikes:
the step must deliver a fixed fraction of the decrease the model promised,

$$f(z) \ \le\ f(\hat{x}) + m\, v, \qquad m \in (0, 1).$$

If it does, the centre moves — a **serious step**. If it does not, the centre
stays, and the iteration is nonetheless not wasted: the trial point contributes
its own subgradient to the collection, so the model is strictly better next time.
This is a **null step**, and a run reporting many of them is refining its picture
of a kink rather than failing.

Left alone the collection grows without bound. It is capped by **aggregation**
rather than by discarding: when the bundle is full, the oldest planes are
replaced by the single affine function the dual solution defines,

$$\ell_{\text{agg}}(x) = f(\hat{x}) - \bar\alpha + \bar{v}^\top (x - \hat{x}),$$

which lies below $f$ because it is a convex combination of functions that do, and
which summarises everything the discarded planes contributed. Simply deleting them
loses that information and can stall the method; keeping the aggregate is what
makes a bounded bundle safe.

### Methods that use no derivatives at all {#sec-direct-search}

There is a second repair, which is to stop using derivatives. It costs a great
deal and buys a weaker guarantee, but it asks nothing of the objective beyond a
number, which is sometimes all there is.

The idea with a theorem behind it is the **positive spanning set**, and the
theory around it is collected in @conn2009 and surveyed in @kolda2003.

::: {#def-spanning}
## Positive spanning set

A finite set $D \subset \mathbb{R}^{p}$ positively spans $\mathbb{R}^{p}$ when
every $w \in \mathbb{R}^{p}$ can be written $w = \sum_{d \in D} c_d\, d$ with all
$c_d \ge 0$.
:::

The relevance is immediate: if $D$ positively spans the space and $f$ is
differentiable at $x$ with $\nabla f(x) \neq 0$, then some $d \in D$ has
$\nabla f(x)^\top d < 0$, since otherwise $-\nabla f(x)$, being a non-negative
combination of the $d$, would satisfy $-\lVert \nabla f(x) \rVert^{2} \ge 0$.
So **a poll that fails in every direction is evidence about the point, not about
the directions**, and that is what licenses shrinking the poll radius. The limit
points of a run whose radius goes to zero are stationary. `compass()` polls the
$2p$ signed coordinate axes, which positively span the space.

For a merely Lipschitz $f$ the argument breaks, and instructively. A kink whose
ridge runs diagonally is descended by no coordinate direction: the poll fails at
a point that is not stationary, the radius shrinks, and the run stops there. The
repair, which is the idea behind the mesh adaptive direct search of @audet2006,
is to let the set of directions used over the whole run become dense in the
sphere, so that no descent direction can be missed for ever; drawing a fresh
random orthonormal basis at each poll achieves it with probability one.

```{r}
#| label: diagonal-kink
#| echo: false
kf <- function(p) abs(p[1] + p[2]) + 0.1 * sum(p^2)
kg <- function(p) rep(sign(p[1] + p[2]), 2) + 0.2 * p
set.seed(2)
rs <- list(
  `compass, coordinate` = optimizers7::minimize(
    optimizers7::compass(directions = "coordinate"), kf, c(1, 0.5)),
  `compass, mads` = optimizers7::minimize(
    optimizers7::compass(directions = "mads"), kf, c(1, 0.5)),
  `nelder-mead` = optimizers7::minimize(optimizers7::nelder_mead(), kf, c(1, 0.5)),
  `bundle` = optimizers7::minimize(optimizers7::bundle(), kf, c(1, 0.5), gr = kg)
)
knitr::kable(
  data.frame(method = names(rs),
             `value reached` =
               formatC(vapply(rs, function(r) r@value, numeric(1)),
                       format = "e", digits = 2),
             `evaluations` =
               as.integer(vapply(rs, function(r) r@counts[["f"]], numeric(1))),
             check.names = FALSE),
  row.names = FALSE, align = "lrr",
  # \lvert and \lVert rather than bare pipes: a | inside math in a markdown
  # table splits the cell, and kable escapes it to an entity that MathJax then
  # prints verbatim.
  caption = paste("The minimum of $\\lvert p_1 + p_2 \\rvert +",
                  "0.1 \\lVert p \\rVert^{2}$ is zero, at the origin.",
                  "The ridge runs diagonally, so no coordinate direction",
                  "descends it.")
)
```

The table repays reading in two columns rather than one.

The first two rows are the argument just made. A fixed set of directions stops
on the ridge, at a point that is not stationary, having exhausted its poll
without finding a descent direction that exists; drawing them at random does an
order of magnitude better and still does not arrive, a finite number of
directions from a fixed centre being the wrong instrument for a kink.

The third row is a genuine surprise, and worth not glossing. Nelder–Mead solves
this to machine precision, better than either poll and better than the bundle
method — because a simplex is not a poll. It has no fixed centre and no fixed set
of offsets: it deforms, and what it deforms into, on a problem whose descent
direction lies along a ridge, is a long thin simplex aligned with that ridge.
The adaptation is exactly what the fixed pattern lacks. What it lacks in turn is
any theorem, since the positive-spanning argument does not apply to it and the
counterexample of @mckinnon1998 shows it can converge to a point that is not a
minimiser at all — @lagarias1998 is what can be proved, and it covers one and two
dimensions only. It also needs 343 evaluations to get there.

The last row is the one to build on. The bundle method arrives, with a theorem,
in a seventh of the evaluations — because it is the only method in the table that
uses the derivative information it was given without asking any single derivative
to vanish.

```{r}
#| label: consistency-optimizers
#| include: false
# Internal consistency gate: fails the render if an accepted step violates the
# condition claimed for it, if the BFGS update does not satisfy the secant
# equation, if the two-loop recursion disagrees with the explicit inverse, if
# the bound transform's chain rule disagrees with a numerical derivative, if the
# aggregate subgradient fails to vanish where an ordinary one does not, or if
# any method fails its own contract.
assert_optimizers_ok()
```
