## Box constraints and stochastic gradients {#sec-constraints}

```{r}
#| label: setup-04d
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Two topics remain, and neither fits inside the descent framework of the
previous sections. The first is the handling of box constraints, which the
toolkit removes by a change of variable rather than enforcing. The second is a
method that gives up the requirement of descent altogether, and is useful for
stochastic objectives precisely because it does.

### The box reparameterisation {#sec-box}

Suppose each parameter is confined to an interval, $\theta_j \in (L_j, U_j)$,
with either endpoint possibly infinite. Passing such constraints to a general
optimiser creates the difficulties described in @sec-fit-scale: every proposal
must be tested for admissibility, a step that overshoots must be shortened
rather than taken, and near an endpoint the objective may be unbounded or
undefined.

The toolkit removes the constraint instead of enforcing it. The alternative,
keeping the box and working with an active set, is the approach of @byrd1995,
and it is the right choice when an optimum lies exactly on a bound; the
comparison appears below. Each bounded coordinate is reparameterised by a
diffeomorphism $h_j : \mathbb{R} \to (L_j, U_j)$,

$$\theta_j \ =\ h_j(\eta_j) \ =\
\begin{cases}
L_j + e^{\eta_j}
  & \text{lower bound only,} \\[4pt]
U_j - e^{\eta_j}
  & \text{upper bound only,} \\[4pt]
L_j + (U_j - L_j)\,\dfrac{1}{1 + e^{-\eta_j}}
  & \text{both finite,}
\end{cases}$$

and the optimiser runs unconstrained in $\eta$. These are the bounded links of
@sec-links, reused here for the purpose they were written for. Every point the
optimiser proposes is admissible by construction, so there is no rejection
step and no boundary for a line search to encounter.

The cost of the change of variable is a set of derivative corrections, and
they are cheap because the Jacobian is diagonal: each coordinate carries its
own transform and nothing else. With $F(\eta) = f(h(\eta))$,

$$\frac{\partial F}{\partial \eta_i} \ =\
  \frac{\partial f}{\partial \theta_i}\, h_i'(\eta_i),
\qquad
\frac{\partial^{2} F}{\partial \eta_i \partial \eta_j} \ =\
  \frac{\partial^{2} f}{\partial \theta_i \partial \theta_j}\, h_i'(\eta_i)\, h_j'(\eta_j)
  \ +\ \delta_{ij}\, \frac{\partial f}{\partial \theta_i}\, h_i''(\eta_i).$$

The second-derivative term appears only on the diagonal, and only because $h$
is not affine; it is the difference between transforming a Hessian and merely
rescaling one. In matrix form,

$$\nabla^{2} F \ =\ \operatorname{diag}(h')\, \nabla^{2} f\, \operatorname{diag}(h')
  \ +\ \operatorname{diag}\!\bigl(\nabla f \odot h''\bigr),$$

which is the congruence of @sec-link-scale with one extra diagonal term. Every
algorithm in this chapter receives these corrected derivatives without any
information that a box was involved.

::: {.callout-note}
## Optima on a bound

An optimum lying exactly on a bound cannot be reached, because reaching it
requires $\eta$ to go to infinity. The optimiser keeps moving, improves by
less and less, and stops on its iteration budget at a point close to the
constraint but not on it.

For the statistical use this mechanism serves — a positive variance, a
probability inside the unit interval, a shape parameter that must not vanish —
the optimum is interior and the situation does not arise. For a genuine
box-constrained problem whose solution has active constraints, an active-set
method is the appropriate tool.

A second limitation is arithmetical rather than mathematical: in exact
arithmetic $h$ never attains its endpoints, but in double precision it does.
The inverse logit returns exactly one above $\eta \approx 37$, and
$L + e^{\eta}$ rounds to $L$ once $e^{\eta}$ falls below half an ulp of $L$.
Both cases are guarded by clamping to the adjacent representable double, which
is the closest value to the bound that the machine can express while remaining
strictly inside the interval.
:::

```{r}
#| label: box-example
#| echo: false
set.seed(6)
yv <- rnorm(300, mean = 0, sd = 2)
nll <- function(s) length(yv) * log(s) + sum(yv^2) / (2 * s^2)
gnll <- function(s) length(yv) / s - sum(yv^2) / s^3

rs <- lapply(list(bfgs = optimizers7::bfgs(),
                  newton = optimizers7::newton(),
                  `nelder-mead` = optimizers7::nelder_mead()), function(o)
  optimizers7::minimize(o, nll, par = 1, gr = gnll, lower = 0))

knitr::kable(
  data.frame(method = names(rs),
             estimate = formatC(vapply(rs, function(r) r@par, numeric(1)),
                                format = "f", digits = 10),
             `strictly positive` =
               vapply(rs, function(r) r@par > 0, logical(1)),
             check.names = FALSE),
  row.names = FALSE, align = "lrl",
  caption = paste("The maximum likelihood scale of a centred Gaussian, whose",
                  "answer is the root mean square:",
                  formatC(sqrt(mean(yv^2)), format = "f", digits = 10))
)
```

Every method reaches the same estimate, and none of their implementations
contains any code about constraints.

### Adam {#sec-adam}

The last method of the chapter departs from the rule established in
@sec-descent. Adam, introduced by @kingma2015, uses no line search and makes
no attempt to decrease $f$ at every step; both omissions are deliberate, and
their purpose becomes clear with stochastic gradients.

The state of the method is a pair of exponentially weighted averages, of the
gradient and of its square, both taken coordinatewise:

$$m_t = \beta_1 m_{t-1} + (1 - \beta_1) g_t, \qquad
  v_t = \beta_2 v_{t-1} + (1 - \beta_2)\, g_t \odot g_t ,$$

with $m_0 = v_0 = 0$. The initial zeros bias both averages towards zero in the
early iterations, and the size of the bias is known exactly: under a
stationary gradient,

$$\mathbb{E}[m_t] = (1 - \beta_1^{t})\, \mathbb{E}[g],$$

so dividing by $1 - \beta_1^{t}$ removes the bias, and the same holds for
$v$. The step is then

$$x_{t+1} = x_t - \alpha\, \frac{\hat m_t}{\sqrt{\hat v_t} + \epsilon},
\qquad \hat m_t = \frac{m_t}{1 - \beta_1^{t}},
\quad \hat v_t = \frac{v_t}{1 - \beta_2^{t}},$$

with the division taken elementwise. The elementwise division is the essential
feature: a coordinate whose gradient has been consistently large is divided by
a large number and moves moderately, while a coordinate whose gradient is
small but persistent still moves. The method is a diagonal preconditioner
assembled from the gradients already computed, at no additional cost. The
diagonal form is also its limitation, because a diagonal matrix cannot
represent the correlation between parameters that @sec-bfgs extracts from the
same information.

Ignoring the bias correction and $\epsilon$, at $t = 1$ the ratio
$\hat m_1 / \sqrt{\hat v_1}$ equals $g_1 / \lvert g_1 \rvert$, which has size
exactly one, so the first step has length $\alpha$ in every coordinate
regardless of the steepness of the objective. This scale invariance is what
makes a single value of $\alpha$ usable across problems of very different
magnitudes, and it also makes Adam insensitive to a badly scaled parameter
vector, on which steepest descent performs poorly.

Without a line search there is no guarantee of monotone progress, and a run
may end at a worse point than one it passed through. This freedom is what
allows Adam to tolerate a gradient that is correct only on average: requiring
a decrease at every step would reject most of the steps that make progress
when the gradient is an estimate.

The estimated gradient is the case Adam is designed for. When the objective is
a sum over observations,

$$f(x) = \sum_{i=1}^{n} f_i(x),$$

the gradient of a randomly drawn subset of $m$ terms is, up to a factor of
$n/m$, an unbiased estimate of the full gradient, and it costs a fraction of
the full evaluation. Many inexpensive noisy steps can therefore be taken for
the price of a few exact ones, and Adam is the method built for that regime.
@bottou2018 gives a full account of stochastic optimisation.

::: {.callout-note}
## Stochastic objectives

`adam()` contains no notion of observations. The subsample is drawn by the objective itself, an ordinary function of the
parameter vector that selects a different set of terms at each call. Adam
optimises it without any indication that this is happening, and `set.seed()`
governs the run because the draws occur in the caller's code.

The package once contained machinery for drawing minibatches inside the
optimiser, and it was removed. Supporting it required a second kind of objective that declared its own
terms, a rule for which stopping criteria such an objective admitted, and a
branch through the iteration. All of this served a task the caller can do in
one line and do better, since the caller knows what an observation, a stratum
or a cluster is. The mathematics is unchanged either
way — a stochastic gradient is a stochastic gradient however it is produced —
but the knowledge now lives in the layer that has it.
:::

One consequence of the noise concerns what may be asked of a stopping rule.
When the gradient is an estimate, the objective value is one as well, and a
tolerance applied to either measures the sampling noise rather than the
progress; neither can serve as a stopping rule. A stochastic run is therefore
governed by its iteration budget and reports that it did not converge, which
is the accurate description of a run in which nothing was checked. Whether an
objective is stochastic is a property of the objective, not of the optimiser,
so this remains a caution to the caller rather than something the package can
detect.

Two safeguards address documented failure modes. The learning rate may be
reduced as $\alpha_t = \alpha / (1 + \delta t)$, which is the condition of
@robbins1951 under which a noisy run settles at the optimum instead of
orbiting it at a radius proportional to $\alpha$. And $v_t$ may be replaced by its running maximum, which prevents the second
moment from shrinking and letting a single large gradient dominate the
iterates long after it has passed. @reddi2018 exhibit a convex problem on
which Adam as originally published fails to converge, and this replacement is
their repair.

### Scope {#sec-optimizers-boundary}

Three topics lie outside this chapter.

Every method here finds a **local** minimum, and none can determine whether it
is the global one. This is a property of local optimisation, not of the
implementations; @sec-multistart describes the standard remedy, which is to
run from several starting points and to read the result for the number of
distinct optima found rather than only for the best point.

There are no constraints beyond a box. Linear and nonlinear constraints
require Lagrange multipliers and an active set, and the reparameterisation
used here does not extend to them.

There is nothing Bayesian. The toolkit intends to let the user choose a
Bayesian treatment in place of a frequentist one, but that requires a sampler
rather than an optimiser, and its output is a distribution rather than a
point. It belongs to the model layer, which is the subject of future chapters.
