## Constraints, and steps that are not descent {#sec-constraints}

```{r}
#| label: setup-04d
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Two things remain, and neither belongs inside the descent framework of the
previous sections. The first is a constraint, which is dealt with by making it
disappear; the second is a method that abandons the requirement of descent
altogether, and is useful precisely because it does.

### Removing a box {#sec-box}

Suppose each parameter is confined to an interval, $\theta_j \in (L_j, U_j)$,
with either endpoint possibly infinite. Handing that to a general optimiser is
awkward for exactly the reasons @sec-fit-scale gave: every proposal must be
tested for admissibility, a step that overshoots must be shortened rather than
taken, and near an endpoint the objective may be unbounded or undefined.

The toolkit's answer is the one it gives everywhere: do not enforce the
constraint, remove it. Each bounded coordinate is reparameterised by a
diffeomorphism $h_j : \mathbb{R} \to (L_j, U_j)$,

$$\theta_j \ =\ h_j(\eta_j) \ =\
\begin{cases}
L_j + e^{\eta_j}
  & \text{lower bound only,} \\[4pt]
U_j - e^{\eta_j}
  & \text{upper bound only,} \\[4pt]
L_j + (U_j - L_j)\,\dfrac{1}{1 + e^{-\eta_j}}
  & \text{both finite,}
\end{cases}$$

and the optimiser runs unconstrained in $\eta$. These are the bounded links of
@sec-links, and using them here is not a coincidence but the point of having
written them once: every point the optimiser proposes is admissible by
construction, so there is no rejection step and no boundary for a line search to
trip over.

What it costs is derivatives, and they are cheap because **the Jacobian is
diagonal**. Each coordinate carries its own transform and nothing else, so with
$F(\eta) = f(h(\eta))$,

$$\frac{\partial F}{\partial \eta_i} \ =\
  \frac{\partial f}{\partial \theta_i}\, h_i'(\eta_i),
\qquad
\frac{\partial^{2} F}{\partial \eta_i \partial \eta_j} \ =\
  \frac{\partial^{2} f}{\partial \theta_i \partial \theta_j}\, h_i'(\eta_i)\, h_j'(\eta_j)
  \ +\ \delta_{ij}\, \frac{\partial f}{\partial \theta_i}\, h_i''(\eta_i).$$

The second-derivative term appears only on the diagonal, and only because $h$ is
not affine; it is the whole difference between transforming a Hessian and merely
rescaling one. In matrix form,

$$\nabla^{2} F \ =\ \operatorname{diag}(h')\, \nabla^{2} f\, \operatorname{diag}(h')
  \ +\ \operatorname{diag}\!\bigl(\nabla f \odot h''\bigr),$$

which is @sec-link-scale's congruence with one extra diagonal term, and every
algorithm in this chapter receives it without ever learning a box was involved.

::: {.callout-note}
## The limitation, which is structural

An optimum lying **on** a bound cannot be reached, because getting there needs
$\eta$ to run to infinity. The optimiser marches off, improves by less and less,
and stops on a budget at a point merely close to the constraint.

For the statistical use this serves — a positive variance, a probability inside
the unit interval, a shape parameter that must not vanish — the optimum is
interior and the situation never arises. For a genuine box-constrained problem
whose solution has active constraints, an active-set method is the right tool
and this is not one.

There is a second, quieter limitation, which is arithmetic rather than
mathematical: in exact arithmetic $h$ never attains its endpoints, and in double
precision it does. The inverse logit returns exactly one above $\eta \approx 37$,
and $L + e^{\eta}$ rounds to $L$ once $e^{\eta}$ falls below half an ulp of $L$.
Both are guarded by clamping to the adjacent representable double, which is the
closest approach to the bound the machine can express while remaining strictly
inside it.
:::

```{r}
#| label: box-example
#| echo: false
set.seed(6)
yv <- rnorm(300, mean = 0, sd = 2)
nll <- function(s) length(yv) * log(s) + sum(yv^2) / (2 * s^2)
gnll <- function(s) length(yv) / s - sum(yv^2) / s^3

rs <- lapply(list(bfgs = optimizers7::bfgs(),
                  newton = optimizers7::newton(),
                  `nelder-mead` = optimizers7::nelder_mead()), function(o)
  optimizers7::minimize(o, nll, par = 1, gr = gnll, bounds = list(c(0, Inf))))

knitr::kable(
  data.frame(method = names(rs),
             estimate = formatC(vapply(rs, function(r) r@par, numeric(1)),
                                format = "f", digits = 10),
             `strictly positive` =
               vapply(rs, function(r) r@par > 0, logical(1)),
             check.names = FALSE),
  row.names = FALSE, align = "lrl",
  caption = paste("The maximum likelihood scale of a centred Gaussian, whose",
                  "answer is the root mean square:",
                  formatC(sqrt(mean(yv^2)), format = "f", digits = 10))
)
```

Every method reaches the same estimate, and none of them contains a line of code
about constraints.

### Steps that go uphill {#sec-adam}

The last method in the chapter breaks the rule the first section established.
Adam takes no line search and makes no attempt to decrease $f$ at every step,
and both omissions are deliberate.

Its state is two exponentially weighted averages, of the gradient and of its
square, both taken coordinatewise:

$$m_t = \beta_1 m_{t-1} + (1 - \beta_1) g_t, \qquad
  v_t = \beta_2 v_{t-1} + (1 - \beta_2)\, g_t \odot g_t ,$$

with $m_0 = v_0 = 0$. Those initial zeros bias both averages towards zero in the
early iterations, and the bias is known exactly: taking expectations under a
stationary gradient,

$$\mathbb{E}[m_t] = (1 - \beta_1^{t})\, \mathbb{E}[g],$$

so dividing by $1 - \beta_1^{t}$ removes it, and likewise for $v$. The step is
then

$$x_{t+1} = x_t - \alpha\, \frac{\hat m_t}{\sqrt{\hat v_t} + \epsilon},
\qquad \hat m_t = \frac{m_t}{1 - \beta_1^{t}},
\quad \hat v_t = \frac{v_t}{1 - \beta_2^{t}},$$

the division being elementwise, which is the entire idea. A coordinate whose
gradient has been consistently large is divided by a large number and moves
modestly; one whose gradient is small but persistent still moves. It is a
diagonal preconditioner assembled from the gradients already seen, and it costs
nothing beyond them — which is also its limit, since a diagonal cannot represent
the correlation between parameters that @sec-bfgs learns from the same
information.

Ignoring the bias correction and $\epsilon$, at $t = 1$ the ratio
$\hat m_1 / \sqrt{\hat v_1}$ is $g_1 / \lvert g_1 \rvert$, of size exactly one,
so the first step has length $\alpha$ in every coordinate regardless of how
steep the objective is. That scale invariance is what makes a single $\alpha$
workable across problems of very different magnitudes, and it is also why Adam
is untroubled by a badly scaled parameter vector where steepest descent is
helpless.

The absence of a line search means there is no guarantee of monotone progress,
and the run may end somewhere worse than it passed through. That freedom is most
of why Adam tolerates a gradient computed from a handful of observations rather
than from all of them: a stochastic gradient points downhill only on average, and
insisting on decrease at every step would reject most of the steps that make
progress. When the objective is a sum over observations,

$$f(x) = \sum_{i=1}^{n} f_i(x),$$

the gradient of a random subsample is an unbiased estimate of a scaled gradient
of the whole, and Adam may be run on subsamples at a fraction of the cost per
iteration.

::: {.callout-important}
## What a stochastic run may not test

When the gradient is estimated from a subsample, so is the objective, and a
tolerance applied to either measures the sampling noise rather than the progress.
Neither can serve as a stopping rule; such a run is governed by its iteration
budget, and reports that it did not converge — which is the truth about a run
that nothing checked.

This is the same discipline as everywhere else in the toolkit. A check that
cannot be evaluated must say so rather than pass or fail silently, and the
package refuses an objective-based or gradient-based rule for a subsampling run
by name rather than accepting one that would never fire.
:::

Two safeguards deserve mention because they answer real failures rather than
imagined ones. The learning rate may be reduced as $\alpha_t = \alpha / (1 +
\delta t)$, which is the Robbins–Monro condition a noisy run needs in order to
settle at the optimum rather than orbit it at a radius set by $\alpha$; and $v_t$
may be replaced by its running maximum, which forbids the second moment from
shrinking and letting a single large gradient dominate the iterate long after it
has passed. The second is not cosmetic: without it there is a *convex* problem
on which Adam as published fails to converge.

### Where the chapter ends {#sec-optimizers-boundary}

It is worth being explicit about what this chapter does not reach, since the
apparatus is capable of a good deal more than it has been asked to do.

Every method here finds a **local** minimum, and none can tell whether it is the
global one. That is a property of local optimisation and not a shortcoming of
the implementations; the only general answer is to run from several starting
points, and the most valuable thing that produces is not the best point but the
count of distinct optima found, which is direct evidence about whether the
question has one answer.

There are no constraints beyond a box. Linear and nonlinear constraints need
Lagrange multipliers and an active set, and the reparameterisation trick that
serves so well here has nothing to say about them.

Nor is there anything Bayesian. The toolkit's stated ambition is that the user
should be able to choose a Bayesian treatment in place of a frequentist one, and
that is a different object entirely — a sampler rather than an optimiser, whose
output is a distribution and not a point. It belongs to the model layer, which is
where this book will go next.
