## Starting values and multiple starts {#sec-starting}

```{r}
#| label: setup-04e
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

The previous sections took the starting point as given. This section examines
how to choose it, why the natural choice is wrong in an easily missed way, and
a related question that no single run can address: how many distinct minima
the problem has.

### Starting values on the unconstrained scale {#sec-starting-scale}

With no other information, the natural starting point is a vector of zeros.
For a model with a scale parameter this is not a starting point but a
boundary: at $\sigma = 0$ the log-likelihood is usually infinite, its gradient
certainly is, and an optimizer started there fails before its first step.
Replacing the zeros with ones only moves the problem, because one is a
reasonable scale but a poor probability.

A single constant works for every kind of parameter when it is applied on the
unconstrained scale, using the change of variable of @sec-box. On that scale
there is no boundary, and the point $\eta = 0$ maps to a reasonable interior
value of whatever the parameter is:

$$
h(0) \ =\
\begin{cases}
0 & \text{unbounded,}\\[2pt]
L + 1 & \text{bounded below by } L,\\[2pt]
U - 1 & \text{bounded above by } U,\\[2pt]
(L + U)/2 & \text{bounded on both sides.}
\end{cases}
$$

A variance starts at one, a probability at one half, a correlation at zero,
and a parameter confined to $(2, 5)$ at $3.5$, all from the same instruction.
The same argument applies to a random start: a draw made on the unconstrained
scale and mapped back is always admissible, however wide the range, while a
draw made on the parameter scale must be rejected and redrawn whenever it
falls outside the box.

```{r}
#| label: starters
f <- function(p) sum((p - c(1, 2, 3))^2)

optimizers7::minimize(optimizers7::bfgs(), f, optimizers7::start_zeros(3),
                      lower = 0)@par

set.seed(1)
optimizers7::minimize(optimizers7::bfgs(), f,
                      optimizers7::start_runif(-3, 3, npar = 3),
                      lower = c(0, 0, 0), upper = c(5, 5, 5))@par
```

A starter must be told how many parameters there are, and this is the one
delicate point. The number can be stated directly; it can be read from bounds
supplied one per parameter; or the objective can be probed, by evaluating it
at candidate lengths until one is accepted. Probing works when the objective
rejects a parameter vector of the wrong length, which the objective of a real
model does: `X %*% beta` with a mismatched `beta` is an error, not a number.
It does not work for a function written in terms of the parameter alone,
because R recycles a shorter vector silently whenever its length divides the
longer one. `sum((p - c(1, 2, 3))^2)` is therefore a perfectly finite
function of one parameter as well as of three. In that case the count cannot be inferred,
and the user is asked to supply it.

A second check precedes the first iteration whenever a gradient is supplied.
A `gr` computed from a different model than `fn` — a fixed predictor where
the objective uses the parameter, a stale copy of the data — leaves the
objective flat along a direction the gradient reports as downhill. The run
then ends with a line-search failure at the first iteration, which describes
the symptom and conceals the cause. Two evaluations settle the question in advance. One central difference of
`fn` along the gradient direction at the starting point is compared with the
rate `gr` predicts, which along that direction is $\lVert g \rVert$, and a
gross disagreement between the two produces a warning that names both
rates. The tolerance is deliberately loose, so that
finite-difference error, or a subgradient evaluated at a kink, does not
trigger it: the check detects the wrong function, not the eighth digit.

```{r}
#| label: gradient-guard
#| warning: true
f_wrong <- function(p) 42            # constant: does not depend on p
g_right <- function(p) 2 * (p - c(1, 2))
r <- optimizers7::minimize(optimizers7::bfgs(maxit = 2), f_wrong, c(0, 0),
                           gr = g_right)
```

### Multiple starts and distinct optima {#sec-multistart}

Every method in this chapter finds a local minimum, and none of them can
determine whether it is the global one. This is a property of local
optimization, not of any implementation, and the standard remedy is to run
from several starting points. Global optimization has a large literature of
its own — @rinnooykan1987 is the classical treatment of multi-start methods
and of how many local searches are worth running — but the practical version
is short.

What matters is how the several runs are read. The best point found is the
obvious summary and the least informative one, because a better point may
always lie in a basin that no start reached. The informative summary is the
**number of distinct optima**, because it describes the objective rather than
the optimizer: twenty starts that all report the same optimum are evidence
that the problem has a single answer, which no single convergent run can
establish.

```{r}
#| label: multistart
tw <- function(p) (p[1]^2 - 1)^2 + p[2]^2 + 0.3 * p[1]
set.seed(1)
r <- optimizers7::minimize(
  optimizers7::multistart(optimizers7::bfgs(), n = 12), tw, c(0, 0))
r@message
table(round(r@trace$value, 6))
```

The starts themselves form a **Latin hypercube** rather than independent
draws: each coordinate's range is divided into as many strata as there are
starts and each stratum is used exactly once, so no region of the range is
left empty by chance. The construction is due to @mckay1979, and it is applied
on the unconstrained scale for the reason given in the previous section.

### Parallel execution {#sec-multistart-parallel}

The starts are independent, so they parallelise naturally; they are executed
in separate processes rather than threads for a specific reason. A stopping
criterion in this package is an R object, consulted at every iteration — this
is the design the package is built around — so every iteration returns to R,
and R cannot be called from several threads at once. Separate processes have
no shared interpreter and no such restriction.

Everything else is handled internally. A number of processes is chosen — as
many as there are starts, but never more than the machine can spare — and the
workers are created, given the package and an independent random stream, and
shut down, all inside the call. On Unix-alikes the workers are forks, which
start in microseconds and inherit the session; Windows has no `fork`, so a
socket cluster is used instead. The user supplies at most a number of cores.

One property of this arrangement deserves explicit statement, because it is
easy to lose and difficult to restore. The starting points are drawn in the
calling session, before any work is dispatched, and each worker is seeded from
that session's stream. `set.seed()` therefore reproduces the run, and reproduces it identically
whether the work was split across eight processes or none. A scheme in which
each worker seeded itself would produce answers that depend on how many cores
happened to be free, which is not acceptable in a result that will be
reported.

```{r}
#| label: consistency-starting
#| include: false
# Internal consistency gate for this section: fails the render if the printed
# values of h(0) are not the ones the package computes, if a Latin hypercube
# leaves a stratum empty, or if the gradient guard fails to fire on a
# mismatched pair or fires on a correct one.
assert_starting_ok()
```
