## Where a run begins, and how many runs there are {#sec-starting}

```{r}
#| label: setup-04e
#| include: false
source("R/_setup.R")
source("R/optimizer-certificates.R")
```

Everything so far has taken the starting point as given. It is worth a section of
its own, partly because the natural choice is wrong in a way that is easy to miss,
and partly because the answer to "where should I start?" turns out to be bound up
with a rather different question — how many answers the problem has — which no
single run can address at all.

### The right scale on which to be simple {#sec-starting-scale}

Asked for a starting point with nothing else to go on, one writes a vector of
zeros. For a model with a scale parameter in it that is not a starting point but
a boundary: at $\sigma = 0$ the log-likelihood is usually infinite and its
gradient certainly is, and an optimiser handed such a point fails before it has
taken a step. The obvious repair, writing ones instead, merely moves the problem,
since one is a perfectly good scale and a very poor probability.

What makes a single constant work for every kind of parameter is the change of
variable of @sec-box. On the unconstrained scale there is no boundary to sit on,
and the point $\eta = 0$ maps to a sensible interior value of whatever the
parameter happens to be:

$$
h(0) \ =\
\begin{cases}
0 & \text{unbounded,}\\[2pt]
L + 1 & \text{bounded below by } L,\\[2pt]
U - 1 & \text{bounded above by } U,\\[2pt]
(L + U)/2 & \text{bounded on both sides.}
\end{cases}
$$

A variance starts at one, a probability at one half, a correlation at zero, and a
parameter confined to $(2, 5)$ at $3.5$ — from the same instruction each time.
The same argument applies to a random start: draw on the unconstrained scale and
map back, and the draw cannot be inadmissible however wide the range, whereas a
draw made on the parameter scale has to be rejected and redrawn whenever it falls
outside the box.

```{r}
#| label: starters
f <- function(p) sum((p - c(1, 2, 3))^2)

optimizers7::minimize(optimizers7::bfgs(), f, optimizers7::start_zeros(3),
                      lower = 0)@par

set.seed(1)
optimizers7::minimize(optimizers7::bfgs(), f,
                      optimizers7::start_runif(-3, 3, npar = 3),
                      lower = c(0, 0, 0), upper = c(5, 5, 5))@par
```

There is one thing such an object has to be told, which is how many parameters
there are, and it is the only genuinely awkward part. It may be said outright; it
may be read off bounds given one per parameter; or the objective may be probed,
by evaluating it at candidate lengths until one is accepted. The last works when
the objective rejects the wrong length, which the objective of a real model does,
`X %*% beta` with a parameter of the wrong length being an error rather than a
number. It does not work for a function written in terms of the parameter alone,
because R recycles a shorter vector silently whenever its length divides, so
`sum((p - c(1, 2, 3))^2)` is a perfectly finite function of one parameter as well
as of three; there the count cannot be inferred and is asked for instead.

### Several starts, and what they measure {#sec-multistart}

Every method in this chapter finds a local minimum, and none of them can tell
whether it is the global one. That is a property of local optimisation rather
than a shortcoming of any implementation, and the only general answer is to run
from several starting points. Global optimisation has a substantial literature of
its own — @rinnooykan1987 is the classical treatment of multi-start and of how
many local searches are worth doing — but the practical version is short.

What matters is what the several runs are read for. The best point found is the
obvious answer and the least informative one, since a better point may always sit
in a basin no start happened to land in. The **number of distinct optima** is the
quantity that carries information, because it is evidence about the objective
rather than about the optimiser: a run reporting one optimum from twenty starts
says the question asked has one answer, which no single convergent run can
establish however cleanly it converges.

```{r}
#| label: multistart
tw <- function(p) (p[1]^2 - 1)^2 + p[2]^2 + 0.3 * p[1]
set.seed(1)
r <- optimizers7::minimize(
  optimizers7::multistart(optimizers7::bfgs(), n = 12), tw, c(0, 0))
r@message
table(round(r@trace$value, 6))
```

The starts themselves are a **Latin hypercube** rather than independent draws:
each coordinate's range is cut into as many strata as there are starts and each
stratum is used exactly once, so no corner of the region can be left empty by
chance. The construction is @mckay1979, and it is applied on the unconstrained
scale for the reason the previous section gave.

### Why the parallelism is at the level of processes {#sec-multistart-parallel}

The starts are independent, so they parallelise perfectly, and it is worth saying
why they are given separate processes rather than threads. The reason is not the
objective but the stopping rule. A criterion in this package is an R object,
consulted at every iteration; that is the feature the whole design is arranged
around. Every iteration therefore returns to R, and R is single-threaded, so
calling it from several threads at once is undefined behaviour that crashes
rather than errors. Separate processes have no shared interpreter and no such
problem.

The rest is arranged rather than asked for. A number of processes is chosen —
as many as there are starts, and never more than the machine can spare — the
workers are created, given the package and an independent random stream, and shut
down again, all inside the call. On a Unix-alike they are forks, which cost
microseconds and inherit the session entire; Windows has no `fork`, so a socket
cluster is used instead. The user sees a number.

One property is worth stating because it is easily lost and hard to recover. The
starting points are drawn in the calling session, before anything is dispatched,
and each worker is seeded from that session's stream; so `set.seed()` reproduces
the run, and reproduces it identically whether the work was split eight ways or
not at all. A scheme that let each worker seed itself would give an answer that
depended on how many cores happened to be free, which is not a property one wants
in a result that will be reported.

```{r}
#| label: consistency-starting
#| include: false
# Internal consistency gate for this section: fails the render if the printed
# values of h(0) are not the ones the package computes, or if a Latin hypercube
# leaves a stratum empty.
assert_starting_ok()
```
