## In practice {#sec-optim-practice}

```{r}
#| label: setup-04f
#| include: false
source("R/_setup.R")
```

This section shows the interface of the package: the stopping rules as
objects, a rule written by a user, the validator, and bounds. The examples
run on Rosenbrock's function, defined once:

```{r}
#| label: op-setup
library(optimizers7)
f  <- function(p) 100 * (p[2] - p[1]^2)^2 + (1 - p[1])^2
gr <- function(p) c(-400 * p[1] * (p[2] - p[1]^2) - 2 * (1 - p[1]),
                    200 * (p[2] - p[1]^2))
```

### The stopping rules {#sec-optim-criteria-practice}

A criterion is an object handed to the optimizer, not a number buried in a
loop. The package ships rules on the gradient (`crit_grad`), on the change
of the objective (`crit_rel_obj`, `crit_abs_obj`), on the change of the
parameters (`crit_rel_par`, `crit_abs_par`), on the bundle method's
stationarity measure (`crit_stationary`), and the rule that never fires
(`crit_never`, which leaves the iteration budget in charge). They combine
with `crit_any()` and `crit_all()`:

```{r}
#| label: op-criteria
runs <- list(
  gradient  = minimize(bfgs(criterion = crit_grad(1e-8)), f, c(-1.2, 1), gr = gr),
  objective = minimize(bfgs(criterion = crit_rel_obj(1e-12)), f, c(-1.2, 1), gr = gr),
  either    = minimize(bfgs(criterion = crit_any(crit_grad(1e-8), crit_rel_obj(1e-12))),
                       f, c(-1.2, 1), gr = gr)
)
data.frame(iterations = sapply(runs, function(r) r@iterations),
           criterion_met = sapply(runs, function(r) r@criterion_met),
           gradient_norm = sapply(runs, function(r) max(abs(r@gradient))))
```

The two rules stop the same method at different points: the objective rule
fires while the gradient is still large, which is the reason
@sec-fit-scoring dropped it from the default of `fit_distrib()`.

A rule the package does not ship is one S7 class and one method. The rule
below stops when the objective falls under a target value:

```{r}
#| label: op-custom-criterion
CritTarget <- S7::new_class("CritTarget", parent = criterion,
                            properties = list(target = S7::class_numeric))
S7::method(crit_met, CritTarget) <- function(criterion, state) {
  state$f_new < criterion@target
}

r <- minimize(bfgs(criterion = CritTarget(target = 1e-6)), f, c(-1.2, 1), gr = gr)
c(value = r@value, iterations = r@iterations)
```

An optimizer rejects a rule it cannot evaluate, before the first iteration:

```{r}
#| label: op-reject-rule
#| error: true
minimize(nelder_mead(criterion = crit_grad(1e-8)), f, c(-1.2, 1))
```

### The validator {#sec-optim-check-practice}

`check_optimizer()` verifies what a method reports (that `value` is the
objective at `par`, that `converged` follows the stopping rule, that bounds
hold strictly) and then runs the standard problems, whose gaps are printed
as information rather than as verdicts:

```{r}
#| label: op-check
res <- check_optimizer(bb(), verbose = FALSE)
data.frame(passed = res$checks)          # the twelve numbered checks
res$battery[, c("problem", "gap", "converged")] # the standard problems, as information
```

### Bounds without constraint code {#sec-optim-bounds-practice}

`lower` and `upper` are removed by the change of variable of @sec-box, so
no method contains constraint handling and every iterate is admissible by
construction. The maximum likelihood scale of a centered Gaussian, kept
positive:

```{r}
#| label: op-bounds
set.seed(6)
yv <- rnorm(300, sd = 2)
nll  <- function(s) length(yv) * log(s) + sum(yv^2) / (2 * s^2)

r <- minimize(bfgs(), nll, par = 1, lower = 0)
c(estimate = r@par, closed_form = sqrt(mean(yv^2)))
```

### Reading a multistart {#sec-optim-multistart-practice}

`multistart()` runs one optimizer from a Latin hypercube of starts and its
result is read for the number of distinct optima, which is what several
runs measure and one cannot:

```{r}
#| label: op-multistart
tw <- function(p) (p[1]^2 - 1)^2 + p[2]^2 + 0.3 * p[1]
set.seed(1)
ms <- minimize(multistart(bfgs(), n = 12), tw, c(0, 0))
ms@message
table(distinct_optimum = round(ms@trace$value, 6))
```
