## Basis expansions {#sec-basis-expansions}

```{r}
#| label: setup-05a
#| include: false
source("R/_setup.R")
source("R/basis-certificates.R")
```

### The expansion and its design matrix {#sec-basis-definition}

A basis on an interval $[a, b]$ is a finite collection of known functions
$B_1, \ldots, B_K$ defined there. The functions it represents are the linear
combinations

$$
f(x) \;=\; \sum_{r=1}^{K} \beta_r B_r(x) \;=\; B(x)^\top \beta ,
\qquad x \in [a, b],
$$ {#eq-basis-expansion}

where $B(x)$ is the column vector of the $K$ values and $\beta$ the vector of
coefficients. The functions are fixed and known before any data are seen; the
coefficients are what a model estimates. This is what makes the expansion
useful in the framework of the previous chapters, since $f$ is nonlinear in
$x$ while remaining linear in the parameters, and a linear predictor is what
every routine in the toolkit already handles.

Evaluated at $n$ points $x_1, \ldots, x_n$, the expansion produces the
$n \times K$ **design matrix** $B$ with entries $B_{ir} = B_r(x_i)$, so that
the vector of function values is $B\beta$. The design matrix is the object a
fitting routine consumes, and everything in this chapter is either a way of
producing one or a matrix that acts on its columns.

Three further quantities follow from the same functions, and a modeling
routine needs all of them. Differentiating @eq-basis-expansion term by term
gives

$$
f^{(d)}(x) \;=\; \sum_{r=1}^{K} \beta_r B_r^{(d)}(x),
$$ {#eq-basis-derivative}

so a derivative of the fitted function is the same linear combination of a
different matrix. Integrating gives the same statement for the antiderivative.
And the integrals of the products $B_r^{(d)} B_s^{(d)}$ form the matrix that a
roughness penalty is a quadratic form in, which is the subject of
@sec-basis-gram.

The package therefore asks four questions of a basis, and answers them in the
same shape. `basis_eval()` returns the design matrix, `basis_deriv()` the
matrix of $d$-th derivatives, `basis_int()` the matrix of integrals, and
`basis_gram()` the $K \times K$ matrix of inner products. The first three have
one row per evaluation point and one column per basis function.

```{r}
#| label: basis-object
b <- bspline_basis(lower = 0, upper = 1, dimension = 6, degree = 3)
b

round(basis_eval(b, c(0, 0.25, 0.5, 1)), 4)
```

A basis also knows its own interval and refuses a point outside it. The
refusal happens in the body of the generic, before dispatch, so a basis
written outside the package inherits it without doing anything.

### The integral is anchored at the lower endpoint {#sec-basis-anchor}

The integral of a basis is fixed by the convention that its value at $a$ is
zero:

$$
\left(\textstyle\int B\right)_r(x) \;=\; \int_{a}^{x} B_r(v)\,\mathrm{d}v,
\qquad \left(\textstyle\int B\right)_r(a) \;=\; 0 .
$$ {#eq-basis-anchor}

The convention is part of the contract rather than a detail of the
implementation, because the antiderivative is otherwise determined only up to
a constant. Any choice of constant satisfies the check that differentiating
the integral returns the basis, so two bases could use different constants,
each be internally consistent, and produce a sum whose integral is wrong by
the difference. Fixing the value at one point removes the freedom, and the
lower endpoint is the natural point to fix it at, since it belongs to every
basis interval.

```{r}
#| label: basis-anchor
f <- fourier_basis(lower = -1, upper = 3, dimension = 5)
basis_int(f, -1)
```

### Numerical methods for a basis that supplies only its values {#sec-basis-fallbacks}

Only `basis_eval()` is compulsory. The derivatives, the integral and the Gram
matrix have numerical methods registered on the base class, so a new basis is
a subclass and one method, and a closed form supplied later replaces the
corresponding numerical method through dispatch with no change to calling
code. This is the arrangement `distributions7` offers a distribution defined
by its density alone.

The derivatives are computed from a single finite-difference stencil, whose
weights are determined by the requirement that the stencil be exact on
polynomials. Writing $\delta_1, \ldots, \delta_p$ for the offsets, in units of
a step $h$, the Taylor expansion of the combination
$\sum_j w_j f(x + \delta_j h)$ multiplies $f^{(i)}(x)$ by
$h^i/i! \sum_j w_j \delta_j^i$. Requiring

$$
\sum_{j=1}^{p} w_j \delta_j^{\,i} \;=\;
\begin{cases}
0, & i \neq d,\\
d!, & i = d,
\end{cases}
\qquad i = 0, \ldots, p-1,
$$ {#eq-fd-weights}

therefore leaves $h^d f^{(d)}(x)$ and nothing else, up to the first neglected
term. This is a Vandermonde system in the offsets, solved once per stencil
shape, and the derivative estimate is $h^{-d}\sum_j w_j f(x + \delta_j h)$. The
construction is due to @fornberg1988, who gives a recurrence for the weights
on an arbitrary set of offsets. The stencil library itself lives in
`numericals7` (@sec-numerical-stencils), where every package of the toolkit
draws it from.

Building the weights this way, rather than composing lower-order differences,
is what keeps a high order usable. Each numerical differentiation multiplies
the error of the one before it, so a fourth derivative reached by four nested
first differences is dominated by rounding error rather than by the function.
The same rule governs the validators of `linkfunctions7` and
`distributions7`: one stencil, applied to the highest order available in
closed form, and never a chain.

The step is $h = \varepsilon^{1/(d+2)}\max(1, \lvert x \rvert)$, which balances
truncation error against rounding error at order $d$, and it is capped so that
the whole stencil fits inside the interval. Near an endpoint a symmetric
stencil would ask for points outside the interval, where the basis is not
defined, so those points use a one-sided stencil with the same number of nodes
and the same order, built from the same system @eq-fd-weights with shifted
offsets.

The construction below defines a basis of Gaussian bumps by its values alone.
Its derivative is available immediately, and agrees with the closed form
nothing in the object knows about.

```{r}
#| label: basis-fallback
Bumps <- S7::new_class("Bumps", parent = basis, package = NULL)

S7::method(basis_eval, Bumps) <- function(basis, x, ...) {
  p <- basis@basis_params
  out <- exp(-0.5 * outer(x, p$centers, "-")^2 / p$width^2)
  colnames(out) <- basis_colnames(basis)
  out
}

bumps <- Bumps(
  basis_name = "bumps", dimension = 5L, lower = 0, upper = 1,
  basis_params = list(centers = seq(0.1, 0.9, length.out = 5), width = 0.15)
)

xs <- c(0.2, 0.5)
round(basis_deriv(bumps, xs, order = 1), 6)

p <- bumps@basis_params
round(-outer(xs, p$centers, "-") / p$width^2 * basis_eval(bumps, xs), 6)
```

An object says which of its answers are numerical. A caller can then tell an
exact quantity from an approximated one, and the validator can report a
quantity it has no independent way of checking as unchecked rather than as
passed.

```{r}
#| label: basis-is-numerical
basis_is_numerical(bumps)
```

```{r}
#| label: consistency-05a
#| include: false
assert_expansions_ok()
```
