## The three families {#sec-basis-families}

```{r}
#| label: setup-05b
#| include: false
source("R/_setup.R")
source("R/basis-certificates.R")
```

The package ships three families, chosen so that the common shapes of
covariate effect are covered. One is local, for effects with no global form;
one is periodic, for effects that repeat; and one is a global polynomial, for
effects that do have a form and need only a few terms. Each is generated by a
recurrence or an identity that produces its derivatives and its integral as
well as its values, so no order is computed by differencing.

```{r}
#| label: fig-basis-families
#| fig-cap: "The three families on $[0, 1]$: a cubic B-spline basis of six functions, a Fourier basis of five, and the first five Legendre polynomials."
#| fig-height: 3.0
#| fig-width: 9
par(mfrow = c(1, 3))
plot(bspline_basis(dimension = 6, degree = 3))
plot(fourier_basis(dimension = 5))
plot(poly_basis(dimension = 5))
```

### B-splines {#sec-basis-bspline}

A B-spline basis is determined by a degree $m$ and a non-decreasing knot
vector $t_1 \le t_2 \le \cdots \le t_{K+m+1}$. The package repeats each
endpoint $m+1$ times and spaces the $K - m - 1$ interior knots equally in
between. That configuration makes the basis span all piecewise polynomials of
degree $m$ on $[a, b]$ with $m-1$ continuous derivatives at the interior
knots. The dimension $K$ and the degree
therefore determine the number of interior knots, and $K \ge m + 1$ is
required, with equality giving the polynomials of degree $m$ on the whole
interval and no interior knot at all.

The functions themselves come from the recurrence of @cox1972 and @deboor1972,
which starts from indicators of the knot intervals and raises the degree one
step at a time:

$$
B_{i,0}(x) \;=\;
\begin{cases}
1, & t_i \le x < t_{i+1},\\
0, & \text{otherwise,}
\end{cases}
$$ {#eq-bspline-zero}

$$
B_{i,q}(x) \;=\;
\frac{x - t_i}{t_{i+q} - t_i}\, B_{i,q-1}(x)
\;+\;
\frac{t_{i+q+1} - x}{t_{i+q+1} - t_{i+1}}\, B_{i+1,q-1}(x),
\qquad q = 1, \ldots, m,
$$ {#eq-cox-de-boor}

with the convention that a term whose denominator vanishes is taken to be
zero, which is what repeated knots produce. @deboor2001 is the reference
account of the recurrence and of its numerical behavior, which is why it is
used in preference to any explicit polynomial form. Every coefficient in
@eq-cox-de-boor is a convex weight, so no cancellation occurs and the
evaluation is stable at any degree.

Two properties of the resulting functions matter for what follows. The first
is **local support**: $B_{i,m}$ vanishes outside $[t_i, t_{i+m+1}]$, so at any
$x$ at most $m+1$ of the $K$ functions are non-zero. The design matrix is
therefore banded, and so is the Gram matrix of @sec-basis-gram, which is what
keeps a spline model of a thousand coefficients tractable. The second is that
the functions form a **partition of unity**, $\sum_{i=1}^{K} B_{i,m}(x) = 1$
on $[a, b]$, so the constant function is in the span and the basis needs no
separate intercept.

```{r}
#| label: bspline-local-support
b <- bspline_basis(dimension = 8, degree = 3)
round(basis_eval(b, c(0.1, 0.5, 0.9)), 3)

rowSums(basis_eval(b, seq(0, 1, length.out = 5)))
```

Derivatives are not computed by differencing @eq-cox-de-boor but by a second
recurrence of the same kind, which expresses the derivative of a spline of
degree $q$ through two splines of degree $q-1$:

$$
B_{i,q}'(x) \;=\; q \left(
\frac{B_{i,q-1}(x)}{t_{i+q} - t_i}
\;-\;
\frac{B_{i+1,q-1}(x)}{t_{i+q+1} - t_{i+1}}
\right).
$$ {#eq-bspline-derivative}

Applying it $d$ times gives the $d$-th derivative exactly, and shows why a
spline of degree $m$ has exactly $m$ non-trivial derivatives: each application
lowers the degree by one, and a derivative of order above $m$ is the zero
function. The package returns the zero matrix for such an order, which is the
value of that derivative rather than a refusal. The integral obeys a
recurrence of the same family, and the package takes it in the form that
already satisfies @eq-basis-anchor.

```{r}
#| label: bspline-derivative
round(basis_deriv(b, 0.4, order = 2), 3)

# a cubic spline has no fourth derivative
all(basis_deriv(b, 0.4, order = 4) == 0)
```

### The Fourier basis {#sec-basis-fourier}

The Fourier basis holds the constant function together with $J$ sine-cosine
pairs of increasing frequency. Writing $\omega$ for the period and

$$
z \;=\; \frac{2\pi (x - a)}{\omega}
$$

for the phase, the functions are $1$, and $\sin(jz)$, $\cos(jz)$ for
$j = 1, \ldots, J$, so the dimension is $K = 2J + 1$ and is necessarily odd. A
sine kept without its cosine would give a basis able to represent a wave at
one phase and not at another, so a dimension that would leave half a pair is
refused rather than adjusted.

Every derivative and the antiderivative come from one identity. Differentiating
a sinusoid rescales it and advances its phase by a quarter period, and
iterating that statement gives

$$
\frac{\mathrm{d}^{k}}{\mathrm{d}x^{k}} \sin(jz)
\;=\; \left(\frac{2\pi j}{\omega}\right)^{k}
\sin\!\left(jz + \frac{k\pi}{2}\right),
$$ {#eq-fourier-shift}

and the same with the cosine in both places. No order is special: the identity
is one line of code for every $k$, and it holds for negative $k$ as well,
where it gives an antiderivative. That antiderivative is not the one
@eq-basis-anchor asks for, and the two differ by a constant that is not the
same in every column, since at $x = a$ the raw sine columns take the value
$-\omega/(2\pi j)$ while the cosine columns are already zero. Subtracting the
value at the lower endpoint corrects every column at once.

```{r}
#| label: fourier-shift
fb <- fourier_basis(dimension = 5)
basis_colnames(fb)

round(basis_deriv(fb, c(0.1, 0.6), order = 3), 4)
```

The period defaults to the width of the interval, which is what makes the
functions orthogonal on it and gives the closed-form Gram matrix of
@sec-basis-gram. A different period is accepted, and the basis then spans a
non-integer number of periods, so the orthogonality is lost and the inner
products are computed by quadrature instead.

### The Legendre polynomials {#sec-basis-legendre}

The third family is the polynomials of degree below $K$ on $[a, b]$, expressed
in the Legendre basis rather than in raw powers. The two span the same space,
so the choice between them is entirely numerical. The Gram matrix of the raw
powers on $[0,1]$ has entries $1/(r+s-1)$, which is the Hilbert matrix, whose
condition number grows like $e^{3.5K}$; a basis of ten raw powers is already
close to singular in double precision. The Legendre polynomials are
orthogonal, so their Gram matrix is diagonal and perfectly conditioned.

They are defined on $[-1, 1]$ by the three-term recurrence

$$
(n+1) P_{n+1}(\tau) \;=\; (2n+1)\, \tau\, P_n(\tau) \;-\; n\, P_{n-1}(\tau),
\qquad P_0(\tau) = 1, \quad P_1(\tau) = \tau,
$$ {#eq-legendre-recurrence}

and carried to the basis interval by the affine change of variable
$\tau = 2(x - a)/(b - a) - 1$. @abramowitz1964 collects the standard
identities, of which two more are used here. Differentiating

$$
P_{n+1}'(\tau) - P_{n-1}'(\tau) \;=\; (2n+1) P_n(\tau)
$$

$d-1$ further times gives a recurrence that produces each derivative order
from the one below it,

$$
P_{n+1}^{(d)}(\tau) \;=\; P_{n-1}^{(d)}(\tau) \;+\; (2n+1) P_{n}^{(d-1)}(\tau),
$$ {#eq-legendre-derivative}

so every order is exact and none is reached by differencing; the change of
variable contributes a factor $\left(2/(b-a)\right)^{d}$. Reading the same
identity as an antiderivative gives

$$
\int P_n \,\mathrm{d}\tau \;=\; \frac{P_{n+1} - P_{n-1}}{2n+1},
\qquad n \ge 1,
$$ {#eq-legendre-integral}

which satisfies @eq-basis-anchor without a correction, because at $\tau = -1$
the combination $P_{n+1} - P_{n-1}$ is $(-1)^{n+1} - (-1)^{n-1} = 0$.

```{r}
#| label: legendre
pb <- poly_basis(dimension = 5)
round(basis_eval(pb, c(0, 0.5, 1)), 4)

# the same span as the raw powers, in a better conditioned form
x <- seq(0, 1, length.out = 40)
max(abs(lm.fit(basis_eval(pb, x), x^3)$fitted.values - x^3))
```

```{r}
#| label: consistency-05b
#| include: false
assert_families_ok()
```
