## Inner products {#sec-basis-gram}

```{r}
#| label: setup-05c
#| include: false
source("R/_setup.R")
source("R/basis-certificates.R")
```

### The Gram matrix and the roughness penalty {#sec-gram-definition}

The **Gram matrix** of order $d$ collects the inner products of the $d$-th
derivatives of the basis functions,

$$
\left(G_d\right)_{rs} \;=\;
\int_{a}^{b} B_r^{(d)}(x)\, B_s^{(d)}(x)\, \mathrm{d}x ,
\qquad r, s = 1, \ldots, K .
$$ {#eq-gram}

Its importance comes from @eq-basis-derivative. Squaring that expansion and
integrating gives

$$
\int_{a}^{b} \left\{ f^{(d)}(x) \right\}^2 \mathrm{d}x
\;=\; \beta^\top G_d\, \beta ,
$$ {#eq-gram-quadratic-form}

so the integrated squared $d$-th derivative of the fitted function, which is
the standard measure of its roughness, is a quadratic form in the coefficients
with $G_d$ as its matrix. Penalised estimation adds
$\lambda \beta^\top G_d \beta$ to the negative log-likelihood, and the
quantity the penalty needs is exactly @eq-gram. The formulation is that of
@wahba1990 and @greensilverman1994, and @wood2017 gives the modern account in
the additive model setting.

$G_d$ is symmetric and positive semidefinite by construction, and for
$d \ge 1$ it is singular: any $\beta$ whose expansion is a polynomial of
degree below $d$ contributes nothing to the integral, so the null space has
dimension at least $d$ whenever the basis contains those polynomials. The
null space is the **unpenalised** part of the model, the set of functions the
penalty leaves alone however large $\lambda$ becomes, and for $d = 2$ it is
the straight lines.

```{r}
#| label: gram-quadratic-form
b <- bspline_basis(dimension = 6, degree = 3)

set.seed(1)
beta <- rnorm(6)
f2 <- function(v) (basis_deriv(b, v, order = 2) %*% beta)^2

c(
  quadratic_form = drop(t(beta) %*% basis_gram(b, order = 2) %*% beta),
  integral = integrate(function(v) apply(cbind(v), 1, f2), 0, 1)$value
)
```

A difference penalty on the coefficients,
$\beta^\top \Delta_d^\top \Delta_d \beta$ with $\Delta_d$ the $d$-th
difference operator, is the alternative in common use and is the P-spline
construction of @eilers1996. It is a different matrix, cheaper to form and
approximating @eq-gram for equally spaced knots, and the package accepts
either wherever a penalty is an argument. The distinction matters because
@eq-gram is a property of the basis functions alone, while a difference
penalty is a property of how the coefficients are indexed.

### Exact integration {#sec-gram-exact}

The integrand of @eq-gram is a product of two piecewise polynomials, so the
integral can be computed without approximation. On each knot interval the
order-$d$ derivative of a spline of degree $m$ is a polynomial of degree
$m - d$, so the integrand there has degree $2(m-d)$. A Gauss-Legendre rule
with $p$ nodes integrates polynomials of degree up to $2p - 1$ exactly, so
$p = m - d + 1$ nodes leave no quadrature error at all, and the result carries
only floating-point error. The rule is applied interval by interval, and the
whole matrix is one crossproduct of the weighted derivative matrix evaluated
at the collected nodes.

The nodes and weights are computed rather than tabulated, by the construction
of @golub1969: the nodes are the eigenvalues of the symmetric tridiagonal
Jacobi matrix of the Legendre recurrence @eq-legendre-recurrence, and the
weights come from the first component of each normalised eigenvector.
Computing them keeps any node count available, which the rule above needs,
since the count depends on the degree and on the derivative order.

The other two families do better still. Over a whole period the trigonometric
functions are mutually orthogonal, and they remain so after differentiation,
since @eq-fourier-shift only rescales a function and shifts its phase. The
Gram matrix is therefore diagonal, with entries

$$
\omega \ \text{ for the constant at } d = 0, \qquad
0 \ \text{ for the constant at } d \ge 1, \qquad
\frac{\omega}{2}\left(\frac{2\pi j}{\omega}\right)^{2d}
\ \text{ for both members of pair } j .
$$ {#eq-fourier-gram}

For the Legendre polynomials the orthogonality relation
$\int_{-1}^{1} P_n P_{n'} \,\mathrm{d}\tau$, which is $2/(2n+1)$ when
$n = n'$ and zero otherwise, gives after the change of variable a diagonal
matrix with entries $(b - a)/(2n+1)$ at order zero. Their derivatives are not orthogonal, but the integrand remains a
polynomial of degree at most $2(K - 1 - d)$, so a single rule with $K$ nodes
is exact on the whole interval.

```{r}
#| label: gram-closed-forms
round(basis_gram(fourier_basis(dimension = 5), order = 1), 4)

round(basis_gram(poly_basis(lower = 0, upper = 2, dimension = 4)), 6)
```

The local support of @sec-basis-bspline shows in the spline case as a band:
$\left(G_d\right)_{rs}$ is zero whenever $\lvert r - s \rvert > m$, since the
supports of the two functions do not overlap.

```{r}
#| label: gram-band
round(basis_gram(bspline_basis(dimension = 8, degree = 2), order = 1), 2)
```

### The measure the inner product is taken against {#sec-gram-measure}

@eq-gram integrates against Lebesgue measure on $[a, b]$, which is the measure
a roughness penalty uses: the penalty is a statement about the function on its
whole interval, and does not depend on where the data lie. Two other measures
are needed elsewhere, and both are supplied as arguments to the same generic.

The **empirical measure** of a set of points replaces the integral by the
average over them,

$$
\left(\widehat{G}_d\right)_{rs}
\;=\; \frac{1}{n}\sum_{i=1}^{n} B_r^{(d)}(x_i)\, B_s^{(d)}(x_i)
\;=\; \frac{1}{n} \left(B^{(d)\top} B^{(d)}\right)_{rs} ,
$$ {#eq-gram-empirical}

which is the matrix a design matrix actually produces, and the one that
appears wherever a construction is meant to depend on the design. It is what
@sec-basis-transforms diagonalises. A **weighted** Lebesgue measure, given by
a density $w$, integrates $B_r^{(d)} B_s^{(d)} w$ instead, and is what appears
when the penalty is meant to follow the distribution of the covariate rather
than the interval.

Both are handled in the body of the generic, before dispatch, so a method
never sees them and never has to implement them. A family always returns the
plain Lebesgue matrix, and the two alternatives are computed the same way for
every basis in the package and for every basis outside it.

```{r}
#| label: gram-measures
set.seed(1)
xs <- rbeta(4000, 2, 2)

round(basis_gram(b, at = xs)[1:3, 1:3], 4)
round(basis_gram(b, weight = function(v) dbeta(v, 2, 2))[1:3, 1:3], 4)
```

```{r}
#| label: consistency-05c
#| include: false
assert_gram_ok()
```
