## Linear transformations of a basis {#sec-basis-transforms}

```{r}
#| label: setup-05d
#| include: false
source("R/_setup.R")
source("R/basis-certificates.R")
```

### One operation, three uses {#sec-transform-one}

Making a basis orthonormal, restricting it so that its coefficients satisfy a
linear constraint, and rebuilding it so that it diagonalises an inner product
are all instances of the same operation. In each case the new functions are
fixed linear combinations of the old ones,

$$
\widetilde{B}(x)^\top \;=\; B(x)^\top T ,
\qquad T \in \mathbb{R}^{K \times \widetilde{K}},
$$ {#eq-transform}

and the three differ only in how $T$ is computed. The matrix may have fewer
columns than rows, which is how a constraint reduces the dimension of the
basis, and the coefficients of the new basis are related to the old ones by
$\beta = T\gamma$.

Every quantity of the previous sections transforms with @eq-transform and
nothing else has to be recomputed. Differentiation and integration are linear
and $T$ does not depend on $x$, so

$$
\widetilde{B}^{(d)}(x)^\top = B^{(d)}(x)^\top T ,
\qquad
\left(\textstyle\int \widetilde{B}\right)(x)^\top
= \left(\textstyle\int B\right)(x)^\top T ,
$$

and the anchor @eq-basis-anchor survives, since a linear combination of
columns that are all zero at $a$ is zero at $a$. The Gram matrix transforms by
**congruence**,

$$
\widetilde{G}_d \;=\; T^\top G_d\, T ,
$$ {#eq-transform-congruence}

which follows from @eq-gram by substituting @eq-transform under the integral.
A parent whose inner products are exact therefore passes that exactness on:
the transformed basis is not integrated again, and no quadrature enters at any
point.

Transforms compose by multiplication rather than by nesting. Applying
@eq-transform twice gives $B^\top T_1 T_2$, so a chain of them is stored as
the single product and costs one matrix multiplication per evaluation however
long the chain is.

### Orthonormalisation {#sec-transform-orthonorm}

An orthonormal basis is one whose Gram matrix is the identity, and
@eq-transform-congruence turns that requirement into a matrix factorisation.
Writing $G_d = R^\top R$ for the Cholesky factorisation of the Gram matrix,
the choice $T = R^{-1}$ gives

$$
\widetilde{G}_d \;=\; R^{-\top} R^\top R\, R^{-1} \;=\; I .
$$ {#eq-orthonorm}

The transform is read off the Gram matrix rather than estimated on a grid of
points, so there is no number of points to choose and no scale factor to
correct, and the result is as exact as $G_d$ is. The factorisation exists
whenever $G_d$ is positive definite, which at order zero holds for any basis
whose functions are linearly independent, and fails at higher orders, where
@sec-gram-definition showed the matrix to be singular.

```{r}
#| label: orthonorm
b <- bspline_basis(dimension = 6, degree = 3)
o <- orthonorm_basis(b)
round(basis_gram(o), 12)[1:4, 1:4]
```

### Linear constraints {#sec-transform-constrain}

A model term is often required to satisfy a linear constraint on its
coefficients, $C\beta = 0$ with $C$ of rank $\mathrm{rank}(C)$. The usual case
is a sum-to-zero condition, which removes the constant function from the span
of a term so that it is identified against the intercept of the model;
@wood2017 discusses the choices available and their consequences for the
fitted component.

Writing $\beta = T\gamma$ turns the constraint into $C T \gamma = 0$ for every
$\gamma$, so $T$ must have the null space of $C$ as its column space. The
package takes the last $K - \mathrm{rank}(C)$ right singular vectors of $C$,
which form an orthonormal basis of that null space, so the constrained basis
spans precisely the admissible functions and no reparameterisation of $C$
changes the space it produces. The dimension drops by the rank of the
constraint, not by the number of rows written down.

```{r}
#| label: constrain
grid <- seq(0, 1, length.out = 200)
cs <- constrain_basis(b, colSums(basis_eval(b, grid)))

c(before = b@dimension, after = cs@dimension)
max(abs(colSums(basis_eval(cs, grid))))
```

The mechanics are what the package supplies. Which constraint a term should
carry needs to know what the term means, and belongs to the layer that does.

### The Demmler-Reinsch basis {#sec-transform-dr}

The third transform makes the empirical inner product and the penalty diagonal
at the same time. Its purpose is the separation of a linear from a nonlinear
effect: an additive component $f_j$ is written as the sum of its projection on
$\mathrm{span}\{1, x_j\}$ and a remainder, and the remainder is expanded in a
basis that is empirically orthogonal to those two functions. Without that
orthogonality the sum of the two parts is still estimated correctly, but the
split between them is not, so a statement about the size of the nonlinear
effect depends on the basis rather than on the data. @bach2024 makes the
decomposition precise and proves the consistency of the two parts separately.

The construction has three steps, following @bach2024. The first restricts the
basis to the null space of

$$
C \;=\; (\mathbf{1}, x)^\top B \;\in\; \mathbb{R}^{2 \times K},
$$

by the transform of @sec-transform-constrain, giving $V_0$ of size
$K \times (K-2)$. This is where the empirical orthogonality comes from, since
$CV_0 = 0$ says exactly that the columns of $BV_0$ are orthogonal to
$\mathbf{1}$ and to $x$ at the observed points. The second diagonalises the
pencil formed by the empirical Gram matrix @eq-gram-empirical and the penalty
$P$ on that subspace: with

$$
\widetilde{G} = V_0^\top \left(B^\top B / n\right) V_0 ,
\qquad
\widetilde{P} = V_0^\top P\, V_0 ,
$$

the generalised eigenproblem
$\widetilde{G} \alpha_k = \lambda_k \widetilde{P} \alpha_k$ is solved by
factorising $\widetilde{P} = R^\top R$ and taking the ordinary
eigendecomposition of $R^{-\top} \widetilde{G} R^{-1}$, whose eigenvectors
$e_k$ give $\alpha_k = R^{-1} e_k$. The third step sets $T = V_0 A$, the
columns of $A$ being the $\alpha_k$, and by default rescales so that the
eigenvalues sum to one.

The resulting design matrix $Z = BT$ satisfies the three properties the
construction was asked for:

$$
Z^\top Z \ \text{is diagonal}, \qquad
T^\top P\, T \propto I, \qquad
(\mathbf{1}, x)^\top Z = 0 .
$$ {#eq-dr-properties}

The consequence is that the penalised normal equations in these coordinates
are diagonal, so each coefficient is shrunk on its own, by a factor determined
by its eigenvalue and the smoothing parameter, and the eigen-decomposition of
the smoother is available without forming any matrix of the size of the
sample. The eigenvalues $\lambda_k$ measure how much empirical variance a
direction carries per unit of penalty, and are the Demmler-Reinsch values of
@demmler1975.

```{r}
#| label: dr-basis
set.seed(1)
xd <- sort(runif(300))
d <- dr_basis(b, xd)

z <- basis_eval(d, xd)
c(
  off_diagonal = max(abs(crossprod(z)[upper.tri(crossprod(z))])),
  orthogonal_to_1_and_x = max(abs(crossprod(cbind(1, xd), z)))
)
```

Two properties of this particular route through the algebra matter in
practice. It factorises a $2 \times K$ matrix and a $(K-2) \times (K-2)$
matrix, never anything whose size depends on $n$. And the matrix inverted is
the penalty rather than the design, so a rank-deficient $B$ passes through
unharmed, which is what equally spaced knots produce whenever the data leave a
knot span empty.

```{r}
#| label: dr-rank-deficient
xg <- sort(c(seq(0, 0.25, length.out = 120), seq(0.78, 1, length.out = 120)))
wide <- bspline_basis(dimension = 14)

c(dimension = wide@dimension, rank = qr(basis_eval(wide, xg))$rank)

zg <- basis_eval(dr_basis(wide, xg), xg)
c(
  off_diagonal = max(abs(crossprod(zg)[upper.tri(crossprod(zg))])),
  orthogonal_to_1_and_x = max(abs(crossprod(cbind(1, xg), zg)))
)
```

```{r}
#| label: consistency-05d
#| include: false
assert_transforms_ok()
```
