## In practice {#sec-parameter-practice}

```{r}
#| label: setup-05p
#| include: false
source("R/_setup.R")
```

### The families that ship {#sec-parameter-census}

The census is generated by constructing each family and reading its
dimension and free labels off the object:

```{r}
#| label: pp-census
library(parameters7)
fams <- list(
  `log_cholesky(4)`          = log_cholesky(4),
  `matrix_log(4)`            = matrix_log(4),
  `diagonal_matrix(4)`       = diagonal_matrix(4),
  `scaled_matrix(P)`         = scaled_matrix(crossprod(diff(diag(6), differences = 2))),
  `correlation_matrix(4)`    = correlation_matrix(4),
  `compound_symmetry(6)`     = compound_symmetry(6),
  `ar1(6)`                   = ar1(6),
  `autoregressive(6, 2)`     = autoregressive(6, order = 2),
  `simplex(4)`               = simplex(4),
  `transition_matrix(3)`     = transition_matrix(3)
)
knitr::kable(data.frame(
  constructor = names(fams),
  d = sapply(fams, function(s) s@n_free),
  `free labels` = sapply(fams, function(s)
    paste(head(s@free_names, 3), collapse = ", ")),
  check.names = FALSE
), row.names = FALSE, caption = "The parameter families, read off the objects.")
```

### The contract in five calls {#sec-parameter-calls}

A parameter answers the matrix, its derivatives, the log-determinant with
its derivatives, the solve, and the inverse map:

```{r}
#| label: pp-calls
s <- ar1(5)
eta <- c(log(2), atanh(0.6))

M <- param_value(s, eta)
round(M[1:3, 1:3], 4)
length(param_d1(s, eta))                   # one derivative matrix per free value
c(logdet = param_logdet(s, eta),
  reference = determinant(M)$modulus)
unlist(param_dlogdet(s, eta))
max(abs(param_solve(s, eta) %*% M - diag(5)))
max(abs(param_free(s, M) - eta))           # the inverse map, exact
```

A matrix outside the family is rejected by the inverse map rather than
projected onto it:

```{r}
#| label: pp-reject
#| error: true
param_free(s, diag(c(1, 2, 3, 4, 5)))
```

### Validation and the readable quantities {#sec-parameter-check-practice}

`check_parameter()` compares every closed form against an independent
route, and reports which orders came from a fallback; `param_readable()`
declares the quantities a reader of a fit examines, with the Jacobian the
delta method needs:

```{r}
#| label: pp-check
res <- check_parameter(autoregressive(5, order = 2), verbose = FALSE)
res[, c("check", "status")]

r <- param_readable(autoregressive(5, order = 2), c(log(2), atanh(0.6), atanh(-0.2)))
data.frame(quantity = names(r$value), value = round(unlist(r$value), 4),
           interval_scale = unlist(r$transform))
```
