## Constrained parameters {#sec-parameter-contract}

```{r}
#| label: setup-06a
#| include: false
source("R/_setup.R")
source("R/parameter-certificates.R")
```

### The map and its free vector {#sec-parameter-map}

Let $\mathcal{M}$ be a set of symmetric $p \times p$ matrices defined by a
constraint: positive definite matrices, positive definite matrices with a
prescribed pattern of zeros, correlation matrices, or positive semidefinite
matrices of a fixed null space. A **parameter** is a map

$$
M : \mathbb{R}^{d} \longrightarrow \mathcal{M}, \qquad \eta \longmapsto M(\eta),
$$ {#eq-parameter-map}

whose image is the whole of $\mathcal{M}$ and which is one-to-one, together
with the derivatives of that map and the quantities a likelihood asks of its
value. The vector $\eta$ is called **free** because it carries no constraint:
every point of $\mathbb{R}^{d}$ is admissible, and an optimiser may take any
step from any point without leaving the domain. This is the same bargain a
link function offers for a scalar, and for the same reason: the constraint is
absorbed into the map so that the estimation problem is unconstrained.

The number $d$ of free values is the dimension of $\mathcal{M}$ as a manifold,
not $p^2$ or $p(p+1)/2$ in general. A full positive definite matrix has
$d = p(p+1)/2$; a diagonal one has $d = p$; a matrix that is a fixed matrix
times an unknown scale has $d = 1$. A parameter therefore owns its dimension:
a map for $3 \times 3$ matrices and a map for $4 \times 4$ matrices are
different objects, with different $d$ and different names for the free values,
and both are needed before any data exist because they determine the shape of
the parameter vector.

Each free value carries a label, and the labels are fixed at construction. The
distribution of @sec-multivariate builds its parameter names from them, the
derivative components of the likelihood are keyed by them, and the printed
summary of a fit reports them, so they are part of the interface rather than a
convenience.

### Why this is not a link {#sec-parameter-not-link}

A link function of @sec-links is a scalar map $g$ applied to each parameter
separately. Applied to a vector of parameters, it produces a **diagonal**
Jacobian, which is what makes the chain rule of @sec-link-scale factorise: the
$k$th derivative in the $k$th direction involves only $h^{(j)}$ for that one
coordinate, and Faà di Bruno's formula reduces to a product of partial Bell
polynomials in one variable.

A matrix parametrisation has no such parameter. Changing one free value of a
log-Cholesky map changes an entire row and column of the matrix it produces,
so $\partial M / \partial \eta_k$ is a matrix rather than a number and the
Jacobian $\partial \mathrm{vech}(M) / \partial \eta$ is dense. The chain rule
still holds, but it is a matrix chain rule, and the quantities a consumer
needs from it are traces rather than products. Nothing that `linkfunctions7`
computes would be reusable, and a consumer that treated a matrix parametrisation
as a link would be wrong in a way that produces plausible numbers.

The two constructions therefore stay apart, and meet at the one place where
they agree, which is a diagonal matrix. There the Jacobian **is** diagonal, and
`parameters7` reuses the link objects exactly as they are rather than writing
its own positive maps; see @sec-parameter-diagonal.

### What a consumer asks {#sec-parameter-generics}

A likelihood that carries a matrix parameter asks four things of it, and a
parameter answers all four.

The **matrix** itself, $M(\eta)$, is the only compulsory one; everything else
has a numerical method available from it alone.

The **derivatives** $A_k = \partial M/\partial \eta_k$ and
$A_{kl} = \partial^2 M/\partial \eta_k \partial \eta_l$ are what the score and
the observed information of @sec-mvgaussian are built from. Only the
$d(d+1)/2$ distinct second derivatives are computed, keyed by the same
enumeration that names them. Recovering an index by splitting a name on its
separator is the obvious route and it is wrong, since a free value whose own
label contains the separator splits into the wrong number of pieces; the
package generates indices and names from one enumeration so that the two
cannot disagree.

The **log-determinant** $\log|M|$ and its first two derivatives are the
normalising constant of a gaussian and its derivatives. The identity behind
them is

$$
\frac{\partial \log|M|}{\partial \eta_k}
  = \mathrm{tr}\!\left(M^{-1} A_k\right),
$$ {#eq-parameter-dlogdet}

which is @magnus1999's, and which makes a closed-form log-determinant
derivative something other than an independent claim: it must agree with the
derivative of the matrix through @eq-parameter-dlogdet, and the package's
validator compares the two routes.

The **solve** $M^{-1}B$ is what a quadratic form needs. It is computed through
a factorisation rather than by forming an inverse, which is both cheaper and
better conditioned; a family that knows its own factor supplies it, and the
rest obtain one from the matrix.

```{r}
#| label: parameter-demo
s <- parameters7::log_cholesky(3)
s

eta <- c(0.1, -0.2, 0.3, 0.5, -0.4, 0.2)
round(parameters7::param_value(s, eta), 4)
```

The sign in front of the log-determinant is the consumer's arithmetic and not
the parameter's. A gaussian written in the covariance carries
$-\frac{1}{2}\log|\Sigma|$ and one written in the precision carries
$+\frac{1}{2}\log|\Omega|$, so the parameter returns $\log|M|$ for the matrix
it parametrises and lets the distribution attach the sign. The same holds for
the two derivatives.

### The inverse map {#sec-parameter-free}

A parameter also answers the reverse question: given a matrix in
$\mathcal{M}$, which free vector produces it? This is what allows a fit to be
started from a sample covariance, and what allows a marginal of
@sec-mv-marginal to be expressed in the free values of a smaller parameter.

The map is exact or refused, never obtained by optimisation. An
optimisation-based inverse would return a plausible $\eta$ for a matrix
**outside** $\mathcal{M}$, which is a wrong answer wearing the shape of a right
one, and a caller has no way to tell the two apart. A family whose map has no
closed-form inverse therefore refuses, and so does the base class on behalf of
any parameter that does not implement one.

```{r}
#| label: parameter-roundtrip
m <- parameters7::param_value(s, eta)
max(abs(parameters7::param_free(s, m) - eta))
```

### What a new parameter has to write {#sec-parameter-fallbacks}

Only @eq-parameter-map is compulsory. A parameter that implements nothing but
the matrix receives numerical methods for the derivatives, the log-determinant
and its derivatives, and the solve, registered on the base class exactly as
the numerical fallbacks of @sec-fallbacks are registered for a distribution.
It is therefore usable the moment it exists, and each closed form supplied
later replaces the corresponding numerical method through dispatch, with no
change to anything that consumes it.

The numerical derivatives obey the rule of @sec-fallbacks: one central
difference applied to the highest analytical order available, never a chain of
first differences. A parameter that supplies its own first derivative gets a
second derivative accurate to the square root of machine precision rather than
to its fourth root.

The validator `check_parameter()` puts a parameter through the same checks the
shipped families face. It compares the derivatives against finite differences,
the log-determinant against the determinant of the assembled matrix, its
derivative against @eq-parameter-dlogdet, and the solve against a direct
solution, and it reports which orders came from a closed form and which from
the fallback, for the reason `check_link()` does: an order that comes from a
fallback would otherwise be compared against a numerical differentiation of
the order below, which is the same arithmetic twice and agrees however wrong
the parameter is.

```{r}
#| label: parameter-check
parameters7::check_parameter(s, verbose = TRUE)
```

```{r}
#| label: consistency-06a
#| include: false
assert_parameter_ok()
```
