## The families {#sec-covstruct-families}

```{r}
#| label: setup-06b
#| include: false
source("R/_setup.R")
source("R/covstruct-certificates.R")
```

### The log-Cholesky map {#sec-covstruct-logchol}

The general positive definite case is parametrised through the Cholesky
factor. Every positive definite $M$ has a unique factorisation
$M = L L^\top$ with $L$ lower triangular and $L_{ii} > 0$, and the
correspondence is a bijection. Taking the logarithm of the diagonal removes
the only remaining constraint, so

$$
\eta = \bigl(\log L_{11}, \ldots, \log L_{pp},\;
             L_{21}, L_{31}, \ldots, L_{p,p-1}\bigr)
$$ {#eq-logchol-free}

ranges over the whole of $\mathbb{R}^{d}$ with $d = p(p+1)/2$, and
$M(\eta) = L L^\top$ is onto the positive definite cone. This is the
parametrisation of @pinheiro1996. The ordering above is fixed rather than
incidental: the diagonal comes first, then the below-diagonal entries column
by column, and the labels of the free values follow it.

The logarithm on the diagonal is intrinsic to the map and not a swappable
link. A different positive map would give a different parametrisation of the
same set, but it would not compose with anything, since the constraint being
removed is on the factor and not on an entry of $M$.

Differentiating is straightforward once the factor is written as a function of
the free vector. Let $L_k = \partial L/\partial \eta_k$, which is the matrix
with a single non-zero entry: $L_{ii}$ itself in a diagonal direction, because
the entry is $e^{\eta}$, and $1$ in a below-diagonal direction. Then

$$
A_k = L_k L^\top + L L_k^\top ,
\qquad
A_{kl} = L_k L_l^\top + L_l L_k^\top + \delta_{kl}^{\,\mathrm{diag}}
         \left( L_k L^\top + L L_k^\top \right),
$$ {#eq-logchol-derivs}

where the last term appears only when $k = l$ and the direction is a diagonal
one, since $\partial^2 L/\partial \eta_k^2 = L_k$ there and vanishes
elsewhere. The factor is linear in every below-diagonal free value, which is
why nothing else survives the second derivative.

The log-determinant is the reason this parametrisation is comfortable. Since
$|M| = |L|^2$ and $L$ is triangular,

$$
\log|M| = 2\sum_{i=1}^{p} \log L_{ii} = 2\sum_{i \in \mathrm{diag}} \eta_i ,
$$ {#eq-logchol-logdet}

which is **linear** in the free vector. Its gradient is the constant $2$ in
each diagonal direction and $0$ in every other, and its Hessian vanishes
identically. A term that would otherwise cost a trace per component per
observation costs nothing at all.

```{r}
#| label: logchol-logdet
s <- covstructs7::log_cholesky(3)
covstructs7::struct_dlogdet(s, c(0.1, -0.2, 0.3, 0.5, -0.4, 0.2))
covstructs7::struct_dlogdet(s, c(-2, 3, 0.7, -1.1, 0.4, 2.5))
```

### Diagonal matrices, and where the two packages meet {#sec-covstruct-diagonal}

A diagonal matrix with positive entries is the case in which a matrix
parametrisation and a scalar link agree. The Jacobian of

$$
M(\eta) = \mathrm{diag}\bigl(h(\eta_1), \ldots, h(\eta_p)\bigr)
$$ {#eq-diag-struct}

is diagonal, which is exactly the contract a link satisfies, so `covstructs7`
takes $h$ from `linkfunctions7` instead of writing a positive map of its own.
Any link whose range is the positive half-line serves, and its exact
derivatives to fourth order come with it. The derivatives of the structure are
then $A_k = h'(\eta_k) e_k e_k^\top$ and $A_{kk} = h''(\eta_k) e_k e_k^\top$,
with every mixed second derivative zero, and the log-determinant is
$\sum_i \log h(\eta_i)$.

```{r}
#| label: diag-struct
covstructs7::struct_matrix(covstructs7::diag_struct(3), c(0, 0.5, -0.5))
covstructs7::struct_matrix(
  covstructs7::diag_struct(2, link = linkfunctions7::sqrt_link()), c(1, 2)
)
```

A single free value shared by every entry gives $\sigma^2 I$, which is the
independent case a variance component or a ridge starts from. It is the same
construction with $d = 1$.

### A fixed matrix carried by one scale {#sec-covstruct-scaled}

The third family is different in kind. Here the matrix is

$$
M(\eta) = \lambda P, \qquad \lambda = h(\eta),
$$ {#eq-scaled-struct}

with $P$ a **known** symmetric positive semidefinite matrix and $\lambda$ the
only unknown. This is the commonest penalty in semiparametric regression: $P$
may be the Gram matrix of a basis derivative from @sec-basis-gram, a
difference penalty $\Delta^\top \Delta$, a neighbourhood matrix, or the
identity, in which case the structure is a ridge.

Nothing here needs deriving. With the default log link, $\lambda = e^{\eta}$
and every quantity is a constant times a function of the scale:

$$
\partial_\eta M = M, \qquad \partial^2_\eta M = M,
$$

$$
\log|M|_{+} = r\,\eta + \log|P|_{+}, \qquad
\partial_\eta \log|M|_{+} = r, \qquad
\partial^2_\eta \log|M|_{+} = 0,
$$ {#eq-scaled-logdet}

with $r$ the rank of $P$ and $|\cdot|_{+}$ the pseudo-determinant of
@sec-covstruct-pseudodet. The constant $\log|P|_{+}$ is computed once at
construction, since $P$ does not change.

```{r}
#| label: scaled-struct
# a ridge: the identity, scaled
r <- covstructs7::scaled_struct(diag(4))
c(rank = r@rank, logdet_at_zero = covstructs7::struct_logdet(r, 0))

# a second-difference penalty on six coefficients
dm <- diff(diag(6), differences = 2)
pen <- covstructs7::scaled_struct(crossprod(dm))
c(dimension = pen@dimension, rank = pen@rank)
```

### Which side of a model a family parametrises {#sec-covstruct-role}

The name of a family does not say whether the matrix it produces is a
covariance or a precision, and the two are different models. Imposing a
structure on $\Sigma$ and imposing the same structure on $\Omega = \Sigma^{-1}$
give the same family only when the structure is closed under inversion.
Diagonal matrices are, and so is compound symmetry, whose inverse is again
compound symmetry with different constants. An AR(1) correlation matrix is not:
its inverse is tridiagonal, which is a different pattern of zeros, so an AR(1)
covariance and an AR(1) precision are genuinely different assumptions about
the same data.

A structure therefore carries a label recording which side it is meant for.
Nothing computed depends on it and no method reads it; it exists so that the
object says what it is, and so that a consumer that cares can ask. The
distribution of @sec-mvgaussian takes the structure on either side and adjusts
its own derivatives accordingly, which is the only place the difference has
arithmetic consequences.

```{r}
#| label: consistency-06b
#| include: false
assert_covstruct_families_ok()
```
