## The families {#sec-parameter-families}

```{r}
#| label: setup-06b
#| include: false
source("R/_setup.R")
source("R/parameter-certificates.R")
```

### The log-Cholesky map {#sec-parameter-logchol}

The general positive definite case is parametrised through the Cholesky
factor. Every positive definite $M$ has a unique factorisation
$M = L L^\top$ with $L$ lower triangular and $L_{ii} > 0$, and the
correspondence is a bijection. Taking the logarithm of the diagonal removes
the only remaining constraint, so

$$
\eta = \bigl(\log L_{11}, \ldots, \log L_{pp},\;
             L_{21}, L_{31}, \ldots, L_{p,p-1}\bigr)
$$ {#eq-logchol-free}

ranges over the whole of $\mathbb{R}^{d}$ with $d = p(p+1)/2$, and
$M(\eta) = L L^\top$ is onto the positive definite cone. This is the
parametrisation of @pinheiro1996. The ordering above is fixed rather than
incidental: the diagonal comes first, then the below-diagonal entries column
by column, and the labels of the free values follow it.

The logarithm on the diagonal is intrinsic to the map and not a swappable
link. A different positive map would give a different parametrisation of the
same set, but it would not compose with anything, since the constraint being
removed is on the factor and not on an entry of $M$.

Differentiating is straightforward once the factor is written as a function of
the free vector. Let $L_k = \partial L/\partial \eta_k$, which is the matrix
with a single non-zero entry: $L_{ii}$ itself in a diagonal direction, because
the entry is $e^{\eta}$, and $1$ in a below-diagonal direction. Then

$$
A_k = L_k L^\top + L L_k^\top ,
\qquad
A_{kl} = L_k L_l^\top + L_l L_k^\top + \delta_{kl}^{\,\mathrm{diag}}
         \left( L_k L^\top + L L_k^\top \right),
$$ {#eq-logchol-derivs}

where the last term appears only when $k = l$ and the direction is a diagonal
one, since $\partial^2 L/\partial \eta_k^2 = L_k$ there and vanishes
elsewhere. The factor is linear in every below-diagonal free value, which is
why nothing else survives the second derivative.

The log-determinant is the reason this parametrisation is comfortable. Since
$|M| = |L|^2$ and $L$ is triangular,

$$
\log|M| = 2\sum_{i=1}^{p} \log L_{ii} = 2\sum_{i \in \mathrm{diag}} \eta_i ,
$$ {#eq-logchol-logdet}

which is **linear** in the free vector. Its gradient is the constant $2$ in
each diagonal direction and $0$ in every other, and its Hessian vanishes
identically. A term that would otherwise cost a trace per component per
observation costs nothing at all.

```{r}
#| label: logchol-logdet
s <- parameters7::log_cholesky(3)
parameters7::param_dlogdet(s, c(0.1, -0.2, 0.3, 0.5, -0.4, 0.2))
parameters7::param_dlogdet(s, c(-2, 3, 0.7, -1.1, 0.4, 2.5))
```

### Diagonal matrices, and where the two packages meet {#sec-parameter-diagonal}

A diagonal matrix with positive entries is the case in which a matrix
parametrisation and a scalar link agree. The Jacobian of

$$
M(\eta) = \mathrm{diag}\bigl(h(\eta_1), \ldots, h(\eta_p)\bigr)
$$ {#eq-diag-struct}

is diagonal, which is exactly the contract a link satisfies, so `parameters7`
takes $h$ from `linkfunctions7` instead of writing a positive map of its own.
Any link whose range is the positive half-line serves, and its exact
derivatives to fourth order come with it. The derivatives of the parameter are
then $A_k = h'(\eta_k) e_k e_k^\top$ and $A_{kk} = h''(\eta_k) e_k e_k^\top$,
with every mixed second derivative zero, and the log-determinant is
$\sum_i \log h(\eta_i)$.

```{r}
#| label: diag-struct
parameters7::param_value(parameters7::diagonal_matrix(3), c(0, 0.5, -0.5))
parameters7::param_value(
  parameters7::diagonal_matrix(2, link = linkfunctions7::sqrt_link()), c(1, 2)
)
```

A single free value shared by every entry gives $\sigma^2 I$, which is the
independent case a variance component or a ridge starts from. It is the same
construction with $d = 1$.

### A fixed matrix carried by one scale {#sec-parameter-scaled}

The third family is different in kind. Here the matrix is

$$
M(\eta) = \lambda P, \qquad \lambda = h(\eta),
$$ {#eq-scaled-struct}

with $P$ a **known** symmetric positive semidefinite matrix and $\lambda$ the
only unknown. This is the commonest penalty in semiparametric regression: $P$
may be the Gram matrix of a basis derivative from @sec-basis-gram, a
difference penalty $\Delta^\top \Delta$, a neighbourhood matrix, or the
identity, in which case the parameter is a ridge.

Nothing here needs deriving. With the default log link, $\lambda = e^{\eta}$
and every quantity is a constant times a function of the scale:

$$
\partial_\eta M = M, \qquad \partial^2_\eta M = M,
$$

$$
\log|M|_{+} = r\,\eta + \log|P|_{+}, \qquad
\partial_\eta \log|M|_{+} = r, \qquad
\partial^2_\eta \log|M|_{+} = 0,
$$ {#eq-scaled-logdet}

with $r$ the rank of $P$ and $|\cdot|_{+}$ the pseudo-determinant of
@sec-parameter-pseudodet. The constant $\log|P|_{+}$ is computed once at
construction, since $P$ does not change.

```{r}
#| label: scaled-struct
# a ridge: the identity, scaled
r <- parameters7::scaled_matrix(diag(4))
c(rank = r@rank, logdet_at_zero = parameters7::param_logdet(r, 0))

# a second-difference penalty on six coefficients
dm <- diff(diag(6), differences = 2)
pen <- parameters7::scaled_matrix(crossprod(dm))
c(dimension = pen@dimension, rank = pen@rank)
```

### Correlation matrices {#sec-parameter-correlation}

A correlation matrix is a covariance whose diagonal is fixed at one, and that
fixing is what makes it awkward: the constraint is not a bound on each entry
but a relation among them, so a parametrisation that produced the entries
directly would have to repair the diagonal afterwards. The spherical
parametrisation avoids the repair by building the Cholesky factor rather than
the matrix. Row $i$ of $L$ is a point on the unit sphere of $\mathbb{R}^{i}$
written in angular coordinates,
$$L_{ij} = \cos\theta_{ij} \prod_{k<j} \sin\theta_{ik} \quad (j < i),
  \qquad L_{ii} = \prod_{k<i} \sin\theta_{ik},$$
with $\theta_{ik} \in (0, \pi)$. The squares of the entries of a row telescope
to one, so $R = LL^{\top}$ has a unit diagonal whatever the angles are, and it
is positive definite because $L$ is triangular with a positive diagonal. The
construction is that of @rapisarda2007, and the angles reach the real line
through the bounded link of @sec-links, so the free values are unconstrained
and no combination of them leaves the set.

The derivatives are the Leibniz rule of @sec-parameter-logchol applied to a
different factor. Each entry of $L$ is a product of sines and cosines of
angles that each depend on one free value, so differentiating it replaces one
factor by its derivative and leaves the rest alone. The rows of $L$ are
independent, which makes a derivative of the factor across two rows vanish;
the same is not true of $R$, whose entry $(i, j)$ is the inner product of rows
$i$ and $j$, and a second derivative across those rows is supported on exactly
those two entries.

The log-determinant is where the parametrisation pays again. The factor is
triangular, so $\lvert R \rvert = \prod_i L_{ii}^{2}$ and
$$\log\lvert R \rvert = 2 \sum_{i,k} \log \sin \theta_{ik},$$
a sum with one term per free value. It is therefore separable: every mixed
derivative is exactly zero and the pure ones are a logarithm composed with a
sine.

```{r}
#| label: correlation
s <- correlation_matrix(4)
eta <- c(0.4, -0.2, 0.6, 0.1, -0.5, 0.3)
r <- param_value(s, eta)
round(r, 4)
diag(r)

# the log-determinant needs no factorisation
c(closed = param_logdet(s, eta),
  spectral = sum(log(eigen(r, symmetric = TRUE, only.values = TRUE)$values)))
```

### Structure, and the two free values it leaves {#sec-parameter-economical}

An unstructured covariance over $p$ occasions has $p(p+1)/2$ free values,
which at $p = 20$ is 210 and beyond what a handful of subjects identifies. The
two classical structured families collapse that count to two whatever $p$ is,
which is the reason they are used at all.

Compound symmetry is the exchangeable covariance
$$M = \sigma^{2}\{(1-\rho)I + \rho J\},$$
every variance equal and every correlation equal. Its eigenvalues are
$\sigma^{2}\{1 + (p-1)\rho\}$ once and $\sigma^{2}(1-\rho)$ with multiplicity
$p-1$, so it is positive definite exactly when
$$-\frac{1}{p-1} < \rho < 1,$$
a bound that depends on the dimension and is the reason the correlation is
carried by a link onto that interval rather than onto $(-1, 1)$: a chart that
allowed $\rho = -0.9$ at $p = 5$ would hand a consumer an indefinite matrix at
an ordinary free value. The eigenvalues also give the log-determinant,
$$\log\lvert M \rvert = p \log \sigma^{2} + \log\{1 + (p-1)\rho\}
  + (p-1)\log(1-\rho),$$
a sum of a function of one free value and a function of the other, so its
mixed derivatives vanish identically. The inverse is compound symmetric again,
by the Sherman-Morrison identity, so the family is closed under the choice of
side: an exchangeable covariance has an exchangeable precision.

The AR(1) family is the covariance of a first-order autoregression,
$M_{ij} = \sigma^{2}\rho^{\lvert i-j \rvert}$, and it is definite for every
$\lvert \rho \rvert < 1$, so its correlation needs no dimension-dependent
bound. Its determinant is $\sigma^{2p}(1-\rho^{2})^{p-1}$ and its inverse is
**tridiagonal**, which is the property the family is used for rather than an
incidental fact: an autoregression of order one is a Markov process, and a
Markov process has no partial correlation beyond the lag, so the precision has
no entries beyond the first off-diagonal.

That inverse is also what makes an AR(1) covariance and an AR(1) precision
different models. The inverse of a compound-symmetric matrix is compound
symmetric, so imposing the structure on either side gives the same family; the
inverse of an AR(1) is tridiagonal but not AR(1), so the two impositions are
genuinely different assumptions about the same data. This is what the `role`
label of @sec-parameter-role records.

The two families differ in one respect that matters for the derivatives. The
compound-symmetric pattern $I + \rho(J - I)$ is *linear* in the correlation, so
every derivative in it is the matching derivative of the correlation times a
constant matrix. The AR(1) pattern is not: an entry is $\rho^{m}$ for the lag
$m$, and its derivatives are a power composed with the link, which is the
chain rule of Faà di Bruno taken to fourth order.

```{r}
#| label: economical
c(unstructured = log_cholesky(20)@n_free,
  compound_symmetry = compound_symmetry(20)@n_free,
  ar1 = ar1(20)@n_free)

a <- ar1(5)
eta <- c(log(2), atanh(0.6))
round(param_solve(a, eta), 4)
```

### Autoregressions of any order {#sec-parameter-arp}

The AR(1) family generalises, and the generalisation is instructive because
the obvious parametrisation fails. A stationary autoregression of order $q$
has coefficients $\phi$ whose polynomial $1 - \phi_1 z - \cdots - \phi_q z^q$
has all its roots outside the unit circle, and that region is not a box:
already at $q = 2$ it is the triangle $\lvert\phi_2\rvert < 1$,
$\phi_2 \pm \phi_1 < 1$. No collection of scalar links onto intervals covers
a set whose constraints tie the coordinates together, so the chart cannot be
built one coefficient at a time.

The partial autocorrelations can. Each $r_k$ lies in $(-1,1)$ independently
of the others, and the Levinson-Durbin recursion carries them onto the
stationary coefficients bijectively, which is the transformation of
@barndorffnielsen1973. Writing $\phi^{(k)}$ for the coefficients of the
best linear predictor of order $k$,
$$\phi^{(k)}_k = r_k, \qquad
  \phi^{(k)}_j = \phi^{(k-1)}_j - r_k \phi^{(k-1)}_{k-j},$$
and the Yule-Walker equations at order $k$ give the autocorrelations
$$\rho_k = \sum_{j<k} \phi^{(k)}_j \rho_{k-j} + r_k,$$
after which $\rho_h = \sum_j \phi_j \rho_{h-j}$ at every lag beyond the
order. Every free vector is then stationary by construction, and each
partial autocorrelation takes the same link the AR(1) correlation takes.

The recursion is built from sums and products alone, so the map from the free
values to the matrix is a polynomial one. Its derivatives are obtained by
carrying **jets** through it -- each intermediate quantity held together with
all its partial derivatives to fourth order, multiplied by the same Leibniz
rule of @sec-parameter-logchol in its scalar form -- rather than by expanding
the recursion, which would be pages of algebra checkable only against the
finite differences this construction exists to avoid.

Two quantities stay closed. The innovation variances of the recursion fall by
$1 - r_k^2$ at each of the first $q$ steps and are constant after, so
$$\log\lvert M \rvert = p\log\gamma_0 + \sum_{k=1}^{q}(p-k)\log(1-r_k^2),$$
one term per free value and therefore separable. And the precision is
**banded of width $q$**: an autoregression of order $q$ is Markov of order
$q$, so no partial correlation survives beyond the lag, and the inverse is
assembled from the prediction form rather than factorised.

```{r}
#| label: arp
s <- autoregressive(7, order = 2)
eta <- c(log(2), atanh(0.7), atanh(-0.3))

# the implied coefficients are stationary whatever the free values are
round(param_solve(s, eta), 4)
```

`ar1()` remains as the order-one case with its formulas written out, and the
two agree to machine precision -- which is a check on both, two
implementations of one object being the only comparison that needs no
tolerance argument.

### The simplex and the transition matrix {#sec-parameter-simplex}

A probability vector $\pi$ on the open simplex satisfies $\pi_a > 0$ and
$\sum_a \pi_a = 1$, so a vector of $K$ categories has $K - 1$ free values.
The package carries it by the additive log-ratio map with the last category
as reference,
$$\pi_a = \frac{e^{\eta_a}}{1 + \sum_b e^{\eta_b}}, \qquad
  \pi_K = \frac{1}{1 + \sum_b e^{\eta_b}},$$
whose inverse is $\eta_a = \log(\pi_a / \pi_K)$, exact. The derivatives of
this map close over the value itself: writing
$Z = 1 + \sum_b e^{\eta_b}$, each $\pi_a$ with $a < K$ is a first derivative
of $\log Z$, so the derivative arrays of $\pi$ are the higher derivatives of
$\log Z$ -- the cumulants of a categorical indicator -- and every order
follows from the single rule
$$\partial_b\, \pi_a = \pi_a(\delta_{ab} - \pi_b)$$
applied repeatedly with the product rule. All four orders are closed form,
and one identity comes free: differentiating $\sum_a \pi_a = 1$ shows that
every derivative component sums to zero over the value index, at every order.

```{r}
#| label: simplex-orders
s <- simplex(4)
eta <- c(0.5, -0.2, 0.8)
round(param_value(s, eta), 4)

# the free identity: each component of each order sums to zero
range(vapply(param_d3(s, eta), sum, numeric(1)))
```

A transition matrix of a Markov chain on $K$ states is a $K \times K$ matrix
whose rows lie on the simplex, and `transition_matrix()` treats it as exactly
that: $K$ independent additive log-ratio charts, one per row, $K(K-1)$ free
values in all. The independence carries to the derivatives, which are block
diagonal by row -- a component pairing free values of two different rows is
exactly zero, not small -- so the cost of an order grows with the row, not
with the matrix.

### The matrix logarithm {#sec-parameter-matrixlog}

The log-Cholesky map is not the only chart on the positive definite cone.
`matrix_log()` writes $M = \exp(S)$ with $S$ symmetric and genuinely free:
the free values fill the triangle of $S$ directly, and no entry is
transformed. Two quantities are then exact by construction. The
log-determinant is the trace,
$$\log\lvert M \rvert = \operatorname{tr}(S),$$
linear in the diagonal free values with every higher derivative zero, and the
inverse is $M^{-1} = \exp(-S)$, through the same eigendecomposition that
evaluates the map. This is the chart to reach for when the covariance and
the precision are wanted at once; log-Cholesky remains the cheaper one for
derivatives, which there are sparse products.

The derivatives of $\exp(S)$ in the free values are its Fréchet derivatives,
and the package computes them by the representation of @daleckii1965: with
$S = Q \Lambda Q^\top$ and the directions rotated by $Q$, the $k$-th
derivative contracts chains of directions against divided differences of
$e^x$ of order $k + 1$ at the eigenvalues, summed over the orderings of the
directions. The divided differences themselves are computed by the theorem of
@opitz1964 -- the exponential of a small upper bidiagonal matrix, read off
its corner entry -- which remains exact when eigenvalues repeat or nearly
repeat, where the recursive quotient formula cancels catastrophically.

```{r}
#| label: matrixlog
m <- matrix_log(3)
eta <- c(0.2, -0.3, 0.4, 0.5, -0.1, 0.3)

# the log-determinant is the trace of S, read off the free vector
c(param_logdet(m, eta), sum(eta[1:3]))

# and the inverse costs no factorisation
max(abs(param_solve(m, eta) - solve(param_value(m, eta))))
```

### Third and fourth derivatives {#sec-parameter-higher}

Every family above carries its derivatives to fourth order in closed form:
log-Cholesky by the Leibniz rule on $M = LL^\top$, the diagonal and scaled
families through the link derivatives that `linkfunctions7` already supplies
at fourth order, the simplex by its cumulant recursion, and the matrix
logarithm by the divided-difference contraction. A parameter defined outside
the package is served by numerical fallbacks that apply one product stencil
directly to the map for each component -- never a stencil on a lower-order
numerical derivative -- so the accuracy degrades honestly with the order and
the no-nesting rule of the toolkit holds structurally.

The higher orders exist for a consumer: the multivariate gaussian of
@sec-mvgaussian builds its own third and fourth derivatives from these, one
Leibniz expansion away, and a mixed-effects model that profiles or corrects
its likelihood asks for exactly that.

### Which side of a model a family parametrises {#sec-parameter-role}

The name of a family does not say whether the matrix it produces is a
covariance or a precision, and the two are different models. Imposing a
parameter on $\Sigma$ and imposing the same parameter on $\Omega = \Sigma^{-1}$
give the same family only when the parameter is closed under inversion.
Diagonal matrices are, and so is compound symmetry, whose inverse is again
compound symmetry with different constants. An AR(1) correlation matrix is not:
its inverse is tridiagonal, which is a different pattern of zeros, so an AR(1)
covariance and an AR(1) precision are genuinely different assumptions about
the same data.

A parameter therefore carries a label recording which side it is meant for.
Nothing computed depends on it and no method reads it; it exists so that the
object says what it is, and so that a consumer that cares can ask. The
distribution of @sec-mvgaussian takes the parameter on either side and adjusts
its own derivatives accordingly, which is the only place the difference has
arithmetic consequences.

```{r}
#| label: consistency-06b
#| include: false
assert_parameter_families_ok()
```
