## The families {#sec-parameter-families}

```{r}
#| label: setup-06b
#| include: false
source("R/_setup.R")
source("R/parameter-certificates.R")
```

### The log-Cholesky map {#sec-parameter-logchol}

The general positive definite case is parametrised through the Cholesky
factor. Every positive definite $M$ has a unique factorisation
$M = L L^\top$ with $L$ lower triangular and $L_{ii} > 0$, and the
correspondence is a bijection. Taking the logarithm of the diagonal removes
the only remaining constraint, so

$$
\eta = \bigl(\log L_{11}, \ldots, \log L_{pp},\;
             L_{21}, L_{31}, \ldots, L_{p,p-1}\bigr)
$$ {#eq-logchol-free}

ranges over the whole of $\mathbb{R}^{d}$ with $d = p(p+1)/2$, and
$M(\eta) = L L^\top$ is onto the positive definite cone. This is the
parametrisation of @pinheiro1996. The ordering above is fixed rather than
incidental: the diagonal comes first, then the below-diagonal entries column
by column, and the labels of the free values follow it.

The logarithm on the diagonal is intrinsic to the map and not a swappable
link. A different positive map would give a different parametrisation of the
same set, but it would not compose with anything, since the constraint being
removed is on the factor and not on an entry of $M$.

Differentiating is straightforward once the factor is written as a function of
the free vector. Let $L_k = \partial L/\partial \eta_k$, which is the matrix
with a single non-zero entry: $L_{ii}$ itself in a diagonal direction, because
the entry is $e^{\eta}$, and $1$ in a below-diagonal direction. Then

$$
A_k = L_k L^\top + L L_k^\top ,
\qquad
A_{kl} = L_k L_l^\top + L_l L_k^\top + \delta_{kl}^{\,\mathrm{diag}}
         \left( L_k L^\top + L L_k^\top \right),
$$ {#eq-logchol-derivs}

where the last term appears only when $k = l$ and the direction is a diagonal
one, since $\partial^2 L/\partial \eta_k^2 = L_k$ there and vanishes
elsewhere. The factor is linear in every below-diagonal free value, which is
why nothing else survives the second derivative.

The log-determinant is the reason this parametrisation is comfortable. Since
$|M| = |L|^2$ and $L$ is triangular,

$$
\log|M| = 2\sum_{i=1}^{p} \log L_{ii} = 2\sum_{i \in \mathrm{diag}} \eta_i ,
$$ {#eq-logchol-logdet}

which is **linear** in the free vector. Its gradient is the constant $2$ in
each diagonal direction and $0$ in every other, and its Hessian vanishes
identically. A term that would otherwise cost a trace per component per
observation costs nothing at all.

```{r}
#| label: logchol-logdet
s <- parameters7::log_cholesky(3)
parameters7::param_dlogdet(s, c(0.1, -0.2, 0.3, 0.5, -0.4, 0.2))
parameters7::param_dlogdet(s, c(-2, 3, 0.7, -1.1, 0.4, 2.5))
```

### Diagonal matrices, and where the two packages meet {#sec-parameter-diagonal}

A diagonal matrix with positive entries is the case in which a matrix
parametrisation and a scalar link agree. The Jacobian of

$$
M(\eta) = \mathrm{diag}\bigl(h(\eta_1), \ldots, h(\eta_p)\bigr)
$$ {#eq-diag-struct}

is diagonal, which is exactly the contract a link satisfies, so `parameters7`
takes $h$ from `linkfunctions7` instead of writing a positive map of its own.
Any link whose range is the positive half-line serves, and its exact
derivatives to fourth order come with it. The derivatives of the parameter are
then $A_k = h'(\eta_k) e_k e_k^\top$ and $A_{kk} = h''(\eta_k) e_k e_k^\top$,
with every mixed second derivative zero, and the log-determinant is
$\sum_i \log h(\eta_i)$.

```{r}
#| label: diag-struct
parameters7::param_value(parameters7::diagonal_matrix(3), c(0, 0.5, -0.5))
parameters7::param_value(
  parameters7::diagonal_matrix(2, link = linkfunctions7::sqrt_link()), c(1, 2)
)
```

A single free value shared by every entry gives $\sigma^2 I$, which is the
independent case a variance component or a ridge starts from. It is the same
construction with $d = 1$.

### A fixed matrix carried by one scale {#sec-parameter-scaled}

The third family is different in kind. Here the matrix is

$$
M(\eta) = \lambda P, \qquad \lambda = h(\eta),
$$ {#eq-scaled-struct}

with $P$ a **known** symmetric positive semidefinite matrix and $\lambda$ the
only unknown. This is the commonest penalty in semiparametric regression: $P$
may be the Gram matrix of a basis derivative from @sec-basis-gram, a
difference penalty $\Delta^\top \Delta$, a neighbourhood matrix, or the
identity, in which case the parameter is a ridge.

Nothing here needs deriving. With the default log link, $\lambda = e^{\eta}$
and every quantity is a constant times a function of the scale:

$$
\partial_\eta M = M, \qquad \partial^2_\eta M = M,
$$

$$
\log|M|_{+} = r\,\eta + \log|P|_{+}, \qquad
\partial_\eta \log|M|_{+} = r, \qquad
\partial^2_\eta \log|M|_{+} = 0,
$$ {#eq-scaled-logdet}

with $r$ the rank of $P$ and $|\cdot|_{+}$ the pseudo-determinant of
@sec-parameter-pseudodet. The constant $\log|P|_{+}$ is computed once at
construction, since $P$ does not change.

```{r}
#| label: scaled-struct
# a ridge: the identity, scaled
r <- parameters7::scaled_matrix(diag(4))
c(rank = r@rank, logdet_at_zero = parameters7::param_logdet(r, 0))

# a second-difference penalty on six coefficients
dm <- diff(diag(6), differences = 2)
pen <- parameters7::scaled_matrix(crossprod(dm))
c(dimension = pen@dimension, rank = pen@rank)
```

### The simplex and the transition matrix {#sec-parameter-simplex}

A probability vector $\pi$ on the open simplex satisfies $\pi_a > 0$ and
$\sum_a \pi_a = 1$, so a vector of $K$ categories has $K - 1$ free values.
The package carries it by the additive log-ratio map with the last category
as reference,
$$\pi_a = \frac{e^{\eta_a}}{1 + \sum_b e^{\eta_b}}, \qquad
  \pi_K = \frac{1}{1 + \sum_b e^{\eta_b}},$$
whose inverse is $\eta_a = \log(\pi_a / \pi_K)$, exact. The derivatives of
this map close over the value itself: writing
$Z = 1 + \sum_b e^{\eta_b}$, each $\pi_a$ with $a < K$ is a first derivative
of $\log Z$, so the derivative arrays of $\pi$ are the higher derivatives of
$\log Z$ -- the cumulants of a categorical indicator -- and every order
follows from the single rule
$$\partial_b\, \pi_a = \pi_a(\delta_{ab} - \pi_b)$$
applied repeatedly with the product rule. All four orders are closed form,
and one identity comes free: differentiating $\sum_a \pi_a = 1$ shows that
every derivative component sums to zero over the value index, at every order.

```{r}
#| label: simplex-orders
s <- simplex(4)
eta <- c(0.5, -0.2, 0.8)
round(param_value(s, eta), 4)

# the free identity: each component of each order sums to zero
range(vapply(param_d3(s, eta), sum, numeric(1)))
```

A transition matrix of a Markov chain on $K$ states is a $K \times K$ matrix
whose rows lie on the simplex, and `transition_matrix()` treats it as exactly
that: $K$ independent additive log-ratio charts, one per row, $K(K-1)$ free
values in all. The independence carries to the derivatives, which are block
diagonal by row -- a component pairing free values of two different rows is
exactly zero, not small -- so the cost of an order grows with the row, not
with the matrix.

### The matrix logarithm {#sec-parameter-matrixlog}

The log-Cholesky map is not the only chart on the positive definite cone.
`matrix_log()` writes $M = \exp(S)$ with $S$ symmetric and genuinely free:
the free values fill the triangle of $S$ directly, and no entry is
transformed. Two quantities are then exact by construction. The
log-determinant is the trace,
$$\log\lvert M \rvert = \operatorname{tr}(S),$$
linear in the diagonal free values with every higher derivative zero, and the
inverse is $M^{-1} = \exp(-S)$, through the same eigendecomposition that
evaluates the map. This is the chart to reach for when the covariance and
the precision are wanted at once; log-Cholesky remains the cheaper one for
derivatives, which there are sparse products.

The derivatives of $\exp(S)$ in the free values are its Fréchet derivatives,
and the package computes them by the representation of @daleckii1965: with
$S = Q \Lambda Q^\top$ and the directions rotated by $Q$, the $k$-th
derivative contracts chains of directions against divided differences of
$e^x$ of order $k + 1$ at the eigenvalues, summed over the orderings of the
directions. The divided differences themselves are computed by the theorem of
@opitz1964 -- the exponential of a small upper bidiagonal matrix, read off
its corner entry -- which remains exact when eigenvalues repeat or nearly
repeat, where the recursive quotient formula cancels catastrophically.

```{r}
#| label: matrixlog
m <- matrix_log(3)
eta <- c(0.2, -0.3, 0.4, 0.5, -0.1, 0.3)

# the log-determinant is the trace of S, read off the free vector
c(param_logdet(m, eta), sum(eta[1:3]))

# and the inverse costs no factorisation
max(abs(param_solve(m, eta) - solve(param_value(m, eta))))
```

### Third and fourth derivatives {#sec-parameter-higher}

Every family above carries its derivatives to fourth order in closed form:
log-Cholesky by the Leibniz rule on $M = LL^\top$, the diagonal and scaled
families through the link derivatives that `linkfunctions7` already supplies
at fourth order, the simplex by its cumulant recursion, and the matrix
logarithm by the divided-difference contraction. A parameter defined outside
the package is served by numerical fallbacks that apply one product stencil
directly to the map for each component -- never a stencil on a lower-order
numerical derivative -- so the accuracy degrades honestly with the order and
the no-nesting rule of the toolkit holds structurally.

The higher orders exist for a consumer: the multivariate gaussian of
@sec-mvgaussian builds its own third and fourth derivatives from these, one
Leibniz expansion away, and a mixed-effects model that profiles or corrects
its likelihood asks for exactly that.

### Which side of a model a family parametrises {#sec-parameter-role}

The name of a family does not say whether the matrix it produces is a
covariance or a precision, and the two are different models. Imposing a
parameter on $\Sigma$ and imposing the same parameter on $\Omega = \Sigma^{-1}$
give the same family only when the parameter is closed under inversion.
Diagonal matrices are, and so is compound symmetry, whose inverse is again
compound symmetry with different constants. An AR(1) correlation matrix is not:
its inverse is tridiagonal, which is a different pattern of zeros, so an AR(1)
covariance and an AR(1) precision are genuinely different assumptions about
the same data.

A parameter therefore carries a label recording which side it is meant for.
Nothing computed depends on it and no method reads it; it exists so that the
object says what it is, and so that a consumer that cares can ask. The
distribution of @sec-mvgaussian takes the parameter on either side and adjusts
its own derivatives accordingly, which is the only place the difference has
arithmetic consequences.

```{r}
#| label: consistency-06b
#| include: false
assert_parameter_families_ok()
```
