## Rank and the null space {#sec-parameter-rank}

```{r}
#| label: setup-06c
#| include: false
source("R/_setup.R")
source("R/parameter-certificates.R")
```

### Why a singular matrix is admitted {#sec-parameter-singular}

A covariance matrix must be positive definite for the density it describes to
exist, and a precision matrix must be positive definite for the same reason: a
singular covariance puts the whole distribution on a subspace, and a singular
precision leaves the density unnormalisable. Both readings of a multivariate
gaussian therefore require full rank, and the distribution of @sec-mvgaussian
refuses a rank-deficient parameter.

A penalty is not a density. Writing a quadratic penalty
$\frac{\lambda}{2}\beta^\top P \beta$ as a negative log prior makes it a
gaussian with precision $\lambda P$, and the whole point of a smoothing
penalty is that $P$ is singular: its null space is the set of functions the
penalty leaves alone. A second-difference penalty on a spline has the
constants and the straight lines in its null space, which is what makes the
penalty shrink towards a straight line rather than towards zero. That prior is
improper, and its impropriety is the modelling assumption rather than a defect.

The package therefore admits rank-deficient matrices, and everything a
consumer asks of one has to be redefined so that it still means something.

### The pseudo-determinant {#sec-parameter-pseudodet}

The determinant of a singular matrix is zero, so $\log|M|$ is $-\infty$ and
useless. The quantity that takes its place is the **pseudo-determinant**, the
product of the non-zero eigenvalues,

$$
\log|M|_{+} \;=\; \sum_{j : \, \nu_j \neq 0} \log \nu_j ,
$$ {#eq-pseudodet}

which is the normalising constant of the proper gaussian the penalty induces
on the complement of its null space. One generic returns both, because a
consumer asks the same question of either matrix — what normalising constant
does this contribute — and the object knows which answer is right, through the
rank it declares.

The gradient identity @eq-parameter-dlogdet survives with the Moore-Penrose
inverse in place of $M^{-1}$, and for the scaled family of
@sec-parameter-scaled it collapses to the constant of @eq-scaled-logdet.

That constant is what makes the scale estimable. The penalised objective seen
as a negative log posterior contributes

$$
\frac{\lambda}{2}\,\beta^\top P \beta \;-\; \frac{r}{2}\log\lambda ,
$$ {#eq-scaled-objective}

and setting its derivative in $\lambda$ to zero gives

$$
\hat{\lambda} = \frac{r}{\beta^\top P \beta}.
$$ {#eq-scaled-lambda}

Dropping the second term of @eq-scaled-objective, which is what a treatment
that regards the penalty as a penalty rather than as a prior does, leaves a
derivative of one sign and sends $\lambda$ to zero. The normalising constant is
therefore kept, and the rank in @eq-scaled-lambda is the reason the rank has
to be right.

```{r}
#| label: pseudodet-lambda
dm <- diff(diag(8), differences = 2)
pen <- parameters7::scaled_matrix(crossprod(dm))
p_mat <- crossprod(dm)

set.seed(3)
beta <- as.numeric(rnorm(8))
quad <- drop(t(beta) %*% p_mat %*% beta)

# the stationary point of (lambda/2) b'Pb - (r/2) log lambda, and the same
# point found by minimising the objective numerically
obj <- function(lambda) 0.5 * lambda * quad - 0.5 * pen@rank * log(lambda)
c(closed_form = pen@rank / quad,
  numerical = stats::optimize(obj, c(1e-8, 1e8))$minimum)
```

### Determining the rank {#sec-parameter-rank-determination}

The rank of a matrix is a statement about the matrix, and a numerical
procedure that reads it off an assembled matrix does not always make that
statement. Counting eigenvalues above a relative tolerance is the obvious
route, and it fails on exactly the object this package exists to carry.

Consider a tensor-product penalty $\lambda_1 P_1 + \lambda_2 P_2$ on a
$4 \times 8$ grid of coefficients, where $P_1$ penalises second differences
along one margin and $P_2$ along the other. Its true rank is 28 out of 32.
Counting eigenvalues of the assembled matrix above a relative tolerance gives
28 while the two smoothing parameters are comparable and 24 once they are ten
orders of magnitude apart, because the contributions of the smaller term sink
below the tolerance and are read as exact zeros. Smoothing parameters that far
apart are an ordinary fitted model rather than a pathology, so the count is
not a property of the family at all.

What is stable is the null space. The null space of a sum of positive
semidefinite matrices is the **intersection** of their null spaces, since
$\beta^\top(P_1 + P_2)\beta = 0$ with both terms non-negative forces both to
vanish. The intersection is the null space of the stacked matrix
$\begin{pmatrix} P_1 \\ P_2 \end{pmatrix}$, so the rank comes from the
components, individually normalised and stacked, and never from an assembled
combination. Membership of the null space is then tested against the stored
basis rather than by counting eigenvalues again.

```{r}
#| label: rank-scale-invariance
d2 <- function(k) crossprod(diff(diag(k), differences = 2))
p1 <- kronecker(d2(4), diag(8))
p2 <- kronecker(diag(4), d2(8))

count_ev <- function(m, tol = 1e-10) {
  ev <- eigen(m, symmetric = TRUE, only.values = TRUE)$values
  sum(ev > tol * max(ev))
}

# the stored null basis, and the residual it leaves in the assembled matrix
ns <- parameters7::param_null_basis(list(p1, p2))

ratios <- c(1, 1e4, 1e8, 1e10, 1e12)
data.frame(
  ratio = ratios,
  eigen_count = vapply(ratios, function(f) count_ev(p1 + f * p2), numeric(1)),
  stacked_rank = ns$rank,
  null_residual = vapply(ratios, function(f) {
    m <- p1 + f * p2
    max(abs(m %*% ns$null_basis)) / max(abs(m))
  }, numeric(1))
)
```

The count moves and the null space does not. The residual $\lVert M N \rVert /
\lVert M \rVert$, with $N$ the stored basis, stays at the level of rounding
across the whole range, which is the statement that the four directions really
are annihilated by the penalty at every combination of smoothing parameters.

The same doubt applies wherever a Cholesky factorisation is used to decide a
branch. On a matrix with an exactly zero eigenvalue the pivot that ought to be
zero comes out positive or negative according to rounding, so a
`tryCatch(chol(...))` used as a rank test answers differently on different
machines for the same matrix. A verdict of that kind belongs to the
eigenvalues, which describe the matrix, rather than to the arithmetic that was
performed on it.

### The solve, which a singular matrix does not have {#sec-parameter-singular-solve}

The remaining quantity is the solve, and it is the one that is refused rather
than redefined. A Moore-Penrose solve $M^{+}B$ exists and is easy to compute,
but no consumer of an improper prior asks for it. What a penalised fit inverts
is $X^\top X + \lambda P$, which is non-singular even when $P$ is not, and
which the consumer assembles from the penalty and its own design; what it asks
of the penalty itself is the quadratic form $\beta^\top P \beta$ and the log
pseudo-determinant. Returning a pseudo-inverse would be a plausible matrix
answering a question nobody asked, and a caller that used it would be wrong
without being told.

```{r}
#| label: singular-solve
#| error: true
parameters7::param_solve(pen, 0)
```

What the null space is for, instead, is membership. A basis of it is stored at
construction, so a consumer can ask whether a coefficient vector lies in the
space the penalty leaves alone — for a second-difference penalty, whether it
is a straight line — without touching an eigenvalue.

```{r}
#| label: nullspace-membership
n_basis <- pen@null_basis
line <- 1 + 2 * seq_len(8)
quad_of <- function(b) drop(t(b) %*% p_mat %*% b)

c(
  columns = ncol(n_basis),
  penalty_of_a_line = quad_of(line),
  penalty_of_a_curve = quad_of(seq_len(8)^2),
  line_is_in_the_null_space =
    max(abs(line - n_basis %*% crossprod(n_basis, line))) < 1e-10
)
```

```{r}
#| label: consistency-06c
#| include: false
assert_parameter_rank_ok()
```
