## In practice {#sec-basis-practice}

```{r}
#| label: setup-06p
#| include: false
source("R/_setup.R")
```

### A penalized smooth from the pieces {#sec-basis-smooth-practice}

The package supplies the design matrix and the exact penalty matrix; a
penalized least-squares fit is then two lines of linear algebra. The data
are a noisy sine, the basis a cubic B-spline of twenty functions, and the
penalty the integrated squared second derivative of @sec-gram-definition:

```{r}
#| label: bp-smooth
#| fig-height: 4
#| fig-cap: "A B-spline smooth at three values of the smoothing parameter. The penalty matrix is the exact Gram matrix of the second derivatives."
library(basis7)
set.seed(2)
x <- sort(runif(150))
yy <- sin(2 * pi * x) + rnorm(150, sd = 0.35)

b <- bspline_basis(lower = 0, upper = 1, dimension = 20, degree = 3)
B <- basis_eval(b, x)
P <- basis_gram(b, order = 2)

fit_smooth <- function(lambda)
  drop(solve(crossprod(B) + lambda * P, crossprod(B, yy)))

xg <- seq(0, 1, length.out = 300)
Bg <- basis_eval(b, xg)
plot(x, yy, col = "grey55", pch = 16, cex = 0.6, xlab = "x", ylab = "y")
for (lam in c(1e-6, 1e-3, 1)) lines(xg, Bg %*% fit_smooth(lam))
lines(xg, sin(2 * pi * xg), lty = 2)
```

The smallest smoothing parameter follows the noise and the largest is close
to the straight line the null space of $G_2$ contains; the dashed curve is
the function that generated the data.

The derivative of the fitted function costs one more matrix, since
@eq-basis-derivative applies the same coefficients to the differentiated
basis:

```{r}
#| label: bp-deriv
beta <- fit_smooth(1e-3)
d1 <- basis_deriv(b, c(0.25, 0.5, 0.75), order = 1) %*% beta
cbind(estimate = drop(d1),
      truth = 2 * pi * cos(2 * pi * c(0.25, 0.5, 0.75)))
```

### The transforms composed {#sec-basis-transform-practice}

The three transformations of @sec-basis-transforms return bases, so they
compose with everything above. A constrained, orthonormalized basis in two
calls:

```{r}
#| label: bp-transform
grid <- seq(0, 1, length.out = 200)
cb <- constrain_basis(b, colSums(basis_eval(b, grid)))   # sum-to-zero
ob <- orthonorm_basis(cb)

c(dimension = ob@dimension)
max(abs(basis_gram(ob) - diag(ob@dimension)))            # orthonormal
max(abs(colSums(basis_eval(ob, grid))))                  # still constrained
```
