## The definition {#sec-penalty-definition}

```{r}
#| label: setup-08a
#| include: false
source("R/_setup.R")
source("R/penalty-certificates.R")
```

A penalty on a coefficient vector $\beta \in \mathbb{R}^q$ is a function

$$
\rho(D\beta;\,\theta),
$$

built from three parts: an $m \times q$ matrix $D$ selecting what is
penalized -- the coefficients themselves, their successive differences, the
jumps of a segmented term -- a scalar function $\rho$ of the resulting
$m$-vector, and hyperparameters $\theta$ governing how strongly. Penalized
estimation adds $\rho$ to the negative log-likelihood; Bayesian estimation
reads $\exp(-\rho)$ as a prior density on $D\beta$. The two readings use the
same object, and the package takes the Bayesian one as the organizing
definition because it forces a decision the frequentist reading lets one
avoid: what to do with the normalizing constant.

`penalties7` keeps the constant. A proper penalty evaluates to exactly the
negative logarithm of a normalized density, so that, for the ridge penalty
with precision $\lambda$ -- the Gaussian prior $N(0, \lambda^{-1}I_q)$,
written by the parameter that measures the shrinkage rather than by its
scale, so that a larger value shrinks harder as it does in every other
penalty here --

$$
\rho(\beta;\,\lambda)
  \;=\; \frac{\lambda}{2}\,\beta'\beta
  \;-\; \frac{q}{2}\log\lambda \;+\; \frac{q}{2}\log 2\pi .
$$

Penalized-likelihood software drops the two trailing terms, because at fixed
$\lambda$ they do not move the coefficient estimate. They do, however,
determine everything that treats $\lambda$ as unknown: joint maximization
over $(\beta, \lambda)$ without the constant improves without bound as the
prior degenerates, $\lambda$ running to zero with nothing to hold it, and
the degrees of freedom $\nu$ of a heavy-tailed prior
appears only in its constant, so dropping it makes $\nu$ inestimable.
Keeping the constant costs one closed expression per penalty and is what
allows the hyperparameters to be estimated rather than tuned.

An improper penalty has no constant to keep. A quadratic penalty whose
matrix is rank deficient is a Gaussian that is flat along the null space,
and the bounded penalties of @sec-penalty-branches make $\exp(-\rho)$
non-integrable outright. The quadratic case keeps the proper constant of its
non-degenerate part -- the rank replaces the dimension, and the log
pseudo-determinant replaces the log determinant -- which is precisely the
constant a marginal-likelihood criterion uses; the bounded case returns the
bare $\rho$ and `is_proper()` says so.

Each hyperparameter travels with a link function from @sec-links, so that a
consumer optimizes $\theta$ on the unconstrained scale; the derivatives with
respect to $\theta$ are returned on the parameter scale by every method and
carried onto the link scale by the chain rule in one place, exactly as
`distributions7` does for its parameters. Beyond the value, every penalty
answers the derivatives an estimation routine asks: the gradient and Hessian
in $\beta$, the gradient and Hessian in $\theta$, and the mixed block
$\partial^2 \rho / \partial\beta\,\partial\theta_k$, which joint estimation
of coefficients and hyperparameters consumes and which the implicit function
theorem turns into the gradient of a profiled objective. A penalty also
declares its kink set (the points where $\rho$ is not differentiable in its
argument, empty for the smooth penalties and $\{0\}$ for the sparsity ones),
so that a validator or a solver can avoid the points where $\rho$ is not
smooth.

```{r}
#| label: penalty-first-look
pen <- ridge_penalty(n_coef = 3)
beta <- c(0.8, -0.4, 1.1)

penalty_value(pen, beta, list(lambda = 0.44))
penalty_gradient(pen, beta, list(lambda = 0.44))

# the mixed block: how the beta-gradient moves when lambda does
penalty_cross(pen, beta, list(lambda = 0.44))$lambda
```

```{r}
#| label: consistency-08a
#| include: false
assert_penalty_definition_ok()
```
