## Jets {#sec-numerical-jets}

```{r}
#| label: setup-07a
#| include: false
source("R/_setup.R")
source("R/numerical-certificates.R")
```

A jet is a value carried together with every partial derivative up to a
fixed order -- here the fourth, because that is the order the toolkit's
derivative contracts run to. Arithmetic on jets propagates the derivatives
exactly: the sum of two jets adds componentwise, the product combines the
components by the Leibniz rule, and a smooth function of a jet composes its
own derivatives with the jet's through the chain rule. A map written in
ordinary R, evaluated on jets instead of numbers, therefore returns its own
derivatives without any of them having been transcribed. The construction is
the forward mode of algorithmic differentiation, truncated at order four,
and @griewank2008 is the standard reference.

The value of the construction lies in maps that are recursions rather than
formulas. The Levinson-Durbin recursion of @sec-parameter-families carries
the partial autocorrelations of an autoregressive process onto its
coefficients through repeated sums and products; its fourth derivative
expanded by hand is pages of algebra whose only check would be the nested
finite differences the toolkit refuses everywhere. Carried in jets, the
recursion is typed exactly as the textbook writes it and every derivative
order falls out exact. The centered skew normal parametrization of
@sec-distributions7 goes through the same machinery for the same reason.

A jet over $d$ free values stores the value and one component for each
multi-index of order one to four. `jet_layout(d)` fixes the indexing once,
`jet_var(k, ...)` seeds the $k$-th coordinate -- its own first derivative
one, everything else zero -- and `jet_const()` embeds a constant, whose
derivatives all vanish. The ordinary operators then dispatch on the class:

```{r}
#| label: jet-weibull
lay <- jet_layout(2)
mu    <- jet_var(1, list(4.0, 1, 0, 0, 0), lay)
sigma <- jet_var(2, list(1.7, 1, 0, 0, 0), lay)

# the map from a Weibull's scale parametrization: scale = mu / gamma(1 + 1/sigma)
scale <- mu / gamma(1 + 1 / sigma)

# d scale / d sigma, read off the jet and written by hand
c(jet = scale$d[[1]][2],
  hand = 4.0 * digamma(1 + 1 / 1.7) / (1.7^2 * gamma(1 + 1 / 1.7)))
```

The product of two jets is the Leibniz rule over the positions of a
multi-index: the component of order $I$ collects, over every subset $S$ of
the positions of $I$, the product of one factor's component at $S$ and the
other's at the complement. This is the same subset enumeration the Gram
matrices of @sec-basis-gram use, and the package implements it once. The
partitions themselves -- of a set of derivative positions into blocks, which
is what a Faà di Bruno formula sums over -- are enumerated by
`set_partitions()`, and the multi-indices that key a derivative list by
`tuple_indices()`; both exist in exactly one copy, so no two consumers of
the enumeration can disagree about it [@comtet1974].

Comparisons and non-smooth functions are refused on jets rather than
approximated. A branch taken on a jet would keep the derivatives of one arm
and report them as the derivatives of the whole expression, which is wrong
whenever the branch point is in the interior of the domain; `abs()` fails
for the same reason. A map that needs a branch takes it outside the jet
computation, on plain numbers, and seeds the jets afterwards -- which is how
the centered skew normal handles the sign of its skewness.

```{r}
#| label: consistency-07a
#| include: false
assert_jets_ok()
```
