## The four constructions {#sec-penalty-branches}

```{r}
#| label: setup-08b
#| include: false
source("R/_setup.R")
source("R/penalty-certificates.R")
```

Every penalty the package ships is one of four shapes, and the four cover
the penalties regression practice actually uses.

### The quadratic penalty {#sec-penalty-quadratic}

The quadratic penalty is the Gaussian prior
$\beta \sim N\!\left(0,\,(\lambda\,D'PD)^{-}\right)$ written as its exact
negative log-density,

$$
\rho \;=\; \frac{\lambda}{2}\,(D\beta)'P(D\beta)
  \;-\; \frac{r}{2}\log\lambda
  \;+\; \frac{r}{2}\log 2\pi
  \;-\; \frac{1}{2}\log\operatorname{pdet}(P),
$$ {#eq-penalty-quadratic}

with $P$ a fixed positive semidefinite matrix -- a Gram matrix from
@sec-basis-gram, a difference penalty $D_k'D_k$ in the manner of
@eilers1996, or anything the caller supplies -- and $r$ its rank. One
eigendecomposition at construction fixes $r$ by the relative rule of
@sec-parameter-rank and stores the exact null basis, so that later
membership questions never go through a rank recomputed from assembled
arithmetic. Everything is then closed form in $\lambda$: the gradient is
$\lambda D'PD\beta$, the Hessian $\lambda D'PD$ constant in $\beta$, the
$\lambda$-derivative of the value $\tfrac{1}{2}(D\beta)'P(D\beta) -
r/(2\lambda)$, and the mixed block $D'PD\beta$.

```{r}
#| label: quadratic-example
P <- crossprod(diff(diag(6), differences = 2))
pen <- quadratic_penalty(P)

# an improper prior: flat along the linear polynomials
penalty_rank(pen)
is_proper(pen)
```

### The sum of quadratics {#sec-penalty-additive}

A single matrix under a single scale forces every direction to be
smoothed alike. A tensor-product smooth wants the opposite: one
smoothing parameter per margin, so that a surface may be rough in one
direction and smooth in another. `additive_penalty()` carries

$$
S(\lambda) \;=\; \sum_{k} \lambda_k P_k,
$$

and everything a marginal criterion reads follows from one
eigendecomposition of that sum:
$\partial_k \log\mathrm{pdet}\,S = \operatorname{tr}(S^{+}P_k)$ and
$\partial^2_{kl} \log\mathrm{pdet}\,S =
-\operatorname{tr}(S^{+}P_kS^{+}P_l)$.

The rank needs care, and @sec-parameter-rank is the reason. Counting the
eigenvalues of $S(\lambda)$ above a tolerance answers correctly only
while the parameters are comparable: once they differ by orders of
magnitude the small contributions sink below any fixed tolerance and are
read as zeros, so the rank appears to fall as the fit is smoothed. What
does not move is the null space, because the null space of a sum of
positive semidefinite matrices is the intersection of theirs and does
not depend on the parameters at all. The rank is therefore fixed once,
at construction, from the components stacked and individually
normalized.

```{r}
#| label: penalty-additive
P1 <- crossprod(diff(diag(4), differences = 2))
pen <- additive_penalty(list(kronecker(diag(4), P1),
                             kronecker(P1, diag(4))))
penalty_rank(pen)

# the count taken from the assembled matrix, as the parameters spread
sapply(c(1, 1e6, 1e12), function(ratio) {
  S <- kronecker(diag(4), P1) + ratio * kronecker(P1, diag(4))
  ev <- eigen(S, symmetric = TRUE, only.values = TRUE)$values
  sum(ev > 1e-10 * max(ev))
})
```

The stored rank is the first number at every value of the parameters;
the counted one is not.

### The separable penalty {#sec-penalty-separable}

The third construction applies a univariate log-density from
@sec-distributions7 coordinatewise:

$$
\rho(t;\,\theta) \;=\; -\sum_{j=1}^{m} \log f(t_j;\,\theta),
\qquad t = D\beta .
$$

The hyperparameters are the distribution's free parameters, with their
bounds and links read off the distribution object, and every derivative is
the distribution's, reassembled through the map: the gradient is
$-D'\ell^{(y)}$, the Hessian $-D'\,\mathrm{diag}(\ell^{(yy)})\,D$, the
$\theta$ blocks are the summed score and Hessian, and the mixed block is
$-D'\ell^{(y\theta_k)}$ -- the response-parameter derivative of
@sec-distributions7, which exists in closed form for every continuous
family and is the reason no separable penalty differences anything.

The classical penalties are named instances of this construction. Ridge is
the Gaussian held at zero with its scale free; the lasso is the Laplace
in location and rate (`laplace2_distrib()`) held at zero, so its free
hyperparameter is the rate $\lambda$ and the value is
$\lambda\lVert D\beta\rVert_1$ up to the constant, with the kink at the
origin declared by the object; the
heavy-tailed prior is the Student t held at zero, whose $\nu$ is estimable
exactly because the constant travels with the density. Ridge therefore
exists twice in the package -- once as @eq-penalty-quadratic with the
identity matrix, once as the separable Gaussian -- and the duplication is
deliberate: two independent implementations of one object can be compared
componentwise with no tolerance to choose, and the certification below does
exactly that at $\lambda = 1/\sigma^2$.

```{r}
#| label: separable-example
# the fused lasso: Laplace increments through a difference map
pen <- lasso_penalty(map = diff(diag(4)))
beta <- c(0.5, 1.4, -0.3, 2.0)
penalty_gradient(pen, beta, list(lambda = 0.8))

penalty_kinks(pen, list(lambda = 0.8))
```

### The families defined by their derivative {#sec-penalty-scad}

SCAD [@fan2001] and MCP [@zhang2010] are stated in the literature through
$\rho'$, because the derivative carries their design: a constant slope near
zero, so that small coefficients are penalized as under the lasso, and a
slope that falls to zero beyond a shoulder, so that large coefficients are
left unbiased. On $t \ge 0$, extended evenly,

$$
\rho'_{\mathrm{SCAD}}(t) =
\begin{cases}
\lambda & t \le \lambda \\[2pt]
\dfrac{a\lambda - t}{a - 1} & \lambda < t \le a\lambda \\[2pt]
0 & t > a\lambda,
\end{cases}
\qquad
\rho'_{\mathrm{MCP}}(t) = \left(\lambda - \frac{t}{\gamma}\right)_{+},
$$ {#eq-penalty-scad}

with $a > 2$ and $\gamma > 1$. The value follows by the closed piecewise
antiderivative anchored at $\rho(0) = 0$, and the $\theta$ and mixed blocks
are closed piecewise forms as well. Both penalties are bounded, so
$\exp(-\rho)$ does not integrate and no constant exists: they are improper
by construction, and the value is the bare $\rho$. The kink set contains
the origin, and the second derivative additionally jumps at the region
boundaries, which the object reports so that a numerical comparison never
straddles them.

```{r}
#| label: scad-example
pen <- scad_penalty(n_coef = 1)
th <- list(lambda = 1, a = 3.7)

# flat beyond the shoulder: no bias on large coefficients
penalty_value(pen, 5, th) == penalty_value(pen, 50, th)
```

```{r}
#| label: consistency-08b
#| include: false
assert_penalty_branches_ok()
```
