## The four constructions {#sec-penalty-branches}

```{r}
#| label: setup-08b
#| include: false
source("R/_setup.R")
source("R/penalty-certificates.R")
```

Every penalty the package ships is one of four shapes, and the four cover
the penalties regression practice actually uses. A fifth object closes the
section and is not a penalty at all: a smoother, which replaces the kink two
of the four carry, and which the model terms of @sec-modelterms7 use for the
same purpose on a break-point.

### The quadratic penalty {#sec-penalty-quadratic}

The quadratic penalty is the Gaussian prior
$\beta \sim N\!\left(0,\,(\lambda\,D'PD)^{-}\right)$ written as its exact
negative log-density,

$$
\rho \;=\; \frac{\lambda}{2}\,(D\beta)'P(D\beta)
  \;-\; \frac{r}{2}\log\lambda
  \;+\; \frac{r}{2}\log 2\pi
  \;-\; \frac{1}{2}\log\operatorname{pdet}(P),
$$ {#eq-penalty-quadratic}

with $P$ a fixed positive semidefinite matrix -- a Gram matrix from
@sec-basis-gram, a difference penalty $D_k'D_k$ in the manner of
@eilers1996, or anything the caller supplies -- and $r$ its rank. One
eigendecomposition at construction fixes $r$ by the relative rule of
@sec-parameter-rank and stores the exact null basis, so that later
membership questions never go through a rank recomputed from assembled
arithmetic. Everything is then closed form in $\lambda$: the gradient is
$\lambda D'PD\beta$, the Hessian $\lambda D'PD$ constant in $\beta$, the
$\lambda$-derivative of the value $\tfrac{1}{2}(D\beta)'P(D\beta) -
r/(2\lambda)$, and the mixed block $D'PD\beta$.

```{r}
#| label: quadratic-example
P <- crossprod(diff(diag(6), differences = 2))
pen <- quadratic_penalty(P)

# an improper prior: flat along the linear polynomials
penalty_rank(pen)
is_proper(pen)
```

#### One block, repeated {#sec-penalty-blocks}

A smooth fitted once per level of a factor gives the same penalty $P$ on
each of $m$ blocks, so the penalty of the whole is $I_m \otimes P$. Building
that from the assembled matrix decomposes something $m$ times larger than it
needs to be, and every quantity the constructor wants is available from $P$
alone: the eigenvalues of $I_m \otimes P$ are $P$'s repeated $m$ times, so
the rank is $mr$, the log pseudo-determinant is $m\log\operatorname{pdet}P$
and the null space is $I_m \otimes N$.

`quadratic_penalty(P, blocks = m)` takes that route, and the large matrix is
never formed or decomposed.

```{r}
#| label: penalty-blocks
P <- crossprod(diff(diag(6), differences = 2))
pb <- quadratic_penalty(P, blocks = 4)
pa <- quadratic_penalty(kronecker(diag(4), P))
b <- seq(-1, 1, length.out = 24)

c(rank_blocks = penalty_rank(pb), rank_assembled = penalty_rank(pa),
  value_gap = abs(penalty_value(pb, b, list(lambda = 2)) -
                  penalty_value(pa, b, list(lambda = 2))))
```

The branch keeps its sparse storage on purpose: `penalty_hessian()` and the
other matrix-valued returns come back as `Matrix` objects rather than base
ones, a construction built to avoid the assembled form having no reason to
densify at its own boundary. A consumer writing such a Hessian into a block
of its own information has to accept both classes, which is the one place
the uniform contract of the other branches is relaxed.

`blocks` does not combine with a map, and the constructor rejects the pair:
a map mixes the blocks, so $D'(I_m \otimes P)D$ is not block diagonal with
the structure `blocks` names.

### The sum of quadratics {#sec-penalty-additive}

A single matrix under a single scale forces every direction to be
smoothed alike. A tensor-product smooth wants the opposite: one
smoothing parameter per margin, so that a surface may be rough in one
direction and smooth in another. `additive_penalty()` carries

$$
S(\lambda) \;=\; \sum_{k} \lambda_k P_k,
$$

and everything a marginal criterion reads follows from one
eigendecomposition of that sum:
$\partial_k \log\mathrm{pdet}\,S = \operatorname{tr}(S^{+}P_k)$ and
$\partial^2_{kl} \log\mathrm{pdet}\,S =
-\operatorname{tr}(S^{+}P_kS^{+}P_l)$.

The rank needs care, and @sec-parameter-rank is the reason. Counting the
eigenvalues of $S(\lambda)$ above a tolerance answers correctly only
while the parameters are comparable: once they differ by orders of
magnitude the small contributions sink below any fixed tolerance and are
read as zeros, so the rank appears to fall as the fit is smoothed. What
does not move is the null space, because the null space of a sum of
positive semidefinite matrices is the intersection of theirs and does
not depend on the parameters at all. The rank is therefore fixed once,
at construction, from the components stacked and individually
normalized.

```{r}
#| label: penalty-additive
P1 <- crossprod(diff(diag(4), differences = 2))
pen <- additive_penalty(list(kronecker(diag(4), P1),
                             kronecker(P1, diag(4))))
penalty_rank(pen)

# the count taken from the assembled matrix, as the parameters spread
sapply(c(1, 1e6, 1e12), function(ratio) {
  S <- kronecker(diag(4), P1) + ratio * kronecker(P1, diag(4))
  ev <- eigen(S, symmetric = TRUE, only.values = TRUE)$values
  sum(ev > 1e-10 * max(ev))
})
```

The stored rank is the first number at every value of the parameters;
the counted one is not.

### The separable penalty {#sec-penalty-separable}

The third construction applies a univariate log-density from
@sec-distributions7 coordinatewise:

$$
\rho(t;\,\theta) \;=\; -\sum_{j=1}^{m} \log f(t_j;\,\theta),
\qquad t = D\beta .
$$

The hyperparameters are the distribution's free parameters, with their
bounds and links read off the distribution object, and every derivative is
the distribution's, reassembled through the map: the gradient is
$-D'\ell^{(y)}$, the Hessian $-D'\,\mathrm{diag}(\ell^{(yy)})\,D$, the
$\theta$ blocks are the summed score and Hessian, and the mixed block is
$-D'\ell^{(y\theta_k)}$ -- the response-parameter derivative of
@sec-distributions7, which exists in closed form for every continuous
family and is the reason no separable penalty differences anything.

The classical penalties are named instances of this construction. The lasso
is the Laplace in location and rate (`laplace2_distrib()`) held at zero, so
its free hyperparameter is the rate $\lambda$ and the value is
$\lambda\lVert D\beta\rVert_1$ up to the constant, with the kink at the
origin declared by the object; the
heavy-tailed prior is the Student t held at zero, whose $\nu$ is estimable
exactly because the constant travels with the density.

Ridge is the exception, and it is an exception to the branch rather than to
the rule. The Gaussian held at zero is a separable penalty like the others,
but that prior written by its PRECISION is exactly @eq-penalty-quadratic at
the identity matrix -- the same value to the last bit -- so it is built
there, and its hyperparameter is the $\lambda$ that construction already
carries. Two names for one number is what the choice removes: every penalty
here is written on the chart whose hyperparameter measures the shrinkage,
and a larger value shrinks harder in all of them. The separable Gaussian
remains as the twin the quadratic one is held to, where two independent
implementations of one object are compared componentwise with no tolerance
to choose, and the certification below does exactly that.

The elastic net is the same construction over the family that is the
product of those first two, `enet_distrib()` held at zero. Writing
$a = \lambda\alpha$ and $c = \lambda(1-\alpha)$, its value is

$$
\rho(D\beta; \lambda, \alpha) =
  \lambda\left\{\alpha\lVert D\beta\rVert_1
  + \frac{1-\alpha}{2}\lVert D\beta\rVert_2^2\right\}
  + p\log Z(\lambda, \alpha),
\qquad
Z = \frac{2M(a/\sqrt{c})}{\sqrt{c}},
$$ {#eq-penalty-enet}

with $M$ the Mills ratio. That the constant is available in closed form is
what makes both hyperparameters estimable: a penalty written as a formula
would have the same minimizer at fixed $(\lambda, \alpha)$ and nothing to
say about which pair to choose, because a criterion that compares them
compares normalized densities. The mixing weight lives in the open unit
interval, the two ends being the lasso and ridge of the paragraph above,
which remain penalties of their own.

Two things follow from the mixture that do not from either part. The
penalized objective is not differentiable, since the $\ell_1$ term carries
its kink through whatever $\alpha$ multiplies it, so the elastic net is a
non-smooth penalty at every admissible $\alpha$; and its proximal operator
is the composition of the two elementary ones,

$$
\operatorname{prox}_{t\rho}(v) =
  \frac{\mathcal{S}(v,\; t\lambda\alpha)}{1 + t\lambda(1-\alpha)},
$$ {#eq-prox-enet}

the soft threshold of the Laplace part followed by the shrinkage of the
Gaussian one, which follows from the stationary condition
$(\beta - v)/t + a\,\mathrm{sgn}(\beta) + c\beta = 0$ separating
coordinatewise.

```{r}
#| label: enet-example
pen <- elasticnet_penalty(n_coef = 4)
beta <- c(0.5, 1.4, -0.3, 2.0)
th <- list(lambda = 0.8, alpha = 0.6)

# the value, less its constant, is the elastic net of the coefficients
penalty_value(pen, beta, th) - penalty_value(pen, rep(0, 4), th)
0.8 * (0.6 * sum(abs(beta)) + 0.4 * sum(beta^2) / 2)

# and the operator is the threshold followed by the shrinkage
penalty_prox(pen, beta, step = 0.5, th)
```


```{r}
#| label: separable-example
# the fused lasso: Laplace increments through a difference map
pen <- lasso_penalty(map = diff(diag(4)))
beta <- c(0.5, 1.4, -0.3, 2.0)
penalty_gradient(pen, beta, list(lambda = 0.8))

penalty_kinks(pen, list(lambda = 0.8))
```

#### The operator as a table {#sec-penalty-prox-spec}

A proximal step is applied once per coordinate per sweep, at a point that
moves every time, so a coordinate descent that called back into R for each
one would spend its gain on the calls. Every closed-form operator this
branch produces is **piecewise linear** in the coordinate -- the soft
threshold has two pieces, the elastic net two, MCP three and SCAD four -- so
it can be handed over as a table instead of as a function.

`penalty_prox_spec()` returns that table: one row per coefficient, and per
row the cuts, the slopes and the intercepts of the pieces. A compiled loop
then evaluates the operator by locating a cut and applying an affine map,
naming no family at all.

```{r}
#| label: prox-spec
sp <- penalty_prox_spec(lasso_penalty(n_coef = 3), list(lambda = 1.2),
                        step = 0.5)
rbind(cut = sp$cut[1, ], slope = sp$slope[1, ], icept = sp$icept[1, ])
```

The first piece is the interval the threshold sets to zero, of half-width
$t\lambda = 0.6$, and the second is the shrinkage $v - 0.6$ beyond it. The
table is read in the **magnitude** of the coordinate and the sign restored
afterwards, every one of these operators being odd, which is also why one
row of cuts describes both tails.

The step enters the table rather than the lookup, so a table built at one
step does not serve another; in a coordinate descent the step is
$1/v_j$ and does not move while the working weights are held, which is what
makes one table serve a whole sweep. The quadratic branch returns `NULL`
here and is not a gap: its proximal operator is a linear solve rather than a
piecewise map, and there is nothing to tabulate.

### The families defined by their derivative {#sec-penalty-scad}

SCAD [@fan2001] and MCP [@zhang2010] are stated in the literature through
$\rho'$, because the derivative carries their design: a constant slope near
zero, so that small coefficients are penalized as under the lasso, and a
slope that falls to zero beyond a shoulder, so that large coefficients are
left unbiased. On $t \ge 0$, extended evenly,

$$
\rho'_{\mathrm{SCAD}}(t) =
\begin{cases}
\lambda & t \le \lambda \\[2pt]
\dfrac{a\lambda - t}{a - 1} & \lambda < t \le a\lambda \\[2pt]
0 & t > a\lambda,
\end{cases}
\qquad
\rho'_{\mathrm{MCP}}(t) = \left(\lambda - \frac{t}{\gamma}\right)_{+},
$$ {#eq-penalty-scad}

with $a > 2$ and $\gamma > 1$. The value follows by the closed piecewise
antiderivative anchored at $\rho(0) = 0$, and the $\theta$ and mixed blocks
are closed piecewise forms as well. Both penalties are bounded, so
$\exp(-\rho)$ does not integrate and no constant exists: they are improper
by construction, and the value is the bare $\rho$. The kink set contains
the origin, and the second derivative additionally jumps at the region
boundaries, which the object reports so that a numerical comparison never
straddles them.

```{r}
#| label: scad-example
pen <- scad_penalty(n_coef = 1)
th <- list(lambda = 1, a = 3.7)

# flat beyond the shoulder: no bias on large coefficients
penalty_value(pen, 5, th) == penalty_value(pen, 50, th)
```

### Smoothing an absolute value {#sec-penalty-smoother}

A kink is what makes the last two constructions non-smooth, and what makes a
break-point term's contribution non-differentiable in its position. Both can
be removed the same way, by replacing $\lvert u\rvert$ with a smooth
surrogate. An `abs_smoother()` is that surrogate as an object: a function
$s(u)$ standing in for the absolute value, carrying its own derivatives to
fifth order and a width that says over what scale the replacement acts.

The derivatives are carried as **functions** rather than as expressions, and
that is a decision the quintic forces: a piecewise smoother has branches
`stats::deriv()` cannot read, and writing the derivatives out in the
constructor makes those branches ordinary code. Only $s$ is contracted for;
the sign, the step and the hinge follow by composition,
$\operatorname{sign}(u) \approx s'(u)$, $\mathbb{1}(u>0) \approx (1+s'(u))/2$
and $(u)_+ \approx (u + s(u))/2$.

```{r}
#| label: smoother
sm <- smooth_probit()
u <- c(-1, -0.3, 0, 0.3, 1)
rbind(s = smoother_deriv(sm, u, width = 0.3), abs = abs(u))
```

Three widths out the surrogate and the absolute value agree to four
decimals; at the corner $s(0) = 2h\varphi(0)$, which is the excess the
smoothing buys and the scale on which the bias lives.

Three are shipped. `smooth_probit()` has Gaussian tails, `smooth_hyperbolic()`
is $\sqrt{u^2+c}$, and `smooth_quintic()` is exact outside $[-h,h]$ and $C^3$
within it. The probit earns something the others do not: smoothing with width
$h$ is exactly convolving with $N(0, h^2)$, so a random quantity smoothed this
way has its scale inflated in a known way and the object declares the
correction $\tau^2_{\text{true}} = \tau^2 - h^2$. That is a property of the
mollifier and of no model, which is why it belongs on the smoother; the other
two declare none and a consumer then reports the apparent scale alone.

The width cannot be arbitrarily small, and the floor is derived rather than
chosen. A Jacobian column carries $s''(0)/2 \sim 1/h$ against columns of the
order of the covariate's range $D$, so holding the design's condition number
below $\epsilon^{-1/2}$ requires $h \gtrsim \sqrt{\epsilon}\,D$. A consumer
that resolves a width from the data checks it against that floor.

`check_abs_smoother()` validates one the way `check_link()` validates a link,
comparing each analytic order against one numerical differentiation of the
order below. It earned itself immediately: the quintic's own third derivative
was written with a denominator of 4 where the algebra gives 2, and the check
reported a relative error of 0.5 at order three before anything shipped.

```{r}
#| label: consistency-08b
#| include: false
assert_penalty_branches_ok()
```
