## In practice {#sec-penalty-practice}

```{r}
#| label: setup-07d
#| include: false
source("R/_setup.R")
```

This section assembles a complete penalized fit from the objects the
preceding sections defined, with no fitting function in between: the
likelihood comes from the model, the penalty from `ridge_penalty()`, and the
optimization from `optimizers7`. The example is a Gaussian linear model with
a ridge prior on the coefficients, and the point it demonstrates is
@sec-penalty-definition's: with the normalizing constant kept, the prior
scale $\sigma$ is estimated jointly with the coefficients by ordinary
maximization, with no cross-validation and no separate criterion.

The data are simulated with three active coefficients among eight, and the
error standard deviation is treated as known:

```{r}
#| label: practice-data
set.seed(42)
n <- 120; p <- 8
X <- matrix(rnorm(n * p), n, p)
beta_true <- c(2, -1.5, 1, rep(0, p - 3))
sd_e <- 1.2
y <- drop(X %*% beta_true + rnorm(n, sd = sd_e))
```

The negative log posterior is the negative log-likelihood plus the penalty,
a function of the $p$ coefficients and of the prior scale. The scale is
optimized on its link scale, $\eta = \log\sigma$, and
`penalty_grad_theta(scale = "link")` returns the matching derivative, so the
gradient of the whole objective is assembled from quantities the objects
already provide:

```{r}
#| label: practice-objective
library(penalties7)
library(optimizers7)
pen <- ridge_penalty(n_coef = p)

neg_post <- function(par) {
  b <- par[1:p]; th <- list(sigma = exp(par[p + 1]))
  -sum(dnorm(y, drop(X %*% b), sd_e, log = TRUE)) + penalty_value(pen, b, th)
}
neg_post_gr <- function(par) {
  b <- par[1:p]; th <- list(sigma = exp(par[p + 1]))
  c(-crossprod(X, y - drop(X %*% b)) / sd_e^2 + penalty_gradient(pen, b, th),
    penalty_grad_theta(pen, b, th, scale = "link")$sigma)
}

fit <- minimize(lbfgs(), neg_post, par = rep(0, p + 1), gr = neg_post_gr)
fit@converged
```

The maximizer shrinks the coefficients towards zero relative to least
squares, and the estimated prior scale lands near the root mean square of
the coefficients that generated the data, which is the quantity it
estimates:

```{r}
#| label: practice-results
b_map <- fit@par[1:p]
sigma_hat <- exp(fit@par[p + 1])
b_ols <- drop(solve(crossprod(X), crossprod(X, y)))

round(rbind(true = beta_true, ols = b_ols, ridge = b_map), 3)
c(sigma_hat = sigma_hat, rms_beta_true = sqrt(mean(beta_true^2)))
```

At a fixed prior scale the maximizing coefficients solve the ridge normal
equations with $\lambda = \mathrm{sd}_e^2/\sigma^2$, so the joint maximum
must agree with the closed form evaluated at the estimated scale:

```{r}
#| label: practice-closed-form
lam <- sd_e^2 / sigma_hat^2
b_closed <- drop(solve(crossprod(X) + lam * diag(p), crossprod(X, y)))
max(abs(b_map - b_closed))
```

Dropping the constant from `penalty_value()` would remove the
$q\log\sigma$ term of @sec-penalty-definition, the objective would improve
without bound as $\sigma$ grows, and the joint maximization above would
drive the penalty to zero. The constant is what makes this section's
one-line estimation possible.

A sparsity penalty replaces the smooth objective with one that has a kink
at zero in every coordinate, which `penalty_kinks()` reports; a fit through
`lasso_penalty()` therefore uses the bundle method of @sec-nonsmooth in
place of a gradient method. The marginal-likelihood route for a smoothing
parameter, which integrates the coefficients out instead of maximizing over
them, consumes the pieces of @sec-penalty-marginal and belongs to the model
layer.
