## Special functions {#sec-numerical-special}

```{r}
#| label: setup-07d
#| include: false
source("R/_setup.R")
source("R/numerical-certificates.R")
```

The functions of this section were each born inside a distribution of
@sec-distributions7 and carry the numerical lesson learned there. What
makes them worth a section is not the formulas, which are classical, but
the overflow discipline attached to each: written naively, every one of
them returns `0/0`, `Inf` or an exact zero over a range of arguments that
ordinary statistical work reaches.

### The Mills ratio {#sec-numerical-mills}

Every derivative of a skew normal log-density is built from the inverse
Mills ratio $R(t) = \phi(t)/\Phi(t)$ and its derivative
$R'(t) = -R(t)\{t + R(t)\}$, the latter following from
$\phi'(t) = -t\,\phi(t)$. Below about $t = -38$ both the density and the
distribution function underflow, so the direct quotient is $0/0$ while the
ratio itself is finite and close to $-t$; `mills_ratio()` therefore forms
it on the log scale, as
$\exp\{\log\phi(t) - \log\Phi(t)\}$.

```{r}
#| label: mills
m <- mills_ratio(c(-60, -40, 0))
m$r                        # finite in the deep tail, sqrt(2/pi) at zero
```

### Owen's T function {#sec-numerical-owen}

The skew normal distribution function is
$F(z) = \Phi(z) - 2\,T(z, \alpha)$ with
$$
T(h, a) \;=\; \frac{1}{2\pi} \int_0^{a}
  \frac{e^{-h^2 (1 + x^2)/2}}{1 + x^2}\, dx,
$$
Owen's T function [@owen1956]. The integrand is bounded and smooth over a
finite range, which is a better object to integrate than the density over a
half line, and `owen_t()` sends every element of its input into one batched
`quad_vec()` call. Two identities keep the extremes exact rather than
quadrature-bound: $T(h, -a) = -T(h, a)$, and
$T(h, \infty) = \tfrac{1}{2}\Phi(-\lvert h \rvert)$.

```{r}
#| label: owen
c(owen_t(0, 1) * 2 * pi, atan(1))            # T(0, a) = arctan(a) / (2 pi)
c(owen_t(1.3, Inf), 0.5 * pnorm(-1.3))
```

### The Bessel ratio and its inverse {#sec-numerical-bessel}

The von Mises distribution of @sec-distributions7 is written in
$A(\kappa) = I_1(\kappa)/I_0(\kappa)$, a strictly increasing bijection from
$(0, \infty)$ onto $(0, 1)$ -- for that family, the mean resultant length.
The unscaled Bessel functions overflow past about $\kappa = 700$, so
`bessel_i_ratio()` takes both exponentially scaled, the shared factor
$e^{\kappa}$ cancelling in the ratio. The scaled functions themselves
underflow to an exact zero between $10^5$ and $10^6$, and past
$\kappa = 10^4$ the ratio is taken from its asymptotic expansion
$$
A(\kappa) \;=\; 1 - \frac{1}{2\kappa} - \frac{1}{8\kappa^2}
  - \frac{1}{8\kappa^3} + O(\kappa^{-4}),
$$
whose next term is already below the resolution of a double at the switch
point [@abramowitz1964], so the ratio is finite and accurate at any
argument.

The derivatives follow from the Bessel recurrences $I_0' = I_1$ and
$I_1' = I_0 - I_1/\kappa$, which combine into
$A' = 1 - A/\kappa - A^2$; differentiating that identity repeatedly writes
each order in the orders below it, so the whole table to $A''''$ costs the
two Bessel evaluations of $A$ and nothing more
(`bessel_i_ratio_derivs()`). $A'$ is the variance of a cosine and therefore
positive, which is what keeps the inverse well defined: `bessel_i_ratio_inverse()`
finds $\kappa = A^{-1}(\rho)$ by root finding and differentiates it by the
inverse function rule,
$$
\kappa' = \frac{1}{A'}, \qquad
\kappa'' = -\frac{A''}{(A')^3}, \qquad
\kappa''' = \frac{3(A'')^2 - A'A'''}{(A')^5},
$$
with the fourth order in the same pattern. These derivatives are what let
the von Mises in its mean resultant length carry exact link-scale
derivatives in @sec-distributions7.

```{r}
#| label: bessel
c(bessel_i_ratio(1000),                       # past the unscaled overflow
  bessel_i_ratio(1e8))                        # past the scaled underflow
inv <- bessel_i_ratio_inverse(0.7)
c(kappa = inv$kappa, round_trip = bessel_i_ratio(inv$kappa))
```

```{r}
#| label: consistency-07d
#| include: false
assert_special_ok()
```
