## The term contract {#sec-terms-contract}

```{r}
#| label: setup-08
#| include: false
source("R/_setup.R")
source("R/term-certificates.R")
```

A term has two states. As written in a formula it is a specification: the
object `ridge(R)` returns records what was asked and holds no design
matrix. `term_build(spec, data)` returns the built term, a copy of the
specification carrying the $n \times k$ design block, the $k$ coefficient
names, the penalty over those coefficients (or `NULL` for an unpenalized
term), and a blueprint. The blueprint is the part of the contract that
concerns new data: it records whatever reapplying the mapping requires --
the terms object, the factor levels, the contrasts, and later knots or
centering constants -- so that `term_predict(built, newdata)` reproduces
the mapping exactly instead of re-deriving it from the rows it happens to
receive. A factor level absent from the new data is encoded against the
levels recorded at build time, and a level the blueprint has never seen is
rejected.

```{r}
#| label: contract-linpar
library(modelterms7)

dd <- data.frame(x = rnorm(8), g = factor(rep(c("a", "b", "c", "d"), 2)))
spec <- linpar(~ x + g)
built <- term_build(spec, dd)
term_coef_names(built)

# a subset that loses two levels keeps the encoding of build time
sub <- droplevels(dd[dd$g %in% c("a", "b"), ])
dim(term_predict(built, sub))
```

A consumer reads a built term through a small set of generics:
`term_matrix()` for the block, `term_npar()` and `term_coef_names()` for
its size and labels, `term_penalty()` for the penalty object, and
`term_predict()` for the block on new data. One flag deserves its own
statement. `term_smooth()` says whether the term's contribution to the
penalized objective is differentiable in the coefficients, and the answer
is read from the penalty rather than declared by the term: an unpenalized
term is smooth, and a penalized one is smooth exactly when its penalty's
kink set (@sec-penalties7) is empty. The model layer uses this flag to
split the coefficient vector into the block the optimizers of
@sec-optimizers7 handle on the unconstrained scale and the block that
needs non-smooth strategies, and deriving it from the kink set means a
term cannot disagree with its own penalty.

`check_term()` validates a term against a data frame: that it builds,
that the block's dimensions, names and count agree, that prediction on
the same data reproduces the block exactly, and that prediction on a
subset of rows equals the corresponding rows. The subset is chosen to
drop a factor level whenever the data carry one, and the dropped level is
what gives the check its power: a term that re-derives the encoding from
the new data instead of reusing its blueprint loses a column there, while
a subset that keeps every level would leave the defect invisible.
