## The term contract {#sec-terms-contract}

```{r}
#| label: setup-08
#| include: false
source("R/_setup.R")
source("R/term-certificates.R")
```

A term has two states. As written in a formula it is a specification: the
object `ridge(R)` returns records what was asked and holds no design
matrix. `term_build(spec, data)` returns the built term, a copy of the
specification carrying the $n \times k$ design block, the $k$ coefficient
names, the penalty over those coefficients (or `NULL` for an unpenalized
term), and a blueprint. The blueprint is the part of the contract that
concerns new data: it records whatever reapplying the mapping requires --
the terms object, the factor levels, the contrasts, and later knots or
centering constants -- so that `term_predict(built, newdata)` reproduces
the mapping exactly instead of re-deriving it from the rows it happens to
receive. A factor level absent from the new data is encoded against the
levels recorded at build time, and a level the blueprint has never seen is
rejected.

```{r}
#| label: contract-linpar
library(modelterms7)

dd <- data.frame(x = rnorm(8), g = factor(rep(c("a", "b", "c", "d"), 2)))
spec <- linpar(~ x + g)
built <- term_build(spec, dd)
term_coef_names(built)

# a subset that loses two levels keeps the encoding of build time
sub <- droplevels(dd[dd$g %in% c("a", "b"), ])
dim(term_predict(built, sub))
```

A consumer reads a built term through a small set of generics:
`term_matrix()` for the block, `term_npar()` and `term_coef_names()` for
its size and labels, `term_penalties()` for what the term declares it
wants penalized, and `term_predict()` for the block on new data.

`term_penalties()` returns one entry per penalty, each naming the
parameters of the term it covers and the penalty over them, which is more
than one penalty per term and less than the whole block for each. Both
directions are used: a segmented term penalizes its changes and not its
break-points, a joint one penalizes its slope changes and its jumps under
hyperparameters of their own, and a filter on a panel penalizes the
deviations of its groups while the population values stay free. Where a
term carries a single penalty over the whole of itself, which is every
parametric penalized block, the entry names all of it and
`term_penalty()` returns that penalty directly.

One flag deserves its own statement. `term_smooth()` says whether the
term's contribution to the penalized objective is differentiable in the
coefficients, and the answer is read from the penalties rather than
declared by the term: an unpenalized term is smooth, and a penalized one
is smooth exactly when no penalty it carries has a kink
(@sec-penalties7). The model layer uses this flag to
split the coefficient vector into the block the optimizers of
@sec-optimizers7 handle on the unconstrained scale and the block that
needs non-smooth strategies, and deriving it from the kink set means a
term cannot disagree with its own penalty.

`check_term()` validates a term against a data frame: that it builds,
that the block's dimensions, names and count agree, that prediction on
the same data reproduces the block exactly, and that prediction on a
subset of rows equals the corresponding rows. The subset is chosen to
drop a factor level whenever the data carry one, and the dropped level is
what gives the check its power: a term that re-derives the encoding from
the new data instead of reusing its blueprint loses a column there, while
a subset that keeps every level would leave the defect invisible.
