## The terms {#sec-terms-catalog}

### The parametric block

`linpar()` is the unpenalized block: the design matrix of a one-sided
formula with the usual `model.matrix()` conventions for factors,
contrasts, interactions and the intercept. The formula interpreter of
@sec-terms-formula collects the bare covariates of a model formula into
one term of this kind, so the default model is exactly a GLM design
matrix; the explicit constructor exists for callers who want several
parametric blocks with distinct labels.

### The penalized blocks

`ridge()`, `lasso()`, `scad()` and `mcp()` take their block as a
one-sided formula or as a numeric matrix, and differ only in the
`penalties7` object they attach to its coefficients at build time:
`ridge_penalty()`, `lasso_penalty()`, `scad_penalty()` and
`mcp_penalty()` of @sec-penalties7. The term adds no mathematics of its
own. The hyperparameters, their bounds and links, every derivative in the
coefficients and the hyperparameters, the mixed block and the kink set
are the penalty object's, so the lasso term's hyperparameter is the rate
$\lambda$ of the Laplace prior and the SCAD term carries $(\lambda, a)$
with the bounds the penalty declares.

```{r}
#| label: terms-penalized
dd <- data.frame(x1 = rnorm(20), x2 = rnorm(20), x3 = rnorm(20))
bl <- term_build(lasso(~ x1 + x2 + x3), dd)
term_penalty(bl)@params
c(lasso = term_smooth(bl),
  ridge = term_smooth(term_build(ridge(~ x1 + x2), dd)))
```

A formula input removes the intercept -- a penalized block does not
penalize an intercept, which lives in the parametric block -- and keeps
the blueprint discipline of @sec-terms-contract. A matrix input is used
as given, and prediction re-evaluates the expression that produced it in
the new data alone, so the intended pattern is a matrix column of the
model data frame: a subset of the data then carries the matching rows,
while an expression that resolves only outside the new data is rejected
rather than silently reusing the build-time rows.

### Grouped random effects

`random(~ 1 | g)` builds one coefficient per level of the grouping
factor, and `random(~ x | g)` one intercept and one slope per level, the
bar's left side being an ordinary one-sided formula for the within-group
design with the usual intercept convention. The block interacts that
design with the group indicators, ordered group by group, so with $m$
groups and $d$ within-group columns the coefficients of one group occupy
$d$ adjacent positions.

The distribution of the effects is the penalty on those coefficients,
which is the penalized-likelihood reading of a random effect. By default
the effects are Gaussian with one covariance shared by every group,
carried as a precision through the structured penalty of
@sec-penalties7 on the block-replicated parameter

$$
\Omega_{\mathrm{full}}(\eta) \;=\; I_m \otimes \Omega_d(\eta),
$$

built by `parameters7::kron_identity()`: $m$ identical diagonal blocks
sharing one free vector, with every contract quantity a linear lift of
the per-group parameter's ($\partial_k \Omega_{\mathrm{full}} = I_m
\otimes \partial_k \Omega_d$, $\log\lvert \Omega_{\mathrm{full}}
\rvert = m \log\lvert \Omega_d \rvert$, blockwise solves). The per-group
parameter is the unstructured `log_cholesky(d)` chart of
@sec-parameters7 when `correlated = TRUE`, so intercepts and slopes may
correlate, and `diagonal_matrix(d)` when `correlated = FALSE`; a
`precision` argument replaces either with any `parameters7` matrix
parameter of dimension $d$. With `distrib`, a univariate
`distributions7` object holding its own location is applied
coordinatewise instead, which gives joint-mode estimation of the effects;
integrating them out is a different computation and is not provided here.

```{r}
#| label: terms-random
dd$g <- factor(rep(letters[1:4], 5))
br <- term_build(random(~ x1 | g), dd)
term_coef_names(br)
term_penalty(br)@params
```

### The censored response

`cens(y, lwr = , upr = )` marks a response as censored on the left-hand
side of a model formula. The bounds recycle to the length of the
response, and the status of each observation follows from the values: at
or below the lower bound the observation is left-censored, at or above
the upper bound right-censored, strictly inside observed, and an `NA`
value with two finite bounds interval-censored. The object carries the
values, the bounds and the statuses; the likelihood assembler of the
model layer consumes it, contributing a density where the observation is
exact and a difference of distribution functions -- through the cdf
derivatives of @sec-distributions7 -- where it is censored.

```{r}
#| label: terms-cens
r <- cens(c(0, 0.7, 2.4, NA), lwr = 0, upr = c(5, 5, 5, 4))
r@status
```
