## The terms {#sec-terms-catalog}

### The parametric block

`linpar()` is the unpenalized block: the design matrix of a one-sided
formula with the usual `model.matrix()` conventions for factors,
contrasts, interactions and the intercept. The formula interpreter of
@sec-terms-formula collects the bare covariates of a model formula into
one term of this kind, so the default model is exactly a GLM design
matrix; the explicit constructor exists for callers who want several
parametric blocks with distinct labels.

### The penalized blocks

`ridge()`, `lasso()`, `scad()` and `mcp()` take their block as a
one-sided formula or as a numeric matrix, and differ only in the
`penalties7` object they attach to its coefficients at build time:
`ridge_penalty()`, `lasso_penalty()`, `scad_penalty()` and
`mcp_penalty()` of @sec-penalties7. The term adds no mathematics of its
own. The hyperparameters, their bounds and links, every derivative in the
coefficients and the hyperparameters, the mixed block and the kink set
are the penalty object's, so the lasso term's hyperparameter is the rate
$\lambda$ of the Laplace prior and the SCAD term carries $(\lambda, a)$
with the bounds the penalty declares.

```{r}
#| label: terms-penalized
dd <- data.frame(x1 = rnorm(20), x2 = rnorm(20), x3 = rnorm(20))
bl <- term_build(lasso(~ x1 + x2 + x3), dd)
term_penalty(bl)@params
c(lasso = term_smooth(bl),
  ridge = term_smooth(term_build(ridge(~ x1 + x2), dd)))
```

A formula input removes the intercept -- a penalized block does not
penalize an intercept, which lives in the parametric block -- and keeps
the blueprint discipline of @sec-terms-contract. A matrix input is used
as given, and prediction re-evaluates the expression that produced it in
the new data alone, so the intended pattern is a matrix column of the
model data frame: a subset of the data then carries the matching rows,
while an expression that resolves only outside the new data is rejected
rather than silently reusing the build-time rows.

### Grouped random effects

`random(~ 1 | g)` builds one coefficient per level of the grouping
factor, and `random(~ x | g)` one intercept and one slope per level, the
bar's left side being an ordinary one-sided formula for the within-group
design with the usual intercept convention. The block interacts that
design with the group indicators, ordered group by group, so with $m$
groups and $d$ within-group columns the coefficients of one group occupy
$d$ adjacent positions.

The distribution of the effects is the penalty on those coefficients,
which is the penalized-likelihood reading of a random effect. By default
the effects are Gaussian with one covariance shared by every group,
carried as a precision through the structured penalty of
@sec-penalties7 on the block-replicated parameter

$$
\Omega_{\mathrm{full}}(\eta) \;=\; I_m \otimes \Omega_d(\eta),
$$

built by `parameters7::kron_identity()`: $m$ identical diagonal blocks
sharing one free vector, with every contract quantity a linear lift of
the per-group parameter's ($\partial_k \Omega_{\mathrm{full}} = I_m
\otimes \partial_k \Omega_d$, $\log\lvert \Omega_{\mathrm{full}}
\rvert = m \log\lvert \Omega_d \rvert$, blockwise solves). The per-group
parameter is the unstructured `log_cholesky(d)` chart of
@sec-parameters7 when `correlated = TRUE`, so intercepts and slopes may
correlate, and `diagonal_matrix(d)` when `correlated = FALSE`; a
`precision` argument replaces either with any `parameters7` matrix
parameter of dimension $d$. With `distrib`, a univariate
`distributions7` object holding its own location is applied
coordinatewise instead, which gives joint-mode estimation of the effects;
integrating them out is a different computation and is not provided here.

```{r}
#| label: terms-random
dd$g <- factor(rep(letters[1:4], 5))
br <- term_build(random(~ x1 | g), dd)
term_coef_names(br)
term_penalty(br)@params
```

### Smooth terms

`s(x)` expands a covariate in a `basis7` basis and penalizes it for
roughness. The default construction is the Demmler-Reinsch
reparametrization of @sec-basis7 applied to a cubic B-spline basis, and
what it buys is a decomposition: the transform makes the reparametrized
functions empirically orthogonal to a constant and to $x$, and carries
the roughness penalty onto the identity. The block is therefore one
column for the linear effect followed by the deviation from it, and the
penalty is

$$
P \;=\; \operatorname{diag}(0, 1, \dots, 1),
$$

rank deficient by exactly one. The linear effect is unpenalized and the
deviation is shrunk towards zero, so as the smoothing parameter grows
the fit approaches a straight line rather than a constant, and the
effective degrees of freedom of @sec-terms-edf fall towards one.

```{r}
#| label: terms-smooth
set.seed(11)
dd <- data.frame(x = sort(runif(120)))
dd$y <- sin(2 * pi * dd$x) + rnorm(120, sd = 0.2)

bs <- term_build(s(x, k = 10), dd)
H <- crossprod(term_matrix(bs))
sapply(c(1e-8, 1, 1e8), function(lam)
  edf(bs, coef = rep(0, term_npar(bs)), hessian = H,
      theta = list(lambda = lam)))
```

The three values are the number of columns in the block, an intermediate
fit, and one: the count of a straight line, which is what an unpenalized
linear column inside a penalized block means. The columns are one fewer
than the basis dimension asked for, the reparametrization having spent
the constant and the linear direction on the constraints that separate
them.

`te(x, z)` is the tensor-product smooth of several covariates, built on
`tensor_basis()`. Its penalty is the sum of the marginal roughness
penalties carried into the product,

$$
P \;=\; \sum_{v} I \otimes \cdots \otimes P_v \otimes \cdots \otimes I,
$$

which penalizes curvature in each direction under one smoothing
parameter. A separate parameter per margin would need a penalty that is
a sum of quadratics with its own coefficient on each, whose log
pseudo-determinant is not separable; that construction is not in
`penalties7`.

Both take a `by` argument. With a factor the term carries one copy per
level, the block being the smooth multiplied by each level's indicator
and the penalty the same matrix repeated blockwise, so one smoothing
parameter governs every level. With a numeric the term is a
varying-coefficient one: the smooth multiplies that variable, and the
function being estimated is the coefficient of `by` as it changes with
the covariate.

### Score-driven dynamics

`gas(p, q)` is the first term of the structural branch, and it produces
no block at all. It adds a level $f_t$ to the linear predictor, driven
by the score of the observation density at the previous times,

$$
f_t \;=\; \omega + \sum_{i=1}^{p} a_i\, s_{t-i}
  + \sum_{j=1}^{q} b_j\, f_{t-j},
\qquad s_t = \frac{\partial \ell_t}{\partial \eta_t},
$$

the recursion of @creal2013 and @harvey2013. The predictor at one time
depends on the data at the previous ones, so the contribution cannot be
written as columns and `term_filter()` runs the recursion instead. What
drives it is the score of whatever distribution the model carries, which
is why one term covers several familiar models: entering the scale of a
Gaussian it is a volatility model of the GARCH kind, entering the mean
of a Poisson a dynamic count model, and entering a Student t a robust
filter, the heavy-tailed score being bounded in the observation so that
an outlier moves the level by a bounded amount rather than in proportion
to its size.

The filter returns the derivative of the predictor with respect to the
term's parameters alongside the predictor itself. That is not a
convenience: the recursion is the only place the derivative can be
computed, since each state depends on the previous one, and a model
layer differencing the filter would pay a pass per parameter and inherit
the error of the difference.

The persistence is carried by partial autocorrelations rather than by
the coefficients $b_j$, for the reason @sec-parameters7 gives for an
autoregressive covariance: the stationary region of an autoregression is
not a box, so no collection of scalar links covers it, while the partial
autocorrelations each range over $(-1, 1)$ independently and the
Levinson-Durbin recursion carries them onto the coefficients
bijectively. The coordinate is named for the chart it lives on.

```{r}
#| label: terms-gas
set.seed(8)
ts_dd <- data.frame(t = 1:80, y = as.numeric(arima.sim(list(ar = 0.7), 80)))
gt <- term_build(gas(p = 1, q = 1, time = t), ts_dd)
term_params(gt)

out <- term_filter(gt, eta = rep(0, 80), y = ts_dd$y,
                   score = function(e, i) ts_dd$y[i] - e,
                   curvature = function(e, i) -1,
                   psi = list(omega = 0.1, a1 = 0.4, pacf1 = 0.6))
c(predictor = length(out$eta), jacobian = ncol(out$jacobian))
```

### The censored response

`cens(y, lwr = , upr = )` marks a response as censored on the left-hand
side of a model formula. The bounds recycle to the length of the
response, and the status of each observation follows from the values: at
or below the lower bound the observation is left-censored, at or above
the upper bound right-censored, strictly inside observed, and an `NA`
value with two finite bounds interval-censored. The object carries the
values, the bounds and the statuses; the likelihood assembler of the
model layer consumes it, contributing a density where the observation is
exact and a difference of distribution functions -- through the cdf
derivatives of @sec-distributions7 -- where it is censored.

```{r}
#| label: terms-cens
r <- cens(c(0, 0.7, 2.4, NA), lwr = 0, upr = c(5, 5, 5, 4))
r@status
```
