## The terms {#sec-terms-catalog}

### The parametric block

`linpar()` is the unpenalized block: the design matrix of a one-sided
formula with the usual `model.matrix()` conventions for factors,
contrasts, interactions and the intercept. The formula interpreter of
@sec-terms-formula collects the bare covariates of a model formula into
one term of this kind, so the default model is exactly a GLM design
matrix; the explicit constructor exists for callers who want several
parametric blocks with distinct labels.

### The penalized blocks

`ridge()`, `lasso()`, `enet()`, `scad()` and `mcp()` take their block as
a one-sided formula or as a numeric matrix, and differ only in the
`penalties7` object they attach to its coefficients at build time:
`ridge_penalty()`, `lasso_penalty()`, `elasticnet_penalty()`,
`scad_penalty()` and `mcp_penalty()` of @sec-penalties7. The term adds no
mathematics of its own. The hyperparameters, their bounds and links,
every derivative in the coefficients and the hyperparameters, the mixed
block and the kink set are the penalty object's, so the lasso term's
hyperparameter is the rate $\lambda$ of the Laplace prior, the elastic net's
are $(\lambda, \alpha)$ and the SCAD term carries $(\lambda, a)$
with the bounds the penalty declares.

```{r}
#| label: terms-penalized
dd <- data.frame(x1 = rnorm(20), x2 = rnorm(20), x3 = rnorm(20))
bl <- term_build(lasso(~ x1 + x2 + x3), dd)
term_penalty(bl)@params
term_penalty(term_build(enet(~ x1 + x2 + x3), dd))@params
c(lasso = term_smooth(bl),
  enet = term_smooth(term_build(enet(~ x1 + x2), dd)),
  ridge = term_smooth(term_build(ridge(~ x1 + x2), dd)))
```

A formula input removes the intercept -- a penalized block does not
penalize an intercept, which lives in the parametric block -- and keeps
the blueprint discipline of @sec-terms-contract. A matrix input is used
as given, and prediction re-evaluates the expression that produced it in
the new data alone, so the intended pattern is a matrix column of the
model data frame: a subset of the data then carries the matching rows,
while an expression that resolves only outside the new data is rejected
rather than silently reusing the build-time rows.

### Grouped random effects

`random(~ 1 | g)` builds one coefficient per level of the grouping
factor, and `random(~ x | g)` one intercept and one slope per level, the
bar's left side being an ordinary one-sided formula for the within-group
design with the usual intercept convention. The block interacts that
design with the group indicators, ordered group by group, so with $m$
groups and $d$ within-group columns the coefficients of one group occupy
$d$ adjacent positions.

The distribution of the effects is the penalty on those coefficients,
which is the penalized-likelihood reading of a random effect. By default
the effects are Gaussian with one covariance shared by every group,
carried as a precision through the structured penalty of
@sec-penalties7 on the block-replicated parameter

$$
\Omega_{\mathrm{full}}(\eta) \;=\; I_m \otimes \Omega_d(\eta),
$$

built by `parameters7::kron_identity()`: $m$ identical diagonal blocks
sharing one free vector, with every contract quantity a linear lift of
the per-group parameter's ($\partial_k \Omega_{\mathrm{full}} = I_m
\otimes \partial_k \Omega_d$, $\log\lvert \Omega_{\mathrm{full}}
\rvert = m \log\lvert \Omega_d \rvert$, blockwise solves). The per-group
parameter is the unstructured `log_cholesky(d)` chart of
@sec-parameters7 when `correlated = TRUE`, so intercepts and slopes may
correlate, and `diagonal_matrix(d)` when `correlated = FALSE`; a
`precision` argument replaces either with any `parameters7` matrix
parameter of dimension $d$. With `distrib`, a univariate
`distributions7` object holding its own location is applied
coordinatewise instead, which gives joint-mode estimation of the effects;
integrating them out is a different computation and is not provided here.

```{r}
#| label: terms-random
dd$g <- factor(rep(letters[1:4], 5))
br <- term_build(random(~ x1 | g), dd)
term_coef_names(br)
term_penalty(br)@params
```

### Smooth terms

`s(x)` expands a covariate in a `basis7` basis and penalizes it for
roughness. The default construction is the Demmler-Reinsch
reparametrization of @sec-basis7 applied to a cubic B-spline basis, and
what it buys is a decomposition: the transform makes the reparametrized
functions empirically orthogonal to a constant and to $x$, and carries
the roughness penalty onto the identity. The block is therefore one
column for the linear effect followed by the deviation from it, and the
penalty is

$$
P \;=\; \operatorname{diag}(0, 1, \dots, 1),
$$

rank deficient by exactly one. The linear effect is unpenalized and the
deviation is shrunk towards zero, so as the smoothing parameter grows
the fit approaches a straight line rather than a constant, and the
effective degrees of freedom of @sec-terms-edf fall towards one.

```{r}
#| label: terms-smooth
set.seed(11)
dd <- data.frame(x = sort(runif(120)))
dd$y <- sin(2 * pi * dd$x) + rnorm(120, sd = 0.2)

bs <- term_build(s(x, k = 10), dd)
H <- crossprod(term_matrix(bs))
sapply(c(1e-8, 1, 1e8), function(lam)
  edf(bs, coef = rep(0, term_npar(bs)), hessian = H,
      theta = list(lambda = lam)))
```

The three values are the number of columns in the block, an intermediate
fit, and one: the count of a straight line, which is what an unpenalized
linear column inside a penalized block means. The columns are one fewer
than the basis dimension asked for, the reparametrization having spent
the constant and the linear direction on the constraints that separate
them.

`te(x, z)` is the tensor-product smooth of several covariates, built on
`tensor_basis()`. Its penalty is the sum of the marginal roughness
penalties carried into the product,

$$
P_v \;=\; I \otimes \cdots \otimes P_v \otimes \cdots \otimes I,
$$

each penalizing curvature in one direction. By default those components
keep a smoothing parameter each, through the additive penalty of
@sec-penalty-branches, so the surface may be rough in one direction and
smooth in another; with `anisotropic = FALSE` they are summed first and
one parameter governs the total.

```{r}
#| label: terms-te
d2 <- data.frame(x = runif(150), z = runif(150))
bt <- term_build(te(x, z, k = 4), d2)
term_penalty(bt)@params
```

Anisotropy is the usual reason for fitting a tensor smooth rather than
an isotropic one, and it is what the second parameter buys: a surface
whose curvature in $x$ is penalized separately from its curvature in
$z$.

Both take a `by` argument. With a factor the term carries one copy per
level, the block being the smooth multiplied by each level's indicator
and the penalty the same matrix repeated blockwise, so one smoothing
parameter governs every level. With a numeric the term is a
varying-coefficient one: the smooth multiplies that variable, and the
function being estimated is the coefficient of `by` as it changes with
the covariate.

### Score-driven dynamics

`gas(p, q)` is the first term of the structural branch, and it produces
no block at all. It adds a level $f_t$ to the linear predictor, driven
by the score of the observation density at the previous times,

$$
f_t \;=\; \omega + \sum_{i=1}^{p} a_i\, s_{t-i}
  + \sum_{j=1}^{q} b_j\, f_{t-j},
\qquad s_t = \frac{\partial \ell_t}{\partial \eta_t},
$$

the recursion of @creal2013 and @harvey2013. The predictor at one time
depends on the data at the previous ones, so the contribution cannot be
written as columns and `term_filter()` runs the recursion instead. What
drives it is the score of whatever distribution the model carries, which
is why one term covers several familiar models: entering the scale of a
Gaussian it is a volatility model of the GARCH kind, entering the mean
of a Poisson a dynamic count model, and entering a Student t a robust
filter, the heavy-tailed score being bounded in the observation so that
an outlier moves the level by a bounded amount rather than in proportion
to its size.

The filter returns the derivative of the predictor with respect to the
term's parameters alongside the predictor itself. That is not a
convenience: the recursion is the only place the derivative can be
computed, since each state depends on the previous one, and a model
layer differencing the filter would pay a pass per parameter and inherit
the error of the difference.

The persistence is carried by partial autocorrelations rather than by
the coefficients $b_j$, for the reason @sec-parameters7 gives for an
autoregressive covariance: the stationary region of an autoregression is
not a box, so no collection of scalar links covers it, while the partial
autocorrelations each range over $(-1, 1)$ independently and the
Levinson-Durbin recursion carries them onto the coefficients
bijectively. The coordinate is named for the chart it lives on.

```{r}
#| label: terms-gas
set.seed(8)
ts_dd <- data.frame(t = 1:80, y = as.numeric(arima.sim(list(ar = 0.7), 80)))
gt <- term_build(gas(p = 1, q = 1, time = t), ts_dd)
term_params(gt)

out <- term_filter(gt, eta = rep(0, 80), y = ts_dd$y,
                   score = function(e, i) ts_dd$y[i] - e,
                   curvature = function(e, i) -1,
                   psi = list(omega = 0.1, a1 = 0.4, pacf1 = 0.6))
c(predictor = length(out$eta), jacobian = ncol(out$jacobian))
```

### Nonlinear parametric terms

`nl()` carries a contribution $f(x;\theta)$ that is nonlinear in its own
parameters. It is not a branch of its own, and the reason is worth
stating: while $f$ is differentiable the contribution is linearized as

$$
f(x;\theta(\beta)) \;\approx\; f(x;\theta(\beta_0))
  + J(\beta_0)\,(\beta - \beta_0),
\qquad J = \frac{\partial f}{\partial \beta},
$$

so the design block is the Jacobian and the term is an ordinary additive
one at every point. What distinguishes it is only that the block moves
with the parameters: `term_refresh()` recomputes it, and `term_value()`
reports $f$ itself, which the Jacobian does not give and a Gauss-Newton
step needs beside it.

```{r}
#| label: terms-nl
set.seed(13)
nd <- data.frame(x = seq(0, 3, length.out = 60))
nd$y <- 2 * exp(-1.3 * nd$x) + rnorm(60, sd = 0.05)

bn <- term_build(nl(~ a * exp(-r * x), start = list(a = 1, r = 0.5)), nd)
b <- c(1, 0.5)
for (it in 1:25) {
  cur <- term_refresh(bn, b)
  J <- term_matrix(cur)
  b <- b + solve(crossprod(J) + 1e-10 * diag(2),
                 crossprod(J, nd$y - term_value(cur)))
}
round(as.numeric(b), 3)
```

The function may be given as a formula or as a function of the
covariates and the parameters, and the two do not reach equally far. A
formula is read symbolically: the names it uses that the data do not
supply are the parameters, and the derivatives come from `deriv()` where
that manages the expression and from a central difference where it does
not. An opaque function is always differenced. More than that, a
parameter can be given a link, and developed with covariates as
$\theta_j = g_j^{-1}(Z\gamma_j)$, only on the formula route: substituting
inside $f$ requires knowing where the parameter enters, which a formula
says and a function does not.

### Break-points

`seg()`, `jump()` and `jseg()` carry a covariate whose effect changes at
positions estimated with everything else: a change of slope that keeps
the relationship continuous, a jump in level that does not, and both at
the same positions. They share the device of the previous section, and
extend it to a case in which the contribution is not differentiable in
the quantity being estimated.

The continuous case is differentiable and needs nothing new. With
contribution $\delta (x-\psi)_+$,

$$
\frac{\partial}{\partial \delta}\,\delta (x-\psi)_+ = (x-\psi)_+,
\qquad
\frac{\partial}{\partial \psi}\,\delta (x-\psi)_+
  = -\delta\,\mathbb{1}(x>\psi)
$$

away from the break-point, so the block is the Jacobian in
$(\delta, \psi)$ and $\psi$ is an ordinary coefficient updated by the
Gauss-Newton increment. Those two columns are the working variables $U$
and $V$ of @muggeo2003, and his update
$\psi \leftarrow \psi + \gamma/\delta$ is that increment, so the term
reproduces his algorithm by supplying its block to the same
`term_refresh()` loop that serves `nl()`.

A jump admits no such derivative, and yet needs no search over candidate
positions either. The identity

$$
\mathbb{1}(x>\psi) \;=\; \frac{1}{2}
  + \frac{x-\psi}{2\lvert x-\psi \rvert}
$$

is exact for $x \neq \psi$ and is linear in $\psi$ once the weight
$W = 1/(2\lvert x - \psi^{0}\rvert)$ is held at the previous iterate:
writing $Z = xW + 1/2$, a jump of size $\kappa$ at $\psi$ is
$\kappa Z + g W$ with $g = -\kappa\psi$. A linear fit on $(Z, W)$
therefore returns the break-point as $\psi = -g/\kappa$, and iterating
the pair is the algorithm of @fasola2018. The break-point is not a
coefficient in this construction but a quantity read off two of them,
which is why `seg_psi()` exists as the function that reports it under
either construction.

```{r}
#| label: terms-seg
set.seed(14)
sd <- data.frame(x = sort(runif(500, 0, 10)))
sd$y <- 0.3 * sd$x + 1.5 * pmax(sd$x - 5, 0) + 2 * (sd$x > 5) +
  rnorm(500, sd = 0.3)

fit_jseg <- function(spec, iters = 200) {
  cur <- term_build(spec, sd)
  b <- cur@blueprint$coef
  for (it in seq_len(iters)) {
    cur <- term_refresh(cur, b)
    if (seg_converged(cur)) break
    b <- as.numeric(qr.coef(qr(term_matrix(cur)), sd$y))
  }
  c(psi = seg_psi(cur, b), slope = b[1], change = b[2], jump = b[3])
}
round(rbind(`from 2`    = fit_jseg(jseg(x, psi = 2)),
            `from the grid` = fit_jseg(seg_start(jseg(x), sd, sd$y))), 3)
```

The weight grows without bound as $x$ approaches $\psi$, and since
$Z - \psi W$ is the step itself, $Z$ is $\psi W$ plus a quantity of order
one: letting $W$ grow makes the two columns numerically collinear and
drowns the signal the fit reads. @fasola2018 answer this by moving the
observations rather than by capping the weight. With a scaling factor
$c$ the two intervals on either side of the break-point are shrunk
towards the ends of the range,

$$
[x_{(1)}, \psi] \;\mapsto\; [x_{(1)},\, \psi - c(\psi - x_{(1)})],
\qquad
(\psi, x_{(n)}] \;\mapsto\; (\psi + c(x_{(n)} - \psi),\, x_{(n)}],
$$

which leaves a gap of relative width $c$ around $\psi$, bounds $W$, and
changes no model: the working covariates are computed on the rescaled
covariate while the contribution stays on the original one, exactly as
the additional covariates of the paper's multiple-break-point model do.

The factor is not a constant, and that is what makes the construction an
algorithm rather than a formula. It governs how far the break-point may
travel in one step, so a large value lets the estimate leave a spurious
optimum and a small one is faithful to the step function; the factor is
halved whenever the break-point reverses direction, the signal that the
iteration has begun to circle an optimum instead of travelling towards
one. A break-point outside the data is a separate matter and not merely
ill-conditioned: the indicator is constant there, so the position is
confined to the interval between the fifth and the ninety-fifth
percentile of the covariate.

The two runs above differ only in where they begin, and they end in
different places: the objective has local optima in the break-point, and
the schedule widens the basin the iteration converges from without
removing them. The first run does not fail, it converges -- to a minimum
of its own, on a plateau of the exact profile that is worse than the
global one. Where a run begins therefore decides what it finds, and
`seg_start()` decides it by scoring an equally spaced grid on the
least-squares profile of the term's own columns and returning the
specification with the best of them, which is the initialization
@fasola2018 recommend. `by` gives an independent set of break-points per
level of a factor, and a penalty on the changes -- the slope changes for
`seg()`, the jump sizes for `jump()` -- selects how many break-points
are really there when it is the lasso.

A run is declared finished when no break-point moves by more than a
hundredth of the distance between consecutive observations, below which
the objective of a discontinuous term cannot change. The rule is one of
resolution and tightens as the sample grows, while the precision at
which the fixed point is reached does not, so a caller that can evaluate
the objective is better served stopping on its relative change. That is
also the rule for a continuous term, whose iteration can settle into a
cycle of period two in the break-point long after the fit has stopped
moving.

### Markov regimes

`regime(k)` is the other dynamic term, and it is the complement of the
first: a score-driven level moves continuously, while a regime chain
moves in jumps between a finite number of states. A latent Markov chain
visits $K$ regimes, each shifting the predictor by a level of its own,
and the likelihood is the mixture over the unobserved path, evaluated by
the forward recursion

$$
\tilde\alpha_t(j) \;=\; f_j(t) \sum_i \alpha_{t-1}(i)\, P_{ij},
\qquad c_t = \sum_j \tilde\alpha_t(j),
\qquad \alpha_t = \tilde\alpha_t / c_t,
$$

started at the chain's stationary distribution, with
$\log L = \sum_t \log c_t$. Here $f_j(t)$ is the density of observation
$t$ at the predictor shifted by the level of regime $j$ — whatever
density the model carries, which is why the term is not tied to any
family.

The normalization at each step is not a refinement. The unnormalized
quantities are products of $t$ densities, so they decay geometrically:
on the series below they are of order $10^{-142}$ by the two hundredth
observation and exactly zero in double precision before the five
hundredth, while the normalized recursion is unaffected.

The contribution of this term is a likelihood and not a predictor, so it
cannot be reported through `term_filter()`; `term_loglik()` is the second
shape of the structural branch, returning the contributions $\log c_t$
with their exact derivatives.

```{r}
#| label: terms-regime
set.seed(4)
m <- 400
P <- matrix(c(0.95, 0.05, 0.08, 0.92), 2, 2, byrow = TRUE)
state <- integer(m); state[1] <- 1L
for (t in 2:m) state[t] <- sample(1:2, 1, prob = P[state[t - 1], ])
yy <- rnorm(m, c(0, 4)[state])
rd <- data.frame(t = seq_len(m), y = yy)

rt <- term_build(regime(2, time = t), rd)
nll <- function(v) {
  -sum(term_loglik(rt, rep(0, m), yy,
                   logdens = function(e, i) dnorm(yy[i], e, log = TRUE),
                   score = function(e, i) yy[i] - e,
                   psi = list(level1 = v[1], gap2 = exp(v[2]),
                              alr1.1 = v[3], alr2.1 = v[4]))$loglik)
}
fit <- optim(c(0.5, log(2), 2, -2), nll, method = "BFGS")
round(c(level1 = fit$par[1], level2 = fit$par[1] + exp(fit$par[2])), 2)
```

The levels are ordered by construction: the first is free and each of
the others is the previous one plus a positive gap on a log link.
Without an ordering the regimes are exchangeable, and the likelihood has
$K!$ identical maxima — a problem that is no harder to fit and whose
answer cannot be reported. The transition matrix is the row-wise simplex
chart of @sec-parameters7, so every row is a probability vector at every
value of the free coordinates.

### The censored response

`cens(y, lwr = , upr = )` marks a response as censored on the left-hand
side of a model formula. The bounds recycle to the length of the
response, and the status of each observation follows from the values: at
or below the lower bound the observation is left-censored, at or above
the upper bound right-censored, strictly inside observed, and an `NA`
value with two finite bounds interval-censored. The object carries the
values, the bounds and the statuses; the likelihood assembler of the
model layer consumes it, contributing a density where the observation is
exact and a difference of distribution functions -- through the cdf
derivatives of @sec-distributions7 -- where it is censored.

```{r}
#| label: terms-cens
r <- cens(c(0, 0.7, 2.4, NA), lwr = 0, upr = c(5, 5, 5, 4))
r@status
```
