## The terms {#sec-terms-catalog}

### The parametric block

`linpar()` is the unpenalized block: the design matrix of a one-sided
formula with the usual `model.matrix()` conventions for factors,
contrasts, interactions and the intercept. The formula interpreter of
@sec-terms-formula collects the bare covariates of a model formula into
one term of this kind, so the default model is exactly a GLM design
matrix; the explicit constructor exists for callers who want several
parametric blocks with distinct labels.

### The penalized blocks

`ridge()`, `lasso()`, `enet()`, `scad()` and `mcp()` take their block as
a one-sided formula or as a numeric matrix, and differ only in the
`penalties7` object they attach to its coefficients at build time:
`ridge_penalty()`, `lasso_penalty()`, `elasticnet_penalty()`,
`scad_penalty()` and `mcp_penalty()` of @sec-penalties7. The term adds no
mathematics of its own. The hyperparameters, their bounds and links,
every derivative in the coefficients and the hyperparameters, the mixed
block and the kink set are the penalty object's, so the lasso term's
hyperparameter is the rate $\lambda$ of the Laplace prior, the elastic net's
are $(\lambda, \alpha)$ and the SCAD term carries $(\lambda, a)$
with the bounds the penalty declares.

```{r}
#| label: terms-penalized
dd <- data.frame(x1 = rnorm(20), x2 = rnorm(20), x3 = rnorm(20))
bl <- term_build(lasso(~ x1 + x2 + x3), dd)
term_penalty(bl)@params
term_penalty(term_build(enet(~ x1 + x2 + x3), dd))@params
c(lasso = term_smooth(bl),
  enet = term_smooth(term_build(enet(~ x1 + x2), dd)),
  ridge = term_smooth(term_build(ridge(~ x1 + x2), dd)))
```

A formula input removes the intercept -- a penalized block does not
penalize an intercept, which lives in the parametric block -- and keeps
the blueprint discipline of @sec-terms-contract. A matrix input is used
as given, and prediction re-evaluates the expression that produced it in
the new data alone, so the intended pattern is a matrix column of the
model data frame: a subset of the data then carries the matching rows,
while an expression that resolves only outside the new data is rejected
rather than silently reusing the build-time rows.

### Grouped random effects

`random(~ 1 | g)` builds one coefficient per level of the grouping
factor, and `random(~ x | g)` one intercept and one slope per level, the
bar's left side being an ordinary one-sided formula for the within-group
design with the usual intercept convention. The block interacts that
design with the group indicators, ordered group by group, so with $m$
groups and $d$ within-group columns the coefficients of one group occupy
$d$ adjacent positions.

The distribution of the effects is the penalty on those coefficients,
which is the penalized-likelihood reading of a random effect. The term
says two things and nothing else, the formula and that distribution,
because everything else follows from it: the penalty of @sec-penalties7
reads a `distributions7` density one block at a time,

$$
\rho(b; \theta) \;=\; -\sum_{i=1}^{m} \log f(b_i; \theta),
$$

so the chart the hyperparameters ride, their names and their number, and
where the log-density has a kink are properties of $f$ rather than
declarations of the term.

A multivariate $f$ of dimension $d$ lets the effects of one group depend
on each other, its matrix parameter carrying the dependence; a univariate
one makes them independent, and with several within-group columns it is a
template, one copy per column with its own hyperparameters, an intercept
and a slope being quantities of different units. The default is Gaussian:
`gaussian1_distrib` at one column, so the hyperparameter is the standard
deviation of the effects, and the multivariate Gaussian on an unstructured
covariance when there are several and `correlated = TRUE`. Whatever it is,
$f$ is centered, a free mean in the effects being confounded with the
intercept of the equation the term sits in.

Using a distribution as a penalty gives joint-mode estimation of the
effects. With Gaussian effects and a Gaussian response the Laplace
approximation of @sec-penalty-marginal is exact, so the variance component a
marginal criterion returns is the marginal estimate; with any other prior
it is an approximation, and integrating the effects out is a different
computation that is not provided here.

```{r}
#| label: terms-random
dd$g <- factor(rep(letters[1:4], 5))
br <- term_build(random(~ x1 | g), dd)
term_coef_names(br)
term_penalty(br)@params
```

### Smooth terms

`s(x)` expands a covariate in a `basis7` basis and penalizes it for
roughness. The default construction is the Demmler-Reinsch
reparametrization of @sec-basis7 applied to a cubic B-spline basis, and
what it buys is a decomposition: the transform makes the reparametrized
functions empirically orthogonal to a constant and to $x$, and carries
the roughness penalty onto the identity. The block is therefore one
column for the linear effect followed by the deviation from it, and the
penalty is

$$
P \;=\; \operatorname{diag}(0, 1, \dots, 1),
$$

rank deficient by exactly one. The linear effect is unpenalized and the
deviation is shrunk towards zero, so as the smoothing parameter grows
the fit approaches a straight line rather than a constant, and the
effective degrees of freedom of @sec-terms-edf fall towards one.

```{r}
#| label: terms-smooth
set.seed(11)
dd <- data.frame(x = sort(runif(120)))
dd$y <- sin(2 * pi * dd$x) + rnorm(120, sd = 0.2)

bs <- term_build(s(x, k = 10), dd)
H <- crossprod(term_matrix(bs))
sapply(c(1e-8, 1, 1e8), function(lam)
  edf(bs, coef = rep(0, term_npar(bs)), hessian = H,
      theta = list(lambda = lam)))
```

The three values are the number of columns in the block, an intermediate
fit, and one: the count of a straight line, which is what an unpenalized
linear column inside a penalized block means. The columns are one fewer
than the basis dimension asked for, the reparametrization having spent
the constant and the linear direction on the constraints that separate
them.

`te(x, z)` is the tensor-product smooth of several covariates, built on
`tensor_basis()`. Its penalty is the sum of the marginal roughness
penalties carried into the product,

$$
P_v \;=\; I \otimes \cdots \otimes P_v \otimes \cdots \otimes I,
$$

each penalizing curvature in one direction. By default those components
keep a smoothing parameter each, through the additive penalty of
@sec-penalty-branches, so the surface may be rough in one direction and
smooth in another; with `anisotropic = FALSE` they are summed first and
one parameter governs the total.

```{r}
#| label: terms-te
d2 <- data.frame(x = runif(150), z = runif(150))
bt <- term_build(te(x, z, k = 4), d2)
term_penalty(bt)@params
```

Anisotropy is the usual reason for fitting a tensor smooth rather than
an isotropic one, and it is what the second parameter buys: a surface
whose curvature in $x$ is penalized separately from its curvature in
$z$.

The tensor product of the marginal bases contains the constant, and so
does the null space of the penalty above, so beside an intercept the
block would be rank deficient by exactly one with no penalty covering
the deficiency. The block therefore carries the sum-to-zero constraint
over the observed covariates, which is `constrain_basis()` of
@sec-transform-constrain applied to the product: the term has one
column fewer than the product of its marginal dimensions, every column
sums to zero, and the penalty follows by congruence with its rank
unchanged, the direction removed having been one of its null
directions. The level of the surface is then the model's intercept, as
the level of an `s()` term is.

Both take a `by` argument. With a factor the term carries one copy per
level, the block being the smooth multiplied by each level's indicator
and the penalty the same matrix repeated blockwise, so one smoothing
parameter governs every level. With a numeric the term is a
varying-coefficient one: the smooth multiplies that variable, and the
function being estimated is the coefficient of `by` as it changes with
the covariate.

### Score-driven dynamics

`gas(p, q)` is the first term of the structural branch, and it produces
no block at all. It adds a level $f_t$ to the linear predictor, driven
by the score of the observation density at the previous times,

$$
f_t \;=\; \omega + \sum_{i=1}^{p} \alpha_i\, s_{t-i}
  + \sum_{j=1}^{q} \beta_j\, f_{t-j},
\qquad s_t = \frac{\partial \ell_t}{\partial \eta_t},
$$

the recursion of @creal2013 and @harvey2013, with the parameters written
as the literature writes them and as the term names them: `omega`,
`alpha1`, and, for the persistence, `beta1`. The last is the one name a
coordinate does not carry. The stationary region of an autoregression is
not a box, so no collection of scalar links covers it, and the term is
parametrized by the partial autocorrelations instead, from which
Levinson-Durbin recovers the coefficients; `term_readable()` is what
carries them back, with the Jacobian a standard error for $\beta_j$
needs. At $q = 1$ the two coincide, and above it they do not. The predictor at one time
depends on the data at the previous ones, so the contribution cannot be
written as columns and `term_filter()` runs the recursion instead. What
drives it is the score of whatever distribution the model carries, which
is why one term covers several familiar models: entering the scale of a
Gaussian it is a volatility model of the GARCH kind, entering the mean
of a Poisson a dynamic count model, and entering a Student t a robust
filter, the heavy-tailed score being bounded in the observation so that
an outlier moves the level by a bounded amount rather than in proportion
to its size.

The filter returns the derivative of the predictor with respect to the
term's parameters alongside the predictor itself. That is not a
convenience: the recursion is the only place the derivative can be
computed, since each state depends on the previous one, and a model
layer differencing the filter would pay a pass per parameter and inherit
the error of the difference.

That derivative is what estimating the term's own parameters needs, and
it is not what estimating the coefficients of the same equation needs.
Writing $\eta^{0}_t$ for the static predictor the filter is handed, the
level $f_t$ depends on it at every earlier time, through the scores it
was driven by, so $\partial \eta_t / \partial \beta$ carries a term the
design block does not. Propagating it forward would cost one derivative
array per coefficient. `term_adjoint()` runs the recursion in reverse
instead, which costs one pass whatever their number: writing
$\bar{f}_t$ and $\bar{s}_t$ for the derivatives of the objective in the
level and in the score at time $t$, and taking $t$ downwards,

$$
\bar{e}_t = g_t + \bar{s}_t\, \ell''_t, \qquad
\bar{f}_t \mathrel{+}= \bar{e}_t, \qquad
\bar{f}_{t-j} \mathrel{+}= b_j \bar{f}_t, \qquad
\bar{s}_{t-i} \mathrel{+}= a_i \bar{f}_t,
$$

with $g_t$ the derivative of the objective in the predictor at $t$ taken
directly. The first of those is $\partial L/\partial \eta^{0}_t$.

Two derivatives are returned rather than one, and the second is what
keeps the contract between the term and the model minimal. The score a
caller supplies depends on more than the predictor it is read at: in a
model with several distribution parameters it depends on all of them.
Reporting $\bar{s}_t$ lets the caller write

$$
\frac{\partial L}{\partial \theta}
  = \left.\frac{\partial L}{\partial \theta}\right|_{\text{direct}}
  + \sum_t \bar{s}_t \frac{\partial s_t}{\partial \theta},
$$

so the derivative in another equation's predictor is $\bar{s}_t$ times a
mixed second derivative of the log-density, and the term never has to
know how many equations there are.

The persistence is carried by partial autocorrelations rather than by
the coefficients $b_j$, for the reason @sec-parameters7 gives for an
autoregressive covariance: the stationary region of an autoregression is
not a box, so no collection of scalar links covers it, while the partial
autocorrelations each range over $(-1, 1)$ independently and the
Levinson-Durbin recursion carries them onto the coefficients
bijectively. The coordinate is named for the chart it lives on.

```{r}
#| label: terms-gas
set.seed(8)
ts_dd <- data.frame(t = 1:80, y = as.numeric(arima.sim(list(ar = 0.7), 80)))
gt <- term_build(gas(p = 1, q = 1, time = t), ts_dd)
term_params(gt)

out <- term_filter(gt, eta = rep(0, 80), y = ts_dd$y,
                   score = function(e, i) ts_dd$y[i] - e,
                   curvature = function(e, i) -1,
                   psi = list(omega = 0.1, alpha1 = 0.4, pacf1 = 0.6))
c(predictor = length(out$eta), jacobian = ncol(out$jacobian))
```

Every parameter of the term takes a subformula: a two-sided formula whose
left side names the parameter develops it as

$$
\psi_{j,t} \;=\; g_j^{-1}\!\bigl(z_t^\top \gamma_j\bigr),
$$

with $g_j$ the link of @sec-links that carries the parameter to its
unconstrained scale and $Z$ the design of the right-hand side through the
interpreter of @sec-terms-formula, so the subformula takes any additive
term of the package. The development acts on the unconstrained scale,
which is what keeps every per-observation value in the parameter's own
set whatever the coefficients are: a loading on the log link is positive
at every observation, and a persistence on the rhobit chart stays inside
$(-1, 1)$ at every observation. The coefficients $\gamma_j$ are the
term's parameters, unconstrained on the identity link, and a parameter
that varies by observation changes the recursion itself,
$f_t = \omega_t + \sum_i a_{i,t} s_{t-i} + \sum_j b_{j,t} f_{t-j}$, with
$b_t$ from the Levinson-Durbin map of that observation's partial
autocorrelations.

On a panel, `by` filters each group independently and
`omega ~ random(~1 | id)` gives each group a level of its own: the
intercept of the development is the population value and the indicator
coefficients are unconstrained departures from it, shrunk by the random
intercept's own ridge. The penalties the sub-terms carry are reported
through `term_penalties()` under the key `parameter::subterm`, so a
fitting layer estimates their hyperparameters as it does any other
term's; `lasso(...)` in the subformula sets the departures of the groups
that do not need one exactly to zero.

The penalty is also what identifies the departures. A population value
and $m$ departures are $m+1$ numbers describing $m$ group values, so
adding a constant to the intercept's coordinate and subtracting it from
every departure leaves the filter and its likelihood exactly unchanged,
and the likelihood is flat along one direction per developed parameter.
This is the parametrization of a random effect and is identified in the
same way: there by a variance component, here by the penalty, which
among the descriptions of the same panel selects the departures of
smallest size.

```{r}
#| label: terms-gas-panel
panel <- data.frame(id = factor(rep(c("a", "b", "c"), each = 40)),
                    t = rep(1:40, 3), y = rnorm(120))
gp <- term_build(gas(p = 1, q = 1, omega ~ random(~1 | id),
                     by = id, time = t), panel)
term_params(gp)
lapply(term_penalties(gp), function(e) term_params(gp)[e$index])
```

### Nonlinear parametric terms

`nl()` carries a contribution $f(x;\theta)$ that is nonlinear in its own
parameters. It is not a branch of its own, and the reason is worth
stating: while $f$ is differentiable the contribution is linearized as

$$
f(x;\theta(\beta)) \;\approx\; f(x;\theta(\beta_0))
  + J(\beta_0)\,(\beta - \beta_0),
\qquad J = \frac{\partial f}{\partial \beta},
$$

so the design block is the Jacobian and the term is an ordinary additive
one at every point. What distinguishes it is only that the block moves
with the parameters: `term_refresh()` recomputes it, and `term_value()`
reports $f$ itself, which the Jacobian does not give and a Gauss-Newton
step needs beside it.

```{r}
#| label: terms-nl
set.seed(13)
nd <- data.frame(x = seq(0, 3, length.out = 60))
nd$y <- 2 * exp(-1.3 * nd$x) + rnorm(60, sd = 0.05)

bn <- term_build(nl(~ a * exp(-r * x), start = list(a = 1, r = 0.5)), nd)
b <- c(1, 0.5)
for (it in 1:25) {
  cur <- term_refresh(bn, b)
  J <- term_matrix(cur)
  b <- b + solve(crossprod(J) + 1e-10 * diag(2),
                 crossprod(J, nd$y - term_value(cur)))
}
round(as.numeric(b), 3)
```

The function may be given as a formula or as a function of the
covariates and the parameters, and the two do not reach equally far. A
formula is read symbolically: the names it uses that the data do not
supply are the parameters, and the derivatives come from `deriv()` where
that manages the expression and from a central difference where it does
not. An opaque function is always differenced. More than that, a
parameter can be given a link, and developed with covariates as
$\theta_j = g_j^{-1}(Z\gamma_j)$, only on the formula route: substituting
inside $f$ requires knowing where the parameter enters, which a formula
says and a function does not.

### Break-points

`seg()`, `jump()` and `jseg()` carry a covariate whose effect changes at
$K$ positions estimated with everything else: a change of slope that
keeps the relationship continuous, a change of level that does not, and
both at the same positions. Written into the equation of any parameter of
any distribution, with the rest of that equation supplying
$z_i'\alpha$, the three contribute

$$
\begin{aligned}
\texttt{seg}: \quad \eta_i &= z_i'\alpha + \beta x_i
  + \sum_{k=1}^{K} \gamma_k (x_i - \psi_k)_+, \\
\texttt{jump}: \quad \eta_i &= z_i'\alpha
  + \sum_{k=1}^{K} \delta_k \mathbb{1}(x_i \geq \psi_k), \\
\texttt{jseg}: \quad \eta_i &= z_i'\alpha + \beta x_i
  + \sum_{k=1}^{K} \bigl[\delta_k \mathbb{1}(x_i \geq \psi_k)
      + \gamma_k (x_i - \psi_k)_+\bigr],
\end{aligned}
$$

with $(u)_+ = \max(0, u)$. $\beta$ is the slope before the first
break-point and $\gamma_k$ the change of slope at $\psi_k$, so the slope
over the $k$-th segment is $\beta + \gamma_1 + \cdots + \gamma_k$;
$\delta_k$ is the change of level there. A step model has no linear
effect at all, its relationship with the covariate being a step function
whose level before the first break-point is the equation's intercept.
The three share the device of the previous section and extend it to a
case in which the contribution is not differentiable in the quantity
being estimated.

The continuous case is differentiable and needs nothing new. With
contribution $\gamma (x-\psi)_+$,

$$
\frac{\partial}{\partial \gamma}\,\gamma (x-\psi)_+ = (x-\psi)_+,
\qquad
\frac{\partial}{\partial \psi}\,\gamma (x-\psi)_+
  = -\gamma\,\mathbb{1}(x>\psi)
$$

away from the break-point, so the block is the Jacobian in
$(\gamma, \psi)$ and $\psi$ is an ordinary coefficient updated by the
Gauss-Newton increment. Those two columns are the working variables $U$
and $V$ of @muggeo2003, and his update is that increment, so the term
reproduces his algorithm by supplying its block to the same
`term_refresh()` loop that serves `nl()`.

The term owns $\beta x_i$, which is what makes `y ~ seg(x)` the whole
relationship rather than half of it. Writing the covariate again in the
same equation is then the same column twice, so `interpret_formula()`
removes it and says so; `seg(x, linear = FALSE)` is the way to keep the
linear effect outside the term instead.

A jump admits no such derivative, and yet needs no search over candidate
positions either. The identity

$$
\mathbb{1}(x>\psi) \;=\; \frac{1}{2}
  + \frac{x-\psi}{2\lvert x-\psi \rvert}
$$

is exact for $x \neq \psi$ and is linear in $\psi$ once the weight
$W = 1/(2\lvert x - \psi^{0}\rvert)$ is held at the previous iterate:
writing $Z = xW + 1/2$, a change of level $\delta$ at $\psi$ is
$\delta Z + g W$ with $g = -\delta\psi$. A linear fit on $(Z, W)$
therefore returns the break-point as $\psi = -g/\delta$, and iterating
the pair is the algorithm of @fasola2018. The break-point is not a
coefficient in this construction but a quantity read off two of them,
which is why `seg_psi()` exists as the function that reports it under
either construction.

```{r}
#| label: terms-seg
set.seed(14)
sd <- data.frame(x = sort(runif(500, 0, 10)))
sd$y <- 0.3 * sd$x + 1.5 * pmax(sd$x - 5, 0) + 2 * (sd$x > 5) +
  rnorm(500, sd = 0.3)

fit_jseg <- function(spec, iters = 200) {
  cur <- term_build(spec, sd)
  b <- cur@blueprint$coef
  for (it in seq_len(iters)) {
    cur <- term_refresh(cur, b)
    if (seg_converged(cur)) break
    b <- as.numeric(qr.coef(qr(term_matrix(cur)), sd$y))
  }
  c(psi = seg_psi(cur, b), beta = b[1], gamma = b[2], delta = b[3])
}
round(rbind(`from 2`    = fit_jseg(jseg(x, psi = 2)),
            `from the grid` = fit_jseg(seg_start(jseg(x), sd, sd$y))), 3)
```

The weight grows without bound as $x$ approaches $\psi$, and since
$Z - \psi W$ is the step itself, $Z$ is $\psi W$ plus a quantity of order
one: letting $W$ grow makes the two columns numerically collinear and
drowns the signal the fit reads. @fasola2018 answer this by moving the
observations rather than by capping the weight. With a scaling factor
$c$ the two intervals on either side of the break-point are shrunk
towards the ends of the range,

$$
[x_{(1)}, \psi] \;\mapsto\; [x_{(1)},\, \psi - c(\psi - x_{(1)})],
\qquad
(\psi, x_{(n)}] \;\mapsto\; (\psi + c(x_{(n)} - \psi),\, x_{(n)}],
$$

which leaves a gap of relative width $c$ around $\psi$, bounds $W$, and
changes no model: the working covariates are computed on the rescaled
covariate while the contribution stays on the original one, exactly as
the additional covariates of the paper's multiple-break-point model do.

The factor is not a constant, and that is what makes the construction an
algorithm rather than a formula. It governs how far the break-point may
travel in one step, so a large value lets the estimate leave a spurious
optimum and a small one is faithful to the step function; the factor is
halved whenever the break-point reverses direction, the signal that the
iteration has begun to circle an optimum instead of travelling towards
one. A break-point outside the data is a separate matter and not merely
ill-conditioned: the indicator is constant there, so the position is
confined to the interval between the fifth and the ninety-fifth
percentile of the covariate.

The two runs above differ only in where they begin, and they end in
different places: the objective has local optima in the break-point, and
the schedule widens the basin the iteration converges from without
removing them. The first run does not fail, it converges -- to a minimum
of its own, on a plateau of the exact profile that is worse than the
global one. Where a run begins therefore decides what it finds, and
`seg_start()` decides it by scoring an equally spaced grid on the
least-squares profile of the term's own columns and returning the
specification with the best of them, which is the initialization
@fasola2018 recommend.

Every coefficient of a break-point term is a parameter like any other and
takes a development on covariates, through a two-sided formula naming it:
`seg(x, psi ~ random(~1 | id))` is the random-changepoint model of
@muggeo2014, and `by = ~0 + g` is the shorthand giving every coefficient
an independent value per level of a factor. A penalty on the changes is
the development on a penalized intercept, `seg(x, npsi = 4,
gamma ~ 0 + lasso(~1))`, which selects how many break-points are really
there; the entries of a subformula shared by every coefficient of a kind
are pooled into one block under one hyperparameter, and a penalty names
the coefficients it covers rather than selecting them from the block with
a map, for the reason the panel filter above gives. A joint term takes one
subformula per kind of change: a change of slope and a change of level are
not comparable quantities and cannot share a hyperparameter.

What the discontinuous constructions can carry is narrower, and the reason
is the read-off rather than the model. The auxiliary coefficient is
$g_k = -\delta_k\psi_k$, a product of two of the unknowns, so it stays
linear in them only where one factor is a single number -- a developed
break-point with a shared change of level, exact on any design -- or where
the design has one column per group and each observation belongs to one,
which collapses the product group by group. Anything else is rejected
rather than approximated.

A run is declared finished when no break-point moves by more than a
hundredth of the distance between consecutive observations, below which
the objective of a discontinuous term cannot change. The rule is one of
resolution and tightens as the sample grows, while the precision at
which the fixed point is reached does not, so a caller that can evaluate
the objective is better served stopping on its relative change. That is
also the rule for a continuous term, whose iteration can settle into a
cycle of period two in the break-point long after the fit has stopped
moving.

### Markov regimes

`regime(k)` is the other dynamic term, and it is the complement of the
first: a score-driven level moves continuously, while a regime chain
moves in jumps between a finite number of states. A latent Markov chain
visits $K$ regimes, each shifting the predictor by a level of its own,
and the likelihood is the mixture over the unobserved path, evaluated by
the forward recursion

$$
\tilde\alpha_t(j) \;=\; f_j(t) \sum_i \alpha_{t-1}(i)\, P_{ij},
\qquad c_t = \sum_j \tilde\alpha_t(j),
\qquad \alpha_t = \tilde\alpha_t / c_t,
$$

started at the chain's stationary distribution, with
$\log L = \sum_t \log c_t$. Here $f_j(t)$ is the density of observation
$t$ at the predictor shifted by the level of regime $j$ — whatever
density the model carries, which is why the term is not tied to any
family.

The normalization at each step is not a refinement. The unnormalized
quantities are products of $t$ densities, so they decay geometrically:
on the series below they are of order $10^{-142}$ by the two hundredth
observation and exactly zero in double precision before the five
hundredth, while the normalized recursion is unaffected.

The contribution of this term is a likelihood and not a predictor, so it
cannot be reported through `term_filter()`; `term_loglik()` is the second
shape of the structural branch, returning the contributions $\log c_t$
with their exact derivatives.

```{r}
#| label: terms-regime
set.seed(4)
m <- 400
P <- matrix(c(0.95, 0.05, 0.08, 0.92), 2, 2, byrow = TRUE)
state <- integer(m); state[1] <- 1L
for (t in 2:m) state[t] <- sample(1:2, 1, prob = P[state[t - 1], ])
yy <- rnorm(m, c(0, 4)[state])
rd <- data.frame(t = seq_len(m), y = yy)

rt <- term_build(regime(2, time = t), rd)
nll <- function(v) {
  -sum(term_loglik(rt, rep(0, m), yy,
                   logdens = function(e, i) dnorm(yy[i], e, log = TRUE),
                   score = function(e, i) yy[i] - e,
                   psi = list(level1 = v[1], gap2 = exp(v[2]),
                              alr1.1 = v[3], alr2.1 = v[4]))$loglik)
}
fit <- optim(c(0.5, log(2), 2, -2), nll, method = "BFGS")
round(c(level1 = fit$par[1], level2 = fit$par[1] + exp(fit$par[2])), 2)
```

The levels are ordered by construction: the first is free and each of
the others is the previous one plus a positive gap on a log link.
Without an ordering the regimes are exchangeable, and the likelihood has
$K!$ identical maxima — a problem that is no harder to fit and whose
answer cannot be reported. The transition matrix is the row-wise simplex
chart of @sec-parameters7, so every row is a probability vector at every
value of the free coordinates.

Two quantities beside the likelihood make a regime term fittable, and they
are the answers to two different questions. `term_posterior()` returns the
smoothed probability of each regime at each observation, from the forward
pass with a backward pass beside it. By Fisher's identity the derivative of
the mixed likelihood in any predictor the model carries is the
posterior-weighted derivative of the ordinary one,

$$
\frac{\partial L}{\partial \eta_{q,t}}
  = \sum_k \gamma_t(k)\,
    \frac{\partial \ell(y_t; \theta_t(k))}{\partial \eta_q},
$$

so a model layer differentiates its own log-density $K$ times, vectorized
over observations, and weights the results.

The second derivative does not follow from the same weighting, and the
distinction matters for what a standard error means. Weighting the ordinary
information by the smoothed probabilities gives the complete-data
information, which is the matrix an expectation-maximization step inverts;
it is the information that would be available if the states were observed,
and the states are not. By the missing-information principle the observed
information is smaller, the difference being the conditional variance of the
complete-data score, so a standard error read off the complete-data matrix
is too small. `term_hessian()` returns the observed one. It is obtained by
propagating first and second derivatives through the same scaled forward
recursion that computes the likelihood: writing $c_t$ for the normalizing
constant of one step, $\log L = \sum_t \log c_t$ exactly, so

$$
\frac{\partial^2 \log c_t}{\partial u \partial u^\top}
  = \frac{\ddot c_t}{c_t}
    - \frac{\dot c_t}{c_t}\frac{\dot c_t^\top}{c_t},
$$

and the filtered distribution is renormalized at every step by the quotient
rule applied twice. The cost is $O(nK^2m^2)$ in $n$ observations, $K$
regimes and $m$ unknowns.

### The censored response

`cens(y, lwr = , upr = )` marks a response as censored on the left-hand
side of a model formula. The bounds recycle to the length of the
response, and the status of each observation follows from the values: at
or below the lower bound the observation is left-censored, at or above
the upper bound right-censored, strictly inside observed, and an `NA`
value with two finite bounds interval-censored. The object carries the
values, the bounds and the statuses; the likelihood assembler of the
model layer consumes it, contributing a density where the observation is
exact and a difference of distribution functions -- through the cdf
derivatives of @sec-distributions7 -- where it is censored.

```{r}
#| label: terms-cens
r <- cens(c(0, 0.7, 2.4, NA), lwr = 0, upr = c(5, 5, 5, 4))
r@status
```
