## The marginal pieces and the structured prior {#sec-penalty-marginal}

```{r}
#| label: setup-08c
#| include: false
source("R/_setup.R")
source("R/penalty-certificates.R")
```

A smoothing parameter is not well estimated by maximizing jointly over
coefficients and hyperparameters, even with the constant kept; the
criterion of choice is the restricted or marginal likelihood, in which the
coefficients are integrated out [@wood2011]. For a quadratic penalty that
criterion is assembled from three quantities as functions of $\theta$: the
matrix $\lambda D'PD$ added to the weighted cross-product of the design,
its rank, and its log pseudo-determinant

$$
\log\operatorname{pdet}\!\left(\lambda D'PD\right)
  \;=\; r\log\lambda \;+\; \log\operatorname{pdet}(P),
$$ {#eq-penalty-logpdet}

linear in $\log\lambda$ with slope exactly $r$. The package exposes the
three -- together with the exact null basis, which is what the model layer
intersects across terms when several penalties act on one coefficient
block, the rank of a sum being a statement about intersected null spaces
rather than about eigenvalues of the assembled matrix
(@sec-parameter-rank). Only the quadratic penalties answer these
questions; the separable and derivative-defined ones refuse, because a
marginal criterion for a non-Gaussian prior is not a determinant, and a
base class that pretended otherwise would produce numbers silently.

### The structured prior {#sec-penalty-structured}

In @eq-penalty-quadratic one scale multiplies a constant matrix. The prior
of a random effect is richer: its precision has a structure -- compound
symmetry, an autoregression, a full covariance -- whose parameters reach
every entry of the matrix. This is the correlated Gaussian prior, and it is
the point where `penalties7` composes with @sec-parameters7:

$$
\rho \;=\; \frac{1}{2}\,\beta'\,\Omega(\theta)\,\beta
  \;-\; \frac{1}{2}\log\operatorname{pdet}\,\Omega(\theta)
  \;+\; \frac{r}{2}\log 2\pi,
$$ {#eq-penalty-structured}

with $\Omega$ a `parameters7` matrix parameter read as the precision. The
hyperparameters are the structure's free vector, which is unconstrained by
construction, so every link is the identity -- the flattening convention of
@sec-multivariate, for the same reason: the constraint lives in the
structure, where a scalar link cannot express it. Every derivative comes
from the structure's own contract. Writing $A_k$ for the structure's first
derivative matrices,

$$
\frac{\partial \rho}{\partial \theta_k}
  \;=\; \frac{1}{2}\,\beta' A_k \beta
  \;-\; \frac{1}{2}\,\frac{\partial \log\operatorname{pdet}\Omega}{\partial\theta_k},
\qquad
\frac{\partial^2 \rho}{\partial\beta\,\partial\theta_k} \;=\; A_k\beta,
$$ {#eq-penalty-structured-grad}

and the second $\theta$ derivatives add the structure's $A_{kl}$. Nothing
in the interface changes: the structured penalty answers the marginal
pieces through the structure's rank, null basis and log-determinant
derivatives, so a consumer cannot tell whether the matrix it was handed
came from a constant times a scale or from a chart on the positive
definite cone.

```{r}
#| label: structured-example
pen <- structured_penalty(parameters7::ar1(4, role = "precision"))
theta <- list(log_scale = 0.2, z_rho = 0.5)
beta <- c(0.3, -0.1, 0.4, 0.2)

# two hyperparameters reach every entry of the precision
round(penalty_hessian(pen, beta, theta), 3)

unlist(penalty_grad_theta(pen, beta, theta))
```

At a free vector of zeros the log-Cholesky structure is the identity
matrix, so the structured prior collapses onto the plain ridge at
$\lambda = 1$ -- one more pair of independent constructions of a single
object, compared exactly in the certification below.

```{r}
#| label: consistency-08c
#| include: false
assert_penalty_marginal_ok()
```
