## The formula interpreter {#sec-terms-formula}

`interpret_formula(formula, data)` splits a model formula into a response
specification and a list of term specifications. Its recognition rule is
by evaluation, not by name: each call on the right-hand side is evaluated
in the data, and it becomes a term of its own exactly when its value
inherits from `model_term`. `log(x)` evaluates to a numeric vector and
stays a covariate; `poly(x, 2)` evaluates to a matrix and stays a
covariate; `lasso(L)` evaluates to a term specification and is routed as
one. Bare symbols and interaction labels are never evaluated directly --
`x1:x2` parses as a call to the sequence operator, which on numeric
vectors returns something unrelated to a model matrix column -- and flow,
together with every unrecognized entry, into one collected `linpar()`
block with the formula's own intercept convention.

The consequence of recognizing by value is that the interpreter carries
no list of special names: a term class defined outside the package works
in a formula the day it is written. The demonstration below defines a
minimal term inline and uses it next to the shipped ones.

```{r}
#| label: interpreter-user
MatTerm <- S7::new_class("MatTerm", parent = additive_term, package = NULL,
                         properties = list(mat = S7::class_any))
mat_term <- function(m) {
  MatTerm(label = "mat", mat = as.matrix(m), X = NULL,
          coef_names = character(0), blueprint = list(), penalty = NULL)
}
S7::method(term_build, MatTerm) <- function(term, data, ...) {
  X <- term@mat
  colnames(X) <- paste(term@label, seq_len(ncol(X)), sep = ".")
  term@X <- X
  term@coef_names <- colnames(X)
  term@blueprint <- list(ncol = ncol(X))
  term
}

set.seed(4)
dd <- data.frame(y = rnorm(12), x1 = rnorm(12), x2 = runif(12),
                 g = factor(rep(letters[1:3], 4)))
dd$L <- matrix(rnorm(24), 12, 2)

out <- interpret_formula(y ~ x1 + log(x2) + lasso(L) + mat_term(dd$L) +
                           random(~ 1 | g), dd)
names(out$terms)
```

The left-hand side, when present, is evaluated in the data as well: a
plain expression gives the numeric response and a response constructor
such as `cens()` gives its object. The intercept belongs to the collected
parametric block, so a formula whose right-hand side holds only special
terms still produces an intercept-only `linpar()` and `- 1` removes it.

```{r}
#| label: interpreter-intercept
names(interpret_formula(y ~ lasso(L), dd)$terms)
names(interpret_formula(y ~ lasso(L) - 1, dd)$terms)
```
