## Quadrature and series over many parameters {#sec-numerical-quadrature}

```{r}
#| label: setup-07c
#| include: false
source("R/_setup.R")
source("R/numerical-certificates.R")
```

The integrals a statistical model needs come in families. An expectation
under a fitted distribution is one integral per parameter value, and in a
regression model the parameters vary by observation, so a single quantity
-- a mean, a normalizing constant, an expected information entry -- becomes
hundreds of integrals of one integrand at as many parameter rows. A scalar
integrator called in a loop pays its adaptive machinery once per row.
`quad_vec()` treats the parameter dimension as a matrix dimension instead:
the integrand receives a matrix of evaluation points together with a row
index saying which parameter set each row belongs to, and one call
evaluates every node of every panel of every row.

### The Gauss-Kronrod pair {#sec-numerical-gk}

Each panel is estimated by the 15-point Kronrod extension of the 7-point
Gauss rule: fifteen nodes carrying two sets of weights, so a single matrix
of function values yields both an estimate and, from the difference of the
two rules, an error for it. The embedded Gauss rule is exact on polynomials
to degree 13 and the Kronrod extension to degree 22; the pair is the
workhorse of QUADPACK [@piessens1983], and the Gauss nodes themselves are
the classical construction of @golub1969.

```{r}
#| label: gk-pair
r <- gauss_kronrod15()
c(sum(r$wk), sum(r$wg))          # weights on [-1, 1] sum to its length
sum(r$wg > 0)                    # the Gauss rule uses seven of the fifteen nodes
```

### Batched adaptivity {#sec-numerical-adaptive}

A row converges when the sum of its panel errors fits the budget
$\max(\mathrm{atol}, \mathrm{rtol}\,\lvert I_i \rvert)$; until then its
worst panels are bisected, and the splits from every row join the next
single evaluation, so one hard row refines its own panels without
serializing the others. Judging the sum matters for integrable endpoint
singularities. Near a point where the integrand behaves like
$x^{s-1}$ with $0 < s < 1$ the error is concentrated in the innermost panel
however deep the bisection goes; a rule that gave each panel a share of the
budget proportional to its length would demand of that panel an accuracy no
depth can reach, while the sum converges as soon as the concentrated error
is small enough in absolute terms.

```{r}
#| label: quad-singular
# a beta(0.8, 1.2) density diverges at zero and still normalizes in one call
f <- function(x, i) dbeta(x, c(2, 0.8, 3)[i], c(5, 1.2, 0.7)[i])
quad_vec(f, 0, rep(1, 3))
```

Two failure rules complete the scheme. A row whose panels at the maximum
depth already carry more error than the whole budget can never converge,
whatever happens elsewhere, and fails at once; a row that cannot reach the
budget otherwise fails when nothing is left to split. Either way the row
returns `NA` with a warning naming it, so the failure is explicit.

Infinite endpoints are brought onto a finite interval by the rational maps
$x = a + t/(1-t)$, $x = b - t/(1-t)$ and $x = t/(1-t^2)$, whose Jacobians
multiply the integrand; rows of different kinds share one call.

```{r}
#| label: quad-infinite
shp <- c(0.5, 1, 2, 5, 50)
g <- function(x, i) x * dgamma(x, shape = shp[i], rate = 1)
range(quad_vec(g, 0, rep(Inf, 5)) - shp)   # the means, against the closed form
```

### Series {#sec-numerical-series}

`series_vec()` sums $\sum_{k \ge k_0} t(k; \theta_i)$ for every parameter
row at once, in blocks of terms evaluated as one matrix, with rows retiring
as they converge. A row retires when three conditions hold together: the
last block contributed less than the budget, its final term is itself below
the budget, and the terms are not growing across the block. The third
condition is what keeps a hump-shaped term alive before its mode -- a
Poisson mass at rate 300 opens with a block over $k = 0, \dots, 63$ that
sums to nearly nothing and ends on a term of about $10^{-52}$, so the first
two conditions alone would retire a series that has not begun.

```{r}
#| label: series-hump
lam <- c(0.5, 4, 300)
series_vec(function(k, i) dpois(k, lam[i]), n = 3)
```

```{r}
#| label: consistency-07c
#| include: false
assert_quadrature_ok()
```
