## Degrees of freedom and the fitted views {#sec-terms-edf}

The effective degrees of freedom of a term follow from its penalty, and
`edf()` implements one counting rule per case. An unpenalized term counts
its coefficients exactly. A term with a smooth penalty uses the trace of
the penalized smoother restricted to its block,

$$
\mathrm{edf} \;=\; \operatorname{tr}\!\big[(H + S(\hat\theta))^{-1} H\big],
$$

where $H$ is the term's unpenalized curvature at the fit -- the weighted
crossproduct of its design block -- and $S(\hat\theta)$ the penalty's
Hessian in the coefficients at the estimated hyperparameters. Without a
penalty the trace reduces to the coefficient count, and as the penalty
grows it falls toward the dimension of the penalty's null space, which is
the standard smoother reading [@wood2017]. A term with a non-smooth
penalty counts its nonzero coefficients, which for the lasso is an
unbiased estimator of its degrees of freedom [@zou2007].

The method takes the fit pieces as arguments because no term can know its
edf without them; what the term contributes is the rule, and the choice
of rule is derived from the penalty exactly as `term_smooth()` is.

```{r}
#| label: edf-ridge
set.seed(9)
dd <- data.frame(x1 = rnorm(40), x2 = rnorm(40), x3 = rnorm(40))
br <- term_build(ridge(~ x1 + x2 + x3), dd)
H <- crossprod(term_matrix(br))
beta <- c(0.4, -1, 2)

sapply(c(400, 4, 0.04, 4e-6), function(lambda)
  edf(br, coef = beta, hessian = H, theta = list(lambda = lambda)))
```

The sweep shows the two limits: a hard penalty removes the block and a
vanishing one recovers the count of three. The hyperparameter is a
PRECISION, so the hard end is the large value. For the non-smooth terms the
count is the support:

```{r}
#| label: edf-lasso
bl <- term_build(lasso(~ x1 + x2 + x3), dd)
edf(bl, coef = c(0.8, 0, -0.3))
```

A built term also displays itself. `print()` names the class, the label
and the block, and for a penalized term the penalty and its
hyperparameters; `plot()` draws the term at supplied coefficients, one
needle per coefficient with its name, which is the fitted-point view a
model summary composes term by term.

```{r}
#| label: fitted-plot
#| fig-height: 3.4
plot(br, coef = beta)
```
