## In practice {#sec-terms-practice}

The functions above compose into the workflow the model layer will run:
interpret a formula, build every term, and read the blocks, penalties and
flags off the built objects. The formula below is the shape `statmod()`
will take -- a censored response, bare covariates, a penalized block and
grouped random effects, all in one right-hand side.

```{r}
#| label: practice-interpret
set.seed(12)
n <- 60
dd <- data.frame(x1 = rnorm(n), x2 = runif(n),
                 id = factor(sample(letters[1:6], n, TRUE)))
dd$L <- matrix(rnorm(n * 4), n, 4)
dd$y <- pmax(0, 1 + 0.5 * dd$x1 + rnorm(n))

out <- interpret_formula(cens(y, lwr = 0) ~ x1 + log(x2) + lasso(L) +
                           random(~ x1 | id), dd)
out$response@status[1:8]
names(out$terms)
```

Building every term gives the blocks a fit assembles, and each built term
reports its size, its smoothness and its hyperparameters:

```{r}
#| label: practice-build
built <- lapply(out$terms, term_build, data = dd)

data.frame(
  npar = sapply(built, term_npar),
  smooth = sapply(built, term_smooth),
  theta = sapply(built, function(b) {
    paste(sapply(term_penalties(b), function(e)
      paste(e$penalty@params, collapse = ", ")), collapse = "; ")
  })
)
```

`check_term()` runs the structural battery of @sec-terms-contract on each
specification, and every shipped term passes it on these data:

```{r}
#| label: practice-check
sapply(out$terms, function(sp)
  all(check_term(sp, dd, verbose = FALSE)$status == "OK"))
```

Prediction reapplies each blueprint: the rows of a subset reproduce the
corresponding rows of every block, factor encodings included, which is
what makes fitted values on new data trustworthy before any fitting
exists.

```{r}
#| label: practice-predict
idx <- c(3, 17, 41)
all(vapply(built, function(b)
  isTRUE(all.equal(unname(term_predict(b, dd[idx, , drop = FALSE])),
                   unname(term_matrix(b)[idx, , drop = FALSE]),
                   tolerance = 1e-12, check.attributes = FALSE)),
  logical(1)))
```

```{r}
#| label: terms-gate
#| include: false
assert_terms_ok()
```
