## In practice {#sec-numerical-practice}

```{r}
#| label: setup-08e
#| include: false
source("R/_setup.R")
```

### Batched quadrature and series {#sec-numerical-batched-practice}

`quad_vec()` evaluates one integrand at many parameter rows in a single
adaptive pass, and `series_vec()` does the same for sums. Four Gaussian
mean absolute deviations in one call, against the closed form
$\sigma\sqrt{2/\pi}$:

```{r}
#| label: np-quad
library(numericals7)
sig <- c(0.5, 1, 2, 10)
mad_g <- quad_vec(function(x, i) abs(x) * dnorm(x, 0, sig[i]),
                  lower = rep(-Inf, 4), upper = rep(Inf, 4))
rbind(quad_vec = mad_g, closed = sig * sqrt(2 / pi))

# three negative binomial means summed as series, rows retiring as they converge
mu <- c(2, 15, 80)
series_vec(function(k, i) k * dnbinom(k, size = 5, mu = mu[i]), n = 3)
```

### The stencil library {#sec-numerical-stencil-practice}

`fd_weights()` solves the Vandermonde system of @sec-numerical-stencils for
any derivative order at any offsets, and `fd_derivative()` applies the
resulting rule with the balanced step:

```{r}
#| label: np-stencils
fd_weights(-2:2, 2L) * 12          # the five-point second-derivative weights
fd_offsets(3L, accuracy = 2L)      # the grid a third derivative uses

# lgamma and its first four derivatives at x = 2.5, against the polygammas
x <- 2.5
est <- vapply(1:4, function(m) fd_derivative(lgamma, x, order = m), numeric(1))
ref <- c(digamma(x), trigamma(x), psigamma(x, 2), psigamma(x, 3))
signif(abs(est - ref), 2)
```

### The special functions at their extremes {#sec-numerical-special-practice}

Each special function exists for an argument range where the textbook
expression fails; the table evaluates both routes at such a point:

```{r}
#| label: np-special
naive_mills <- function(t) dnorm(t) / pnorm(t)
naive_logi0 <- function(x) log(besselI(x, 0, expon.scaled = TRUE)) + x

rbind(
  c(naive = naive_mills(-40), numericals7 = mills_ratio(-40)$r),
  c(naive = naive_logi0(5e5), numericals7 = log_bessel_i(5e5, 0)),
  c(naive = log(besselK(1500, 2)), numericals7 = log_bessel_k(1500, 2))
)

c(bessel_i_ratio(1e8),                       # past the scaled underflow
  bessel_i_ratio_inverse(0.7)$kappa)         # and the inverse map
c(owen_t(1.3, Inf), 0.5 * pnorm(-1.3))       # the closed tail identity
```

### The enumerations {#sec-numerical-enum-practice}

The chain rules of @sec-link-scale and the Bartlett identities enumerate
set partitions, and the multinomial support of @sec-mv-simplex enumerates
weak compositions; both enumerations live here, written once:

```{r}
#| label: np-enum
length(set_partitions(4))          # the Bell number B4
compositions(3, 2)                 # the support of a multinomial(3) on 2 cells
```
