## The model {#sec-statmod-model}

```{r}
#| label: setup-09
#| include: false
source("R/_setup.R")
source("R/statmod-certificates.R")
```

A response distribution with $K$ parameters is written here with an equation
for each of them. Observation $i$ has its own parameter vector
$\theta_i = (\theta_{i1}, \ldots, \theta_{iK})$, and each coordinate is
carried onto the whole real line by that parameter's own link, exactly as
@sec-link-scale carries it for a single distribution:

$$
g_k(\theta_{ik}) \;=\; \eta_{ik} \;=\; o_{ik} + x_{ik}'\beta_k ,
\qquad k = 1, \ldots, K .
$$

Here $x_{ik}'$ is the $i$th row of the $k$th equation's design $X_k$, which
is the blocks of the terms that equation names set side by side, and
$\beta_k$ the coefficients they carry; $o_{ik}$ is an offset, a column of the
predictor whose coefficient is known to be one. The log-likelihood is the
weighted sum of the family's own,

$$
\ell(\beta) \;=\; \sum_{i=1}^{n} w_i \, \log f\!\left(y_i; \theta_i(\beta)\right),
$$

with the prior weights $w_i$ taken as given and never normalized. Dividing by
their sum would turn the log-likelihood into a mean, which shrinks every
standard error by $\sqrt{n}$ and makes an information criterion incomparable
with the unweighted fit of the same model.

Writing an equation for every parameter rather than for the mean alone is the
distributional regression of @rigby2005; what differs here is that the terms
those equations may name are the whole of @sec-modelterms7 rather than a fixed
catalog, and that the parameters they are written for are whatever the family
of @sec-distributions7 declares.

### The formula {#sec-statmod-formula}

The equations are separated by a bar, the first carrying the response:

``` r
y ~ x1 + s(x2) + ridge(R)  |  sigma ~ z  |  nu ~ 1
```

The first equation belongs to the family's first parameter and each bar
introduces the next. The parameters may be named in any order, and one with
no equation of its own is given `~ 1`, an intercept, so a formula with no bar
at all fits an ordinary regression with every other parameter constant.

Recovering those equations is not a matter of splitting the right-hand side.
R's `~` binds looser than `|` and associates to the left, so

``` r
y ~ a | sigma ~ b | nu ~ c
```

parses as `((y ~ (a | sigma)) ~ (b | nu)) ~ c`, whose right-hand side is `c`
alone: splitting it on `|` returns one piece and drops two equations without
an error. What the associativity calls for is a walk down the left spine of
the nested `~` calls, collecting each level's right-hand side and reversing
at the end, with the innermost left operand as the response. The walk
descends through `~` and through a `|` at the top of a collected piece, and
through nothing else, so a bar written inside a call is left alone and
`random(1 | id)` reaches its constructor whole.

```{r}
#| label: equations
eq <- statmod_equations(y ~ a | sigma ~ b | nu ~ cc,
                        c("mu", "sigma", "nu"))
deparse(eq$response)
vapply(eq$equations, function(q) deparse(q[[2]]), character(1))
```

The same call on a formula that names one parameter fills the others in:

```{r}
#| label: equations-filled
vapply(statmod_equations(y ~ x, c("mu", "sigma", "nu"))$equations,
       function(q) deparse(q[[2]]), character(1))
```

Each equation is then read by `modelterms7`'s interpreter, whose recognition
rule is the one of @sec-terms-formula: a call whose value inherits from the
term class is a term, and everything else is a covariate, so the bare
covariates of an equation collapse into a single parametric block. The
interpretation happens in an environment where the term constructors sit in
front of the search path, which is what makes `s()` this toolkit's smooth
even in a session with `mgcv` attached.

An `offset()` written among an equation's terms is taken out before the
interpreter sees the rest, and its expression is kept rather than its value,
which is what lets an offset survive prediction at new data. One written
inside a term's own formula is rejected rather than moved: in a penalized
term's formula it would be a contribution to the equation's predictor, and in
a subformula a contribution to that parameter's own chart, and choosing one
reading would fit the wrong model in half the cases.

### The specification {#sec-statmod-spec}

`statmod_spec()` performs everything above and stops short of fitting: it
splits the formula, interprets each equation, and builds the terms against
the data, giving one object that carries the equations, the built terms, the
response, the weights and the offsets.

```{r}
#| label: spec
set.seed(9)
dd <- data.frame(x = runif(120, -2, 2), z = runif(120, -2, 2))
dd$y <- rnorm(120, 1 + 0.8 * dd$x, exp(-0.5 + 0.3 * dd$z))

spec <- statmod_spec(y ~ x + s(z, k = 6) | sigma ~ z,
                     gaussian1_distrib(), dd)
names(spec@equations)
vapply(spec@terms, function(tm) paste(names(tm), collapse = ", "),
       character(1))
```

`statmod_design()` assembles from it, per parameter, the terms' blocks side
by side, together with the names of the coefficients they carry and the range
of columns each term occupies:

```{r}
#| label: design
des <- statmod_design(spec)
vapply(des, function(e) e$npar, integer(1))
des$mu$coef_names
```

The mean's equation carries an intercept and a slope from its parametric
block, and from the smooth a linear column together with the reparametrized
basis of @sec-transform-dr, one column fewer than the basis dimension it was
given; the scale's equation carries an intercept and a slope. The block a
term contributes is sparse whenever the term builds it
that way -- a grouping indicator has density $1/m$ over $m$ groups -- and an
equation's design is sparse when any one of its blocks is, sparsity being a
property of the assembled matrix and never of its pieces.

A fitted model keeps this specification rather than the data. That is what
makes prediction at new data a matter of reapplying each term's blueprint,
with the knots of a smooth and the levels of a factor the ones the model was
fitted with, instead of rebuilding a basis against rows the fit never saw.

### The intercept and a structural level {#sec-statmod-intercept}

Most terms add columns to a predictor. A structural term of
@sec-terms-catalog rewrites the likelihood instead, and one of them, the
score-driven filter, contributes a level that evolves with the data:

$$
f_t \;=\; \omega + \sum_{r=1}^{p} a_r\, s_{t-r} + \sum_{j=1}^{q} b_j\, f_{t-j} ,
$$

added to whatever the equation's design already produces. A constant in that
same equation and the level's own $\omega$ are then exactly confounded.
Replacing the intercept by $\alpha + c$ and $\omega$ by
$\omega - c\,(1 - \sum_j b_j)$ leaves the recursion's stationary level lower
by exactly $c$ and every predictor unchanged, because the recursion is affine
in the level given the score path and the score path depends on the predictor
alone. A fit written that way is not identified: moving along that
direction changes neither a predictor nor the likelihood.

Where an equation's design spans the constant and it also carries a
structural term, the intercept keeps the level and the term's own parameter
is held at zero. What is given up by holding $\omega$ is a quantity the
intercept already reports, while the loadings and the persistence, which a
constant cannot express, stay free. The question is asked of the span of the
equation's design rather than of a column called `(Intercept)`, since a
factor coded without one still sums to the constant.
