## Choosing the hyperparameters {#sec-statmod-outer}

A penalty carries hyperparameters: a smoothing parameter, the scale of a
prior, the $\lambda$ and $\alpha$ of an elastic net. A term holds any of them
it was given a value for, and every one left free is estimated from the same
data the coefficients are. That is a second estimation problem sitting
outside the first, and the model is refitted at every value it visits.

Two families of criterion answer it, and they ask different questions. A
marginal criterion integrates the coefficients out and asks which
hyperparameter makes the data most likely; a prediction-error criterion asks
which one predicts best. For a smooth penalty either applies. For a penalty
with a kink only the second does, and the reason is the subject of the last
part of this section.

### The marginal criterion {#sec-statmod-marginal}

The coefficients are integrated out by a Laplace approximation at the
penalized mode. Writing $\hat\beta(\theta)$ for that mode, $H$ for the
information of the log-likelihood, $S$ for the penalty's second derivative in
the coefficients and $A$ for an orthonormal basis of the subspace integrated
over,

$$
\log L(\theta) \;=\; \ell(\hat\beta) - \rho(\hat\beta; \theta)
  + \frac{q}{2}\log 2\pi - \frac12 \log\bigl\lvert A'(H+S)A \bigr\rvert ,
$$

with $q$ the dimension of that subspace. Nothing is added to $\rho$ to make
this work. A `penalties7` penalty keeps its normalizing constant, as
@sec-penalty-definition sets out, so $\rho$ is exactly minus a log prior
density, and for a quadratic penalty that constant carries the
$-\frac{r}{2}\log\lambda$ and the log pseudo-determinant of
@sec-penalty-marginal that a marginal criterion needs. Written out, the
expression reproduces the REML criterion of @wood2011 term for term.

The two members differ in what they integrate. `reml()` takes $A = I$: every
coefficient is integrated, the unpenalized ones under the flat prior their
absence of a penalty amounts to. `ml()` takes $A$ spanning the range of the
penalty, so an unpenalized direction is profiled instead — an ordinary
covariate is one, and so is the linear column of a Demmler-Reinsch smooth,
which that smooth's penalty leaves alone. This is the distinction between
REML and ML for a variance component, and `reml()` is the default for the
same reason: profiling a fixed effect leaves the variance estimate biased
downwards.

```{r}
#| label: reml-ml
set.seed(11)
n_out <- 250
d_out <- data.frame(z = sort(runif(n_out, -3, 3)))
d_out$y <- sin(1.4 * d_out$z) + rnorm(n_out, sd = 0.4)

f_reml <- statmod(y ~ s(z, k = 8), gaussian1_distrib(), d_out)
f_ml   <- statmod(y ~ s(z, k = 8), gaussian1_distrib(), d_out,
                  outer_criterion = ml())

c(reml = hyper(f_reml)$estimate, ml = hyper(f_ml)$estimate)
c(reml = sum(f_reml@edf$edf), ml = sum(f_ml@edf$edf))
```

The unpenalized directions here are the two intercepts and the smooth's
linear column, three of the model's nine coefficients, the Demmler-Reinsch
penalty being $\operatorname{diag}(0, 1, \ldots, 1)$ and leaving that column
alone. `ml()` profiles them rather than integrating them, so it shrinks
slightly less, and the gap widens with the number of such directions.

### The exact gradient {#sec-statmod-outer-gradient}

The criterion is differentiable in the hyperparameters, and its derivative is
available in closed form rather than by differencing. Every piece comes from
the implicit function theorem at the penalized mode: differentiating the
mode's own condition $\partial(-\ell + \rho)/\partial\beta = 0$ in a
hyperparameter gives

$$
(H + S)\,\frac{\partial\hat\beta}{\partial\theta_k}
  \;=\; -\frac{\partial^2\rho}{\partial\beta\,\partial\theta_k} ,
$$

and the determinant's derivative follows from the same solve. What the
gradient needs of the model is that the information be the observed one and
that each penalty's Hessian be linear in its hyperparameters, which covers
`s()`, `te()` and any quadratic penalty. Where it exists it is handed to the
search and `lbfgs()` becomes the default optimizer; where it does not, the
search compares values.

Measured in evaluations of the criterion, each of them a whole refit of the
coefficients, against a derivative-free search: 40 against 32 with one
smoothing parameter, 40 against 135 with two, 41 against 269 with three, and
12 against 283 with three and a modeled scale. It does not pay in one
dimension and pays from two on, a simplex needing a vertex per dimension
where a quasi-Newton method does not.

### Prediction error {#sec-statmod-pe}

The other family scores an estimate of how well the fit would predict:

$$
C(\theta) \;=\; -2\,\ell(\hat\beta(\theta)) + \kappa\,\tau(\theta),
\qquad \tau(\theta) = \operatorname{tr}\bigl[(H+S)^{-1}H\bigr] ,
$$

minimized, with $\kappa$ the price of one degree of freedom: `aic()` charges
2 and `bic()` charges $\log n$. `cv()` scores the deviance on held-out folds
instead, refitting on each.

These are differentiated exactly as well, and out of the same pieces, but one
thing separates the two families. The envelope theorem does not apply here:
$\ell$ alone is not stationary at the penalized mode, only the penalized
objective is, so $\mathrm{d}\ell/\mathrm{d}\theta$ carries
$\mathrm{d}\hat\beta/\mathrm{d}\theta$ from the first order. What makes that
computable is the mode's own condition again,
$\partial\ell/\partial\beta = \partial\rho/\partial\beta$.

Generalized cross-validation is not among them. The classical form divides a
residual sum of squares by $(n - \tau)^2$, which estimates an unknown scale;
here every distribution parameter carries its own equation and nothing is
profiled, so there is no unknown scale for that ratio to estimate and the
criterion degenerates to `aic()`.

### A penalty with a kink is swept, not solved {#sec-statmod-path}

A marginal criterion expands around the penalized mode. For a lasso that mode
sits *at* the kink for every coefficient the penalty sets to zero, and the
second derivative the expansion wants does not exist there. Such a
hyperparameter is therefore not estimated by `reml()` or `ml()` at all; it is
swept along a path of its own values and scored by a prediction-error
criterion, `bic()` by default.

The path is geometric in the **size of the kink** rather than in the
hyperparameter, since the two need not be the same quantity: the size is read
from the penalty by probing its subdifferential, which covers an elastic
net's $\lambda\alpha$ and a Laplace prior written by its scale, whose kink
narrows as the hyperparameter grows. It starts at the value that empties the
block and descends to a fraction of it, and that top depends on the data and
on the rest of the model, being found at the coefficients in hand rather than
at a refitted null.

The reason for a path rather than a search is the shape of the problem. The
penalized mode is only piecewise smooth in such a hyperparameter, turning a
corner whenever a coefficient joins the active set or leaves it, so the
criterion inherits the corners and a gradient search would read a slope that
is about to change.

```{r}
#| label: lasso-path
set.seed(12)
n_l <- 200
X <- matrix(rnorm(n_l * 12), n_l, 12)
d_lasso <- data.frame(y = X %*% c(1.4, -0.9, 0.7, rep(0, 9)) + rnorm(n_l, sd = 1))
d_lasso$X <- X

f_lasso <- statmod(y ~ lasso(X), gaussian1_distrib(), d_lasso)
hyper(f_lasso)[, c("name", "estimate", "held", "source")]
sum(coef(f_lasso)$mu != 0)
```

Of the twelve columns offered, the chosen point keeps the three that carry
signal and the unpenalized intercept beside them.

Where the penalty has a kink, $\tau$ is restricted to the coordinates away
from it, so for a lasso it is the number of surviving coefficients, which is
the unbiased count of @zou2007. No special case is written for it: the
penalty is linear away from the kink, so its second derivative vanishes there
and the trace reduces to that number on its own.

### What the count leaves out {#sec-statmod-edf}

$\tau = \operatorname{tr}[(H+S)^{-1}H]$ reads the smoothing parameters as
though they were known, and they were not: they were chosen from the same
data, so a criterion built on the count is too generous. Propagating their
uncertainty into the coefficients gives the corrected variance and count

$$
V' \;=\; V_\beta + J V_\theta J', \qquad
J = \frac{\partial\hat\beta}{\partial\theta}, \qquad
\tau' = \operatorname{tr}(V' H),
$$

with $J$ the solve above and $V_\theta$ the inverse of the negated Hessian of
the outer criterion, everything on the hyperparameter's link scale, which is
where that criterion optimizes and therefore the only scale on which its
Hessian is a variance. This is the first of the two terms @woodpya2016 sum
into their corrected count; the second exists there because a Gaussian scale
is profiled out of the fit and its uncertainty has to be added back, and here
that scale carries an equation and coefficients of its own, so its
uncertainty is already inside $H$.

For a kinked penalty the correction is exactly zero, and that is a refusal
rather than an approximation: the map from the hyperparameter to the
penalized mode turns a corner whenever the active set changes, and a delta
method needs a derivative that is not there.

The same distinction governs what a hyperparameter's own uncertainty can be.
One a marginal criterion maximized is the root of a derivative, so the
curvature of that criterion is a variance and `summary()` prints a standard
error and an interval, built on the free scale and mapped back. One chosen
along a path is the argument of a minimum over a grid, and no standard error
follows from it. `hyper()`'s `source` column is what tells the two apart.
