## Structural terms in the fit {#sec-statmod-structural}

Most terms of @sec-modelterms7 add columns to a design. A structural term
rewrites the likelihood instead, and two shapes of it are fitted here. A
**filter** produces a level of its own, evolving with the data, which is
added to the predictor of the equation it sits in. A term of the **mixture**
shape reports no predictor at all: its contribution is a likelihood averaged
over latent states, and the model's log-likelihood comes from the term rather
than from the density at any single point.

At most one such term is allowed in a formula, whatever equations they would
sit in. Two filters in one equation are two levels adding up, which is one
level written twice; in two equations each would be driven by a score that
depends on the other's state, which is one recursion written as two and
implemented by neither.

### A filter inside the system {#sec-statmod-filter}

The level of @sec-statmod-intercept,
$f_t = \omega + \sum_r a_r s_{t-r} + \sum_j b_j f_{t-j}$, is driven by scores
read at the predictors the coefficients also enter. So the derivative of the
log-likelihood in a coefficient is not the design column times the score: the
level at time $t$ carries the coefficient through every earlier step of the
recursion.

Propagating that forward would cost one derivative array per coefficient. It
is taken backwards instead, in one pass, and what the term returns is two
things: the derivative of the objective in the static predictor, and the
derivative in the **score sequence**. The second is what keeps the contract
small. The term knows its recursion and the model knows its likelihood, so
multiplying the second by a mixed second derivative of the log-density gives
the derivative in another equation's predictor without the term ever
learning that there are several.

The curvature is a separate matter and was for a while the weaker half. The
matrix a scoring step inverts is assembled as though the level were an
offset, which is a legitimate matrix to take a step with and is not the
information. Writing $u$ for the coefficients followed by the term's own
parameters, $V_{q,t}$ for the derivative of equation $q$'s predictor in $u$,
and $E_t$ for the second derivative of the level,

$$
-\frac{\partial^2 \ell}{\partial u\, \partial u'}
  \;=\; -\sum_t w_t \sum_{q,r} \ell_{qr,t}\, V_{q,t}'\,V_{r,t}
     \;-\; \sum_t w_t\, \ell_{p,t}\, E_t .
$$

Only the equation carrying the filter has a $V$ that is not simply its own
design; there it is the forward Jacobian of the recursion, returned beside
the contracted $E$. The third derivatives the second sum needs are the
family's, in closed form for every one of them, so the exact curvature of a
score-driven model rests on the third-order surface of @sec-higher-orders.
`vcov()` inverts that joint matrix and returns the coefficient block, which
is why a standard error here is not the one a fit with the level frozen
would report.

```{r}
#| label: gas-fit
set.seed(21)
n_gas <- 600
om <- 0.3; a1 <- 0.4; b1 <- 0.7
lev <- numeric(n_gas); yg <- numeric(n_gas); lev[1] <- om / (1 - b1)
for (t in seq_len(n_gas)) {
  yg[t] <- rnorm(1, lev[t], 1)
  if (t < n_gas) lev[t + 1] <- om + a1 * (yg[t] - lev[t]) + b1 * lev[t]
}

f_gas <- statmod(y ~ gas(p = 1, q = 1) - 1, gaussian1_distrib(),
                 data.frame(y = yg))
round(f_gas@structural[[1]]$parameter, 4)
```

against $\omega = 0.3$, $a_1 = 0.4$ and $b_1 = 0.7$. The persistence is
reported here as `pacf1` because that is the coordinate the fit runs on: the
stationary region of an autoregression is not a box, so the persistence rides
a partial autocorrelation chart and no collection of scalar links would keep
it inside. At $q = 1$ the coordinate and the coefficient coincide; above it
they do not, and the term reports the coefficients under the literature's
names separately.

The `- 1` is not decoration. Written with an intercept the equation would
carry a constant and a level that are exactly confounded, and
@sec-statmod-intercept is where that is set out; the fit would hold the
term's $\omega$ at zero and let the intercept carry it.

### A likelihood mixed over states {#sec-statmod-regime}

A latent Markov term gives each observation a level drawn from one of $K$
regimes, the regime following a chain of its own. What the model maximizes is
the mixed likelihood, computed by a normalized forward recursion, and every
derivative follows from **Fisher's identity**: the derivative of that mixed
likelihood in any predictor the model carries is the posterior-weighted
derivative of the ordinary one. So the layer differentiates its own
log-density once per state, vectorized over the sample, and weights by the
smoothed state probabilities. No second recursion is needed for the gradient,
and the predictor reported for the equation is the posterior-weighted one,
which is what a fitted value means here.

The observed information is not Louis's identity. That route writes the
observed information as the complete-data one minus the variance of the
complete-data score, and the second term involves every pair of times, which
the smoothed marginals do not give. What is used instead is second-order
propagation through the recursion that already computes the likelihood:
$\log L = \sum_t \log c_t$ exactly, so differentiating that arithmetic twice
gives the observed Hessian with no identity invoked and no pairwise smoothed
probabilities. Louis's relation becomes the check rather than the method, the
missing-information principle requiring the complete-data information to
dominate the observed one, with the gap going to zero as the regimes
separate and the states become effectively observed.

The distinction is not academic. The complete-data matrix is what a scoring
step inverts, and it is positive definite and legitimate for that; used as a
variance it understates a standard error by up to 30 per cent, and by most in
the middle of the range, where the states are neither well separated nor
hopelessly overlapped.

```{r}
#| label: regime-fit
set.seed(22)
n_reg <- 400
state <- numeric(n_reg); state[1] <- 1
for (t in 2:n_reg) {
  state[t] <- if (runif(1) < 0.05) 3 - state[t - 1] else state[t - 1]
}
d_reg <- data.frame(y = rnorm(n_reg, c(-1.5, 1.5)[state], 0.6))

f_reg <- statmod(y ~ regime(k = 2) - 1, gaussian1_distrib(), d_reg)
round(f_reg@structural[[1]]$parameter, 4)
```

The levels are parametrized as a first level and a positive gap, so they come
out ordered, and that is a requirement rather than a convenience: $K$
exchangeable regimes give $K!$ identical maxima, and an unordered
parametrization would let a search wander between them. The chain's own
transition probabilities are carried on the simplex charts of
@sec-parameter-simplex, one per row.

### Break-points integrated out {#sec-statmod-marginal-bp}

A break-point that varies by group can be given a distribution rather than a
coefficient, and integrated out. The obvious cost is prohibitive: with $K$
break-points and $n$ observations the positions partition the sample in
$(n+1)^K$ ways.

The structure removes the exponent in $n$. The latent positions are
independent, and the indicator $\mathbf{1}(x \ge \psi_k)$ is monotone in the
sorted covariate, so the process of *which break-points are already passed*
is itself a hidden Markov chain on the $2^K$ side patterns, whose transitions
factorize over the coordinates. The exact forward recursion then costs
$nK2^K$ rather than $(n+1)^K$, and everything the term reports — the
likelihood, the Jacobian, the posteriors and the observed Hessian —
propagates through that one recursion. What remains exponential is $K$ alone,
which is bounded at eight and in practice small.

Such a term is of the mixture shape, so it reaches the fit through the same
branch `regime()` does, with the component levels read per observation rather
than per state. The prior's own parameters are estimated by plain maximum
likelihood alongside everything else: the prior is part of the likelihood
here, not a penalty, so no hyperparameter of it is declared and no criterion
of @sec-statmod-outer is put to it.
