## Reading and using a fit {#sec-statmod-reading}

A fitted model holds three things: the specification, kept whole so the model
can be read back and reapplied; the coefficients, hyperparameters and
structural parameters the fit reached; and the record of how it got there.
The accessors below read those, and each of them answers per distribution
parameter rather than over one flat vector, because the coefficients of a
mean and of a scale are different quantities on different scales.

```{r}
#| label: read-fit
set.seed(13)
n_rd <- 300
d_rd <- data.frame(x = sort(runif(n_rd, -3, 3)),
                   g = factor(rep(1:15, length.out = 300)))
d_rd$y <- rnorm(n_rd, 1 + sin(1.2 * d_rd$x) + rnorm(15, sd = 0.5)[d_rd$g],
                exp(-0.7 + 0.2 * d_rd$x))

f_rd <- statmod(y ~ s(x, k = 8) + random(~ 1 | g) | sigma ~ x,
                gaussian1_distrib(), d_rd)
lapply(coef(f_rd), head, 3)
```

### The two variances {#sec-statmod-vcov}

Writing $H$ for the information of the log-likelihood and $S$ for the second
derivative of the penalty,

$$
V_\mathrm{b} = (H+S)^{-1}, \qquad
V_\mathrm{f} = (H+S)^{-1} H (H+S)^{-1} .
$$

The first is the posterior variance under the prior the penalty is the
negative logarithm of, and it is what an interval around a penalized term
should be built from: it carries the smoothing bias as though it were
variance, which is why such intervals cover at about their nominal rate. The
second is the sampling variance of the penalized estimator at a fixed
penalty, which is smaller and covers less. With nothing penalized $S = 0$ and
the two coincide.

`vcov()` gives the first by default, on the same information the fit itself
inverted. A coefficient a kinked penalty has set to zero has no row at all:
the penalty is not twice differentiable there, so no curvature can be read,
and the entry is missing rather than zero. The coefficients such a penalty
left alone do get a variance, and it is conditional on that selection.

```{r}
#| label: intervals
head(confint(f_rd), 4)
```

Every interval is built on the scale that keeps its quantity in its own set
and mapped back, exactly as @sec-fit-delta builds one for a single
distribution, so a scale's interval cannot contain a negative number.

### Which likelihood {#sec-statmod-loglik}

`logLik()` answers with the **conditional** log-likelihood: the density at the
fitted coefficients, a penalized coefficient among them, paired with the
effective degrees of freedom of @sec-statmod-edf. A criterion built on that
pair is the conditional AIC, and it asks how well the model describes the
groups and curves actually observed.

A mixed-model package reports the **marginal** likelihood instead, with the
random effects integrated out and the count the number of estimated
parameters. That asks about the population, and its AIC is a different
number. Neither convention allows the halves to be mixed, a marginal
likelihood against an effective count or a conditional one against a
parameter count [@vaida2005].

```{r}
#| label: loglik-two
c(conditional = as.numeric(logLik(f_rd)),
  marginal    = as.numeric(logLik(f_rd, type = "marginal")))
c(edf = attr(logLik(f_rd), "df"),
  npar = attr(logLik(f_rd, type = "marginal"), "df"))
```

The marginal value is the one the outer criterion maximized, so it exists
only where a marginal criterion actually ran. Where the hyperparameters were
held or chosen along a path there is none, and asking for one signals an
error rather than returning a number that would look like one.

Measured on a random intercept over 120 groups, the two readings of one fit
are $-4148.59$ on 115.72 effective degrees of freedom and $-4372.79$ on 4,
against `lme4`'s marginal $-4371.71$ on 4: the marginal halves agree and the
conditional pair is a different quantity.

### Prediction {#sec-statmod-predict}

`predict()` answers per parameter, on the parameter scale or the predictor's,
and will give a moment of the family instead if asked for one. Each term's
blueprint is reapplied rather than rebuilt, so the knots of a smooth and the
levels of a factor are those the model was fitted with. The identity that
pins it is that predicting on rows the model saw returns the fitted values
there.

```{r}
#| label: predict-se
nd <- data.frame(x = c(-1, 0, 1),
                 g = factor(c(1, 2, 3), levels = levels(d_rd$g)))
predict(f_rd, "mu", newdata = nd, se = TRUE)
```

A standard error is the delta method over the variance above, so it inherits
which of the two was asked for. It is `NA` for an observation whose predictor
reads a coefficient that has no variance, which is the truth about such a fit
rather than a gap in the arithmetic.

A term carrying state is continued past the end of its series by handing
`predict()` time indices that were not observed. The continuation is
deterministic: a score has zero conditional mean, so past the data the
driving term sits at its expectation and the recursion runs on with the
loadings contributing only while they still reach an observed time. No
standard error is reported there, because the uncertainty of a forecast is
mostly the uncertainty of the future scores and not of the parameters, and
reporting the second under the name of the first would understate it.

### Simulating from a model {#sec-statmod-rstatmod}

`rstatmod()` draws from a model at coefficients the caller supplies, which is
what a simulation study needs. The truth comes back beside the data rather
than attached to it: a list holding the data, the coefficients, the
per-observation parameters and whatever a term with state drew.

```{r}
#| label: simulate
sim <- rstatmod(y ~ x | sigma ~ x, gaussian1_distrib(),
                data.frame(x = runif(200, -2, 2)),
                par = list(mu = c(1, 0.8), sigma = c(-0.7, 0.2)))
names(sim)
round(coef(statmod(y ~ x | sigma ~ x, gaussian1_distrib(), sim$data))$mu, 3)
```

The predictor is assembled as a fit assembles it, so a term whose block moves
with its coefficients contributes what the term computes and not its block
times them; the two differ by the whole nonlinearity.

Two choices in it decide what a study measures. Passing `data` holds the
covariates fixed, so what is measured is the estimator's behavior
*conditional* on that design; passing `covariates`, one function per column,
draws them afresh at every replicate and measures over the design as well.
Neither is more correct and a study should say which it ran. And with several
replicates the truth is drawn **once** and shared, since a study over
replicates measures the variability of an estimator at a set of parameters,
so the replicates differ in what is random rather than in what is being
estimated.

```{r}
#| label: consistency-statmod
#| include: false
assert_statmod_ok()
```
