## In practice {#sec-statmod-practice}

Two models fitted and read from end to end. The first is the one the
framework exists for, an equation on every parameter of the response; the
second carries a term whose design block moves with its own coefficients.

### A distributional model {#sec-statmod-practice-dist}

Four hundred observations in twenty groups. The mean is a smooth function of
a covariate plus a group effect; the log scale is linear in the same
covariate, which is a fact the model is not told.

```{r}
#| label: practice-data
set.seed(31)
n_p <- 400
u_grp <- rnorm(20, sd = 0.5)
dp <- data.frame(x = sort(runif(n_p, -3, 3)),
                 g = factor(rep(1:20, length.out = n_p)))
truth_mu <- 1 + sin(1.2 * dp$x) + u_grp[dp$g]
truth_sd <- exp(-0.7 + 0.25 * dp$x)
dp$y <- rnorm(n_p, truth_mu, truth_sd)

fp <- statmod(y ~ s(x, k = 8) + random(~ 1 | g) | sigma ~ s(x, k = 5),
              gaussian1_distrib(), dp)
```

The whole fit is read by one call:

```{r}
#| label: practice-summary
summary(fp)
```

Four things in that report answer questions the earlier sections raised. The
hyperparameters stand at the head of the block they govern rather than after
the coefficients they shrink, since a table opening with a hundred selected
coefficients buries the one number that produced the selection. The
log-likelihood is the conditional one and the degrees of freedom beside it
are effective, so the information criteria built from that pair are the
conditional ones. Every interval is on the scale that keeps its quantity in
its own set. And the last line is the certificate.

```{r}
#| label: practice-hyper
hyper(fp)[, c("parameter", "term", "name", "estimate", "source")]
statmod_certificate(fp)$state
```

The random effect's prior scale is estimated near the 0.5 the groups were
drawn at. The scale equation's smoothing parameter runs to the edge, for the
reason @sec-statmod-certificate gives: the truth there is a straight line, the
smooth is being pushed onto its own null space, and the criterion is flat.
That is a `boundary` fit, which is a certified state.

What the model recovers is best seen against the truth it was drawn from:

```{r}
#| label: practice-recovery
c(cor_mean  = cor(predict(fp, "mu"), truth_mu),
  cor_scale = cor(predict(fp, "sigma"), truth_sd),
  rmse_mean = sqrt(mean((predict(fp, "mu") - truth_mu)^2)))
```

### A model with a break-point {#sec-statmod-practice-seg}

The second carries a `seg()` term, whose block is the Jacobian of a
continuous broken line at the current coefficients and therefore moves as the
fit moves, in the sense of @sec-statmod-moving.

```{r}
#| label: practice-seg
set.seed(23)
n_s <- 300
xs <- sort(runif(n_s, 0, 10))
ds <- data.frame(x = xs,
                 y = 1 + 0.4 * xs + 1.6 * pmax(xs - 6, 0) + rnorm(n_s, sd = 0.5))

fs <- statmod(y ~ seg(x), gaussian1_distrib(), ds)
round(coef(fs)$mu, 4)
```

against an intercept of 1, a slope of 0.4, a change of slope of 1.6 and a
break-point at 6. The position is an ordinary coefficient here, which is what
the continuous construction buys: `seg()`'s block is a genuine Jacobian, so
the scoring step's increment is the Gauss-Newton one and $\psi$ is estimated
beside everything else rather than read off a pair of coefficients.

```{r}
#| label: practice-seg-cert
c(state = statmod_certificate(fs)$state,
  mode_error = signif(statmod_certificate(fs)$mode_error, 3))
```

The model carries no penalty, so there is no hyperparameter for an outer
gradient to be about and the certificate rests on the mode error alone, which
is the case @sec-statmod-certificate describes.
