## Finite-difference stencils {#sec-numerical-stencils}

```{r}
#| label: setup-07b
#| include: false
source("R/_setup.R")
source("R/numerical-certificates.R")
```

A finite-difference rule approximates the $m$-th derivative of $f$ at $x$ by
a weighted combination of values on a grid of offsets,
$$
f^{(m)}(x) \;\approx\; \frac{1}{h^m} \sum_{j=1}^{n} w_j\, f(x + o_j h),
$$
and the weights are determined by exactness on polynomials: substituting
$f(t) = (t - x)^p$ for $p = 0, \dots, n-1$ gives the linear system
$$
\sum_{j=1}^{n} w_j\, o_j^{\,p} \;=\; p!\,[p = m],
\qquad p = 0, \dots, n - 1,
$$
whose matrix is a Vandermonde in the offsets. `fd_weights(offsets, order)`
solves exactly this system, for any derivative order at any distinct
offsets, so central, one-sided and staggered rules are all one construction;
@fornberg1988 gives the recursive solution and the classical background.

```{r}
#| label: stencil-weights
# the five-point first-derivative weights, times 12: (1, -8, 0, 8, -1)
fd_weights(-2:2, 1L) * 12

# a third derivative of exp at 1, against the function itself
fd_derivative(exp, 1, order = 3L, accuracy = 4L) - exp(1)
```

`fd_offsets(order, accuracy)` chooses the grid: a central stencil of
$2r + 1$ points with reach $r = \max\{1, \lceil (m + a)/2 \rceil - 1\}$
evaluates a derivative of order $m$ to accuracy $O(h^a)$, and the one-sided
variants shift the same number of points to one side of $x$ for use near a
boundary. `fd_step(x, order, accuracy)` supplies the step,
$\varepsilon^{1/(m+a)} \max(1, |x|)$ with $\varepsilon$ the machine
precision, which balances the $O(h^a)$ truncation error against the
$O(\varepsilon h^{-m})$ amplification of rounding [@press2007]; when bounds
are supplied the step is clamped so the whole stencil stays inside the
domain.

What stays out of this package is the policy on how stencils may be used.
The rule of @sec-links and @sec-distributions7 -- one stencil of order
$k - m$ applied to the highest analytically available order $m$, never a
chain of first differences -- belongs to the packages that enforce it,
because it is a statement about their derivative contracts and not about
the weights. `numericals7` supplies the stencil; the consumer decides what
it may legitimately be applied to.

```{r}
#| label: consistency-07b
#| include: false
assert_stencils_ok()
```
