# Notation {#sec-notation}

```{r}
#| label: setup-a1
#| include: false
source("R/_setup.R")
```

## Links {.unnumbered}

| Symbol | Meaning |
|---|---|
| $\theta$ | a parameter, on its natural (constrained) scale |
| $\Theta$ | the open interval $\theta$ lives in, `link_bounds` |
| $g$ | the link function, $\eta = g(\theta)$ |
| $h$ | its inverse, $\theta = h(\eta)$, $h = g^{-1}$ |
| $\eta$ | the linear predictor, unconstrained |
| $g^{(k)}, h^{(k)}$ | $k$-th derivatives, in $\theta$ and $\eta$ respectively |
| $B_{m,j}$ | the partial (incomplete) Bell polynomials |

: {.striped}

## Distributions {.unnumbered}

| Symbol | Meaning |
|---|---|
| $f(y;\theta)$ | density or mass function |
| $\ell(y) = \log f(y;\theta)$ | log-density |
| $\ell^{(i)}(y) = \partial\ell/\partial\theta_i$ | score component |
| $\ell^{(ij)}(y) = \partial^{2}\ell/\partial\theta_i\partial\theta_j$ | observed second derivative |
| $\mathcal{I}(\theta) = -\mathbb{E}[\ell^{(ij)}]$ | Fisher information |
| $\Pi_k$ | the set partitions of a $k$-element index set |
| $\mathbb{E}_f[\cdot]$ | expectation under the parent distribution |

: {.striped}

Note the sign convention: the package returns $\ell^{(ij)}$ and
$\mathbb{E}[\ell^{(ij)}]$, the **derivatives**. The informations are the
negatives of these.

## Transformations {.unnumbered}

| Symbol | Meaning |
|---|---|
| $f_0 = f(0;\theta)$ | the parent's mass at zero |
| $\zeta$ | zero-inflation probability, parameter `zi` |
| $\pi$ | zero-adjustment probability, parameter `za` |
| $L_0 = \zeta + (1-\zeta)f_0$ | total mass at zero of a zero-inflated model |
| $w_0 = (1-\zeta)f_0/L_0$ | probability that an observed zero is a sampling zero |
| $C_{ij}$ | the hurdle's truncation correction (@prp-hurdle-hessian) |
| $g$, $g^{-1}$, $\lvert J\rvert$ | a transformer, its inverse and the absolute Jacobian $\lvert\mathrm{d}g^{-1}/\mathrm{d}y\rvert$ |
| $\nu = \delta_0 + \lambda$ | the dominating measure of a mixed distribution |
| $L$, $U$ | the truncation points, both included in the support |
| $Z(\theta)$ | the retained mass of a truncated distribution |
| $m_i,\ M_{ij}$ | $\mathbb{E}_T[\ell^{(i)}]$ and $\mathbb{E}_T[\ell^{(ij)}+\ell^{(i)}\ell^{(j)}]$ under truncation |

: {.striped}

## Optimisation {.unnumbered}

| Symbol | Meaning |
|---|---|
| $x$, $d$, $s$ | the point, the direction, and the step $s d$ taken along it |
| $g = \nabla f(x)$ | the gradient of the objective; see the note below |
| $c_1$, $c_2$ | the Armijo and Wolfe constants, $0 < c_1 < c_2 < 1$ |
| $\theta_k$ | the angle between $d_k$ and $-g_k$, in @thm-zoutendijk only |
| $B$, $M$ | a positive definite curvature model, and its role in $d = -M^{-1}g$ |
| $H$ | an approximation to the **inverse** Hessian; $\nabla^{2} f$ is written out |
| $s$, $y$ | the secant pair, $x_{+} - x$ and $\nabla f(x_{+}) - \nabla f(x)$ |
| $\rho = 1/(y^\top s)$ | the BFGS scalar |
| $\partial f(x)$ | the subdifferential; $v$ denotes one of its elements |
| $v_j$, $\alpha_j$ | a collected subgradient and its linearisation error |
| $\bar v$, $\bar\alpha$ | their aggregates, $V\lambda^{\star}$ and $\alpha^\top\lambda^{\star}$ |
| $\hat x$ | the stability centre of the bundle method |
| $t$ | the proximity weight, a step length on the parameter scale |
| $\Delta_k$ | the unit simplex in $\mathbb{R}^{k}$ |
| $L_j$, $U_j$ | the endpoints of the box constraining $\theta_j$ |
| $m_t$, $v_t$ | Adam's first and second moment estimates |

: {.striped}

## Bases {.unnumbered}

| Symbol | Meaning |
|---|---|
| $[a, b]$ | the interval a basis is defined on |
| $K$ | the number of basis functions; $B_1, \ldots, B_K$ |
| $r$, $s$ | indices of basis functions |
| $B(x)$, $B$ | the vector of values at one point, and the $n \times K$ design matrix |
| $\beta$ | the coefficients of an expansion, $f = B^\top\beta$ |
| $d$ | a derivative order; $(d_1, \ldots, d_D)$ a multi-index |
| $G_d$, $\widehat{G}_d$ | the Gram matrix @eq-gram and its empirical counterpart @eq-gram-empirical |
| $m$, $t_i$ | the degree of a spline and its knots |
| $q$ | the intermediate degree in the recurrence @eq-cox-de-boor |
| $\omega$, $z$, $j$ | the Fourier period, phase and frequency |
| $\tau$ | the Legendre variable, $\tau = 2(x-a)/(b-a) - 1$ |
| $\delta_j$, $w_j$, $h$ | a finite-difference offset, its weight and the step |
| $T$ | a linear transform of a basis, $\widetilde{B}^\top = B^\top T$ |
| $C$, $V_0$ | a linear constraint on the coefficients and a basis of its null space |
| $P$, $\lambda_k$ | a penalty matrix and the Demmler-Reinsch eigenvalues |
| $Z = BT$ | the design matrix of a transformed basis |
| $D$, $K_j$, $B_j$ | the number of variables of a product basis, and the marginals |
| $\ast$ | the row-wise Kronecker product @eq-khatri-rao |
| $\Gamma_j$, $\gamma_{j,f}$, $F$ | the factor matrices of a polyadic form, their columns and the rank |

: {.striped}

Five symbols are reused across chapters, without ambiguity in
context, and it is better to say so than to invent notation nobody uses.

$g$ is the link function in @sec-links and the gradient here; $h$ is the inverse
link throughout, and in @sec-box it is exactly that, since the map removing a box
constraint *is* a bounded link. The two meanings of $g$ never appear in the same
section: where @sec-box needs both, the gradient is written $\nabla f$ in full.

$\theta$ is a parameter everywhere except @thm-zoutendijk, where $\theta_k$ is
the angle appearing in the theorem's classical statement.

$\pi$ is the zero-adjustment probability of @sec-hurdle and nothing else.
The bundle method's aggregate subgradient, which much of the literature calls
$p$ or $\pi$, is written $\bar v$ here. That choice is consistent with $v_j$
for the subgradients it aggregates, and it avoids the collision.

$B$ is a curvature model in @sec-optimizers7 and a design matrix in
@sec-basis7. The two never appear together, and both are the letter their own
literature uses. For the same reason the interval of a basis is written
$[a, b]$ rather than with the $\ell$ that would collide with the
log-likelihood.

$m$ is the degree of a spline in @sec-basis7, while it is a truncated
expectation in @sec-truncation and a moment estimate in @sec-adam. Each
carries the index its own section gives it.

## Session {.unnumbered}

```{r}
#| label: session
#| echo: false
si <- sessionInfo()
cat(si$R.version$version.string, "\n")
cat("Platform: ", si$platform, "\n", sep = "")
pkgs <- c("linkfunctions7", "distributions7", "optimizers7", "basis7",
          "S7", "numDeriv", "knitr")
for (p in pkgs) {
  cat(sprintf("%-16s %s\n", p, as.character(utils::packageVersion(p))))
}
```
