# Notation {#sec-notation}

```{r}
#| label: setup-a1
#| include: false
source("R/_setup.R")
```

## Links {.unnumbered}

| Symbol | Meaning |
|---|---|
| $\theta$ | a parameter, on its natural (constrained) scale |
| $\Theta$ | the open interval $\theta$ lives in, `link_bounds` |
| $g$ | the link function, $\eta = g(\theta)$ |
| $h$ | its inverse, $\theta = h(\eta)$, $h = g^{-1}$ |
| $\eta$ | the linear predictor, unconstrained |
| $g^{(k)}, h^{(k)}$ | $k$-th derivatives, in $\theta$ and $\eta$ respectively |
| $B_{m,j}$ | the partial (incomplete) Bell polynomials |

: {.striped}

## Distributions {.unnumbered}

| Symbol | Meaning |
|---|---|
| $f(y;\theta)$ | density or mass function |
| $\ell(y) = \log f(y;\theta)$ | log-density |
| $s_i(y) = \partial\ell/\partial\theta_i$ | score component |
| $H_{ij}(y) = \partial^{2}\ell/\partial\theta_i\partial\theta_j$ | observed second derivative |
| $\mathcal{I}(\theta) = -\mathbb{E}[H]$ | Fisher information |
| $\Pi_k$ | the set partitions of a $k$-element index set |
| $\mathbb{E}_f[\cdot]$ | expectation under the parent distribution |

: {.striped}

Note the sign convention: the package returns $H$ and $\mathbb{E}[H]$, the
**derivatives**. The informations are the negatives of these.

## Transformations {.unnumbered}

| Symbol | Meaning |
|---|---|
| $f_0 = f(0;\theta)$ | the parent's mass at zero |
| $\zeta$ | zero-inflation probability, parameter `zi` |
| $\pi$ | zero-adjustment probability, parameter `za` |
| $L_0 = \zeta + (1-\zeta)f_0$ | total mass at zero of a zero-inflated model |
| $w_0 = (1-\zeta)f_0/L_0$ | probability that an observed zero is a sampling zero |
| $C_{ij}$ | the hurdle's truncation correction (@prp-hurdle-hessian) |
| $g$, $g^{-1}$, $|J|$ | a transformer, its inverse and the absolute Jacobian $|\mathrm{d}g^{-1}/\mathrm{d}y|$ |
| $\nu = \delta_0 + \lambda$ | the dominating measure of a mixed distribution |
| $Z(\theta)$ | the retained mass of a truncated distribution |
| $m_i,\ M_{ij}$ | $\mathbb{E}_T[s_i]$ and $\mathbb{E}_T[H_{ij}+s_is_j]$ under truncation |

: {.striped}

## Session {.unnumbered}

```{r}
#| label: session
#| echo: false
si <- sessionInfo()
cat(si$R.version$version.string, "\n")
cat("Platform: ", si$platform, "\n", sep = "")
for (p in c("linkfunctions7", "distributions7", "S7", "numDeriv", "knitr")) {
  cat(sprintf("%-16s %s\n", p, as.character(utils::packageVersion(p))))
}
```
